### 2014-07-09 Paper_v1.tex
My first publication commit

\documentclass[12pt]{article}
% Load packages
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
\usepackage{setspace}
%\usepackage{longtable}
\usepackage{colortbl}
%\usepackage{palatino,lettrine}
%\usepackage{times}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage[wide]{sidecap}
%\usepackage[authoryear,round,comma,sort&compress]{natbib}
\usepackage[square,sort,comma,numbers]{natbib}
%\usepackage[authoryear,round]{natbib}
\usepackage{supertabular}
\usepackage{simplemargins}
\usepackage{comment}
\usepackage{lineno}

\urlstyle{rm}

%\textwidth = 6.50 in
%\textheight = 9.5 in
%\oddsidemargin =  0.0 in
%\evensidemargin = 0.0 in
%\topmargin = -0.50 in
%\headheight = 0.0 in
%\headsep = 0.25 in
%\parskip = 0.15in
%\linespread{1.75}
\doublespace

%\usepackage{geometry}
\usepackage{fullpage}

%\bibliographystyle{plain}
\bibliographystyle{msb}

\makeatletter
\renewcommand\subsection{\@startsection
	{subsection}{2}{0mm}
	{-0.05in}
	{-0.5\baselineskip}
	{\normalfont\normalsize\bfseries}}
\renewcommand\subsubsection{\@startsection
	{subsubsection}{2}{0mm}
	{-0.05in}
	{-0.5\baselineskip}
	{\normalfont\normalsize\itshape}}
\renewcommand\section{\@startsection
	{subsection}{2}{0mm}
	{-0.2in}
	{0.05\baselineskip}
	{\normalfont\large\bfseries}}	
\renewcommand\paragraph{\@startsection
	{paragraph}{2}{0mm}
	{-0.05in}
	{-0.5\baselineskip}
	{\normalfont\normalsize\itshape}}
\makeatother

%Review style settings
%\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}

%Publication style settings

% Single space'd bib -
\setlength\bibsep{0pt}

\renewcommand{\rmdefault}{phv}\renewcommand{\sfdefault}{phv}

% Change the number format in the ref list -
\renewcommand{\bibnumfmt}[1]{#1.}

% Change Figure to Fig.
\renewcommand{\figurename}{Fig.}

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Dynamic Modeling of Cell Free Metabolic Networks using Effective Kinetic Models}}
\vspace{0.05in}
{\par \centering \large{ Joseph A. Wayman and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering \large{School of Chemical and Biomolecular Engineering}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Modeling cell free metabolism}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Processes}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
{\par \centering Email: jdv27@cornell.edu} 
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}

{\noindent \textbf{Keywords:}~Cell free metabolism, Mathematical modeling}

\pagebreak

\setcounter{page}{1}

\linenumbers

\section*{Introduction}
Whole-cell bacterial processes are widely used in biotechnology to produce an array of products including high-value protein therapeutics.
However, whole-cell processes share the central limitation of requiring cell growth, 
which redirects resources away from product synthesis, and cell walls, which complicate interrogation and control of intracellular metabolic processes.
On the other hand, cell-free metabolic systems offer many advantages over traditional in vivo production methods. 
For example, cell-free systems can direct scarce metabolic resources exclusively towards a single product of interest. 
Moreover, with no cell wall, cell free systems can more easily be interrogated, and substrates of the metabolite processes directly controlled.
Cell free production offers the unique opportunity to study metabolism without the complication of cell growth and gene expression processes.
For modeling, this implies that we need only consider allosteric regulation of enzyme activity when building and testing cell free metabolic models.
Of course, modeling allosteric mechanisms is itself a difficult problem when the model is at a whole genome scale. To address this problem, we have developed a
an approach based upon the constrained fuzzy logic framework of Morris and Lauffenburger [REFHERE].

In this study, we present an effective cell free metabolic modeling framework, and test this framework using two proof of concept metabolic models. 
[FINISH].

\clearpage

\section*{Results}
We developed two proof of concept metabolic networks to investigate the features of our effective cell free modeling approach (Fig. \ref{fig-networks}).
In both examples, substrate $S$ was converted to the end-products $P_{1}$ and $P_{2}$ through a series of enzymatically catalyzed reactions, 
including a branch point at hypothetical metabolite $M_{2}$. 
Several of these reactions involved cofactor dependence ($AH$ or $A$), and various allosteric regulation mechanisms. 
Network A included feedback inhibition of the initial pathway enzyme ($E_{1}$) by pathway end products $P_{1}$ and $P_{2}$ (Fig. \ref{fig-networks}A).
On the other hand, network B involved feedback inhibition of $E_{1}$ by $P_{2}$ and $E_{6}$ by $P_{1}$ (Fig. \ref{fig-networks}B). 
In both networks, branch point enzymes $E_{3}$ and $E_{6}$ were subject to feed-forward activation by cofactor $AH$.
Lastly, in both cases pathway enzyme activity was assumed to decay according to a first-order rate law. 
Allosteric regulation of enzyme activity was represented using a novel rule-based strategy, similar in spirit to the 
Constrained Fuzzy Logic (cFL) approach of Lauffenberger and coworkers [REFHERE]. 
In this formulation, Hill-like transfer functions were used to calculate the influence of metabolite abundance upon target enzyme activity. 
When an enzyme was potentially sensitive to more than one regulatory influence, logical rules were used to select which transfer function regulated enzyme 
activity at any given time (Fig. \ref{fig-control-schematic}).      
Thus, our test networks involved important features such as cofactor recycling, enzyme activity and metabolite dynamics, 
as well as multiple overlapping allosteric regulatory mechanisms.  
As such, developing our effective modeling approach using these problems gave us valuable insight into how to develop larger network models, 
without the complication of network size.

The rule based regulatory strategy captured classic patterns of allosteric activation and inhibition of enzyme activity (Fig. \ref{fig-kinetics-simulations}).

End product yield was controlled by feedback inhibition, while selectivity was controlled by branch point enzyme inhibition (Fig. \ref{fig-onoff-simulations}).
A critical test of our modeling approach was to simulate networks with known behavior. If we cannot reproduce simple network expected behavior, then our
modeling strategy, and particularly the rule-based approximation of allosteric regulation, will not be feasible for large scale problems.
We considered two cases,  control on/off, for each network configuration. 
Each of these cases had identical kinetic parameters and initial conditions; 
the \textit{only} differences between the cases was the allosteric regulation rules, and the control parameters associated with these rules. 
As expected, end product accumulation was larger for network A when the control was off (no feedback inhibition of $E_{1}$ by $P_{1}$ and $P_{2}$),
as compared to the on case (Fig. \ref{fig-onoff-simulations}A). We found this behavior was robust to the choice of underlying kinetic parameters, 
as we observed that same qualitative response across an ensemble of randomized parameter sets (N = 100). 
The control on/off response of network B was more subtle. In the off case, the behavior was qualitatively similar to network A. 
However, for the on case, flux was diverted away from $P_{2}$ formation by feedback inhibition of $E_{6}$ activity at the $M_{2}$ branch point by $P_{1}$ (Fig. \ref{fig-onoff-simulations}B).
Lower $E_{6}$ activity at the $M_{2}$ branch point allowed more flux toward $P_{1}$ formation, hence the yield of $P_{1}$ also increased (Fig. \ref{fig-onoff-simulations}C).
Again, the control on/off behavior was robust to the values of the kinetic parameters, as the same qualitative trend was conserved across 
an ensemble of possible randomized kinetic parameters (N = 100). Taken together, these simulations suggested that the rule based allosteric control 
concept could robustly capture expected feedback behavior.

The estimation of kinetic parameters was sensitive to the choice of control structure (Fig. \ref{fig-parameter-fit}).
A critical challenge for any dynamic model effort is the estimation of kinetic parameters. 
However, beyond this challenge is the identification of model structure, i.e., the biological connectivity of the system you are exploring.
Of course these two issues are not independent as any description of the control of enzyme activity will be a function of the system state, which
in turn is a function of the kinetic parameters.

Discrimination between competing model formulations can be achieved by optimal experimental design (Fig. ZZ). [FINISH]

\clearpage

\section*{Discussion}
In this study, we proposed a dynamic modeling strategy to simulate cell free metabolic networks. We demonstrated this strategy using two proof of concept metabolic networks 
with the same enzymatic connectivity, but differing control structures. [FINISH]

Cybernetic models, other dynamic models of metabolism.

While the results of this study were encouraging, there are several critical next steps that must be accomplished before we can model genome scale cell free metabolic networks.
[FINISH]

\section*{Materials and Methods}

\subsection*{Formulation and solution of the model equations.}
We used ordinary differential equations (ODEs) to model the time evolution of metabolite ($x_{i}$) and scaled enzyme ($\epsilon_{i}$) abundance in hypothetical cell free
metabolic networks:
\begin{eqnarray}
	\frac{dx_{i}}{dt} & = & \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)\qquad{i=1,2,\hdots,\mathcal{M}}\\
	\frac{d\epsilon_{i}}{dt} & = & -\lambda_{i}\epsilon_{i}\qquad{i = 1,2,\hdots,\mathcal{E}}
\end{eqnarray}where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of metabolites and $\mathcal{E}$ denotes the number of
enzymes in the model. The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ 
denotes the rate of reaction $j$. Typically, reaction $j$ is a non-linear function of metabolite and enzyme abundance, as well as unknown kinetic parameters 
$\mathbf{k}$ ($\mathcal{K}\times{1}$).
The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$. If 
$\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$. Conversely, if $\sigma_{ij}>0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates
metabolite $i$ is not connected with reaction $j$. 
Each reaction rate was written as the product of a reaction term ($\bar{r}_{j}$) and a regulatory term ($v_{j}$):
\begin{equation}
	r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right) = \bar{r}_{j}v_{j}
\end{equation}We used multiple saturation kinetics to model the reaction term $\bar{r}_{j}$:
\begin{equation}\label{eqn:rate-bar}
	\bar{r}_{j} = k_{j}^{max}\epsilon_{i}\left(\prod_{s\in{m_{j}^{-}}}\frac{x_{s}}{K_{js} + x_{s}}\right)
\end{equation}where $k_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the scaled enzyme activity which catalyzes reaction $j$, and
$K_{js}$ denotes the saturation constant for species $s$ in reaction $j$. 
The product in Eqn. \eqref{eqn:rate-bar} was carried out over the set of \textit{reactants} for reaction $j$ (denoted as $m_{j}^{-}$). 
The allosteric regulation term $v_{j}$ depended upon the combination of factors which influenced the activity of enzyme $i$.
For each enzyme, we used a rule based approach to select from competing control factors (Fig. \ref{fig-control-schematic}). 
If an enzyme was activated by $m$ metabolites, we modeled this activation as:
\begin{equation}
	v_{j} = \max\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right)
\end{equation}Conversely, if enzyme activity was inhibited by a $m$ metabolites, we modeling this inhibition as:
\begin{equation}
	v_{j} = 1 - \max\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right)
\end{equation}Lastly, if an enzyme had both $m$ activating and $n$ inhibitory factors, we modeled the regulatory term as:
\begin{equation}
	v_{j} = \min\left(u_{j},d_{j}\right)
\end{equation}where:
\begin{eqnarray}
	u_{j} &=& \max_{j^{+}}\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right) \\
	d_{j} &=& 1 - \max_{j^{-}}\left(f_{1j}\left(x\right),\hdots,f_{nj}\left(x\right)\right)
\end{eqnarray}where $j^{+}$ and $j^{-}$ denote the sets of activating, and inhibitory factors for enzyme $j$. 
If an enzyme had no allosteric factors, we set $v_{j} = 1$.
In this study, each individual factor had the form:
\begin{equation}
	f_{i}\left(\mathbf{x}\right) = \frac{\kappa_{ij}^{\eta}x_{j}^{\eta}}{1 + \kappa_{ij}^{\eta}x_{j}^{\eta}}
\end{equation}where $x_{j}$ denotes the abundance of metabolite $j$, and $\kappa_{ij}$ and $\eta$ are control parameters. 
The model equations were encoded using the Octave programming language, and solved using the LSODE routine in Octave [REFHERE].

\subsection*{Estimation of model parameters from synthetic experimental data}
Model parameters were estimated by minimizing the difference between simulations and synthetic experimental data using particle swarm optimization [REFHERE].
In this study, the parameter estimation problem took the form:
\begin{equation}\label{eqn:objective-function}
	\min_{\mathbf{k}} \sum_{\tau=1}^{\mathcal{T}}\sum_{j=1}^{\mathcal{S}}\left(\frac{\hat{x}_{j}\left(\tau\right) - x_{j}\left(\tau,\mathbf{k}\right)}{\omega_{j}\left(\tau\right)}\right)^{2}
\end{equation}where $\hat{x}_{j}\left(\tau\right)$ denotes the measured value of species $j$ at time $\tau$, $x_{j}\left(\tau,\mathbf{k}\right)$ denotes the simulated 
value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$. The outer summation is respect to
time, while the inner summation is with respect to state. We wanted to approximate a realistic model identification scenario, thus we assumed noisy experimental data, 
limited sampling resolution (approximately 20 minutes per sample) and that only a subset of metabolites were measurable.

We minimized Eqn. \eqref{eqn:objective-function} using Particle swarm optimization (PSO).
PSO is global optimization procedure which uses a swarming metaheuristic to explore high-dimensional parameter spaces. 
A strength of PSO is its ability to find the global minimum, even in the presence of potentially many local minima, by communicating the local
error landscape experienced by each particle collectively to the swarm using the update rules:
\begin{eqnarray}
	\mathbf{\Delta}_{i} &=&\theta_{1}\mathbf{\Delta}_{i} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{k}_{i}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G} - \mathbf{k}_{i}\right) \\
	\mathbf{k}_{i} &=& \mathbf{k}_{i} + \mathbf{\Delta}_{i}
\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right) = (1.0,0.05564,0.02886)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes local best solution found by particle $i$, and
$\mathcal{G}$ denotes the best solution found over the entire population of particles. The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
parameters ($\mathcal{K}\times{1}$). The PSO algorithm, and the objective function were encoded and solved in the Octave programming language [REFHERE].

\subsection{Model discrimination}

\section*{Acknowledgements}
This study was supported by the National Science Foundation GK12 award (DGE-1045513) 
and by the National Science Foundation CAREER award (FILLMEIN).

\clearpage
%\bibliographystyle{plain}
%\bibliographystyle{IEEEbib}

\bibliography{References_v1}

%\begin{comment}

\clearpage

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-1-Networks.pdf}
\caption{Proof of concept cell-free metabolic networks considered in this study. Substrate $S$ is converted to products $P_{1}$ and $P_{2}$ through a series of chemical conversions
catalyzed by enzyme(s) $E_{j}$. The activity of the pathway enzymes is subject to both positive and negative allosteric regulation. }\label{fig-networks}
\end{figure}

\clearpage

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-2-ControlSchematic.pdf}
\caption{Schematic of the rule based allosteric enzyme activity control laws. }\label{fig-control-schematic}
\end{figure}

\clearpage

\begin{figure}
\centering
\includegraphics[width=0.55\textwidth]{./figs/Figure-3-OnOffSimulations.pdf}
\caption{On/off control simulations for network A and network B for an ensemble of kinetic parameter sets. 
For each case, N = 100 simulations were conducted using kinetic and initial conditions randomly generated from a hypothetical true parameter set. 
The gray area represents $\pm$ one standard deviation surrounding the mean. Control parameters were fixed during the ensemble calculations.}\label{fig-onoff-simulations}
\end{figure}

\clearpage

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-4-ParameterFit.pdf}
\caption{Parameter estimation from synthetic data for the same and mismatched allosteric control logic. }\label{fig-parameter-fit}
\end{figure}

\clearpage

% Supplemental figures -
% Set the S- 
\renewcommand\thefigure{S\arabic{figure}}
\renewcommand\thetable{T\arabic{table}}
\renewcommand\thepage{S-\arabic{page}}
\renewcommand\theequation{S\arabic{equation}}

% Reset the counters -
\setcounter{equation}{0}
\setcounter{table}{0}
\setcounter{figure}{0}
\setcounter{page}{1}

\end{document}

### 2014-07-09 Paper_v1.tex
Small update

@@
 The rule based regulatory strategy captured classic patterns of allosteric activation and inhibition of enzyme activity (Fig. \ref{fig-kinetics-simulations}).  
+The first case we explored was substrate activation of enzyme activity (positive cooperativity). 
+Classically, we expected a sigmoidal relationship substrate abundance and reaction rate for positive cooperativity.
 
 End product yield was controlled by feedback inhibition, while selectivity was controlled by branch point enzyme inhibition (Fig. \ref{fig-onoff-simulations}).
-A critical test of our modeling approach was to simulate networks with known behavior. If we cannot reproduce simple network expected behavior, then our
+A critical test of our modeling approach was to simulate networks with known behavior. If we cannot reproduce the expected behavior of simple networks, then our
 modeling strategy, and particularly the rule-based approximation of allosteric regulation, will not be feasible for large scale problems.

### 2014-07-09 Paper_v1.tex
going home ...

@@
 
-The rule based regulatory strategy captured classic patterns of allosteric activation and inhibition of enzyme activity (Fig. \ref{fig-kinetics-simulations}).  
-The first case we explored was substrate activation of enzyme activity (positive cooperativity). 
-Classically, we expected a sigmoidal relationship substrate abundance and reaction rate for positive cooperativity.
+The rule based regulatory strategy captured the kinetic behavior of classic allosteric activation and inhibition mechanisms (Fig. \ref{fig-kinetics-simulations}).  
+We first explored feed-forward substrate activation of enzyme activity (for both positive and negative cooperativity). 
+Consistent with classical data, the rule based strategy predicted a sigmoidal relationship between substrate abundance 
+and reaction rate as a function of the cooperativity parameter (Fig. \ref{fig-kinetics-simulations}A). 
+For cooperativity parameters less than unity, increased substrate abundance \textit{decreased} the reaction rate. This was consistent with the idea that
+substrate binding \textit{decreases} at regulatory sites negatively impacts the ability of the enzyme to bind substrate at the active site. 
+On the other hand, as the cooperativity parameter increased past unity, the rate of conversion of substrate $S$ to product $P$ by enzyme $E$ approached a step function.
+In the presence of an inhibitor, the rule based strategy predicted non-competitive like behavior as a function of the 
+cooperativity parameter (Fig. \ref{fig-kinetics-simulations}B). 
+When the control gain parameter, $\kappa_{ij}$ in Eqn. \eqref{eqn:control-factor}, was greater than unity, the inhibitory force was
+directly proportional to the cooperativity parameter, $\eta$ in Eqn. \eqref{eqn:control-factor}. 
+Thus, as the cooperativity parameter increased, the maximum reaction rate decreased (Fig. \ref{fig-kinetics-simulations}B, orange).
+However, when the gain parameter was less than unity, enzyme inhibition increased with \textit{decreasing} cooperativity, 
+i.e., smaller $\eta$ yielded increased inhibition (Fig. \ref{fig-kinetics-simulations}B).
+Interestingly, our rule based approach was not able to directly simulate competitive inhibition of enzyme activity. 
+Taken together, the rule based strategy captured classical regulatory patterns for both enzyme activation and inhibition.
 
@@
 In this study, each individual factor had the form:
-\begin{equation}
+\begin{equation}\label{eqn:control-factor}
 	f_{i}\left(\mathbf{x}\right) = \frac{\kappa_{ij}^{\eta}x_{j}^{\eta}}{1 + \kappa_{ij}^{\eta}x_{j}^{\eta}}
 \end{equation}where $x_{j}$ denotes the abundance of metabolite $j$, and $\kappa_{ij}$ and $\eta$ are control parameters. 
+The $\kappa_{ij}$ parameter was species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).
 The model equations were encoded using the Octave programming language, and solved using the LSODE routine in Octave [REFHERE].
@@
 \centering
-\includegraphics[width=0.55\textwidth]{./figs/Figure-3-OnOffSimulations.pdf}
-\caption{On/off control simulations for network A and network B for an ensemble of kinetic parameter sets. 
+\includegraphics[width=1.0\textwidth]{./figs/Figure-3-EnzymeKinetics.pdf}
+\caption{Kinetics of simple transformations in the presence of activation and inhibition. 
+A:The conversion of substrate $S$ to product $P$ by enzyme $E$ was activated by $S$. 
+B:The conversion of substrate $S$ to product $P$ by enzyme $E$ was inhibited by inhibitor $I$. 
+}\label{fig-kinetics-simulations}
+\end{figure}
+
+\clearpage
+
+\begin{figure}
+\centering
+\includegraphics[width=0.55\textwidth]{./figs/Figure-4-OnOffSimulations.pdf}
+\caption{On/off control simulations for network A and network B for an ensemble of kinetic parameter sets versus time. 
 For each case, N = 100 simulations were conducted using kinetic and initial conditions randomly generated from a hypothetical true parameter set. 
@@
 \centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-4-ParameterFit.pdf}
+\includegraphics[width=1.0\textwidth]{./figs/Figure-5-ParameterFit.pdf}
 \caption{Parameter estimation from synthetic data for the same and mismatched allosteric control logic. }\label{fig-parameter-fit}

### 2014-07-09 Paper_v1.tex
going home ...

@@
 A critical test of our modeling approach was to simulate networks with known behavior. If we cannot reproduce the expected behavior of simple networks, then our
-modeling strategy, and particularly the rule-based approximation of allosteric regulation, will not be feasible for large scale problems.
+effect modeling strategy, and particularly the rule-based approximation of allosteric regulation, will not be feasible for large scale problems.
 We considered two cases,  control on/off, for each network configuration. 

### 2014-07-09 Paper_v1.tex
going home ...

@@
 
-In this study, we present an effective cell free metabolic modeling framework, and test this framework using two proof of concept metabolic models. 
+In this study, we present an effective cell free metabolic modeling framework, and test this framework using two proof of concept metabolic networks. 
 [FINISH].
@@
 \section*{Discussion}
-In this study, we proposed a dynamic modeling strategy to simulate cell free metabolic networks. We demonstrated this strategy using two proof of concept metabolic networks 
+In this study, we proposed a effective modeling strategy to simulate cell free metabolic networks. 
+We tested this strategy using two proof of concept metabolic networks 
 with the same enzymatic connectivity, but differing control structures. [FINISH]

### 2014-07-10 Paper_v1.tex
going home ...
Added references

@@
 %\bibliographystyle{plain}
-\bibliographystyle{msb}
+\bibliographystyle{plos2009}
 
@@
 Allosteric regulation of enzyme activity was represented using a novel rule-based strategy, similar in spirit to the 
-Constrained Fuzzy Logic (cFL) approach of Lauffenberger and coworkers [REFHERE]. 
+Constrained Fuzzy Logic (cFL) approach of Lauffenberger and coworkers \citep{Morris:2011aa}. 
 In this formulation, Hill-like transfer functions were used to calculate the influence of metabolite abundance upon target enzyme activity. 
@@
 The $\kappa_{ij}$ parameter was species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).
-The model equations were encoded using the Octave programming language, and solved using the LSODE routine in Octave [REFHERE].
+The model equations were encoded using the Octave programming language, and solved using the LSODE routine in Octave \citep{Octave:2014}.
 
 \subsection*{Estimation of model parameters from synthetic experimental data}
-Model parameters were estimated by minimizing the difference between simulations and synthetic experimental data using particle swarm optimization [REFHERE].
+Model parameters were estimated by minimizing the difference between simulations and synthetic experimental data using particle swarm optimization \citep{PSO}.
 In this study, the parameter estimation problem took the form:
@@
 We minimized Eqn. \eqref{eqn:objective-function} using Particle swarm optimization (PSO).
-PSO is global optimization procedure which uses a swarming metaheuristic to explore high-dimensional parameter spaces. 
+PSO is global optimization procedure which uses a swarming metaheuristic to explore parameter spaces. 
 A strength of PSO is its ability to find the global minimum, even in the presence of potentially many local minima, by communicating the local
@@
 The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
-parameters ($\mathcal{K}\times{1}$). The PSO algorithm, and the objective function were encoded and solved in the Octave programming language [REFHERE].
+parameters ($\mathcal{K}\times{1}$). The PSO algorithm, and the objective function were encoded and solved in the Octave programming language \citep{Octave:2014}.
 
-\subsection{Model discrimination}
+\subsection*{Model discrimination}
 

### 2014-07-10 Paper_v1.tex
Updated enzyme kinetics section

@@
 In both networks, branch point enzymes $E_{3}$ and $E_{6}$ were subject to feed-forward activation by cofactor $AH$.
-Lastly, in both cases pathway enzyme activity was assumed to decay according to a first-order rate law. 
+Lastly, enzyme activity was assumed to decay according to a first-order rate law in both cases. 
 Allosteric regulation of enzyme activity was represented using a novel rule-based strategy, similar in spirit to the 
@@
 as well as multiple overlapping allosteric regulatory mechanisms.  
-As such, developing our effective modeling approach using these problems gave us valuable insight into how to develop larger network models, 
+As such, developing our effective modeling approach using these simple problems gave us valuable insight into the development of larger network models, 
 without the complication of network size.
 
-The rule based regulatory strategy captured the kinetic behavior of classic allosteric activation and inhibition mechanisms (Fig. \ref{fig-kinetics-simulations}).  
+The rule based regulatory strategy approximated the behavior of classical allosteric activation and inhibition mechanisms (Fig. \ref{fig-kinetics-simulations}).  
 We first explored feed-forward substrate activation of enzyme activity (for both positive and negative cooperativity). 
@@
 i.e., smaller $\eta$ yielded increased inhibition (Fig. \ref{fig-kinetics-simulations}B).
-Interestingly, our rule based approach was not able to directly simulate competitive inhibition of enzyme activity. 
+Interestingly, our rule based approach was unable to directly simulate competitive inhibition of enzyme activity. 
+For competitive inhibitors, the kinetic component of our rate, $\bar{r}_{j}$ in \eqref{eqn:rate-factor}, could be modified to account 
+for the inhibition (data not shown). 
 Taken together, the rule based strategy captured classical regulatory patterns for both enzyme activation and inhibition. 
+Thus, we are able to model complex kinetic phenomena such as ultrasensitivity, despite an effective description of reaction kinetics.
 
@@
 
-The estimation of kinetic parameters was sensitive to the choice of control structure (Fig. \ref{fig-parameter-fit}).
-A critical challenge for any dynamic model effort is the estimation of kinetic parameters. 
-However, beyond this challenge is the identification of model structure, i.e., the biological connectivity of the system you are exploring.
-Of course these two issues are not independent as any description of the control of enzyme activity will be a function of the system state, which
-in turn is a function of the kinetic parameters.
+The estimation of kinetic parameters was sensitive to the choice of regulatory structure (Fig. \ref{fig-parameter-fit}).
+A critical challenge for any dynamic model is the estimation of kinetic parameters. 
+For metabolic processes, there is also the added challenge of identifying the regulation and control structures that manage metabolism.  
+Of course, these issues are not independent; any description of enzyme activity regulation will be a function of system state, which
+in turn depends upon the kinetic parameters. For cell free systems, regulated gene expression has been removed, however, enzyme activity regulation is still operational. 
+We explored this linkage by estimating model parameters from synthetic data using both network
+structures. We generated noise-corrupted synthetic measurements of the substrate $S$, intermediate $M_{5}$ and end-product $P_1$ approximately every 20 min using
+network A. We then generated an ensemble of model parameter estimates by minimizing the difference between model simulations and the synthetic data using particle swarm
+optimization, starting from random initial guesses. 
+PSO identified an ensemble of parameters that bracketed the mean of the synthetic measurements in less than 1000 iterations when the control structure was correct (Fig.
+\ref{fig-parameter-fit}A and B).
+However, when there was model mismatch (network B fit against network A data), PSO was not able to identify an ensemble, all else being the same (Fig. \ref{fig-parameter-fit}C and D).
 
-Discrimination between competing model formulations can be achieved by optimal experimental design (Fig. ZZ). [FINISH]
+Discrimination between competing regulatory formulations can be achieved by optimal experimental design (Fig. ZZ). [FINISH]
 
@@
 \section*{Discussion}
-In this study, we proposed a effective modeling strategy to simulate cell free metabolic networks. 
-We tested this strategy using two proof of concept metabolic networks 
-with the same enzymatic connectivity, but differing control structures. [FINISH]
+In this study, we proposed a effective modeling strategy to dynamically simulate cell free metabolic networks. 
+We tested this strategy using two proof of concept metabolic networks.
+In both networks, substrate $S$ was converted to the end-products $P_{1}$ and $P_{2}$ through a series of enzymatically catalyzed reactions, 
+including a branch point at hypothetical metabolite $M_{2}$. While both networks had the same enzymatic connectivity, that had differing control structures.
+[FINISH]
 
@@
 Each reaction rate was written as the product of a reaction term ($\bar{r}_{j}$) and a regulatory term ($v_{j}$):
-\begin{equation}
+\begin{equation}\label{eqn:rate-factor}
 	r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right) = \bar{r}_{j}v_{j}

### 2014-07-10 Paper_v1.tex
Updated enzyme kinetics section

@@
 \subsection*{Estimation of model parameters from synthetic experimental data}
-Model parameters were estimated by minimizing the difference between simulations and synthetic experimental data using particle swarm optimization \citep{PSO}.
-In this study, the parameter estimation problem took the form:
+Model parameters were estimated by minimizing the difference between simulations and synthetic experimental data:
 \begin{equation}\label{eqn:objective-function}
@@
 
-We minimized Eqn. \eqref{eqn:objective-function} using Particle swarm optimization (PSO).
+We minimized Eqn. \eqref{eqn:objective-function} using Particle swarm optimization (PSO) \citep{PSO}.
 PSO is global optimization procedure which uses a swarming metaheuristic to explore parameter spaces. 

### 2014-07-10 Paper_v1.tex
Updated enzyme kinetics section

@@
 time, while the inner summation is with respect to state. We wanted to approximate a realistic model identification scenario, thus we assumed noisy experimental data, 
-limited sampling resolution (approximately 20 minutes per sample) and that only a subset of metabolites were measurable.
+limited sampling resolution (approximately 20 minutes per sample) and a limited number of measurable metabolites.
 

### 2014-07-10 Paper_v1.tex
Updated materials and methods

@@
 \subsection*{Formulation and solution of the model equations.}
-We used ordinary differential equations (ODEs) to model the time evolution of metabolite ($x_{i}$) and scaled enzyme ($\epsilon_{i}$) abundance in hypothetical cell free
+We used ordinary differential equations (ODEs) to model the time evolution of metabolite ($x_{i}$) and scaled enzyme abundance ($\epsilon_{i}$) in hypothetical cell free
 metabolic networks:
@@
 $\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$. Conversely, if $\sigma_{ij}>0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates
-metabolite $i$ is not connected with reaction $j$. 
-Each reaction rate was written as the product of a reaction term ($\bar{r}_{j}$) and a regulatory term ($v_{j}$):
+metabolite $i$ is not connected with reaction $j$. Lastly, $\lambda_{i}$ denotes the scaled enzyme degradation constant. The system material balances are subject to the
+initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$ and $\mathbf{\epsilon}\left(t_{o}\right) = \mathbf{1}$ (initially we have 100\% cell-free enzyme abundance).
+
+Each reaction rate was written as the product of two terms, a kinetic term ($\bar{r}_{j}$) and a regulatory term ($v_{j}$):
 \begin{equation}\label{eqn:rate-factor}

### 2014-07-11 Paper_v1.tex
New sections

@@
 \section*{Results}
+
+\subsection*{Formulation and properties of cell free effective models.}
 We developed two proof of concept metabolic networks to investigate the features of our effective cell free modeling approach (Fig. \ref{fig-networks}).
@@
 
-The estimation of kinetic parameters was sensitive to the choice of regulatory structure (Fig. \ref{fig-parameter-fit}).
+\subsection*{Estimating parameters and effective allosteric regulatory structures.}
 A critical challenge for any dynamic model is the estimation of kinetic parameters. 
@@
 optimization, starting from random initial guesses. 
+The estimation of kinetic parameters was sensitive to the choice of regulatory structure (Fig. \ref{fig-parameter-fit}).
 PSO identified an ensemble of parameters that bracketed the mean of the synthetic measurements in less than 1000 iterations when the control structure was correct (Fig.
 \ref{fig-parameter-fit}A and B).
-However, when there was model mismatch (network B fit against network A data), PSO was not able to identify an ensemble, all else being the same (Fig. \ref{fig-parameter-fit}C and D).
+However, when there was network mismatch (network B fit against network A synthetic data), PSO unable to identify an ensemble, all else being the same (Fig. \ref{fig-parameter-fit}C and D).
+Interestingly, the particle swarm generated a \textit{sloppy} parameter ensemble, in the absence of network mismatch (Fig. ZZZ).
+
+We modified our particle swarm identification strategy to simultaneously search over control structures and kinetic parameters (Fig. ZZZ). 
+We constructed five putative network models, each with the same enzymatic connectivity but different allosteric structures (Fig. ZZZ). 
+We then initialized a population of particles, each with one of the five putative regulatory programs, and randomized kinetic parameters.
+We biased the distribution of the particle population according to \textit{a~prior} belief of the correct regulatory program.
+To this end, we considered three different priors, a uniform distribution where each of the regulatory structures represented 20\% of the population, and
+two mixed distributions that were positively and negatively biased towards the correct structure (Table ZZ).
 

### 2014-07-11 Paper_v1.tex
New sections

@@
 We biased the distribution of the particle population according to \textit{a~prior} belief of the correct regulatory program.
-To this end, we considered three different priors, a uniform distribution where each of the regulatory structures represented 20\% of the population, and
-two mixed distributions that were positively and negatively biased towards the correct structure (Table ZZ).
+To this end, we considered three different priors, a uniform distribution where each regulatory structure represented 20\% of the population, and
+two mixed distributions that were positively or negatively biased towards the correct structure (Table ZZ).
 

### 2014-07-15 Paper_v1.tex
New sections
Update

@@
 \renewcommand{\rmdefault}{phv}\renewcommand{\sfdefault}{phv}
+\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
 
@@
 \vspace{0.05in}
-{\par \centering \large{ Joseph A. Wayman and Jeffrey D. Varner$^{*}$}}
+{\par \centering \large{ Joseph A. Wayman, Adithya Sagar and Jeffrey D. Varner$^{*}$}}
 \vspace{0.10in}
@@
 Interestingly, the particle swarm generated a \textit{sloppy} parameter ensemble, in the absence of network mismatch (Fig. ZZZ). 
+Taken together, ... [FINISH].
 
-We modified our particle swarm identification strategy to simultaneously search over control structures and kinetic parameters (Fig. ZZZ). 
-We constructed five putative network models, each with the same enzymatic connectivity but different allosteric structures (Fig. ZZZ). 
-We then initialized a population of particles, each with one of the five putative regulatory programs, and randomized kinetic parameters.
-We biased the distribution of the particle population according to \textit{a~prior} belief of the correct regulatory program.
-To this end, we considered three different priors, a uniform distribution where each regulatory structure represented 20\% of the population, and
-two mixed distributions that were positively or negatively biased towards the correct structure (Table ZZ).
-
-Discrimination between competing regulatory formulations can be achieved by optimal experimental design (Fig. ZZ). [FINISH]
+We modified our particle swarm identification strategy to simultaneously search over both kinetic parameters and putative control structures.
+In addition to our initial networks, we constructed three additional presumptive network models, each with the same enzymatic connectivity but different allosteric 
+regulation of the initial pathway enzyme (Fig. \ref{fig-alternative-networks}). 
+We then initialized a population of particles, each with one of the five potential regulatory programs, and randomized kinetic parameters.
+Thus, we generated an initial population of particles that had \textit{both} different kinetic parameters as well as different control structures. 
+We biased the distribution of the particle population according to our \textit{a~prior} belief of the correct regulatory program.
+To this end, we considered three different priors, a uniform distribution where each putative regulatory structure represented 20\% of the population, and
+two mixed distributions that were positively or negatively biased towards the correct structure (network A).
+In both the positively biased, and uniform cases the particle swarm clearly differentiated between the true or closely related structures and those that were materially different (Fig.
+\ref{fig-control-search}).
+As expected, the positively biased population (40\% of the initial particle population seeded with network A) gave the best results, where the correct structure was preferentially 
+identified (Fig. \ref{fig-control-search}A). 
+On the other hand, the uniform distribution identified a combination of network A (ZZ) and network C (YY) as the most likely control structures (Fig. \ref{fig-control-search}B).
+Network A and C differ by the regulatory connection between the end-product $P_2$ and enzyme $E_{1}$; in network A $P_{2}$ was assumed to inhibit $E_{1}$ while in network C $P_{2}$ was assumed to
+activate $E_{1}$.  Lastly, when the initial population was biased towards a completely incorrect structure (initial population seeded with 40\% network B), 
+the particle swarm misidentified the correct structure (Fig. \ref{fig-control-search}C). [FINISH]
 
@@
 [FINISH]
+
+\clearpage
 
@@
 
-\subsection*{Estimation of model parameters from synthetic experimental data}
-Model parameters were estimated by minimizing the difference between simulations and synthetic experimental data:
+\subsection*{Estimation of model parameters and structures from synthetic experimental data.}
+Model parameters were estimated by minimizing the difference between simulations and synthetic experimental data (squared residual):
 \begin{equation}\label{eqn:objective-function}
@@
 value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$. The outer summation is respect to
-time, while the inner summation is with respect to state. We wanted to approximate a realistic model identification scenario, thus we assumed noisy experimental data, 
+time, while the inner summation is with respect to state. We approximated a realistic model identification scenario, assuming noisy experimental data, 
 limited sampling resolution (approximately 20 minutes per sample) and a limited number of measurable metabolites.
 
-We minimized Eqn. \eqref{eqn:objective-function} using Particle swarm optimization (PSO) \citep{PSO}.
-PSO is global optimization procedure which uses a swarming metaheuristic to explore parameter spaces. 
+We minimized the model residual using Particle swarm optimization (PSO) \citep{PSO}.
+PSO uses a \textit{swarming} metaheuristic to explore parameter spaces. 
 A strength of PSO is its ability to find the global minimum, even in the presence of potentially many local minima, by communicating the local
-error landscape experienced by each particle collectively to the swarm using the update rules:
+error landscape experienced by each particle collectively to the swarm. Thus, PSO acts both as a local and a global search algorithm. 
+For each iteration, particles in the swarm compute their local error by evaluating the model equations using their specific parameter vector realization.
+From each of these local points, a globally best error is identified. Both the local and global error 
+are then used to update the parameter estimates of each particle using the rules:
 \begin{eqnarray}
@@
 	\mathbf{k}_{i} &=& \mathbf{k}_{i} + \mathbf{\Delta}_{i}
-\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right) = (1.0,0.05564,0.02886)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes local best solution found by particle $i$, and
+\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes local best solution found by particle $i$, and
 $\mathcal{G}$ denotes the best solution found over the entire population of particles.
 The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
-parameters ($\mathcal{K}\times{1}$). The PSO algorithm, and the objective function were encoded and solved in the Octave programming language \citep{Octave:2014}.
+parameters ($\mathcal{K}\times{1}$). In thus study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$, which was taken from XXX. The quality of parameter
+estimates was measured using two criteria, goodness of fit (model residual) and angle between the estimated parameter vector $\mathbf{k}_{j}$ and the true parameter set $\mathbf{k}^{*}$:
+\begin{equation}
+	\alpha_{j} = cos^{-1}\left(\frac{\mathbf{k}_{j}\cdot{\mathbf{k^{*}}}}{\norm{\mathbf{k}_{j}}\norm{\mathbf{k^{*}}}}\right)
+\end{equation}If the candidate parameter set $\mathbf{k}_{j}$ were perfect, the residual between the model and synthetic data 
+and the angle between $\mathbf{k}_{j}$ and the true parameter set $\mathbf{k}^{*}$ would be equal to zero.
 
-\subsection*{Model discrimination}
+We modified our PSO implementation to simultaneously search over kinetic parameters and putative model control structures.
+In the combined case, each particle potentially carried a different model realization in addition to a different kinetic parameter vector. 
+We kept the update rules the same (along with the update parameters). 
+Thus, each particle competed on the basis of goodness of fit, which allowed different model structures to contribute to the overall behavior of the swarm. 
+We considered five possible model structures (A through E), where network A was the correct formulation (used to generate the synthetic data). 
+We considered a population N = 100 particles, where each particle in the swarm was assigned a model structure, and a random parameter vector.
+The PSO algorithm, model equations, and the objective function were encoded and solved in the Octave programming language \citep{Octave:2014}.
 
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure-6-AlternativeNetworks.pdf}
+\caption{Schematic of the alternative allosteric control programs used in the structural particle swarm computation. 
+Each network had the same enzymatic connectivity, initial conditions and kinetic parameters, but alternative feedback control structures for the first enzyme in the pathway.}\label{fig-alternative-networks}
+\end{figure}
+
+\clearpage
+
+\begin{figure}
+\centering
+\includegraphics[width=0.6\textwidth]{./figs/Figure-7-ControlSearch.pdf}
+\caption{Combined control structure and kinetic parameter search results. Think about this ...}\label{fig-control-search}
+\end{figure}
+
+\clearpage
+
 % Supplemental figures -

### 2014-07-15 Paper_v1.tex
Going home 7/15/14 - good day

@@
 $\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$. Conversely, if $\sigma_{ij}>0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates
-metabolite $i$ is not connected with reaction $j$. Lastly, $\lambda_{i}$ denotes the scaled enzyme degradation constant. The system material balances are subject to the
+metabolite $i$ is not connected with reaction $j$. Lastly, $\lambda_{i}$ denotes the scaled enzyme degradation constant. The system material balances were subject to the
 initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$ and $\mathbf{\epsilon}\left(t_{o}\right) = \mathbf{1}$ (initially we have 100\% cell-free enzyme abundance).
@@
 The product in Eqn. \eqref{eqn:rate-bar} was carried out over the set of \textit{reactants} for reaction $j$ (denoted as $m_{j}^{-}$).
+
 The allosteric regulation term $v_{j}$ depended upon the combination of factors which influenced the activity of enzyme $i$.
@@
 	v_{j} = \max\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right)
-\end{equation}Conversely, if enzyme activity was inhibited by a $m$ metabolites, we modeling this inhibition as:
+\end{equation}where $0\leq f_{ij}\left(x\right)\leq 1$ was a regulatory transfer function that calculated the influence of metabolite $i$ on the activity of enzyme $j$. 
+Conversely, if enzyme activity was inhibited by a $m$ metabolites, we modeling this inhibition as:
 \begin{equation}
@@
 	d_{j} &=& 1 - \max_{j^{-}}\left(f_{1j}\left(x\right),\hdots,f_{nj}\left(x\right)\right)
-\end{eqnarray}where $j^{+}$ and $j^{-}$ denote the sets of activating, and inhibitory factors for enzyme $j$. 
+\end{eqnarray}The quantities $j^{+}$ and $j^{-}$ denoted the sets of activating, and inhibitory factors for enzyme $j$. 
 If an enzyme had no allosteric factors, we set $v_{j} = 1$.
-In this study, each individual factor had the form:
+There are many possible functional forms for $0\leq f_{ij}\left(x\right)\leq 1$. 
+However, in this study, each individual transfer function took the form:
 \begin{equation}\label{eqn:control-factor}

### 2014-07-21 Paper_v1.tex
First commit

--- 2014-07-11 Paper_v1.tex
@@
 \begin{titlepage}
-{\par\centering\textbf{\Large Dynamic Modeling of Cell Free Metabolic Networks using Effective Kinetic Models}}
+{\par\centering\textbf{\Large Dynamically Dimensioned Particle Swarm Optimization}}
 \vspace{0.05in}
-{\par \centering \large{ Joseph A. Wayman and Jeffrey D. Varner$^{*}$}}
+{\par \centering \large{Adithya Sagar, Christine Shoemaker and Jeffrey D. Varner$^{*}$}}
 \vspace{0.10in}
@@
 \vspace{0.1in}
-{\par \centering \textbf{To be submitted:}~\emph{Processes}}
+{\par \centering \textbf{To be submitted:}~\emph{BMC~Systems~Biology}}
 \vspace{0.5in}
@@
 
-{\noindent \textbf{Keywords:}~Cell free metabolism, Mathematical modeling}
+{\noindent \textbf{Keywords:}~Parameter identification, Mathematical modeling}
 
@@
 \section*{Introduction}
-Whole-cell bacterial processes are widely used in biotechnology to produce an array of products including high-value protein therapeutics.
-However, whole-cell processes share the central limitation of requiring cell growth, 
-which redirects resources away from product synthesis, and cell walls, which complicate interrogation and control of intracellular metabolic processes.
-On the other hand, cell-free metabolic systems offer many advantages over traditional in vivo production methods. 
-For example, cell-free systems can direct scarce metabolic resources exclusively towards a single product of interest. 
-Moreover, with no cell wall, cell free systems can more easily be interrogated, and substrates of the metabolite processes directly controlled.
-Cell free production offers the unique opportunity to study metabolism without the complication of cell growth and gene expression processes.
-For modeling, this implies that we need only consider allosteric regulation of enzyme activity when building and testing cell free metabolic models.
-Of course, modeling allosteric mechanisms is itself a difficult problem when the model is at a whole genome scale. To address this problem, we have developed a
-an approach based upon the constrained fuzzy logic framework of Morris and Lauffenburger [REFHERE].
-
-In this study, we present an effective cell free metabolic modeling framework, and test this framework using two proof of concept metabolic networks. 
-[FINISH].
 
@@
 
-\subsection*{Formulation and properties of cell free effective models.}
-We developed two proof of concept metabolic networks to investigate the features of our effective cell free modeling approach (Fig. \ref{fig-networks}).
-In both examples, substrate $S$ was converted to the end-products $P_{1}$ and $P_{2}$ through a series of enzymatically catalyzed reactions, 
-including a branch point at hypothetical metabolite $M_{2}$. 
-Several of these reactions involved cofactor dependence ($AH$ or $A$), and various allosteric regulation mechanisms. 
-Network A included feedback inhibition of the initial pathway enzyme ($E_{1}$) by pathway end products $P_{1}$ and $P_{2}$ (Fig. \ref{fig-networks}A).
-On the other hand, network B involved feedback inhibition of $E_{1}$ by $P_{2}$ and $E_{6}$ by $P_{1}$ (Fig. \ref{fig-networks}B). 
-In both networks, branch point enzymes $E_{3}$ and $E_{6}$ were subject to feed-forward activation by cofactor $AH$.
-Lastly, enzyme activity was assumed to decay according to a first-order rate law in both cases. 
-Allosteric regulation of enzyme activity was represented using a novel rule-based strategy, similar in spirit to the 
-Constrained Fuzzy Logic (cFL) approach of Lauffenberger and coworkers \citep{Morris:2011aa}. 
-In this formulation, Hill-like transfer functions were used to calculate the influence of metabolite abundance upon target enzyme activity. 
-When an enzyme was potentially sensitive to more than one regulatory influence, logical rules were used to select which transfer function regulated enzyme 
-activity at any given time (Fig. \ref{fig-control-schematic}).      
-Thus, our test networks involved important features such as cofactor recycling, enzyme activity and metabolite dynamics, 
-as well as multiple overlapping allosteric regulatory mechanisms.  
-As such, developing our effective modeling approach using these simple problems gave us valuable insight into the development of larger network models, 
-without the complication of network size.
+Coagulation is an archetype biochemical network that is highly interconnected and tightly regulated with multiple positive and negative feedback loops. 
+The biochemistry underlying coagulation has been well studied [REFHERE], and reliable experimental coagulation models have been developed [REFHERE].
 
-The rule based regulatory strategy approximated the behavior of classical allosteric activation and inhibition mechanisms (Fig. \ref{fig-kinetics-simulations}).  
-We first explored feed-forward substrate activation of enzyme activity (for both positive and negative cooperativity). 
-Consistent with classical data, the rule based strategy predicted a sigmoidal relationship between substrate abundance 
-and reaction rate as a function of the cooperativity parameter (Fig. \ref{fig-kinetics-simulations}A). 
-For cooperativity parameters less than unity, increased substrate abundance \textit{decreased} the reaction rate. This was consistent with the idea that
-substrate binding \textit{decreases} at regulatory sites negatively impacts the ability of the enzyme to bind substrate at the active site. 
-On the other hand, as the cooperativity parameter increased past unity, the rate of conversion of substrate $S$ to product $P$ by enzyme $E$ approached a step function.
-In the presence of an inhibitor, the rule based strategy predicted non-competitive like behavior as a function of the 
-cooperativity parameter (Fig. \ref{fig-kinetics-simulations}B). 
-When the control gain parameter, $\kappa_{ij}$ in Eqn. \eqref{eqn:control-factor}, was greater than unity, the inhibitory force was
-directly proportional to the cooperativity parameter, $\eta$ in Eqn. \eqref{eqn:control-factor}. 
-Thus, as the cooperativity parameter increased, the maximum reaction rate decreased (Fig. \ref{fig-kinetics-simulations}B, orange).
-However, when the gain parameter was less than unity, enzyme inhibition increased with \textit{decreasing} cooperativity, 
-i.e., smaller $\eta$ yielded increased inhibition (Fig. \ref{fig-kinetics-simulations}B).
-Interestingly, our rule based approach was unable to directly simulate competitive inhibition of enzyme activity. 
-For competitive inhibitors, the kinetic component of our rate, $\bar{r}_{j}$ in \eqref{eqn:rate-factor}, could be modified to account 
-for the inhibition (data not shown). 
-Taken together, the rule based strategy captured classical regulatory patterns for both enzyme activation and inhibition. 
-Thus, we are able to model complex kinetic phenomena such as ultrasensitivity, despite an effective description of reaction kinetics.
-
-End product yield was controlled by feedback inhibition, while selectivity was controlled by branch point enzyme inhibition (Fig. \ref{fig-onoff-simulations}).
-A critical test of our modeling approach was to simulate networks with known behavior. If we cannot reproduce the expected behavior of simple networks, then our
-effect modeling strategy, and particularly the rule-based approximation of allosteric regulation, will not be feasible for large scale problems.
-We considered two cases,  control on/off, for each network configuration. 
-Each of these cases had identical kinetic parameters and initial conditions; 
-the \textit{only} differences between the cases was the allosteric regulation rules, and the control parameters associated with these rules. 
-As expected, end product accumulation was larger for network A when the control was off (no feedback inhibition of $E_{1}$ by $P_{1}$ and $P_{2}$),
-as compared to the on case (Fig. \ref{fig-onoff-simulations}A). We found this behavior was robust to the choice of underlying kinetic parameters, 
-as we observed that same qualitative response across an ensemble of randomized parameter sets (N = 100). 
-The control on/off response of network B was more subtle. In the off case, the behavior was qualitatively similar to network A. 
-However, for the on case, flux was diverted away from $P_{2}$ formation by feedback inhibition of $E_{6}$ activity at the $M_{2}$ branch point by $P_{1}$ (Fig. \ref{fig-onoff-simulations}B).
-Lower $E_{6}$ activity at the $M_{2}$ branch point allowed more flux toward $P_{1}$ formation, hence the yield of $P_{1}$ also increased (Fig. \ref{fig-onoff-simulations}C).
-Again, the control on/off behavior was robust to the values of the kinetic parameters, as the same qualitative trend was conserved across 
-an ensemble of possible randomized kinetic parameters (N = 100). Taken together, these simulations suggested that the rule based allosteric control 
-concept could robustly capture expected feedback behavior.
-
-\subsection*{Estimating parameters and effective allosteric regulatory structures.}
-A critical challenge for any dynamic model is the estimation of kinetic parameters. 
-For metabolic processes, there is also the added challenge of identifying the regulation and control structures that manage metabolism.  
-Of course, these issues are not independent; any description of enzyme activity regulation will be a function of system state, which
-in turn depends upon the kinetic parameters. For cell free systems, regulated gene expression has been removed, however, enzyme activity regulation is still operational. 
-We explored this linkage by estimating model parameters from synthetic data using both network
-structures. We generated noise-corrupted synthetic measurements of the substrate $S$, intermediate $M_{5}$ and end-product $P_1$ approximately every 20 min using
-network A. We then generated an ensemble of model parameter estimates by minimizing the difference between model simulations and the synthetic data using particle swarm
-optimization, starting from random initial guesses. 
-The estimation of kinetic parameters was sensitive to the choice of regulatory structure (Fig. \ref{fig-parameter-fit}).
-PSO identified an ensemble of parameters that bracketed the mean of the synthetic measurements in less than 1000 iterations when the control structure was correct (Fig.
-\ref{fig-parameter-fit}A and B).
-However, when there was network mismatch (network B fit against network A synthetic data), PSO unable to identify an ensemble, all else being the same (Fig. \ref{fig-parameter-fit}C and D).
-Interestingly, the particle swarm generated a \textit{sloppy} parameter ensemble, in the absence of network mismatch (Fig. ZZZ).
-
-We modified our particle swarm identification strategy to simultaneously search over control structures and kinetic parameters (Fig. ZZZ). 
-We constructed five putative network models, each with the same enzymatic connectivity but different allosteric structures (Fig. ZZZ). 
-We then initialized a population of particles, each with one of the five putative regulatory programs, and randomized kinetic parameters.
-We biased the distribution of the particle population according to \textit{a~prior} belief of the correct regulatory program.
-To this end, we considered three different priors, a uniform distribution where each regulatory structure represented 20\% of the population, and
-two mixed distributions that were positively or negatively biased towards the correct structure (Table ZZ).
-
-Discrimination between competing regulatory formulations can be achieved by optimal experimental design (Fig. ZZ). [FINISH]
+This makes it an ideal system for mathematical modeling and parameter estimation.
+Coagulation is regulated by a set of serine proteases also known as coagulation factors and blood platelets.
+The coagulation factors are generally in an inactive state and are known as zymogens.
+These zymogens are activated when through certain triggers.
+These trigger events like injury or trauma or sepsis exposure factors like collagen, tissue factor and von Willebrand factor (vWF) to blood.
+The exposure of these factors to blood kick-starts a series of convergent cascades that lead to conversion of zymogen prothrombinase to thrombin.
+Luan et al.
+modeled coagulation using coupled non-linear ordinary differential equations with 301 reactions and 192 species.
 
@@
 \section*{Discussion}
-In this study, we proposed a effective modeling strategy to dynamically simulate cell free metabolic networks. 
-We tested this strategy using two proof of concept metabolic networks.
-In both networks, substrate $S$ was converted to the end-products $P_{1}$ and $P_{2}$ through a series of enzymatically catalyzed reactions, 
-including a branch point at hypothetical metabolite $M_{2}$. While both networks had the same enzymatic connectivity, that had differing control structures.
-[FINISH]
 
-Cybernetic models, other dynamic models of metabolism.
-
-While the results of this study were encouraging, there are several critical next steps that must be accomplished before we can model genome scale cell free metabolic networks.
-[FINISH]
+\clearpage
 
@@
 
-\subsection*{Formulation and solution of the model equations.}
-We used ordinary differential equations (ODEs) to model the time evolution of metabolite ($x_{i}$) and scaled enzyme abundance ($\epsilon_{i}$) in hypothetical cell free
-metabolic networks:
-\begin{eqnarray}
-	\frac{dx_{i}}{dt} & = & \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)\qquad{i=1,2,\hdots,\mathcal{M}}\\
-	\frac{d\epsilon_{i}}{dt} & = & -\lambda_{i}\epsilon_{i}\qquad{i = 1,2,\hdots,\mathcal{E}}
-\end{eqnarray}where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of metabolites and $\mathcal{E}$ denotes the number of
-enzymes in the model. The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ 
-denotes the rate of reaction $j$. Typically, reaction $j$ is a non-linear function of metabolite and enzyme abundance, as well as unknown kinetic parameters 
-$\mathbf{k}$ ($\mathcal{K}\times{1}$).
-The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$. If 
-$\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$. Conversely, if $\sigma_{ij}>0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates
-metabolite $i$ is not connected with reaction $j$. Lastly, $\lambda_{i}$ denotes the scaled enzyme degradation constant. The system material balances are subject to the
-initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$ and $\mathbf{\epsilon}\left(t_{o}\right) = \mathbf{1}$ (initially we have 100\% cell-free enzyme abundance).
-
-Each reaction rate was written as the product of two terms, a kinetic term ($\bar{r}_{j}$) and a regulatory term ($v_{j}$):
-\begin{equation}\label{eqn:rate-factor}
-	r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right) = \bar{r}_{j}v_{j}
-\end{equation}We used multiple saturation kinetics to model the reaction term $\bar{r}_{j}$:
-\begin{equation}\label{eqn:rate-bar}
-	\bar{r}_{j} = k_{j}^{max}\epsilon_{i}\left(\prod_{s\in{m_{j}^{-}}}\frac{x_{s}}{K_{js} + x_{s}}\right)
-\end{equation}where $k_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the scaled enzyme activity which catalyzes reaction $j$, and
-$K_{js}$ denotes the saturation constant for species $s$ in reaction $j$. 
-The product in Eqn. \eqref{eqn:rate-bar} was carried out over the set of \textit{reactants} for reaction $j$ (denoted as $m_{j}^{-}$). 
-The allosteric regulation term $v_{j}$ depended upon the combination of factors which influenced the activity of enzyme $i$.
-For each enzyme, we used a rule based approach to select from competing control factors (Fig. \ref{fig-control-schematic}). 
-If an enzyme was activated by $m$ metabolites, we modeled this activation as:
-\begin{equation}
-	v_{j} = \max\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right)
-\end{equation}Conversely, if enzyme activity was inhibited by a $m$ metabolites, we modeling this inhibition as:
-\begin{equation}
-	v_{j} = 1 - \max\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right)
-\end{equation}Lastly, if an enzyme had both $m$ activating and $n$ inhibitory factors, we modeled the regulatory term as:
-\begin{equation}
-	v_{j} = \min\left(u_{j},d_{j}\right)
-\end{equation}where:
-\begin{eqnarray}
-	u_{j} &=& \max_{j^{+}}\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right) \\
-	d_{j} &=& 1 - \max_{j^{-}}\left(f_{1j}\left(x\right),\hdots,f_{nj}\left(x\right)\right)
-\end{eqnarray}where $j^{+}$ and $j^{-}$ denote the sets of activating, and inhibitory factors for enzyme $j$. 
-If an enzyme had no allosteric factors, we set $v_{j} = 1$.
-In this study, each individual factor had the form:
-\begin{equation}\label{eqn:control-factor}
-	f_{i}\left(\mathbf{x}\right) = \frac{\kappa_{ij}^{\eta}x_{j}^{\eta}}{1 + \kappa_{ij}^{\eta}x_{j}^{\eta}}
-\end{equation}where $x_{j}$ denotes the abundance of metabolite $j$, and $\kappa_{ij}$ and $\eta$ are control parameters. 
-The $\kappa_{ij}$ parameter was species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).
-The model equations were encoded using the Octave programming language, and solved using the LSODE routine in Octave \citep{Octave:2014}.
-
-\subsection*{Estimation of model parameters from synthetic experimental data}
-Model parameters were estimated by minimizing the difference between simulations and synthetic experimental data:
-\begin{equation}\label{eqn:objective-function}
-	\min_{\mathbf{k}} \sum_{\tau=1}^{\mathcal{T}}\sum_{j=1}^{\mathcal{S}}\left(\frac{\hat{x}_{j}\left(\tau\right) - x_{j}\left(\tau,\mathbf{k}\right)}{\omega_{j}\left(\tau\right)}\right)^{2}
-\end{equation}where $\hat{x}_{j}\left(\tau\right)$ denotes the measured value of species $j$ at time $\tau$, $x_{j}\left(\tau,\mathbf{k}\right)$ denotes the simulated 
-value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$. The outer summation is respect to
-time, while the inner summation is with respect to state. We wanted to approximate a realistic model identification scenario, thus we assumed noisy experimental data, 
-limited sampling resolution (approximately 20 minutes per sample) and a limited number of measurable metabolites.
-
-We minimized Eqn. \eqref{eqn:objective-function} using Particle swarm optimization (PSO) \citep{PSO}.
-PSO is global optimization procedure which uses a swarming metaheuristic to explore parameter spaces. 
-A strength of PSO is its ability to find the global minimum, even in the presence of potentially many local minima, by communicating the local
-error landscape experienced by each particle collectively to the swarm using the update rules:
-\begin{eqnarray}
-	\mathbf{\Delta}_{i} &=&\theta_{1}\mathbf{\Delta}_{i} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{k}_{i}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G} - \mathbf{k}_{i}\right) \\
-	\mathbf{k}_{i} &=& \mathbf{k}_{i} + \mathbf{\Delta}_{i}
-\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right) = (1.0,0.05564,0.02886)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes local best solution found by particle $i$, and
-$\mathcal{G}$ denotes the best solution found over the entire population of particles.
-The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
-parameters ($\mathcal{K}\times{1}$). The PSO algorithm, and the objective function were encoded and solved in the Octave programming language \citep{Octave:2014}.
-
-\subsection*{Model discrimination}
+\subsection*{Formulation and solution of coagulation model equations.}
+\subsection*{Optimization methods.}
 
@@
 \clearpage
-%\bibliographystyle{plain}
-%\bibliographystyle{IEEEbib}
 
@@
 %\begin{comment}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-1-Networks.pdf}
-\caption{Proof of concept cell-free metabolic networks considered in this study. Substrate $S$ is converted to products $P_{1}$ and $P_{2}$ through a series of chemical conversions
-catalyzed by enzyme(s) $E_{j}$. The activity of the pathway enzymes is subject to both positive and negative allosteric regulation. }\label{fig-networks}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-2-ControlSchematic.pdf}
-\caption{Schematic of the rule based allosteric enzyme activity control laws. }\label{fig-control-schematic}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-3-EnzymeKinetics.pdf}
-\caption{Kinetics of simple transformations in the presence of activation and inhibition. 
-A:The conversion of substrate $S$ to product $P$ by enzyme $E$ was activated by $S$. 
-B:The conversion of substrate $S$ to product $P$ by enzyme $E$ was inhibited by inhibitor $I$. 
-}\label{fig-kinetics-simulations}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=0.55\textwidth]{./figs/Figure-4-OnOffSimulations.pdf}
-\caption{On/off control simulations for network A and network B for an ensemble of kinetic parameter sets versus time. 
-For each case, N = 100 simulations were conducted using kinetic and initial conditions randomly generated from a hypothetical true parameter set. 
-The gray area represents $\pm$ one standard deviation surrounding the mean. Control parameters were fixed during the ensemble calculations.}\label{fig-onoff-simulations}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-5-ParameterFit.pdf}
-\caption{Parameter estimation from synthetic data for the same and mismatched allosteric control logic. }\label{fig-parameter-fit}
-\end{figure}
 

### 2014-07-21 Paper_v1.tex
First commit on Publications repository

@@
 %\bibliographystyle{plain}
-\bibliographystyle{plos2009}
+\bibliographystyle{msb}
 
@@
 \renewcommand{\rmdefault}{phv}\renewcommand{\sfdefault}{phv}
-\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
 
@@
 \begin{titlepage}
-{\par\centering\textbf{\Large Dynamically Dimensioned Particle Swarm Optimization}}
+{\par\centering\textbf{\Large Modeling and Analysis of Hormone and Mitogenic Signal Integration in Prostate Cancer}}
 \vspace{0.05in}
-{\par \centering \large{Adithya Sagar, Christine Shoemaker and Jeffrey D. Varner$^{*}$}}
+{\par \centering \large{ Katharine V. Rogers, Joseph A. Wayman, Ryan Tasseff, Timothy Chu, Kyriakos Tsahalis, Spencer Davis, Rohit Yallamraju, Gongcheng Lu, Matthew P. DeLisa, 
+and Jeffrey D. Varner$^{*}$}}
 \vspace{0.10in}
@@
 \vspace{0.1in}
-{\par \centering \textbf{Running Title:}~Modeling cell free metabolism}
+{\par \centering \textbf{Running Title:}~Modeling signal transduction in prostate cancer}
 \vspace{0.1in}
-{\par \centering \textbf{To be submitted:}~\emph{BMC~Systems~Biology}}
+{\par \centering \textbf{To be submitted:}~\emph{Molecular Systems Biology}}
 \vspace{0.5in}
@@
 \section*{Abstract}
+Prostate cancer is the most common cancer in men and the second leading cause of cancer related death in the United States. 
+Androgens, such as testosterone, are required for prostate cancer growth. 
+Androgen ablation in combination with radiation or chemotherapy remains the primary non-surgical treatment for androgen-dependent prostate cancer. 
+However, androgen ablation typically fails to permanently arrest cancer progression, often resulting in castration resistant prostate cancer (CRPC). 
+CRPC is closely related to metastasis and decreased survival. 
+In this study, we developed and analyzed a population of mathematical models describing growth factor and hormone signal integration in androgen dependent, intermediate and resistant prostate cancer cells.
+The model describes the integration of two simultaneous extracellular signaling inputs, namely, androgen and growth factors into a G1/S cell cycle checkpoint decision. 
+Model parameters were identified from 43 studies in androgen dependent and resistant LNCaP cell lines. 
+The model was validated by comparing simulations with an additional 29 data sets from LNCaP cell lines that were not used during training. 
+Additionally, data from four drug trials was also used to evaluate the model’s performance.
+Sensitivity analysis, conducted over an ensemble of prostate signaling models, suggested that in an androgen free environment general translation and transcription was more sensitive in androgen dependent cells, while in androgen independent cells the PI3K and MAPK pathway species were more sensitive.
+In a constant DHT environment sensitive species were conserved between the cell lines. 
+These results suggest targeting the PI3K and MAPK pathways in addition to anti-androgen therapies as a treatment for AIPC.
 
-{\noindent \textbf{Keywords:}~Parameter identification, Mathematical modeling}
+{\noindent \textbf{Keywords:}~Prostate cancer, signal transduction, mathematical modeling}
 
@@
 
+%
+%Sipuleucel-T is a first generation cancer vaccine \cite{Kantoff2010}, while radium-223 is a targeted alpha emitter for bone metastasis \cite{Parker2013}. 
+%Abiraterone is an inhibitor of cytochrome P450 17A1 (CYP17), a critical enzyme in androgen biosynthesis \cite{Sartor2013} while enzalutamide inhibits androgen receptor (AR) translocation to the nucleus, as well as DNA binding and coactivator recruitment \cite{Scher2012}.
+%Thus, understanding the operation of the biological pathways which ultimately facilitate the transition to CRPC is vital to the discovery of new therapeutic strategies against this disease.
+%The model focuses on the outlaw pathway, in which constitutively or overly-activated receptor tyrosine kinases (RTKs) stimulate AKT and mitogen-activated protein kinases (MAPKs), which in turn activate cytosolic androgen receptor (AR) in the absence of androgen.
+Interestingly, among the few genes activated AR represses is cellular prostatic acid phosphatase (cPAcP), itself a key regulatory of RTK activation.
+Thus, the outlaw pathway encodes crosstalk between RTK-dependent kinase and AR activation inside of a positive feedback loop.
+The model was validated with multiple experiments in LNCaP cell lines.
+In addition  For example, in response to enzalutamide, an AR inhibitor, the model predicted a decline in PSA expression similar to patient response.
+%and there is limited data on combination therapies or the use of biomarkers to develop personalized treatment strategies.
+
+Prostate cancer (PCa) is the most commonly diagnosed cancer and the second leading cause of cancer-related death in men in the United States \cite{Siegel2013}. 
+Initially, PCa cells depend upon the activation of cytosolic androgen receptors (AR) by androgen hormones, such as testosterone, for survival and growth. 
+Androgen ablation in combination with radiation or chemotherapy remains the primary non-surgical treatment for androgen-dependent prostate cancer (ADPC) \cite{Huggins1967}. 
+However, androgen ablation typically fails to permanently arrest cancer progression as malfunctioning cells eventually lose androgen sensitivity and proliferate without hormone. 
+The loss of androgen sensitivity results in castration resistant prostate cancer (CRPC), a phenotype closely linked with metastasis and greatly reduced survival \cite{Harris2009}.
+Currently, there are six approved treatments that demonstrate a survival advantage in patients with metastatic CRPC, each of these target diverse aspects of the disease \cite{Sartor2013}.
+The taxane family members docetaxel and cabazitaxel interact with microtubule stability \cite{Tannock2004,Bono2010}, while abiraterone \cite{Sartor2013} or enzalutamide \cite{Scher2012} interfere with androgen signaling by blocking androgen formation or nuclear translocation, respectively.
+Other approved treatments are non-specific to PCa. 
+For example, general treatments such as sipuleucel-T, a first generation cancer vaccine \cite{Kantoff2010}, 
+and radium-223, an alpha emitter which targets bone metastasis \cite{Parker2013}, are both approved to treat CRPC.
+Unfortunately, regardless of the therapeutic approach, the survival advantage of these treatments is typically only a few months. 
+Thus, understanding the molecular basis of the loss of androgen sensitivity in CRPC could be an important step for the development of the next generation of therapies with a prolonged survival
+advantage.
+
+Androgen-induced proliferation and survival depends upon many coordinated signal transduction and gene expression events. 
+Androgen Receptor (AR) is part of the nuclear hormone receptor superfamily, which includes other important cancer targets such as progesterone receptor (PR) and estrogen receptor (ER) in breast cancer \cite{Aranda2001}.
+Nuclear hormone receptors act as ligand dependent transcription factors interacting with specific DNA sequences of target genes as either monomers, heterodimers, or homodimers; AR, PR, and ER act as homodimers.
+In the case of AR these specific DNA sequences are known as androgen response elements (ARE) \cite{Mangelsdorf1995}.
+In the absence of androgen, AR is predominately found in the cytoplasm bound to heat shock proteins (HSP) \cite{Prescott2006}.
+%Nuclear Hormone Receptors (Reviewed \cite{Aranda2001}) and AR in disease reviewed in \cite{Matsumoto2013}
+Androgen, either testosterone or testosterone metabolites such as 5$\alpha$-dihydrotestosterone (DHT), enter prostate cells and interact with the cytosolic androgen receptor (AR). 
+The interaction of DHT with AR promotes the dissociation of AR from chaperones such as HSP \citep{Pratt:1997kx} and its subsequent dimerization, phosphorylation and translocation to the nucleus (reviewed by Brinkmann \textit{et al.} \citep{Brinkmann1999}).
+Activated nuclear AR drives a gene expression program broadly referred to as androgen action, that promotes both proliferation and survival.
+In addition to many genes including itself, activated nuclear AR promotes the expression and secretion of prostate specific antigen (PSA), arguably the best known PCa biomarker \citep{Feldman2001}.
+PSA is commonly used as a prostate cancer indicator, although its prognostic ability is controversial \citep{Attard2009,Hoffman2011,Moyer2012}. 
+In CRPC, AR signals in the absence of androgens. 
+Androgen dependent (AD) prostate cells can become castration resistant (CR) through several possible mechanisms, including constitutively amplified AR expression and altered AR sensitivity to testosterone or other non-androgenic molecules \citep{Feldman2001}.
+In this study, we focused on the aberrant activation of AR by kinase signaling cascades, sometimes called the outlaw pathway. 
+Outlaw pathway activation is driven by over-activated receptor tyrosine kinases (RTKs), a common pathology in many cancer types including PCa \citep{Slamon1989,Craft1999}. 
+RTKs stimulate downstream kinases, including the AKT and mitogen-activated protein kinase (MAPK) pathways, which promote AR phosphorylation and dimerization in the absence of an androgen signal \citep{Craft1999,Yeh1999}.
+Interestingly, among the few genes activated AR represses is cellular prostatic acid phosphatase (cPAcP), itself a key regulatory of RTK activation \cite{Veeramani2005}. 
+Thus, in CRPC the androgen program is initiated without the corresponding extracellular hormone cue, potentially from crosstalk between growth factor and hormone receptor pathways.
+
+In this study, we developed a mathematical model of growth factor and hormone signal integration in androgen dependent, intermediate and resistant prostate cancer cells. 
+We used this model to better understand which components and processes were differentially important in AD versus CR cells.
+The new model architecture was a significant advance over our previous prostate signaling model \citep{Tasseff2010}. 
+We added the regulated expression of ten additional proteins, including the cell cycle restriction point proteins cyclin D (and the differentially spliced variants cyclin D1a and cyclin D1b), cyclin E, cyclin-dependent kinase inhibitor 1A (p21Cip1), and cyclin-dependent kinase inhibitor 1B (p27Kip1).
+Also, we included the Rb/E2F pathway, expanded our description of the activation of the mammalian target of rapamycin (mTOR) protein and its role in translation initiation, and included the regulation of AR action by cyclin D1a and E2F.
+However, this upgraded architecture, while increasing the biological scope of the model, also expanded the number of unknown model parameters. 
+To estimate these parameters, we used multiobjective optimization in combination with dynamic and steady-state data sets generated in AD, intermediate and CR LNCaP cell lines. 
+We identified a population of approximately N = 5000 models (from well over a million candidate models) which described both AD and CR data sets using a single model structure.
+Furthermore, we tested the model using an additional 29 LNCaP data sets not used for model training, along with four drug studies.
+We analyzed the model population using sensitivity and robustness analysis to uncover differentially important mechanisms in AD versus CR cell lines. 
+In the presence of androgen, the sensitivity profile was similar between AD and CR cells. 
+Components of the MAPK and PI3K pathways were highly fragile, irrespective of the level of androgen dependence. 
+However, in the absence of androgen, there were 609 statistically significant shifts in species sensitivity between AD and CR cells.
+Of these, 108 were larger than one standard deviation above the mean.
+In CR cells, HER2 activation of the MAPK and PI3K pathways was significantly more important, as was AR activation through the MAPK pathway. 
+On the other hand, components of the translation and transcription infrastructure were differentially more important in AD cells in the absence of androgen.  
+Taken together, our analysis suggested that independently targeting the PI3K or MAPK pathways in combination with anti-androgen therapies could perhaps be an effective treatment strategy for CRPC.
+
+%(323 more and 287 less sensitive)
+%In this study, we rank-ordered the importance of the molecular components controlling the transition from ADPC to CRPC.
+%Toward this, we developed and analyzed an ensemble of mechanistic mathematical models of the signal transduction processes governing the 
+%transition from androgen dependent to castration resistant prostate cancer.
+%The objective of this study was to rank-order the importance of components in CRPC and ADPC using computational tools. 
+%We formulated and analyzed an ensemble of mechanistic mathematical models of the signal transduction processes in CRPC and ADPC 
+%that was an extension of our previous work .
+%In particular, we focused on the outlaw pathway and the possible crosstalk between RTK-dependent kinases and AR activation. 
+%First, we expanded the original model by adding ten  
+%Third, we added additional co-activators, such as cyclin E, and inhibitors of AR, such as
+%ODE models typically contain a large number of unknown model parameters, which must be estimated  
+%Due to uncertainty in parameter estimation and to account for differences between individual cells and cell sublines, an ensemble of parameter sets was used to model the system. 
+%Proteins that were not modeled at a gene regulatory level were generated using zeroth order reactions. 
+%All protein and RNA species were degraded with first order reaction rates. 
+%To obtain an ensemble of parameter sets, we used multiobjective optimization. 
+%Relative concentration values were quantified 
+%Training data came from experiments on LNCaP sublines. 
+%We used , previously developed in the lab \cite{Song2010}, to determine a diverse ensemble of models. 
+%C-33 cells, having undergone fewer than 33 passages, are considered AD while C-51 cells, having between 34 and 80 passages, are considered an intermediate prostate cancer cell line. 
+%C-81 (passage numbers beyond 80) display a CR phenotype \cite{Lin2000}. 
+%http://endo.endojournals.org/content/140/11/5054.long or http://endo.endojournals.org/content/140/11/5054.long#ref-19
+
 \clearpage
@@
 
-Coagulation is an archetype biochemical network that is highly interconnected and tightly regulated with multiple positive and negative feedback loops. 
-The biochemistry underlying coagulation has been well studied [REFHERE], and reliable experimental coagulation models have been developed [REFHERE].
+\subsection*{Estimating an ensemble of prostate signaling models.} 
+We modeled the integration of growth factor, cell cycle and hormone signaling pathways in AD and CR LNCaP cells (Fig. \ref{fg:NetworkSchematic}). 
+The signaling architecture was hand curated from over 80 primary literature sources in combination with biological databases.
+The model equations were formulated as a system of ordinary differential equations (ODEs), where biochemical reaction rates were modeled using mass action kinetics.
+ODEs and mass action kinetics are common modeling tools \citep{Wayman2013}, however, ODEs have the disadvantage of requiring estimates for unknown model parameters.
+Many techniques have been developed to estimate ODE model parameters, often from noisy and sparse experimental data \citep{Moles:2003dn}.
+Typically these identification problems are underdetermined, hence no unique parameter values can be estimated \citep{Villaverde:2014kl}.
+Thus, instead of estimating a single yet highly uncertain parameter set, we estimated an ensemble of possible parameter sets using the Pareto Optimal Ensemble Techniques (POETs) algorithm \citep{Song2010}.
+POETs uses a combination of simulated annealing and local optimization techniques coupled with Pareto optimality-based ranking to simultaneously optimize multiple objective functions.
+Starting from an initial best fit set, we estimated the 1687 unknown model parameters (1674 kinetic parameters and 13 non-zero initial conditions) using 43 \textit{in vitro} data sets taken from AD, intermediate and CR LNCaP cells (Table~\ref{objective_table}).
+Each of the 43 training data sets was a separate objective in the multiobjective calculation. 
+The training data were steady-state or dynamic immunoblots from which we extracted relative species abundance from their optical density profiles. 
+POETs sampled well over a million possible parameter sets, from which we selected N = 5000 sets for further analysis. 
+Over the ensemble, the coefficient of variation (CV) of the kinetic parameters spanned 0.5 - 5.8, with 33\% of the parameters having a CV of less than one. 
+As a control, we also performed simulations for R = 100 random parameter sets to compare against the ensemble generated by POETs.
 
-This makes it an ideal system for mathematical modeling and parameter estimation.
-Coagulation is regulated by a set of serine proteases also known as coagulation factors and blood platelets.
-The coagulation factors are generally in an inactive state and are known as zymogens.
-These zymogens are activated when through certain triggers.
-These trigger events like injury or trauma or sepsis exposure factors like collagen, tissue factor and von Willebrand factor (vWF) to blood.
-The exposure of these factors to blood kick-starts a series of convergent cascades that lead to conversion of zymogen prothrombinase to thrombin.
-Luan et al.
-modeled coagulation using coupled non-linear ordinary differential equations with 301 reactions and 192 species.
+The ensemble of PCa models recapitulated training data in both AD and CR cell lines with only two experimentally justified parameter changes (Fig. \ref{fg:ColorPlot} and Fig. \ref{fg:Training}). 
+Data from the LNCaP clones C-33 (dependent), C-51 (intermediate), and C-81 (resistant) \cite{Lin2000,Igawa2002,Horoszewicz1983} along with the CR LNCaP cell lines LNCaP-Rf \cite{Murillo2001}, LNCaP-AI \cite{Chen2011} and LNAI \cite{Graff2009} were used for model identification.
+To simulate the effective difference between LNCaP cell lines, the parameter controlling the maximum rate of PAcP gene expression was scaled by 0.1 and 0.5, respectively, for the C-81 and C-51 cell-lines compared to C-33.
+This modification was based upon steady-state PAcP data from the three LNCaP clones \cite{Lee2003}. 
+Similarly, the expression of p16INK4 was adjusted in accordance with the study of Lu \textit{et al.} \citep{Lu1997}. 
+These two parameters were the only adjustable parameter differences between AD and CR cells. 
+To simulate an increased mTOR activation in the presence of a DHT stimulus, we added a first order activation term for mTOR activation with a DHT stimulus.
+Androgens have been shown to increase expression of proteins involved in cellular metabolism, which may lead to an increase in mTOR activation \cite{Xu2006}.
+The model fit 36 of the 43 training objectives for greater than 40\% of the ensemble members (Fig. \ref{fg:ColorPlot}A). 
+Conversely, only 10 of the 43 training objectives were captured with the random parameter control (Fig. \ref{fg:ColorPlot}B). 
+The model captured the crosstalk between RTK activation and androgen action (Fig. \ref{fg:Training}). 
+The model described DHT-induced PSA expression in both C-33 (Fig. \ref{fg:Training}A) and C-81 (Fig. \ref{fg:Training}B) cells. 
+Interestingly, simulations with the HER2 inhibitor AG879 recapitulated decreased PSA expression in C-81 cells (Fig. \ref{fg:Training}C) in the absence of androgen stimulation.
+AR action decreased the PAcP mRNA message (Fig. \ref{fg:Training}D), presumably leading to increased HER2 activity. 
+The model also recapitulated the integration of androgen action with AR expression, G1/S cell cycle protein expression and AKT phosphorylation. 
+For example, the model captured AR-induced AR expression following a DHT stimulus (Fig. \ref{fg:Training}H).
+Conversely, the transcription factor E2F inhibits AR transcription in LNCaP cells (Fig. \ref{fg:Training}I). 
+Other cell cycle proteins were also integrated with androgen action.
+For example, the cyclin D1 abundance increased in CR compared to AD cells in the absence of androgen (Fig.
+\ref{fg:Training}E), while DHT induced p21Cip1 expression in C-33 cells (Fig.
+\ref{fg:Training}F).
+The level of phosphorylated AKT was also increased in higher passage number cells (Fig. \ref{fg:Training}G).
 
-\clearpage
+%We used a single model architecture to describe the behavior of both AD and CR LNCaP cells, 
+%Two new additions to the model were the cell cycle protein p21Cip1 and transcriptional regulation of AR. 
+%We were able to capture training data for both mechanisms. 
+%AR transcriptionally upregulates its own expression \cite{Lin2003}\cite{Grad1999}. 
+%We captured the increase in AR concentration due to AR activation, and thus increased transcriptional activity, upon  
+%Higher basal AKT phosphorylation leads to increased activation of the PI3K/AKT pathway in PCa cells with less androgen dependence \cite{Lin2003}. 
+%\cite{Davis2006}
+
+%An additional protein we considered in validation that was present in training was phosphorylated SHC-transforming protein 1 (Shc).
+%Simulations predicted a similar increase in Shc phosphorylation after DHT stimulus (Fig. \ref{fg:Validation}D) \cite{Veeramani2005Onc}. 
+%The Prostate Cancer Working Group (PCWG) updated outcome measurements for castration resistant prostate cancer studies in 2008 \cite{Scher2008}. 
+%The primary endpoint in prostate cancer studies is overall survival, while PSA response is often a secondary endpoint. 
+%Figure \ref{fg:Validation}I shows fold change of PSA of ensemble members due to the application of sorafenib.
+
+%(Fig. \ref{fg:Validation}H)
+
+\subsection*{Validation revealed missing network structure.} 
+The model was validated against 29 \textit{in vitro} and four \textit{in vivo} studies (Table~\ref{prediction_table}). 
+For 15 of the 29 cases, greater than 40\% of the ensemble was qualitatively consistent with the experimental data (Fig. \ref{fg:ColorPlot}C). 
+However, for the random parameter control, only 7 of the 29 cases were satisfied (Fig. \ref{fg:ColorPlot}D). 
+We correctly predicted positive feedback between HER2 auto-activation and androgen action (Fig. \ref{fg:Validation}A and Fig. \ref{fg:Validation}B). 
+We also captured the dose-dependence of AR abundance on DHT (Fig. \ref{fg:Validation}C). 
+In addition to the cell line studies, we simulated the outcome of enzalutamide, lapatinib, and sorafenib clinical trials in AD and CRPC patients. 
+The trial end points were the reduction in PSA expression relative to an untreated baseline.  
+Enzalutamide acts on AR by inhibiting its nuclear translocation, DNA binding, and coactivator recruitment \cite{Scher2012}. 
+In the enzalutamide trial, 54\% of the patients that received the drug showed a PSA decline of $\geq$ 50\% while 25\% showed a decline $\geq$ 90\%. 
+We simulated enzalutamide exposure by reducing the rate constants governing activated AR binding to nuclear importer, cyclin E, and CDK6 to 1\% of their initial values. 
+Consistent with the trial, 62\% of ensemble members showed a $\geq$ 50\% decline in PSA abundance, while 14\% showed a $\geq$ 90\% decline (Fig. \ref{fg:Validation}G). 
+The second trial we simulated involved exposure of CRPC patients to sorafenib. 
+Sorafenib is a kinase inhibitor with activity against Raf, vascular endothelial growth factor receptor (VEGFR), platelet-derived growth factor receptor (PDGFR), c-kit and c-Ret \cite{Dahut2008}. 
+We considered only the effects of sorafenib on the protein kinase Raf, as VEGFR, PDGFR, c-kit and c-Ret were not included in the model. 
+None of the 22 patients in the sorafenib study showed a PSA decline of $>$ 50\%. 
+However, our simulations showed that approximately 55\% of the ensemble members had a PSA decline of $\geq$ 50\%. 
+The last drug we considered was lapatinib, an inhibitor of epidermal growth factor receptor (EGFR) and HER2 tyrosine kinase activity \cite{Liu2013}. 
+Two lapatinib drug trials were considered: one in which patients had CRPC and one in which patients had biochemically relapsed ADPC \cite{Whang2013,Liu2013}. 
+In the CRPC lapatinib drug trial, two of the 21 enrolled patients had a PSA response $\geq$ 47\%  \cite{Whang2013}. 
+For the CRPC case, our model showed 26.5\% of ensemble members with a PSA response $\geq$ 47\%. 
+Of the 35 patients enrolled in the ADPC lapatinib study, no PSA decreases was observed \cite{Liu2013}. 
+In this case, our model showed 9.2\% of ensemble members with a PSA response $\geq$ 50\%.
+Although no response to lapatinib was seen in ADPC clinical trials, \textit{in vitro} AD LNCaP experiments showed decreased PSA expression in response to lapatinib, most notably with the addition of DHT \cite{Liu2005}.
+
+Validation and training failures suggested the original signaling architecture was missing critical components. 
+Several of the failed training and validation simulations involved the response of the network to epidermal growth factor (EGF) stimulation. 
+For example, Chen \textit{et al.} showed that HER2 phosphorylation increased within five minutes following EGF stimulation of LNCaP-AI cells \citep{Chen2011}. 
+We predicted no connection between HER2 phosphorylation and EGF stimulation on this short timescale (Fig. \ref{fg:Validation}E). 
+Interestingly, we initially neglected the heterodimerization of HER2 with other ErbB family members in order to simplify the model.
+However, Chen \textit{et al.} suggested that HER2-EGFR heterodimerization could be an important factor in EGF-driven activation of HER2 \cite{Chen2011}. 
+We tested this hypothesis by developing a new model that included HER2 and EGFR heterodimerization. 
+We set the rate constants governing the assembly of HER2/EGFR heterodimers equal to EGFR homodimer assembly; all other parameters were unchanged.
+This was a reasonable first approximation, as the affinity of HER2/EGFR heterodimerization and EGRF homodimerization is thought to be similar \cite{Hendriks2003}. 
+With the inclusion of HER2-EGFR heterodimerization, we qualitatively fit the EGF-induced HER2 activation case and improved our training for experiments that involved an EGF stimulus, e.g., cyclin D mRNA and protein abundance following an EGF stimulus in C-33 cells (Fig \ref{fg:ColorPlot}A and C, white pixels).
+
+%in diverse environments (i.e., DHT, no DHT, drug) 
+%Stimuli included the presence or absence of DHT, presence of therapeutics, and also EGF. 
+%PIP3 recruits AKT and PDK1 to the membrane where PDK1 activates AKT \cite{Vivanco2002}. 
+%Thus, the PIP3 and AKT interaction is required in the model for AKT phosphorylation to occur.
+
+\subsection*{Sensitivity analysis identified differentially important features of the prostate architecture.} 
+We used sensitivity analysis to identify important signaling components in AD versus CR cells (Fig. \ref{fg:Sensitivity}). 
+We calculated first order steady-state sensitivity coefficients under different stimuli for 500 parameter sets selected from the ensemble. 
+Signaling components were rank-ordered based upon analysis of their sensitivity coefficient values. 
+In the presence of DHT, the sensitivity profile was similar for AD versus CR cells, with only a few differences (Fig. \ref{fg:Sensitivity}B). 
+The top 2\% of sensitive species, regardless of androgen dependence, involved components from the MAPK and PI3K pathways. 
+In particular, activated Ras, Raf, phosphorylated MEK, as well as PIP3 localized AKT, phosphorylated AKT, and PI3K were sensitive in both AD and CR cells. 
+Species involving PAcP and p16INK4 were more sensitive in AD cells, which was expected since the expression of these two proteins were the only parameters changed between AD and CR cells. 
+Other species such as E2F, cyclin E, and DHT-activated AR were also more sensitive in AD cells. 
+On the other hand, HER2-Grb2-Gab activation of PI3K and AKT inhibition of RAF were more sensitive in CR cells.
+
+The importance of signaling components varied with androgen dependence in the absence of DHT (Fig. \ref{fg:Sensitivity}A). 
+There were 609 statistically significant shifts in species sensitivity (318 more and 291 less sensitive) between CR and AD cells in a non-androgen environment. 
+However, only 108 of these shifts were greater than one standard deviation above the mean. 
+In CR cells, HER2 activation of ERK and PI3K was more sensitive, as was AR activation through the MAPK pathway. 
+This was expected, as outlaw pathway activity was elevated in castration resistant cells. 
+Species in the MAPK pathway were in general more sensitive in CR cells (128 out of 140 significant), with all forms of sPAcP more robust in CR cells.
+On the other hand, infrastructure pathways encoding transcription and translation were more sensitive in AD cells.
+PSA  and cyclin D1b (mRNA and mRNA complexes) were the only species involved in translation that were more robust in AD cells (14 out of 116).
+The transcription factor, E2F was more fragile in AD cells, while the transcription factors ETS and AP1 were more robust. 
+ETS and AP1 are activated by phosphorylated ERK, and ETS is also activated by active PKC \cite{Wilkinson2000,Lindemann2003}. 
+E2F is deactivated through binding to Rb, which is deactivated by cyclin D1 and CDK phosphorylation \cite{Lapenna2009}. 
+The model also included AP1 suppression of AR transcriptional activity (more sensitive in CR) \cite{Sato1997}, as well as inhibition of transcription of the AR gene by E2F (more sensitive in AD) \cite{Davis2006}.
+Species in the PI3K pathway that were more fragile in AD cells included Rheb and TOR complexes. 
+Interestingly, these species were included as the last step in the PI3K pathway prior to translation, with the phosphorylation of 4E-BP1 by TOR being considered the beginning of translation in this model.
+This again indicates that in the absence of DHT general translation is more fragile in AD cells.
+
+There was a large shift in sensitive species between an androgen and a non-androgen environment in both AD and CR cell lines (Fig. \ref{fg:Sensitivity}C and Fig. \ref{fg:Supp_Sensitivity}). 
+Of the 664 statistically significant shifts in AD cells, 288 were more sensitive between androgen versus non-androgen environments. 
+However, only 119 shifts were larger than one standard deviation above the mean. 
+Unsurprisingly, AR activation through DHT binding, with and without coactivators, in a DHT environment was more sensitive, as was AR inhibition of PAcP transcription (repressed by AR in the model). 
+Species further upstream, such as HER2 activation of the MAPK and PI3K/AKT pathways, were also more sensitive in a DHT environment. 
+Cell cycle species that were more fragile in the presence of DHT, included complexes involving p21Cip1 and CDC25A.
+In a non-androgen environment, basal transcription (68 out of 72) and translation (114 out of 120) were more fragile. 
+Other fragile species in the absence of DHT included Rb, E2F, Sam68, cyclin D1a complexes, phosphatases in the MAPK pathway, Rheb complexes, and TOR complexes.
+
+We also considered the sensitivity of CR cells following the application of the AR inhibitor enzalutamide in the presence of DHT (Fig. \ref{fg:Sensitivity}D). 
+In an androgen environment with enzalutamide, species which were more fragile included cytosolic AR, cPAcP, and p21Cip1. 
+As we would expect, AR species found in the nucleus and/or bound to coactivators, were more robust in the absence of enzalutamide. 
+The top two percent of sensitive species with and without enzalutamide were conserved. 
+In a CR cell, enzalutamide had no effect on the sensitivity of PI3K/AKT species as well as many MAPK species (ERK, Raf, and MEK). 
+Next, we looked at the effect of enzalutamide on a CR cell in both a non-androgen and DHT environment (Fig. \ref{fg:Supp_Sensitivity}). 
+More sensitive species in a non-androgen environment included dimerized HER2, ERK, and PAcP. 
+Species which were more robust in the non-androgen environment included, AR activated by DHT, AKT, p70, and AR bound to HSP.
+The results of our sensitivity analysis indicate that instead of inhibiting solely the AR pathway (enzalutamide), a combination therapy targeting the PI3K or MAPK pathways in addition to AR may be more effective.
+
+\subsection*{Robustness analysis identified key regulators of prostate cancer.} 
+Robustness analysis was conducted for 80 proteins to quantify the effects of amplifying or knocking down key model components in both AD and CR cells. 
+Gene expression parameters were altered by a factor 10, 0.5, and 0 for knock-in, knock-down, or knock-out perturbations, respectively.
+We calculated the effect of these perturbations on different protein markers, such as PSA, AR, and cyclin D. 
+The results of the model suggest that an inhibition of either the PI3K pathway or the MAPK pathway in combination with an AR inhibitor as a possible therapy for CRPC. 
+Sensitivity analysis revealed no change in the top sensitive species in the presence or absence of the AR inhibitor, enzalutamide.
+PI3K/AKT and MAPK species continued to fall in the top two percent of sensitive species.
+A study by Carver \textit{et al.} looked at dual inhibition of AR and PI3K signaling in LNCaP cells and in a Pten-deficient murine prostate cancer model \cite{Parker2013}. 
+Using both the PI3K inhibitor, BEZ235, and the AR inhibitor, MDV3100 (enzalutamide), the group saw a drastic decrease in the total number of cells. 
+Each inhibitor on it’s own had a much smaller effect on total cell number. 
+They saw an increase in the cell death marker, c-PARP, in the dual inhibition case. 
+The group hypothesized that AKT inhibition leads to increased AR signaling activity through increased protein concentrations of HER3. 
+On the other hand, AR inhibition leads to increased AKT activity due to the down regulation of PHLPP, a protein phosphatase that regulates AKT. 
+For simplicity, the HER3 pathway and also cell death were not included in the model. 
+Dual knock-out studies of PI3K and AR in our model show no additive effect on any cell cycle proteins through the dual knock-out. 
+The pathways in our model appear to be uncoupled and therefore no synergy is shown in the dual knock-out case. 
+This could indicate that the combined decrease in cell population is entirely due to cell death. 
+The Carver \textit{et al.} study did not consider cell cycle proteins or cell growth. 
+Our model does show a decrease in cell cycle proteins in the PI3K knock-out as well as in the dual knock-out case. 
+This result seems to be consistent with the decreased cell growth in the PI3K knock-out case which is not dependent on cell death, as c-PARP levels are low. 
+The decrease in cell cycle proteins is due to a decrease in general translation, including free eIF4E levels and activated 40S ribosome subunit. 
+The decrease in p70 (S6) activation due to inhibition of PI3K is shown in both the model and in the Carver \textit{et al.} study (Supplementary Figure), indicating this result is due to the PI3K pathway.
+It would be interesting to repeat the experiment looking at cell cycle proteins and also performing the experiment in CR LNCaP cells, instead of AD cells.
+
+A knock-out of Raf, MEK or ERK showed an overall increase in cyclin D levels in CR cells (Fig. \ref{fg:Supp_Robustness}). 
+This was unexpected and we saw a similar increase in cyclin D due to the knock-in of Raf, MEK or ERK. 
+We found that individual ensemble members showed different response to a Raf knock-out, in both cyclin D concentration and PSA concentration. 
+Of the 500 ensemble members, 126 members saw an increase in PSA concentration and 62 members saw an increase in cyclin D concentration due to the knock-out of Raf (Fig. \ref{fg:Robustness}). 
+We saw three distinct regions: (1) increase in PSA concentration, (2) increase in cyclin D concentration, and (3) a decrease in both PSA and cyclin D. 
+We explored the flux vectors of the outlying parameter sets to understand the mechanistic effect of Raf knock-out on PSA and cyclin D. 
+Outlying parameter sets in region 1 displayed high activation of PI3K through HER2 signaling as well as high association of AP1 with AR. 
+AP1 binds and suppresses AR transcriptional activity in LNCaP cells \cite{Sato1997}. 
+Knocking out Raf lowered AP1 levels and, therefore, freed AR for increased transcription of PSA.
+In region 2, parameter sets also had high activation of PI3K through HER2. 
+They also had higher association of E2F with Rb and cyclin D1a with AR. 
+Cyclin D levels in this region increase due to an increase in E2F levels caused by the Raf knock-out. 
+Parameters in region 3 have high association of TOR. 
+Interestingly, the drug sorafenib, a multi-kinase inhibitor that has activity against Raf, showed no measurable PSA decline in prostate cancer patients in clinical trials \cite{Dahut2008}. 
+The robustness analysis showed that network perturbation can result in unexpected responses due to cell-to-cell heterogeneity in gene expression. 
+These outlying cell types could be critical for understanding when designing drug targets and combination therapies.
 
 \section*{Discussion}
+In this study, we developed a population of mathematical models describing growth factor and hormone signal integration in androgen dependent, intermediate and resistant prostate cancer cells. 
+These models described the regulation of androgen receptor expression and activation through androgen binding as well as a ligand-independent, MAPK-driven mechanism referred to as the outlaw pathway. 
+An ensemble of model parameters was estimated using 43 steady-state and dynamic data sets taken from androgen dependent, intermediate and independent LNCaP cell lines using multiobjective optimization.
+Further, we tested the predictive power of the model by comparing model predictions against 33 novel data sets (including four \textit{in vivo} drug studies) not used during model training. 
+The model ensemble captured 84\% of the training data and 52\% of the validation data relative to 23\% and 24\% for a random control population. 
+Interestingly, during the the initial round of parameter estimation, we identified several potentially missing structural components not present in the original connectivity. 
+One such component, EGF-induced HER2/EGFR heterodimerization, was added to the current generation model. 
+Inclusion of this structural component significantly improved both training and validation performance using the same rate constants as the EGFR-homodimer case (no additional parameter fitting). 
+We then analyzed the population of signaling models, using both sensitivity and robustness analysis, to identify the critical components controlling network performance in a variety of conditions.
+
+%Hendrick's \textit{et al.} hypothesized that HER2 heterodimerization amplified signals by allowing just one molecule of EGF to activate two receptors \cite{Hendriks2003}. 
+%In our case, we see an amplified cyclin D concentration due to an equal amount of EGF stimulus. %
+%Model connectivity involving EGF and the downstream effect on AR and AR-regulated genes also came up as an open question. 
+%Whether EGF downregulates or upregulates AR activity is debatable and the biology is still unknown \cite{Cai2009,Cinar2005,Lin2003,Lin2001PNAS, Hakariya2006,Gregory2004,Ponguta2008,Guo2006,Weber2004}.
+%The current model assumed no direct effect of EGF on AR, and instead we saw increased AR activity through increased HER2 activation. 
+%Further studies will need to be done to update this section of the model.
+
+In addition, three of the validation cases involved the effect of EGF on AR and AR-activated genes, i.e., PSA. 
+Cai \textit{et al.} showed decreased expression of endogenous AR as well as androgen-regulated PSA in AD LNCaP cells due to an EGF stimulus \cite{Cai2009}. 
+Cinar \textit{et al.} also showed decreased AR protein expression due to EGF, an effect reversed by the mTOR inhibitor, Rapamycin \cite{Cinar2005}. 
+Model simulations show either the opposite trend or no effect due to EGF stimulus (Fig. \ref{fg:Validation}F) \cite{Cinar2005}. 
+These results suggest missing network structure. 
+From additional literature searches, the inhibition of AR activation through EGF is still an open question, with many groups debating the biology involved, predominately in the PI3K/AKT pathway. 
+Lin \textit{et al.} found that in low passage number LNCaP cells (C-33), AKT negatively regulates AR by destabilizing it and marking it for ubiquitylation. 
+In high passage number LNCaP (C-81), AKT levels are high which contribute to AR stability and less degradation \cite{Lin2003}. 
+Wen \textit{et al.} showed that HER2 could induce AKT activation and LNCaP cell growth in the presence and absence of androgen \cite{Wen2000}. 
+Another study shows AKT phosphorylation of AR at S213 and S790 suppresses AR transactivation and AR-mediated apoptosis of LNCaP \cite{Lin2001PNAS}. 
+The study from Cai \textit{et al.} showed the reduction in AR was not due to degradation or PI3K/AKT signaling, but instead was due to decreased AR mRNA levels \cite{Cai2009}. 
+They found that AR protein levels in CR cells were not affected by EGF. 
+Others though have found that PSA expression, even in C-81 cells, is decreased by EGF \cite{Hakariya2006}. 
+In other prostate cell lines, EGF has been shown to increase AR transactivation \cite{Gregory2004,Ponguta2008}. 
+The MAPK pathway, which is downstream of EGFR, may also enhance AR responses to low levels of androgen \cite{Guo2006,Weber2004}. 
+Due to the discrepancies in the literature, experiments should be performed before adding additional network connectivity to the model.
+
+The population of PCa models was analyzed using sensitivity analysis to identify key signaling components and processes in both AD and CR cells. 
+There was very little difference between sensitive and robust components in AD versus CR cells in the presence of androgen. 
+MAPK and PI3K pathway components were consistently ranked in the top 2\% of sensitive species in the presence of androgen for both AD and CR cells.
+On the other hand, cell cycle species, such as cyclin D-CDK4/6 complexes bound to cell cycle inhibitors (p27Kip1, p21Cip1, p16INK4), were consistently robust.
+However, this profile changed considerably in the absence of androgen.
+The activation of PI3K and ERK by HER2 dimerization and autophosphorylation was significantly more important in CR versus AD cells.
+Interestingly, AR activation by ERK was also more sensitive in CR versus AD cells in the absence of androgen.
+Lastly, although AR-regulated transcriptional processes were equally sensitive between the cell types, general translational and transcriptional components were more robust in CR versus AD cells.
+This evidence supports the current theory that CR cells will still respond to androgen and, thus, AR is still an active target in therapeutic against CRPC \cite{Karantanos2013}. 
+Supporting the argument that AR can be activated in the absence of androgens by MAPK activation \cite{Feldman2001}. 
+Advanced prostate cancers often have higher levels of E2F and other transcription factors \cite{Davis2006}.
+Interestingly, E2F was more sensitive in AD cells, while other transcription factors (ETS and AP1) were more robust. 
+The drug enzalutamide had no effect on the top 2\% of sensitive species. 
+Species in the PI3K/AKT and MAPK pathways in the presence of enzalutamide were still highly sensitive. 
+The application of enzalutamide, increased sensitivity of AR species found outside of the nucleus as well as PAcP species.
+Robustness analysis indicated diverse effects of Raf knock-out on PSA and cyclin D concentrations. 
+Clinical studies of sorafenib, a multi-kinase inhibitor that has activity against Raf, showed increase PSA levels in patients \cite{Dahut2008}. 
+Our results indicate that cell-to cell heterogeneity in gene expression can play a significant role in determining cell response. 
+Thus, combination therapies need to be considered even in the case of a Raf knock-out.
+
+The results of the model suggest that an inhibition of either the PI3K pathway or the MAPK pathway in combination with an AR inhibitor as a possible therapy for CRPC. 
+Sensitivity analysis revealed no change in the top sensitive species in the presence or absence of the AR inhibitor, enzalutamide. 
+PI3K/AKT and MAPK species continued to fall in the top two percent of sensitive species. 
+A study by Carver \textit{et al.} looked at dual inhibition of AR and PI3K signaling in LNCaP cells and in a Pten-deficient murine prostate cancer model \cite{Parker2013}. 
+Using both the PI3K inhibitor, BEZ235, and the AR inhibitor, MDV3100 (enzalutamide), the group saw a drastic decrease in the total number of cells. 
+Each inhibitor on it’s own had a much smaller effect on total cell number. They saw an increase in the cell death marker, c-PARP, in the dual inhibition case. 
+The group hypothesized that AKT inhibition leads to increased AR signaling activity through increased protein concentrations of HER3. 
+On the other hand AR inhibition leads to increased AKT activity due to the down regulation of PHLPP, a protein phosphatase that regulates AKT. 
+For the simplicity of this model, the HER3 pathway and also cell death were not included in the model. 
+Dual knock-out studies of PI3K and AR in our model show no additive effect on any cell cycle proteins through the dual knock-out. 
+The pathways in our model appear to be uncoupled and therefore no synergy is shown in the dual knock-out case. 
+This could indicate that the combined decrease in cell population is entirely due to cell death. 
+The Carver \textit{et al.} study did not consider cell cycle proteins or cell growth. 
+Our model does show a decrease in cell cycle proteins in the PI3K knock-out as well as in the dual knock-out case. 
+This result seems to be consistent with the decreased cell growth in the PI3K knock-out case which is not dependent on cell death, as c-PARP levels are low. 
+The decrease in cell cycle proteins is due to a decrease in general translation, including free eIF4E levels and activated 40S ribosome subunit. 
+The decrease in p70 (S6) activation due to inhibition of PI3K is shown in both the model and in the Carver \textit{et al.} study (Supplementary Figure), indicating this result is due to the PI3K pathway.
+It would be interesting to repeat the experiment looking at cell cycle proteins and also performing the experiment in CR LNCaP cells, instead of AD cells.
+
+%and  in the CR CWR-R1 prostate cell line.
+%a steroid receptor coactivator family member along with other important AR coregulators such as 
+%transcriptional intermediary factor 2 (TIF-2)
+
+The PCa signaling architecture was assembled after extensive literature review and hand curation of the biochemical interactions.
+However, there are a number of areas where model connectivity could be refined, e.g., the regulation of AR phosphorylation.
+We assumed a single canonical activating AR phosphorylation site (S515), with ERK being the major kinase and PP2A or PP1 being the major phosphatases responsible for regulating this site.  
+MAPK activation following EGF treatment increases AR transcription and cell growth, partially through AR phosphorylation on MAPK consensus site S515 \cite{Ponguta2008}.
+However, there are at least 13 phosphorylation sites identified on AR, with phosphorylation at six of these being androgen induced \cite{Gioeli2012}. 
+Moreover, other kinases such as AKT, protein kinase C (PKC) family members, as well as Src-family kinases can all phosphorylate AR in prostate cells \cite{Guo2006,Ponguta2008}. 
+For example, AKT activation leads to AR  phosphorylation at both S213 and S791 (however, the role of these sites remains unclear) \cite{Wen2000,Lin2001PNAS,Taneja2005,Lin2003}.  
+AKT effects on AR may also be passage number dependent, with AKT repressing AR transcription in low passage number cells and enhancing transcription in higher passage numbers \cite{Lin2003}. 
+Androgen independent phosphorylation of AR by Src family kinases (not currently in model) at Y534 \cite{Guo2006} or by protein kinase C (PKC) family members at the consensus site S578 could also be important for understanding the regulation of AR activity.
+A second area we will revisit is the gene expression program associated with androgen action, and particularly the role of AR coregulators.
+Currently, we included only two AR coactivators, cyclin E and CDK6 \cite{Yamamoto2000,Lim2005} and three corepressors AP1, Cdc25A, and cyclin D1a in the model \cite{Sato1997,Chiu2009,Petre-Draviam2003}.
+However, there are at least 169 proteins classified as potential AR coregulators \cite{Heinlein:2002qy,Heemers2007} with many of these being differentially expressed in malignant cells.
+For example, the expression of steroid receptor coactivator-1 (Src-1) and transcriptional intermediary factor 2 (Tif-2), both members of the steroid receptor coactivator family, are elevated in prostate cancer \cite{Gregory2001,Gregory2004}.
+Src-1 is phosphorylated by MAPK and interacts directly with AR to enhance AR-mediated transcription \cite{Heemers2007}. 
+Another class of potentially important AR coregulators are the cell cycle proteins Cdc25 and Rb.
+Unlike Cdc25A, Cdc25B (not in the model) can act as an AR coactivator leading to enhanced AR transcription activity \cite{Ngan2003}. 
+The Rb protein, in addition to being a key cell cycle regulator, has been shown to be an AR coactivator in an androgen-independent manner in DU145 cells \cite{Yeh1998}.
+However, there is some uncertainty about the role of Rb as Sharma \textit{et al.} showed that Rb decreased AR activation in multiple prostate cancer cell lines and xenografts \cite{Sharma2010}. 
+Forkhead proteins have also been shown to activate as well as repress AR function. 
+In prostate cancer, AKT suppresses AFX/Forkhead proteins, which diminishes expression of AFX target genes, such as p27Kip1 \cite{Graff2000, Brunet1999, Medema2000, Takaishi1999}. 
+Lastly, undoubtedly there are several other signaling axes important in PCa, such as cytokine or insulin- and insulin-like growth factor signaling \citep{Cardillo2003,Heinlein2004,Tam2007,Seaton2008}.
+Understanding the pathways associated with these signals and how they relate to the current model, may give us a more complete picture of CR prostate cancer.
+
+%Lastly, with an extended model, it would be beneficial to include possible biomarkers for drug outcomes, as currently there is no biomarker that is used to select treatments for patients  \cite{Sartor2013}.
 
@@
 
-\subsection*{Formulation and solution of coagulation model equations.}
-\subsection*{Optimization methods.}
+\subsection*{Prostate model signaling architecture.}
+We modeled the expression, translation and post-translational modifications of key components of the signaling architecture.  
+The model, which consisted of 780 protein, lipid or mRNA species interconnected by 1674 interactions, was a significant extension to our previous model \cite{Tasseff2010} in several important areas. 
+First, we included well-mixed nuclear, cytosolic, membrane and extracellular compartments (including transfer terms between compartments).
+Next, we expanded the description of growth factor receptor signaling, considering both homo- and heterodimer formation between ErbB family members and the role of cellular and secreted prostatic acid phosphatase (cPAcP and sPAcP, respectively).
+Both forms of PAcP were included because cPAcP downregulates HER2 activity, while sPAcP promotes modest HER2 activation \cite{Veeramani2005}. 
+Third, we expanded the description of the G1/S transition of the cell cycle (restriction point). 
+The previous model used the abundance of cyclin D as a proliferation marker, but did not include other proteins or interactions potentially important to the restriction point.
+Toward this shortcoming, we included cyclin E expression (and its role as a coregulator of androgen receptor expression), enhanced the description of cyclin D expression and the alternative splicing of cyclin D mRNA (including the role of the splice variants in androgen action), included the Rb/E2F pathway as well as E2F inhibition of androgen receptor expression \cite{Davis2006}, and the cyclin-dependent kinases cyclin-dependent kinase 4 (CDK4) and cyclin-dependent kinase 6 (CDK6).
+We also included key inhibitors of the restriction point including cyclin-dependent kinase inhibitor 1 (p21Cip1), cyclin-dependent kinase inhibitor 1B (p27Kip1), and cyclin-dependent kinase inhibitor 2A (p16INK4) \cite{Sherr1999}.
+Fourth, we enhanced the description of growth factor induced translation initiation.
+One of the key findings of the previous model was that growth factor induced translation initiation was globally sensitive (important in both androgen dependent and independent conditions). 
+However, the description of this important subsystem was simplified in the previous model. 
+Here, we expanded this subsystem, using connectivity similar to previous study of Lequieu \textit{et al.} \cite{Lequieu:2011eu}, and re-examined the importance of key components of this axis, such as mammalian target of rapamycin (mTOR), phosphatidylinositide 3-kinase (PI3K) and AKT.
+Lastly, we significantly expanded the description of the role of androgen receptor. 
+The previous model assumed constant AR expression, consistent with studies in androgen dependent and independent LNCaP sublines \cite{Lee2003}.
+However, other prostate cancer cell lines vary in their AR expression \cite{Sobel2005}. 
+Thus, to capture androgen signaling in a variety of prostate cancer cells, we included the transcriptional regulation governing androgen receptor expression, updated our description of the regulation of androgen receptor activity and androgen action (gene expression program driven by activated androgen receptor).
+At the expression level, we included AR auto-regulation in combination with the co-activators cyclin E and CDK6 \cite{Yamamoto2000,Lim2005}. 
+We also assumed androgen receptor could be activated through androgen binding or a ligand-independent, MAPK-driven mechanism referred to as the outlaw pathway \cite{Feldman2001,Yeh1999}. 
+We assumed a single canonical activating AR phosphorylation site (S515), with phosphorylated extracellular-signal-regulated kinase 1/2 (ppERK1/2) being the major kinase and protein phosphatase 2 (PP2A) or phosphoprotein phosphatase 1 (PP1) being the major phosphatases responsible for regulating this site.
+Finally, we modeled androgen receptor induced gene expression, including prostate specific antigen (PSA), cPAcP and p21Cip1.
+
+%This description includes  were added to the model because of their importance in co-regulating AR transcription regulation activity in prostate cells . 
+%With the addition of cell cycle proteins, we also included 
+%only considered cyclin D as a cell marker and 
+%In both models we considered nuclear hormone (AR) and transmembrane growth factor receptor (homo- and heterodimers of ErbB1/2) activation through both ligand-dependent and independent mechanisms. 
+%The model investigated the activation of AR through androgen binding or a ligand-independent, MAPK-driven mechanism. 
+%In addition to MAPK-dependent outlaw activation of AR, the model also included transcriptional activation via MAPK signaling \cite{Feldman2001,Yeh1999}. 
+%The PI3K/AKT/mTOR pathway for translation initiation was also included in the model, as well as the regulated expression of PSA, cyclin D,  
+%Thus, 
+%The model of Tasseff \textit{et al.}  
+%In order 
+%Additionally, expression of transcription factors AP-1 and ETS were also added to the model. 
+%Another significant addition to the model was the inclusion of a nuclear compartment separate from the cytosol, membrane and extracellular environments. 
+%The prostate model was developed to determine the shift in important signaling components between CR and AD prostate cancer cells.
+
+\subsection*{Formulation and solution of the model equations.}
+The prostate model was formulated as a coupled set of non-linear ordinary differential equations (ODEs):
+\begin{equation}\label{asseassertain_massbalance}
+\frac{d\mathbf{x}}{dt}=\mathbf{S}\cdot\mathbf{r}\left(\mathbf{x},\mathbf{k}\right)\qquad\mathbf{x}\left(t_{o}\right)=\mathbf{x}_{o}
+\end{equation}
+The quantity $\mathbf{x}$ denotes the vector describing the abundance of protein, mRNA, and other species in the model ($780\times 1$).
+The stoichiometric matrix $\mathbf{S}$ encodes the signaling architecture considered in the model ($780\times 1674$).  
+Each row of $\mathbf{S}$ describes a signaling component while each column describes a particular interaction.
+The $(i,j)$ element of $\mathbf{S}$, denoted by $\sigma_{ij}$, describes how species $i$ is involved with interaction $j$. 
+If $\sigma_{ij}>0$, species $i$ is produced by interaction $j$. Conversely, If $\sigma_{ij}<0$, then species $i$ is consumed in interaction $j$. 
+Lastly, if $\sigma_{ij}=0$, then species $i$ is not involved in interaction $j$.
+The term  $\mathbf{r}\left(\mathbf{x},\mathbf{k}\right)$ denotes the vector of interactions rates ($1674\times{1}$).
+Gene expression and translation processes as well as all biochemical transformations were decomposed into simple elementary steps, 
+where all reversible interactions were split into two irreversible steps (supplemental materials). 
+We modeled each network interaction using elementary rate laws where all reversible interactions were split into two irreversible steps. 
+Thus, the rate expression for interaction $q$ was given by:
+\begin{equation}\label{eq-mass-action}
+r_{q}\left(\mathbf{x},k_{q}\right)=k_{q}\prod_{j\in\left\{\mathbf{R}_{q}\right\}}x_{j}^{-\sigma_{jq}}
+\end{equation}
+The set $\left\{\mathbf{R}_{q}\right\}$ denotes reactants for reaction $q$, while $\sigma_{jq}$ denotes the stoichiometric coefficient (element of the matrix $\mathbf{S}$) governing species $j$ in reaction $q$.
+The quantity $k_{q}$ denotes the rate constant (unknown) governing reaction $q$.
+Model equations were generated in the C-programming language using the UNIVERSAL code generator, starting from an text-based input file (available in supplemental materials). 
+UNIVERSAL, an open source Objective-C/Java code generator, is freely available as a Google Code project (http://code.google.com/p/universal-code-generator/). 
+Model equations were solved using the CVODE solver in the SUNDIALS library \cite{Hindmarsh2005} on an Apple workstation (Apple, Cupertino, CA; OS X v10.6.8).
+
+We ran the model to steady-state before calculating the response to DHT or growth factor inputs.
+The steady-state was estimated numerically by repeatedly solving the model equations and estimating the difference between subsequent time points:
+\begin{equation}
+\Vert\mathbf{x}\left(t+\Delta{t}\right) - \mathbf{x}\left(t\right)\Vert_{2}\leq\gamma
+\end{equation}
+The quantities $\mathbf{x}\left(t\right)$ and $\mathbf{x}\left(t+\Delta{t}\right)$ denote the simulated abundance vector at time $t$ and $t + \Delta{t}$, respectively.
+The $L_{2}$ vector-norm was used as the distance metric, where $\Delta{t} = 100$ hr of simulated time and $\gamma$ = 0.001 for all simulations.
+
+We estimated an ensemble of model parameter sets using the Pareto Optimal Ensemble Techniques (POETs) multiobjective optimization routine \cite{Song:2009dq,Song2010,Lequieu:2011eu}. 
+POETs minimized the residual between model simulations and 43 separate training objectives taken from protein and mRNA signaling data generated in androgen dependent, intermediate and independent LNCaP cell lines (Table \ref{objective_table}).
+From these training objectives, POETs generated $>10^6$ candidate parameter vectors from which we selected N = 5000 Pareto rank-zero vectors for further analysis. 
+The set-to-set correlation between selected sets was approximately 0.60, suggesting only modest similarity between ensemble members.
+Approximately 33\%, or 560 of the 1674 parameters had a coefficient of variation (CV) of less than 1.0, where the CV ranged from 0.59 to 5.8 over the ensemble.
+Details of the parameter estimation problem and POETs are given in the supplemental materials.
+
+\subsection*{Sensitivity and robustness analysis.}
+Steady-state sensitivity coefficients were calculated for N = 500 parameter sets selected from the ensemble by solving the augmented kinetic-sensitivity equations \cite{Dickinson1976}:
+\begin{equation}
+	\begin{bmatrix}
+		\mathbf{S}\cdot\mathbf{r}(\mathbf{x},\mathbf{k}) \\
+		\mathbf{A}(t_s)\mathbf{s}_j + \mathbf{b}_j(t_s)
+	\end{bmatrix} = \begin{pmatrix}
+		\mathbf{0} \\
+		\mathbf{0}
+	\end{pmatrix}
+	   \qquad j = 1,2,\dots,\mathcal{P}
+\end{equation}
+where
+\begin{equation}
+s_{ij}(t_s)= \left. \frac{\partial{x_i}}{\partial{k_j}} \right|_{t_s}
+\end{equation} for each parameter set.
+Steady-state was calculated as described previously.
+The quantity $j$ denotes the parameter index, $\mathbf{A}$ denotes the Jacobian matrix, and $\mathcal{P}$ denotes the number of parameters in the model. 
+The vector $\mathbf{b}_j$ denotes the $j$th column of the matrix of first-derivatives of the mass balances with respect to the parameters. 
+Steady-state sensitivity coefficients were used because of the computational burden associated with sampling several hundred parameters sets for each of the 1674 parameters. 
+The steady-state sensitivity coefficients $\mathcal{N}_{ij} \equiv s_{ij}$ were organized into an array for each parameter set in the ensemble: 
+\begin{equation}
+	\mathcal{N}^{\left(\epsilon\right)} = 
+	\begin{pmatrix}
+		\mathcal{N}_{11}^{\left(\epsilon\right)} & \mathcal{N}_{12}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{1j}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{1P}^{\left(\epsilon\right)} \\
+		\mathcal{N}_{21}^{\left(\epsilon\right)} & \mathcal{N}_{22}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{2j}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{2P}^{\left(\epsilon\right)} \\
+		\vdots & \vdots & & \vdots & & \vdots \\
+		\mathcal{N}_{M1}^{\left(\epsilon\right)} & \mathcal{N}_{M2}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{Mj}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{MP}^{\left(\epsilon\right)} \\
+	\end{pmatrix}\qquad\epsilon = 1,2,\hdots,N_{\epsilon}
+\end{equation}
+where $\epsilon$ denotes the index of the ensemble member, $P$ denotes the number of parameters, $N_{\epsilon}$ denotes the number of parameter sets sampled (N = 500) and $M$ denotes the number of model species.
+To estimate the relative fragility or robustness of species and reactions in the network, we decomposed $\mathcal{N}^{\left(\epsilon\right)}$ 
+using Singular Value Decomposition (SVD):
+
+\begin{equation}
+	\mathcal{N}^{\left(\epsilon\right)} = \mathbf{U}^{\left(\epsilon\right)}\Sigma^{\left(\epsilon\right)}\mathbf{V}^{T,{\left(\epsilon\right)}}
+\end{equation}
+Coefficients of the left singular vectors corresponding to largest $\theta$ $\leq$ 15 singular values of $\mathcal{N}^{\left(\epsilon\right)}$ were rank-ordered to estimate important species combinations, while coefficients of the right singular vectors were used to rank important reaction combinations.
+Only coefficients with magnitude greater than a threshold ($\delta$ = 0.001) were considered. 
+The fraction of the $\theta$ vectors in which a reaction or species index occurred was used to quantify its importance (sensitivity ranking). 
+We compared the sensitivity ranking between different conditions to understand how control in the network shifted in different cellular environments.
+
+Robustness coefficients were calculated as shown previously \cite{Tasseff2011}. 
+Robustness coefficients denoted by $\alpha\left(i,j,t_{o},t_{f}\right)$ are defined as:
+\begin{equation}
+\alpha\left(i,j,t_{o},t_{f}\right)=\left(\displaystyle\int_{t_{o}}^{t_{f}}x_{i}\left(t\right)dt\right)^{-1}\left(\displaystyle\int_{t_{o}}^{t_{f}}x^{(j)}_{i}\left(t\right)dt\right)
+\end{equation}
+Robustness coefficients quantify the response of a marker to a structural or operational perturbation to the network architecture.
+Here $t_o$ and $t_f$ denote the initial and final simulation time respectively, while $i$ and $j$ denote the indices for the marker and the perturbation respectively. 
+A value of $\alpha\left(i,j,t_{o},t_{f}\right)>1$, indicates increased marker abundance, while $\alpha\left(i,j,t_{o},t_{f}\right)<1$ indicates decreased marker abundance following perturbation $j$. 
+If $\alpha\left(i,j,t_{o},t_{f}\right)\sim{1}$ the $j$th perturbation does not influence the abundance of marker $i$.
+Robustness coefficients were calculated (starting from steady-state) from $t_{o} = 0$ hr to $t_{f} = 72$ hr following the addition of 10nM DHT at $t_{o}$.
+Robustness coefficients were calculated for the same N = 500 models selected for sensitivity analysis.
 
 \section*{Acknowledgements}
-This study was supported by the National Science Foundation GK12 award (DGE-1045513) 
-and by the National Science Foundation CAREER award (FILLMEIN).
+This study was supported by award number U54CA143876 from the National Cancer Institute.
+The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Cancer Institute or the National Institutes of Health.
+This study was also supported by the National Science Foundation GK12 award number DGE-1045513.
 
 \clearpage
+%\bibliographystyle{plain}
+%\bibliographystyle{IEEEbib}
 
@@
 %\begin{comment}
+
+\clearpage
+
+\begin{figure}\centering
+%\includegraphics[width = 7in]{Figure1_LNCAPModel.pdf}
+\caption{Schematic overview of the prostate signaling network.
+The model describes hormone and growth factor induced expression of several proteins, including PSA.
+In the absence of outside hormones/growth factors, overactive HER2 can stimulate the MAPK and AKT pathways.
+AR can be activated directly by the MAPK pathway.
+%Addition of DHT leads to androgen receptor binding and ultimately increased transcription and proliferation.
+HER2 stimulate the MAPK and AKT pathway in the absence of DHT.
+EGF binding leads to EGFR dimerization and activation of the MAPK and AKT pathways.
+AR can be activated directly by the MAPK pathway.
+}
+\label{fg:NetworkSchematic}
+\end{figure}
+
+\begin{figure}\centering
+%\includegraphics[width = 7in]{Figure3_ColorPlot.pdf}
+\caption{Simulation results versus experimental results for training and validation data.
+Experiment numbers 1 through 43 were used for training, while experiments 44 through 72 were validation.
+Gray means the ensemble member qualitatively fit experimental data in both models.
+White means the the ensemble member only fit the data using the new model that included HER2 heterodimerization.
+Red means the ensemble member fit using only the old model.
+Black corresponds to an incorrect cellular response in both models.
+$\bf{A.,C.}$ Training and validation results, respectively, for entire ensemble population using both the original model and an updated model including HER2 heterodimerization (N = 5000).
+$\bf{B.,D.}$ Simulation results for training and validation of a random set of 100 members using both models.
+ }
+\label{fg:ColorPlot}
+\end{figure}
+
+\begin{figure}\centering
+%\includegraphics[width = 7in]{Figure2_Training.pdf}
+\caption{Ensemble performance against selected training objectives (N = 5000).
+A, B.
+Time course data for PSA concentration due to a stimulus of 10 nM DHT in LNCaP C33 cells and LNCaP C81 cells, respectively (O2, O3).
+C.
+PSA levels in the presence and absence of a HER2 inhibitor (LNCaP C81 cells, O7).
+D.
+PAcP mRNA levels at 72 hours in the presence and absence of DHT (LNCaP C51 cells, O14).
+E.
+Steady-state cyclin D levels in LNCaP C33 vs.
+C81 (O17).
+F.
+p21Cip1 levels at 48 hrs in the presence and absence of DHT (LNCaP C33, O25).
+G.
+Steady-state AKT phosphorylation levels in LNCaP C33 vs.
+C51 (O30).
+H.
+AR levels at 24 hours in the presence and absence of DHT (LNCaP C51, O31).
+I.
+AR mRNA levels in the presence and absence of E2F over expression (LNCaP C33, O34).
+}
+\label{fg:Training}
+\end{figure}
+
+\begin{figure}\centering
+%\includegraphics[width = 7in]{Figure4_Validation.pdf}
+\caption{Blind model predictions for the ensemble (N = 5000).
+The model ensemble’s predictive ability was assessed by comparing simulation versus experimental data not used for training.
+A.
+Time course data for HER phosphorylation due to a stimulus of 10 nM DHT (LNCaP C33, P1).
+B.
+ERK phosphorylation levels in the presence and absence of a PAcP inhibitor (LNCaP C33 cells, P3).
+C.
+AR levels at 24 hrs in varying levels of DHT (LNCaP C33, P17).
+D.
+Shc phosphorylation levels at 24 hrs in the presence and absence of DHT (LNCaP C33, P22).
+E.
+Time course data for HER phosphorylation due to a stimulus of 1.6 nM EGF (LNCaP C33, P7).
+F.
+AR levels in varying levels of EGF (LNCaP C33, P14).
+G, H, I.
+Fold change in PSA concentration due to drug stimulus: enzalutamide, lapatinib, and sorafenib.}
+\label{fg:Validation}
+\end{figure}
+
+\begin{figure}\centering
+%\includegraphics[width = 7in]{Figure5_Sensitivity.pdf}
+\caption{Sensitivity analysis of a population of prostate models (N = 500).
+Species with a low sensitivity are considered robust, while species with a high sensitivity ranking are considered fragile.
+$\bf{A,B.}$ Sensitivity ranking of network species in AD versus CR cells in the absence (presence) of DHT.
+$\bf{C.}$ Sensitivity ranking of network species in AD cells in the absence and presence of DHT.
+$\bf{D.}$ Sensitivity ranking of network species in CR cells in the presence and absence of enzalutamide with a DHT stimulus.
+ }
+\label{fg:Sensitivity}
+\end{figure}
+
+\begin{figure}\centering
+%\includegraphics[width = 7in]{Figure6_RAF_K0_C81.pdf}
+\caption{Robustness analysis of a population of CR prostate models with Raf knock-out (N = 500).
+A log fold change of greater then zero implies that the concentration of the protein increased with the knock-out of Raf, while a log fold change of less then zero indicates that the concentration of protein decreased.
+A log of fold change equal to 0, shows no response due to Raf knock-out.
+Three distinct regions emerge in Raf knock-out case: (1) PSA increases, (2) cyclin D concentration increases, and (3) PSA and cyclin D concentration decrease.
+}
+\label{fg:Robustness}
+\end{figure}
 
@@
 
+\section*{Supplementary materials}
+
+\input{Prostate_Obj_Table}
+\clearpage
+
+\input{Prostate_Pred_Table}
+\clearpage
+
+\begin{figure}\centering
+%\includegraphics[width = 7in]{Supp_Figure_Sensitivity.pdf}
+\caption{Sensitivity analysis of a population of prostate models (N = 500).
+Species with a low sensitivity are considered robust, while species with a high sensitivity ranking are considered fragile.
+$\bf{A}$ Sensitivity ranking of network species in CR cells in the absence and presence of DHT.
+$\bf{B.}$ Sensitivity ranking of network species in CR cells in the presence and absence of enzalutamide in the absence of a DHT stimulus.
+$\bf{C.,D.}$ Sensitivity ranking of network species in CR cells in the presence and absence of sorafenib and lapatinib, respectively, with a DHT stimulus.}
+\label{fg:Supp_Sensitivity}
+\end{figure}
+
+\begin{figure}\centering
+%\includegraphics[width = 7in]{Supp_Figure_Robustness.pdf}
+\caption{Robustness analysis of protein markers.
+Expression level of key proteins was altered by a factor of 2, 0.1, or 0 (knock-in, knock-down, or knock-out) and robustness coefficients were calculated for five key protein markers.
+Simulations shown were from CR cells, with indicated perturbation.
+Mean of 500 ensemble members is shown.
+}
+\label{fg:Supp_Robustness}
+\end{figure}
+
+%\end{comment}
+
 \end{document}

### 2014-07-21 Supplemental.tex
First commit on Publications repository

\subsection*{Estimation of a population of models using Pareto Optimal Ensemble Techniques (POETs).}

We used multiobjective optimization to estimate an ensemble of prostate models. Although computationally more complex than single-objective formulations, multiobjective optimization can be used to address qualitative conflicts in training data arising from experimental error or cell-line artifacts \cite{Handl2007}. In this study we used the Pareto Optimal Ensemble Technique (POETs) to perform the optimization. POETs integrates standard search strategies, e.g., Simulated Annealing (SA) or Local Pattern Search (PS) with a Pareto-rank fitness assignment \cite{Song2010}. The mean squared error, $\eta$, of parameter set $\mathbf{k}$ for training objective $j$ was defined as:

\begin{equation}
\eta_j(\mathbf{p}_k) = \frac{1}{N}\sum_{i}^{N}\frac{(\hat{x}_{i,j} - \beta_j x(\mathbf{p_k})_{i,j})^2}{\hat{\sigma}^2_{i,j}}
\end{equation}
The symbol $\hat{x}_{i,j}$ denotes scaled experimental observations (from training objective j) while $x(\mathbf{p_k})_{i,j}$ denotes the simulation output (from training objective j). The quantity $i$ denotes the sampled time-index or condition, and $N$ denotes the number of time points or conditions for experiment j. The standard deviation, $\hat{\sigma}_{i,j}$, was assumed to be equal to 10\% of the reported observation, if no experimental error was reported. $\beta_j$ is a scaling factor which is required when considering experimental data that is accurate only to a multiplicative constant. In this study, the experimental data used for training and validation was typically band intensity from immunoblots, where intensity was estimated using the ImageJ software package \cite{Abramoff2004}. The scaling factor used was chosen to minimize the normalized squared error \cite{Brown2003}:

\begin{equation}
\beta_j = \frac{\sum_{i}(\hat{x}_{i,j}x_{i,j}/\hat{\sigma}_{i,j}^2)}{\sum_{i}(x_{i,j}/\hat{\sigma}_{i,j})^2}
\end{equation} 
By using the scaling factor, the concentration units on simulation results were arbitrary, which was consistent with the arbitrary units on the experimental training data. All simulation data was scaled by the corresponding $\beta_j$.

We computed the Pareto rank of parameter set $\mathbf{k}_{i+1}$ by comparing the simulation error at iteration $i+1$ against the simulation archive, denoted as $\mathbf{K}_i$. We used the Fonseca and Fleming ranking scheme \cite{Fonseca1993} to estimate the rank of the parameter set $\mathbf{k}_{i+1}$. Parameter sets with increasing rank are progressively further away from the optimal trade-off surface. The parameter set $\mathbf{k}_{i+1}$ was accepted or rejected by the SA with probability $\mathcal{P}\left(\mathbf{k}_{i+1}\right)$: 
\begin{equation}\label{eqn_costMOSA}
\mathcal{P}(\mathbf{k}_{i+1}) \equiv \exp{\{-rank\left(\mathbf{k}_{i+1} \mid \mathbf{K}_{i} \right)/T\}}
\end{equation}
where $T$ is the computational annealing temperature. The Pareto rank for $\mathbf{k}_{i+1}$ is denoted by $rank\left(\mathbf{k}_{i+1}\mid \mathbf{K}_{i}\right)$. The annealing temperature was  adjusted according to the schedule
$T_k = \beta^k T_0$
where $\beta$ was defined as $\beta = \left(\frac{T_{f}}{T_{o}}\right)^{1/10}$. The initial temperature was given by $T_0 = n/log(2)$, with $n=4$ and the final temperature $T_f = 0.1$ used in this study.
The epoch-counter $k$ was incremented after the addition of 50 members to the ensemble. As the ensemble grew, the likelihood of accepting a high rank set decreased. Parameter sets were generated by applying a random perturbation in log space:

\begin{equation}
\log\mathbf{k}_{i+1} = \log\mathbf{k}_i + \mathcal{N}(0,\nu)
\end{equation}
where $\mathcal{N}(0,\nu)$ is a normally distributed random number with zero mean and variance $\nu$, set as 0.1 in this model. The perturbation was applied in log space to account for large variation in parameter scales and to ensure positive parameter values. We used a local pattern search every $q$ steps, in our case 20, to minimize error for a single randomly selected objective. The local pattern-search algorithm used has been described previously \cite{Gadkar2003}.

### 2014-07-21 Build.sh

@@ -1,0 +1,10 @@
+#!/bin/sh
+
+# Tex this mofo -
+latex $1.tex
+bibtex $1
+latex $1.tex
+latex $1.tex
+
+# Make the pdf -
+dvipdfm -p "letter" $1.dvi

### 2014-07-21 Figures_v1.tex
First commit

\begin{figure}\centering
\epsfig{file=./Main_Figures/Pathway_F1.eps,width=0.8\textwidth}
	\caption{An array of cellular stressors can perturb the folding environment in the endoplasmic reticulum (ER) leading to unfolded or misfolded protein. In response to the folding imbalance, cells initiate the cytoprotective unfolded protein response (UPR). The problem of unfolded or misfolded proteins in the ER is addressed by increasing the folding capacity through the up-regulation of the expression of chaperone proteins, attenuating translation by regulating eIF2$\alpha$, and promoting the degradation of misfolded proteins through ER-associated degradation (ERAD). If UPR is unable to restore the folding balance, ER stress will eventually lead to apoptotic cell-death. The three signal transduction pathways mediating the unfolded protein response in higher eukaryotes. First, the PRKR-like ER kinase (PERK) pathway is initiated after BiP dissociation from PERK. While PERK transduces both pro- and anti-apoptotic signals, its main function is translation attenuation through the phosphorylation of eIF2$\alpha$. Next, the activating transcription factor 6 (ATF6) pathway is activated following BiP dissociation. ATF6 induces the expression of chaperones e.g., BiP as well as apoptosis effectors such as CHOP. Lastly, the inositol-requiring kinase 1 (IRE1) pathway is activated following BiP dissociation from IRE1. Activated IRE1 has both an endoribonuclease and a serine-threonine kinase activity that drive can pro-apoptotic signals. Inset: The UPR network consisted of 636 protein or mRNA species interconnected by 1090 interactions.}
	\label{fg:fig1}
\end{figure}

\begin{figure}\centering
\epsfig{file=./Main_Figures/OBJ_F2.eps,width=1.0\textwidth}
	\caption{Objective function plot for selected training constrains (O1,O2,$\hdots$O33) for the UPR model population generated using POETs. Points denote separate models in the population. Several objectives exhibit clear Pareto fronts, e.g., O29 $\times$ O25. This suggests an inability to model both training constraints simultaneously or conflicts in the training data.}
	\label{fg:objective_plot}
\end{figure}

\begin{figure}\centering
\epsfig{file=./Main_Figures/TRPR_F3.eps,width=1.0\textwidth}
	\caption{Simulations versus experimental data for selected objective functions following exposure to the ER-stress inducers Thapsigargin (Tg or Thaps) or Tunicamycin (TM). The first-column (A - D) denotes adaptation components, the second column (E - H) denotes alarm phase components, while the third column (I - L) denotes apoptosis phase components. Bars denote the scaled mean concentration computed over the ensemble, while the error bars describe one standard error.}
	\label{fg:simulation_comparison}
\end{figure}

\begin{figure}\centering
\epsfig{file=./Main_Figures/Proof_F4.eps,width=1.0\textwidth}
	\caption{Proof of concept simulation unfolded protein response activation. 
	\textbf{A:}UPR induction was controlled by manipulating the generation rate of unfolded or misfolded protein (qP) in the ER compartment. A step-change in qP from qP = 0.1 to qP = 100 was issued at approximately t = 0.1 hrs and then adjusted back to qP = 0.1 at t = 20 hrs.
	\textbf{B:}Flux through the PERK, ATF6 and IRE1 stress sensing branches as a function of time following a step change in misfolded protein generation.
	\textbf{C:}Simulated expression profile for the 59 genes in the model. The symbol UPROT denotes the level of unfolded protein.}
	\label{fg:proof_of_concept_simulation}
\end{figure}

\begin{figure}\centering
\epsfig{file=./Main_Figures/Flux_F5.eps,width=0.8\textwidth}
	\caption{Cross plot of the fluxes at P1-P5 as denoted in Figure 4: We tried to see is how the system behaves and how the system can recuperate from UPR dose when it is in the adaptation phase as compared to the apoptosis phase. (D) As compared to P1 (No-UPR Steady State), we see that early on at 1 hr after UPR dose there is a marked increase in ATF4 and CHOP regulation, ATF6 signaling along with unfolded protein sensing and degradation. These are hallmarks of the adaptation-alarm phase of the UPR response. (A) If we continue with the dose of UPR till around 25 hrs, we see the fluxes reach a steady state. This state is marked by increased BiP regulation, enhanced ATF4 transcriptional activity, increased mitochondrial membrane permeability and increased apoptotic fluxes. This state is similar to the Apoptotic phase of UPR, where in the cell has committed itself to apoptosis mediated cell death. (B) and (C) If we reduce the UPR load after the cell has committed to apoptosis (as in P3), we find that the cell continues to function similar to the UPR state even upon UPR load reduction after 25 hrs. There are certain aspects which are seen to reduce like IRE1-TRAF2 signaling, ASK1 activation. However not much difference is seen in terms of apoptotic fluxes, denoting the cell has committed itself to death and is in a point of no return. (E)-(F) On the contrary if we reduce the load of UPR in the adaptation-alarm phase (P4), we see that the cell can recuperate using its ERAD machinery and the regulation of BiP.}
	\label{fg:cross_plot_flux}
\end{figure}

\begin{figure}\centering
\includegraphics[width = 6.0in,height=5.5in]{./Main_Figures/Sens_F1.pdf}
	\caption{Rank-ordering of species sensitivities in the presence of UPR as a function of time.
	\textbf{Inset:} Rank-ordering of parameter sensitivity for UPR-induced versus normal conditions.
Points denote the mean ranking computed over N = 5 parameter sets from the model population, while error bars denote one standard deviation. Points are color-coded based based upon biological function.}
	\label{fg:fig5}
\end{figure}

\begin{figure}\centering
\epsfig{file=./Main_Figures/Rob_F7.eps,width=0.8\textwidth}
	\caption{Robustness analysis of the UPR network: A-B Phenotypic phase plane analysis for the the UPR model following structural perturbations. Coupling coefficients (area under the curve from the simulation with species removed dived by the wild-type simulation) for all 636 model species were calculated for the nominal parameter set following gene overexpression/knockout (A) and deletion of single network edges (B). Coupling coefficients of one indicate no change in a marker level following a perturbation, while values less (greater) than one denote decreased (increases) marker levels. C Structural distinguishability analysis: We computed the dendrogram of the coupling coefficients for single GKO of model species. Individual coupling coefficients were clustered, where the euclidean norm was used as the distance metric and the linkage function was the inner square product (variance minimization algorithm). Each additional cluster was chosen to reduce the overall variance (y-axis). A general description of the biological function of the clusters were indicated by each group. \textbf{Insets:} Distinguishability as the magnitude of the orthogonal components for all knockout species.  Species were ordered from largest to smallest magnitudes.  Red markers indicate species which were statistically significant.}
	\label{fg:robustness_figure}
\end{figure}

% Supplementary Figures over here.. 
\setcounter{figure}{0}
\makeatletter
\makeatletter \renewcommand{\fnum@figure}
{\figurename~S\thefigure}
\makeatother

\begin{figure}\centering
	\epsfig{file=./Supp_Figures/Para_S1.eps,width=1.0\textwidth}
	\caption{ POETs generated an ensemble of models that predicted approximately 94\% of the objective functions with a significantly higher likelihood than a random control. (A) The coefficient of variation (CV) for the model parameters ranged from 0.5 - 1.6, where approximately 65\% of the parameters were constrained with a CV $\leq$ 1.0 (black dots). (B) We selected five parameter sets (red dots in A) for further analysis based on CV and distance from the nominal parameter set (based on second norm).}
	\label{fg:Para_S1}
\end{figure}

\begin{figure}\centering
	\epsfig{file=./Supp_Figures/Flux_S1.eps,width=1.0\textwidth}
	\caption{Signal flow analysis using simulated knockout (KO) of key proteins on the UPR system: Simulation results suggest that the three branches in UPR fire simultaneously with varying rates and the state of the cell in terms of adaptation, alarm or apoptosis is a result of counteracting effects of these three prongs of UPR signaling. 
	(A-C) The counteracting effects is seen when knockout of one ER stress transducer leads to enhancement of the other branches of UPR. 
	(D) ATF4, cleaved ATF6 and XBP1s act as integrators of the signals coming from all the three branches of UPR and furthermore leads to regulation of BiP, thereby leading to a negative feedback or control of UPR signal. PERK and ATF4 KO studies revealed a slower and lower amount of BiP production ($\sim$ 50\%) as compared to WT. However, ATF6 or IRE1 KO did not affect BiP regulation as compared to WT. 
	(E) Regulation of BiP was the critical regulator of spliced XBP1 (XBP1s), which in turn acts as a key marker of progression through different stages of UPR. 
	(F) PERK and ATF4 KO lead to delay in the onset of apoptosis (marked by slower and lower reduction of Bcl2 levels. This effect could be attributed to the lack of CHOP mediated branch of Bcl2 regulation. On the other hand, IRE1 and CHOP KO leads to drastic reduction in apoptosis (marked by little or no change of Bcl2 levels). CHOP KO, implicated the importance of CHOP in the down-regulation of Bcl2. IRE1 KO implicated the critical role of IRE1-TRAF2 mediated route of apoptosis. 
Overall flux analysis highlighted the extensive amount of crosstalk within the three branches of the UPR network.}
\label{fg:Flux_Supp_Fig1}
\end{figure}

\begin{figure}\centering
	\epsfig{file=./Supp_Figures/Sens_S1.eps,width=1.0\textwidth}
	\caption{Plot of species sensitivity at earlier (0-2 hrs) versus later (6-8 hrs) time points: Sensitivity analysis was conducted over discrete two hour time windows thereby revealing the time evolution of the importance of UPR network modules. We found that signal integration via the transcriptional activity of ATF6, ATF4 and XBP1s along with RCAN1 and cIAP role in apoptosis were significantly more important at 6-8 hrs as compared to 0-2 hrs time window. This is consistent with the dominant role of the negative feedback via the transcriptional regulation of BiP in UPR. Interestingly, the majority of species rankings were similar as seen in the cluster in the grey box.}
	\label{fg:Sens_S1}
\end{figure}

\begin{figure}\centering
	\epsfig{file=./Supp_Figures/Sens_S3.eps,width=0.6\textwidth}
	\caption{Plot of parameter sensitivity with APAF-1 feedback KO and all BiP feedback KO: Upon knockout of any individual feedback branch like that of ATF4, ATF6 and XBP1s, the system overall remains equally robust. However the sensitivity of the alternate feedback components increases. This was most evident upon ATF4 feedback KO. 
	(A) We saw increase in sensitivity of feedback components associated with XBP1s and ATF6. Upon ATF6 and XBP1s feedback KO, there wasn't much change in terms of sensitivity of the system (data not shown). This further attests the key regulatory effect of ATF4 in mediating the positive BiP feedback which is an essential component of the adaptation phase of UPR. 
	(B) When we completely knockout all the feedback branches of BiP in the adaptation phase, the system overall becomes relatively more robust. We distinctly saw a major shift of sensitivity of BiP upon removal of positive feedback. Overall $\sim$ 54 \% of the parameters were differentially less sensitive upon removal of BiP feedback as compared to WT. This brings to light how the presence of BiP feedback makes the system more susceptible/sensitive to perturbations.}
	\label{fg:Sens_S3}
\end{figure}

\begin{figure}\centering
	\epsfig{file=./Supp_Figures/TRANS_S1.eps,width=1.0\textwidth}
	\caption{Simulations with translation attenuation built in the model: One of the key aspects which was not included in the current model was translation attenuation. So we simulated that to identify that there isnt much of a change overall in the system except for the tad bit delay in the onset of the responses. }
	\label{fg:Trans_S1}
\end{figure}

\begin{figure}\centering
	\epsfig{file=./Supp_Figures/COUP_S1.eps,width=1.0\textwidth}
	\caption{Survival-death phenotypic plane for APAF-1 and CHOP KOs over the entire ensemble: (A) With APAF-1 KO, we found that there were two populations of cells in the ensemble: population 1 where APAF-1 was the dominant regulator of cell-death (marked by enhanced reduction in caspase 3 upon APAF-1 KO) and population 2 where APAF-1 is not the most dominant regulator (marked by reduced effect on Caspase 3 upon APAF-1 KO). (B) Upon CHOP KO, we identified two distinct populations within the ensembles. One with a strong effect of CHOP mediated down-regulation of Bcl2 (marked by $\sim$ 10 fold increase in Bcl2 levels) and the other with very little effect of CHOP on Bcl2 levels. This behavior could be attributed to other conflicting means of regulation of Bcl2 levels.}
	\label{fg:COUP_S1}
\end{figure}

\begin{figure}\centering
	\epsfig{file=./Supp_Figures/COUP_S2.eps,width=1.0\textwidth}
	\caption{To further investigate the implications of the feedback regulation of BiP via ATF4/ATF6/XBP1s, we simulated KOs of these components over the entire ensemble. (A) Upon KO of all branches of BiP feedback, we found overall reductions of BiP levels. However, there were two distinct sub-populations. One with a $\sim$ 10 fold reduction in BiP levels while the other had $\sim$ 1000 fold reduction in BiP levels. These two populations could resemble two distinct operational paradigms within UPR. In the first mode of operation feedback regulation of BiP is really strong so when we knockout BiP feedback we have drastic reductions in BiP levels and ultimately a stronger and faster UPR response. (B) ATF4 mediated feedback KO led to significant amount of reduction in BiP levels thereby highlighting the significance of ATF4 in BiP feedback. (C-D) However, KO of ATF6 and XBP1s mediated feedback of BiP was seen to have little effect (as marked by robustness coefficients for BiP).}
	\label{fg:COUP_S2}
\end{figure}

### 2014-07-21 Paper_v1.tex
First commit

--- 2014-07-21 Paper_v1.tex
@@
 \usepackage{multicol}   %Columns
-\usepackage[utf8]{inputenc} %unicode support
+
 \usepackage{amsmath}
@@
 \usepackage{epsfig}
-\usepackage{epstopdf}
 \usepackage{graphicx}
 \usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
+
+\usepackage[round,numbers,comma,sort&compress]{natbib}
+\bibliographystyle{biophysj}
+
+\usepackage{longtable}
+\usepackage{supertabular}
 \usepackage{setspace}
-%\usepackage{longtable}
+%\usepackage{lineno}
+
+%*** extra packages ****
 \usepackage{colortbl}
-%\usepackage{palatino,lettrine}
-%\usepackage{times}
-%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
-%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
-\usepackage[wide]{sidecap}
-%\usepackage[authoryear,round,comma,sort&compress]{natbib}
-\usepackage[square,sort,comma,numbers]{natbib}
-%\usepackage[authoryear,round]{natbib}
-\usepackage{supertabular}
+\usepackage{color}
 \usepackage{simplemargins}
-\usepackage{comment}
-\usepackage{lineno}
-
-\urlstyle{rm}
 
@@
 
-%\usepackage{geometry}
-\usepackage{fullpage}
-
-%\bibliographystyle{plain}
-\bibliographystyle{plos2009}
+\setallmargins{1in}
 
@@
 %Publication style settings
-
 % Single space'd bib -
@@
 \begin{titlepage}
-{\par\centering\textbf{\Large Dynamically Dimensioned Particle Swarm Optimization}}
+{\par\centering\textbf{\Large Physiochemical Modeling and Analysis of the Mammalian Unfolded Protein Response}}
 \vspace{0.05in}
-{\par \centering \large{Adithya Sagar, Christine Shoemaker and Jeffrey D. Varner$^{*}$}}
-\vspace{0.10in}
+{\par \centering \large{Anirikh Chakrabarti and Jeffrey D. Varner$^{*}$}}
+\vspace{0.1in}
 {\par \centering \large{School of Chemical and Biomolecular Engineering}}
@@
 \vspace{0.1in}
-{\par \centering \textbf{Running Title:}~Modeling cell free metabolism}
+{\par \centering \textbf{Running Title:}~Mammalian unfolded protein response and cellular stress}
 \vspace{0.1in}
-{\par \centering \textbf{To be submitted:}~\emph{BMC~Systems~Biology}}
+{\par \centering \textbf{To be submitted:}~\emph{Bio Phys J}}
 \vspace{0.5in}
@@
 {\par \centering Jeffrey D. Varner,}
-{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
+{\par \centering Assistant Professor, School of Chemical and Biomolecular Engineering,}
 {\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
@@
 \section*{Abstract}
+Cellular stressors routinely lead to malfunctions in the folding state of critical proteins, thereby playing a vital role in disease states like cancer, diabetes and cardiovascular ailments.
+Newly synthesized proteins must fold and assemble into unique three-dimensional structures, in order to become functionally active.
+Cells monitor protein folding by an inbuilt quality-control system involving both the Endoplasmic Reticulum (ER) and the Golgi apparatus.
+Incorrectly folded proteins are tagged for degradation via ER associated degradation (ERAD) or sent back through a refolding cycle.
+However, accumulation of incorrectly folded proteins can trigger a cascade of events, termed the Unfolded Protein Response (UPR).
+UPR leads to re-establishing the cellular homeostasis.
+%This is done by increased transcription of chaperones, reduced translation and cellular apoptosis.
+In this study, we developed a family of mechanistic models of eukaryotic UPR, which was composed of a system of ordinary differential equations.
+The objective of this study was to assemble a series of molecular modules describing different aspects of UPR and subsequently analyze these models for fragility and robustness.
+Kinetic parameters for these models were estimated by comparing simulations with experimental data.
+Using POETs and a cross-validation scheme, we developed an ensemble of models, consistent with literature data.
+Model analysis highlighted the presence of a sequential order for firing of the ER stress transducers in UPR.
+The counteracting effects of the ER stress transducers converge at feedback regulation of the molecular chaperone BiP.
+The three main regulators of feedback, ATF4, cleaved ATF6 and XBP1s share the load ensuring the phased response of UPR.
+However these regulators add on to the overall fragility of the system and allow scope for manipulation of UPR as seen by sensitivity/robustness analysis.
+Downstream effects of UPR include a balance between cell survival and cell death based on the magnitude of the stress.
+Model analysis suggested that the cell-death axis was relatively robust owing to redundant routes e.g., APAF-1 dependent and APAF-1 independent routes of apoptosis.
+However the cell-survival axis was relatively susceptible to perturbations owing to multiple levels of regulation, thereby highlighting importance of key proteins like CHOP.
 
-{\noindent \textbf{Keywords:}~Parameter identification, Mathematical modeling}
+\vspace{0.1in}
+{\noindent \textbf{Keywords:}~ Protein Folding, UPR, PERK, ATF6, IRE1, Apoptosis, Mathematical Modeling, Multi-objective thermal annealing, Sensitivity Analysis.}
 
@@
 
-\linenumbers
+\section*{Introduction}
+Protein folding is strategically important to cellular function in all organisms.
+In eukaryotes, secreted, membrane-bound and organelle-targeted proteins are typically processed and folded in the endoplasmic reticulum (ER) \cite{naidoo2009er, ron2002translational, kaufman2002unfolded}.
+Intracellular perturbations caused by a variety of stressors disturb the specialized environment of the ER leading to the accumulation of misfolded or unfolded proteins \cite{ellgaard2003qce,Fonseca:2009fk}.
+Shifts in folding capacity have been associated with diseases such as cancer, diabetes and cardiovascular disorders \cite{ellgaard2003qce}.
+Physiological processes such as aging can also influence protein folding \cite{naidoo2009er}.
+Normally, cells ensure proper protein folding using a combination of molecular chaperones, foldases and lectins \cite{naidoo2009er}.
+However, when proper folding can not be restored, unfolded or misfolded proteins are targeted to ER Associated Degradation (ERAD) pathways for processing \cite{kaufman2002unfolded}.
+If unfolded or misfolded proteins continue to accumulate, eukaryotes induce the unfolded protein response (UPR).
+In mammalian cells, UPR is a complex signaling program mediated by three ER transmembrane receptors: activating transcription factor 6 (ATF6), inositol requiring kinase 1 (IRE1) and double-stranded RNA-activated protein kinase (PKR)-like endoplasmic reticulum kinase (PERK).
+UPR performs three functions, adaptation, alarm and apoptosis.
+During adaptation, the UPR tries to reestablish folding homeostasis by inducing the expression of chaperones that enhance protein folding.
+Simultaneously, translation is globally attenuated to reduce the ER folding load while the degradation of unfolded proteins is increased.
+If these steps fail, the UPR induces a cellular alarm and apoptosis program.
+The alarm phase involves several signal transduction events, ultimately leading to the removal of the translational block and the down-regulation of the expression and activity of pro-survival factors such as the B-cell lymphoma 2 (Bcl2) protein.
+After the alarm phase, cells can undergo apoptosis, although ER stress can also initiate autophagy \cite{ogata2006aac, yorimitsu2006ers, bernales2006ace, kamimoto2006iic, hoyerhansen2007cmc, kouroku2006esp, fujita2007ter}.
+Thus, ER folding homeostasis strongly influences mammalian physiology \cite{Fonseca:2009fk}.
 
-\section*{Introduction}
+In this study we used physiochemical computer modeling as a tool to study the mammalian unfolded protein response.
+Physiochemical models which describe system-level responses can prioritize experimental directions, generate testable hypothesis and perhaps identify and validate potential therapeutic targets \cite{KITANO2007}.
+For mammalian UPR, physiochemical modeling might tell us the critical components of each branch, and ultimately what must be manipulated to get a desired network response, for example, enhanced death or survival.
+However, there are critical limitations with physiochemical models.
+Foremost amongst these is the large number of unknown model parameters and uncertainty in the model structure.
+It is typically impossible to uniquely identify model parameters, even with extensive training data and perfect models \cite{Gadkar:2005ad}.
+Alternatively, ensemble approaches, which use uncertain model families, have emerged as a promising strategy in systems biology and other fields like climate prediction \cite{Battogtokh:2002uq,Kuepfer:2007,Brown:2003qy,Palmer:2005,Song:2009uq}.
+Their central value has been the ability to quantify simulation uncertainty and to experimentally constrain model predictions.
+For example, Gutenkunst \emph{et al.} showed that predictions were possible using signal transduction model ensembles, despite sometimes only order of magnitude parameter estimates \cite{Gutenkunst:2007oq}.
+More recently, Luan \emph{et al.} predicted patient response to therapeutic intervention using an ensemble of human coagulation models where approximately 40\% of the model parameters had a coefficient of variation (CV) greater than 0.5 \cite{Luan:2010fc}.
+Model ensembles have also been used with parameter dependent analysis techniques, e.g., sensitivity analysis to robustly estimate important signaling network features.
+For example, Tasseff \emph{et al.} characterized emergent behavior between androgen and growth factor signaling in prostate cancer cell-lines using an ensemble of models and sensitivity analysis \cite{Tasseff:2010rq}.
+Thus, despite uncertainty, ensembles of physiochemical models can be used to understand qualitative properties of complex biochemical networks.
+
+We developed a population of physiochemical models describing the adaptation, alarm and apoptosis phases of mammalian UPR.
+Traditionally, it has been hypothesized that there is a sequential order for firing of the ER stress transducers in UPR.
+The PERK and ATF6 branches are thought to be activated before IRE1 \cite{szegezdi2006mediators} and largely promote ER adaptation to misfolding, while IRE1 is the last one to fire as it has a dual role, transmitting both survival and pro-apoptotic signals.
+However our modeling analysis suggests that these three branches fire simultaneously with varying rates and the state of the cell in terms of adaptation, alarm or apoptosis is a result of counteracting effects of these three prongs of UPR signaling.
+The counteracting effects is further substantiated by simulated knockout studies, where in knockout of one ER stress transducer lead to enhancement of the other branches of UPR, e.g, if we want more protein production even with ER stress, PERK KO might suggest as a viable route as that shall lead to no translational block.
+However the IRE1 branch shall amplify and so will the ATF6 branch and ultimately still lead to death (less compared to WT).
+So a viable strategy could be to regulate CHOP, as it shall lead to reduced death by regulation of Bcl2.
+One common playground for the three ER stress transducers is the feedback regulation of BiP expression.
+Model analysis substantiated the significance of BiP feedback in the overall robustness of the system.
+Removal of any one branch of BiP feedback lead to increased sensitivity of the other branches.
+So there was load/responsibility sharing within the system.
+Interestingly, removal of all nodes of BiP feedback increased the overall robustness of the system.
+So while BiP feedback is crucial in terms of the cell to adapt to small perturbations, at the same time this makes the system more fragile and susceptible to manipulations.
+Model analysis further highlighted the presence of redundant routes of regulation of apoptosis such as APAF-1 dependent/independent strategies.
+While manipulation of pro-death axis of the system (cleaved caspases) was relatively robust to perturbations, manipulation of the pro-survival axis (Bcl2) was relatively feasible via CHOP dependent/independent strategies.
+Overall UPR was seen to be robust to perturbations (simulated knockout/overexpression studies), thereby highlighting the redundancy and crosstalk within the three branches of UPR.
+However, BiP regulation at the transcriptional level via intermediates of the ER-stress transducer signaling cascades (e.g.
+ATF4, XBP1s) was seen to be the key in the regulation of UPR.
+The EUPR model is available in SBML in the supplemental materials.
+
+\section*{Results}
+\subsection*{Formulation of the UPR network architecture.}
+The UPR network described the ER folding cycle, ER-associated degradation (ERAD), ER-stress transducer (PERK, IRE1 and ATF6) signaling cascades and stress-induced caspase activation (Fig.
+\ref{fg:fig1}).
+The network consisted of 636 protein or mRNA species interconnected by 1090 interactions (Fig.
+\ref{fg:fig1} Inset).
+Connectivity was formulated from a comprehensive review of the primary literature  \cite{naidoo2009er, ron2002translational, kaufman2002unfolded,ellgaard2003qce,Fonseca:2009fk,Schroder:2005vn,gotoh:hdc,mccullough2001gsc, belmont2008cga,Hetz:2009qc,Urano:2000uq,szegezdi2006mediators}, and from on-line databases; String-8 \cite{Jensen:2009fk}, NetworKIN \cite{Linding:2007fk} and TRANSFAC.
+Model connectivity was not specific to a single cell-line.
+Rather, it was a canonical representation of the pathways involved in monitoring and controlling the folding capacity of a generic well-mixed ER compartment.
+UPR induction was modeled as the release of BiP from the ER stress transducers, PERK, IRE1$\alpha$ and ATF6 leading initially to adaptation of the folding cycle and then subsequently to alarm and apoptosis.
+The adaption phase of UPR was marked by general translation attenuation, selective transcriptional programs for key species like bZIP transcription factor ATF4 \cite{lu2004tra}, cellular inhibitor of apoptosis (cIAP) \cite{hamanaka2008pdr}, molecular chaperones e.g., BiP \cite{harding2003isr} and enhanced clearance of accumulated proteins via ERAD.
+The alarm and apoptosis phases were mediated by the induction of CHOP \cite{ron1992cnd}, regulation of Bcl2, Bcl2-antagonist of cell death (BAD) \cite {wang1999cia} and (TNF) receptor associated factor 2 (TRAF2) \cite{lei2003jpb,putcha2003jmb, yamamoto1999bpa, szegezdi2006mediators} activation (Fig.
+\ref{fg:fig1}).
+Model connectivity is available in SBML format from the supplemental materials.
+Complete details of the interactions in the model are also enlisted in the supplementary materials.
+
+%\clearpage
+\subsection*{Estimating a population of canonical UPR models using POETs.}
+Despite a significant identifiability challenge, the multiobjective POETs algorithm generated a predictive UPR model population.
+The three phases of UPR were modeled using mass action kinetics within an ordinary differential equation (ODE) framework.
+While ODEs and mass-action kinetics are common methods of modeling biological pathways \cite{Schoeberl:2002rv, Chen:2009uq, Helmy:2009kx}, this modeling strategy resulted in a large number of unknown model parameters.
+These parameters were not uniquely identifiable (data not shown).
+Instead, we estimated an experimentally constrained population of parameters using multiobjective optimization.
+A population of the 1726 unknown model parameters (1090 kinetic parameters and 636 initial conditions) was estimated from 33 dynamic and steady state data-sets taken from literature (Table \ref{obj-table}).
+The residual between model simulations and each of the experimental constraints was simultaneously minimized using the multiobjective POETs algorithm \cite{Song:2010ij}.
+A leave-eight-out cross-validation strategy was used to independently estimate the training and prediction error over the 33 data sets; we estimated four different model families, where eight of the 33 objectives were reserved for validation and 25 were used model training each family.
+Thus, each model family was trained and validated on different experimental data.
+Starting from an initial best-fit initial parameter set (nominal set), more than 25,000 probable models were estimated by POETs from which we selected N = 100 models (25 from each training family) with a Pareto rank of one or less (from approximately 1200 possible choices) for further study.
+For each objective, the nominal, training (75 models) and prediction (25 models) error was calculated for each objective (Table \ref{obj-table}).
+Models used for prediction error calculations for a particular objective were \emph{not} trained on that objective.
+The prediction likelihood was statistically significantly better for 31 of the 33 objective functions at a 95\% confidence level, compared with random parameter sets generated from the nominal set (Table \ref{obj-table}).
+Thus, POETs generated model families that predicted approximately 94\% of the objective functions with a significantly higher likelihood than a random control.
+However, the specific value of any given parameter was likely not well described.
+The coefficient of variation (CV) for the model parameters ranged from 0.5 - 1.6, where approximately 65\% of the parameters were constrained with a CV $\leq$ 1.0 (supplementary materials Fig.
+S\ref{fg:Para_S1}).
+The most constrained parameters involved a wide-array of functions e.g., regulation of PERK, eIF2$\alpha$, ATF4, Calcineurin, BiP, CHOP and ATF6 signaling.
+However, the least constrained parameters involved JNK and apoptosis interactions.
+
+POETs identified Pareto fronts between several objectives, e.g., O13$\times$O14, O25$\times$O29, O11$\times$O29 and O27$\times$O2 in the training data  (Fig.
+\ref{fg:objective_plot}).
+Strong Pareto fronts suggested an inability to simultaneously model different aspects of the training data.
+However, fronts could also result from experimental artifacts, e.g., variation between cell-lines, time-scale differences or from functional relationships in the data.
+Globally, adaptation and alarm phase training constraints conflicted with those involving apoptosis.
+For example, objectives involving caspase-7 or caspase-9 activity conflicted with phosphorylated eIF2$\alpha$ levels.
+Phosphorylation of eIF2$\alpha$ by activated PERK attenuates translation, which decreases the ER folding load.
+Thus, eIF2$\alpha$ phosphorylation is a key early adaptive event in UPR.
+On the other hand, caspase-9 is a stress-induced death marker activated only after UPR has failed to restore folding homeostasis.
+Conflicts between these early and late phase markers suggested the UPR time-scale was perhaps cell-line or perturbation dependent.
+Similarly, negative feedback may lead to conflicting objectives.
+For example, XBP1 mRNA measurements (O14) conflicted with CHOP protein measurements (O13), even though these data-sets were taken from the same study and were collected in the same cell-line.
+XBP1 splicing increased BiP levels, which in turn reduced CHOP protein levels, hence the trade-off.
+Lastly, in addition to fronts, we also observed strong correlation between objectives.
+For example, models that performed well for the CHOP protein (O11), also performed well against Procaspase-12 (O22) measurements, even though these were not in the same cell-line or from the same study.
+Both CHOP and Procaspase-12 are downstream of IRE1/TRAF2/JNK signaling cascade, so these errors were directly correlated.
+
+%\clearpage
+\subsection*{The population of UPR models recapitulated adaptation, alarm and apoptotic events across multiple cell-lines and timescales.} 
+We trained the UPR ensemble using data from UPR initiation events (e.g., PERK activation) and the downstream activation of proteins involved in apoptosis (BiP, caspase-3 and caspase-7).
+ER stress induced by exposure to thapsigargin (Tg), a non-competitive inhibitor of SERCA Ca$^{2+}$ transporters, leads to the dissociation of BiP from the ER-stress transducers.
+For theses initial proof-of-concept simulations, we assumed that the action of Tg and other stress-inducing agents such as dithiothreitol (DTT) was identical, i.e., induction of BiP dissociation.
+Within 20 min after Tg exposure, PERK was activated and transmitting adaptation signals downstream.
+The population of UPR models recapitulated the timescale of PERK phosphorylation (Fig.
+\ref{fg:simulation_comparison}A) as well as its downstream signaling activity, for example, the phosphorylation of eIF2$\alpha$ (Fig.
+\ref{fg:simulation_comparison}H).
+The nuclear fraction of ATF4 increased from approximately zero (untreated cells) to a maximum value 4 hrs after Tg exposure.
+While the model ensemble generally predicted the correct trend, there was significant error in the early time points for ATF4 (Fig.
+\ref{fg:simulation_comparison}G).
+The phosphorylation of eIF2$\alpha$ by PERK is required for ATF4 activation.
+Interestingly, when we compared model simulations of p-eIF2$\alpha$ levels following Tg (1$\mu$M) exposure in MEFs with measurements (O15 \cite{yamamoto2007tim}), the model correctly captured the appropriate behavior.
+To test the functioning of the ATF6 branch of UPR model, we compared simulations with measurements of cleaved ATF6 in tunicamycin-treated MEFs \cite{lee2002imu}.
+ER stress leads to the release of the BiP from ATF6.
+Cleaved ATF6 is then translocated to the nucleus where it up-regulates gene expression \cite{silver1999mtf,hai1989tfa}.
+Simulations of cleaved ATF6 levels following UPR initiation were consistent with measurements (Fig.
+\ref{fg:simulation_comparison}C).
+The formation of p-PERK or p-IRE1 initiates a complex series of events that operate on both short and long time-scales.
+Signals from the ER stress transducers converge downstream to regulate BiP transcription \cite{malhotra2007era, rao2004mpe,kokame2001iei, yoshida2000aap}.
+Model ensemble recapitulated correct trends of BiP mRNA with maximum levels $\sim$ 8 hrs, similar to what was seen in experiments done on HEK293 cells (Fig.
+\ref{fg:simulation_comparison}D) \cite{lin2007ire1}.
+One of the long-term outcomes of PERK/IRE1 activation is apoptotic cell death.
+In the proof-of-concept ER-model the link between UPR and apoptosis occurred through the action of eIF2$\alpha$, the dual role of the ATF4 transcription factor and caspase-12 activation by IRE1-TRAF2 signaling axis.
+We constrained model parameters associated with the activation of the cell-death program using measurements of pro/caspase-7 levels, pro/caspase-9 levels, pro/caspase-3 levels, pro/caspase-12 levels and PARP cleavage mediated by executioner caspases following treatment with 0.5$\mu$M Tg \cite{rao2002cer}.
+These experiments were performed in Sak2 cells that lack Apaf-1 protein expression \cite{rao2002cer}.
+Thus, these data allowed us to include a non-Apaf-1 mediated stress-induced caspase activation pathway into the model.
+The population of models recapitulated caspase-3 (Fig.
+\ref{fg:simulation_comparison}K), caspase-9 (Fig.
+\ref{fg:simulation_comparison}J) as well as cleaved Parp levels (Fig.
+\ref{fg:simulation_comparison}L) following exposure to ER stress-inducers.
+Interestingly, while PERK activation occurred on the timescale of minutes, initiator and executioner caspase activation occurred over 36 hrs.
+Thus, the population of UPR models captured complex signaling events occurring across multiple timescales.
+
+\subsection*{Signal flow analysis of UPR highlighted modes of crosstalk and redundancy in BiP and Bcl2 regulation.}
+Traditionally, it has been hypothesized that there is a sequential order for firing of the ER stress transducers in UPR.
+The PERK and ATF6 branches are thought to be activated before IRE1 \cite{szegezdi2006mediators} and largely promote ER adaptation to misfolding, while IRE1 is the last one to fire as it has a dual role, transmitting both survival and pro-apoptotic signals.
+However our modeling analysis suggests that these three branches fire simultaneously with varying rates and the state of the cell in terms of adaptation, alarm or apoptosis is a result of counteracting effects of these three prongs of UPR signaling (Fig.
+\ref{fg:proof_of_concept_simulation} and supplementary materials Fig.
+S\ref{fg:Flux_Supp_Fig1}).
+The counteracting effects is further substantiated in simulated knockout studies, where in knockout of one ER stress transducer leads to enhancement of the other branches of UPR (supplementary materials Fig.
+S\ref{fg:Flux_Supp_Fig1}A-C).
+UPR induction in the model was controlled by manipulation of the generation rate of unfolded or misfolded protein (qP) in the ER compartment.
+Upon UPR induction, initially ($\leq$ 1hr) the response is damped marking the adaptation phase of UPR.
+This is followed by an increase in the activity of the IRE1$\alpha$, PERK and ATF6 cascades at $\sim$ 1 hr marking the onset of the alarm phase, which leads to a final steady state $\sim$ 8-10 hrs marking the onset of the commitment or apoptosis phase of UPR (Fig.
+\ref{fg:proof_of_concept_simulation}).
+This time frame was consistent with the maximum levels of BiP mRNA upon UPR induction \cite{lin2007ire1}.
+This analysis was substantiated further by looking at the fluxes at different phases of UPR induction.
+As compared to P1 (No-UPR Steady State Fig.
+\ref{fg:proof_of_concept_simulation}), we saw that early on at 1 hr after unfolded protein dose there was a marked increase in ATF4 and CHOP regulation, ATF6 signaling along with unfolded protein sensing and degradation.
+These are hallmarks of the adaptation-alarm phase of the UPR response (Fig.
+\ref{fg:cross_plot_flux}D).
+The apoptosis phase ($\geq$ 8-10 hrs P2), was marked by increased BiP regulation, enhanced ATF4 transcriptional activity, increased mitochondrial membrane permeability and increased apoptotic fluxes.
+So the cell has committed itself to apoptosis mediated cell death.
+We compared the ability of the system to recuperate upon reversal of unfolded protein loads from the alarm phase (P4, Fig.
+\ref{fg:proof_of_concept_simulation} ) and the apoptosis phase (P2, Fig.
+\ref{fg:proof_of_concept_simulation} ).
+If we reduced the protein load after the cell was committed to apoptosis (as in P3), we found that the cell continued to function similar to the UPR state even upon protein load reduction after 25 hrs (Fig.
+\ref{fg:cross_plot_flux}C).
+Certain modules were seen to reduce like IRE1-TRAF2 signaling, ASK1 activation.
+However not much difference was seen in terms of apoptotic fluxes, denoting the commitment of the cell to death and to a point of no return.
+On the contrary if we reduced the load of proteins in the adaptation-alarm phase (P4), we saw that the cell could recuperate using its ERAD machinery and the regulation of BiP (Fig.
+\ref{fg:cross_plot_flux}E-F).
+
+We further investigated the effect of simulated knockout (KO) and overexpression (OX) of key proteins on the UPR system.
+Interestingly, PERK and ATF4 KO studies revealed a slower and lower amount of BiP production ($\sim$ 50\%) as compared to WT.
+However, ATF6 or IRE1 KO did not affect BiP regulation as compared to WT.
+This highlighted the dominant role of ATF4 in regulation of BiP, which is consistent with experimental evidence \cite{ma2003delineation}.
+Regulation of BiP was the critical regulator of spliced XBP1 (XBP1s), which in turn acts as a key marker of progression through different stages of UPR (supplementary materials Fig.
+S\ref{fg:Flux_Supp_Fig1}E).
+ATF4, cleaved ATF6 and XBP1s act as integrators of the signals coming from all the three branches of UPR and furthermore leads to regulation of BiP, thereby leading to a negative feedback or control of UPR signal.
+Another interesting note was the regulation of pro-apoptosis phenotype via regulation of Bcl2.
+PERK and ATF4 KO lead to delay in the onset of apoptosis (marked by slower and lower reduction of Bcl2 levels, supplementary materials Fig.
+S\ref{fg:Flux_Supp_Fig1}F).
+This effect could be attributed to the lack of CHOP mediated branch of Bcl2 regulation.
+On the other hand, IRE1 and CHOP KO leads to drastic reduction in apoptosis (marked by little or no change of Bcl2 levels, supplementary materials Fig.
+S\ref{fg:Flux_Supp_Fig1}F).
+CHOP KO, implicated the importance of CHOP in the down-regulation of Bcl2.
+IRE1 KO implicated the critical role of IRE1-TRAF2 mediated route of apoptosis.
+Overall signal flow analysis highlighted the extensive amount of crosstalk within the three branches of the UPR network.
+Taken together, the population of EUPR models recapitulated both short- and long timescale behavior following overload of unfolded proteins in a variety of cell types.
+It also captured the integration between multiple pathways and generated specific and testable hypothesis about the role of network components in signal propagation.
+
+\subsection*{Sensitivity analysis stratified locally and globally important components of the UPR architecture.} 
+First-order sensitivity coefficients were computed, time-averaged (over approximately 8 hrs of simulated time following ER stress) and rank-ordered for the 1090 models parameters and 636 species under normal and UPR induced conditions.
+Five parameter sets were selected from the parent ensemble based on rightful representation of the diversity for the sensitivity calculations (further details in the supplementary materials).
+We selected only a few parameter sets for the sensitivity analysis, because of the computational cost of computing all 1090 coefficients over several hours of simulated time.
+However, we did select diverse sets as shown by the scatter in the CV values (supplementary materials Fig.
+S\ref{fg:Para_S1}).
+Infrastructure parameters e.g.
+nuclear transport, RNA polymerase or ribosome binding were globally critical, independent of stress (black points, Fig.
+\ref{fg:fig5}).
+Additionally, apoptotic species and parameters were also important, both in the presence and absence of UPR (yellow points, Fig.
+\ref{fg:fig5}).
+Thus, as expected, components such as RNA polymerase or caspase activation were globally important irrespective of the folding state of the ER.
+More interesting however were coefficients that shifted above or below the 45$^{o}$-line in the presence of UPR.
+These points denote differentially important network components.
+While the majority of parameters and species became more important in the presence of stress, we found a band of parameters (Fig.
+\ref{fg:fig5} Inset) that were differentially important under stress.
+For example, the rank-ordering of the sensor and stress-transducer modules clearly increased in the presence of UPR.
+Approximately, 172 or 15\% of the parameters were significantly more important in the presence of UPR.
+These parameters were largely associated with adaptation and processing of unfolded or misfolded proteins, e.g., unfolded protein degradation, cleaved ATF6-induced gene expression, IRE1-TRAF2 mediated apoptosis regulation and RCAN1 regulation.
+Likewise, 75 or 12\% of the species were significantly more important in UPR compared with normal protein loads (data not shown).
+
+Sensitivity analysis conducted over discrete two hour time windows revealed the time evolution of the importance of UPR network modules (Fig.
+\ref{fg:fig5}).
+Comparison of the 0 - 2 hrs time window with itself (top panel, first column of Fig.
+\ref{fg:fig5}), supported the earlier results that infrastructure components were globally critical followed by ERAD species.
+These species remained important in all time windows.
+On the other hand, during the initial 0 - 2 hrs window, ER stress transduction pathway components were robust.
+Comparison of the 0 - 2 hrs time window with later time points (working down the first column, Fig.
+\ref{fg:fig5}), showed the increasing importance of different modules as a function of time.
+For example, components of the PERK and IRE1 modules were more important in the 2 - 4 hrs window compared to the earlier time points, while alarm and apoptotic phase species were more important in the 6 - 8 hrs window compared to the earlier time points.
+Specifically, signal integration via the transcriptional activity of ATF6, ATF4 and XBP1s along with RCAN1 and cIAP role in apoptosis were significantly more important at 6-8 hrs as compared to 0-2 hrs time window.
+This is consistent with the dominant role of the negative feedback via the transcriptional regulation of BiP in UPR.
+Interestingly, the majority of species rankings were similar after 6 hrs (bottom row,  Fig.
+\ref{fg:fig5} and supplementary materials Fig.
+S\ref{fg:Sens_S1}).
+This analysis supports the structural analysis results, wherein the signal integrators like XBP1s and transcriptional regulation of BiP via ATF6 and ATF4 are the key regulators of the phased response of UPR.
+
+To further investigate the role of positive BiP regulation via ATF6, ATF4 and XBP1s, we conducted sensitivity analysis upon knocking out the feedback branches of BiP for the nominal parameter set (supplementary materials Fig.
+S\ref{fg:Sens_S3}).
+Interestingly, upon knockout of any individual feedback branch like that of ATF4, ATF6 and XBP1s, the system overall remains equally robust.
+However the sensitivity of the alternate feedback components increases.
+This was most evident upon ATF4 feedback KO.
+We distinctly saw increase in sensitivity of feedback components associated with XBP1s and ATF6 (supplementary materials Fig.
+S\ref{fg:Sens_S3}).
+Upon ATF6 and XBP1s feedback KO, there wasn't much change in terms of sensitivity of the system.
+This further attests the key regulatory effect of ATF4 in mediating the positive BiP feedback which is an essential component of the adaptation phase of UPR.
+Another interesting observation was that when we completely knockout all the feedback branches of BiP in the adaptation phase, the system overall becomes relatively more robust (supplementary materials Fig.
+S\ref{fg:Sens_S3}).
+We distinctly saw a major shift of sensitivity of BiP upon removal of positive feedback.
+Overall $\sim$ 54 \% of the parameters were differentially less sensitive upon removal of BiP feedback as compared to WT.
+This brings to light how the presence of BiP feedback makes the system more susceptible/sensitive to perturbations.
+This highlights the essential role and relevance of targeting the BiP feedback in manipulating UPR and  specific importance of ATF4 amongst the three feedback branches.
+
+\subsection*{Robustness analysis predicted the phenotypic consequence of structural perturbations to the UPR network.} We calculated the direction of unfolded protein load induced concentration shifts for 636 markers following single parameter knockouts (Edge KO), single gene knockouts (GKO) and single gene overexpression (GOX) to the UPR network for the nominal parameter set (Fig.
+\ref{fg:robustness_figure}).
+Robustness coefficients were used to quantify the effect of structural perturbations on network markers.
+Coefficients with values $>$ 1 ($<$ 1) indicated a marker increased (decreased) compared to the basal state, while a value $\sim$ 1 indicated approximately no change following a perturbation.
+Phenotypic behavior of the UPR models were analyzed as a result of perturbation, using downstream markers like Caspase 3 (marker for cell death), Bcl2 (marker for cell-survival) and other key signal integrators of the signals from ER stress transducers e.g., XBP1s, CHOP and ATF4.
+
+Overall, we found that in the survival-death phenotypic plane, the pro-death phenotype (marked by robustness coefficients $>$ 1 for Caspase 3) was seen to be relatively robust to structural (both GKO, GOX and Edge KO) perturbations.
+Few perturbations lead to increased Caspase 3 levels, e.g., overexpression of Procaspases 9/3 (Fig \ref{fg:robustness_figure} A, Table \ref{KO-O-table}).
+This robustness of the apoptotic marker caspase 3, can be attributed to the redundant sources of cell death (e.g., APAF-1 dependent and APAF-1 independent strategies).
+To confirm this we ran simulated APAF-1 KOs over the entire ensemble (supplementary materials Fig.
+S\ref{fg:COUP_S1}).
+We found that there were two populations of cells in the ensemble: population 1 where APAF-1 was the dominant regulator of cell-death (marked by enhanced reduction in caspase 3 upon APAF-1 KO) and population 2 where APAF-1 is not the most dominant regulator (marked by reduced effect on Caspase 3 upon APAF-1 KO) (supplementary materials Fig.
+S\ref{fg:COUP_S1}).
+This behavior is consistent based on the training data from Sak2 cells (APAF-1 $-ve$ cells) \cite{rao2002cer}.
+Interestingly, manipulation of the pro-survival axis via regulation of Bcl2 was possible (Fig \ref{fg:robustness_figure}A, Table \ref{KO-O-table}).
+For example, KO of PERK/ATF4 signaling components lead to increased Bcl2 marker levels (Fig \ref{fg:robustness_figure}A).
+This behavior is attributed to the dominant role of PERK-ATF4 mediated regulation of CHOP which downstream leads to down-regulation of Bcl2 levels (seen earlier in signal analysis results, supplementary information).
+To further investigate the extent of CHOP mediated down-regulation of Bcl2, we simulated CHOP KOs over the entire ensemble (supplementary materials Fig.
+S\ref{fg:COUP_S1}).
+We identified two distinct populations within the ensembles.
+One with a strong effect of CHOP mediated down-regulation of Bcl2 (marked by $\sim$ 10 fold increase in Bcl2 levels) and the other with very little effect of CHOP on Bcl2 levels.
+This behavior could be attributed to other conflicting means of regulation of Bcl2 levels.
+Other schemes of regulating Bcl2 levels include perturbations involving eIF2$\alpha$, BiP, JNK, cIAP and RCAN1 (Fig \ref{fg:robustness_figure}A, Table \ref{KO-O-table}).
+
+Robustness analysis allowed us to investigate complex network properties like redundancy and crosstalk.
+For example, the direct correlation between ATF4 and CHOP was further noted in the ATF4-CHOP phenotypic plane.
+Any perturbations affecting ATF4 affected CHOP levels in the same manner (reduction in ATF4 levels lead to reduced levels of CHOP) (Fig.
+\ref{fg:robustness_figure}A-B).
+However, owing to redundant sources of CHOP regulation (e.g., via XBP1s), effect on CHOP is damped as compared to significant changes in ATF4 levels.
+In the XBP1s-CHOP plane, we see at lower levels of XBP1s and CHOP, there is a direct relation between XBP1s levels and CHOP levels.
+However, there exists very few strategies of having both high XBP1s levels and CHOP levels indicating that higher XBP1s doesn't necessarily mean higher CHOP levels.
+To further investigate the implications of the feedback regulation of BiP via ATF4/ATF6/XBP1s, we simulated KOs of these components over the entire ensemble (supplementary materials Fig.
+S\ref{fg:COUP_S2}).
+KO of ATF6 and XBP1s mediated feedback of BiP was seen to have little effect (as marked by robustness coefficients for BiP, supplementary materials Fig.
+S\ref{fg:COUP_S2}).
+However, ATF4 mediated feedback KO led to significant amount of reduction in BiP levels (supplementary materials Fig.
+S\ref{fg:COUP_S2}) thereby highlighting the significance of ATF4 in BiP feedback.
+Upon KO of all branches of BiP feedback, we found overall reductions of BiP levels.
+However, there were two distinct populations.
+One with a $\sim$ 10 fold reduction in BiP levels while the other had $\sim$ 1000 fold reduction in BiP levels.
+These two populations could resemble two distinct operational paradigms within UPR.
+In the first mode of operation feedback regulation of BiP is really strong so when we knockout BiP feedback we have drastic reductions in BiP levels and ultimately a stronger and faster UPR response.
+
+Global analysis of the gene knockout robustness coefficients by clustering to minimize the variance, provides a systems-level insight into the UPR network.
+As in Fig.
+\ref{fg:robustness_figure}C, dendrogram of the single GKOs revealed clusters with inherent functional relationships.
+For example, the most distinct separation was between unfolded protein sensing and IRE1/PERK signal initiation from the rest of the knockouts.
+PERK/ATF4 branch as seen earlier, plays a dominant role in the regulation of BiP and CHOP upon the onset of UPR.
+Similarly, IRE1/TRAF2 signaling axis is valuable to the apoptosis module.
+Another interesting functional module was that of CHOP, involving p38MAPK which leads to down-regulation of Bcl2 levels which considerably affects the apoptosis module.
+We computed the magnitude of the orthogonal components of the single GKO coefficients (Fig.
+\ref{fg:robustness_figure}C Inset).
+The orthogonal component was used to establish the uniqueness of the knockout.
+All the single GKOs were found to have orthogonal components greater than one with a 95\% confidence.
+KO of BiP, Procaspase 9, IRE1, PERK and TRAF2 were some of the knockouts which produced the most unique effects.
+This is in accord to the critical role of BiP in initiating and further regulating the time scale of progression and the ultimate result of the UPR response.
+So PERK via ATF4 plays a key role in regulation of BiP, is rightly seen to have a major effect as compared to other GKOs.
+Similarly, regulation of the apoptosis branch via IRE1-TRAF2 and Procaspase 9 rightly were seen to have the most effect as compared to other GKOs.
 
 \clearpage
+\section*{Discussion}
 
-\section*{Results}
+% Structure of discussion: 4 Paragraphs: General, Para 1, Para 2, Weaknesses
+% General Paragraph
+Proteins requiring post-translational modifications such as N-linked glycosylation or disulfide bond formation are processed in the endoplasmic reticulum (ER).
+A diverse array of cellular stresses can lead to dysfunction of the ER, and ultimately to an imbalance between protein-folding capacity and protein-folding load.
+Unfolded or misfolded proteins are tagged for degradation via ER associated degradation (ERAD) or sent back through the folding cycle.
+Continued accumulation of incorrectly folded proteins can also trigger the Unfolded Protein Response (UPR).
+In this study, we formulated a mechanistic model of the cellular response to stress (protein overload) and studied the core regulatory aspects and downstream effects of UPR induction.
+PRKR-like ER kinase (PERK), inositol-requiring kinase 1 (IRE1) and activating transcription factor 6 (ATF6) were modeled as the key UPR initiators.
+While UPR has been extensively studied \cite{naidoo2009er, ron2002translational, kaufman2002unfolded,ellgaard2003qce,Fonseca:2009fk,Schroder:2005vn,gotoh:hdc,mccullough2001gsc, belmont2008cga,Hetz:2009qc,Urano:2000uq,szegezdi2006mediators}, a detailed mathematical model to investigate the complexities involved is lacking.
+The UPR network architecture is based on extensive review of the literature (details of connectivity in supplementary materials) \cite{naidoo2009er, ron2002translational, kaufman2002unfolded,ellgaard2003qce,Fonseca:2009fk,Schroder:2005vn,gotoh:hdc,mccullough2001gsc, belmont2008cga,Hetz:2009qc,Urano:2000uq,szegezdi2006mediators}.
+Mass balance equations describing 636 species interconnected by 1090 interactions were formulated using mass-action kinetics within an ordinary differential equation (ODE) framework.
+Four model populations were estimated using multi-objective optimization (33 objective functions) in conjunction with a leave-eight out cross-validation strategy using POETs \cite{Song:2010ij}.
+These model populations were then analyzed using population-based sensitivity and robustness analysis.
+Overall we identified sources of network crosstalk and redundancy within the UPR module which could be the reason for emergent properties and be the links for aberrations in cellular adaptation to stress.
 
-Coagulation is an archetype biochemical network that is highly interconnected and tightly regulated with multiple positive and negative feedback loops. 
-The biochemistry underlying coagulation has been well studied [REFHERE], and reliable experimental coagulation models have been developed [REFHERE].
+% Paragraph 1: 
+A key finding of our study was that the overall outcome of UPR was as a result of simultaneous firing and competition between signaling mediated by the three ER-stress transducers: PERK, IRE1 and ATF6.
+This is in contrast to the traditional belief that PERK and ATF6 branches are activated before IRE1 \cite{szegezdi2006mediators}.
+So what we hypothesize is that instead of a sequential ordering of these branches, the state of the cell in terms of adaptation, alarm or apoptosis is a result of counteracting effects of these three prongs of UPR signaling.
+The counteracting/competing effects was further substantiated in simulated knockout studies, where in knockout of one ER stress transducer lead to enhancement of the other branches of UPR.
+Signal transduction architectures frequently contain redundancy, feedback and crosstalk.
+These topological features ensure signal propagation is adaptable, efficient and robust.
+However, they also make reprogramming signal flow challenging.
+This was highlighted remarkably in case of UPR.
+Signals from the three ER-stress transducers converged at the level of up-regulation of BiP.
+This is the key junction which regulates the three stages (onset and time) of adaptation/alarm and apoptosis.
+In this regard, regulators of this feedback cleaved ATF6, ATF4 and XBP1s were seen as highly sensitive components of UPR.
+Interestingly these components put-together increased the overall fragility of the system and presenting more scope of manipulation of the UPR response.
+This was substantiated by sensitivity analysis upon KO of the feedback loops, where we saw increased stability of the UPR module.
+When these feedback components were knocked out individually, the system overall remained stable thanks to increased activity and load sharing via the other feedback branches.
+Amongst the three components of feedback, we identified ATF4 as the key load bearer/regulator.
+This was substantiated by signal flow, robustness and sensitivity analysis.
+This is really interesting as ATF4 protein has shown to be present greater levels in cancer compared to normal tissue, and it is up-regulated by signals of the tumor microenvironment such as hypoxia/anoxia, oxidative stress, and ER stress \cite{Ameri:2008fk}.
+So any aberrations in regulation of ATF4 could potentially serve as a specific target in cancer therapy.
+As a target ATF4 is attractive because it is also potentially involved in angiogenesis and adaptation of cancer cells to hypoxia/anoxia, which are major problems in cancer progression \cite{Ameri:2008fk}.
 
-This makes it an ideal system for mathematical modeling and parameter estimation.
-Coagulation is regulated by a set of serine proteases also known as coagulation factors and blood platelets.
-The coagulation factors are generally in an inactive state and are known as zymogens.
-These zymogens are activated when through certain triggers.
-These trigger events like injury or trauma or sepsis exposure factors like collagen, tissue factor and von Willebrand factor (vWF) to blood.
-The exposure of these factors to blood kick-starts a series of convergent cascades that lead to conversion of zymogen prothrombinase to thrombin.
-Luan et al.
-modeled coagulation using coupled non-linear ordinary differential equations with 301 reactions and 192 species.
+% Paragraph 2: 
+Downstream effects of UPR ranges from cellular adaptation/survival (low stress) to the cell committing to apoptosis mediated death (high stress).
+Our modeling analysis suggested that the cell-death phenotype (marked by increased levels of Caspase 3 as compared to WT) was relatively robust.
+This robustness could be attributed to redundant routes of APAF-1 dependent and APAF-1 independent routes of apoptosis.
+This claim is supported by experimental evidence in Sak2 cells \cite{rao2002cer} and as seen by our simulated knockout studies where we identified two distinct populations representing clear distinctions in APAF-1 dependent and independent routes of apoptosis.
+Interestingly, manipulation of the pro-survival phenotype (marked by increased levels of Bcl2 as compared to WT) was feasible.
+The most effective route was via manipulation of PERK/ATF4/CHOP branch.
+This was substantiated by simulated CHOP KO experiments over the entire ensemble, wherein we identified two distinct populations within the ensembles.
+One with a strong effect of CHOP mediated down-regulation of Bcl2 (marked by $\sim$ 10 fold increase in Bcl2 levels) and the other with very little effect of CHOP on Bcl2 levels.
+This complex network behavior could be attributed to other conflicting means of regulation of Bcl2 levels.
+Rightfully so, induction of CHOP is involved in the development of various diseases and several therapeutic interventions \cite{oyadomari2003rcg}.
+For instance, suppression of CHOP by RNA interference, decoy oligodeoxynucleotides or drug inhibitors have a significant therapeutic potential to modulate type I diabetes and brain ischemia.
+On the other hand, overexpression of CHOP may represent a new class of anticancer therapy.
+Since induction of BiP has been observed in a variety of tumor cells, overexpression of CHOP directed by the BiP promoter may be used as a highly specific therapy for cancer \cite{oyadomari2003rcg}.
+Model analysis also highlighted the essential role of RCAN1 and IRE1-TRAF2 routes of apoptosis.
+ATF6 induces regulator of calcineurin 1 (RCAN1) expression \cite{belmont2008cga}.
+RCAN1 sequesters calcineurin \cite{belmont2008cga}, a calcium activated protein-phosphotase B, that dephosphorylates Bcl2-antagonist of cell death (BAD) at S75 or S99 \cite{wang1999cia}.
+This leads to sequestering of Bcl2 by Bad, which inhibits its downstream anti-apoptotic activity \cite{wang1999cia}.
+Recently, a number of ATF6 homologs have been identified, e.g., OASIS, CREBH, LUMAN/CREB3, CREB4 and BBF2H7 that are processed in a similar way as ATF6, yet their function remains unknown \cite{Ron:2007rq}.
+Thus, ER-stress induced ATF6 signaling may be responsible for additional undiscovered functionality.
+
+% Network corrections/corrections for future models:  
+Given the structural and parametric uncertainty associated with the current version of the UPR model, we were still able to extract interesting insight into the complex inner workings and present falsifiable hypothesis regarding manipulating the UPR program.
+While we did an extensive search of the literature to formulate the model, we were missing certain key structural aspects of UPR which could provide a more comprehensive analysis for further generations of the model.
+One key missing aspect is the negative regulation of the three ER-stress transducers.
+Given PERK's central role in translation attenuation, cells have evolved multiple axes to regulate PERK activity.
+First, the cytosolic kinase domain of PERK can be inhibited by the action of the DNAJ family member P58$^{IPK}$.
+P58$^{IPK}$ was initially discovered as an inhibitor of the eIF2$\alpha$ protein kinase PKR \cite{Lee:1990ye}.
+P58$^{IPK}$, whose expression is induced following ATF6 activation, binds to the cytosolic kinase domain of PERK, inhibiting its activity \cite{Yan:2002fk,Huizen:2003qo}.
+Inhibition of PERK kinase activity relieves eIF2$\alpha$ phosphorylation, thereby removing the translational block.
+Interestingly, P58$^{IPK}$ expression occurs several hours after PERK activation and eIF2$\alpha$ phosphorylation.
+Thus, P58$^{IPK}$ induction may mark the end of UPR adaptation, and the beginning of the alarm/apoptosis phase of the response \cite{szegezdi2006mediators}.
+Second, PERK induces a negative feedback loop, through its downstream effector CHOP, involving the direct de-phosphorylation of eIF2$\alpha$.
+CHOP induces the expression of GADD34 which, in conjunction with protein phosphatase 1 (PP1), assembles into a phosphatase which dephosphorylates the S51 residue of eIF2$\alpha$ \cite{Novoa:2001mb}.
+GADD34 is a member of the GADD family of genes which are induced by DNA damage and a variety of other cellular stresses \cite{Zhan:1994cq}.
+The GADD34 binding partner in this complex appears to be responsible for PP1$\alpha$ recognition and targeting of the phosphatase complex to the ER.
+Association between GADD34 and PP1 is encoded by a C-terminal canonical PP1 binding motif, KVRF, while approximately 180 residues, near the N-terminus of GADD34, appear to be responsible for ER localization \cite{Brush:2003kh}.
+Currently, little is known about deactivation of ATF6.
+Recently, XBP1u, the unspliced form of XBP1, has been implicated as a negative regulator for ATF6 \cite{Yoshida:2009bs}.
+Following, the induction of ER stress, two versions of XBP1 exist: XBP1u and sXBP1 \cite{Yoshida:2009bs}.
+In the recovery phase following ER stress, high levels of XBP1u may play a dual role.
+First, XBP1u binds sXBP1, promoting complex degradation \cite{Yoshida:2006dz, Tirosh:2006fv}.
+Second, XBP1u can bind ATF6$\alpha$ rendering it more prone to proteasomal degradation \cite{Yoshida:2009bs}.
+Taken together, these two steps may slow the transcription of ER chaperones and ERAD components during the recovery phase following ER stress.
+IRE1$\alpha$ activity is regulated by several proteins, including tyrosine phosphatase 1B (PTP-1B), ASK1-interactive protein 1 (AIP1) and members of the Bcl2 protein family.
+PTP-1B has been implicated in a number of IRE1$\alpha$ signaling events.
+The absence of PTP-1B reduced IRE1$\alpha$ dependent JNK activation, XBP1 splicing and EDEM transcription in immortalized and primary mouse embryonic fibroblasts \cite{Gu:2004kx}.
+However, no physical interaction between IRE1$\alpha$ and PTP-1B was established.
+On the other hand, AIP1 physically interacts with both TRAF2 and IRE1$\alpha$, suggesting a model in which AIP1 facilitates IRE1$\alpha$ dimerization and activation \cite{Luo:2008ly}.
+The C-terminal period-like domain (PER) of AIF1 binds the N-terminal RING finger domain of TRAF2, followed by ASK1-JNK signaling \cite{Zhang:2004ve}.
+Thus, based on these findings, Luo \emph{et al.} postulated that AIF1 may be directly involved in the IRE1$\alpha$-TRAF2 complex and its activation of the ASK1-JNK signaling axis \cite{Luo:2008ly}.
+This hypothesis was validated in AIP1-KO mouse studies; AIP1-knockout mouse embryonic fibroblasts and vascular endothelial cells showed significant reductions in ER-stress induced ASK1-JNK activation that was rescued in AIP1 knock-in cells \cite{Luo:2008ly}.
+IRE1$\alpha$ has also been shown to directly interact with Bcl-2 family members Bax and Bak.
+Hetz \emph{et al.} showed that Bax and Bak complex with the cytosolic domain of IRE1$\alpha$ and modulate IRE1$\alpha$ signaling \cite{Hetz:2006qf}.
+Bax and Bak double knockout mice failed to signal through the IRE1$\alpha$ UPR branch following tunicamycin-induced ER stress; however, PERK signaling markers, e.g., eIF2$\alpha$ phosphorylation, responded normally \cite{Hetz:2006qf}.
+This pro-activation role of Bak and Bax may be modulated by one of the few negative regulators of IRE1$\alpha$ activity, Bax inhibitor 1 (BI-1).
+BI-1 is an anti-apoptotic protein that enhances cell survival following several intrinsic death stimuli \cite{Xu:1998dq}.
+Bailly-Maitre \emph{et al.} were the first to suggest that BI-1 may downregulate IRE1$\alpha$ and possibly ATF6 activity \cite{Bailly-Maitre:2006bh}.
+BI-1 deficient mice displayed increased XBP1s and enhanced JNK activity in the liver and kidney, while eIF2$\alpha$ phosphorylation remained normal under ER-stress conditions \cite{Bailly-Maitre:2006bh}.
+Lisbona \emph{et al.} later showed that BI-1 directly interacts with the cytosolic domain of IRE1$\alpha$, inhibiting its endoribonuclease activity \cite{Lisbona:2009cr}.
+Interestingly, BI-1 interacts with several members of the Bcl2 protein family e.g., Bcl2 and Bcl-X$_{L}$, even though it has no homology \cite{Xu:1998dq}.
+Members of the HSP family of proteins have also been shown to regulate IRE1$\alpha$.
+For example, HSP90 interacts with the cytosolic domain of IRE1$\alpha$, potentially protecting it from degradation by the proteasome \cite{Marcu:2002oq}.
+HSP72 interaction with the cytosolic IRE1$\alpha$ domain has also recently been shown to enhance IRE1$\alpha$ endoribonuclease activity \cite{Gupta:2010kl}.
+Taken together, these modes of IRE1$\alpha$ regulation with the exception of B1-1, largely promote or enhance IRE1$\alpha$ signaling.
+Given the importance of CHOP in regulation of Bcl2, it is vital to establish the exact connectivity.
+However, while CHOP expression is negatively correlated with Bcl2 levels, there is no CHOP binding site in the \emph{bcl2} promoter \cite{mccullough2001gsc}.
+McCullough \emph{et al.} have suggested that the bZIP domain of CHOP could act with other bZIP transcription factors to regulate \emph{bcl2} expression \cite{mccullough2001gsc}.
+Thus, it's likely that the connection between CHOP expression and apoptosis is more complex than simple down-regulation of Bcl2 expression.
+These missing structural connections shall allow us to establish a detailed model and extract more relevant insights into manipulating UPR.
 
 \clearpage
+\section*{Materials and Methods}
 
-\section*{Discussion}
+\subsection*{Formulation and solution of the model equations.}
+The unfolded protein response model was formulated as a set of coupled ordinary differential equations (ODEs):
+\begin{equation}\label{asseassertain_massbalance}
+\frac{d\mathbf{x}}{dt}=\mathbf{S}\cdot\mathbf{r}\left(\mathbf{x},\mathbf{p}\right)\qquad\mathbf{x}\left(t_{o}\right)=\mathbf{x}_{o}
+\end{equation}
+The symbol $\mathbf{S}$ denotes the stoichiometric matrix ($636\times 1090$). The quantity $\mathbf{x}$ denotes the concentration vector of proteins or protein complexes ($636\times 1$). 
+The term $\mathbf{r}\left(\mathbf{x},\mathbf{p}\right)$ denotes the vector of reaction rates ($1090\times{1}$). 
+Each row in $\mathbf{S}$ described a protein or protein-protein complex,
+while each column described the stoichiometry of network interactions. 
+Thus, the $(i,j)$ element of $\mathbf{S}$, denoted by $\sigma_{ij}$, described how protein $i$ was involved in rate $j$. 
+If $\sigma_{ij}<0$, then protein $i$ was consumed in $r_j$. Conversely, if $\sigma_{ij}>0$, protein $i$ was
+produced by $r_j$. Lastly, if $\sigma_{ij}=0$, there was no protein $i$ in rate $j$. All of these interactions were obtained from the literature.
+
+We assumed mass-action kinetics for each interaction in the network. 
+The rate expression for interaction $q$ was given by:
+\begin{equation}\label{eq-mass-action}
+r_{q}\left(\mathbf{x},k_{q}\right)=k_{q}\prod_{j\in\left\{\mathbf{R}_{q}\right\}}x_{j}^{-\sigma_{jq}}
+\end{equation}
+The set $\left\{\mathbf{R}_{q}\right\}$ denotes reactants for reaction $q$ while $\sigma_{jq}$ denotes the stoichiometric coefficient (element of the matrix $\mathbf{S}$) governing species $j$ in reaction $q$.
+All reversible interactions were split into two irreversible steps. The mass-action formulation, while expanding the dimension of the UPR model, 
+regularized the mathematical structure; this allowed automatic generation of the model code using UNIVERSAL and regularized the unknown model parameters (parameters were one of only three types, association, dissociation or catalytic rate constants).
+UNIVERSAL, an open source Java code generator, generates multiple code types from text and SBML inputs.
+UNIVERSAL is freely available as a Google Code project (http://code.google.com/p/universal-code-generator/). 
+Thus, although mass-action kinetics increased the number of parameters and species, they reduced the complexity of model analysis. 
+In this study, we considered well-mixed nuclear, cytosolic and extracellular compartments. 
+The model equations were solved using the LSODE routine in OCTAVE (v 3.1.0; www.octave.org) on an Apple workstation (Apple, Cupertino, CA; OS X v10.6.4).
+
+Unfolded protein response conditions were simulated by running the model to steady state and then providing a dose of proteins. 
+The steady-state was estimated numerically by repeatedly solving the model equations and estimating the difference between subsequent time points:
+\begin{equation}
+\Vert\mathbf{x}\left(t+\Delta{t}\right) - \mathbf{x}\left(t\right)\Vert_{2}\leq\gamma
+\end{equation}
+The quantities $\mathbf{x}\left(t\right)$ and $\mathbf{x}\left(t+\Delta{t}\right)$ denote the simulated concentration vector at time $t$ and $t + \Delta{t}$, respectively.
+The $L_{2}$ vector-norm was used as the distance metric. 
+We used $\Delta{t} = 1$ s and $\gamma$ = 0.001 for all simulations.
+
+\subsection*{Estimation and cross-validation of a population of models using Pareto Optimal Ensemble Techniques (POETs).}
+POETs is a multiobjective optimization strategy which integrates several local search strategies e.g., Simulated Annealing (SA) or Pattern Search (PS) 
+with a Pareto-rank-based fitness assignment \cite{Song:2010ij}.
+Denote a candidate parameter set at iteration $i+1$ as $\mathbf{k}_{i+1}$. 
+The squared error for $\mathbf{k}_{i+1}$ for training set $j$ was defined as:
+\begin{equation}\label{eqn_cost2}
+	E_{j}(\mathbf{k}) = \sum_{i=1}^{\mathcal{T}_{j}}\biggl(\hat{\mathcal{M}}_{ij}-\hat{y}_{ij}(\mathbf{k})\biggr)^2
+\end{equation}
+The symbol $\hat{\mathcal{M}}_{ij}$ denotes scaled experimental observations (from training set $j$) while the symbol $\hat{y}_{ij}$ denotes the scaled simulation output (from training set $j$). 
+The quantity $i$ denotes the sampled time-index and $\mathcal{T}_{j}$ denotes the number of time points for experiment $j$. 
+The read-out from the training immunoblots was band intensity where we assumed intensity was only loosely proportional to concentration.
+Suppose we have the intensity for species $x$ at time $i=\{t_{1},t_{2},..,t_{n}\}$ in condition $j$. The scaled measurement would then be given by:
+\begin{equation}\label{norm_exp_data}
+\hat{\mathcal{M}}_{ij} = \frac{\mathcal{M}_{ij} - \min_{i}\mathcal{M}_{ij}}{\max_{i}{\mathcal{M}_{ij}}-\min_{i}{\mathcal{M}_{ij}}}
+\end{equation}
+Under this scaling, the lowest intensity band equaled zero while the highest intensity band equaled one. 
+A similar scaling was defined for the simulation output.
+
+We computed the Pareto rank of $\mathbf{k}_{i+1}$ by comparing the simulation error at iteration $i+1$ against the simulation archive $\mathbf{K}_{i}$. 
+We used the Fonseca and Fleming ranking scheme \cite{fonseca1993genetic}:
+\begin{equation}\label{eqn_rank}
+rank\left(\mathbf{k}_{i+1}\mid \mathbf{K}_{i}\right)=p
+\end{equation}
+where $p$ denotes the number of parameter sets that dominate parameter set $\mathbf{k}_{i+1}$.
+Parameter sets on or near the optimal trade-off surface have small rank. 
+Sets with increasing rank are progressively further away from the optimal trade-off surface. 
+The parameter set $\mathbf{k}_{i+1}$ was accepted or rejected by the SA with probability $\mathcal{P}\left(\mathbf{k}_{i+1}\right)$: 
+\begin{equation}\label{eqn_costMOSA}
+\mathcal{P}(\mathbf{k}_{i+1}) \equiv \exp{\{-rank\left(\mathbf{k}_{i+1} \mid \mathbf{K}_{i} \right)/T\}}
+\end{equation}
+where $T$ is the computational annealing temperature. 
+The initial temperature $T_{o} = \mathit{n}/log(2)$, where $n$ is user defined ($n = 4$ for this study). The final temperature was $T_{f} = 0.1$.
+The annealing temperature was discretized into 10 quanta between $T_{o}$ and $T_{f}$ and adjusted according to the schedule
+$T_k = \beta^k T_0$
+where $\beta$ was defined as:
+\begin{equation}
+	\beta = \left(\frac{T_{f}}{T_{o}}\right)^{1/10}
+\end{equation}
+The epoch-counter $k$ was incremented after the addition of 100 members to the ensemble. 
+Thus, as the ensemble grew, the likelihood of accepting parameter sets with a large Pareto rank decreased.
+To generate parameter diversity, we randomly perturbed each parameter by $\leq\pm25\%$. 
+We performed a local pattern-search every $q$ steps to minimize the residual for a single randomly selected objective. 
+The local pattern-search algorithm has been described previously \cite{gadkar2003cybernetic, varner2000large}.
+The parameter ensemble used in the simulation and sensitivity studies was generated from the low-rank parameter sets in $\mathbf{K}_{i}$.
+
+We simultaneously calculated training and prediction error during the parameter estimation procedure using leave-eight-out cross-validation \cite{kohavi1995study}.
+The complete set of training data (33 objectives) was subdivided into four bins; in each bin 25 data sets were reserved for training while eight were reserved for prediction.  
+In the first bin DS$_{1}\hdots$DS$_{8}$ were used for validation while DS$_{9}\hdots$DS$_{33}$ were used for training.
+In the second bin DS$_{9}\hdots$DS$_{16}$ were used for validation while DS$_{1}\hdots$DS$_{8}~$DS$_{17}\hdots$DS$_{33}$ were used for training, etc.
+Thus, we formulated four ensembles from which we evenly selected parameter sets for the \emph{parent} ensemble (Fig. \ref{fg:fig2}).
+While cross-validation required that we generate additional model populations, we trained and tested against all the data sets.
+
+\subsection*{Sensitivity and robustness analysis of the population of EHR models.}
+Sensitivity coefficients were calculated as shown previously \cite{Song:2010ij} using five models selected from the ensemble (red points, supplementary materials Fig. S\ref{fg:Para_S1}). 
+The resulting sensitivity coefficients were scaled and time-averaged (Trapezoid rule):
+\begin{equation}\label{eqn_INsen}
+	\mathcal{N}_{ij} \equiv \frac{1}{T}\int^{T}_{0} dt \cdot |\alpha_{ij}\left(t\right)s_{ij}(t)|
+\end{equation}
+where $T$ denotes the final simulation time and $\alpha_{ij}=1$.
+The time-averaged sensitivity coefficients were then organized into an array for each ensemble member:
+\begin{equation}
+	\mathcal{N}^{\left(\epsilon\right)} = 
+	\begin{pmatrix}
+		\mathcal{N}_{11}^{\left(\epsilon\right)} & \mathcal{N}_{12}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{1j}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{1P}^{\left(\epsilon\right)} \\
+		\mathcal{N}_{21}^{\left(\epsilon\right)} & \mathcal{N}_{22}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{2j}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{2P}^{\left(\epsilon\right)} \\
+		\vdots & \vdots & & \vdots & & \vdots \\
+		\mathcal{N}_{M1}^{\left(\epsilon\right)} & \mathcal{N}_{M2}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{Mj}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{MP}^{\left(\epsilon\right)} \\
+	\end{pmatrix}\qquad\epsilon = 1,2,\hdots,N_{\epsilon}
+\end{equation}
+where $\epsilon$ denotes the index of the ensemble member, $P$ denotes the number of parameters, $N_{\epsilon}$ denotes the number of ensemble samples and $M$ denotes the number of model species.
+To estimate the relative fragility or robustness of species and reactions in the network, we decomposed the $\mathcal{N}^{\left(\epsilon\right)}$ matrix 
+using Singular Value Decomposition (SVD):
+\begin{equation}
+	\mathcal{N}^{\left(\epsilon\right)} = \mathbf{U}^{\left(\epsilon\right)}\Sigma^{\left(\epsilon\right)}\mathbf{V}^{T,{\left(\epsilon\right)}}
+\end{equation}
+Coefficients of the left (right) singular vectors corresponding to largest $\beta$ singular values of $\mathcal{N}^{\left(\epsilon\right)}$ 
+were rank-ordered to estimate important species (reaction) combinations. Only coefficients with magnitude greater than a threshold ($\delta$ = 0.1) were considered. 
+The fraction of the $\beta$ vectors in which a reaction or species index occurred was used to rank its importance.
+
+Robustness coefficients of the form:
+\begin{equation}
+\alpha\left(i,j,t_{o},t_{f}\right)=\left(\displaystyle\int_{t_{o}}^{t_{f}}x_{i}\left(t\right)dt\right)^{-1}\left(\displaystyle\int_{t_{o}}^{t_{f}}x^{(j)}_{i}\left(t\right)dt\right)
+\end{equation}
+were calculated to understand the robustness of the network. 
+The robustness coefficient $\alpha\left(i,j,t_{o},t_{f}\right)$ is the ratio of the integrated concentration of a network marker 
+in the presence (numerator) and absence (denominator) of structural or operational perturbation.
+The quantities $t_0$ and $t_f$ denote the initial and final simulation time respectively, while
+$i$ and $j$ denote the indices for the marker and the perturbation respectively. 
+If $\alpha\left(i,j,t_{o},t_{f}\right)>1$, then the perturbation \emph{increased} the marker concentration. Conversely,
+if $\alpha\left(i,j,t_{o},t_{f}\right)\ll{1}$ the perturbation \emph{decreased} the marker concentration. Lastly, if
+$\alpha\left(i,j,t_{o},t_{f}\right)\sim{1}$ the perturbation did not influence the marker concentration.
+
+\subsection*{Species clustering and dendrogram.}
+A dendrogram was derived by considering each of the knockouts(over-expressions) as variables and the average log of robustness coefficient (LRC) for each of the species as observations.
+We used the Euclidean norm in LRC space as the distance metric.
+The linkage function (objective function for identifying variable clusters) was the inner squared distance (minimum variance algorithm).
+The Statistical Toolbox of Matlab (The Mathworks, Natick, MA) was used to generate the distances, linkages and the final dendrogram.
+
+\subsection*{Identification of distinguishable species.}
+Robustness coefficients were used to rank-order knockout(overexpression) experiments in terms of the greatest unique responses and identify species which were linearly distinguishable.
+The response of the knockout(overexpression) was measured in terms of the robustness coefficients.
+The LRC had desirable linear properties, such that no response (no change in trajectories from wild-type) returns a value of zero and similar negative and positive responses have different directions but similar magnitudes.
+We considered the unique component of the response to be the orthogonal component in LRC space and the magnitude of the response to be the Euclidean norm.
+The orthogonal components and there magnitude were identified for each parameter set in the ensemble by first choosing the knockout(overexpression) with the greatest magnitude, $x_{1}$ and placing it in the empty set $\mathcal{V}$.
+The knockout(overexpression) $x_{1}$ defines the orthogonal directions in the LRC space.
+We then calculated the orthogonal components for all remaining knockouts(overexpressions) relative to $x_{1}$, and added the knockout(overexpression) species with the greatest orthogonal magnitude to set $\mathcal{V}$.
+In general the components of all remaining $x_{i}$ orthogonal to set $\mathcal{V}$ were calculated and the largest was moved into set $\mathcal{V}$.
+This process was continued until all knockout(overexpression) species, $x_{i}$ were added to set $\mathcal{V}$.
+Mathematically two species were considered distinguishable if and only if they were linearly independent (the orthogonal components were non-zero).
+We considered a threshold value of one or five and performed a student t-test (Matlab Statistical Toolbox, The Mathworks, Natick, MA) to identify which species had orthogonal components above the threshold with a 95\% confidence over the ensemble.
+
+\section*{Acknowledgements}
+The project described was supported by Award Number \#U54CA143876 from the National Cancer Institute.  
+The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Cancer Institute or the National Institutes of Health. 
+In addition, special thanks to Gaurav Chakravorty and Mansi Singhal for their generous support. 
+Lastly, the authors thank the anonymous reviewers for their helpful suggestions.
 
 \clearpage
-
-\section*{Materials and Methods}
-
-\subsection*{Formulation and solution of coagulation model equations.}
-\subsection*{Optimization methods.}
-
-\section*{Acknowledgements}
-This study was supported by the National Science Foundation GK12 award (DGE-1045513) 
-and by the National Science Foundation CAREER award (FILLMEIN).
+\bibliography{References_v1}
 
 \clearpage
-
-\bibliography{References_v1}
-
-%\begin{comment}
+\input{Figures_v1}
 
 \clearpage
+\input{Tables_v1}
 
-% Supplemental figures -
-% Set the S- 
-\renewcommand\thefigure{S\arabic{figure}}
-\renewcommand\thetable{T\arabic{table}}
-\renewcommand\thepage{S-\arabic{page}}
-\renewcommand\theequation{S\arabic{equation}}
-
-% Reset the counters -
-\setcounter{equation}{0}
-\setcounter{table}{0}
-\setcounter{figure}{0}
-\setcounter{page}{1}
+%\clearpage
+%\input{UPR_Modeling_Supplement_Material}
 

### 2014-07-21 Supplementary_Write_Up/UPR_Modeling_Supplement_Material.tex
First commit

\documentclass[12pt]{article}
% Load packages
%\usepackage{url}  % Formatting web addresses  
%\usepackage{ifthen}  % Conditional 
%\usepackage{multicol}   %Columns
%\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{epsfig}
%\usepackage{graphicx}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
%\usepackage{palatino,lettrine}
%\usepackage{times}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage[wide]{sidecap}
\usepackage[sort&compress,comma]{natbib}
\usepackage{longtable}
\usepackage{supertabular}
\usepackage{setspace}
%\usepackage{lineno}
%\urlstyle{rm}
%*** extra packages ****
%\usepackage{colortbl}
%\usepackage{color}

\textwidth = 6.50 in
\textheight = 9.5 in
\oddsidemargin =  0.0 in
\evensidemargin = 0.0 in
\topmargin = -0.50 in
\headheight = 0.0 in
\headsep = 0.25 in
%\parskip = 0.15in
%\linespread{1.75}
\doublespace

\bibliographystyle{biophysj}

\makeatletter
\renewcommand\subsection{\@startsection
	{subsection}{2}{0mm}
	{-0.05in}
	{-0.5\baselineskip}
	{\normalfont\normalsize\bfseries}}
\renewcommand\subsubsection{\@startsection
	{subsubsection}{2}{0mm}
	{-0.05in}
	{-0.5\baselineskip}
	{\normalfont\normalsize\itshape}}
\renewcommand\section{\@startsection
	{subsection}{2}{0mm}
	{-0.2in}
	{0.05\baselineskip}
	{\normalfont\normalsize\bfseries}}	
\renewcommand\paragraph{\@startsection
	{paragraph}{2}{0mm}
	{-0.05in}
	{-0.5\baselineskip}
	{\normalfont\normalsize\itshape}}
\makeatother

% Single space'd bib -
\setlength\bibsep{0pt}

\renewcommand{\rmdefault}{phv}\renewcommand{\sfdefault}{phv}

% Change the number format in the ref list -
\renewcommand{\bibnumfmt}[1]{#1.}

% Change Figure to Fig.
\renewcommand{\figurename}{Fig.}

% Begin ...
\begin{document}
\setcounter{page}{1}

\section*{Molecular basis of Eukaryotic Unfolded Protein Response (UPR).}
Protein folding is strategically important to cellular function. Secreted, membrane-bound and organelle-targeted proteins are typically processed and folded in the endoplasmic reticulum (ER) in eukaryotes \citep{naidoo2009er,ron2002translational,kaufman2002unfolded}. Intracellular perturbations caused by a variety of stressors disturb the specialized environment of the ER leading to the accumulation of unfolded proteins \citep{ellgaard2003qce,Fonseca:2009fk}. Normally, cells ensure that proteins are correctly folded using a combination of molecular chaperones, foldases and lectins \citep{naidoo2009er}. However, when proper folding can not be restored, incorrectly folded proteins are targeted to ER Associated Degradation (ERAD) pathways for processing \citep{kaufman2002unfolded}. If unfolded or misfolded proteins continue to accumulate, eukaryotes induce the unfolded protein response (UPR).

In mammalian cells, UPR is a complex signaling program mediated by three ER transmembrane receptors: activating transcription factor 6 (ATF6), inositol requiring kinase 1 (IRE1) and double-stranded RNA-activated protein kinase (PKR)-like endoplasmic reticulum kinase (PERK). UPR performs three functions, adaptation, alarm and apoptosis. During adaptation, the UPR tries to reestablish folding homeostasis by inducing the expression of chaperones that enhance protein folding. Simultaneously, translation is globally attenuated to reduce the ER folding load while the degradation of unfolded proteins is increased. If these steps fail, the UPR induces a cellular alarm and apoptosis program. The alarm phase involves several signal transduction events, ultimately leading to the removal of the translational block and the down-regulation of the expression and activity of pro-survival factors such as the B-cell lymphoma 2 (Bcl2) protein. After the alarm phase, cells can undergo apoptosis, although ER stress can also initiate autophagy \citep{ogata2006aac,yorimitsu2006ers,bernales2006ace,kamimoto2006iic,hoyerhansen2007cmc,kouroku2006esp,fujita2007ter}. Thus, ER folding homeostasis strongly influences physiology \citep{Fonseca:2009fk}. Aberrant protein folding and UPR have been implicated in a number of pathologies. For example, the onset of diabetes \citep{schnell2009model} as well as myocardial ischaemia, cardiac hypertrophy, atherosclerosis and heart failure \citep{glembotski2007ers} have all been linked with aberrant folding or UPR signaling.

\subsection*{The folding cycle, quality control and ER associated degradation (ERAD).} 
Newly synthesized polypeptide chains enter the ER through a peptide translocon in the ER membrane composed of four proteins, Sec61$\alpha$,$\beta$,$\gamma$ and TRAM \citep{Matlack:1998kx}. 
Upon entering the ER, these nascent chains begin to fold, often as they are being co-translationally modified \citep{Fedorov:1997ys}. The folding quality of proteins in the ER is maintained by an in-built quality control (QC) system which ensures proteins are in their native folded state before exiting the ER \citep{ellgaard1999ssq,ellgaard2003qce}. A protein is correctly folded, if it has attained its native conformation after required co- or post-translational modifications. On the other hand, exposed hydrophobic regions, unpaired cysteine residues, or aggregation are all markers of an unfolded or misfolded conformation \citep{ellgaard1999ssq}, which leads to subsequent retro-translocation to the cytosol. Once in the cytosol, these unfolded or misfolded proteins are degraded by the ubiquitin proteasome system \citep{Hershko:2000kl}. Hydrophobic unfolded or misfolded queues are recognized in the ER by molecular chaperones which bind these queues and increase the probability of correct folding \citep{fra1993qce,helenius1997cca,hellman1999vab}. For example, the HSP70 family of chaperones recognize, in an ATP-dependent manner, exposed hydrophobic patches on a broad spectrum of unfolded or misfolded proteins \citep{kaufman2002unfolded}. Repeated binding and release of HSP70 chaperones ensures that incorrectly folded proteins do not exit the ER \citep{kaufman2002unfolded}. One critical member of the HSP70 family is BiP or GRP78. BiP consists of an N-terminal ATPase domain and a C-terminal peptide binding domain \citep{GETHING1999}. BiP also regulates the activation of the three transmembrane ER stress transducers: PERK, ATF6 and IRE1. Normally, BiP is bound to these ER receptors, blocking their activation. However, in the presence of exposed hydrophobic residues BiP disassociates, allowing PERK, ATF6 and IRE1 activation. Overexpression of BiP leads to reduced activation of IRE1 and PERK \citep{bertolotti2000dib,kohno1993pry}. The PERK and ATF6 branches are thought to be activated before IRE1 \citep{szegezdi2006mediators}; this ordering is consistent with the signals that each branch transduces. The PERK and ATF6 pathways largely promote ER adaptation to misfolding, while IRE1 has a dual role, transmitting both survival and pro-apoptotic signals.

\subsection*{Double-stranded RNA-activated protein kinase (PKR)-like endoplasmic reticulum kinase (PERK) pathway:} The PERK branch of UPR transduces both pro-survival as well as pro-apoptotic signals following the accumulation of unfolded or misfolded protein in the ER. PERK is a type I transmembrane protein, composed of a ER luminal stress sensor and a cytosolic protein kinase domain. Dissociation of BiP from the N-terminus of PERK initiates dimerization and autophosphorylation of the kinase domain at T981 \citep{Kebache:2004kx}. The eIF2$\alpha$ protein, which is composed of three subunits, is critical to translation initiation in eukaryotes, including GTP-dependent start-site recognition \citep{Merrick:2004lr}. Activated PERK can phosphorylate eIF2$\alpha$ at S51 \citep{harding1999pta,Raven:2008fu}, which leads to three downstream effects. First, phosphorylated eIF2$\alpha$ globally attenuates translation initiation (Not included in the current model). Decreased translation reduces the influx of protein into the ER, hence diminishing the folding load. Translation attenuation is followed by increased clearance of the accumulated proteins from the ER by ERAD and expression of pro-survival genes. For example, PERK activation induces of the expression of cellular inhibitor of apoptosis (cIAP) \citep{hamanaka2008pdr}. Interestingly, decreased protein translation is not universal; genes with internal ribosome entry site (IRES) sequences in the 5$^{\prime}$ untranslated regions bypass the eIF2$\alpha$ translational block \citep{Schroder:2005vn}. One of the most well-studied of these, \emph{ATF4}, encodes a cAMP response element-binding transcription factor (C/EBP) \citep{lu2004tra} ATF4 that drives the expression of pro-survival functions such as amino acid transport and synthesis, redox reactions and protein secretion \citep{harding2003isr}. Taken together, these effects seem to be largely pro-survival. However, ATF4 can also induce the expression of pro-apoptotic factors. For example, ATF4 induces the expression of the transcription factor C/EBP homologous protein (CHOP), which is associated with apoptotic cell-death. CHOP (also known as GADD153) is 29 kDa protein composed of an N-terminal transcriptional activation domain and a C-terminal basic-leucine zipper (bZIP) domain that is normally present at low levels in mammalian cells \citep{ron1992cnd}. The transcriptional activator domain is positively regulated by phosphorylation at S78 and S81 by p38 MAPK family members \citep{wang272sip,maytin2001sit} while the bZIP domain plays a key role in the homodimerization of the protein \citep{matsumoto1996eec,maytin2001sit}. CHOP activity promotes apoptosis primarily by repression of Bcl2 expression and the sensitization of cells to ER-stress inducing agents \citep{gotoh:hdc,mccullough2001gsc}.

\subsection*{Activating transcription factor 6 (ATF6) pathway:} ATF6 activation involves a complex series of translocation and irreversible proteolytic processing steps, ultimately leading to the up-regulation of a pro-survival transcriptional program, in the presence of unfolded or misfolded proteins. ATF6 is a 90 kDa ER transmembrane protein with two homologs: ATF6$\alpha$ \citep{silver1999mtf,hai1989tfa} and ATF6$\beta$ \citep{min1995ncf,khanna1996ggc,kyosuke2001igc}. In the current model, only ATF6$\alpha$ is included. Similar to IRE1 and PERK, ER stress leads to the dissociation of BIP from the N-terminus of ATF6, followed by translocation and activation. N-terminal golgi localization sequences (GLS1 and GLS2) seem to be involved with BiP regulation of ATF6. BiP binding to the N-terminal GLS1 promotes the retention of ATF6 in the ER \citep{shen2002esr}. On the other hand, the GLS2 domain was required to target ATF6 to the golgi body following BiP dissociation from GLS1 \citep{shen2002esr}. Unlike the previous two kinase pathways, ATF6 activation does not involve phosphorylation of a C-terminal kinase domain. Rather, after translocated to the golgi, ATF6 undergoes regulated intramembrane proteolysis (RIP); the luminal domain is first cleaved by serine protease site-1 protease (S1P) followed by metalloprotease site-2 protease (S2P) cleavage \citep{silver1999mtf,ye2000esi,chen2002lda,shen2004dsp}. Cleavage at the juxtamembrane site allows the 50 kDa transcriptional domain of ATF6 to be translocated to the nucleus where it regulates the expression of genes with ATF/cAMP response elements (CREs) \citep{wang2000aaa} and  ER stress response elements (ERSE) in their promoters \citep{yoshida1998ica, kokame2001iei}. Cleaved ATF6 induces a gene expression program, in conjunction with other bZIP transcription factors and required co-regulators, such as nuclear factor Y (NF-Y) \citep{kokame2001iei,yoshida2000aap}, that increases chaperone activity as well as the degradation of unfolded proteins \citep{yamamoto2007tim,wu2007aol}. For example, ATF6 upregulates BiP, protein disulfide isomerase (PDI) and ER degradation-enhancing alpha-mannosidase-like protein 1 (EDEM1) expression. Additionally, ATF6 induces the expression of the X box-binding protein 1 (XBP1) which, after processing by activated IRE1$\alpha$, induces the expression of chaperones. The ATF6-induced gene expression program is also cytoprotective. For example, ATF6 induces regulator of calcineurin 1 (RCAN1) expression \citep{belmont2008cga}. RCAN1 sequesters calcineurin \citep{belmont2008cga}, a calcium activated protein-phosphotase B, that dephosphorylates Bcl2-antagonist of cell death (BAD) at S75 or S99 \citep{wang1999cia}. This leads to sequestering of Bcl2 by Bad, which inhibits its downstream anti-apoptotic activity \citep{wang1999cia}.

\subsection*{Inositol-requiring kinase 1 (IRE1) pathway:}
IRE1 initiates a program with both pro-survival and pro-apoptotic components in the presence of misfolded or unfolded proteins. IRE1 is a 100 kDa type I ER transmembrane protein with both an endoribonuclease and a serine-threonine kinase domain \citep{kaufman2002unfolded}. IRE1 has two homologs, IRE1$\alpha$ and IRE1$\beta$; IRE1$\alpha$ is expressed in a variety of tissues \citep{tirasophon1998srp} while IRE1$\beta$ is found only in the intestinal epithelia \citep{tirasophon1998srp,wang1998cmi}. In the current model only IRE1$\alpha$ has been considered. The N-terminus of IRE1, located in the ER lumen, senses unfolded or misfolded proteins through its interaction with BiP \citep{cox1993tig,shamu1996oap,sidrauski1997tki}. Normally BiP is bound to the N-terminus of IRE1 \citep{bertolotti2000dib,okamura2000dkb,liu2003sai}. However, in the presence of unfolding queues BiP dissociates and is sequestered by the unfolded or misfolded proteins \citep{kimata2003ger}. Subsequently, IRE1 is activated by homooligomerization followed by autophosphorylation of the C-terminal kinase domain at S724 \citep{shamu1996oap,welihinda1996unfolded,weiss1998sso,papa2003bka}. IRE1 activation enables both its kinase and endoribonuclease activities to transduce signals simultaneously through two distinct signaling axes. The endoribonuclease activity cleaves a 26-nucleotide intron from the XBP1-mRNA \citep{shen2001csp,yoshida2001xmi,lee2002imu} which generates a 41 kDa frameshift variant (sXBP1) that acts as a potent transcription factor. sXBP1 homodimers, along with co-regulators such as nuclear factor Y (NF-Y), regulate the expression of a variety of ER chaperones and protein degradation related genes \citep{malhotra2007era,rao2004mpe}. Cytosolic IRE1$\alpha$ dimers interact with adaptors such as tumor necrosis factor receptor-associated factor 2 (TRAF2) to drive signal-regulating kinase (ASK1) activation and then subsequently cJUN NH$_{2}$-terminal kinase (JNK) and p38MAPK activation \citep{Urano:2000uq}. ASK1 activity is regulated by phosphorylation/de-phosphorylation at several sites as well as by physical interaction with other proteins. ASK1 phosphorylates and activates two downstream kinases, MMK4 and MMK3 which in turn activate JNK and p38 MAP kinase, respectively. JNK is activated by dual phosphorylation at T183 and Y185 by MMK4 \citep{Derijard:1995fk}. Activated JNK activates the proapoptotic Bcl-2 family member Bim by phosphorylation at S65 \citep{lei2003jpb,putcha2003jmb}. JNK activation also regulates the activity of anti-apoptotic protein Bcl2 \citep{yamamoto1999bpa, Wei:2008hc}. Active JNK1 inhibits Bcl2 via phosphorylation at sites T69, S70 and S87 \citep{Wei:2008hc}. Ultimately, inhibition of Bcl2 and the activation of Bim leads to BAX/BAK dependent apoptosis. Thus, signals initiated from the cytosolic kinase domain of IRE1$\alpha$ are largely pro-apoptotic. IRE1$\alpha$ activity is regulated by protein serine/threonine phosphatase (PTC2P).

\subsection*{ER stress-induced apoptosis:} Ultimately, if UPR fails to restore ER homeostasis, cells initiate terminal programs such as apoptosis. A common biomarker of apoptosis is the activation of aspartate-specific proteases, collectively known as caspases \citep{alnemri1996hic}. Caspases rapidly dismantle cell cycle, cytoskeletal and organelle proteins by proteolytic cleavage. There are two pathways that result in caspase activation in response to apoptotic signals; the death-receptor and the stress mediated pathways. The death-receptor pathway is marked by ligand-mediated activation of death receptors on the plasma membrane. The alternative pathway for caspase activation is mediated by cellular stress e.g., ER stress. Caspases are activated from their zymogens (procaspases), in response to various death cues. First, the initiator caspases, caspase-8 and caspase-9, are activated in response to death cues \citep{muzio1998ipm}. This is followed by the activation of executioner caspases, such as caspase-3, caspase-6 and caspase-7. Activated executioner caspases proteolytically process several substrates, facilitating cell death. They also activate initiator caspases, forming a positive feedback loop. Activation of both the PERK and IRE1 pathways modulate stress-induced apoptosis through their regulation of Bcl2 expression and activity. Overall, stress induced apoptosis can occur through both mitochondrial-dependent and independent pathways. Stress signals cause oligomerization of pro-apoptotic proteins, such as Bax and Bak. These proteins are normally sequestered at the mitochondrial outer membrane by the survival protein Bcl2, under non-apoptotic conditions \citep{wei2001proapoptotic}. Once Bax and Bak oligomerize, they insert into the mitochondrial membrane and breach membrane integrity \citep{nechushtan1999conformation}. This results in a net efflux of cytochrome-c from the mitochondria to the cytosol and the initiation of the well-studied Apaf-1 mediated caspase-9 activation pathway. Stress induced mitochondrial-independent apoptotic pathways are not well understood. Currently, caspase 12 has been suggested as a possible ER-stress apoptotic mediator \citep{szegezdi2006mediators, yoneda2001ace, nakagawa2000caspase}. However, caspase 12 is not expressed in human. Moreover, there is considerable debate about its role in stress-induced apoptotic cell-death \citep{Saleh:2006ys}.

\clearpage
% \bibliographystyle{plain}

\bibliography{References}

\end{document}

### 2014-07-21 Tables_v1.tex
First commit

\begin{center}
\tablefirsthead{%
\hline
\rowcolor[gray]{0.8}\textbf{Obj\#}  & \textbf{species} & \textbf{cell~type} & \textbf{nominal} & \textbf{training} & \textbf{prediction}  & \textbf{random} & \textbf{source} \\
\hline}
\tablehead{%
\multicolumn{8}{l}{\small \emph{continued from previous page}} \\
\hline
\rowcolor[gray]{0.8}\textbf{Obj\#}  & \textbf{species} & \textbf{cell~type} & \textbf{nominal} & \textbf{training} & \textbf{prediction}  & \textbf{random} & \textbf{source} \\
\hline}
\tabletail{%
\hline
\multicolumn{8}{r}{\small\emph {continued on next page}}\\
}
\tablelasttail{\hline}
\topcaption{Objective function list along with species, cell-type, nominal error, training error, prediction error, random error with a randomly generated parameter set and the corresponding literature reference.}\label{obj-table}
\begin{scriptsize}
\begin{supertabular}{l l l l l l l l l l}  
\hline
O1  & ATF6 free & HEK293 & 0.25 & 0.25 $\pm$ 0.04 & 0.26 $\pm$ 0.08 & 0.49 $\pm$ 0.05 & \cite{lin2007ire1} \\
O2  & peIF2$\alpha$ & HEK293 & 0.14 & 0.23 $\pm$ 0.08 & 0.25 $\pm$ 0.11 & 0.51 $\pm$ 0.11 & \cite{lin2007ire1} \\
O3  & BiP mRNA & HEK293 & 0.24 & 0.43 $\pm$ 0.17 & 0.49 $\pm$ 0.15 &  0.72 $\pm$ 0.13 & \cite{lin2007ire1} \\
O4  & CHOP mRNA & HEK293 & 0.55 & 0.56 $\pm$ 0.07 & 0.57 $\pm$ 0.06 & 0.68 $\pm$ 0.20 & \cite{lin2007ire1} \\
O5  & Cleaved PARP & HEK293 & 0.24 & 0.23 $\pm$ 0.14 & 0.23 $\pm$ 0.17 & 0.50 $\pm$ 0.19 & \cite{lin2009divergent} \\
O6  & pIRE1$\alpha$ & AR42J & 0.37 & 0.36 $\pm$ 0.05 & 0.35 $\pm$ 0.07 & 0.63 $\pm$ 0.09 & \cite{bertolotti2000dib} \\
O7  & PERK & AR42J & 0.19 & 0.36 $\pm$ 0.27 & 0.38 $\pm$ 0.29 & 0.46 $\pm$ 0.20  & \cite{bertolotti2000dib} \\
O8  & pPERK & AR42J & 0.10 & 0.14 $\pm$ 0.09 & 0.15 $\pm$ 0.15 & 0.28 $\pm$ 0.08  & \cite{bertolotti2000dib} \\
O9  & BIP mRNA & MEF & 0.19 & 0.30 $\pm$ 0.14 & 0.26 $\pm$ 0.07 & 0.66 $\pm$ 0.24 & \cite{yamamoto2004Differential} \\
O10  & BIP Protein & P19 EC & 0.42 & 0.34 $\pm$ 0.15 & 0.35 $\pm$ 0.11 & 0.68 $\pm$ 0.15 & \cite{jimbo2003er} \\
O11 & CHOP Protein & P19 EC & 0.24 & 0.42 $\pm$ 0.20 & 0.38 $\pm$ 0.18 & 0.59 $\pm$ 0.20 & \cite{jimbo2003er} \\
O12 & sXBP1 protein & MEF & 0.34 & 0.29 $\pm$ 0.08 & 0.30 $\pm$ 0.06 & 0.40 $\pm$ 0.11  & \cite{calfon2002ire1} \\
O13 & CHOP Protein & MEF & 0.46 & 0.43 $\pm$ 0.12 & 0.45 $\pm$ 0.13 & 0.56 $\pm$ 0.18  & \cite{calfon2002ire1} \\
O14 & XBP1 mRNA & MEF &  0.57 & 0.48 $\pm$ 0.13 & 0.49 $\pm$ 0.13 & 0.47 $\pm$ 0.05  & \cite{calfon2002ire1} \\
O15 & peIF2$\alpha$ & MEF  & 0.38 & 0.37 $\pm$ 0.08 & 0.38 $\pm$ 0.03 & 0.44 $\pm$ 0.11 & \cite{yamamoto2007tim} \\
O16 & ATF4 nuclear & MEF & 0.63 & 0.67 $\pm$ 0.06 & 0.67 $\pm$ 0.07 & 0.68 $\pm$ 0.16  & \cite{yamamoto2007tim} \\
O17 & CHOP Protein & MEF & 0.52 & 0.51 $\pm$ 0.12 & 0.53 $\pm$ 0.08 & 0.64 $\pm$ 0.19 & \cite{yamamoto2007tim} \\
O18 & BIP Protein & MEF & 0.18 & 0.34 $\pm$ 0.18 & 0.34 $\pm$ 0.23 & 0.55 $\pm$ 0.16 & \cite{yamamoto2007tim} \\
O19 & Spliced XBP1 & MEF & 0.49 & 0.49 $\pm$ 0.04 & 0.48 $\pm$ 0.04 & 0.50 $\pm$ 0.10 & \cite{lee2002imu} \\
O20 & Cleaved ATF6 & MEF & 0.45 & 0.44 $\pm$ 0.06 & 0.44 $\pm$ 0.05 & 0.49 $\pm$ 0.11 & \cite{lee2002imu} \\
O21 & cIAP Protein & NIH 3T3 & 0.36 & 0.25 $\pm$ 0.15 & 0.30 $\pm$ 0.10 & 0.52 $\pm$ 0.15 & \cite{hamanaka2008pdr} \\
O22 & Caspase 12 Protein & Transfected HEK 293T & 0.44 & 0.57 $\pm$ 0.11 & 0.52 $\pm$ 0.11 & 0.75 $\pm$ 0.11 & \cite{Rao:2001fk} \\
O23 & Phospho eIF2$\alpha$ & Liver of Mice & 0.50 & 0.51 $\pm$ 0.07 & 0.51 $\pm$ 0.04 & 0.50 $\pm$ 0.12  & \cite{hetz2006pba} \\
O24 & Procaspase 12 & SAK2 & 0.28 & 0.46 $\pm$ 0.28 & 0.38 $\pm$ 0.26 & 0.78 $\pm$ 0.14 & \cite{rao2002cer} \\
O25 & Caspase 12 & SAK2 & 0.12 & 0.22 $\pm$ 0.12 & 0.25 $\pm$ 0.13 & 0.73 $\pm$ 0.07 & \cite{rao2002cer} \\
O26 & Procaspase 9 & SAK2 & 0.14 & 0.28 $\pm$ 0.22 & 0.33 $\pm$ 0.26 & 0.78 $\pm$ 0.13  & \cite{rao2002cer} \\
O27 & Caspase 9 & SAK2 & 0.03 & 0.19 $\pm$ 0.16 & 0.24 $\pm$ 0.18 & 0.68 $\pm$ 0.14  & \cite{rao2002cer} \\
O28 & Procaspase 7 & SAK2 &  0.31 & 0.44 $\pm$ 0.24 & 0.43 $\pm$ 0.23 & 0.74 $\pm$ 0.14 & \cite{rao2002cer} \\
O29 & Caspase 7 & SAK2 & 0.64 & 0.61 $\pm$ 0.07 & 0.58 $\pm$ 0.11 & 0.75 $\pm$ 0.07 & \cite{rao2002cer} \\
O30 & Procaspase 3 & SAK2 & 0.34 & 0.46 $\pm$ 0.19 & 0.46 $\pm$ 0.21 & 0.82 $\pm$ 0.13 & \cite{rao2002cer} \\
O31 & Caspase 3 & SAK2 & 0.34 & 0.39 $\pm$ 0.09 & 0.42 $\pm$ 0.13 & 0.74 $\pm$ 0.10 & \cite{rao2002cer} \\
O32 & PARP Protein & SAK2 & 0.37 & 0.51 $\pm$ 0.17 & 0.51 $\pm$ 0.14 & 0.85 $\pm$ 0.13 & \cite{rao2002cer} \\
O33 & Cleaved PARP & SAK2 & 0.11 & 0.21 $\pm$ 0.17 & 0.31 $\pm$ 0.21 & 0.67 $\pm$ 0.15 & \cite{rao2002cer} \\

\end{supertabular}
\end{scriptsize}
\end{center}

\clearpage

\begin{center}
\tablefirsthead{%
\hline
\rowcolor[gray]{0.8}\textbf{Increased marker levels}  & \textbf{Phenotype} & \textbf{G.O.S (\% of cases)} & \textbf{Example genes} & \textbf{G.K.S (\% of cases)} & \textbf{Example genes}   \\
\hline}
\tablehead{%
\multicolumn{6}{l}{\small \emph{continued from previous page}} \\
\hline
\rowcolor[gray]{0.8}\textbf{Increased marker levels}  & \textbf{Phenotype} & \textbf{G.O.S (\% of cases)} & \textbf{Example genes} & \textbf{G.K.S (\% of cases)} & \textbf{Example genes}   \\
\hline}
\tabletail{%
\hline
\multicolumn{6}{r}{\small\emph {continued on next page}}\\
}
\tablelasttail{\hline}
\topcaption{Phenotypic response of simulated Gene knockout/overexpression. (\textbf{G.O.S} - Gene Overexpression Studies, \textbf{G.K.S} - Gene Knockout Studies)}
\label{KO-O-table}
\begin{scriptsize}
\begin{supertabular}{|p{2cm}| p{2cm}| p{1.5cm}| p{3cm}| p{1.5cm}| p{3cm}|}  
\hline

BCL2 & Cell Survival & 16 & PERK, ATF4, cIAP, Bcl2, ATF6, RCAN1, Bad, UB & 24 & HSP40, BiP, PERK, eIF2$\alpha$, ATF4, NFY, JNK, Bad, CHOP, CREB, Procaspase 8  \\
\hline
CHOP & -- & 5 & eIF2$\alpha$, CHOP & 3.5 & CEBP, CREB  \\
\hline
Caspase 3 & Cell Death & 9 & Procaspase 9/8/3/6/7 & 0 & --  \\
\hline
XBP1 & -- & 12 & BiP, cIAP, XBP1, RCAN1 & 3.5 & NFY, TRAF2  \\
\hline
BiP & -- & 12 & PERK, IRE1, NFY, ATF6, CEBP, CREB, UB & 12 & HSP40, BiP, eIF2$\alpha$, ATF4, NFY, S1P, S2P  \\
\hline
PARP & DNA damage & 3.5 & Procaspase 9, PARP & 0 & -- \\
\hline
ATF4 & -- & 20 & HSP40, PERK, BiP, eIF2$\alpha$, ATF4, IRE1, BAK, TRAF2, ASK1, JNK, p38MAPK, Calcinuerin, CHOP, CREB & 5 & NFY, CEBP, CREB \\

\end{supertabular}
\end{scriptsize}
\end{center}

### 2014-07-24 Paper_v1.tex
Trying to fix this ...

--- 2014-07-21 Paper_v1.tex
@@
 \begin{titlepage}
-{\par\centering\textbf{\Large Dynamically Dimensioned Particle Swarm Optimization}}
+{\par\centering\textbf{\Large An Effective Model of the Retinoic Acid induced HL-60 Differentiation Circuit}}
 \vspace{0.05in}
-{\par \centering \large{Adithya Sagar, Christine Shoemaker and Jeffrey D. Varner$^{*}$}}
+{\par \centering \large{Ryan Tasseff$^{1}$, 
+Holly A. Jensen$^{1}$, 
+Johanna Congleton$^{2}$, Andrew Yen$^{2}$ and Jeffrey D. Varner$^{1,*}$}}
 \vspace{0.10in}
-{\par \centering \large{School of Chemical and Biomolecular Engineering}}
+{\par \centering \large{$^{1}$~School of Chemical and Biomolecular Engineering}}
 {\par \centering \large{Cornell University, Ithaca NY 14853}}
 \vspace{0.1in}
-{\par \centering \textbf{Running Title:}~Modeling cell free metabolism}
+{\par \centering \large{$^{2}$~Department of Biomedical Sciences}}
+{\par \centering \large{Cornell University, Ithaca NY 14853}}
 \vspace{0.1in}
-{\par \centering \textbf{To be submitted:}~\emph{BMC~Systems~Biology}}
+{\par \centering \textbf{Running Title:}~Effective HL60 differentiation circuit}
+\vspace{0.1in}
+{\par \centering \textbf{To be submitted:}~\emph{Processes}}
 \vspace{0.5in}
@@
 
-{\noindent \textbf{Keywords:}~Parameter identification, Mathematical modeling}
+{\noindent \textbf{Keywords:}~Cell free metabolism, Mathematical modeling}
 
@@
 
+Understanding differentiation, 
+the process by which precursor cells become more specialized cell types, 
+is an important challenge facing biology. 
+If differentiation programs could be rationally manipulated, 
+advanced therapies could be developed to treat a spectrum of cancers, spinal cord injuries and neurodegenerative disorders. 
+However, to rationally reprogram differentiation networks, 
+we must first understand their connectivity and regulation \cite{Young2011}. 
+Lessons learned in model systems, such as the lineage-uncommitted human myeloblastic cell line HL-60,
+could inform our analysis of more complex programs. 
+HL-60 has been a durable experimental model since the 1970's to study differentiation \cite{Breitman1980}.
+Depending upon the stimulus, HL-60 cells undergo G1/G0-arrest and myeloid or monocytic differentiation. 
+All-Trans Retinoic Acid (ATRA) induces G1/G0-arrest and myeloid differentiation in HL-60 cells, 
+whereas 1,25-dihydroxy vitamin D3 induces arrest with monocytic differentiation.
+Commitment to cell cycle arrest and terminal differentiation requires approximately 48 hr of treatment,
+during which HL-60 cells undergo approximately two division cycles.  
+Interestingly, cells treated with ATRA for time periods shorter than the commitment phase retain a limited inheritable memory, 
+which reduces the time required to reach commitment during subsequent ATRA exposure \cite{YEN1984}.
+
+Sustained Mitogen-Activated Protein Kinase (MAPK) cascade
+activation is a defining feature of ATRA-induced differentiation in HL-60 cells. 
+ATRA drives slow yet sustained MEK-dependent activation of the Raf/MEK/ERK pathway, 
+leading to arrest and functional differentiation \cite{Yen1998}.
+MEK inhibition results in the loss of both ERK and Raf phosphorylation, 
+as well as the failure to arrest and terminally differentiate \cite{Yen1998,Hong2001}.
+At the transcriptional level, 
+ATRA (and its metabolic products) are ligands for the hormone activated nuclear transcription factors Retinoic Acid Receptor (RAR) and Retinoid X Receptor (RXR) \cite{Mangelsdorf1990}.
+Activation of both RAR and RXR is necessary for ATRA-induced Raf phosphorylation and MAPK activation,
+suggesting that the initiation of MAPK signaling is partially transcriptionally regulated \cite{Hong2001}.
+ATRA, through activation of a transcription factor complex including RAR and RXR, induces the expression of many proteins.
+BLR1 also known as CXCR5, is a putative heterotrimeric Gq protein-coupled receptor 
+that is necessary for MAPK activation, growth arrest and functional differentiation \cite{YEN1990,LIPP1994,WANG2004}.
+BLR1 was identified as an early ATRA (or D3)-inducible gene in HL-60 cells using differential display \cite{YEN1990}. 
+Studies of the BLR1 promoter identified a 5' 17bp GT box approximately 1 kb upstream of the transcriptional start that conferred ATRA responsiveness \cite{WANG2004}.
+Additionally, members of the BLR1 transcriptional activator complex, e.g. NFATc3 and CREB, 
+can be phosphorylated by ERK, JNK or p38 MAPK family members \cite{Yang2002}.
+This suggests positive feedback between BLR1 expression and MAPK activation. 
+BLR1 overexpression enhanced Raf phosphorylation and accelerated terminal differentiation, 
+while BLR1 knock-out HL-60 cells failed to activate Raf or differentiate in the presence of ATRA \cite{Wang2008}.
+Interestingly, both the knockdown or inhibition of Raf, also reduced BLR1 expression and functional differentiation \cite{Wang2008}.
+A recent computational study of ATRA-induced differentiation in HL-60 cells suggested that the BLR1-MAPK positive feedback circuit was sufficient to explain ATRA-induced sustained MAPK activation and the expression of differentiation markers \cite{Tasseff2011}.
+Model analysis also suggested that Raf was the most distinct of the MAPK proteins.
+
+A critical question is what other components of the MAPK positive feedback circuit are required to drive ATRA-induced functional differentiation of HL-60 cells. 
+Wang and Yen showed that ectopic expression of the constitutively active CR3 domain of Raf1 restored ATRA-induced G0 arrest and differentiation in BLR1 knock-out cells \cite{Wang2008}. 
+However, ectopic expression of Raf1 CR3 alone, in the absence of ATRA, failed to induce arrest or differentiation.
+Thus, additional ATRA-inducible components must exist, 
+which independently promote arrest and differentiation in the absence of BLR1.
+In this study, we explored this hypothesis using a combination of experimental and computational tools. 
+First, we explored the ATRA-inducible Raf interactome 
+by surveying a panel of 19 possible binding partners using immunoprecipitation (IP), 
+with and without ATRA and the Raf inhibitor GW5074.  
+Initially, we expected increased ATRA-dependent association between Raf and kinases linked to BLR1 activity; 
+however, this was not supported by data.
+Instead, we found that the interaction between the guanine nucleotide exchange factor Vav1 and Raf was both ATRA-inducible and simultaneously sensitive to Raf inhibition. 
+Next, we considered how MAPK activation and differentiation were affected by the inhibition of Raf kinase activity in the presence and absence of ATRA.
+We showed that Raf activity was directly proportional to ERK phosphorylation and to functional differentiation processes such as the generation of reactive oxygen species (ROS). 
+Moreover, interactions between Raf and kinase partners such as Akt or CK2, or the scaffolding protein 14-3-3 were largely insensitive to ATRA treatment. 
+These studies established the working hypothesis that Vav1 (or potentially other ATRA-inducible proteins) acted as limiting members of a constitutively assembled trigger complex that propelled sustained MAPK activation, arrest and differentiation.
+We tested this hypothesis by constructing a mechanistic mathematical model of the Raf-Vav1 circuit, 
+based on the IP and Western blot data presented in this study, 
+and from previous literature. 
+The proposed model architecture was consistent with the ATRA-induced sustained MAPK activation observed experimentally. 
+Additionally, we found the Raf-Vav1 circuit possessed interesting dynamic features such as bistability, 
+that could explain ATRA-induction memory effects.
+
 \clearpage
@@
 \section*{Results}
-
-Coagulation is an archetype biochemical network that is highly interconnected and tightly regulated with multiple positive and negative feedback loops. 
-The biochemistry underlying coagulation has been well studied [REFHERE], and reliable experimental coagulation models have been developed [REFHERE].
-
-This makes it an ideal system for mathematical modeling and parameter estimation.
-Coagulation is regulated by a set of serine proteases also known as coagulation factors and blood platelets.
-The coagulation factors are generally in an inactive state and are known as zymogens.
-These zymogens are activated when through certain triggers.
-These trigger events like injury or trauma or sepsis exposure factors like collagen, tissue factor and von Willebrand factor (vWF) to blood.
-The exposure of these factors to blood kick-starts a series of convergent cascades that lead to conversion of zymogen prothrombinase to thrombin.
-Luan et al.
-modeled coagulation using coupled non-linear ordinary differential equations with 301 reactions and 192 species.
 
@@
 
-\subsection*{Formulation and solution of coagulation model equations.}
-\subsection*{Optimization methods.}
+\subsection*{Cell Culture and Treatment.}
+Human myeloblastic leukemia cells (HL-60 cells) were grown in a humidified atmosphere of 5\% CO$_2$ at 37$^{o}$C and maintained in RPMI 1640 from Gibco (Carlsbad, CA) 
+supplemented with 5\% fetal bovine serum from Hyclone (Logan, UT) and 1x antibiotic/antimicotic (Sigma, St. Louis, MO).
+Cells were cultured in constant exponential growth as described previously \cite{Brooks1996}. 
+Experimental cultures were initiated at $0.1\times10^6$ cells/mL 24 hr prior to 1$\mu$M ATRA treatment;
+if indicated, cells were also treated with GW5074 (2$\mu$M) 18 hr before ATRA treatment.   
+For cell culture washout experiments HL-60 cells were treated with ATRA for 24 hr, 
+washed 3x with prewarmed serum supplemented culture medium to remove ATRA exposure, 
+and reseeded in ATRA-free media as described. 
+Western blot analysis was performed at incremental time points after removal of ATRA.
+
+\subsection*{Chemicals.}
+All-Trans Retinoic Acid from Sigma-Aldrich (St. Louis, MO) was dissolved in 100\% ethanol with a stock concentration of 5mM, 
+and used at a final concentration of 1$\mu$M.
+The Raf inhibitor GW5074 from Sigma-Aldrich (St. Louis, MO) was dissolved in DMSO with a stock concentration of 10mM, 
+and used at a final concentration of 2$\mu$M.
+HL-60 cells were treated with 2$\mu$M GW5074 with or without ATRA (1$\mu$M) at 0 hr.  
+This GW5074 dosage had a negligible effect on the cell cycle distribution, compared to ATRA treatment alone (Fig. \ref{F:MAPKGWWB}A).
+
+\subsection*{CD11b expression studies by flow cytometry.}
+Approximately $1.0\times10^6$ HL-60 cells were harvested  by centrifugation. 
+Cells were resuspended in 200$\mu$L PBS containing 5$\mu$L of allophycocyanin (APC)–conjugated anti-CD11b antibody from BD Biosciences (San Jose, CA). 
+Following incubation for 1hr at 37$^{o}$C, cells were analyzed by flow cytometry (LSRII flow cytometer, BD Biosciences; San Jose, CA) using 633nm red laser excitation. 
+The threshold on experimental groups was set to exclude 95\% of control, the untreated sample.
+
+\subsection*{Measurement of inducible oxidative metabolism.}
+Approximately $0.5\times10^6$ Cells were harvested by centrifugation.
+Cells were resuspended in 200$\mu$L 37$^{o}$C PBS containing 10$\mu$mol/L 
+5-(and-6)-chloromethyl-2,7-dichlorodihydro-fluorescein diacetate acetyl ester (DCF; Invitrogen Carlsbad, CA) 
+and 0.4$\mu$g/mL 12-O-tetradecanoylphorbol-13-acetate (TPA; Sigma-Aldrich St. Louis, MO)
+with incubation for 20 min in a humidified atmosphere of 5\% CO$_2$ at 37$^{o}$C. 
+Flow cytometric analysis was done as described previously \cite{Reiterer2007}.
+To determine TPA inducible ROS, the threshold on experimental groups was set to exclude 95\% of control,
+samples not treated with TPA.
+
+\subsection*{Cell cycle analysis.}
+Approximately $1.0\times10^6$ cells were collected by centrifugation.
+Cells were resuspended in 500$\mu$L hypotonic staining solution containing 50$\mu$g/mL propidium iodine, 
+1$\mu$L/mL Triton X-100, and 1 mg/mL sodium citrate. 
+Cells were incubated at room temperature for 1 hr and analyzed by flow cytometry (BD LSRII) using 488-nm excitation.
+
+\subsection*{Immunoprecipitation and Western blot.}
+Approximately $1.2 \times 10^7$ cells were lysed using 400$\mu$L of M-Per lysis buffer from Thermo Scientific (Waltham, MA). 
+Lysates were cleared by centrifugation at 16,950 $\times$ g in a micro-centrifuge for 20 min at 4$^{o}$C.  
+Lysates were pre-cleared using 100$\mu$L protein A/G Plus agarose beads from Santa Cruz Biotechnology (Santa Cruz, CA) by 
+inverting overnight at 4$^{o}$C.  Beads were cleared by centrifugation and total protein concentration
+was determined by a BCA assay (Thermo Scientific, Waltham, MA).  Immunoprecipitations were setup by bringing 
+lysate to a concentration of 1.0g/L in a total volume of 300$\mu$L (M-Per buffer was used for dilution).     
+The anti-Raf antibody was added at 3 $\mu$L.
+A negative control with no bait protein was also used to exclude the direct interaction of proteins with the A/G beads.   
+After 1 hr of inversion at 4$^{o}$C, 20$\mu$L of agarose beads was added and samples were left 
+to invert overnight at 4$^{o}$C.  
+Samples were then washed three times with M-Per buffer by centrifugation.
+Finally proteins were eluted from agarose beads using a laemmli loading buffer.
+Eluted proteins were resolved by SDS-PAGE and Western blotting.  Total 
+lysate samples were normalized by total protein concentration (20$\mu$g per sample)
+and resolved by SDS-PAGE and Western blotting.  Secondary HRP bound antibody was used for visualization.
+All antibodies were purchased from Cell Signaling (Boston, MA) with the exception of anti-p621 Raf which was purchased from Biosource/Invitrogen  (Carlsbad, CA), and
+anti-pS338 Raf which was purchased from Santa Cruz Biotechnology (Santa Cruz, CA); anti-retinoblastoma from Zymed (South San Francisco, CA); and anti-CK2 from BD Biosciences (San Jose, CA).
+
+\subsection*{Formulation and solution of the model equations.}
+We used ordinary differential equations (ODEs) to model the time evolution of metabolite ($x_{i}$) and scaled enzyme abundance ($\epsilon_{i}$) in hypothetical cell free
+metabolic networks:
+\begin{eqnarray}
+	\frac{dx_{i}}{dt} & = & \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)\qquad{i=1,2,\hdots,\mathcal{M}}\\
+	\frac{d\epsilon_{i}}{dt} & = & -\lambda_{i}\epsilon_{i}\qquad{i = 1,2,\hdots,\mathcal{E}}
+\end{eqnarray}where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of metabolites and $\mathcal{E}$ denotes the number of
+enzymes in the model. The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ 
+denotes the rate of reaction $j$. Typically, reaction $j$ is a non-linear function of metabolite and enzyme abundance, as well as unknown kinetic parameters 
+$\mathbf{k}$ ($\mathcal{K}\times{1}$).
+The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$. If 
+$\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$. Conversely, if $\sigma_{ij}>0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates
+metabolite $i$ is not connected with reaction $j$. Lastly, $\lambda_{i}$ denotes the scaled enzyme degradation constant. The system material balances were subject to the
+initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$ and $\mathbf{\epsilon}\left(t_{o}\right) = \mathbf{1}$ (initially we have 100\% cell-free enzyme abundance).
+
+Each reaction rate was written as the product of two terms, a kinetic term ($\bar{r}_{j}$) and a regulatory term ($v_{j}$):
+\begin{equation}\label{eqn:rate-factor}
+	r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right) = \bar{r}_{j}v_{j}
+\end{equation}We used multiple saturation kinetics to model the reaction term $\bar{r}_{j}$:
+\begin{equation}\label{eqn:rate-bar}
+	\bar{r}_{j} = k_{j}^{max}\epsilon_{i}\left(\prod_{s\in{m_{j}^{-}}}\frac{x_{s}}{K_{js} + x_{s}}\right)
+\end{equation}where $k_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the scaled enzyme activity which catalyzes reaction $j$, and
+$K_{js}$ denotes the saturation constant for species $s$ in reaction $j$. 
+The product in Eqn. \eqref{eqn:rate-bar} was carried out over the set of \textit{reactants} for reaction $j$ (denoted as $m_{j}^{-}$).
+
+The allosteric regulation term $v_{j}$ depended upon the combination of factors which influenced the activity of enzyme $i$.
+For each enzyme, we used a rule based approach to select from competing control factors (Fig. \ref{fig-control-schematic}). 
+If an enzyme was activated by $m$ metabolites, we modeled this activation as:
+\begin{equation}
+	v_{j} = \max\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right)
+\end{equation}where $0\leq f_{ij}\left(x\right)\leq 1$ was a regulatory transfer function that calculated the influence of metabolite $i$ on the activity of enzyme $j$. 
+Conversely, if enzyme activity was inhibited by a $m$ metabolites, we modeling this inhibition as:
+\begin{equation}
+	v_{j} = 1 - \max\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right)
+\end{equation}Lastly, if an enzyme had both $m$ activating and $n$ inhibitory factors, we modeled the regulatory term as:
+\begin{equation}
+	v_{j} = \min\left(u_{j},d_{j}\right)
+\end{equation}where:
+\begin{eqnarray}
+	u_{j} &=& \max_{j^{+}}\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right) \\
+	d_{j} &=& 1 - \max_{j^{-}}\left(f_{1j}\left(x\right),\hdots,f_{nj}\left(x\right)\right)
+\end{eqnarray}The quantities $j^{+}$ and $j^{-}$ denoted the sets of activating, and inhibitory factors for enzyme $j$. 
+If an enzyme had no allosteric factors, we set $v_{j} = 1$.
+There are many possible functional forms for $0\leq f_{ij}\left(x\right)\leq 1$. 
+However, in this study, each individual transfer function took the form:
+\begin{equation}\label{eqn:control-factor}
+	f_{i}\left(\mathbf{x}\right) = \frac{\kappa_{ij}^{\eta}x_{j}^{\eta}}{1 + \kappa_{ij}^{\eta}x_{j}^{\eta}}
+\end{equation}where $x_{j}$ denotes the abundance of metabolite $j$, and $\kappa_{ij}$ and $\eta$ are control parameters. 
+The $\kappa_{ij}$ parameter was species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).
+The model equations were encoded using the Octave programming language, and solved using the LSODE routine in Octave \citep{Octave:2014}.
+
+\subsection*{Estimation of model parameters from experimental data.}
 
@@
 \clearpage
+%\bibliographystyle{plain}
+%\bibliographystyle{IEEEbib}
 
 \bibliography{References_v1}
-
-%\begin{comment}
 

### 2014-07-25 Paper_v1.tex
Updated text

@@
 \begin{titlepage}
-{\par\centering\textbf{\Large An Effective Model of the Retinoic Acid induced HL-60 Differentiation Circuit}}
+{\par\centering\textbf{\Large An Effective Model of the Core Retinoic Acid induced HL-60 Differentiation Circuit}}
 \vspace{0.05in}
@@
 \section*{Abstract}
+The complexity of human differentiation networks makes analysis of their architecture and function challenging.  
+Analysis of model systems, such as the lineage-uncommitted human myeloblastic leukemia cell line HL-60, 
+could inform the analysis of more complex programs. 
+HL-60 cells undergo myeloid differentiation along with G1/G0-arrest when exposed to All-Trans Retinoic Acid (ATRA). 
+One of the defining features of this program is slowly induced, persistent MAPK activation.
+The molecular mechanisms of ATRA-induced commitment, arrest and functional differentiation are only partially understood.
+In this study, we explored the ATRA-inducible c-Raf, also known as Raf1 (Raf), 
+interactome to determine the functional and regulatory architecture responsible for persistent MAPK activation in HL-60 cells. 
+To better understand the role of Raf in the ATRA-induced differentiation program, 
+we constructed an ensemble of mechanistic, mathematical models of the transcriptional and post-translational events driving ATRA-induced MAPK activation. 
+Bifurcation analysis of this model predicted bistability in ppERK levels as a function of ATRA forcing. 
+A functional consequence of this was the ability to lock the MAPK cascade into a self-sustaining activated state, even after ATRA removal.
+These simulations were then qualitatively validated with ATRA washout experiments. 
+Taken together, this study provided further details on sustained MAPK activation, mechanistic insight into cellular memory,
+and proof-of-concept that a
+combination of experimental and computational methods is an effective strategy for dissecting complex intracellular signaling programs.
 
@@
 \section*{Discussion}
+In this study, we explored the functional and regulatory architecture responsible for ATRA-induced MAPK activation in HL-60 cells.
+Retinoic acid is an embryonic morphogen, a cancer chemotherapeutic agent, as well as potent regulator of cell differentiation and proliferation. 
+Using a human lineage-uncommitted myelo-monocytic precursor cell, 
+it has been shown that ATRA-induced phenotypic conversion and G0 arrest depend on eliciting a hyperactive prolonged signal along the Raf/MEK/ERK axis \cite{Yen1998}. 
+In particular, signaling involves two positive loops, which incorporates the BLR1 receptor and Raf kinase, putatively activated by S621 phosphorylation, 
+that is seminal to sustaining the signal needed for differentiation \cite{Wang2008}. 
+In genetically engineered BLR1 KO cells crippled for signaling and consequentially ability to differentiate, ectopic expression of Raf activity restores the cell's ability to differentiate. 
+While the canonical paradigm of MAPK signaling is a transient signal that is mitogenic, a prolonged signal elicits cell differentiation and G0 arrest. 
+Interest thus becomes focused on whether there is a Raf partner that is responsible for the prolonged signal in ATRA-treated cells. 
+Surveying known MAPK regulatory molecules using Western blotting and IP we found that ATRA increased the expression of Vav1 as well as its immunoprecipitation with Raf phosphorylated at S621. 
+A parsimonious effective mathematical model showed that a Raf-Siganlsome-mediated positive feedback loop accounted for sustained MAPK signaling and aspects of phenotypic conversion and G0 arrest. 
+The significant finding was thus that ATRA-regulated Vav1 expression and partnering with Raf could drive sustained signaling and phenotypic shift.
+The model also predicted the observed hysteresis effect of ATRA where the cell retains a memory of transient ATRA exposure, betrayed by activated signaling and by accelerated differentiation upon re-exposure to ATRA.
 
@@
 \subsection*{Formulation and solution of the model equations.}
-We used ordinary differential equations (ODEs) to model the time evolution of metabolite ($x_{i}$) and scaled enzyme abundance ($\epsilon_{i}$) in hypothetical cell free
-metabolic networks:
-\begin{eqnarray}
-	\frac{dx_{i}}{dt} & = & \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)\qquad{i=1,2,\hdots,\mathcal{M}}\\
-	\frac{d\epsilon_{i}}{dt} & = & -\lambda_{i}\epsilon_{i}\qquad{i = 1,2,\hdots,\mathcal{E}}
-\end{eqnarray}where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of metabolites and $\mathcal{E}$ denotes the number of
-enzymes in the model. The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ 
-denotes the rate of reaction $j$. Typically, reaction $j$ is a non-linear function of metabolite and enzyme abundance, as well as unknown kinetic parameters 
-$\mathbf{k}$ ($\mathcal{K}\times{1}$).
+We used ordinary differential equations (ODEs) to model the time evolution of the circuit components:
+\begin{equation}
+	\frac{dx_{i}}{dt}  = \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right) - \left(\mu + k_{d,i}\right)x_{i}\qquad{i=1,2,\hdots,\mathcal{M}}\\
+\end{equation}where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of circuit components, $k_{d,i}$ denotes a first-order non-specific degradation constant
+for species i, and $\mu$ denotes the specific growth rate of HL-60 cells.
 The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$. If 
 $\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$. Conversely, if $\sigma_{ij}>0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates
-metabolite $i$ is not connected with reaction $j$. Lastly, $\lambda_{i}$ denotes the scaled enzyme degradation constant. The system material balances were subject to the
-initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$ and $\mathbf{\epsilon}\left(t_{o}\right) = \mathbf{1}$ (initially we have 100\% cell-free enzyme abundance).
+metabolite $i$ is not connected with reaction $j$. 
+The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of reaction $j$. 
+Typically, reaction $j$ is a non-linear function of network state, as well as unknown kinetic parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$).
+The system material balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$.
 
@@
 
-The allosteric regulation term $v_{j}$ depended upon the combination of factors which influenced the activity of enzyme $i$.
-For each enzyme, we used a rule based approach to select from competing control factors (Fig. \ref{fig-control-schematic}). 
-If an enzyme was activated by $m$ metabolites, we modeled this activation as:
+The control term $v_{j}$ depended upon the combination of factors which influenced the abundance or activity of $x_{i}$.
+For each species, we used a rule based approach to select from competing control factors (Fig. \ref{fig-control-schematic}). 
+We modeled species activation as:
 \begin{equation}
 	v_{j} = \max\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right)
-\end{equation}where $0\leq f_{ij}\left(x\right)\leq 1$ was a regulatory transfer function that calculated the influence of metabolite $i$ on the activity of enzyme $j$. 
-Conversely, if enzyme activity was inhibited by a $m$ metabolites, we modeling this inhibition as:
+\end{equation}where $0\leq f_{ij}\left(x\right)\leq 1$ was a regulatory transfer function that calculated the influence of species $x_{i}$ on rate process $j$. 
+Conversely, we modeled the inhibition (or repression) of rate process $j$ as:
 \begin{equation}
 	v_{j} = 1 - \max\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right)
-\end{equation}Lastly, if an enzyme had both $m$ activating and $n$ inhibitory factors, we modeled the regulatory term as:
+\end{equation}Lastly, if rate process $j$ was subject to had both $m$ activating and $n$ inhibitory factors, we modeled the control term as:
 \begin{equation}
@@
 	d_{j} &=& 1 - \max_{j^{-}}\left(f_{1j}\left(x\right),\hdots,f_{nj}\left(x\right)\right)
-\end{eqnarray}The quantities $j^{+}$ and $j^{-}$ denoted the sets of activating, and inhibitory factors for enzyme $j$. 
-If an enzyme had no allosteric factors, we set $v_{j} = 1$.
+\end{eqnarray}The quantities $j^{+}$ and $j^{-}$ denoted the sets of activating, and inhibitory factors for process $j$. 
+If a process had no control factors, we set $v_{j} = 1$.
 There are many possible functional forms for $0\leq f_{ij}\left(x\right)\leq 1$. 
@@
 	f_{i}\left(\mathbf{x}\right) = \frac{\kappa_{ij}^{\eta}x_{j}^{\eta}}{1 + \kappa_{ij}^{\eta}x_{j}^{\eta}}
-\end{equation}where $x_{j}$ denotes the abundance of metabolite $j$, and $\kappa_{ij}$ and $\eta$ are control parameters. 
+\end{equation}where $x_{j}$ denotes the abundance (or activity) of component $j$.  
 The $\kappa_{ij}$ parameter was species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).
@@
 
-\subsection*{Estimation of model parameters from experimental data.}
+\subsection*{Estimation of an ensemble of model parameters from experimental data.}
+Model parameters were estimated by minimizing the squared difference between simulations and experimental data taken from ATRA-induced HL-60 cells:
+\begin{equation}\label{eqn:objective-function}
+	E_{j}(\mathbf{k}) = \sum_{i=1}^{\mathcal{T}_{j}}\biggl(\hat{\mathcal{M}}_{ij}-\hat{y}_{ij}(\mathbf{k})\biggr)^2 + \left(\frac{{\mathcal{M}^{\prime}_{ij}}-\max{y_{ij}}}{{\mathcal{M}^{\prime}_{ij}}}\right)^{2}
+\end{equation}
+The terms $\hat{\mathcal{M}}_{ij}$ and $\hat{y}_{ij}$ denote scaled experimental observations and simulation outputs from training set $j$. 
+The quantity $i$ denoted the sampled time-index and $\mathcal{T}_{j}$ denoted the number of time points for experiment j. 
+The first term in Eqn. \eqref{eqn:objective-function} quantified the relative error in the simulation. 
+We used only immunoblot measurements for model training. 
+Thus, we trained the model on the \emph{relative} change between bands within each training data set.
+The read-out from the training immunoblots was band intensity where we assumed intensity was only loosely proportional to concentration.
+Suppose we have the intensity for species $x$ at time $\{t_{1},t_{2},..,t_{n}\}$ in condition $j$. The scaled-value $\hat{\mathcal{M}}_{ij}$ would then be given by:
+\begin{equation}\label{norm_exp_data}
+\hat{\mathcal{M}}_{ij} = \frac{\mathcal{M}_{ij} - \min_{i}\mathcal{M}_{ij}}{\max_{i}{\mathcal{M}_{ij}}-\min_{i}{\mathcal{M}_{ij}}}
+\end{equation}
+Under this scaling $0\leq\hat{\mathcal{M}}_{ij}\leq{1}$ where $\hat{\mathcal{M}}_{ij}=0$ describes the lowest intensity band and $\hat{\mathcal{M}}_{ij}=1$ describes the highest intensity band. 
+A similar scaling was defined for the simulation output. The second-term in the objective function ensured the proper concentration scale was estimated by the model.
+In this study, we set the highest intensity band to $\mathcal{M}^{\prime}_{ij} = 10$ [AU] for all simulations.
+
+We minimized the total model residual ($\sum{E_{j}}$) using Particle swarm optimization (PSO) \citep{PSO}.
+PSO uses a \textit{swarming} metaheuristic to explore parameter spaces. 
+A strength of PSO is its ability to find the global minimum, even in the presence of potentially many local minima, by communicating the local
+error landscape experienced by each particle collectively to the swarm. Thus, PSO acts both as a local and a global search algorithm. 
+For each iteration, particles in the swarm compute their local error by evaluating the model equations using their specific parameter vector realization.
+From each of these local points, a globally best error is identified. Both the local and global error 
+are then used to update the parameter estimates of each particle using the rules:
+\begin{eqnarray}
+	\mathbf{\Delta}_{i} &=&\theta_{1}\mathbf{\Delta}_{i} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{k}_{i}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G} - \mathbf{k}_{i}\right) \\
+	\mathbf{k}_{i} &=& \mathbf{k}_{i} + \mathbf{\Delta}_{i}
+\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes local best solution found by particle $i$, and
+$\mathcal{G}$ denotes the best solution found over the entire population of particles.
+The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
+parameters ($\mathcal{K}\times{1}$). In thus study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$, which was taken from XXX.
 
 \section*{Acknowledgements}
-This study was supported by the National Science Foundation GK12 award (DGE-1045513) 
-and by the National Science Foundation CAREER award (FILLMEIN).
+The authors thank the anonymous reviewers for their helpful suggestions. 
+We acknowledge the gracious financial support to J.V. by the National Science Foundation CAREER (CBET-0846876) for the support of R.T. and H.J.
+In addition, we acknowledge support to A.Y. from the National Institutes of Health (CA 30555, CA152870) 
+and a grant from New York State Stem Cell Science.  Lastly, we acknowledge the financial support to J.V. and A.Y. 
+from the National Cancer Institute (\#U54CA143876). The content is solely the responsibility of the authors and does not necessarily 
+represent the official views of the National Cancer Institute or the National Institutes of Health.
 

### 2014-07-25 Paper_v1.tex
Updated text and figures

@@
 
+\subsection*{Inhibition of Raf kinase activity modulated MAPK activation and differentiation markers.}
+Next, we considered the effect of the Raf kinase inhibitor GW5074 on functional markers of ATRA-induced growth arrest and differentiation.
+For more details on GW5074 including dose response see Supplemental Materials including Fig. \ref{F:MAPKGWWB}A.
+ATRA treatment alone statistically significantly increased the G1/G0 percentage over the untreated control, 
+while GW5074 alone had a negligible effect on the cell cycle distribution (Fig. \ref{F:GWdiff}A).
+Secondly, ATRA treatment alone increased both Raf phosphorylation at S621/S259 and Raf expression 24 hr post-treatment,
+while ATRA-induced phosphorylation of ERK was blocked by GW5074, consistent with the loss of Raf kinase activity (Fig. \ref{F:MAPKGWWB}B).
+Lastly, both the phosphorylation and expression of Raf was inhibited by GW5074.
+This suggested a positive feedback between Raf kinase activity, Raf phosphorylation and ERK action on Raf transcription factors.
+Surprisingly, the combination of GW5074 and ATRA statistically significantly increased the G1/G0 population (82 $\pm$ 1\%)
+compared with ATRA alone (61 $\pm$ 0.5\%).
+Increased G1/G0 arrest following the combined treatment with GW5074 and ATRA was unexpected, 
+as the combination of ATRA and the MEK inhibitor (PD98059) has been shown previously to decrease ATRA-induced growth arrest \cite{Yen1998}.
+However, growth arrest is not the sole indication of functional differentiation.
+Expression of the cell surface marker CD11b has also been shown to coincide with HL-60 cells myeloid differentiation \cite{Hickstein1989}.  
+We measured CD11b expression, for the various treatment groups, using immuno-fluorescence flow cytometry 48 hr post-treatment.
+As with G1/G0 arrest, ATRA alone increased CD11b expression over the untreated control, 
+while GW5074 further enhanced ATRA-induced CD11b expression (Fig. \ref{F:GWdiff}B).
+GW5074 alone had no statistically significant effect on CD11b expression, 
+compared with the untreated control.
+
+The inducible reactive oxygen species (ROS) response was used as a functional marker of differentiated neutrophils \cite{Congleton2011}.
+We measured the ROS response induced by the phorbol ester 12-O-tetradecanoylphorbol-13-acetate (TPA) using flow cytometry.
+Untreated cells showed no discernible TPA response, 
+with only 7.0 $\pm$ 3.0\% ROS induction (Fig. \ref{F:GWdiff}C).
+Cells treated with ATRA had a significantly increased TPA response, 53 $\pm$ 7\% ROS induction 48 hr post-treatment.
+Treatment with both ATRA and GW5074 statistically significantly reduced ROS induction (22 $\pm$ 0.6\%) compared to ATRA alone. 
+Interestingly, Western blot analysis did not detect a GW5074 effect on ATRA-induced expression of p47phox, 
+a required upstream component of the ROS response  (Fig. \ref{F:GWdiff}C, bottom).
+Thus, the inhibitory effect of GW5074 on inducible ROS might occur downstream of p47phox expression.
+However, the ROS producing complex is MAPK dependent, 
+therefore it is also possible that GW5074 inhibited ROS production by interfering 
+with MAPK activation (in which case the p47phox marker might not accurately reflect phenotypic conversion and differentiation).
+
 \clearpage

### 2014-07-25 Paper_v1.tex
Updated text and figures

@@
 
+\subsection*{Mathematical modeling of ATRA-induced MAPK activation in HL-60 cells.}
+We constructed a mechanistic mathematical model of a putative Raf-Vav1 activation circuit (Fig. \ref{F:network}).  
+The immunoprecipitation and kinase inhibitor studies suggested that Raf kinase activity was required for ATRA-induced ERK phosphorylation and ROS response. 
+Furthermore, the phosphorylation, expression, and Vav1 binding of Raf itself decreased in response to Raf inhibition. 
+These results suggested a possible positive feedback loop between Raf kinase activity and increased Raf expression, 
+possibly through ppERK-dependent transcription factors such as Sp1 and STAT1 \cite{Kim2009,Zhang2004}.
+Both Sp1 and STAT1 have putative binding domains in the Raf promoter (TRANSFAC).
+However, this loop is not ATRA-responsive, without an ATRA-inducible trigger, which we hypothesize is Vav1.
+The Vav1 gene contains an upstream canonical Retinoic Acid Response Element (RARE) sequence motif (TRANSFAC) which could allow
+Vav1 expression to be at least partially regulated by ATRA. 
+By contrast, the Src promoter does not have a discernible canonical putative RARE.
+(Tables \ref{T:params}, \ref{T:species} and \ref{T:rxns} respectively).
+The levels of ATRA and GW5074 were introduced as user defined parameters depending on the experiment being simulated.
+In general Dilution due to growth was 
+More details on the model architecture including additional references can be found in Materials and Methods.
+
+To address parametric uncertainty, 
+an ensemble of models was estimated using the eight datasets taken from this study (38 total data points),
+and the previous study of Wang and Yen \cite{Wang2008}.
+Here ensemble refers to the use of different parametric sets for the same system of reactions.
+Cross-validation (data randomly split into 80\% training and 20\% validation) demonstrated that, 
+on average, the model identification scheme (described in Materials and Methods) decreased both training 
+and prediction error (Fig. \ref{F:parameters}) over successive iterations of the parameter selection procedure. 
+This is suggestive of the predictive ability of the model and also demonstrated that the nominal parameter set, 
+the set with the lowest training error, did not result in over-fitting. 
+Subsequently, we considered an ensemble of likely parameter sets (N = 99) with relative mean squared training errors within 20\% of the nominal set. 
+The model ensemble recapitulated both time-series data (including sustained MAPK activation) and the GW5074 results (Fig. \ref{F:Train}).
+
 \subsection*{Inhibition of Raf kinase activity modulated MAPK activation and differentiation markers.}
-Next, we considered the effect of the Raf kinase inhibitor GW5074 on functional markers of ATRA-induced growth arrest and differentiation.
+We considered the effect of the Raf kinase inhibitor GW5074 on functional markers of ATRA-induced growth arrest and differentiation.
 For more details on GW5074 including dose response see Supplemental Materials including Fig. \ref{F:MAPKGWWB}A.
@@
 Lastly, both the phosphorylation and expression of Raf was inhibited by GW5074.
-This suggested a positive feedback between Raf kinase activity, Raf phosphorylation and ERK action on Raf transcription factors.
+This suggested a positive feedback between Raf kinase activity, Raf phosphorylation and possible ERK action on Raf transcription factors.
 Surprisingly, the combination of GW5074 and ATRA statistically significantly increased the G1/G0 population (82 $\pm$ 1\%)

### 2014-07-25 Paper_v1.tex
Figs 1,2 and 3

--- 2014-07-25 Paper_v1.tex
@@
 \begin{titlepage}
-{\par\centering\textbf{\Large An Effective Model of the Core Retinoic Acid induced HL-60 Differentiation Circuit}}
+{\par\centering\textbf{\Large Dynamically Dimensioned Particle Swarm Optimization}}
 \vspace{0.05in}
-{\par \centering \large{Ryan Tasseff$^{1}$, 
-Holly A. Jensen$^{1}$, 
-Johanna Congleton$^{2}$, Andrew Yen$^{2}$ and Jeffrey D. Varner$^{1,*}$}}
+{\par \centering \large{Adithya Sagar, Christine Shoemaker and Jeffrey D. Varner$^{*}$}}
 \vspace{0.10in}
-{\par \centering \large{$^{1}$~School of Chemical and Biomolecular Engineering}}
+{\par \centering \large{School of Chemical and Biomolecular Engineering}}
 {\par \centering \large{Cornell University, Ithaca NY 14853}}
 \vspace{0.1in}
-{\par \centering \large{$^{2}$~Department of Biomedical Sciences}}
-{\par \centering \large{Cornell University, Ithaca NY 14853}}
+{\par \centering \textbf{Running Title:}~Modeling cell free metabolism}
 \vspace{0.1in}
-{\par \centering \textbf{Running Title:}~Effective HL60 differentiation circuit}
-\vspace{0.1in}
-{\par \centering \textbf{To be submitted:}~\emph{Processes}}
+{\par \centering \textbf{To be submitted:}~\emph{BMC~Systems~Biology}}
 \vspace{0.5in}
@@
 \section*{Abstract}
-The complexity of human differentiation networks makes analysis of their architecture and function challenging.  
-Analysis of model systems, such as the lineage-uncommitted human myeloblastic leukemia cell line HL-60, 
-could inform the analysis of more complex programs. 
-HL-60 cells undergo myeloid differentiation along with G1/G0-arrest when exposed to All-Trans Retinoic Acid (ATRA). 
-One of the defining features of this program is slowly induced, persistent MAPK activation.
-The molecular mechanisms of ATRA-induced commitment, arrest and functional differentiation are only partially understood.
-In this study, we explored the ATRA-inducible c-Raf, also known as Raf1 (Raf), 
-interactome to determine the functional and regulatory architecture responsible for persistent MAPK activation in HL-60 cells. 
-To better understand the role of Raf in the ATRA-induced differentiation program, 
-we constructed an ensemble of mechanistic, mathematical models of the transcriptional and post-translational events driving ATRA-induced MAPK activation. 
-Bifurcation analysis of this model predicted bistability in ppERK levels as a function of ATRA forcing. 
-A functional consequence of this was the ability to lock the MAPK cascade into a self-sustaining activated state, even after ATRA removal.
-These simulations were then qualitatively validated with ATRA washout experiments. 
-Taken together, this study provided further details on sustained MAPK activation, mechanistic insight into cellular memory,
-and proof-of-concept that a
-combination of experimental and computational methods is an effective strategy for dissecting complex intracellular signaling programs.
 
-{\noindent \textbf{Keywords:}~Cell free metabolism, Mathematical modeling}
+{\noindent \textbf{Keywords:}~Parameter identification, Mathematical modeling}
 
@@
 
-Understanding differentiation, 
-the process by which precursor cells become more specialized cell types, 
-is an important challenge facing biology. 
-If differentiation programs could be rationally manipulated, 
-advanced therapies could be developed to treat a spectrum of cancers, spinal cord injuries and neurodegenerative disorders. 
-However, to rationally reprogram differentiation networks, 
-we must first understand their connectivity and regulation \cite{Young2011}. 
-Lessons learned in model systems, such as the lineage-uncommitted human myeloblastic cell line HL-60,
-could inform our analysis of more complex programs. 
-HL-60 has been a durable experimental model since the 1970's to study differentiation \cite{Breitman1980}.
-Depending upon the stimulus, HL-60 cells undergo G1/G0-arrest and myeloid or monocytic differentiation. 
-All-Trans Retinoic Acid (ATRA) induces G1/G0-arrest and myeloid differentiation in HL-60 cells, 
-whereas 1,25-dihydroxy vitamin D3 induces arrest with monocytic differentiation.
-Commitment to cell cycle arrest and terminal differentiation requires approximately 48 hr of treatment,
-during which HL-60 cells undergo approximately two division cycles.  
-Interestingly, cells treated with ATRA for time periods shorter than the commitment phase retain a limited inheritable memory, 
-which reduces the time required to reach commitment during subsequent ATRA exposure \cite{YEN1984}.
-
-Sustained Mitogen-Activated Protein Kinase (MAPK) cascade
-activation is a defining feature of ATRA-induced differentiation in HL-60 cells. 
-ATRA drives slow yet sustained MEK-dependent activation of the Raf/MEK/ERK pathway, 
-leading to arrest and functional differentiation \cite{Yen1998}.
-MEK inhibition results in the loss of both ERK and Raf phosphorylation, 
-as well as the failure to arrest and terminally differentiate \cite{Yen1998,Hong2001}.
-At the transcriptional level, 
-ATRA (and its metabolic products) are ligands for the hormone activated nuclear transcription factors Retinoic Acid Receptor (RAR) and Retinoid X Receptor (RXR) \cite{Mangelsdorf1990}.
-Activation of both RAR and RXR is necessary for ATRA-induced Raf phosphorylation and MAPK activation,
-suggesting that the initiation of MAPK signaling is partially transcriptionally regulated \cite{Hong2001}.
-ATRA, through activation of a transcription factor complex including RAR and RXR, induces the expression of many proteins.
-BLR1 also known as CXCR5, is a putative heterotrimeric Gq protein-coupled receptor 
-that is necessary for MAPK activation, growth arrest and functional differentiation \cite{YEN1990,LIPP1994,WANG2004}.
-BLR1 was identified as an early ATRA (or D3)-inducible gene in HL-60 cells using differential display \cite{YEN1990}. 
-Studies of the BLR1 promoter identified a 5' 17bp GT box approximately 1 kb upstream of the transcriptional start that conferred ATRA responsiveness \cite{WANG2004}.
-Additionally, members of the BLR1 transcriptional activator complex, e.g. NFATc3 and CREB, 
-can be phosphorylated by ERK, JNK or p38 MAPK family members \cite{Yang2002}.
-This suggests positive feedback between BLR1 expression and MAPK activation. 
-BLR1 overexpression enhanced Raf phosphorylation and accelerated terminal differentiation, 
-while BLR1 knock-out HL-60 cells failed to activate Raf or differentiate in the presence of ATRA \cite{Wang2008}.
-Interestingly, both the knockdown or inhibition of Raf, also reduced BLR1 expression and functional differentiation \cite{Wang2008}.
-A recent computational study of ATRA-induced differentiation in HL-60 cells suggested that the BLR1-MAPK positive feedback circuit was sufficient to explain ATRA-induced sustained MAPK activation and the expression of differentiation markers \cite{Tasseff2011}.
-Model analysis also suggested that Raf was the most distinct of the MAPK proteins.
-
-A critical question is what other components of the MAPK positive feedback circuit are required to drive ATRA-induced functional differentiation of HL-60 cells. 
-Wang and Yen showed that ectopic expression of the constitutively active CR3 domain of Raf1 restored ATRA-induced G0 arrest and differentiation in BLR1 knock-out cells \cite{Wang2008}. 
-However, ectopic expression of Raf1 CR3 alone, in the absence of ATRA, failed to induce arrest or differentiation.
-Thus, additional ATRA-inducible components must exist, 
-which independently promote arrest and differentiation in the absence of BLR1.
-In this study, we explored this hypothesis using a combination of experimental and computational tools. 
-First, we explored the ATRA-inducible Raf interactome 
-by surveying a panel of 19 possible binding partners using immunoprecipitation (IP), 
-with and without ATRA and the Raf inhibitor GW5074.  
-Initially, we expected increased ATRA-dependent association between Raf and kinases linked to BLR1 activity; 
-however, this was not supported by data.
-Instead, we found that the interaction between the guanine nucleotide exchange factor Vav1 and Raf was both ATRA-inducible and simultaneously sensitive to Raf inhibition. 
-Next, we considered how MAPK activation and differentiation were affected by the inhibition of Raf kinase activity in the presence and absence of ATRA.
-We showed that Raf activity was directly proportional to ERK phosphorylation and to functional differentiation processes such as the generation of reactive oxygen species (ROS). 
-Moreover, interactions between Raf and kinase partners such as Akt or CK2, or the scaffolding protein 14-3-3 were largely insensitive to ATRA treatment. 
-These studies established the working hypothesis that Vav1 (or potentially other ATRA-inducible proteins) acted as limiting members of a constitutively assembled trigger complex that propelled sustained MAPK activation, arrest and differentiation.
-We tested this hypothesis by constructing a mechanistic mathematical model of the Raf-Vav1 circuit, 
-based on the IP and Western blot data presented in this study, 
-and from previous literature. 
-The proposed model architecture was consistent with the ATRA-induced sustained MAPK activation observed experimentally. 
-Additionally, we found the Raf-Vav1 circuit possessed interesting dynamic features such as bistability, 
-that could explain ATRA-induction memory effects.
-
 \clearpage
@@
 
+Coagulation is an archetype biochemical network that is highly interconnected and tightly regulated with multiple positive and negative feedback loops. 
+The biochemistry underlying coagulation has been well studied [REFHERE], and reliable experimental coagulation models have been developed [REFHERE].
+
+This makes it an ideal system for mathematical modeling and parameter estimation.
+Coagulation is regulated by a set of serine proteases also known as coagulation factors and blood platelets.
+The coagulation factors are generally in an inactive state and are known as zymogens.
+These zymogens are activated when through certain triggers.
+These trigger events like injury or trauma or sepsis exposure factors like collagen, tissue factor and von Willebrand factor (vWF) to blood.
+The exposure of these factors to blood kick-starts a series of convergent cascades that lead to conversion of zymogen prothrombinase to thrombin.
+Luan et al.
+modeled coagulation using coupled non-linear ordinary differential equations with 301 reactions and 192 species.
+
 \clearpage
@@
 \section*{Discussion}
-In this study, we explored the functional and regulatory architecture responsible for ATRA-induced MAPK activation in HL-60 cells.
-Retinoic acid is an embryonic morphogen, a cancer chemotherapeutic agent, as well as potent regulator of cell differentiation and proliferation. 
-Using a human lineage-uncommitted myelo-monocytic precursor cell, 
-it has been shown that ATRA-induced phenotypic conversion and G0 arrest depend on eliciting a hyperactive prolonged signal along the Raf/MEK/ERK axis \cite{Yen1998}. 
-In particular, signaling involves two positive loops, which incorporates the BLR1 receptor and Raf kinase, putatively activated by S621 phosphorylation, 
-that is seminal to sustaining the signal needed for differentiation \cite{Wang2008}. 
-In genetically engineered BLR1 KO cells crippled for signaling and consequentially ability to differentiate, ectopic expression of Raf activity restores the cell's ability to differentiate. 
-While the canonical paradigm of MAPK signaling is a transient signal that is mitogenic, a prolonged signal elicits cell differentiation and G0 arrest. 
-Interest thus becomes focused on whether there is a Raf partner that is responsible for the prolonged signal in ATRA-treated cells. 
-Surveying known MAPK regulatory molecules using Western blotting and IP we found that ATRA increased the expression of Vav1 as well as its immunoprecipitation with Raf phosphorylated at S621. 
-A parsimonious effective mathematical model showed that a Raf-Siganlsome-mediated positive feedback loop accounted for sustained MAPK signaling and aspects of phenotypic conversion and G0 arrest. 
-The significant finding was thus that ATRA-regulated Vav1 expression and partnering with Raf could drive sustained signaling and phenotypic shift.
-The model also predicted the observed hysteresis effect of ATRA where the cell retains a memory of transient ATRA exposure, betrayed by activated signaling and by accelerated differentiation upon re-exposure to ATRA.
 
@@
 
-\subsection*{Cell Culture and Treatment.}
-Human myeloblastic leukemia cells (HL-60 cells) were grown in a humidified atmosphere of 5\% CO$_2$ at 37$^{o}$C and maintained in RPMI 1640 from Gibco (Carlsbad, CA) 
-supplemented with 5\% fetal bovine serum from Hyclone (Logan, UT) and 1x antibiotic/antimicotic (Sigma, St. Louis, MO).
-Cells were cultured in constant exponential growth as described previously \cite{Brooks1996}. 
-Experimental cultures were initiated at $0.1\times10^6$ cells/mL 24 hr prior to 1$\mu$M ATRA treatment;
-if indicated, cells were also treated with GW5074 (2$\mu$M) 18 hr before ATRA treatment.   
-For cell culture washout experiments HL-60 cells were treated with ATRA for 24 hr, 
-washed 3x with prewarmed serum supplemented culture medium to remove ATRA exposure, 
-and reseeded in ATRA-free media as described. 
-Western blot analysis was performed at incremental time points after removal of ATRA.
-
-\subsection*{Chemicals.}
-All-Trans Retinoic Acid from Sigma-Aldrich (St. Louis, MO) was dissolved in 100\% ethanol with a stock concentration of 5mM, 
-and used at a final concentration of 1$\mu$M.
-The Raf inhibitor GW5074 from Sigma-Aldrich (St. Louis, MO) was dissolved in DMSO with a stock concentration of 10mM, 
-and used at a final concentration of 2$\mu$M.
-HL-60 cells were treated with 2$\mu$M GW5074 with or without ATRA (1$\mu$M) at 0 hr.  
-This GW5074 dosage had a negligible effect on the cell cycle distribution, compared to ATRA treatment alone (Fig. \ref{F:MAPKGWWB}A).
-
-\subsection*{CD11b expression studies by flow cytometry.}
-Approximately $1.0\times10^6$ HL-60 cells were harvested  by centrifugation. 
-Cells were resuspended in 200$\mu$L PBS containing 5$\mu$L of allophycocyanin (APC)–conjugated anti-CD11b antibody from BD Biosciences (San Jose, CA). 
-Following incubation for 1hr at 37$^{o}$C, cells were analyzed by flow cytometry (LSRII flow cytometer, BD Biosciences; San Jose, CA) using 633nm red laser excitation. 
-The threshold on experimental groups was set to exclude 95\% of control, the untreated sample.
-
-\subsection*{Measurement of inducible oxidative metabolism.}
-Approximately $0.5\times10^6$ Cells were harvested by centrifugation.
-Cells were resuspended in 200$\mu$L 37$^{o}$C PBS containing 10$\mu$mol/L 
-5-(and-6)-chloromethyl-2,7-dichlorodihydro-fluorescein diacetate acetyl ester (DCF; Invitrogen Carlsbad, CA) 
-and 0.4$\mu$g/mL 12-O-tetradecanoylphorbol-13-acetate (TPA; Sigma-Aldrich St. Louis, MO)
-with incubation for 20 min in a humidified atmosphere of 5\% CO$_2$ at 37$^{o}$C. 
-Flow cytometric analysis was done as described previously \cite{Reiterer2007}.
-To determine TPA inducible ROS, the threshold on experimental groups was set to exclude 95\% of control,
-samples not treated with TPA.
-
-\subsection*{Cell cycle analysis.}
-Approximately $1.0\times10^6$ cells were collected by centrifugation.
-Cells were resuspended in 500$\mu$L hypotonic staining solution containing 50$\mu$g/mL propidium iodine, 
-1$\mu$L/mL Triton X-100, and 1 mg/mL sodium citrate. 
-Cells were incubated at room temperature for 1 hr and analyzed by flow cytometry (BD LSRII) using 488-nm excitation.
-
-\subsection*{Immunoprecipitation and Western blot.}
-Approximately $1.2 \times 10^7$ cells were lysed using 400$\mu$L of M-Per lysis buffer from Thermo Scientific (Waltham, MA). 
-Lysates were cleared by centrifugation at 16,950 $\times$ g in a micro-centrifuge for 20 min at 4$^{o}$C.  
-Lysates were pre-cleared using 100$\mu$L protein A/G Plus agarose beads from Santa Cruz Biotechnology (Santa Cruz, CA) by 
-inverting overnight at 4$^{o}$C.  Beads were cleared by centrifugation and total protein concentration
-was determined by a BCA assay (Thermo Scientific, Waltham, MA).  Immunoprecipitations were setup by bringing 
-lysate to a concentration of 1.0g/L in a total volume of 300$\mu$L (M-Per buffer was used for dilution).     
-The anti-Raf antibody was added at 3 $\mu$L.
-A negative control with no bait protein was also used to exclude the direct interaction of proteins with the A/G beads.   
-After 1 hr of inversion at 4$^{o}$C, 20$\mu$L of agarose beads was added and samples were left 
-to invert overnight at 4$^{o}$C.  
-Samples were then washed three times with M-Per buffer by centrifugation.
-Finally proteins were eluted from agarose beads using a laemmli loading buffer.
-Eluted proteins were resolved by SDS-PAGE and Western blotting.  Total 
-lysate samples were normalized by total protein concentration (20$\mu$g per sample)
-and resolved by SDS-PAGE and Western blotting.  Secondary HRP bound antibody was used for visualization.
-All antibodies were purchased from Cell Signaling (Boston, MA) with the exception of anti-p621 Raf which was purchased from Biosource/Invitrogen  (Carlsbad, CA), and
-anti-pS338 Raf which was purchased from Santa Cruz Biotechnology (Santa Cruz, CA); anti-retinoblastoma from Zymed (South San Francisco, CA); and anti-CK2 from BD Biosciences (San Jose, CA).
-
-\subsection*{Formulation and solution of the model equations.}
-We used ordinary differential equations (ODEs) to model the time evolution of the circuit components:
-\begin{equation}
-	\frac{dx_{i}}{dt}  = \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right) - \left(\mu + k_{d,i}\right)x_{i}\qquad{i=1,2,\hdots,\mathcal{M}}\\
-\end{equation}where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of circuit components, $k_{d,i}$ denotes a first-order non-specific degradation constant
-for species i, and $\mu$ denotes the specific growth rate of HL-60 cells.
-The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$. If 
-$\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$. Conversely, if $\sigma_{ij}>0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates
-metabolite $i$ is not connected with reaction $j$. 
-The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of reaction $j$. 
-Typically, reaction $j$ is a non-linear function of network state, as well as unknown kinetic parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$).
-The system material balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$.
-
-Each reaction rate was written as the product of two terms, a kinetic term ($\bar{r}_{j}$) and a regulatory term ($v_{j}$):
-\begin{equation}\label{eqn:rate-factor}
-	r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right) = \bar{r}_{j}v_{j}
-\end{equation}We used multiple saturation kinetics to model the reaction term $\bar{r}_{j}$:
-\begin{equation}\label{eqn:rate-bar}
-	\bar{r}_{j} = k_{j}^{max}\epsilon_{i}\left(\prod_{s\in{m_{j}^{-}}}\frac{x_{s}}{K_{js} + x_{s}}\right)
-\end{equation}where $k_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the scaled enzyme activity which catalyzes reaction $j$, and
-$K_{js}$ denotes the saturation constant for species $s$ in reaction $j$. 
-The product in Eqn. \eqref{eqn:rate-bar} was carried out over the set of \textit{reactants} for reaction $j$ (denoted as $m_{j}^{-}$).
-
-The control term $v_{j}$ depended upon the combination of factors which influenced the abundance or activity of $x_{i}$.
-For each species, we used a rule based approach to select from competing control factors (Fig. \ref{fig-control-schematic}). 
-We modeled species activation as:
-\begin{equation}
-	v_{j} = \max\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right)
-\end{equation}where $0\leq f_{ij}\left(x\right)\leq 1$ was a regulatory transfer function that calculated the influence of species $x_{i}$ on rate process $j$. 
-Conversely, we modeled the inhibition (or repression) of rate process $j$ as:
-\begin{equation}
-	v_{j} = 1 - \max\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right)
-\end{equation}Lastly, if rate process $j$ was subject to had both $m$ activating and $n$ inhibitory factors, we modeled the control term as:
-\begin{equation}
-	v_{j} = \min\left(u_{j},d_{j}\right)
-\end{equation}where:
-\begin{eqnarray}
-	u_{j} &=& \max_{j^{+}}\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right) \\
-	d_{j} &=& 1 - \max_{j^{-}}\left(f_{1j}\left(x\right),\hdots,f_{nj}\left(x\right)\right)
-\end{eqnarray}The quantities $j^{+}$ and $j^{-}$ denoted the sets of activating, and inhibitory factors for process $j$. 
-If a process had no control factors, we set $v_{j} = 1$.
-There are many possible functional forms for $0\leq f_{ij}\left(x\right)\leq 1$. 
-However, in this study, each individual transfer function took the form:
-\begin{equation}\label{eqn:control-factor}
-	f_{i}\left(\mathbf{x}\right) = \frac{\kappa_{ij}^{\eta}x_{j}^{\eta}}{1 + \kappa_{ij}^{\eta}x_{j}^{\eta}}
-\end{equation}where $x_{j}$ denotes the abundance (or activity) of component $j$.  
-The $\kappa_{ij}$ parameter was species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).
-The model equations were encoded using the Octave programming language, and solved using the LSODE routine in Octave \citep{Octave:2014}.
-
-\subsection*{Estimation of an ensemble of model parameters from experimental data.}
-Model parameters were estimated by minimizing the squared difference between simulations and experimental data taken from ATRA-induced HL-60 cells:
-\begin{equation}\label{eqn:objective-function}
-	E_{j}(\mathbf{k}) = \sum_{i=1}^{\mathcal{T}_{j}}\biggl(\hat{\mathcal{M}}_{ij}-\hat{y}_{ij}(\mathbf{k})\biggr)^2 + \left(\frac{{\mathcal{M}^{\prime}_{ij}}-\max{y_{ij}}}{{\mathcal{M}^{\prime}_{ij}}}\right)^{2}
-\end{equation}
-The terms $\hat{\mathcal{M}}_{ij}$ and $\hat{y}_{ij}$ denote scaled experimental observations and simulation outputs from training set $j$. 
-The quantity $i$ denoted the sampled time-index and $\mathcal{T}_{j}$ denoted the number of time points for experiment j. 
-The first term in Eqn. \eqref{eqn:objective-function} quantified the relative error in the simulation. 
-We used only immunoblot measurements for model training. 
-Thus, we trained the model on the \emph{relative} change between bands within each training data set.
-The read-out from the training immunoblots was band intensity where we assumed intensity was only loosely proportional to concentration.
-Suppose we have the intensity for species $x$ at time $\{t_{1},t_{2},..,t_{n}\}$ in condition $j$. The scaled-value $\hat{\mathcal{M}}_{ij}$ would then be given by:
-\begin{equation}\label{norm_exp_data}
-\hat{\mathcal{M}}_{ij} = \frac{\mathcal{M}_{ij} - \min_{i}\mathcal{M}_{ij}}{\max_{i}{\mathcal{M}_{ij}}-\min_{i}{\mathcal{M}_{ij}}}
-\end{equation}
-Under this scaling $0\leq\hat{\mathcal{M}}_{ij}\leq{1}$ where $\hat{\mathcal{M}}_{ij}=0$ describes the lowest intensity band and $\hat{\mathcal{M}}_{ij}=1$ describes the highest intensity band. 
-A similar scaling was defined for the simulation output. The second-term in the objective function ensured the proper concentration scale was estimated by the model.
-In this study, we set the highest intensity band to $\mathcal{M}^{\prime}_{ij} = 10$ [AU] for all simulations.
-
-We minimized the total model residual ($\sum{E_{j}}$) using Particle swarm optimization (PSO) \citep{PSO}.
-PSO uses a \textit{swarming} metaheuristic to explore parameter spaces. 
-A strength of PSO is its ability to find the global minimum, even in the presence of potentially many local minima, by communicating the local
-error landscape experienced by each particle collectively to the swarm. Thus, PSO acts both as a local and a global search algorithm. 
-For each iteration, particles in the swarm compute their local error by evaluating the model equations using their specific parameter vector realization.
-From each of these local points, a globally best error is identified. Both the local and global error 
-are then used to update the parameter estimates of each particle using the rules:
-\begin{eqnarray}
-	\mathbf{\Delta}_{i} &=&\theta_{1}\mathbf{\Delta}_{i} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{k}_{i}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G} - \mathbf{k}_{i}\right) \\
-	\mathbf{k}_{i} &=& \mathbf{k}_{i} + \mathbf{\Delta}_{i}
-\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes local best solution found by particle $i$, and
-$\mathcal{G}$ denotes the best solution found over the entire population of particles.
-The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
-parameters ($\mathcal{K}\times{1}$). In thus study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$, which was taken from XXX.
+\subsection*{Formulation and solution of coagulation model equations.}
+\subsection*{Optimization methods.}
 
 \section*{Acknowledgements}
-The authors thank the anonymous reviewers for their helpful suggestions. 
-We acknowledge the gracious financial support to J.V. by the National Science Foundation CAREER (CBET-0846876) for the support of R.T. and H.J.
-In addition, we acknowledge support to A.Y. from the National Institutes of Health (CA 30555, CA152870) 
-and a grant from New York State Stem Cell Science.  Lastly, we acknowledge the financial support to J.V. and A.Y. 
-from the National Cancer Institute (\#U54CA143876). The content is solely the responsibility of the authors and does not necessarily 
-represent the official views of the National Cancer Institute or the National Institutes of Health.
+This study was supported by the National Science Foundation GK12 award (DGE-1045513) 
+and by the National Science Foundation CAREER award (FILLMEIN).
 
 \clearpage
-%\bibliographystyle{plain}
-%\bibliographystyle{IEEEbib}
 
 \bibliography{References_v1}
+
+%\begin{comment}
 

### 2014-08-29 Paper_v1.tex
New fig and Fig 8

@@
 \begin{titlepage}
-{\par\centering\textbf{\Large Dynamically Dimensioned Particle Swarm Optimization}}
+{\par\centering\textbf{\Large Dynamic Modeling of Biochemical Networks using Effective Kinetic Models}}
 \vspace{0.05in}
-{\par \centering \large{Adithya Sagar, Christine Shoemaker and Jeffrey D. Varner$^{*}$}}
+{\par \centering \large{Joseph A. Wayman, Adithya Sagar and Jeffrey D. Varner$^{*}$}}
 \vspace{0.10in}
@@
 \vspace{0.1in}
-{\par \centering \textbf{Running Title:}~Modeling cell free metabolism}
+{\par \centering \textbf{Running Title:}~Effective models of metabolism}
 \vspace{0.1in}
-{\par \centering \textbf{To be submitted:}~\emph{BMC~Systems~Biology}}
+{\par \centering \textbf{To be submitted:}~\emph{Processes}}
 \vspace{0.5in}
@@
 \section*{Introduction}
+Biochemical processes are widely used in biotechnology to produce an array of products including commodity chemicals and protein therapeutics.
+However, whole-cell processes share the central limitation of requiring cell growth, 
+which redirects resources away from product synthesis, and cell walls, which complicate interrogation and control of intracellular metabolic processes.
+On the other hand, cell-free metabolic systems offer many advantages over traditional in vivo production methods. 
+For example, cell-free systems can direct scarce metabolic resources exclusively towards a single product of interest. 
+Moreover, with no cell wall, cell free systems can more easily be interrogated, and substrates of the metabolite processes directly controlled.
+Cell free production offers the unique opportunity to study metabolism without the complication of cell growth and gene expression processes.
+For modeling, this implies that we need only consider allosteric regulation of enzyme activity when building and testing cell free metabolic models.
+Of course, modeling allosteric mechanisms is itself a difficult problem when the model is at a whole genome scale. To address this problem, we have developed a
+an approach based upon the constrained fuzzy logic framework of Morris and Lauffenburger [REFHERE].
+
+In this study, we present an effective biochemical network modeling framework that integrates classical kinetic modeling with rules based approaches.
+[FINISH ME]
 
@@
 
-Coagulation is an archetype biochemical network that is highly interconnected and tightly regulated with multiple positive and negative feedback loops. 
-The biochemistry underlying coagulation has been well studied [REFHERE], and reliable experimental coagulation models have been developed [REFHERE].
+\subsection*{Formulation and properties of cell free effective models.}
+We developed two proof-of-concept metabolic networks to investigate the features of our effective biochemical network modeling approach (Fig. \ref{fig-networks}).
+In both examples, substrate $S$ was converted to the end-products $P_{1}$ and $P_{2}$ through a series of enzymatically catalyzed reactions, 
+including a branch point at hypothetical metabolite $M_{2}$. 
+Several of these reactions involved cofactor dependence ($AH$ or $A$), and various allosteric regulation mechanisms. 
+Network A included feedback inhibition of the initial pathway enzyme ($E_{1}$) by pathway end products $P_{1}$ and $P_{2}$ (Fig. \ref{fig-networks}A).
+On the other hand, network B involved feedback inhibition of $E_{1}$ by $P_{2}$ and $E_{6}$ by $P_{1}$ (Fig. \ref{fig-networks}B). 
+In both networks, branch point enzymes $E_{3}$ and $E_{6}$ were subject to feed-forward activation by cofactor $AH$.
+Lastly, enzyme activity was assumed to decay according to a first-order rate law in both cases. 
+Allosteric regulation of enzyme activity was represented using a novel rule-based strategy, similar in spirit to the 
+Constrained Fuzzy Logic (cFL) approach of Lauffenberger and coworkers \citep{Morris:2011aa}. 
+In this formulation, Hill-like transfer functions were used to calculate the influence of metabolite abundance upon target enzyme activity. 
+When an enzyme was potentially sensitive to more than one regulatory influence, logical rules were used to select which transfer function regulated enzyme 
+activity at any given time (Fig. \ref{fig-control-schematic}).      
+Thus, our test networks involved important features such as cofactor recycling, enzyme activity and metabolite dynamics, 
+as well as multiple overlapping allosteric regulatory mechanisms.  
+As such, developing our effective modeling approach using these simple problems gave us valuable insight into the development of larger network models, 
+without the complication of network size.
 
-This makes it an ideal system for mathematical modeling and parameter estimation.
-Coagulation is regulated by a set of serine proteases also known as coagulation factors and blood platelets.
-The coagulation factors are generally in an inactive state and are known as zymogens.
-These zymogens are activated when through certain triggers.
-These trigger events like injury or trauma or sepsis exposure factors like collagen, tissue factor and von Willebrand factor (vWF) to blood.
-The exposure of these factors to blood kick-starts a series of convergent cascades that lead to conversion of zymogen prothrombinase to thrombin.
-Luan et al.
-modeled coagulation using coupled non-linear ordinary differential equations with 301 reactions and 192 species.
+The rule based regulatory strategy approximated the behavior of classical allosteric activation and inhibition mechanisms (Fig. \ref{fig-kinetics-simulations}).  
+We first explored feed-forward substrate activation of enzyme activity (for both positive and negative cooperativity). 
+Consistent with classical data, the rule based strategy predicted a sigmoidal relationship between substrate abundance 
+and reaction rate as a function of the cooperativity parameter (Fig. \ref{fig-kinetics-simulations}A). 
+For cooperativity parameters less than unity, increased substrate abundance \textit{decreased} the reaction rate. This was consistent with the idea that
+substrate binding \textit{decreases} at regulatory sites negatively impacts the ability of the enzyme to bind substrate at the active site. 
+On the other hand, as the cooperativity parameter increased past unity, the rate of conversion of substrate $S$ to product $P$ by enzyme $E$ approached a step function.
+In the presence of an inhibitor, the rule based strategy predicted non-competitive like behavior as a function of the 
+cooperativity parameter (Fig. \ref{fig-kinetics-simulations}B). 
+When the control gain parameter, $\kappa_{ij}$ in Eqn. \eqref{eqn:control-factor}, was greater than unity, the inhibitory force was
+directly proportional to the cooperativity parameter, $\eta$ in Eqn. \eqref{eqn:control-factor}. 
+Thus, as the cooperativity parameter increased, the maximum reaction rate decreased (Fig. \ref{fig-kinetics-simulations}B, orange).
+However, when the gain parameter was less than unity, enzyme inhibition increased with \textit{decreasing} cooperativity, 
+i.e., smaller $\eta$ yielded increased inhibition (Fig. \ref{fig-kinetics-simulations}B).
+Interestingly, our rule based approach was unable to directly simulate competitive inhibition of enzyme activity. 
+For competitive inhibitors, the kinetic component of our rate, $\bar{r}_{j}$ in \eqref{eqn:rate-factor}, could be modified to account 
+for the inhibition (data not shown). 
+Taken together, the rule based strategy captured classical regulatory patterns for both enzyme activation and inhibition. 
+Thus, we are able to model complex kinetic phenomena such as ultrasensitivity, despite an effective description of reaction kinetics.
+
+End product yield was controlled by feedback inhibition, while selectivity was controlled by branch point enzyme inhibition (Fig. \ref{fig-onoff-simulations}).
+A critical test of our modeling approach was to simulate networks with known behavior. If we cannot reproduce the expected behavior of simple networks, then our
+effect modeling strategy, and particularly the rule-based approximation of allosteric regulation, will not be feasible for large scale problems.
+We considered two cases,  control on/off, for each network configuration. 
+Each of these cases had identical kinetic parameters and initial conditions; 
+the \textit{only} differences between the cases was the allosteric regulation rules, and the control parameters associated with these rules. 
+As expected, end product accumulation was larger for network A when the control was off (no feedback inhibition of $E_{1}$ by $P_{1}$ and $P_{2}$),
+as compared to the on case (Fig. \ref{fig-onoff-simulations}A). We found this behavior was robust to the choice of underlying kinetic parameters, 
+as we observed that same qualitative response across an ensemble of randomized parameter sets (N = 100). 
+The control on/off response of network B was more subtle. In the off case, the behavior was qualitatively similar to network A. 
+However, for the on case, flux was diverted away from $P_{2}$ formation by feedback inhibition of $E_{6}$ activity at the $M_{2}$ branch point by $P_{1}$ (Fig. \ref{fig-onoff-simulations}B).
+Lower $E_{6}$ activity at the $M_{2}$ branch point allowed more flux toward $P_{1}$ formation, hence the yield of $P_{1}$ also increased (Fig. \ref{fig-onoff-simulations}C).
+Again, the control on/off behavior was robust to the values of the kinetic parameters, as the same qualitative trend was conserved across 
+an ensemble of possible randomized kinetic parameters (N = 100). Taken together, these simulations suggested that the rule based allosteric control 
+concept could robustly capture expected feedback behavior.
+
+\subsection*{Estimating parameters and effective allosteric regulatory structures.}
+A critical challenge for any dynamic model is the estimation of kinetic parameters. 
+For metabolic processes, there is also the added challenge of identifying the regulation and control structures that manage metabolism.  
+Of course, these issues are not independent; any description of enzyme activity regulation will be a function of system state, which
+in turn depends upon the kinetic parameters. For cell free systems, regulated gene expression has been removed, however, enzyme activity regulation is still operational. 
+We explored this linkage by estimating model parameters from synthetic data using both network
+structures. We generated noise-corrupted synthetic measurements of the substrate $S$, intermediate $M_{5}$ and end-product $P_1$ approximately every 20 min using
+network A. We then generated an ensemble of model parameter estimates by minimizing the difference between model simulations and the synthetic data using particle swarm
+optimization, starting from random initial guesses. The estimation of kinetic parameters was sensitive to the choice of regulatory structure (Fig. \ref{fig-parameter-fit}).
+PSO identified an ensemble of parameters that bracketed the mean of the synthetic measurements in less than 1000 iterations when the control structure was correct (Fig.
+\ref{fig-parameter-fit}A and B).However, when there was network mismatch (network B fit against network A synthetic data), PSO unable to identify an ensemble, all else being the same (Fig.
+\ref{fig-parameter-fit}C and D).
+Interestingly, the particle swarm generated a \textit{sloppy} parameter ensemble, in the absence of network mismatch (Fig.
+ZZZ).
+Taken together, ... [FINISH].
+
+We modified our particle swarm identification strategy to simultaneously search over both kinetic parameters and putative control structures.
+In addition to our initial networks, we constructed three additional presumptive network models, each with the same enzymatic connectivity but different allosteric 
+regulation of the pathway enzymes (Fig. \ref{fig-alternative-networks}). 
+We then initialized a population of particles, each with one of the five potential regulatory programs, and randomized kinetic parameters.
+Thus, we generated an initial population of particles that had \textit{both} different kinetic parameters as well as different control structures. 
+We biased the distribution of the particle population according to our \textit{a~prior} belief of the correct regulatory program.
+To this end, we considered three different priors, a uniform distribution where each putative regulatory structure represented 20\% of the population, and
+two mixed distributions that were either positively or negatively biased towards the correct structure (network A).
+In both the positively biased, and uniform cases the particle swarm clearly differentiated between the true or closely related structures and those that were materially different (Fig.
+\ref{fig-control-search}).
+As expected, the positively biased population (40\% of the initial particle population seeded with network A) gave the best results, where the correct structure was preferentially identified (Fig.
+\ref{fig-control-search}A).
+On the other hand, when given a uniform distribution, 
+the PSO approach identified a combination of network A (ZZ) and network C (YY) as the most likely control structures (Fig. \ref{fig-control-search}B).
+Network A and C differ by the regulatory connection between the end-product $P_2$ and enzyme $E_{1}$; in network A $P_{2}$ was assumed to inhibit $E_{1}$ while in network C $P_{2}$ was assumed to activate $E_{1}$.
+Lastly, when the initial population was biased towards a completely incorrect structure (initial population seeded with 40\% network B),
+the particle swarm misidentified the correct structure (Fig. \ref{fig-control-search}C).
 
@@
 \section*{Discussion}
+In this study, we proposed a effective modeling strategy to dynamically simulate cell free metabolic networks. 
+We tested this strategy using two proof of concept metabolic networks.
+In both networks, substrate $S$ was converted to the end-products $P_{1}$ and $P_{2}$ through a series of enzymatically catalyzed reactions, 
+including a branch point at hypothetical metabolite $M_{2}$. While both networks had the same enzymatic connectivity, that had differing control structures.
+[FINISH]
+
+Cybernetic models, other dynamic models of metabolism.
+
+While the results of this study were encouraging, there are several critical next steps that must be accomplished before we can model genome scale cell free metabolic networks.
+[FINISH]
 
@@
 
-\subsection*{Formulation and solution of coagulation model equations.}
-\subsection*{Optimization methods.}
+\subsection*{Formulation and solution of the model equations.}
+We used ordinary differential equations (ODEs) to model the time evolution of metabolite ($x_{i}$) and scaled enzyme abundance ($\epsilon_{i}$) in hypothetical cell free
+metabolic networks:
+\begin{eqnarray}
+	\frac{dx_{i}}{dt} & = & \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)\qquad{i=1,2,\hdots,\mathcal{M}}\\
+	\frac{d\epsilon_{i}}{dt} & = & -\lambda_{i}\epsilon_{i}\qquad{i = 1,2,\hdots,\mathcal{E}}
+\end{eqnarray}where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of metabolites and $\mathcal{E}$ denotes the number of
+enzymes in the model. The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ 
+denotes the rate of reaction $j$. Typically, reaction $j$ is a non-linear function of metabolite and enzyme abundance, as well as unknown kinetic parameters 
+$\mathbf{k}$ ($\mathcal{K}\times{1}$).
+The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$. If 
+$\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$. Conversely, if $\sigma_{ij}>0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates
+metabolite $i$ is not connected with reaction $j$. Lastly, $\lambda_{i}$ denotes the scaled enzyme degradation constant. The system material balances were subject to the
+initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$ and $\mathbf{\epsilon}\left(t_{o}\right) = \mathbf{1}$ (initially we have 100\% cell-free enzyme abundance).
+
+Each reaction rate was written as the product of two terms, a kinetic term ($\bar{r}_{j}$) and a regulatory term ($v_{j}$):
+\begin{equation}\label{eqn:rate-factor}
+	r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right) = \bar{r}_{j}v_{j}
+\end{equation}We used multiple saturation kinetics to model the reaction term $\bar{r}_{j}$:
+\begin{equation}\label{eqn:rate-bar}
+	\bar{r}_{j} = k_{j}^{max}\epsilon_{i}\left(\prod_{s\in{m_{j}^{-}}}\frac{x_{s}}{K_{js} + x_{s}}\right)
+\end{equation}where $k_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the scaled enzyme activity which catalyzes reaction $j$, and
+$K_{js}$ denotes the saturation constant for species $s$ in reaction $j$. 
+The product in Eqn. \eqref{eqn:rate-bar} was carried out over the set of \textit{reactants} for reaction $j$ (denoted as $m_{j}^{-}$).
+
+The allosteric regulation term $v_{j}$ depended upon the combination of factors which influenced the activity of enzyme $i$.
+For each enzyme, we used a rule based approach to select from competing control factors (Fig. \ref{fig-control-schematic}). 
+If an enzyme was activated by $m$ metabolites, we modeled this activation as:
+\begin{equation}
+	v_{j} = \max\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right)
+\end{equation}where $0\leq f_{ij}\left(x\right)\leq 1$ was a regulatory transfer function that calculated the influence of metabolite $i$ on the activity of enzyme $j$. 
+Conversely, if enzyme activity was inhibited by a $m$ metabolites, we modeling this inhibition as:
+\begin{equation}
+	v_{j} = 1 - \max\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right)
+\end{equation}Lastly, if an enzyme had both $m$ activating and $n$ inhibitory factors, we modeled the regulatory term as:
+\begin{equation}
+	v_{j} = \min\left(u_{j},d_{j}\right)
+\end{equation}where:
+\begin{eqnarray}
+	u_{j} &=& \max_{j^{+}}\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right) \\
+	d_{j} &=& 1 - \max_{j^{-}}\left(f_{1j}\left(x\right),\hdots,f_{nj}\left(x\right)\right)
+\end{eqnarray}The quantities $j^{+}$ and $j^{-}$ denoted the sets of activating, and inhibitory factors for enzyme $j$. 
+If an enzyme had no allosteric factors, we set $v_{j} = 1$.
+There are many possible functional forms for $0\leq f_{ij}\left(x\right)\leq 1$. 
+However, in this study, each individual transfer function took the form:
+\begin{equation}\label{eqn:control-factor}
+	f_{i}\left(\mathbf{x}\right) = \frac{\kappa_{ij}^{\eta}x_{j}^{\eta}}{1 + \kappa_{ij}^{\eta}x_{j}^{\eta}}
+\end{equation}where $x_{j}$ denotes the abundance of metabolite $j$, and $\kappa_{ij}$ and $\eta$ are control parameters. 
+The $\kappa_{ij}$ parameter was species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).
+The model equations were encoded using the Octave programming language, and solved using the LSODE routine in Octave \citep{Octave:2014}.
+
+\subsection*{Estimation of model parameters and structures from synthetic experimental data.}
+Model parameters were estimated by minimizing the difference between simulations and synthetic experimental data (squared residual):
+\begin{equation}\label{eqn:objective-function}
+	\min_{\mathbf{k}} \sum_{\tau=1}^{\mathcal{T}}\sum_{j=1}^{\mathcal{S}}\left(\frac{\hat{x}_{j}\left(\tau\right) - x_{j}\left(\tau,\mathbf{k}\right)}{\omega_{j}\left(\tau\right)}\right)^{2}
+\end{equation}where $\hat{x}_{j}\left(\tau\right)$ denotes the measured value of species $j$ at time $\tau$, $x_{j}\left(\tau,\mathbf{k}\right)$ denotes the simulated 
+value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$. The outer summation is respect to
+time, while the inner summation is with respect to state. We approximated a realistic model identification scenario, assuming noisy experimental data, 
+limited sampling resolution (approximately 20 minutes per sample) and a limited number of measurable metabolites.
+
+We minimized the model residual using Particle swarm optimization (PSO) \citep{PSO}.
+PSO uses a \textit{swarming} metaheuristic to explore parameter spaces. 
+A strength of PSO is its ability to find the global minimum, even in the presence of potentially many local minima, by communicating the local
+error landscape experienced by each particle collectively to the swarm. Thus, PSO acts both as a local and a global search algorithm. 
+For each iteration, particles in the swarm compute their local error by evaluating the model equations using their specific parameter vector realization.
+From each of these local points, a globally best error is identified. Both the local and global error 
+are then used to update the parameter estimates of each particle using the rules:
+\begin{eqnarray}
+	\mathbf{\Delta}_{i} &=&\theta_{1}\mathbf{\Delta}_{i} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{k}_{i}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G} - \mathbf{k}_{i}\right) \\
+	\mathbf{k}_{i} &=& \mathbf{k}_{i} + \mathbf{\Delta}_{i}
+\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes local best solution found by particle $i$, and
+$\mathcal{G}$ denotes the best solution found over the entire population of particles.
+The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
+parameters ($\mathcal{K}\times{1}$). In thus study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$, which was taken from XXX. The quality of parameter
+estimates was measured using two criteria, goodness of fit (model residual) and angle between the estimated parameter vector $\mathbf{k}_{j}$ and the true parameter set $\mathbf{k}^{*}$:
+\begin{equation}
+	\alpha_{j} = cos^{-1}\left(\frac{\mathbf{k}_{j}\cdot{\mathbf{k^{*}}}}{\norm{\mathbf{k}_{j}}\norm{\mathbf{k^{*}}}}\right)
+\end{equation}If the candidate parameter set $\mathbf{k}_{j}$ were perfect, the residual between the model and synthetic data 
+and the angle between $\mathbf{k}_{j}$ and the true parameter set $\mathbf{k}^{*}$ would be equal to zero.
+
+We modified our PSO implementation to simultaneously search over kinetic parameters and putative model control structures.
+In the combined case, each particle potentially carried a different model realization in addition to a different kinetic parameter vector. 
+We kept the update rules the same (along with the update parameters). 
+Thus, each particle competed on the basis of goodness of fit, which allowed different model structures to contribute to the overall behavior of the swarm. 
+We considered five possible model structures (A through E), where network A was the correct formulation (used to generate the synthetic data). 
+We considered a population N = 100 particles, where each particle in the swarm was assigned a model structure, and a random parameter vector.
+The PSO algorithm, model equations, and the objective function were encoded and solved in the Octave programming language \citep{Octave:2014}.
 
@@
 \clearpage
+%\bibliographystyle{plain}
+%\bibliographystyle{IEEEbib}
 
@@
 %\begin{comment}
+
+\clearpage
+
+\begin{figure}
+\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure-1-Networks.pdf}
+\caption{Proof of concept cell-free metabolic networks considered in this study. Substrate $S$ is converted to products $P_{1}$ and $P_{2}$ through a series of chemical conversions
+catalyzed by enzyme(s) $E_{j}$. The activity of the pathway enzymes is subject to both positive and negative allosteric regulation. }\label{fig-networks}
+\end{figure}
+
+\clearpage
+
+\begin{figure}
+\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure-2-ControlSchematic.pdf}
+\caption{Schematic of the rule based allosteric enzyme activity control laws. }\label{fig-control-schematic}
+\end{figure}
+
+\clearpage
+
+\begin{figure}
+\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure-3-EnzymeKinetics.pdf}
+\caption{Kinetics of simple transformations in the presence of activation and inhibition. 
+A:The conversion of substrate $S$ to product $P$ by enzyme $E$ was activated by $S$. 
+B:The conversion of substrate $S$ to product $P$ by enzyme $E$ was inhibited by inhibitor $I$. 
+}\label{fig-kinetics-simulations}
+\end{figure}
+
+\clearpage
+
+\begin{figure}
+\centering
+\includegraphics[width=0.55\textwidth]{./figs/Figure-4-OnOffSimulations.pdf}
+\caption{On/off control simulations for network A and network B for an ensemble of kinetic parameter sets versus time. 
+For each case, N = 100 simulations were conducted using kinetic and initial conditions randomly generated from a hypothetical true parameter set. 
+The gray area represents $\pm$ one standard deviation surrounding the mean. Control parameters were fixed during the ensemble calculations.}\label{fig-onoff-simulations}
+\end{figure}
+
+\clearpage
+
+\begin{figure}
+\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure-5-ParameterFit.pdf}
+\caption{Parameter estimation from synthetic data for the same and mismatched allosteric control logic. }\label{fig-parameter-fit}
+\end{figure}
+
+\clearpage
+
+\begin{figure}
+\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure-6-AlternativeNetworks.pdf}
+\caption{Schematic of the alternative allosteric control programs used in the structural particle swarm computation. 
+Each network had the same enzymatic connectivity, initial conditions and kinetic parameters, but alternative feedback control structures for the first enzyme in the pathway.}\label{fig-alternative-networks}
+\end{figure}
+
+\clearpage
+
+\begin{figure}
+\centering
+\includegraphics[width=0.6\textwidth]{./figs/Figure-7-ControlSearch.pdf}
+\caption{Combined control structure and kinetic parameter search results. Think about this ...}\label{fig-control-search}
+\end{figure}
 

### 2014-08-30 Paper_v1.tex
New fig and Fig 8

@@
 Lastly, when the initial population was biased towards a completely incorrect structure (initial population seeded with 40\% network B),
-the particle swarm misidentified the correct structure (Fig. \ref{fig-control-search}C).
+the particle swarm \textit{misidentified} the correct allosteric structure (Fig. \ref{fig-control-search}C).
 
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=1.0\textwidth,height=0.6\textheight]{./figs/Figure-8-Flux.pdf}
+\caption{Flux and control pattern for estimated model parameters and structures. }\label{fig-flux-pattern}
+\end{figure}
+
+\clearpage
+
 % Supplemental figures -

### 2014-08-31 Paper_v1.tex
Updated captions for Fig 1 - 6

@@
 \includegraphics[width=1.0\textwidth]{./figs/Figure-2-ControlSchematic.pdf}
-\caption{Schematic of the rule based allosteric enzyme activity control laws. }\label{fig-control-schematic}
+\caption{Schematic of rule based allosteric enzyme activity control laws. 
+Traditional enzyme kinetic expressions e.g., Michaelis–Menten or multiple saturation kinetics are multiplied by an enzyme activity control variable $0 \leq v_{j} \leq 1 $. 
+Control variables are functions of many possible regulatory factors encoded by arbitrary functions of the form $0\leq f_{j}\left(\mathcal{Z}\right)\leq 1$.
+At each simulation time step, the $v_{j}$ variables are calculated by evaluating integration rules such as the max or min of the set of factors $f_{1},\hdots$ 
+influencing the activity of enzyme $E_{j}$. }\label{fig-control-schematic}
 \end{figure}
@@
 \caption{Kinetics of simple transformations in the presence of activation and inhibition. 
-A:The conversion of substrate $S$ to product $P$ by enzyme $E$ was activated by $S$. 
-B:The conversion of substrate $S$ to product $P$ by enzyme $E$ was inhibited by inhibitor $I$. 
+\textbf{A}:The conversion of substrate $S$ to product $P$ by enzyme $E$ was activated by $S$. 
+For a fixed control gain parameter $k_{control}$, the reaction rate approached a step for increasing control order $N$. 
+\textbf{B}:The conversion of substrate $S$ to product $P$ by enzyme $E$ with inhibitor $I$. 
+For a fixed control gain parameter $k_{control}$, the reaction rate approximated non-competitive inhibition for increasing control order $N$. 
 }\label{fig-kinetics-simulations}
@@
 \centering
-\includegraphics[width=0.55\textwidth]{./figs/Figure-4-OnOffSimulations.pdf}
-\caption{On/off control simulations for network A and network B for an ensemble of kinetic parameter sets versus time. 
-For each case, N = 100 simulations were conducted using kinetic and initial conditions randomly generated from a hypothetical true parameter set. 
-The gray area represents $\pm$ one standard deviation surrounding the mean. Control parameters were fixed during the ensemble calculations.}\label{fig-onoff-simulations}
+\includegraphics[width=0.50\textwidth]{./figs/Figure-4-OnOffSimulations.pdf}
+\caption{ON/OFF control simulations for network A and network B for an ensemble of kinetic parameter sets versus time (N = 100). 
+For each case, N = 100 simulations were conducted using kinetic and initial conditions generated randomly from a hypothetical true parameter set. 
+The gray area represents $\pm$ one standard deviation surrounding the mean. 
+Control parameters were fixed during the ensemble calculations.
+\textbf{A}: End-product P$_{1}$ abundance versus time for Network A. 
+The abundance of P$_{1}$ decreased with end-product inhibition of $E_{1}$ activity (Control-ON) versus the no inhibition case (Control-OFF). 
+\textbf{B}: End-product P$_{2}$ abundance versus time for Network B. Inhibition of branch point $E_{6}$ by end-product P$_{1}$ decreased P$_{2}$ abundance (Control-ON) versus the
+no inhibition case (Control-OFF).
+\textbf{C}: End-product P$_{1}$ abundance versus time for Network A. 
+Inhibition of branch point $E_{6}$ by end-product P$_{1}$ decreased P$_{1}$ abundance (Control-ON) versus the no inhibition case (Control-OFF).
+}\label{fig-onoff-simulations}
 \end{figure}
@@
 \includegraphics[width=1.0\textwidth]{./figs/Figure-5-ParameterFit.pdf}
-\caption{Parameter estimation from synthetic data for the same and mismatched allosteric control logic. }\label{fig-parameter-fit}
+\caption{Parameter estimation from synthetic data for the same and mismatched allosteric control logic using particle swarm optimization (PSO). 
+Synthetic experimental data was generated from a hypothetical parameter set using Network A, 
+where substrate $S$, end-product P$_{1}$ and intermediate M$_5$ were sampled approximately every 20 minutes. 
+For cases $\textbf{A,B}$ 20 particles were initialized with randomized parameters and allowed to search for 300 iterations. 
+\textbf{A,B}: PSO estimated an ensemble of parameters sets (N = 20) consistent with the synthetic experimental data assuming the correct
+enzymatic and control connectivity starting from randomized initial parameters.
+\textbf{C,D}: In the presence of control mismatch (Network B control policy simulated with Network A kinetic parameters) 
+the ensemble of models did not describe the synthetic data. 
+}\label{fig-parameter-fit}
 \end{figure}
@@
 \caption{Schematic of the alternative allosteric control programs used in the structural particle swarm computation. 
-Each network had the same enzymatic connectivity, initial conditions and kinetic parameters, but alternative feedback control structures for the first enzyme in the pathway.}\label{fig-alternative-networks}
+Each network had the same enzymatic connectivity, initial conditions and kinetic parameters, 
+but alternative feedback control structures for the first enzyme in the pathway.}\label{fig-alternative-networks}
 \end{figure}
@@
 \centering
-\includegraphics[width=0.6\textwidth]{./figs/Figure-7-ControlSearch.pdf}
+\includegraphics[width=0.5\textwidth]{./figs/Figure-7-ControlSearch.pdf}
 \caption{Combined control structure and kinetic parameter search results. Think about this ...}\label{fig-control-search}

### 2014-08-31 Paper_v1.tex
Updated captions for Fig 7 - 8

@@
 \begin{titlepage}
-{\par\centering\textbf{\Large Dynamic Modeling of Biochemical Networks using Effective Kinetic Models}}
+{\par\centering\textbf{\Large Dynamic Modeling of Cell-Free Biochemical Networks using Effective Kinetic Models}}
 \vspace{0.05in}
@@
 \section*{Introduction}
-Biochemical processes are widely used in biotechnology to produce an array of products including commodity chemicals and protein therapeutics.
+Biological processes are widely used in industrially to produce an array of products from commodity chemicals and to high-value protein therapeutics.
 However, whole-cell processes share the central limitation of requiring cell growth, 
@@
 \includegraphics[width=0.5\textwidth]{./figs/Figure-7-ControlSearch.pdf}
-\caption{Combined control structure and kinetic parameter search results. Think about this ...}\label{fig-control-search}
+\caption{Combined control and kinetic parameter search using modified particle swarm optimization (PSO). 
+A population of N = 100 particles was initialized with randomized kinetic parameters and one of five possible control configurations (Network A - E).
+Simulation error was minimized for a synthetic data set 
+($S$, end-product P$_{1}$ and intermediate M$_5$ sampled approximately every 20 min) generated using Network A.
+\textbf{A}: Simulation error versus parameter set angle for N = 100 particles biased toward the correct regulatory program (A,B,C,D,E) = (40\%, 30\%, 10\%,10\% and 10\%).
+\textbf{B}: Simulation error versus parameter set angle for N = 100 uniformly distributed particles (A,B,C,D,E) = (20\%, 20\%, 20\%, 20\% and 20\%).
+\textbf{C}: Simulation error versus parameter set angle for N = 100 negatively biased particles (A,B,C,D,E) = (10\%, 10\%, 10\%, 30\% and 40\%).
+Network A (the correct structure) was preferentially identified for positively and uniform biased particle distributions, but misidentified in the presence of
+a large incorrect bias. 
+}\label{fig-control-search}
 \end{figure}
@@
 \includegraphics[width=1.0\textwidth,height=0.6\textheight]{./figs/Figure-8-Flux.pdf}
-\caption{Flux and control pattern for estimated model parameters and structures. }\label{fig-flux-pattern}
+\caption{Metabolic flux and control program output as a function of network type and parameter estimate.
+The population of particles from the modified particle swarm optimization strategy were sorted based upon simulation error. 
+The scaled metabolic flux and control variables governing the activity of $E_{1},E_{3}$ and $E_{6}$ were plotted for each particle in the swarm.
+\textbf{A}: Model performance for the N = 100 positively biased particles.
+\textbf{B}: Model performance for the N = 100 uniformly biased particles.
+\textbf{C}: Model performance for the N = 100 negatively biased particles.
+}\label{fig-flux-pattern}
 \end{figure}

### 2014-08-31 Paper_v1.tex
working on last paragraph of results

@@
 We developed two proof-of-concept metabolic networks to investigate the features of our effective biochemical network modeling approach (Fig. \ref{fig-networks}).
-In both examples, substrate $S$ was converted to the end-products $P_{1}$ and $P_{2}$ through a series of enzymatically catalyzed reactions, 
+In both examples, substrate S was converted to the end-products P$_{1}$ and P$_{2}$ through a series of enzymatically catalyzed reactions, 
 including a branch point at hypothetical metabolite $M_{2}$. 
 Several of these reactions involved cofactor dependence ($AH$ or $A$), and various allosteric regulation mechanisms. 
-Network A included feedback inhibition of the initial pathway enzyme ($E_{1}$) by pathway end products $P_{1}$ and $P_{2}$ (Fig. \ref{fig-networks}A).
-On the other hand, network B involved feedback inhibition of $E_{1}$ by $P_{2}$ and $E_{6}$ by $P_{1}$ (Fig. \ref{fig-networks}B). 
+Network A included feedback inhibition of the initial pathway enzyme ($E_{1}$) by pathway end products P$_{1}$ and P$_{2}$ (Fig. \ref{fig-networks}A).
+On the other hand, network B involved feedback inhibition of $E_{1}$ by P$_{2}$ and $E_{6}$ by P$_{1}$ (Fig. \ref{fig-networks}B). 
 In both networks, branch point enzymes $E_{3}$ and $E_{6}$ were subject to feed-forward activation by cofactor $AH$.
@@
 the \textit{only} differences between the cases was the allosteric regulation rules, and the control parameters associated with these rules. 
-As expected, end product accumulation was larger for network A when the control was off (no feedback inhibition of $E_{1}$ by $P_{1}$ and $P_{2}$),
+As expected, end product accumulation was larger for network A when the control was off (no feedback inhibition of $E_{1}$ by P$_{1}$ and P$_{2}$),
 as compared to the on case (Fig. \ref{fig-onoff-simulations}A). We found this behavior was robust to the choice of underlying kinetic parameters, 
@@
 The control on/off response of network B was more subtle. In the off case, the behavior was qualitatively similar to network A. 
-However, for the on case, flux was diverted away from $P_{2}$ formation by feedback inhibition of $E_{6}$ activity at the $M_{2}$ branch point by $P_{1}$ (Fig. \ref{fig-onoff-simulations}B).
-Lower $E_{6}$ activity at the $M_{2}$ branch point allowed more flux toward $P_{1}$ formation, hence the yield of $P_{1}$ also increased (Fig. \ref{fig-onoff-simulations}C).
-Again, the control on/off behavior was robust to the values of the kinetic parameters, as the same qualitative trend was conserved across 
-an ensemble of possible randomized kinetic parameters (N = 100). Taken together, these simulations suggested that the rule based allosteric control 
+However, for the on case, flux was diverted away from P$_{2}$ formation by feedback inhibition of $E_{6}$ activity at the M$_{2}$ branch point by P$_{1}$ (Fig. \ref{fig-onoff-simulations}B).
+Lower $E_{6}$ activity at the M$_{2}$ branch point allowed more flux toward P$_{1}$ formation, hence the yield of P$_{1}$ also increased (Fig. \ref{fig-onoff-simulations}C).
+Again, the control on/off behavior was robust to changes kinetic parameters, as the same qualitative trend was conserved across 
+an ensemble of randomized parameters (N = 100). Taken together, these simulations suggested that the rule based allosteric control 
 concept could robustly capture expected feedback behavior.
@@
 network A. We then generated an ensemble of model parameter estimates by minimizing the difference between model simulations and the synthetic data using particle swarm
-optimization, starting from random initial guesses. The estimation of kinetic parameters was sensitive to the choice of regulatory structure (Fig. \ref{fig-parameter-fit}).
-PSO identified an ensemble of parameters that bracketed the mean of the synthetic measurements in less than 1000 iterations when the control structure was correct (Fig.
-\ref{fig-parameter-fit}A and B).However, when there was network mismatch (network B fit against network A synthetic data), PSO unable to identify an ensemble, all else being the same (Fig.
-\ref{fig-parameter-fit}C and D).
-Interestingly, the particle swarm generated a \textit{sloppy} parameter ensemble, in the absence of network mismatch (Fig.
-ZZZ).
-Taken together, ... [FINISH].
+optimization, starting from random initial parameter guesses. The estimation of kinetic parameters was sensitive to the choice of regulatory structure (Fig. \ref{fig-parameter-fit}).
+PSO identified an ensemble of parameters that bracketed the mean of the synthetic measurements in less than 1000 iterations 
+when the control structure was correct (Fig. \ref{fig-parameter-fit}A and B).
+However, with control mismatch (network B simulated with network A parameters), the model simulations were not consistent with the synthetic data  (Fig. \ref{fig-parameter-fit}C and D).
 
@@
 On the other hand, when given a uniform distribution, 
-the PSO approach identified a combination of network A (ZZ) and network C (YY) as the most likely control structures (Fig. \ref{fig-control-search}B).
-Network A and C differ by the regulatory connection between the end-product $P_2$ and enzyme $E_{1}$; in network A $P_{2}$ was assumed to inhibit $E_{1}$ while in network C $P_{2}$ was assumed to activate $E_{1}$.
+the PSO approach identified a combination of network A and network C as the most likely control structures (Fig. \ref{fig-control-search}B).
+Network A and C differ by the regulatory connection between the end-product P$_2$ and enzyme $E_{1}$; in network A end-product P$_{2}$ was assumed to inhibit $E_{1}$ while in network C end-product P$_{2}$ activated $E_{1}$.
 Lastly, when the initial population was biased towards a completely incorrect structure (initial population seeded with 40\% network B),
 the particle swarm \textit{misidentified} the correct allosteric structure (Fig. \ref{fig-control-search}C). 
+Interestingly, each particle swarm identified parameter sets and model structures that minimized the simulation error. 
+However, the estimated parameter values were not necessarily similar to the true parameters; the angle between the estimated and true parameters was not consistently small. 
+This suggested that the particle swarm approach identified a \textit{sloppy} ensemble, i.e., parameter estimates were individually incorrect but collectively exhibited the correct 
+model behavior.
+
+We calculated control program output and scaled metabolic flux for the positively, uniformly and negatively biased particle swarms (Fig. \ref{fig-flux-pattern}).
+Network A and network C models from the positively (Fig. \ref{fig-flux-pattern}A) and uniformly (Fig. \ref{fig-flux-pattern}B) 
+biased particle swarms showed similar operational patterns, despite differences in the kinetic parameters and control structures. 
+In all cases, irregardless of network configuration or parameter values, the rate of enzyme decay was small compared to the other fluxes. 
+Consistent with the correct model structure, production of end-product P$_1$ was the preferred branch. 
+High flux through this branch resulted in significant end product inhibition of $E_{1}$ in both network A and network C, when compared to network B,D and E.
+Interestingly, the behavior of network A was degenerate in the presence of high P$_{1}$ production; 
+positive or negative feedback connections from P$_2$ to $E_{1}$ were ignored because of our choice of integration rule (max). 
+Thus, while theoretically distinct, network A and network C were operationally similar.
+On the other hand, networks B, D and E showed distinct behavior that was not consistent with the true network.  
+These architectures exhibited either limited inhibition (network B) or activation (network D and E) of $E_{1}$ activity, resulting a significantly different
+metabolic flux profile. Taken together ... [FINISH ME].
 
@@
 We tested this strategy using two proof of concept metabolic networks.
-In both networks, substrate $S$ was converted to the end-products $P_{1}$ and $P_{2}$ through a series of enzymatically catalyzed reactions, 
+In both networks, substrate $S$ was converted to the end-products P$_{1}$ and P$_{2}$ through a series of enzymatically catalyzed reactions, 
 including a branch point at hypothetical metabolite $M_{2}$. While both networks had the same enzymatic connectivity, that had differing control structures.
@@
 \includegraphics[width=1.0\textwidth]{./figs/Figure-1-Networks.pdf}
-\caption{Proof of concept cell-free metabolic networks considered in this study. Substrate $S$ is converted to products $P_{1}$ and $P_{2}$ through a series of chemical conversions
+\caption{Proof of concept cell-free metabolic networks considered in this study. Substrate $S$ is converted to products P$_{1}$ and P$_{2}$ through a series of chemical conversions
 catalyzed by enzyme(s) $E_{j}$. The activity of the pathway enzymes is subject to both positive and negative allosteric regulation. }\label{fig-networks}
@@
 \includegraphics[width=1.0\textwidth,height=0.6\textheight]{./figs/Figure-8-Flux.pdf}
-\caption{Metabolic flux and control program output as a function of network type and parameter estimate.
-The population of particles from the modified particle swarm optimization strategy were sorted based upon simulation error. 
-The scaled metabolic flux and control variables governing the activity of $E_{1},E_{3}$ and $E_{6}$ were plotted for each particle in the swarm.
-\textbf{A}: Model performance for the N = 100 positively biased particles.
-\textbf{B}: Model performance for the N = 100 uniformly biased particles.
-\textbf{C}: Model performance for the N = 100 negatively biased particles.
+\caption{Metabolic flux and control variables as a function of network type and particle index at t = 100 min.
+The control variables governing $E_{1},E_{3}$ and $E_{6}$ activity and the scaled metabolic flux and were calculated for the positively, 
+uniformly and negatively biased particle swarms (N = 100). The particles from each swarm were sorted based upon simulation error (low to high error). 
+\textbf{A}: Model performance for the positively biased particle swarm as a function of particle index.
+\textbf{B}: Model performance for the uniformly biased particle swarm as a function of particle index.
+\textbf{C}: Model performance for the negatively biased particle swarm as a function of particle index. 
+Models with significant control mismatch showed distinct control and flux patterns versus those models with the correct or closely related control policies.
+In particular, models with the correct control policy showed stronger inhibition of $E_{1}$ activity, leading to decreased flux from S$\rightarrow$P$_{1}$.
+Conversely, models with significant mismatch had increased $E_{1}$ activity, leading to an altered flux distribution. 
+This is especially apparent in the negatively biased particle swarm. 
 }\label{fig-flux-pattern}

### 2014-09-01 Paper_v1.tex
background paragraphs in the introduction

@@
 \section*{Introduction}
-Biological processes are widely used in industrially to produce an array of products from commodity chemicals and to high-value protein therapeutics.
-However, whole-cell processes share the central limitation of requiring cell growth, 
-which redirects resources away from product synthesis, and cell walls, which complicate interrogation and control of intracellular metabolic processes.
-On the other hand, cell-free metabolic systems offer many advantages over traditional in vivo production methods. 
-For example, cell-free systems can direct scarce metabolic resources exclusively towards a single product of interest. 
-Moreover, with no cell wall, cell free systems can more easily be interrogated, and substrates of the metabolite processes directly controlled.
-Cell free production offers the unique opportunity to study metabolism without the complication of cell growth and gene expression processes.
-For modeling, this implies that we need only consider allosteric regulation of enzyme activity when building and testing cell free metabolic models.
-Of course, modeling allosteric mechanisms is itself a difficult problem when the model is at a whole genome scale. To address this problem, we have developed a
-an approach based upon the constrained fuzzy logic framework of Morris and Lauffenburger [REFHERE].
+Mathematical modeling has long contributed to our understanding of metabolism. 
+Decades before the genomics revolution, mechanistically, structured metabolic models arose from the desire to predict microbial phenotypes resulting from 
+changes in intracellular or extracellular states \citep{1976_fredrickson_BiotechBioeng}. 
+The single cell \textit{E. coli} models of Shuler and coworkers pioneered the construction of large-scale, dynamic metabolic models that incorporated multiple, regulated catabolic and anabolic 
+pathways constrained by experimentally determined kinetic parameters \citep{1984_domach_shuler_BiotechBioeng_01}. 
+Shuler and coworkers generated many single cell kinetic models, including single cell models of eukaryotes \citep{1989_steinmeyer_shuler_ChemEngSci,1992_wu_shuler_AnnNYAcadSci}, minimal cell architectures \citep{2004_castellanos_shuler_PNAS}, as well as DNA sequence based whole-cell models of \textit{E.
+coli} \citep{2008_atlas_shuler_IETSysBio}.
+Conversely, highly abstracted kinetic frameworks, such as the cybernetic framework, represented a paradigm shift, viewing cells as growth-optimizing strategists \citep{1985_dhurjati_ramkrishna_tsao_BiotechBioeng}.
+Cybernetic models have been highly successful at predicting metabolic choice behavior, e.g., diauxie behavior \citep{1986_kompala_ramkrishna_tsao_BiotechBioeng}, 
+steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, as well as the cellular response to metabolic engineering modifications \citep{1999_varner_ramkrishna_MetaEng}.
 
-In this study, we present an effective biochemical network modeling framework that integrates classical kinetic modeling with rules based approaches.
-[FINISH ME]
+In the post genomics world, large-scale stoichiometric reconstructions of microbial metabolism popularized by static, constraint-based modeling techniques such as flux balance analysis (FBA)
+have become standard tools \citep{2012_lewis_palsson_NatRevMicrobio}.
+The first genome-scale stoichiometric model of \textit{E. coli}, developed by Edwards and Palsson \citep{2000_edwards_palsson_PNAS}, 
+was quickly followed by many other organism models \citep{2009_feist_palsson_NatRevMicrobio}. 
+Stoichiometric models rely on a pseudo-steady-state assumption to reduce unidentifiable genome-scale kinetic models to an underdetermined linear algebraic system, 
+which can be solved efficiently even for large systems.  
+Traditionally, stoichiometric models have also neglected explicit descriptions of metabolic regulation and control mechanisms, 
+instead opting to describe the choice of pathways by prescribing an objective function on metabolism. 
+Interestingly, similar to early cybernetic models, the most common metabolic objective function has been the optimization of biomass formation \citep{2002_ibarra_edwards_palsson_Nat}, 
+although other metabolic objectives have also been estimated \citep{2007_schuetz_sauer_MolSysBio}. 
+Recent advances in constraints based modeling have overcome the early shortcomings of the platform, including capturing metabolic regulation and control \citep{2013_hyduke_lewis_palsson_MolBioSys}.
+Thus, modern constraints based approaches have proven extremely useful in the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \citep{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}.
+
+Cell free systems offer unique advantages to study and model microbial metabolism when compared to traditional whole cell processes.
+[FINISH ME].
+
+In this study, we present an effective biochemical network modeling framework for building dynamic metabolic models.
+The key innovation of our approach is the seamless integration of simple effective rules encoding complex regulation, with traditional kinetic pathway modeling.
+This integration allows the description of complex regulatory interactions, such as time-dependent allosteric regulation of enzyme activity, in the absence of specific mechanistic information.
+Moreover, the rules are easy to understand, easy to formulate 
+and do not rely on overarching theoretical abstractions, or restrictive assumptions.
+We tested our approach by modeling the time evolution of hypothetical cell free metabolic networks. 
+In particular, we tested whether our effective modeling approach could describe classically expected behavior, 
+and second whether we could simultaneously estimate kinetic parameters and regulatory connectivity, in the absence of specific mechanistic knowledge, 
+from synthetic experimental data. Toward these questions, we explored five hypothetical cell-free networks. 
+Each network shared the same enzymatic connectivity, but had different allosteric regulatory connectivity. 
+We found that simple effective rules, when integrated with traditional enzyme kinetic expressions,
+captured complex allosteric patterns such as ultrasensitivity, or non-competitive inhibition in the absence of specific mechanistic information. 
+Moreover, when integrated into network models, these rules captured classical regulatory patterns such as product induced feedback inhibition. 
+Lastly, we simultaneously estimated kinetic parameters and discriminated between competing regulatory structures, using synthetic data in combination with a
+modified particle swarm approach. While only a proof-of-concept, [FINISH ME].
 
@@
 We explored this linkage by estimating model parameters from synthetic data using both network
-structures. We generated noise-corrupted synthetic measurements of the substrate $S$, intermediate $M_{5}$ and end-product $P_1$ approximately every 20 min using
+structures. We generated noise-corrupted synthetic measurements of the substrate $S$, intermediate M$_{5}$ and end-product P$_1$ approximately every 20 min using
 network A. We then generated an ensemble of model parameter estimates by minimizing the difference between model simulations and the synthetic data using particle swarm
-optimization, starting from random initial parameter guesses. The estimation of kinetic parameters was sensitive to the choice of regulatory structure (Fig. \ref{fig-parameter-fit}).
+optimization (PSO), starting from random initial parameter guesses. The estimation of kinetic parameters was sensitive to the choice of regulatory structure (Fig. \ref{fig-parameter-fit}).
 PSO identified an ensemble of parameters that bracketed the mean of the synthetic measurements in less than 1000 iterations 
@@
 However, with control mismatch (network B simulated with network A parameters), the model simulations were not consistent with the synthetic data  (Fig. \ref{fig-parameter-fit}C and D). 
+Taken together, these results suggested that we could perhaps simultaneously estimate both parameters and network control architectures, as incorrect control structures 
+would be manifest as poor model fits.
 
@@
 Network A and C differ by the regulatory connection between the end-product P$_2$ and enzyme $E_{1}$; in network A end-product P$_{2}$ was assumed to inhibit $E_{1}$ while in network C end-product P$_{2}$ activated $E_{1}$.
-Lastly, when the initial population was biased towards a completely incorrect structure (initial population seeded with 40\% network B),
+Lastly, when the initial population was biased towards a incorrect structures (initial population seeded with 90\% incorrect structures),
 the particle swarm \textit{misidentified} the correct allosteric structure (Fig. \ref{fig-control-search}C). 
-Interestingly, each particle swarm identified parameter sets and model structures that minimized the simulation error. 
-However, the estimated parameter values were not necessarily similar to the true parameters; the angle between the estimated and true parameters was not consistently small. 
-This suggested that the particle swarm approach identified a \textit{sloppy} ensemble, i.e., parameter estimates were individually incorrect but collectively exhibited the correct 
-model behavior.
+Interestingly, while each particle swarm identified parameter sets that minimized the simulation error, the estimated parameter values were not necessarily similar to the true parameters. 
+The angle between the estimated and true parameters was not consistently small across the swarms (identical parameters would give an angle of zero). 
+This suggested that our particle swarm approach identified a \textit{sloppy} ensemble, i.e., 
+parameter estimates that were individually incorrect but collectively exhibited the correct model behavior.
 
@@
 Network A and network C models from the positively (Fig. \ref{fig-flux-pattern}A) and uniformly (Fig. \ref{fig-flux-pattern}B) 
-biased particle swarms showed similar operational patterns, despite differences in the kinetic parameters and control structures. 
-In all cases, irregardless of network configuration or parameter values, the rate of enzyme decay was small compared to the other fluxes. 
-Consistent with the correct model structure, production of end-product P$_1$ was the preferred branch. 
-High flux through this branch resulted in significant end product inhibition of $E_{1}$ in both network A and network C, when compared to network B,D and E.
-Interestingly, the behavior of network A was degenerate in the presence of high P$_{1}$ production; 
-positive or negative feedback connections from P$_2$ to $E_{1}$ were ignored because of our choice of integration rule (max). 
-Thus, while theoretically distinct, network A and network C were operationally similar.
+biased particle swarms showed similar operational patterns, despite differences in kinetic parameters and control structures.
+While models from the negatively biased population had error values similar to the correct structures in the previous swarms,
+they have different flux and control profiles (Fig. \ref{fig-flux-pattern}C). 
+In all cases, irregardless of network configuration or parameter values, the rate of enzyme decay was small compared to the other fluxes,
+and all networks had qualitatively similar trends for $E_{3}$ and $E_{6}$ control.
+Moreover, consistent with the correct model structure, production of end-product P$_1$ was the preferred branch for all model configurations.
+However, there was variability in P$_{2}$ production flux across the population of models, especially for the uniform swarm when compared with the other cases.
+High P$_{1}$ branch flux resulted in end product inhibition of $E_{1}$ in both network A and network C, however in network D and E high P$_1$ flux induced $E_1$ activation. 
+These trends were manifested in different flux profiles, where the negatively biased population appeared more uniform across the population compared with the other swarms, and
+had higher $E_{1}$ specific activity. 
+Interestingly, the behavior of network A and network C highlighted an artifact of our integration rule; 
+both a positive or negative feedback connection from P$_2$ to $E_{1}$ were ignored because the P$_{1}$ inhibition of $E_{1}$ was dominate.  
+Thus, while theoretically distinct, network A and network C appeared operationally to the PSO algorithm to be that same network.
 On the other hand, networks B, D and E showed distinct behavior that was not consistent with the true network.  
-These architectures exhibited either limited inhibition (network B) or activation (network D and E) of $E_{1}$ activity, resulting a significantly different
-metabolic flux profile. Taken together ... [FINISH ME].
+These architectures exhibited either limited inhibition (network B) or activation (network D and E) of $E_{1}$ activity, resulting in significantly different
+metabolic flux profiles. However, the PSO was able to find low error parameter solutions, despite the mismatch in the control structures.
+Taken together, these results suggested that a uniform sampling approach could potentially yield an unbiassed estimate of both kinetic parameters control structures.
 
@@
 \section*{Discussion}
-In this study, we proposed a effective modeling strategy to dynamically simulate cell free metabolic networks. 
-We tested this strategy using two proof of concept metabolic networks.
-In both networks, substrate $S$ was converted to the end-products P$_{1}$ and P$_{2}$ through a series of enzymatically catalyzed reactions, 
-including a branch point at hypothetical metabolite $M_{2}$. While both networks had the same enzymatic connectivity, that had differing control structures.
-[FINISH]
-
-Cybernetic models, other dynamic models of metabolism.
+In this study, we presented an effective kinetic modeling strategy to dynamically simulate biochemical networks. 
+Our proposed strategy integrated traditional kinetic modeling with an effective rules based approach to dynamically describe metabolic regulation and control.
+We tested this approach by developing kinetic models of hypothetical cell free metabolic networks. 
+In particular, we tested whether our effective modeling approach could describe classically expected behavior, 
+and second whether we could simultaneously estimate kinetic parameters and regulatory connectivity, in the absence of specific mechanistic knowledge, 
+from synthetic experimental data. Toward these questions, we explored five hypothetical cell-free networks. In each network, a
+substrate S was converted to the end-products P$_{1}$ and P$_{2}$ through a series of enzymatically catalyzed reactions, 
+including a branch point at a hypothetical metabolite M$_{2}$. Each network also included the same cofactors and cofactor recycle architecture. 
+However, while all five networks shared the same enzymatic connectivity, each had different allosteric regulatory connectivity. 
+We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, could
+capture complex allosteric patterns such as ultrasensitivity, or non-competitive inhibition in the absence of specific mechanistic information. 
+Moreover, when integrated into network models, these rules captured classical regulatory patterns such as product induced feedback inhibition. 
+Lastly, we simultaneously estimated kinetic parameters and discriminated between competing regulatory structures, using synthetic data in combination with a
+modified particle swarm approach.
 
 While the results of this study were encouraging, there are several critical next steps that must be accomplished before we can model genome scale cell free metabolic networks.
-[FINISH]
 
@@
 ($S$, end-product P$_{1}$ and intermediate M$_5$ sampled approximately every 20 min) generated using Network A.
-\textbf{A}: Simulation error versus parameter set angle for N = 100 particles biased toward the correct regulatory program (A,B,C,D,E) = (40\%, 30\%, 10\%,10\% and 10\%).
+\textbf{A}: Simulation error versus parameter set angle for N = 100 particles biased toward the correct regulatory program (A,B,C,D,E) = (40\%, 10\%, 20\%, 20\% and 10\%).
 \textbf{B}: Simulation error versus parameter set angle for N = 100 uniformly distributed particles (A,B,C,D,E) = (20\%, 20\%, 20\%, 20\% and 20\%).
-\textbf{C}: Simulation error versus parameter set angle for N = 100 negatively biased particles (A,B,C,D,E) = (10\%, 10\%, 10\%, 30\% and 40\%).
+\textbf{C}: Simulation error versus parameter set angle for N = 100 negatively biased particles (A,B,C,D,E) = (10\%, 40\%, 10\%, 20\% and 20\%).
 Network A (the correct structure) was preferentially identified for positively and uniform biased particle distributions, but misidentified in the presence of

### 2014-09-01 Paper_v1.tex
still working on the introduction

@@
 steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, as well as the cellular response to metabolic engineering modifications \citep{1999_varner_ramkrishna_MetaEng}. 
+Unfortunately, cybernetic models also suffer from an identifiability challenge, as both the kinetic parameters and an abstracted model of cellular objectives must be estimated simultaneously.
 
@@
 have become standard tools \citep{2012_lewis_palsson_NatRevMicrobio}.
-The first genome-scale stoichiometric model of \textit{E. coli}, developed by Edwards and Palsson \citep{2000_edwards_palsson_PNAS}, 
-was quickly followed by many other organism models \citep{2009_feist_palsson_NatRevMicrobio}. 
+Since the first genome-scale stoichiometric model of \textit{E. coli}, developed by Edwards and Palsson \citep{2000_edwards_palsson_PNAS}, 
+well over 100 organisms, including industrially important prokaryotes such as \emph{E.~coli} \citep{Feist:2007aa} or \emph{B.~subtilis} \citep{Oh:2007aa}, 
+are now available \citep{2009_feist_palsson_NatRevMicrobio}.
 Stoichiometric models rely on a pseudo-steady-state assumption to reduce unidentifiable genome-scale kinetic models to an underdetermined linear algebraic system, 
@@
 Thus, modern constraints based approaches have proven extremely useful in the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \citep{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}.
+However, genome scale kinetic models of industrial important organisms such as \textit{E~coli} have yet to
+be constructed.
 
-Cell free systems offer unique advantages to study and model microbial metabolism when compared to traditional whole cell processes.
-[FINISH ME].
+Cell free systems offer many advantages for the study and modeling of metabolism when compared to traditional \textit{in~vivo} processes.
+Central amongst these advantages, is direct access to the microbial biosynthetic machinery without the interference of a cell wall. 
+This allows us to control as well as interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
+Second, it also allows us to study biological processes without the complications of cell growth that would normally occur with \textit{in~vivo} processes.
+Cell free protein synthesis (CFPS) systems are arguably the most prominent examples of cell free biology in use today \citep{Jewett:2008aa}.
+However, CFPS is not a new idea; CFPS in crude \emph{E.~coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \citep{MATTHAEI:1961aa,NIRENBERG:1961aa}. 
+Today, cell free systems are used in a variety of applications ranging therapeutic protein production \citep{Lu:2014aa} to synthetic biology \citep{Hodgman:2012aa}. 
+Many of the challenges to whole-genome kinetic modeling are overcome in a cell-free system. 
+For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth. 
+In this case, whole cell dynamic metabolic models may be possible for important organisms such as \emph{E.~coli.} 
+Thus, cell-free operation holds several significant advantages for model development, identification and validation.
 

### 2014-09-02 Paper_v1.tex
Need to finish discussion and polish

@@
 \section*{Abstract}
+Cell free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in~vivo} processes.
+Many of the challenges confronting genome scale kinetic modeling can potentially be overcome in a cell free system. 
+For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth. 
+Thus, cell free operation holds several significant advantages for model development, identification and validation.
+Theoretically, genome scale cell free kinetic models may be possible for industrially important organisms such as \emph{E.~coli}, 
+if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated. 
+Toward this unmet need, we present an effective biochemical network modeling framework for building dynamic cell free metabolic models.
+The key innovation of our approach is the integration of simple effective rules encoding complex allosteric regulation, with traditional kinetic pathway modeling.
+We tested our approach by modeling the time evolution of several hypothetical cell free metabolic networks.
+We found that simple effective rules, when integrated with traditional enzyme kinetic expressions,
+captured complex allosteric patterns such as ultrasensitivity, or non-competitive inhibition in the absence of mechanistic information. 
+Second, when integrated into network models, these rules captured classic regulatory patterns such as product induced feedback inhibition. 
+Lastly, we showed, at least for the network architectures considered here, that we could 
+simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data.
+While only an initial proof-of-concept, the framework presented here could be an important first step toward genome scale cell free kinetic modeling of the biosynthetic capacity 
+of industrially important organisms.
 
@@
 
-Cell free systems offer many advantages for the study and modeling of metabolism when compared to traditional \textit{in~vivo} processes.
-Central amongst these advantages, is direct access to the microbial biosynthetic machinery without the interference of a cell wall. 
+Cell free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in~vivo} processes.
+Central amongst these advantages, is direct access to metabolites and the microbial biosynthetic machinery without the interference of a cell wall. 
 This allows us to control as well as interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
-Second, it also allows us to study biological processes without the complications of cell growth that would normally occur with \textit{in~vivo} processes.
-Cell free protein synthesis (CFPS) systems are arguably the most prominent examples of cell free biology in use today \citep{Jewett:2008aa}.
-However, CFPS is not a new idea; CFPS in crude \emph{E.~coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \citep{MATTHAEI:1961aa,NIRENBERG:1961aa}. 
+Second, cell free systems also allow us to study biological processes without the complications associated with cell growth.
+Cell free protein synthesis (CFPS) systems are arguably the most prominent examples of cell free systems used today \citep{Jewett:2008aa}.
+However, CFPS is not new; CFPS in crude \emph{E.~coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \citep{MATTHAEI:1961aa,NIRENBERG:1961aa}. 
 Today, cell free systems are used in a variety of applications ranging therapeutic protein production \citep{Lu:2014aa} to synthetic biology \citep{Hodgman:2012aa}. 
-Many of the challenges to whole-genome kinetic modeling are overcome in a cell-free system. 
+Interestingly, many of the challenges confronting genome scale kinetic modeling can potentially be overcome in a cell free system. 
 For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth. 
-In this case, whole cell dynamic metabolic models may be possible for important organisms such as \emph{E.~coli.} 
-Thus, cell-free operation holds several significant advantages for model development, identification and validation.
+Thus, cell free operation holds several significant advantages for model development, identification and validation.
+Theoretically, genome scale cell free kinetic models may be possible for industrially important organisms such as \emph{E.~coli}, 
+if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.
 
-In this study, we present an effective biochemical network modeling framework for building dynamic metabolic models.
+In this study, we present an effective biochemical network modeling framework for building dynamic cell free metabolic models.
 The key innovation of our approach is the seamless integration of simple effective rules encoding complex regulation, with traditional kinetic pathway modeling.
 This integration allows the description of complex regulatory interactions, such as time-dependent allosteric regulation of enzyme activity, in the absence of specific mechanistic information.
-Moreover, the rules are easy to understand, easy to formulate 
+The regulatory rules are easy to understand, easy to formulate 
 and do not rely on overarching theoretical abstractions, or restrictive assumptions.
-We tested our approach by modeling the time evolution of hypothetical cell free metabolic networks. 
-In particular, we tested whether our effective modeling approach could describe classically expected behavior, 
+We tested our approach by modeling the time evolution of several hypothetical cell free metabolic networks. 
+In particular, we tested whether our effective modeling approach could describe classically expected enzyme kinetic behavior, 
 and second whether we could simultaneously estimate kinetic parameters and regulatory connectivity, in the absence of specific mechanistic knowledge, 
@@
 We found that simple effective rules, when integrated with traditional enzyme kinetic expressions,
-captured complex allosteric patterns such as ultrasensitivity, or non-competitive inhibition in the absence of specific mechanistic information. 
-Moreover, when integrated into network models, these rules captured classical regulatory patterns such as product induced feedback inhibition. 
-Lastly, we simultaneously estimated kinetic parameters and discriminated between competing regulatory structures, using synthetic data in combination with a
-modified particle swarm approach. While only a proof-of-concept, [FINISH ME].
+captured complex allosteric patterns such as ultrasensitivity, or non-competitive inhibition in the absence of mechanistic information. 
+Second, when integrated into network models, these rules captured classical regulatory patterns such as product induced feedback inhibition. 
+Lastly, we showed, at least for the network architectures considered here, that we could 
+simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data.
+While only an initial proof-of-concept, the framework presented here could be an important first step toward genome scale cell free kinetic modeling of the biosynthetic capacity 
+of industrially important organisms such as \emph{E.~coli} or \emph{B.~subtilis}.
 
@@
 
-\subsection*{Formulation and properties of cell free effective models.}
+\subsection*{Formulation and properties of effective cell free metabolic models.}
 We developed two proof-of-concept metabolic networks to investigate the features of our effective biochemical network modeling approach (Fig. \ref{fig-networks}).
 In both examples, substrate S was converted to the end-products P$_{1}$ and P$_{2}$ through a series of enzymatically catalyzed reactions, 
-including a branch point at hypothetical metabolite $M_{2}$. 
-Several of these reactions involved cofactor dependence ($AH$ or $A$), and various allosteric regulation mechanisms. 
+including a branch point at hypothetical metabolite M$_{2}$. 
+Several of these reactions involved cofactor dependence (AH or A), and various allosteric regulatory mechanisms modified the activity of pathway enzymes. 
 Network A included feedback inhibition of the initial pathway enzyme ($E_{1}$) by pathway end products P$_{1}$ and P$_{2}$ (Fig. \ref{fig-networks}A).
 On the other hand, network B involved feedback inhibition of $E_{1}$ by P$_{2}$ and $E_{6}$ by P$_{1}$ (Fig. \ref{fig-networks}B). 
-In both networks, branch point enzymes $E_{3}$ and $E_{6}$ were subject to feed-forward activation by cofactor $AH$.
-Lastly, enzyme activity was assumed to decay according to a first-order rate law in both cases. 
-Allosteric regulation of enzyme activity was represented using a novel rule-based strategy, similar in spirit to the 
-Constrained Fuzzy Logic (cFL) approach of Lauffenberger and coworkers \citep{Morris:2011aa}. 
-In this formulation, Hill-like transfer functions were used to calculate the influence of metabolite abundance upon target enzyme activity. 
-When an enzyme was potentially sensitive to more than one regulatory influence, logical rules were used to select which transfer function regulated enzyme 
-activity at any given time (Fig. \ref{fig-control-schematic}).      
-Thus, our test networks involved important features such as cofactor recycling, enzyme activity and metabolite dynamics, 
+In both networks, branch point enzymes $E_{3}$ and $E_{6}$ were subject to feed-forward activation by reduced cofactor AH.
+Lastly, it is known experimentally that cell free systems have a finite 
+operational lifespan. Loss of biosynthetic capability could be a function of many factors e.g., cofactor or metabolite 
+limitations. However, we modeled the loss of biosynthetic capability as a non-specific first-order decay of enzyme activity.
+
+Allosteric regulation of enzyme activity was modeled by combining individual regulatory contributions to
+the activity of pathway enzymes into a control coefficient using an integration rule (Fig. \ref{fig-control-schematic}).  
+This strategy is similar in spirit to the Constrained Fuzzy Logic (cFL) approach of Lauffenburger 
+and coworkers which has been used to effectively model signal transduction pathways important in human health \citep{Morris:2011aa}.
+In our formulation, Hill-like transfer functions $0 \leq f\left(\mathcal{Z}\right) \leq 1$ were used to calculate the influence of factor abundance upon target enzyme activity. 
+In this context, factors can be individual metabolite levels, or some function e.g., the product of metabolite levels. However, more generally factors can also 
+correspond to non-modeled influences, categorial variables or other abstract quantities. 
+In the current study, we simply let $\mathcal{Z}$ correspond to the abundance of individual metabolites. 
+When an enzyme was potentially sensitive to more than one regulatory input, logical integration rules were used to select which regulatory transfer function influenced enzyme 
+activity at any given time. Thus, our test networks involved important features such as cofactor recycling, enzyme activity and metabolite dynamics, 
 as well as multiple overlapping allosteric regulatory mechanisms.
-As such, developing our effective modeling approach using these simple problems gave us valuable insight into the development of larger network models, 
-without the complication of network size.
 
 The rule based regulatory strategy approximated the behavior of classical allosteric activation and inhibition mechanisms (Fig. \ref{fig-kinetics-simulations}).  
+We considered the enzyme catalyzed conversion of substrate S to a product P, where the overall reaction rate was modeled as the product of a Michaelis-Menten term and an effective 
+allosteric control variable reflecting the particular regulatory interaction.
 We first explored feed-forward substrate activation of enzyme activity (for both positive and negative cooperativity). 
@@
 For cooperativity parameters less than unity, increased substrate abundance \textit{decreased} the reaction rate. This was consistent with the idea that
-substrate binding \textit{decreases} at regulatory sites negatively impacts the ability of the enzyme to bind substrate at the active site. 
-On the other hand, as the cooperativity parameter increased past unity, the rate of conversion of substrate $S$ to product $P$ by enzyme $E$ approached a step function.
+substrate binding \textit{decreased} at regulatory sites, which negatively impacted substrate binding at the active site. 
+On the other hand, as the cooperativity parameter increased past unity, the rate of conversion of substrate S to product P by enzyme $E$ approached a step function.
 In the presence of an inhibitor, the rule based strategy predicted non-competitive like behavior as a function of the 
@@
 directly proportional to the cooperativity parameter, $\eta$ in Eqn. \eqref{eqn:control-factor}. 
-Thus, as the cooperativity parameter increased, the maximum reaction rate decreased (Fig. \ref{fig-kinetics-simulations}B, orange).
-However, when the gain parameter was less than unity, enzyme inhibition increased with \textit{decreasing} cooperativity, 
-i.e., smaller $\eta$ yielded increased inhibition (Fig. \ref{fig-kinetics-simulations}B).
+Thus, as the cooperativity parameter increased, the maximum reaction rate decreased (Fig. \ref{fig-kinetics-simulations}B).
 Interestingly, our rule based approach was unable to directly simulate competitive inhibition of enzyme activity. 
-For competitive inhibitors, the kinetic component of our rate, $\bar{r}_{j}$ in \eqref{eqn:rate-factor}, could be modified to account 
-for the inhibition (data not shown). 
 Taken together, the rule based strategy captured classical regulatory patterns for both enzyme activation and inhibition. 
@@
 
-End product yield was controlled by feedback inhibition, while selectivity was controlled by branch point enzyme inhibition (Fig. \ref{fig-onoff-simulations}).
+End product yield was controlled by feedback inhibition, while product selectivity was controlled by branch point enzyme inhibition (Fig. \ref{fig-onoff-simulations}).
 A critical test of our modeling approach was to simulate networks with known behavior. If we cannot reproduce the expected behavior of simple networks, then our
-effect modeling strategy, and particularly the rule-based approximation of allosteric regulation, will not be feasible for large scale problems.
-We considered two cases,  control on/off, for each network configuration. 
+effective modeling strategy, and particularly the rule-based approximation of allosteric regulation, will not be feasible for genome scale cell free problems.
+We considered two cases,  control ON/OFF, for each network configuration. 
 Each of these cases had identical kinetic parameters and initial conditions; 
 the \textit{only} differences between the cases was the allosteric regulation rules, and the control parameters associated with these rules. 
-As expected, end product accumulation was larger for network A when the control was off (no feedback inhibition of $E_{1}$ by P$_{1}$ and P$_{2}$),
-as compared to the on case (Fig. \ref{fig-onoff-simulations}A). We found this behavior was robust to the choice of underlying kinetic parameters, 
-as we observed that same qualitative response across an ensemble of randomized parameter sets (N = 100). 
-The control on/off response of network B was more subtle. In the off case, the behavior was qualitatively similar to network A. 
-However, for the on case, flux was diverted away from P$_{2}$ formation by feedback inhibition of $E_{6}$ activity at the M$_{2}$ branch point by P$_{1}$ (Fig. \ref{fig-onoff-simulations}B).
+As expected, end product accumulation was larger for network A when the control was OFF (no feedback inhibition of $E_{1}$ by P$_{1}$ and P$_{2}$),
+as compared to the ON case (Fig. \ref{fig-onoff-simulations}A). We found this behavior was robust to the choice of underlying kinetic parameters, 
+as we observed that same qualitative response across an ensemble of randomized parameter sets (N = 100), for fixed control parameters.
+The control ON/OFF response of network B was more subtle. In the OFF case, the behavior was qualitatively similar to network A. 
+However, for the ON case, flux was diverted away from P$_{2}$ formation by feedback inhibition of $E_{6}$ activity at the M$_{2}$ branch point by P$_{1}$ (Fig. \ref{fig-onoff-simulations}B).
 Lower $E_{6}$ activity at the M$_{2}$ branch point allowed more flux toward P$_{1}$ formation, hence the yield of P$_{1}$ also increased (Fig. \ref{fig-onoff-simulations}C).
-Again, the control on/off behavior was robust to changes kinetic parameters, as the same qualitative trend was conserved across 
-an ensemble of randomized parameters (N = 100). Taken together, these simulations suggested that the rule based allosteric control 
-concept could robustly capture expected feedback behavior.
+Again, the control ON/OFF behavior of network B was robust to changes kinetic parameters, as the same qualitative trend was conserved across 
+an ensemble of randomized parameters (N = 100), for fixed control parameters. 
+Taken together, these simulations suggested that the rule based allosteric control concept could robustly capture expected feedback behavior for networks with uncertain
+kinetic parameters.
 
@@
 We explored this linkage by estimating model parameters from synthetic data using both network
-structures. We generated noise-corrupted synthetic measurements of the substrate $S$, intermediate M$_{5}$ and end-product P$_1$ approximately every 20 min using
+structures. We generated noise corrupted synthetic measurements of the substrate S, intermediate M$_{5}$ and end-product P$_1$ approximately every 20 min using
 network A. We then generated an ensemble of model parameter estimates by minimizing the difference between model simulations and the synthetic data using particle swarm
@@
 when the control structure was correct (Fig. \ref{fig-parameter-fit}A and B).
-However, with control mismatch (network B simulated with network A parameters), the model simulations were not consistent with the synthetic data  (Fig. \ref{fig-parameter-fit}C and D). 
+However, with control mismatch (network B simulated with network A parameters), model simulations were not consistent with the synthetic data  (Fig. \ref{fig-parameter-fit}C and D). 
 Taken together, these results suggested that we could perhaps simultaneously estimate both parameters and network control architectures, as incorrect control structures 
@@
 \section*{Discussion}
-In this study, we presented an effective kinetic modeling strategy to dynamically simulate biochemical networks. 
+In this study, we presented an effective kinetic modeling strategy to dynamically simulate cell free biochemical networks. 
 Our proposed strategy integrated traditional kinetic modeling with an effective rules based approach to dynamically describe metabolic regulation and control.
@@
 
+[CONTRAST WITH CYBERNETIC MODELS].
+
 While the results of this study were encouraging, there are several critical next steps that must be accomplished before we can model genome scale cell free metabolic networks.
+[FINISH ME]
 

### 2014-09-03 Paper_v1.tex
final edits - Joe and Adi

@@
 \vspace{0.1in}
-{\par \centering \textbf{To be submitted:}~\emph{Processes}}
+{\par \centering \textbf{To be submitted:}~\textit{Processes}}
 \vspace{0.5in}
@@
 \section*{Abstract}
-Cell free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in~vivo} processes.
-Many of the challenges confronting genome scale kinetic modeling can potentially be overcome in a cell free system. 
+Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
+Many of the challenges confronting genome-scale kinetic modeling can potentially be overcome in a cell-free system. 
 For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth. 
-Thus, cell free operation holds several significant advantages for model development, identification and validation.
-Theoretically, genome scale cell free kinetic models may be possible for industrially important organisms such as \emph{E.~coli}, 
-if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated. 
-Toward this unmet need, we present an effective biochemical network modeling framework for building dynamic cell free metabolic models.
-The key innovation of our approach is the integration of simple effective rules encoding complex allosteric regulation, with traditional kinetic pathway modeling.
-We tested our approach by modeling the time evolution of several hypothetical cell free metabolic networks.
-We found that simple effective rules, when integrated with traditional enzyme kinetic expressions,
-captured complex allosteric patterns such as ultrasensitivity, or non-competitive inhibition in the absence of mechanistic information. 
-Second, when integrated into network models, these rules captured classic regulatory patterns such as product induced feedback inhibition. 
-Lastly, we showed, at least for the network architectures considered here, that we could 
-simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data.
-While only an initial proof-of-concept, the framework presented here could be an important first step toward genome scale cell free kinetic modeling of the biosynthetic capacity 
-of industrially important organisms.
+Thus, cell-free operation holds several significant advantages for model development, identification and validation. 
+Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms, such as \textit{E.
+coli}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.
+Toward this unmet need, we present an effective biochemical network modeling framework for building dynamic cell-free metabolic models.
+The key innovation of our approach is the integration of simple effective rules encoding complex allosteric regulation with traditional kinetic pathway modeling. 
+We tested our approach by modeling the time evolution of several hypothetical cell-free metabolic networks. 
+We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, captured complex allosteric patterns such as ultrasensitivity or non-competitive inhibition in the absence of mechanistic information.
+Second, when integrated into network models, these rules captured classic regulatory patterns such as product-induced feedback inhibition. 
+Lastly, we showed, at least for the network architectures considered here, that we could simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data. 
+While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.
 
-{\noindent \textbf{Keywords:}~Cell free metabolism, Mathematical modeling}
+{\noindent \textbf{Keywords:}~Cell-free metabolism, Mathematical modeling}
 
@@
 Mathematical modeling has long contributed to our understanding of metabolism. 
-Decades before the genomics revolution, mechanistically, structured metabolic models arose from the desire to predict microbial phenotypes resulting from 
-changes in intracellular or extracellular states \citep{1976_fredrickson_BiotechBioeng}. 
-The single cell \textit{E. coli} models of Shuler and coworkers pioneered the construction of large-scale, dynamic metabolic models that incorporated multiple, regulated catabolic and anabolic 
-pathways constrained by experimentally determined kinetic parameters \citep{1984_domach_shuler_BiotechBioeng_01}. 
+Decades before the genomics revolution, mechanistically, structured metabolic models arose from the desire to predict microbial phenotypes resulting from changes in intracellular or extracellular states \citep{1976_fredrickson_BiotechBioeng}.
+The single cell \textit{E.
+coli} models of Shuler and coworkers pioneered the construction of large-scale, dynamic metabolic models that incorporated multiple, regulated catabolic and anabolic pathways constrained by experimentally determined kinetic parameters \citep{1984_domach_shuler_BiotechBioeng_01}.
 Shuler and coworkers generated many single cell kinetic models, including single cell models of eukaryotes \citep{1989_steinmeyer_shuler_ChemEngSci,1992_wu_shuler_AnnNYAcadSci}, minimal cell architectures \citep{2004_castellanos_shuler_PNAS}, as well as DNA sequence based whole-cell models of \textit{E.
@@
 Conversely, highly abstracted kinetic frameworks, such as the cybernetic framework, represented a paradigm shift, viewing cells as growth-optimizing strategists \citep{1985_dhurjati_ramkrishna_tsao_BiotechBioeng}.
-Cybernetic models have been highly successful at predicting metabolic choice behavior, e.g., diauxie behavior \citep{1986_kompala_ramkrishna_tsao_BiotechBioeng}, 
-steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, as well as the cellular response to metabolic engineering modifications \citep{1999_varner_ramkrishna_MetaEng}. 
+Cybernetic models have been highly successful at predicting metabolic choice behavior, e.g., diauxie behavior \citep{1986_kompala_ramkrishna_tsao_BiotechBioeng}, steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, as well as the cellular response to metabolic engineering modifications \citep{1999_varner_ramkrishna_MetaEng}.
 Unfortunately, cybernetic models also suffer from an identifiability challenge, as both the kinetic parameters and an abstracted model of cellular objectives must be estimated simultaneously.
 
-In the post genomics world, large-scale stoichiometric reconstructions of microbial metabolism popularized by static, constraint-based modeling techniques such as flux balance analysis (FBA)
-have become standard tools \citep{2012_lewis_palsson_NatRevMicrobio}.
-Since the first genome-scale stoichiometric model of \textit{E. coli}, developed by Edwards and Palsson \citep{2000_edwards_palsson_PNAS}, 
-well over 100 organisms, including industrially important prokaryotes such as \emph{E.~coli} \citep{Feist:2007aa} or \emph{B.~subtilis} \citep{Oh:2007aa}, 
-are now available \citep{2009_feist_palsson_NatRevMicrobio}.
-Stoichiometric models rely on a pseudo-steady-state assumption to reduce unidentifiable genome-scale kinetic models to an underdetermined linear algebraic system, 
-which can be solved efficiently even for large systems.  
-Traditionally, stoichiometric models have also neglected explicit descriptions of metabolic regulation and control mechanisms, 
-instead opting to describe the choice of pathways by prescribing an objective function on metabolism. 
-Interestingly, similar to early cybernetic models, the most common metabolic objective function has been the optimization of biomass formation \citep{2002_ibarra_edwards_palsson_Nat}, 
-although other metabolic objectives have also been estimated \citep{2007_schuetz_sauer_MolSysBio}. 
-Recent advances in constraints based modeling have overcome the early shortcomings of the platform, including capturing metabolic regulation and control \citep{2013_hyduke_lewis_palsson_MolBioSys}.
-Thus, modern constraints based approaches have proven extremely useful in the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \citep{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}.
-However, genome scale kinetic models of industrial important organisms such as \textit{E~coli} have yet to
-be constructed.
+In the post genomics world, large-scale stoichiometric reconstructions of microbial metabolism popularized by static, constraint-based modeling techniques such as flux balance analysis (FBA) have become standard tools \citep{2012_lewis_palsson_NatRevMicrobio}.
+Since the first genome-scale stoichiometric model of \textit{E.
+coli}, developed by Edwards and Palsson \citep{2000_edwards_palsson_PNAS}, well over 100 organisms, including industrially important prokaryotes such as \textit{E.
+coli} \citep{Feist:2007aa} or \textit{B.
+subtilis} \citep{Oh:2007aa}, are now available \citep{2009_feist_palsson_NatRevMicrobio}.
+Stoichiometric models rely on a pseudo-steady-state assumption to reduce unidentifiable genome-scale kinetic models to an underdetermined linear algebraic system, which can be solved efficiently even for large systems.
+Traditionally, stoichiometric models have also neglected explicit descriptions of metabolic regulation and control mechanisms, instead opting to describe the choice of pathways by prescribing an objective function on metabolism.
+Interestingly, similar to early cybernetic models, the most common metabolic objective function has been the optimization of biomass formation \citep{2002_ibarra_edwards_palsson_Nat}, although other metabolic objectives have also been estimated \citep{2007_schuetz_sauer_MolSysBio}.
+Recent advances in constraint-based modeling have overcome the early shortcomings of the platform, including capturing metabolic regulation and control \citep{2013_hyduke_lewis_palsson_MolBioSys}. 
+Thus, modern constraint-based approaches have proven extremely useful in the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \citep{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}.
+However, genome-scale kinetic models of industrial important organisms such as \textit{E. coli} have yet to be constructed.
 
-Cell free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in~vivo} processes.
-Central amongst these advantages, is direct access to metabolites and the microbial biosynthetic machinery without the interference of a cell wall. 
+Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
+Central amongst these advantages is direct access to metabolites and the microbial biosynthetic machinery without the interference of a cell wall. 
 This allows us to control as well as interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
-Second, cell free systems also allow us to study biological processes without the complications associated with cell growth.
-Cell free protein synthesis (CFPS) systems are arguably the most prominent examples of cell free systems used today \citep{Jewett:2008aa}.
-However, CFPS is not new; CFPS in crude \emph{E.~coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \citep{MATTHAEI:1961aa,NIRENBERG:1961aa}. 
-Today, cell free systems are used in a variety of applications ranging therapeutic protein production \citep{Lu:2014aa} to synthetic biology \citep{Hodgman:2012aa}. 
-Interestingly, many of the challenges confronting genome scale kinetic modeling can potentially be overcome in a cell free system. 
+Second, cell-free systems also allow us to study biological processes without the complications associated with cell growth.
+Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \citep{Jewett:2008aa}. 
+However, CFPS is not new; CFPS in crude \textit{E. coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \citep{MATTHAEI:1961aa,NIRENBERG:1961aa}. 
+Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \citep{Lu:2014aa} to synthetic biology \citep{Hodgman:2012aa}. 
+Interestingly, many of the challenges confronting genome-scale kinetic modeling can potentially be overcome in a cell-free system. 
 For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth. 
-Thus, cell free operation holds several significant advantages for model development, identification and validation.
-Theoretically, genome scale cell free kinetic models may be possible for industrially important organisms such as \emph{E.~coli}, 
-if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.
+Thus, cell-free operation holds several significant advantages for model development, identification and validation.
+Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms, such as \textit{E.
+coli} or \textit{B.
+subtilis}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.
 
-In this study, we present an effective biochemical network modeling framework for building dynamic cell free metabolic models.
-The key innovation of our approach is the seamless integration of simple effective rules encoding complex regulation, with traditional kinetic pathway modeling.
+In this study, we present an effective biochemical network modeling framework for building dynamic cell-free metabolic models.
+The key innovation of our approach is the seamless integration of simple effective rules encoding complex regulation with traditional kinetic pathway modeling. 
 This integration allows the description of complex regulatory interactions, such as time-dependent allosteric regulation of enzyme activity, in the absence of specific mechanistic information. 
-The regulatory rules are easy to understand, easy to formulate 
-and do not rely on overarching theoretical abstractions, or restrictive assumptions.
-We tested our approach by modeling the time evolution of several hypothetical cell free metabolic networks. 
-In particular, we tested whether our effective modeling approach could describe classically expected enzyme kinetic behavior, 
-and second whether we could simultaneously estimate kinetic parameters and regulatory connectivity, in the absence of specific mechanistic knowledge, 
-from synthetic experimental data. Toward these questions, we explored five hypothetical cell-free networks. 
+The regulatory rules are easy to understand, easy to formulate and do not rely on overarching theoretical abstractions or restrictive assumptions. 
+We tested our approach by modeling the time evolution of several hypothetical cell-free metabolic networks. 
+In particular, we tested whether our effective modeling approach could describe classically expected enzyme kinetic behavior, and second whether we could simultaneously estimate kinetic parameters and regulatory connectivity, in the absence of specific mechanistic knowledge, from synthetic experimental data.
+Toward these questions, we explored five hypothetical cell-free networks. 
 Each network shared the same enzymatic connectivity, but had different allosteric regulatory connectivity. 
-We found that simple effective rules, when integrated with traditional enzyme kinetic expressions,
-captured complex allosteric patterns such as ultrasensitivity, or non-competitive inhibition in the absence of mechanistic information. 
-Second, when integrated into network models, these rules captured classical regulatory patterns such as product induced feedback inhibition. 
-Lastly, we showed, at least for the network architectures considered here, that we could 
-simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data.
-While only an initial proof-of-concept, the framework presented here could be an important first step toward genome scale cell free kinetic modeling of the biosynthetic capacity 
-of industrially important organisms such as \emph{E.~coli} or \emph{B.~subtilis}.
+We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, captured complex allosteric patterns such as ultrasensitivity or non-competitive inhibition in the absence of mechanistic information.
+Second, when integrated into network models, these rules captured classical regulatory patterns such as product-induced feedback inhibition. 
+Lastly, we showed, at least for the network architectures considered here, that we could simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data. 
+While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.
 
@@
 
-\subsection*{Formulation and properties of effective cell free metabolic models.}
+\subsection*{Formulation and properties of effective cell-free metabolic models.}
 We developed two proof-of-concept metabolic networks to investigate the features of our effective biochemical network modeling approach (Fig. \ref{fig-networks}). 
-In both examples, substrate S was converted to the end-products P$_{1}$ and P$_{2}$ through a series of enzymatically catalyzed reactions, 
-including a branch point at hypothetical metabolite M$_{2}$. 
+In both examples, substrate S was converted to the end products P$_{1}$ and P$_{2}$ through a series of enzymatically catalyzed reactions, including a branch point at hypothetical metabolite M$_{2}$. 
 Several of these reactions involved cofactor dependence (AH or A), and various allosteric regulatory mechanisms modified the activity of pathway enzymes. 
@@
 In both networks, branch point enzymes $E_{3}$ and $E_{6}$ were subject to feed-forward activation by reduced cofactor AH. 
-Lastly, it is known experimentally that cell free systems have a finite 
-operational lifespan. Loss of biosynthetic capability could be a function of many factors e.g., cofactor or metabolite 
-limitations. However, we modeled the loss of biosynthetic capability as a non-specific first-order decay of enzyme activity.
+Lastly, it is known experimentally that cell-free systems have a finite operational lifespan. 
+Loss of biosynthetic capability could be a function of many factors, e.g., cofactor or metabolite limitations. 
+We modeled the loss of biosynthetic capability as a non-specific first-order decay of enzyme activity.
 
-Allosteric regulation of enzyme activity was modeled by combining individual regulatory contributions to
-the activity of pathway enzymes into a control coefficient using an integration rule (Fig. \ref{fig-control-schematic}).  
-This strategy is similar in spirit to the Constrained Fuzzy Logic (cFL) approach of Lauffenburger 
-and coworkers which has been used to effectively model signal transduction pathways important in human health \citep{Morris:2011aa}.
+Allosteric regulation of enzyme activity was modeled by combining individual regulatory contributions to the activity of pathway enzymes into a control coefficient using an integration rule (Fig.
+\ref{fig-control-schematic}).
+This strategy is similar in spirit to the Constrained Fuzzy Logic (cFL) approach of Lauffenburger and coworkers which has been used to effectively model signal transduction pathways important in human health \citep{Morris:2011aa}.
 In our formulation, Hill-like transfer functions $0 \leq f\left(\mathcal{Z}\right) \leq 1$ were used to calculate the influence of factor abundance upon target enzyme activity. 
-In this context, factors can be individual metabolite levels, or some function e.g., the product of metabolite levels. However, more generally factors can also 
-correspond to non-modeled influences, categorial variables or other abstract quantities. 
+In this context, factors can be individual metabolite levels or some function, e.g., the product of metabolite levels. 
+However, more generally, factors can also correspond to non-modeled influences, categorial variables or other abstract quantities. 
 In the current study, we simply let $\mathcal{Z}$ correspond to the abundance of individual metabolites. 
-When an enzyme was potentially sensitive to more than one regulatory input, logical integration rules were used to select which regulatory transfer function influenced enzyme 
-activity at any given time. Thus, our test networks involved important features such as cofactor recycling, enzyme activity and metabolite dynamics, 
-as well as multiple overlapping allosteric regulatory mechanisms.
+When an enzyme was potentially sensitive to more than one regulatory input, logical integration rules were used to select which regulatory transfer function influenced enzyme activity at any given time.
+Thus, our test networks involved important features such as cofactor recycling, enzyme activity and metabolite dynamics, as well as multiple overlapping allosteric regulatory mechanisms.
 
-The rule based regulatory strategy approximated the behavior of classical allosteric activation and inhibition mechanisms (Fig. \ref{fig-kinetics-simulations}).  
-We considered the enzyme catalyzed conversion of substrate S to a product P, where the overall reaction rate was modeled as the product of a Michaelis-Menten term and an effective 
-allosteric control variable reflecting the particular regulatory interaction.
+The rule-based regulatory strategy approximated the behavior of classical allosteric activation and inhibition mechanisms (Fig. \ref{fig-kinetics-simulations}). 
+We considered the enzyme catalyzed conversion of substrate S to a product P, where the overall reaction rate was modeled as the product of a Michaelis-Menten term and an effective allosteric control variable reflecting the particular regulatory interaction.
 We first explored feed-forward substrate activation of enzyme activity (for both positive and negative cooperativity). 
-Consistent with classical data, the rule based strategy predicted a sigmoidal relationship between substrate abundance 
-and reaction rate as a function of the cooperativity parameter (Fig. \ref{fig-kinetics-simulations}A). 
-For cooperativity parameters less than unity, increased substrate abundance \textit{decreased} the reaction rate. This was consistent with the idea that
-substrate binding \textit{decreased} at regulatory sites, which negatively impacted substrate binding at the active site. 
+Consistent with classical data, the rule-based strategy predicted a sigmoidal relationship between substrate abundance and reaction rate as a function of the cooperativity parameter (Fig.
+\ref{fig-kinetics-simulations}A).
+For cooperativity parameters less than unity, increased substrate abundance \textit{decreased} the reaction rate. 
+This was consistent with the idea that substrate binding \textit{decreased} at regulatory sites, which negatively impacted substrate binding at the active site. 
 On the other hand, as the cooperativity parameter increased past unity, the rate of conversion of substrate S to product P by enzyme $E$ approached a step function. 
-In the presence of an inhibitor, the rule based strategy predicted non-competitive like behavior as a function of the 
-cooperativity parameter (Fig. \ref{fig-kinetics-simulations}B). 
-When the control gain parameter, $\kappa_{ij}$ in Eqn. \eqref{eqn:control-factor}, was greater than unity, the inhibitory force was
-directly proportional to the cooperativity parameter, $\eta$ in Eqn. \eqref{eqn:control-factor}. 
+In the presence of an inhibitor, the rule-based strategy predicted non-competitive like behavior as a function of the cooperativity parameter (Fig. \ref{fig-kinetics-simulations}B). 
+When the control gain parameter, $\kappa_{ij}$ in Eqn.
+\eqref{eqn:control-factor}, was greater than unity, the inhibitory force was directly proportional to the cooperativity parameter, $\eta$ in Eqn.
+\eqref{eqn:control-factor}.
 Thus, as the cooperativity parameter increased, the maximum reaction rate decreased (Fig. \ref{fig-kinetics-simulations}B). 
-Interestingly, our rule based approach was unable to directly simulate competitive inhibition of enzyme activity. 
-Taken together, the rule based strategy captured classical regulatory patterns for both enzyme activation and inhibition. 
+Interestingly, our rule-based approach was unable to directly simulate competitive inhibition of enzyme activity. 
+Taken together, the rule-based strategy captured classical regulatory patterns for both enzyme activation and inhibition. 
 Thus, we are able to model complex kinetic phenomena such as ultrasensitivity, despite an effective description of reaction kinetics.
@@
 End product yield was controlled by feedback inhibition, while product selectivity was controlled by branch point enzyme inhibition (Fig. \ref{fig-onoff-simulations}). 
-A critical test of our modeling approach was to simulate networks with known behavior. If we cannot reproduce the expected behavior of simple networks, then our
-effective modeling strategy, and particularly the rule-based approximation of allosteric regulation, will not be feasible for genome scale cell free problems.
+A critical test of our modeling approach was to simulate networks with known behavior. 
+If we cannot reproduce the expected behavior of simple networks, then our effective modeling strategy, and particularly the rule-based approximation of allosteric regulation, will not be feasible for genome-scale cell-free problems.
 We considered two cases, control ON/OFF, for each network configuration. 
-Each of these cases had identical kinetic parameters and initial conditions; 
-the \textit{only} differences between the cases was the allosteric regulation rules, and the control parameters associated with these rules. 
+Each of these cases had identical kinetic parameters and initial conditions; the \textit{only} differences between the cases were the allosteric regulation rules and the control parameters associated with these rules.
 As expected, end product accumulation was larger for network A when the control was OFF (no feedback inhibition of $E_{1}$ by P$_{1}$ and P$_{2}$),
-as compared to the ON case (Fig. \ref{fig-onoff-simulations}A). We found this behavior was robust to the choice of underlying kinetic parameters, 
-as we observed that same qualitative response across an ensemble of randomized parameter sets (N = 100), for fixed control parameters.
-The control ON/OFF response of network B was more subtle. In the OFF case, the behavior was qualitatively similar to network A. 
+as compared to the ON case (Fig. \ref{fig-onoff-simulations}A). 
+We found this behavior was robust to the choice of underlying kinetic parameters, as we observed that same qualitative response across an ensemble of randomized parameter sets (N = 100), for fixed control parameters.
+The control ON/OFF response of network B was more subtle. 
+In the OFF case, the behavior was qualitatively similar to network A. 
 However, for the ON case, flux was diverted away from P$_{2}$ formation by feedback inhibition of $E_{6}$ activity at the M$_{2}$ branch point by P$_{1}$ (Fig. \ref{fig-onoff-simulations}B). 
 Lower $E_{6}$ activity at the M$_{2}$ branch point allowed more flux toward P$_{1}$ formation, hence the yield of P$_{1}$ also increased (Fig. \ref{fig-onoff-simulations}C). 
-Again, the control ON/OFF behavior of network B was robust to changes kinetic parameters, as the same qualitative trend was conserved across 
-an ensemble of randomized parameters (N = 100), for fixed control parameters. 
-Taken together, these simulations suggested that the rule based allosteric control concept could robustly capture expected feedback behavior for networks with uncertain
-kinetic parameters.
+Again, the control ON/OFF behavior of network B was robust to changes in kinetic parameters, as the same qualitative trend was conserved across an ensemble of randomized parameters (N = 100), for fixed control parameters.
+Taken together, these simulations suggested that the rule-based allosteric control concept could robustly capture expected feedback behavior for networks with uncertain kinetic parameters.
 
@@
 For metabolic processes, there is also the added challenge of identifying the regulation and control structures that manage metabolism. 
-Of course, these issues are not independent; any description of enzyme activity regulation will be a function of system state, which
-in turn depends upon the kinetic parameters. For cell free systems, regulated gene expression has been removed, however, enzyme activity regulation is still operational. 
-We explored this linkage by estimating model parameters from synthetic data using both network
-structures. We generated noise corrupted synthetic measurements of the substrate S, intermediate M$_{5}$ and end-product P$_1$ approximately every 20 min using
-network A. We then generated an ensemble of model parameter estimates by minimizing the difference between model simulations and the synthetic data using particle swarm
-optimization (PSO), starting from random initial parameter guesses. The estimation of kinetic parameters was sensitive to the choice of regulatory structure (Fig. \ref{fig-parameter-fit}).
-PSO identified an ensemble of parameters that bracketed the mean of the synthetic measurements in less than 1000 iterations 
-when the control structure was correct (Fig. \ref{fig-parameter-fit}A and B).
+Of course, these issues are not independent; any description of enzyme activity regulation will be a function of system state, which in turn depends upon the kinetic parameters. 
+For cell-free systems, regulated gene expression has been removed, however, enzyme activity regulation is still operational. 
+We explored this linkage by estimating model parameters from synthetic data using both network structures. 
+We generated noise corrupted synthetic measurements of the substrate S, intermediate M$_{5}$ and end product P$_1$ approximately every 20 min using
+network A. 
+We then generated an ensemble of model parameter estimates by minimizing the difference between model simulations and the synthetic data using particle swarm optimization (PSO), starting from random initial parameter guesses.
+The estimation of kinetic parameters was sensitive to the choice of regulatory structure (Fig. \ref{fig-parameter-fit}). 
+PSO identified an ensemble of parameters that bracketed the mean of the synthetic measurements in less than 1000 iterations when the control structure was correct (Fig.
+\ref{fig-parameter-fit}A and B).
 However, with control mismatch (network B simulated with network A parameters), model simulations were not consistent with the synthetic data  (Fig. \ref{fig-parameter-fit}C and D). 
-Taken together, these results suggested that we could perhaps simultaneously estimate both parameters and network control architectures, as incorrect control structures 
-would be manifest as poor model fits.
+Taken together, these results suggested that we could perhaps simultaneously estimate both parameters and network control architectures, as incorrect control structures would be manifest as poor model fits.
 
 We modified our particle swarm identification strategy to simultaneously search over both kinetic parameters and putative control structures. 
-In addition to our initial networks, we constructed three additional presumptive network models, each with the same enzymatic connectivity but different allosteric 
-regulation of the pathway enzymes (Fig. \ref{fig-alternative-networks}). 
-We then initialized a population of particles, each with one of the five potential regulatory programs, and randomized kinetic parameters.
+In addition to our initial networks, we constructed three additional presumptive network models, each with the same enzymatic connectivity but different allosteric regulation of the pathway enzymes (Fig.
+\ref{fig-alternative-networks}).
+We then initialized a population of particles, each with one of the five potential regulatory programs and randomized kinetic parameters. 
 Thus, we generated an initial population of particles that had \textit{both} different kinetic parameters as well as different control structures. 
-We biased the distribution of the particle population according to our \textit{a~prior} belief of the correct regulatory program.
-To this end, we considered three different priors, a uniform distribution where each putative regulatory structure represented 20\% of the population, and
-two mixed distributions that were either positively or negatively biased towards the correct structure (network A).
-In both the positively biased, and uniform cases the particle swarm clearly differentiated between the true or closely related structures and those that were materially different (Fig.
-\ref{fig-control-search}).
+We biased the distribution of the particle population according to our \textit{a prior} belief of the correct regulatory program. 
+To this end, we considered three different priors, a uniform distribution where each putative regulatory structure represented 20\% of the population and two mixed distributions that were either positively or negatively biased towards the correct structure (network A).
+In both the positively biased and uniform cases the PSO clearly differentiated between the true or closely related structures and those that were materially different (Fig. \ref{fig-control-search}). 
 As expected, the positively biased population (40\% of the initial particle population seeded with network A) gave the best results, where the correct structure was preferentially identified (Fig.
 \ref{fig-control-search}A).
-On the other hand, when given a uniform distribution, 
-the PSO approach identified a combination of network A and network C as the most likely control structures (Fig. \ref{fig-control-search}B).
-Network A and C differ by the regulatory connection between the end-product P$_2$ and enzyme $E_{1}$; in network A end-product P$_{2}$ was assumed to inhibit $E_{1}$ while in network C end-product P$_{2}$ activated $E_{1}$.
-Lastly, when the initial population was biased towards a incorrect structures (initial population seeded with 90\% incorrect structures),
-the particle swarm \textit{misidentified} the correct allosteric structure (Fig. \ref{fig-control-search}C). 
+On the other hand, when given a uniform distribution, the PSO approach identified a combination of network A and network C as the most likely control structures (Fig. \ref{fig-control-search}B). 
+Network A and C differ by the regulatory connection between the end product P$_2$ and enzyme $E_{1}$; in network A, end product P$_{2}$ was assumed to inhibit $E_{1}$, while in network C, end product P$_{2}$ activated $E_{1}$.
+Lastly, when the initial population was biased towards incorrect structures (initial population seeded with 90\% incorrect structures), the particle swarm \textit{misidentified} the correct allosteric structure (Fig.
+\ref{fig-control-search}C).
 Interestingly, while each particle swarm identified parameter sets that minimized the simulation error, the estimated parameter values were not necessarily similar to the true parameters. 
 The angle between the estimated and true parameters was not consistently small across the swarms (identical parameters would give an angle of zero). 
-This suggested that our particle swarm approach identified a \textit{sloppy} ensemble, i.e., 
-parameter estimates that were individually incorrect but collectively exhibited the correct model behavior.
+This suggested that our particle swarm approach identified a \textit{sloppy} ensemble, i.e., parameter estimates that were individually incorrect but collectively exhibited the correct model behavior.
 
 We calculated control program output and scaled metabolic flux for the positively, uniformly and negatively biased particle swarms (Fig. \ref{fig-flux-pattern}). 
-Network A and network C models from the positively (Fig. \ref{fig-flux-pattern}A) and uniformly (Fig. \ref{fig-flux-pattern}B) 
-biased particle swarms showed similar operational patterns, despite differences in kinetic parameters and control structures.
-While models from the negatively biased population had error values similar to the correct structures in the previous swarms,
-they have different flux and control profiles (Fig. \ref{fig-flux-pattern}C). 
-In all cases, irregardless of network configuration or parameter values, the rate of enzyme decay was small compared to the other fluxes,
-and all networks had qualitatively similar trends for $E_{3}$ and $E_{6}$ control.
-Moreover, consistent with the correct model structure, production of end-product P$_1$ was the preferred branch for all model configurations.
+Network A and network C models from the positively (Fig.
+\ref{fig-flux-pattern}A) and uniformly (Fig.
+\ref{fig-flux-pattern}B) biased particle swarms showed similar operational patterns, despite differences in kinetic parameters and control structures.
+While models from the negatively biased population had error values similar to the correct structures in the previous swarms, they have different flux and control profiles (Fig.
+\ref{fig-flux-pattern}C).
+In all cases, regardless of network configuration or parameter values, the rate of enzyme decay was small compared to the other fluxes, and all networks had qualitatively similar trends for $E_{3}$ and $E_{6}$ control.
+Moreover, consistent with the correct model structure, production of end product P$_1$ was the preferred branch for all model configurations. 
 However, there was variability in P$_{2}$ production flux across the population of models, especially for the uniform swarm when compared with the other cases. 
-High P$_{1}$ branch flux resulted in end product inhibition of $E_{1}$ in both network A and network C, however in network D and E high P$_1$ flux induced $E_1$ activation. 
-These trends were manifested in different flux profiles, where the negatively biased population appeared more uniform across the population compared with the other swarms, and
-had higher $E_{1}$ specific activity. 
-Interestingly, the behavior of network A and network C highlighted an artifact of our integration rule; 
-both a positive or negative feedback connection from P$_2$ to $E_{1}$ were ignored because the P$_{1}$ inhibition of $E_{1}$ was dominate.  
+High P$_{1}$ branch flux resulted in end product inhibition of $E_{1}$ in both network A and network C, however in network D and E, high P$_1$ flux induced $E_1$ activation. 
+These trends were manifested in different flux profiles, where the negatively biased population appeared more uniform across the population compared with the other swarms, and had higher $E_{1}$ specific activity.
+Interestingly, the behavior of network A and network C highlighted an artifact of our integration rule; both a positive or negative feedback connection from P$_2$ to $E_{1}$ were ignored because the P$_{1}$ inhibition of $E_{1}$ was dominate.
 Thus, while theoretically distinct, network A and network C appeared operationally to the PSO algorithm to be that same network. 
 On the other hand, networks B, D and E showed distinct behavior that was not consistent with the true network. 
-These architectures exhibited either limited inhibition (network B) or activation (network D and E) of $E_{1}$ activity, resulting in significantly different
-metabolic flux profiles. However, the PSO was able to find low error parameter solutions, despite the mismatch in the control structures.
+These architectures exhibited either limited inhibition (network B) or activation (network D and E) of $E_{1}$ activity, resulting in significantly different metabolic flux profiles. 
+However, the PSO was able to find low error parameter solutions, despite the mismatch in the control structures. 
 Taken together, these results suggested that a uniform sampling approach could potentially yield an unbiassed estimate of both kinetic parameters control structures.
@@
 \section*{Discussion}
-In this study, we presented an effective kinetic modeling strategy to dynamically simulate cell free biochemical networks. 
+In this study, we presented an effective kinetic modeling strategy to dynamically simulate cell-free biochemical networks. 
 Our proposed strategy integrated traditional kinetic modeling with an effective rules based approach to dynamically describe metabolic regulation and control. 
-We tested this approach by developing kinetic models of hypothetical cell free metabolic networks. 
-In particular, we tested whether our effective modeling approach could describe classically expected behavior, 
-and second whether we could simultaneously estimate kinetic parameters and regulatory connectivity, in the absence of specific mechanistic knowledge, 
-from synthetic experimental data. Toward these questions, we explored five hypothetical cell-free networks. In each network, a
-substrate S was converted to the end-products P$_{1}$ and P$_{2}$ through a series of enzymatically catalyzed reactions, 
-including a branch point at a hypothetical metabolite M$_{2}$. Each network also included the same cofactors and cofactor recycle architecture. 
+We tested this approach by developing kinetic models of hypothetical cell-free metabolic networks. 
+In particular, we tested whether our effective modeling approach could describe classically expected behavior, and second whether we could simultaneously estimate kinetic parameters and regulatory connectivity, in the absence of specific mechanistic knowledge, from synthetic experimental data.
+Toward these questions, we explored five hypothetical cell-free networks. 
+In each network, a substrate S was converted to the end products P$_{1}$ and P$_{2}$ through a series of enzymatically catalyzed reactions, including a branch point at a hypothetical metabolite M$_{2}$.
+Each network also included the same cofactors and cofactor recycle architecture. 
 However, while all five networks shared the same enzymatic connectivity, each had different allosteric regulatory connectivity. 
-We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, could
-capture complex allosteric patterns such as ultrasensitivity, or non-competitive inhibition in the absence of specific mechanistic information. 
-Moreover, when integrated into network models, these rules captured classical regulatory patterns such as product induced feedback inhibition. 
-Lastly, we simultaneously estimated kinetic parameters and discriminated between competing regulatory structures, using synthetic data in combination with a
-modified particle swarm approach.
+We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, could capture complex allosteric patterns such as ultrasensitivity, or non-competitive inhibition in the absence of specific mechanistic information.
+Moreover, when integrated into network models, these rules captured classical regulatory patterns such as product-induced feedback inhibition. 
+Lastly, we simultaneously estimated kinetic parameters and discriminated between competing regulatory structures, using synthetic data in combination with a modified particle swarm approach.
 
@@
 
-While the results of this study were encouraging, there are several critical next steps that must be accomplished before we can model genome scale cell free metabolic networks.
+While the results of this study were encouraging, there are several critical next steps that must be accomplished before we can model genome-scale cell-free metabolic networks.
 [FINISH ME]
@@
 \subsection*{Formulation and solution of the model equations.}
-We used ordinary differential equations (ODEs) to model the time evolution of metabolite ($x_{i}$) and scaled enzyme abundance ($\epsilon_{i}$) in hypothetical cell free
-metabolic networks:
+We used ordinary differential equations (ODEs) to model the time evolution of metabolite ($x_{i}$) and scaled enzyme abundance ($\epsilon_{i}$) in hypothetical cell-free metabolic networks:
 \begin{eqnarray}
@@
 	\frac{d\epsilon_{i}}{dt} & = & -\lambda_{i}\epsilon_{i}\qquad{i = 1,2,\hdots,\mathcal{E}}
-\end{eqnarray}where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of metabolites and $\mathcal{E}$ denotes the number of
-enzymes in the model. The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ 
-denotes the rate of reaction $j$. Typically, reaction $j$ is a non-linear function of metabolite and enzyme abundance, as well as unknown kinetic parameters 
-$\mathbf{k}$ ($\mathcal{K}\times{1}$).
-The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$. If 
-$\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$. Conversely, if $\sigma_{ij}>0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates
-metabolite $i$ is not connected with reaction $j$. Lastly, $\lambda_{i}$ denotes the scaled enzyme degradation constant. The system material balances were subject to the
-initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$ and $\mathbf{\epsilon}\left(t_{o}\right) = \mathbf{1}$ (initially we have 100\% cell-free enzyme abundance).
+\end{eqnarray}where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of metabolites and $\mathcal{E}$ denotes the number of enzymes in the model. 
+The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of reaction $j$. 
+Typically, reaction $j$ is a non-linear function of metabolite and enzyme abundance, as well as unknown kinetic parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$). 
+The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$. 
+If $\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$. 
+Conversely, if $\sigma_{ij}>0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates metabolite $i$ is not connected with reaction $j$. 
+Lastly, $\lambda_{i}$ denotes the scaled enzyme degradation constant. 
+The system material balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$ and $\mathbf{\epsilon}\left(t_{o}\right) = \mathbf{1}$ (initially we have 100\% cell-free enzyme abundance).
 
@@
 	r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right) = \bar{r}_{j}v_{j}
-\end{equation}We used multiple saturation kinetics to model the reaction term $\bar{r}_{j}$:
+\end{equation}
+We used multiple saturation kinetics to model the reaction term $\bar{r}_{j}$:
 \begin{equation}\label{eqn:rate-bar}
@@
 The allosteric regulation term $v_{j}$ depended upon the combination of factors which influenced the activity of enzyme $i$.
-For each enzyme, we used a rule based approach to select from competing control factors (Fig. \ref{fig-control-schematic}). 
+For each enzyme, we used a rule-based approach to select from competing control factors (Fig. \ref{fig-control-schematic}). 
 If an enzyme was activated by $m$ metabolites, we modeled this activation as:
@@
 	d_{j} &=& 1 - \max_{j^{-}}\left(f_{1j}\left(x\right),\hdots,f_{nj}\left(x\right)\right)
-\end{eqnarray}The quantities $j^{+}$ and $j^{-}$ denoted the sets of activating, and inhibitory factors for enzyme $j$. 
+\end{eqnarray}
+The quantities $j^{+}$ and $j^{-}$ denoted the sets of activating and inhibitory factors for enzyme $j$. 
 If an enzyme had no allosteric factors, we set $v_{j} = 1$.
@@
 The $\kappa_{ij}$ parameter was species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).
-The model equations were encoded using the Octave programming language, and solved using the LSODE routine in Octave \citep{Octave:2014}.
+The model equations were encoded using the Octave programming language and solved using the LSODE routine in Octave \citep{Octave:2014}.
 
@@
 
-We minimized the model residual using Particle swarm optimization (PSO) \citep{PSO}.
+We minimized the model residual using particle swarm optimization (PSO) \citep{PSO}.
 PSO uses a \textit{swarming} metaheuristic to explore parameter spaces. 
@@
 \includegraphics[width=1.0\textwidth]{./figs/Figure-2-ControlSchematic.pdf}
-\caption{Schematic of rule based allosteric enzyme activity control laws. 
-Traditional enzyme kinetic expressions e.g., Michaelis–Menten or multiple saturation kinetics are multiplied by an enzyme activity control variable $0 \leq v_{j} \leq 1 $. 
+\caption{Schematic of rule-based allosteric enzyme activity control laws. 
+Traditional enzyme kinetic expressions, e.g., Michaelis–Menten or multiple saturation kinetics, are multiplied by an enzyme activity control variable $0 \leq v_{j} \leq 1 $. 
 Control variables are functions of many possible regulatory factors encoded by arbitrary functions of the form $0\leq f_{j}\left(\mathcal{Z}\right)\leq 1$.
@@
 Control parameters were fixed during the ensemble calculations.
-\textbf{A}: End-product P$_{1}$ abundance versus time for Network A. 
-The abundance of P$_{1}$ decreased with end-product inhibition of $E_{1}$ activity (Control-ON) versus the no inhibition case (Control-OFF). 
-\textbf{B}: End-product P$_{2}$ abundance versus time for Network B. Inhibition of branch point $E_{6}$ by end-product P$_{1}$ decreased P$_{2}$ abundance (Control-ON) versus the
+\textbf{A}: End product P$_{1}$ abundance versus time for Network A. 
+The abundance of P$_{1}$ decreased with end product inhibition of $E_{1}$ activity (Control-ON) versus the no inhibition case (Control-OFF). 
+\textbf{B}: End product P$_{2}$ abundance versus time for Network B. Inhibition of branch point $E_{6}$ by end product P$_{1}$ decreased P$_{2}$ abundance (Control-ON) versus the
 no inhibition case (Control-OFF).
-\textbf{C}: End-product P$_{1}$ abundance versus time for Network A. 
-Inhibition of branch point $E_{6}$ by end-product P$_{1}$ decreased P$_{1}$ abundance (Control-ON) versus the no inhibition case (Control-OFF).
+\textbf{C}: End product P$_{1}$ abundance versus time for Network A. 
+Inhibition of branch point $E_{6}$ by end product P$_{1}$ decreased P$_{1}$ abundance (Control-ON) versus the no inhibition case (Control-OFF).
 }\label{fig-onoff-simulations}
@@
 Synthetic experimental data was generated from a hypothetical parameter set using Network A, 
-where substrate $S$, end-product P$_{1}$ and intermediate M$_5$ were sampled approximately every 20 minutes. 
+where substrate $S$, end product P$_{1}$ and intermediate M$_5$ were sampled approximately every 20 minutes. 
 For cases $\textbf{A,B}$ 20 particles were initialized with randomized parameters and allowed to search for 300 iterations. 
@@
 Simulation error was minimized for a synthetic data set 
-($S$, end-product P$_{1}$ and intermediate M$_5$ sampled approximately every 20 min) generated using Network A.
+($S$, end product P$_{1}$ and intermediate M$_5$ sampled approximately every 20 min) generated using Network A.
 \textbf{A}: Simulation error versus parameter set angle for N = 100 particles biased toward the correct regulatory program (A,B,C,D,E) = (40\%, 10\%, 20\%, 20\% and 10\%).

### 2014-09-04 Paper_v1.tex
grant numbers have been added

@@
 Thus, modern constraint-based approaches have proven extremely useful in the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \citep{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}.
-However, genome-scale kinetic models of industrial important organisms such as \textit{E. coli} have yet to be constructed.
+However, genome-scale kinetic models of industrial important organisms such as \textit{E.
+coli} have yet to be constructed.
 
@@
 We then generated an ensemble of model parameter estimates by minimizing the difference between model simulations and the synthetic data using particle swarm optimization (PSO), starting from random initial parameter guesses.
-The estimation of kinetic parameters was sensitive to the choice of regulatory structure (Fig. \ref{fig-parameter-fit}). 
+The estimation of kinetic parameters was sensitive to the choice of regulatory structure (Fig.
+\ref{fig-parameter-fit}).
 PSO identified an ensemble of parameters that bracketed the mean of the synthetic measurements in less than 1000 iterations when the control structure was correct (Fig.
@@
 These architectures exhibited either limited inhibition (network B) or activation (network D and E) of $E_{1}$ activity, resulting in significantly different metabolic flux profiles. 
-However, the PSO was able to find low error parameter solutions, despite the mismatch in the control structures. 
-Taken together, these results suggested that a uniform sampling approach could potentially yield an unbiassed estimate of both kinetic parameters control structures.
+However, the PSO was able to find low error parameter solutions, despite the mismatch in the control structures (error values similar, but not better than the best network A and network C estimates). 
+Taken together, these results suggested that a uniform sampling approach could potentially yield an unbiassed estimate of both kinetic parameters and control structures. 
+However, the negatively biased particle swarm results illustrated a potential shortcoming of the approach, namely convergence to a local error minimum despite a significantly incorrect control structure.
+This suggested that estimated model structures will need to be further evaluated, for example by generating falsifiable experimental designs which could distinguish between low error solutions.
 
@@
 The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
-parameters ($\mathcal{K}\times{1}$). In thus study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$, which was taken from XXX. The quality of parameter
+parameters ($\mathcal{K}\times{1}$). In thus study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$. The quality of parameter
 estimates was measured using two criteria, goodness of fit (model residual) and angle between the estimated parameter vector $\mathbf{k}_{j}$ and the true parameter set $\mathbf{k}^{*}$:
@@
 \section*{Acknowledgements}
-This study was supported by the National Science Foundation GK12 award (DGE-1045513) 
-and by the National Science Foundation CAREER award (FILLMEIN).
+This study was supported by an award from the National Science Foundation (MCB \#1411715) 
+and the Army Research Office (ARO \#59155-LS).
 

### 2014-09-05 Paper_v1.tex
second paragraph in the discussion

@@
 Lastly, we simultaneously estimated kinetic parameters and discriminated between competing regulatory structures, using synthetic data in combination with a modified particle swarm approach.
+If we considered all putative regulatory architectures to be equally likely, we were able to estimate a \textit{sloppy} ensemble of models with the correct architecture and kinetic parameters.
 
-[CONTRAST WITH CYBERNETIC MODELS].
+The proposed modeling strategy shares features with other popular techniques, but has also has several key differences. 
+At its core, our effective modeling approach is similar to regulatory constraints based methods, and to the cybernetic modeling paradigm developed by Ramkrishna and colleagues.
+Covert, Palsson and coworkers drastically improved the predictability of constraints based approaches by integrating boolean rules into the calculation of metabolic fluxes \citep{Covert:2001aa}.
+If the regulated intracellular flux problem is coupled with time-dependent extracellular balances, these models can predict complex behavior such as diauxie growth or the switch between 
+aerobic and anaerobic metabolism. Another important features of this approach is that it scales with biological complexity. 
+For example, Covert \emph{et al.,} showed that a genome scale model of \emph{E.~coli} augmented with a boolean rule layer, 
+correctly predicted 80\% of the outcomes for high-throughput growth phenotyping experiments in \emph{E. coli}.
+Further, they showed that we could learn new biology by iteratively refining the model and its associated rules \citep{Covert:2004aa}. 
+However, while regulated flux balance analysis is a powerful technique, it does not easily allow the calculation of time-resolved metabolite abundance. 
+Additionally, the boolean rules which populate the regulatory layer are limited to ON/OFF decisions; for qualitative predictions of gene expression this is a reasonable limitation. 
+However, boolean rules will likely be less effective at capturing dynamic allosteric regulation in a cell free metabolic system. 
+On the other hand, the strength of cybernetic models is the integration of optimal metabolic control heuristics with traditional kinetic pathway modeling. 
+Cybernetic models are highly predictive; they have successfully predicted mutant behavior from limited wild-type data \citep{1999_varner_ramkrishna_MetaEng,Varner:2000aa,Song:2012aa}, 
+steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, strain specific metabolic function \citep{Song:2011aa} and have been used in bioprocess control applications \citep{Gadkar:2003aa}.
+However, cybernetic control heuristics are not mechanistic, instead they are the output of an optimal decision with respect to a set of hypothetical physiological objectives. 
+Thus, they are abstractions which are difficult to translate into a specific biological mechanism. 
+Our approach addresses the shortcomings of both regulatory constraints based models and cybernetic models. 
+First, similar to cybernetic models, the core of our approach is a kinetic model. Thus, we are able to directly calculate the time evolution of metabolism, for example the dynamic abundance of 
+network metabolites. Second, similar to regulatory flux balance analysis, our control laws describe specific mechanistic motifs such a activation or inhibition of enzyme activity. 
+However, our rules are continuous, thus they potentially allow a finer grained description metabolic regulation and control mechanisms.
+Lastly, we can naturally incorporate non modeled factors and categorical factors or factor combinations into our control law formulations.
 
 While the results of this study were encouraging, there are several critical next steps that must be accomplished before we can model genome-scale cell-free metabolic networks.
-[FINISH ME]
+First, it is unclear how this approach can be scaled to a genome scale cell free network. The central challenges to scaling up our approach will be the estimation of network parameters, 
+and the identification of allosteric regulatory connectivity. The particle swarm approach presented here quickly identified a \emph{sloppy} ensemble of model parameters.
 

### 2014-09-06 Paper_v1.tex
third paragraph in the discussion

@@
 For example, Covert \emph{et al.,} showed that a genome scale model of \emph{E.~coli} augmented with a boolean rule layer, 
-correctly predicted 80\% of the outcomes for high-throughput growth phenotyping experiments in \emph{E. coli}.
-Further, they showed that we could learn new biology by iteratively refining the model and its associated rules \citep{Covert:2004aa}. 
+correctly predicted approximately 80\% of the outcomes of a high-throughput growth phenotyping experiment in \emph{E. coli}.
+Further, they showed that they could learn new biology by iteratively refining the model and its associated rules \citep{Covert:2004aa}. 
 However, while regulated flux balance analysis is a powerful technique, it does not easily allow the calculation of time-resolved metabolite abundance. 
@@
 
-While the results of this study were encouraging, there are several critical next steps that must be accomplished before we can model genome-scale cell-free metabolic networks.
-First, it is unclear how this approach can be scaled to a genome scale cell free network. The central challenges to scaling up our approach will be the estimation of network parameters, 
-and the identification of allosteric regulatory connectivity. The particle swarm approach presented here quickly identified a \emph{sloppy} ensemble of model parameters.
+While the results of this proof of concept study were encouraging, there are several critical questions that should be explored.
+First, it is unclear how this approach will scale to genome scale networks, and second it is unclear how we will design the appropriate experiments 
+to identify model parameters and allosteric connectivity. 
+We showed that particle swarm optimization quickly identified an ensemble of model parameters, at least for proof-of-concept metabolic networks using synthetic data.
+This suggested that we could expect reasonable model predictions, despite only partial knowledge of model parameters. 
+Brown and Sethna showed that good predictions were possible despite only order of magnitude estimates of parameter values in a model of signal transduction \citep{Brown:2003aa}.
+Sethna and coworkers further showed that model behavior is often controlled by only a few parameter combinations, 
+a characteristic seemingly universal to multi-parameter models referred to as ``sloppiness'' \citep{Machta:2013aa}.
+We have also seen \textit{sloppy} behavior in a wide class of signal transduction processes \citep{Luan:2007aa,Song:2009aa,Tasseff:2010aa,Tasseff:2011aa,Nayak:2011aa,Lequieu:2011aa}.
+Thus, we expect good model predictions despite uncertainty in the estimated parameter values.
+Perhaps more critical will be our ability to estimate allosteric connectivity.
 

### 2014-09-12 Paper_v1.tex
First commit of reduced order coagulation materials

--- 2014-09-04 Paper_v1.tex
@@
 %\usepackage[authoryear,round,comma,sort&compress]{natbib}
-\usepackage[square,sort,comma,numbers]{natbib}
+\usepackage[square,sort&compress,comma,numbers]{natbib}
 %\usepackage[authoryear,round]{natbib}
@@
 \begin{titlepage}
-{\par\centering\textbf{\Large Dynamic Modeling of Cell-Free Biochemical Networks using Effective Kinetic Models}}
+{\par\centering\textbf{\Large Dynamic Modeling of the Human Coagulation Cascade using Reduced Order Effective Kinetic Models}}
 \vspace{0.05in}
-{\par \centering \large{Joseph A. Wayman, Adithya Sagar and Jeffrey D. Varner$^{*}$}}
+{\par \centering \large{Adithya Sagar and Jeffrey D. Varner$^{*}$}}
 \vspace{0.10in}
@@
 \vspace{0.1in}
-{\par \centering \textbf{Running Title:}~Effective models of metabolism}
+{\par \centering \textbf{Running Title:}~Modeling blood coagulation}
 \vspace{0.1in}
-{\par \centering \textbf{To be submitted:}~\textit{Processes}}
+{\par \centering \textbf{To be submitted:}~\emph{Processes}}
 \vspace{0.5in}
@@
 \section*{Abstract}
-Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
-Many of the challenges confronting genome-scale kinetic modeling can potentially be overcome in a cell-free system. 
-For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth. 
-Thus, cell-free operation holds several significant advantages for model development, identification and validation. 
-Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms, such as \textit{E.
-coli}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.
-Toward this unmet need, we present an effective biochemical network modeling framework for building dynamic cell-free metabolic models.
-The key innovation of our approach is the integration of simple effective rules encoding complex allosteric regulation with traditional kinetic pathway modeling. 
-We tested our approach by modeling the time evolution of several hypothetical cell-free metabolic networks. 
-We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, captured complex allosteric patterns such as ultrasensitivity or non-competitive inhibition in the absence of mechanistic information.
-Second, when integrated into network models, these rules captured classic regulatory patterns such as product-induced feedback inhibition. 
-Lastly, we showed, at least for the network architectures considered here, that we could simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data. 
-While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.
 
-{\noindent \textbf{Keywords:}~Cell-free metabolism, Mathematical modeling}
+{\noindent \textbf{Keywords:}~Coagulation, Mathematical modeling, Systems Biology}
 
@@
 
+%Developing computational tools to understand the molecular basis of human disease is a grand challenge facing systems biology.
+%Many have suggested that the integration of experimental and computational research is required to unravel
+%critical questions facing molecular medicine \cite{KITANO2002,ASSMUS2006,ARNAUD2006}. 
+%Despite this, mechanistic network modeling has not played a significant role in the development of new therapies for cancer, cardiovascular diseases or for acute events  
+%like thrombosis during surgery.
+%A critical limiting issue often cited has been model uncertainty \cite{BAILEY2001}. 
+%Development and validation of detailed network models requires a significant investment of time and material resources. 
+%Alternatively, development of tools using toy models often fails to demonstrate relevance to biologists or physicians.  
+%What is needed is a model system, that is simultaneously tractable and directly applicable to the study of disease.
+%Human blood is perhaps such an ideal model system \cite{Diamond:2009yf}. 
+%Blood is a complex mixture of proteins and cells which collectively carry out many functions.
+%However, one well studied facet of blood, which could serve as a valuable model system, is the coagulation cascade. 
+%Understanding the mechanism of rFVIIa might provide insight into the therapeutic dose under different physiological conditions.
+%More generally, rFVIIa treatment could also serve an ideal test case to develop strategies which retrospectively estimate operational perturbations associated with drugs.
+%For example, the effect of rFVIIa is directly observable in coagulation outputs such as thrombin concentration.
+%Thus, by analyzing these outputs in a normal versus a perturbed network, it might be possible to reverse engineer the molecular influence associated with a therapeutic intervention.
+
 \section*{Introduction}
-Mathematical modeling has long contributed to our understanding of metabolism. 
-Decades before the genomics revolution, mechanistically, structured metabolic models arose from the desire to predict microbial phenotypes resulting from changes in intracellular or extracellular states \citep{1976_fredrickson_BiotechBioeng}.
-The single cell \textit{E.
-coli} models of Shuler and coworkers pioneered the construction of large-scale, dynamic metabolic models that incorporated multiple, regulated catabolic and anabolic pathways constrained by experimentally determined kinetic parameters \citep{1984_domach_shuler_BiotechBioeng_01}.
-Shuler and coworkers generated many single cell kinetic models, including single cell models of eukaryotes \citep{1989_steinmeyer_shuler_ChemEngSci,1992_wu_shuler_AnnNYAcadSci}, minimal cell architectures \citep{2004_castellanos_shuler_PNAS}, as well as DNA sequence based whole-cell models of \textit{E.
-coli} \citep{2008_atlas_shuler_IETSysBio}.
-Conversely, highly abstracted kinetic frameworks, such as the cybernetic framework, represented a paradigm shift, viewing cells as growth-optimizing strategists \citep{1985_dhurjati_ramkrishna_tsao_BiotechBioeng}.
-Cybernetic models have been highly successful at predicting metabolic choice behavior, e.g., diauxie behavior \citep{1986_kompala_ramkrishna_tsao_BiotechBioeng}, steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, as well as the cellular response to metabolic engineering modifications \citep{1999_varner_ramkrishna_MetaEng}.
-Unfortunately, cybernetic models also suffer from an identifiability challenge, as both the kinetic parameters and an abstracted model of cellular objectives must be estimated simultaneously.
+The human coagulation system is an archetype proteolytic cascade regulated by both positive and negative feedback loops.
+The biology of coagulation is well studied \cite{ROBERTS1998,MANN1999,BUTENAS2002,FURIE2004}. 
+Coagulation is mediated by a family serine proteases, called factors and a key group of blood cells, called platelets. 
+The activation of coagulation factors and platelets, which are normally inactive in the circulation, requires a trigger event.
+Trigger events such as vessel injury expose procoagulant materials like collagen, fibronectin, von Willebrand factor (vWF) or tissue factor (TF).
+These materials drive platelet localization and activation, coagulation factor activation and ultimately clot formation \cite{GIESEN1999}.  
+Once coagulation has been activated, two converging pathways activate the master protease thrombin. 
+The extrinsic cascade is generally believed to be the main mechanism of thrombinogenesis in the blood \cite{MANN1990,ROBERTS1998,MANN1999}.
+Thrombin generation consists of three phases, initiation, propagation and termination \cite{GOLDHABER2006, BRUMMEL2002}.  
+Initially thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa) \cite{BUTENAS2002a}. 
+Picomolar amounts of thrombin can then activate the cofactors factors V and VIII (fV and fVIII) and platelets, 
+resulting in the formation of the \emph{tenase} and \emph{prothrombinase} complexes on the surface of activated platelets.
+These complexes amplify the early coagulation signal by activating downstream and upstream coagulation factors.  
+Termination occurs after prothrombin is consumed or activated thrombin is neutralized by inhibitors such as activated protein C (APC) or antithrombin III (ATIII).
 
-In the post genomics world, large-scale stoichiometric reconstructions of microbial metabolism popularized by static, constraint-based modeling techniques such as flux balance analysis (FBA) have become standard tools \citep{2012_lewis_palsson_NatRevMicrobio}.
-Since the first genome-scale stoichiometric model of \textit{E.
-coli}, developed by Edwards and Palsson \citep{2000_edwards_palsson_PNAS}, well over 100 organisms, including industrially important prokaryotes such as \textit{E.
-coli} \citep{Feist:2007aa} or \textit{B.
-subtilis} \citep{Oh:2007aa}, are now available \citep{2009_feist_palsson_NatRevMicrobio}.
-Stoichiometric models rely on a pseudo-steady-state assumption to reduce unidentifiable genome-scale kinetic models to an underdetermined linear algebraic system, which can be solved efficiently even for large systems.
-Traditionally, stoichiometric models have also neglected explicit descriptions of metabolic regulation and control mechanisms, instead opting to describe the choice of pathways by prescribing an objective function on metabolism.
-Interestingly, similar to early cybernetic models, the most common metabolic objective function has been the optimization of biomass formation \citep{2002_ibarra_edwards_palsson_Nat}, although other metabolic objectives have also been estimated \citep{2007_schuetz_sauer_MolSysBio}.
-Recent advances in constraint-based modeling have overcome the early shortcomings of the platform, including capturing metabolic regulation and control \citep{2013_hyduke_lewis_palsson_MolBioSys}. 
-Thus, modern constraint-based approaches have proven extremely useful in the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \citep{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}.
-However, genome-scale kinetic models of industrial important organisms such as \textit{E.
-coli} have yet to be constructed.
+Malfunctions in coagulation can have serious or potentially even fatal consequences.
+For example, aggressive clotting is involved with Coronary Artery Diseases (CADs), which collectively account for 38\% of all deaths in North America \cite{HANSSON2005}.
+Coagulation management during surgery can also be challenging, particularly because of the increasing clinical use of antithrombotic drugs \cite{Tanaka:2009wo}.   
+Insufficient coagulation due to genetic disorders such as hemophilia can also result in recurrent bleeding. 
+The coagulation factors VIII (fVIII) and IX (fIX) are deficient in Hemophilia A and B, respectively \cite{COOPER1994,MANNUCCI2001,MITCHELL2008}.
+People with mild hemophilia have 5-40\% of the normal clotting factor levels while severe hemophiliacs have $<$1\% \cite{MITCHELL2008}.
+Hemophilia can be controlled with regular infusions of the deficient clotting factors. 
+However, clotting factor replacement sometimes leads to the formation of \emph{in-vivo} fVIII and fIX inhibitors \cite{TOMOKIYO2003}.
+Activated Prothrombin Complex Concentrates (aPCCs) from pooled plasma have been used to treat hemophilia. 
+However, aPCCs can have safety issues \cite{CHAVIN1988,ALLEN2006,SCHNEIDERMAN2007}. 
+Alternatively, recombinant factor VIIa (rFVIIa) has been used to treat bleeding disorders \cite{Hedner:2008rm,Talbot:2009kk} including
+hemophilia with and without factor VIII/IX inhibitors \cite{Lusher:1998bs,Shapiro:1998sh,Shapiro:2000pr,Shapiro:2008px}.
+However, rFVIIa requires frequent administration (every 2-3 hr) because of its short half-life in the circulation. 
+Many questions also remain about its mechanism \cite{KOMIYAMA1990,PAYNE1996,MONROE1997,HOFFMAN1998,BUTENAS2002a,ROBERTS2004,ROBERTS2006,ALLEN2007}
+and effective dose range \cite{TOMOKIYO2003}.
+Thus, despite its wide adoption, the utility of rFVIIa in trauma-associated hemorrhage remains controversial \cite{Duchesne:2008hw}.
 
-Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
-Central amongst these advantages is direct access to metabolites and the microbial biosynthetic machinery without the interference of a cell wall. 
-This allows us to control as well as interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
-Second, cell-free systems also allow us to study biological processes without the complications associated with cell growth.
-Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \citep{Jewett:2008aa}. 
-However, CFPS is not new; CFPS in crude \textit{E. coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \citep{MATTHAEI:1961aa,NIRENBERG:1961aa}. 
-Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \citep{Lu:2014aa} to synthetic biology \citep{Hodgman:2012aa}. 
-Interestingly, many of the challenges confronting genome-scale kinetic modeling can potentially be overcome in a cell-free system. 
-For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth. 
-Thus, cell-free operation holds several significant advantages for model development, identification and validation.
-Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms, such as \textit{E.
-coli} or \textit{B.
-subtilis}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.
+Modeling approaches differ in their degree of mechanistic detail, and the choice of approach is often determined by prior system knowledge \citep{2012_kholodenko_kolch_SciSig}. 
+The ability of ODE modeling to capture dynamics has made it one of the most common tools for studying signaling systems. 
+However, dynamics and ODEs come at the expense of difficult (and often impossible) to solve model parameter identification problems. 
+For example, Gadkar et al., showed that even with near-perfect information and high frequency sampling, 
+it was often impossible to identify (to within a specified uncertainty) all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
+This reality highlights the perhaps under-appreciated role that experimental design could play in generating the best training and validation data sets for model identification \citep{Apgar:2010fk}.
+Moreover, detailed ODE models require significant mechanistic knowledge of the underlying biology. 
+However, as Bailey suggested more than a decade ago, achieving qualitative and quantitative understanding of complex biological systems should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}.
+Since Bailey's complex biology with no parameters hypothesis, Sethna and coworkers showed that model behavior and predictive ability were sensitive to only a few parameter combinations, a characteristic seemingly universal to multi-parameter signaling models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
+Thus, reasonable model predictions could be possible with limited parameter information. 
+Taking advantage of this property of model behavior, we developed sloppy techniques for parameter identification using ensembles of deterministic models. 
+Pareto optimal ensemble techniques (POETs) incorporate principles of competing objectives into a multi-objective optimization framework, enabling the exploration of kinetic parameter space while accounting for uncertainty and potential conflicts in the experimental training data \citep{2010_song_varner_BiotechJ}.
+We have proposed that the sloppy behavior of biological networks may also be a source of cell-to-cell \citep{2011_lequieu_chakrabarti_nayak_varner_PLoSCompBio} or even patient-to-patient heterogeneity \citep{2010_luan_varner_MolBioSys}.
+Recently, Bayesian techniques of parameter identification have also been used to explore cell-to-cell heterogeneity  \citep{2011_kalita_brasier_JBC,2011_hasenauer_allgower_BMCBioinfo}. 
+Thus, a population of cells could be viewed as a dynamic ensemble of networks as the operational biochemical pathways are often context-specific \citep{2012_creixell_linding_NatBiotech}.
 
-In this study, we present an effective biochemical network modeling framework for building dynamic cell-free metabolic models.
-The key innovation of our approach is the seamless integration of simple effective rules encoding complex regulation with traditional kinetic pathway modeling. 
+In this study, we constructed a reduced order effective model of thrombin generation in normal and hemophilic blood.
+We used this model to understand how physiological coagulation was altered by prothrombin levels, the activated protein C pathway and the level of ATIII.  
+Previous models of coagulation have been formulated as systems of nonlinear ordinary differential equations, often using mass-action kinetics to describe the rates of biochemical conversions [REFHERE].
+While this approach has proven highly effective, it does require significant experimental data for model identification and validation.    
+For example, a previous coagulation model from our laboratory, which described 193 proteins and protein complexes interconnected by 301 biochemical interactions, 
+contained 467 unknown parameters (301 kinetic parameters and 166 initial conditions) [REFHERE]. 
+The key innovation of our approach is the seamless integration of simple effective rules encoding complex regulatory motifs with traditional kinetic pathway modeling. 
 This integration allows the description of complex regulatory interactions, such as time-dependent allosteric regulation of enzyme activity, in the absence of specific mechanistic information. 
 The regulatory rules are easy to understand, easy to formulate and do not rely on overarching theoretical abstractions or restrictive assumptions. 
-We tested our approach by modeling the time evolution of several hypothetical cell-free metabolic networks. 
-In particular, we tested whether our effective modeling approach could describe classically expected enzyme kinetic behavior, and second whether we could simultaneously estimate kinetic parameters and regulatory connectivity, in the absence of specific mechanistic knowledge, from synthetic experimental data.
-Toward these questions, we explored five hypothetical cell-free networks. 
-Each network shared the same enzymatic connectivity, but had different allosteric regulatory connectivity. 
-We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, captured complex allosteric patterns such as ultrasensitivity or non-competitive inhibition in the absence of mechanistic information.
-Second, when integrated into network models, these rules captured classical regulatory patterns such as product-induced feedback inhibition. 
-Lastly, we showed, at least for the network architectures considered here, that we could simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data. 
-While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.
+We tested our approach by modeling the time evolution of the human coagulation cascade.
 
@@
 
-\subsection*{Formulation and properties of effective cell-free metabolic models.}
-We developed two proof-of-concept metabolic networks to investigate the features of our effective biochemical network modeling approach (Fig. \ref{fig-networks}). 
-In both examples, substrate S was converted to the end products P$_{1}$ and P$_{2}$ through a series of enzymatically catalyzed reactions, including a branch point at hypothetical metabolite M$_{2}$. 
-Several of these reactions involved cofactor dependence (AH or A), and various allosteric regulatory mechanisms modified the activity of pathway enzymes. 
-Network A included feedback inhibition of the initial pathway enzyme ($E_{1}$) by pathway end products P$_{1}$ and P$_{2}$ (Fig. \ref{fig-networks}A). 
-On the other hand, network B involved feedback inhibition of $E_{1}$ by P$_{2}$ and $E_{6}$ by P$_{1}$ (Fig. \ref{fig-networks}B). 
-In both networks, branch point enzymes $E_{3}$ and $E_{6}$ were subject to feed-forward activation by reduced cofactor AH. 
-Lastly, it is known experimentally that cell-free systems have a finite operational lifespan. 
-Loss of biosynthetic capability could be a function of many factors, e.g., cofactor or metabolite limitations. 
-We modeled the loss of biosynthetic capability as a non-specific first-order decay of enzyme activity.
+\subsection*{Formulation and properties of reduced order effective models.}
+Reduced order effective models are [FINISH ME].
 
-Allosteric regulation of enzyme activity was modeled by combining individual regulatory contributions to the activity of pathway enzymes into a control coefficient using an integration rule (Fig.
-\ref{fig-control-schematic}).
-This strategy is similar in spirit to the Constrained Fuzzy Logic (cFL) approach of Lauffenburger and coworkers which has been used to effectively model signal transduction pathways important in human health \citep{Morris:2011aa}.
-In our formulation, Hill-like transfer functions $0 \leq f\left(\mathcal{Z}\right) \leq 1$ were used to calculate the influence of factor abundance upon target enzyme activity. 
-In this context, factors can be individual metabolite levels or some function, e.g., the product of metabolite levels. 
-However, more generally, factors can also correspond to non-modeled influences, categorial variables or other abstract quantities. 
-In the current study, we simply let $\mathcal{Z}$ correspond to the abundance of individual metabolites. 
-When an enzyme was potentially sensitive to more than one regulatory input, logical integration rules were used to select which regulatory transfer function influenced enzyme activity at any given time.
-Thus, our test networks involved important features such as cofactor recycling, enzyme activity and metabolite dynamics, as well as multiple overlapping allosteric regulatory mechanisms.
+\clearpage
 
-The rule-based regulatory strategy approximated the behavior of classical allosteric activation and inhibition mechanisms (Fig. \ref{fig-kinetics-simulations}). 
-We considered the enzyme catalyzed conversion of substrate S to a product P, where the overall reaction rate was modeled as the product of a Michaelis-Menten term and an effective allosteric control variable reflecting the particular regulatory interaction.
-We first explored feed-forward substrate activation of enzyme activity (for both positive and negative cooperativity). 
-Consistent with classical data, the rule-based strategy predicted a sigmoidal relationship between substrate abundance and reaction rate as a function of the cooperativity parameter (Fig.
-\ref{fig-kinetics-simulations}A).
-For cooperativity parameters less than unity, increased substrate abundance \textit{decreased} the reaction rate. 
-This was consistent with the idea that substrate binding \textit{decreased} at regulatory sites, which negatively impacted substrate binding at the active site. 
-On the other hand, as the cooperativity parameter increased past unity, the rate of conversion of substrate S to product P by enzyme $E$ approached a step function. 
-In the presence of an inhibitor, the rule-based strategy predicted non-competitive like behavior as a function of the cooperativity parameter (Fig. \ref{fig-kinetics-simulations}B). 
-When the control gain parameter, $\kappa_{ij}$ in Eqn.
-\eqref{eqn:control-factor}, was greater than unity, the inhibitory force was directly proportional to the cooperativity parameter, $\eta$ in Eqn.
-\eqref{eqn:control-factor}.
-Thus, as the cooperativity parameter increased, the maximum reaction rate decreased (Fig. \ref{fig-kinetics-simulations}B). 
-Interestingly, our rule-based approach was unable to directly simulate competitive inhibition of enzyme activity. 
-Taken together, the rule-based strategy captured classical regulatory patterns for both enzyme activation and inhibition. 
-Thus, we are able to model complex kinetic phenomena such as ultrasensitivity, despite an effective description of reaction kinetics.
+\subsection*{Identification of reduced order coagulation model parameters using particle swarm optimization.}
+A critical challenge for any dynamic model is the estimation of kinetic parameters. 
+We estimated kinetic and control parameters simultaneously from \emph{in vitro} coagulation data sets with and without the protein C pathway. 
+The residual between model simulations and experimental measurements was minimized using particle swarm optimization (PSO). 
+A population of particles (N = 20) was initialized with randomized kinetic and control parameters and allowed to search for parameter vectors that minimized
+the residual. However, not all parameters were varied simultaneously. 
+We partitioned the parameter estimation problem into two subproblems based upon the biological organization of the training data; 
+(i) estimation of parameters associated with thrombin formation in the absence of the protein C pathway and (ii) estimation of parameters associated with the protein C pathway.
+Only those parameters associated with each subproblem were varied during the optimization procedure for that subproblem e.g., thrombin parameters were \textit{not} varied during the protein C subproblem.
+The PSO procedure was run for 20 generations for each subproblem, where each generation was 1200 iterations. The best particle from each generation was used to generate the particle population
+for the next generation. We rotated the subproblems, starting with subproblem 1 in the first generation.
 
-End product yield was controlled by feedback inhibition, while product selectivity was controlled by branch point enzyme inhibition (Fig. \ref{fig-onoff-simulations}). 
-A critical test of our modeling approach was to simulate networks with known behavior. 
-If we cannot reproduce the expected behavior of simple networks, then our effective modeling strategy, and particularly the rule-based approximation of allosteric regulation, will not be feasible for genome-scale cell-free problems.
-We considered two cases, control ON/OFF, for each network configuration.
-Each of these cases had identical kinetic parameters and initial conditions; the \textit{only} differences between the cases were the allosteric regulation rules and the control parameters associated with these rules.
-As expected, end product accumulation was larger for network A when the control was OFF (no feedback inhibition of $E_{1}$ by P$_{1}$ and P$_{2}$),
-as compared to the ON case (Fig. \ref{fig-onoff-simulations}A). 
-We found this behavior was robust to the choice of underlying kinetic parameters, as we observed that same qualitative response across an ensemble of randomized parameter sets (N = 100), for fixed control parameters.
-The control ON/OFF response of network B was more subtle. 
-In the OFF case, the behavior was qualitatively similar to network A. 
-However, for the ON case, flux was diverted away from P$_{2}$ formation by feedback inhibition of $E_{6}$ activity at the M$_{2}$ branch point by P$_{1}$ (Fig. \ref{fig-onoff-simulations}B). 
-Lower $E_{6}$ activity at the M$_{2}$ branch point allowed more flux toward P$_{1}$ formation, hence the yield of P$_{1}$ also increased (Fig. \ref{fig-onoff-simulations}C). 
-Again, the control ON/OFF behavior of network B was robust to changes in kinetic parameters, as the same qualitative trend was conserved across an ensemble of randomized parameters (N = 100), for fixed control parameters.
-Taken together, these simulations suggested that the rule-based allosteric control concept could robustly capture expected feedback behavior for networks with uncertain kinetic parameters.
+The reduced order coagulation model captured the role of initial prothrombin abundance, and the decay of the thrombin signal following from ATIII activity (Fig. \ref{fig-diff_fII_NO_APC}).
+However, we systematically under-predicted the thrombin peak and the strength of ATIII inhibition in this training data set.
+On the other hand, with fixed thrombin parameters, we captured peak thrombin values and the decay of the thrombin signal (at least for the 150\% fII case) in the presence
+of both ATIII and the protein C pathway (Fig. \ref{fig-diff_fII_APC}). Lastly, we were unable to capture global differences in initiation time \textit{across} separate data 
+sets with a single ensemble of model parameters. These differences likely result from normal experimental variability.
+For example different thrombin generation experiments within our 
+training data (at the same physiological factor levels) have significantly different initiation times (supplemental results). 
+However, this also highlights a potential shortcoming of the initiation module within the model. 
+To capture the variability in initiation time \textit{across} training data sets, we included a constant time-delay parameter ($T_{D}$) for each data group.
+The delay parameter was constant within a data set, but allowed to vary \textit{across} training data sets. Introduction of the delay parameter allowed the model
+to simulate multiple training data sets using a single ensemble of model parameters.
+Taken together, the model identification results suggested 
+that our kinetic-rules based approach could reproduce a panel of thrombin generation data sets conducted at physiological factor and inhibitors concentrations.
+However, it was unclear if the reduced order model could predict new data sets, without changing kinetic or control parameters.
 
-\subsection*{Estimating parameters and effective allosteric regulatory structures.}
-A critical challenge for any dynamic model is the estimation of kinetic parameters. 
-For metabolic processes, there is also the added challenge of identifying the regulation and control structures that manage metabolism. 
-Of course, these issues are not independent; any description of enzyme activity regulation will be a function of system state, which in turn depends upon the kinetic parameters. 
-For cell-free systems, regulated gene expression has been removed, however, enzyme activity regulation is still operational. 
-We explored this linkage by estimating model parameters from synthetic data using both network structures. 
-We generated noise corrupted synthetic measurements of the substrate S, intermediate M$_{5}$ and end product P$_1$ approximately every 20 min using
-network A.
-We then generated an ensemble of model parameter estimates by minimizing the difference between model simulations and the synthetic data using particle swarm optimization (PSO), starting from random initial parameter guesses.
-The estimation of kinetic parameters was sensitive to the choice of regulatory structure (Fig.
-\ref{fig-parameter-fit}).
-PSO identified an ensemble of parameters that bracketed the mean of the synthetic measurements in less than 1000 iterations when the control structure was correct (Fig.
-\ref{fig-parameter-fit}A and B).
-However, with control mismatch (network B simulated with network A parameters), model simulations were not consistent with the synthetic data  (Fig. \ref{fig-parameter-fit}C and D). 
-Taken together, these results suggested that we could perhaps simultaneously estimate both parameters and network control architectures, as incorrect control structures would be manifest as poor model fits.
+We tested the predictive power of the reduced order coagulation model with data sets not used for model training. 
+Two data sets were used from validation, thrombin generation for different initial prothrombin and ATIII concentrations in the absence of the protein C pathway, 
+and normal versus hemophilia thrombin generation in the presence of the protein C pathway. 
+The hemophilia data was an especially difficult test as it was taken from a different laboratory and a completely different
+\textit{in~vitro} assay method involving unmodeled factors such as platelets. 
+All kinetic and control parameters were fixed for the validation simulations. 
+The only globally adjustable parameter $T_{D}$, was fixed within each validation data set but allowed to vary between data sets.
+The reduced order model predicted the thrombin generation profile for ratios of prothrombin and ATIII in the absence of the protein C pathway (Fig. \ref{fig-fII_ATIII}).
+Model predictions for factors levels outside of the physiological range (fII,ATIII) = (50\%, 150\%) or (150\%, 50\%) were only qualitatively consistent with measured thrombin values.
+However, simulations closer to the physiological range (fII,ATIII) = (100\%, 100\%) or (125\%,75\%) tracked the measured thrombin values. 
+Likewise, simulations of thrombin generation in normal versus hemophilia (missing both fVIII and fIX) were consistent with measured thrombin values (Fig. \ref{fig-hemophilia}).
 
-We modified our particle swarm identification strategy to simultaneously search over both kinetic parameters and putative control structures. 
-In addition to our initial networks, we constructed three additional presumptive network models, each with the same enzymatic connectivity but different allosteric regulation of the pathway enzymes (Fig.
-\ref{fig-alternative-networks}).
-We then initialized a population of particles, each with one of the five potential regulatory programs and randomized kinetic parameters. 
-Thus, we generated an initial population of particles that had \textit{both} different kinetic parameters as well as different control structures. 
-We biased the distribution of the particle population according to our \textit{a prior} belief of the correct regulatory program. 
-To this end, we considered three different priors, a uniform distribution where each putative regulatory structure represented 20\% of the population and two mixed distributions that were either positively or negatively biased towards the correct structure (network A).
-In both the positively biased and uniform cases the PSO clearly differentiated between the true or closely related structures and those that were materially different (Fig. \ref{fig-control-search}). 
-As expected, the positively biased population (40\% of the initial particle population seeded with network A) gave the best results, where the correct structure was preferentially identified (Fig.
-\ref{fig-control-search}A).
-On the other hand, when given a uniform distribution, the PSO approach identified a combination of network A and network C as the most likely control structures (Fig. \ref{fig-control-search}B). 
-Network A and C differ by the regulatory connection between the end product P$_2$ and enzyme $E_{1}$; in network A, end product P$_{2}$ was assumed to inhibit $E_{1}$, while in network C, end product P$_{2}$ activated $E_{1}$.
-Lastly, when the initial population was biased towards incorrect structures (initial population seeded with 90\% incorrect structures), the particle swarm \textit{misidentified} the correct allosteric structure (Fig.
-\ref{fig-control-search}C).
-Interestingly, while each particle swarm identified parameter sets that minimized the simulation error, the estimated parameter values were not necessarily similar to the true parameters. 
-The angle between the estimated and true parameters was not consistently small across the swarms (identical parameters would give an angle of zero). 
-This suggested that our particle swarm approach identified a \textit{sloppy} ensemble, i.e., parameter estimates that were individually incorrect but collectively exhibited the correct model behavior.
-
-We calculated control program output and scaled metabolic flux for the positively, uniformly and negatively biased particle swarms (Fig. \ref{fig-flux-pattern}). 
-Network A and network C models from the positively (Fig.
-\ref{fig-flux-pattern}A) and uniformly (Fig.
-\ref{fig-flux-pattern}B) biased particle swarms showed similar operational patterns, despite differences in kinetic parameters and control structures.
-While models from the negatively biased population had error values similar to the correct structures in the previous swarms, they have different flux and control profiles (Fig.
-\ref{fig-flux-pattern}C).
-In all cases, regardless of network configuration or parameter values, the rate of enzyme decay was small compared to the other fluxes, and all networks had qualitatively similar trends for $E_{3}$ and $E_{6}$ control.
-Moreover, consistent with the correct model structure, production of end product P$_1$ was the preferred branch for all model configurations. 
-However, there was variability in P$_{2}$ production flux across the population of models, especially for the uniform swarm when compared with the other cases. 
-High P$_{1}$ branch flux resulted in end product inhibition of $E_{1}$ in both network A and network C, however in network D and E, high P$_1$ flux induced $E_1$ activation. 
-These trends were manifested in different flux profiles, where the negatively biased population appeared more uniform across the population compared with the other swarms, and had higher $E_{1}$ specific activity.
-Interestingly, the behavior of network A and network C highlighted an artifact of our integration rule; both a positive or negative feedback connection from P$_2$ to $E_{1}$ were ignored because the P$_{1}$ inhibition of $E_{1}$ was dominate.
-Thus, while theoretically distinct, network A and network C appeared operationally to the PSO algorithm to be that same network.
-On the other hand, networks B, D and E showed distinct behavior that was not consistent with the true network. 
-These architectures exhibited either limited inhibition (network B) or activation (network D and E) of $E_{1}$ activity, resulting in significantly different metabolic flux profiles. 
-However, the PSO was able to find low error parameter solutions, despite the mismatch in the control structures (error values similar, but not better than the best network A and network C estimates). 
-Taken together, these results suggested that a uniform sampling approach could potentially yield an unbiassed estimate of both kinetic parameters and control structures. 
-However, the negatively biased particle swarm results illustrated a potential shortcoming of the approach, namely convergence to a local error minimum despite a significantly incorrect control structure.
-This suggested that estimated model structures will need to be further evaluated, for example by generating falsifiable experimental designs which could distinguish between low error solutions.
+\subsection*{Global sensitivity analysis of the reduced order coagulation model.}
+We conducted a variance based global sensitivity analysis to better understand which parameters were controlling the performance of the coagulation model. 
+In particular, we calculated the sensitivity of two performance metrics, the time to maximum thrombin and the thrombin exposure (area under the thrombin curve) to
+changes in the model parameters (Fig. ZZ).
 
@@
 \section*{Discussion}
-In this study, we presented an effective kinetic modeling strategy to dynamically simulate cell-free biochemical networks. 
-Our proposed strategy integrated traditional kinetic modeling with an effective rules based approach to dynamically describe metabolic regulation and control. 
-We tested this approach by developing kinetic models of hypothetical cell-free metabolic networks. 
-In particular, we tested whether our effective modeling approach could describe classically expected behavior, and second whether we could simultaneously estimate kinetic parameters and regulatory connectivity, in the absence of specific mechanistic knowledge, from synthetic experimental data.
-Toward these questions, we explored five hypothetical cell-free networks. 
-In each network, a substrate S was converted to the end products P$_{1}$ and P$_{2}$ through a series of enzymatically catalyzed reactions, including a branch point at a hypothetical metabolite M$_{2}$.
-Each network also included the same cofactors and cofactor recycle architecture. 
-However, while all five networks shared the same enzymatic connectivity, each had different allosteric regulatory connectivity. 
-We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, could capture complex allosteric patterns such as ultrasensitivity, or non-competitive inhibition in the absence of specific mechanistic information.
-Moreover, when integrated into network models, these rules captured classical regulatory patterns such as product-induced feedback inhibition. 
-Lastly, we simultaneously estimated kinetic parameters and discriminated between competing regulatory structures, using synthetic data in combination with a modified particle swarm approach.
-
-[CONTRAST WITH CYBERNETIC MODELS].
-
-While the results of this study were encouraging, there are several critical next steps that must be accomplished before we can model genome-scale cell-free metabolic networks.
-[FINISH ME]
+Controlling hemophilia and more generally hemorrhage has been a driving force to understand the molecular basis of blood coagulation.
+Replacement therapy improves bleeding times for most hemophilia patients. However, repeated fVIII/fIX administration can lead to neutralizing inhibitors. 
+Alternatively, rFVIIa has been used as a universal hemostatic agent to initiate clotting with and without fVIII or fIX inhibitors \cite{Hedner:2008rm}.
+Despite its wide adoption, the mechanism and the utility of rFVIIa in trauma-associated hemorrhage remains controversial \cite{Duchesne:2008hw}. 
+In this study, we formulated a model of the human coagulation cascade and studied thrombin formation in normal and rFVIIa treated and untreated
+hemophilic plasma. The model described 193 proteins or protein complexes connected by 301 interactions. 
+The coagulation network architecture was based on literature and a previous mathematical model from our laboratory \cite{VARNER2007}.
+Simulations of thrombin dynamics were independently validated using blood drawn from patients with coronary artery disease (reported here) and previous TF/FVIIa and FVIIa 
+\emph{in-vitro} studies \cite{BUTENAS2002}.
+While the human coagulation cascade is perhaps an ideal model system to develop network analysis tools, not all the model parameters were identifiable given the current training data.
+Instead, an \emph{ensemble} of models (N = 437) was estimated using nine cell-based coagulation training sets \cite{ALLEN2007}.
+We used this family of consistent coagulation models in all simulations instead of a single best-fit but uncertain model.
+Additionally, using sensitivity analysis on the ensemble of models, we identified 
+structurally sensitive components of the coagulation architecture as a function of condition.
+We also demonstrated that the ensemble robustly constrained model predictions of independent validation sets, despite having many poorly constrained parameters.
+Thus, the model ensemble displayed \emph{sloppy} behavior similar to that observed by Sethna and coworkers for other signal transduction networks \cite{Gutenkunst2007}.  
+By studying a family of models, we perhaps partially addressed the uncertainty stemming from the many poorly characterized model parameters.
+However, many other factors could influence our results. For example, missing structural interactions or biophysical factors such as blood flow could play a large role.
+Thus, while the results presented here may be a valuable first step, more studies are required.
 
@@
 \subsection*{Formulation and solution of the model equations.}
-We used ordinary differential equations (ODEs) to model the time evolution of metabolite ($x_{i}$) and scaled enzyme abundance ($\epsilon_{i}$) in hypothetical cell-free metabolic networks:
-\begin{eqnarray}
-	\frac{dx_{i}}{dt} & = & \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)\qquad{i=1,2,\hdots,\mathcal{M}}\\
-	\frac{d\epsilon_{i}}{dt} & = & -\lambda_{i}\epsilon_{i}\qquad{i = 1,2,\hdots,\mathcal{E}}
-\end{eqnarray}where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of metabolites and $\mathcal{E}$ denotes the number of enzymes in the model. 
+We used ordinary differential equations (ODEs) to model the time evolution of proteins ($x_{i}$) in our reduced order coagulation model:
+\begin{equation}
+	\frac{dx_{i}}{dt}  =  \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)\qquad{i=1,2,\hdots,\mathcal{M}}\\
+\end{equation}
+where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of protein species in the model. 
 The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of reaction $j$. 
-Typically, reaction $j$ is a non-linear function of metabolite and enzyme abundance, as well as unknown kinetic parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$). 
+Typically, reaction $j$ is a non-linear function of biochemical species abundance, as well as unknown kinetic parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$). 
 The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$. 
@@
 Lastly, $\lambda_{i}$ denotes the scaled enzyme degradation constant. 
-The system material balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$ and $\mathbf{\epsilon}\left(t_{o}\right) = \mathbf{1}$ (initially we have 100\% cell-free enzyme abundance).
+The system material balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$.
 
@@
 
-The allosteric regulation term $v_{j}$ depended upon the combination of factors which influenced the activity of enzyme $i$.
+The control term $v_{j}$ depended upon the combination of factors which influenced the activity of enzyme $i$.
 For each enzyme, we used a rule-based approach to select from competing control factors (Fig. \ref{fig-control-schematic}). 
@@
 \begin{equation}
-	v_{j} = \max\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right)
-\end{equation}where $0\leq f_{ij}\left(x\right)\leq 1$ was a regulatory transfer function that calculated the influence of metabolite $i$ on the activity of enzyme $j$. 
+	v_{j} = \max\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{mj}\left(\mathcal{Z}\right)\right)
+\end{equation}where $0\leq f_{ij}\left(\mathcal{Z}\right)\leq 1$ was a regulatory transfer function that calculated the influence of metabolite $i$ on the activity of enzyme $j$. 
 Conversely, if enzyme activity was inhibited by a $m$ metabolites, we modeling this inhibition as:
 \begin{equation}
-	v_{j} = 1 - \max\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right)
+	v_{j} = 1 - \max\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{mj}\left(\mathcal{Z}\right)\right)
 \end{equation}Lastly, if an enzyme had both $m$ activating and $n$ inhibitory factors, we modeled the regulatory term as:
@@
 \begin{eqnarray}
-	u_{j} &=& \max_{j^{+}}\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right) \\
-	d_{j} &=& 1 - \max_{j^{-}}\left(f_{1j}\left(x\right),\hdots,f_{nj}\left(x\right)\right)
+	u_{j} &=& \max_{j^{+}}\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{mj}\left(\mathcal{Z}\right)\right) \\
+	d_{j} &=& 1 - \max_{j^{-}}\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{nj}\left(\mathcal{Z}\right)\right)
 \end{eqnarray}
@@
 If an enzyme had no allosteric factors, we set $v_{j} = 1$.
-There are many possible functional forms for $0\leq f_{ij}\left(x\right)\leq 1$. 
+There are many possible functional forms for $0\leq f_{ij}\left(\mathcal{Z}\right)\leq 1$. 
 However, in this study, each individual transfer function took the form:
 \begin{equation}\label{eqn:control-factor}
-	f_{i}\left(\mathbf{x}\right) = \frac{\kappa_{ij}^{\eta}x_{j}^{\eta}}{1 + \kappa_{ij}^{\eta}x_{j}^{\eta}}
-\end{equation}where $x_{j}$ denotes the abundance of metabolite $j$, and $\kappa_{ij}$ and $\eta$ are control parameters. 
+	f_{i}\left(\mathbf{x}\right) = \frac{\kappa_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}{1 + \kappa_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}
+\end{equation}where $\mathcal{Z}_{j}$ denotes the abundance of the $j$ factor (e.g., metabolite abundance), and $\kappa_{ij}$ and $\eta$ are control parameters. 
 The $\kappa_{ij}$ parameter was species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).
-The model equations were encoded using the Octave programming language and solved using the LSODE routine in Octave \citep{Octave:2014}.
+The model equations were encoded using the Python programming language and solved using the ODEINT routine of the SciPy module \citep{SCIPY}.
 
-\subsection*{Estimation of model parameters and structures from synthetic experimental data.}
-Model parameters were estimated by minimizing the difference between simulations and synthetic experimental data (squared residual):
+\subsection*{Estimation of model parameters from experimental data.}
+Model parameters were estimated by minimizing the difference between simulations and experimental thrombin measurements (squared residual):
 \begin{equation}\label{eqn:objective-function}
@@
 value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$. The outer summation is respect to
-time, while the inner summation is with respect to state. We approximated a realistic model identification scenario, assuming noisy experimental data, 
-limited sampling resolution (approximately 20 minutes per sample) and a limited number of measurable metabolites.
-
-We minimized the model residual using particle swarm optimization (PSO) \citep{PSO}.
+time, while the inner summation is with respect to state. We minimized the model residual using Particle swarm optimization (PSO) \citep{PSO}.
 PSO uses a \textit{swarming} metaheuristic to explore parameter spaces. 
@@
 parameters ($\mathcal{K}\times{1}$). In thus study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$. The quality of parameter
-estimates was measured using two criteria, goodness of fit (model residual) and angle between the estimated parameter vector $\mathbf{k}_{j}$ and the true parameter set $\mathbf{k}^{*}$:
+estimates was measured using two criteria, goodness of fit (model residual) and angle between the estimated parameter vector $\mathbf{k}_{j}$ and the best parameter set $\mathbf{k}^{*}$:
 \begin{equation}
@@
 and the angle between $\mathbf{k}_{j}$ and the true parameter set $\mathbf{k}^{*}$ would be equal to zero.
-
-We modified our PSO implementation to simultaneously search over kinetic parameters and putative model control structures.
-In the combined case, each particle potentially carried a different model realization in addition to a different kinetic parameter vector. 
-We kept the update rules the same (along with the update parameters). 
-Thus, each particle competed on the basis of goodness of fit, which allowed different model structures to contribute to the overall behavior of the swarm. 
-We considered five possible model structures (A through E), where network A was the correct formulation (used to generate the synthetic data). 
-We considered a population N = 100 particles, where each particle in the swarm was assigned a model structure, and a random parameter vector.
-The PSO algorithm, model equations, and the objective function were encoded and solved in the Octave programming language \citep{Octave:2014}.
+The particle swarm optimization routine was implemented in the Python programming language. 
+All plots were made using the Matplotlib module of Python \citep{Matplotlib}.
 
 \section*{Acknowledgements}
-This study was supported by an award from the National Science Foundation (MCB \#1411715) 
-and the Army Research Office (ARO \#59155-LS).
+This study was supported by an award from the Army Research Office (ARO \#59155-LS).
 
@@
 \centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-1-Networks.pdf}
-\caption{Proof of concept cell-free metabolic networks considered in this study. Substrate $S$ is converted to products P$_{1}$ and P$_{2}$ through a series of chemical conversions
+\includegraphics[width=1.0\textwidth]{./figs/Figure-1-Network.pdf}
+\caption{Proof of concept cell-free metabolic networks considered in this study. Substrate $S$ is converted to products $P_{1}$ and $P_{2}$ through a series of chemical conversions
 catalyzed by enzyme(s) $E_{j}$. The activity of the pathway enzymes is subject to both positive and negative allosteric regulation. }\label{fig-networks}
@@
 \includegraphics[width=1.0\textwidth]{./figs/Figure-2-ControlSchematic.pdf}
-\caption{Schematic of rule-based allosteric enzyme activity control laws. 
-Traditional enzyme kinetic expressions, e.g., Michaelis–Menten or multiple saturation kinetics, are multiplied by an enzyme activity control variable $0 \leq v_{j} \leq 1 $. 
+\caption{Schematic of rule based effective control laws. 
+Traditional enzyme kinetic expressions e.g., Michaelis–Menten or multiple saturation kinetics are multiplied by an enzyme activity control variable $0 \leq v_{j} \leq 1 $. 
 Control variables are functions of many possible regulatory factors encoded by arbitrary functions of the form $0\leq f_{j}\left(\mathcal{Z}\right)\leq 1$.
@@
 \centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-3-EnzymeKinetics.pdf}
-\caption{Kinetics of simple transformations in the presence of activation and inhibition. 
-\textbf{A}:The conversion of substrate $S$ to product $P$ by enzyme $E$ was activated by $S$. 
-For a fixed control gain parameter $k_{control}$, the reaction rate approached a step for increasing control order $N$. 
-\textbf{B}:The conversion of substrate $S$ to product $P$ by enzyme $E$ with inhibitor $I$. 
-For a fixed control gain parameter $k_{control}$, the reaction rate approximated non-competitive inhibition for increasing control order $N$. 
-}\label{fig-kinetics-simulations}
+\includegraphics[width=1.0\textwidth]{./figs/Figure-3-DIFF_FII_NO_APC.pdf}
+\caption{Reduced order coagulation model training simulations.
+Reduced order coagulation model parameters were estimated using particle swarm optimization (PSO) with and without the protein C pathway as a function of prothrombin. 
+Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data. 
+The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation).
+(A,B,C) training results for 150\%, 100\% and 50\% of physiological prothrombin levels in the absence of the protein C pathway.
+Thrombin generation was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
+All factors and control proteins were at their physiological concentration unless others denoted. 
+The experimental training data was reproduced from the study of Butenas et al. \citep{Butenas:1999aa}.}\label{fig-diff_fII_NO_APC}
 \end{figure}
@@
 \centering
-\includegraphics[width=0.50\textwidth]{./figs/Figure-4-OnOffSimulations.pdf}
-\caption{ON/OFF control simulations for network A and network B for an ensemble of kinetic parameter sets versus time (N = 100). 
-For each case, N = 100 simulations were conducted using kinetic and initial conditions generated randomly from a hypothetical true parameter set. 
-The gray area represents $\pm$ one standard deviation surrounding the mean. 
-Control parameters were fixed during the ensemble calculations.
-\textbf{A}: End product P$_{1}$ abundance versus time for Network A. 
-The abundance of P$_{1}$ decreased with end product inhibition of $E_{1}$ activity (Control-ON) versus the no inhibition case (Control-OFF). 
-\textbf{B}: End product P$_{2}$ abundance versus time for Network B. Inhibition of branch point $E_{6}$ by end product P$_{1}$ decreased P$_{2}$ abundance (Control-ON) versus the
-no inhibition case (Control-OFF).
-\textbf{C}: End product P$_{1}$ abundance versus time for Network A. 
-Inhibition of branch point $E_{6}$ by end product P$_{1}$ decreased P$_{1}$ abundance (Control-ON) versus the no inhibition case (Control-OFF).
-}\label{fig-onoff-simulations}
+\includegraphics[width=1.0\textwidth]{./figs/Figure-4-DIFF_FII_APC.pdf}
+\caption{Reduced order coagulation model training simulations.
+Reduced order coagulation model parameters were estimated using particle swarm optimization (PSO) with and without the protein C pathway as a function of prothrombin. 
+Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data.
+The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation).
+(A,B,C) training results for 150\%, 100\% and 50\% of physiological prothrombin levels in the presence of the protein C pathway. 
+Only APC pathway parameters were allowed to vary in the simulations on the right.
+Thrombin generation was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
+All factors and control proteins were at their physiological concentration unless others denoted.
+The experimental training data was reproduced from the study of Butenas et al. \citep{Butenas:1999aa}.}\label{fig-diff_fII_APC}
 \end{figure}
@@
 \centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-5-ParameterFit.pdf}
-\caption{Parameter estimation from synthetic data for the same and mismatched allosteric control logic using particle swarm optimization (PSO). 
-Synthetic experimental data was generated from a hypothetical parameter set using Network A, 
-where substrate $S$, end product P$_{1}$ and intermediate M$_5$ were sampled approximately every 20 minutes. 
-For cases $\textbf{A,B}$ 20 particles were initialized with randomized parameters and allowed to search for 300 iterations. 
-\textbf{A,B}: PSO estimated an ensemble of parameters sets (N = 20) consistent with the synthetic experimental data assuming the correct
-enzymatic and control connectivity starting from randomized initial parameters.
-\textbf{C,D}: In the presence of control mismatch (Network B control policy simulated with Network A kinetic parameters) 
-the ensemble of models did not describe the synthetic data. 
-}\label{fig-parameter-fit}
+\includegraphics[width=1.0\textwidth]{./figs/Figure-5-FII_ATIII.pdf}
+\caption{Reduced order coagulation model predictions versus experimental data for normal coagulation. 
+The reduced order coagulation model parameter estimates were tested against data not used during model training.
+Simulations of different levels of prothrombin and ATIII were compared with experimental data in the absence of the protein C pathway. 
+Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data.
+The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation). 
+(A,B,C,D) prediction results for (FII,ATIII): (50\%,150\%), (100\%, 100\%), (125\%, 75\%) and (150\%, 50\%) of physiological prothrombin and ATIII levels in the absence of the protein C pathway. 
+Thrombin generation was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
+All factors and control proteins were at their physiological concentration unless others denoted.
+The experimental validation data was reproduced from the study of Butenas et al. \citep{Butenas:1999aa}}\label{fig-fII_ATIII}
 \end{figure}
@@
 \centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-6-AlternativeNetworks.pdf}
-\caption{Schematic of the alternative allosteric control programs used in the structural particle swarm computation. 
-Each network had the same enzymatic connectivity, initial conditions and kinetic parameters, 
-but alternative feedback control structures for the first enzyme in the pathway.}\label{fig-alternative-networks}
+\includegraphics[width=1.0\textwidth]{./figs/Figure-6-Hemophilia.pdf}
+\caption{Reduced order coagulation model predictions versus experimental data with and without FVIII and FIX.
+The reduced order coagulation model parameter estimates were tested against data not used during model training. 
+Simulations of normal thrombin formation with ATIII and the protein C pathway were compared with thrombin formation in the absence of fVIII and fIX.
+Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data.
+The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation). 
+(A,B) prediction results for normal thrombin generation and thrombin generation in hemophilia. 
+All factors and control proteins were at their physiological concentration unless others noted.
+The experimental validation data was reproduced from the study of Allen et al. \citep{ALLEN2006}.}\label{fig-hemophilia}
 \end{figure}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{./figs/Figure-7-ControlSearch.pdf}
-\caption{Combined control and kinetic parameter search using modified particle swarm optimization (PSO). 
-A population of N = 100 particles was initialized with randomized kinetic parameters and one of five possible control configurations (Network A - E).
-Simulation error was minimized for a synthetic data set 
-($S$, end product P$_{1}$ and intermediate M$_5$ sampled approximately every 20 min) generated using Network A.
-\textbf{A}: Simulation error versus parameter set angle for N = 100 particles biased toward the correct regulatory program (A,B,C,D,E) = (40\%, 10\%, 20\%, 20\% and 10\%).
-\textbf{B}: Simulation error versus parameter set angle for N = 100 uniformly distributed particles (A,B,C,D,E) = (20\%, 20\%, 20\%, 20\% and 20\%).
-\textbf{C}: Simulation error versus parameter set angle for N = 100 negatively biased particles (A,B,C,D,E) = (10\%, 40\%, 10\%, 20\% and 20\%).
-Network A (the correct structure) was preferentially identified for positively and uniform biased particle distributions, but misidentified in the presence of
-a large incorrect bias. 
-}\label{fig-control-search}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth,height=0.6\textheight]{./figs/Figure-8-Flux.pdf}
-\caption{Metabolic flux and control variables as a function of network type and particle index at t = 100 min.
-The control variables governing $E_{1},E_{3}$ and $E_{6}$ activity and the scaled metabolic flux and were calculated for the positively, 
-uniformly and negatively biased particle swarms (N = 100). The particles from each swarm were sorted based upon simulation error (low to high error). 
-\textbf{A}: Model performance for the positively biased particle swarm as a function of particle index.
-\textbf{B}: Model performance for the uniformly biased particle swarm as a function of particle index.
-\textbf{C}: Model performance for the negatively biased particle swarm as a function of particle index. 
-Models with significant control mismatch showed distinct control and flux patterns versus those models with the correct or closely related control policies.
-In particular, models with the correct control policy showed stronger inhibition of $E_{1}$ activity, leading to decreased flux from S$\rightarrow$P$_{1}$.
-Conversely, models with significant mismatch had increased $E_{1}$ activity, leading to an altered flux distribution. 
-This is especially apparent in the negatively biased particle swarm. 
-}\label{fig-flux-pattern}
+\includegraphics[width=1.0\textwidth]{./figs/Figure-7-rFVIIa-Trigger.pdf}
+\caption{Reduced order coagulation model predictions of rFVIIa administration. 
+Simulations of thrombin formation in the presence of ATIII and the protein C pathway were conducted for a range of trigger values (1x - 50x) in the absence of fVIII and fIX.
+Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles.
+The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation). 
+All factors and control proteins were at their physiological concentration unless others noted.}\label{fig-rFVII}
 \end{figure}

### 2014-09-17 Paper_v1.tex
Results section taking shape, and some new figures

@@
 that our kinetic-rules based approach could reproduce a panel of thrombin generation data sets conducted at physiological factor and inhibitors concentrations.
-However, it was unclear if the reduced order model could predict new data sets, without changing kinetic or control parameters.
+However, it was unclear whether the reduced order model could predict new data, without updating the model parameters.
 
-We tested the predictive power of the reduced order coagulation model with data sets not used for model training. 
-Two data sets were used from validation, thrombin generation for different initial prothrombin and ATIII concentrations in the absence of the protein C pathway, 
-and normal versus hemophilia thrombin generation in the presence of the protein C pathway. 
-The hemophilia data was an especially difficult test as it was taken from a different laboratory and a completely different
-\textit{in~vitro} assay method involving unmodeled factors such as platelets. 
+\subsection*{Validation of the reduced order coagulation model.}
+We tested the predictive power of the reduced order coagulation model with validation data sets not used during model training. 
+Two validation data sets were used, thrombin generation for various prothrombin and ATIII concentrations with the protein C pathway, 
+and thrombin generation in normal versus hemophilic plasma in the presence of the protein C pathway. 
+Lastly, we compared the qualitative output of the model to rFVIIa addition in the presence of hemophilia.
+The hemophilia case was an especially difficult test as it was taken from a different study which used a plasma-based
+\textit{in~vitro} assay involving platelets instead of phospholipid vesicles (PCPS). 
 All kinetic and control parameters were fixed for the validation simulations. 
@@
 The reduced order model predicted the thrombin generation profile for ratios of prothrombin and ATIII in the absence of the protein C pathway (Fig. \ref{fig-fII_ATIII}).
-Model predictions for factors levels outside of the physiological range (fII,ATIII) = (50\%, 150\%) or (150\%, 50\%) were only qualitatively consistent with measured thrombin values.
-However, simulations closer to the physiological range (fII,ATIII) = (100\%, 100\%) or (125\%,75\%) tracked the measured thrombin values. 
+Simulations near the physiological range (fII,ATIII) = (100\%, 100\%) or (125\%,75\%) tracked the measured thrombin values (Fig. \ref{fig-fII_ATIII}B and C).
+On the other hand, predictions for factors levels outside of the physiological range (fII,ATIII) = (50\%, 150\%) or (150\%, 50\%), while qualitatively consistent with measured thrombin values, did show significant
+deviation from the measurements (Fig. \ref{fig-fII_ATIII}A and D).
 Likewise, simulations of thrombin generation in normal versus hemophilia (missing both fVIII and fIX) were consistent with measured thrombin values (Fig. \ref{fig-hemophilia}).
+We modeled the dependence of thrombin amplification on factor levels using a product rule ($\mathcal{Z} = fV\times fX\times fVIII\times fIX$), 
+which was then was integrated using a $\min$ integration rule into the control variable governing amplification. 
+Thus, in the absence of fVIII or fIX, the amplification control variable evaluated to zero, and the only thrombin produced was from initiation (Fig. \ref{fig-hemophilia}B).
+However, the decay of the thrombin signal was underpredicted in the normal case (Fig. \ref{fig-hemophilia}A), while the activated thrombin level was overpredicted 
+in hemophilia simulations, although thrombin generation was far less than normal (Fig. \ref{fig-hemophilia}B). 
+Taken together, the reduced order model performed well in the physiological range of factors, even with
+unmodeled components such as platelet activation in the hemophilia data set. Thus, our kinetic-rules framework predicted 
+the output of a physiologically important cascades such as coagulation, despite significant unmodeled components.
 
-\subsection*{Global sensitivity analysis of the reduced order coagulation model.}
-We conducted a variance based global sensitivity analysis to better understand which parameters were controlling the performance of the coagulation model. 
+The model ensemble predicted a direct correlation between thrombin generation and rFVIIa addition in hemophilia (Fig. \ref{fig-rFVII}).
+In the current proof of concept model, we cannot distinguish between different initiation sources, e.g., TF/FVIIa and rFVIIa, as we have only a single lumped initiation source (trigger).
+Thus, we simulated the addition of rFVIIa in hemophilia by removing fVIII and fIX from the model, and modulating the level of trigger.
+The baseline level of trigger is consistent with the previous hemophilia simulations, where the only thrombin being produced is from initiation mechanisms  (Fig. \ref{fig-rFVII}, 1$\times$~trigger).
+As we increased the strength of the trigger event, the thrombin peak time and the maximum value of thrombin increased (Fig. \ref{fig-rFVII}, 50$\times$~trigger). 
+However, as the trigger strength increased, the thrombin generated quickly decayed.
+For example, for the highest trigger strength simulated (50$\times$trigger), 95\% of the thrombin was gone by 20 min after initiation.
+
+We performed flux analysis to understand how the reduced order coagulation model balanced initiation, amplification and inhibition of thrombin formation for normal coagulation and hemophilia.
+Analysis of the reaction flux through the reduced order network for thrombin generation in normal, hemophilia and rFVIIa-treated hemophilia identified three distinct operational modes (Fig.
+\ref{fig-flux-analysis}).
+The reduced order network includes four lumped reactions, initiation, amplification, thrombin-induced APC generation and total thrombin inhibition (including both APC and ATIII action).
+Directly after the addition of a trigger (e.g., TF/FVIIa or rFVIIa), the lumped initiation flux was the largest for all three cases.
+However, within a few minutes enough thrombin was generated by the initiation mechanism to induce the amplification stage. 
+During amplification, thrombin catalyzes its own formation and inhibition by generating activated protein C (APC), a potent inhibitor of the coagulation cascade.
+For normal coagulation, amplification and thrombin inhibition are the dominate reactions by 6 min after initiation (Fig. \ref{fig-flux-analysis}, left). 
+After 10 min, the dominate reaction has shifted to thrombin inhibition (both ATIII and APC action). In the current proof-of-principle model, 
+APC inhibits upstream of amplification thus APC activity will slow the rate of thrombin formation directly from the initiation trigger. 
+On the other hand, ATIII inhibits thrombin directly, as well as upstream coagulation factors.
+In hemophilia (missing both fVIII and fIX), the amplification reaction does not occur and the only thrombin produced is from initiation (Fig. \ref{fig-flux-analysis}, center).
+Initiation is quickly inhibited by APC, and the thrombin level stabilizes and eventually decays because of ATIII activity. 
+Lastly, when 50$\times$rFVIIa is used to induce thrombin formation in hemophilia (absence of fVIII/fIX), 
+initiation mechanisms dominate for up to 6 min following initiation (Fig. \ref{fig-flux-analysis}, right). 
+However, similar to hemophilia alone, without amplification the thrombin signal is quickly extinguished by the combined action of ATIII and APC generated by thrombin.
+
+\subsection*{Sensitivity analysis of the ensemble of reduced order coagulation models.}
+We conducted a variance based global sensitivity analysis to better understand which parameters were controlling the performance of the reduced order coagulation model. 
 In particular, we calculated the sensitivity of two performance metrics, the time to maximum thrombin and the thrombin exposure (area under the thrombin curve) to
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure-8-Flux-rFVIIa-Trigger.pdf}
+\caption{Reaction flux distribution as a function of time for thrombin generation under normal (left), hemophilia (center) and rFVIIa treated hemophilia (right).
+Reaction flux was calculated for each particle at $T = 0,4,6,8,10,12,14$ min after the initiation of coagulation. 
+Blue colors denote low flux values red colors denote high flux values. }\label{fig-flux-analysis}
+\end{figure}
+
+\clearpage
+
 % Supplemental figures -

### 2014-09-22 Paper_v1.tex
Fig 9 sensitivity

@@
 The model ensemble predicted a direct correlation between thrombin generation and rFVIIa addition in hemophilia (Fig. \ref{fig-rFVII}).
-In the current proof of concept model, we cannot distinguish between different initiation sources, e.g., TF/FVIIa and rFVIIa, as we have only a single lumped initiation source (trigger).
-Thus, we simulated the addition of rFVIIa in hemophilia by removing fVIII and fIX from the model, and modulating the level of trigger.
-The baseline level of trigger is consistent with the previous hemophilia simulations, where the only thrombin being produced is from initiation mechanisms  (Fig. \ref{fig-rFVII}, 1$\times$~trigger).
+In the current model, we cannot distinguish between different initiation sources, e.g., TF/FVIIa and rFVIIa, as we have only a single lumped initiation source (trigger).
+Thus, we simulated the addition of rFVIIa in hemophilia by removing fVIII and fIX from the model, and modulating the initial level of trigger.
+Simulations with a baseline level of trigger were consistent with the previous hemophilia simulations, where the only thrombin being produced was from initiation (Fig.
+\ref{fig-rFVII}, 1$\times$~trigger).
 As we increased the strength of the trigger event, the thrombin peak time and the maximum value of thrombin increased (Fig. \ref{fig-rFVII}, 50$\times$~trigger). 
@@
 For example, for the highest trigger strength simulated (50$\times$trigger), 95\% of the thrombin was gone by 20 min after initiation. 
-
-We performed flux analysis to understand how the reduced order coagulation model balanced initiation, amplification and inhibition of thrombin formation for normal coagulation and hemophilia.
+We performed flux analysis to understand how the reduced order coagulation model balanced initiation, amplification and inhibition of thrombin formation for normal coagulation and hemophilia.util..
 Analysis of the reaction flux through the reduced order network for thrombin generation in normal, hemophilia and rFVIIa-treated hemophilia identified three distinct operational modes (Fig.
@@
 \subsection*{Sensitivity analysis of the ensemble of reduced order coagulation models.}
-We conducted a variance based global sensitivity analysis to better understand which parameters were controlling the performance of the reduced order coagulation model. 
-In particular, we calculated the sensitivity of two performance metrics, the time to maximum thrombin and the thrombin exposure (area under the thrombin curve) to
-changes in the model parameters (Fig. ZZ).
+We conducted a global sensitivity analysis to determine which parameters controlled the performance of the reduced order coagulation model. 
+We calculated the sensitivity of two performance metrics, the time to maximum thrombin and the thrombin exposure (area under the thrombin curve) with respect to
+changes in parameter values (Fig. \ref{fig-sensitivity}).
 
@@
 \section*{Discussion}
-Controlling hemophilia and more generally hemorrhage has been a driving force to understand the molecular basis of blood coagulation.
+Controlling hemorrhage has been a driving force to understand the molecular basis of blood coagulation.
 Replacement therapy improves bleeding times for most hemophilia patients. However, repeated fVIII/fIX administration can lead to neutralizing inhibitors. 
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure-9-Sensitivity.pdf}
+\caption{Global sensitivity of thrombin peak time and thrombin exposure. }\label{fig-sensitivity}
+\end{figure}
+
+\clearpage
+
 % Supplemental figures -

### 2014-09-29 Paper_v1.tex
Results complete

@@
 \section*{Introduction}
+Developing mathematical models of the biochemical networks important in human disease is a significant facet of systems biology.
+Modeling approaches differ in their degree of mechanistic detail, and the choice of approach is often determined by prior system knowledge \citep{2012_kholodenko_kolch_SciSig}. 
+The ability of ordinary differential equation (ODE) modeling to capture dynamics has made it one of the most common tools for studying signaling systems. 
+However, ODEs come with difficult (and sometimes impossible) to solve parameter identification problems. 
+For example, Gadkar et al., showed that even with near-perfect information and high frequency sampling, 
+it was often impossible to identify (to within a specified uncertainty) all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
+This results highlights the perhaps under-appreciated role that experimental design plays in generating the best training and validation data sets for model identification \citep{Apgar:2010fk}.
+Detailed ODE models also require significant mechanistic knowledge of the underlying biology. 
+However, as Bailey suggested more than a decade ago, achieving qualitative and quantitative understanding of complex biological systems should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}.
+Since Bailey's complex biology with no parameters hypothesis, Sethna and coworkers showed that model behavior and predictive ability were sensitive to only a few parameter combinations, a characteristic seemingly universal to multi-parameter signaling models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
+Thus, reasonable model predictions could be possible with limited parameter information. 
+Taking advantage of this property of model behavior, we developed sloppy techniques for parameter identification using ensembles of deterministic models. 
+Pareto optimal ensemble techniques (POETs) incorporate principles of competing objectives into a multi-objective optimization framework, enabling the exploration of kinetic parameter space while accounting for uncertainty and potential conflicts in the experimental training data \citep{2010_song_varner_BiotechJ}.
+We have proposed that the sloppy behavior of biological networks may also be a source of cell-to-cell \citep{2011_lequieu_chakrabarti_nayak_varner_PLoSCompBio} or even patient-to-patient heterogeneity \citep{2010_luan_varner_MolBioSys}.
+Recently, Bayesian techniques of parameter identification have also been used to explore cell-to-cell heterogeneity  \citep{2011_kalita_brasier_JBC,2011_hasenauer_allgower_BMCBioinfo}. 
+Thus, a population of cells could be viewed as a dynamic ensemble of networks as the operational biochemical pathways are often context-specific \citep{2012_creixell_linding_NatBiotech}.
+
 The human coagulation system is an archetype proteolytic cascade regulated by both positive and negative feedback loops.
 The biology of coagulation is well studied \cite{ROBERTS1998,MANN1999,BUTENAS2002,FURIE2004}. 
-Coagulation is mediated by a family serine proteases, called factors and a key group of blood cells, called platelets. 
+Coagulation is mediated by a family proteases, called factors and a key group of blood cells, called platelets. 
 The activation of coagulation factors and platelets, which are normally inactive in the circulation, requires a trigger event.
@@
 
-Modeling approaches differ in their degree of mechanistic detail, and the choice of approach is often determined by prior system knowledge \citep{2012_kholodenko_kolch_SciSig}. 
-The ability of ODE modeling to capture dynamics has made it one of the most common tools for studying signaling systems. 
-However, dynamics and ODEs come at the expense of difficult (and often impossible) to solve model parameter identification problems. 
-For example, Gadkar et al., showed that even with near-perfect information and high frequency sampling, 
-it was often impossible to identify (to within a specified uncertainty) all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
-This reality highlights the perhaps under-appreciated role that experimental design could play in generating the best training and validation data sets for model identification \citep{Apgar:2010fk}.
-Moreover, detailed ODE models require significant mechanistic knowledge of the underlying biology. 
-However, as Bailey suggested more than a decade ago, achieving qualitative and quantitative understanding of complex biological systems should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}.
-Since Bailey's complex biology with no parameters hypothesis, Sethna and coworkers showed that model behavior and predictive ability were sensitive to only a few parameter combinations, a characteristic seemingly universal to multi-parameter signaling models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
-Thus, reasonable model predictions could be possible with limited parameter information. 
-Taking advantage of this property of model behavior, we developed sloppy techniques for parameter identification using ensembles of deterministic models. 
-Pareto optimal ensemble techniques (POETs) incorporate principles of competing objectives into a multi-objective optimization framework, enabling the exploration of kinetic parameter space while accounting for uncertainty and potential conflicts in the experimental training data \citep{2010_song_varner_BiotechJ}.
-We have proposed that the sloppy behavior of biological networks may also be a source of cell-to-cell \citep{2011_lequieu_chakrabarti_nayak_varner_PLoSCompBio} or even patient-to-patient heterogeneity \citep{2010_luan_varner_MolBioSys}.
-Recently, Bayesian techniques of parameter identification have also been used to explore cell-to-cell heterogeneity  \citep{2011_kalita_brasier_JBC,2011_hasenauer_allgower_BMCBioinfo}. 
-Thus, a population of cells could be viewed as a dynamic ensemble of networks as the operational biochemical pathways are often context-specific \citep{2012_creixell_linding_NatBiotech}.
-
 In this study, we constructed a reduced order effective model of thrombin generation in normal and hemophilic blood.
@@
 
-\subsection*{Formulation and properties of reduced order effective models.}
-Reduced order effective models are [FINISH ME].
+%However, components could also correspond to non-modeled influences, categorial variables or other abstract quantities.
 
-\clearpage
+\subsection*{Formulation of reduced order coagulation models.}
+We developed a reduced order coagulation model to test our hybrid kinetic modeling approach (Fig. \ref{fig-network}).
+A trigger event initiates thrombin formation (FIIa) from prothrombin (fII) through a lumped initiation step.
+This step loosely represents the initial fluid phase activation of thrombin by activated FXa.
+Once activated, thrombin catalyzes its own formation (amplification step), and inhibition via the conversion of protein C to activated protein C (APC).
+APC and tissue factor pathway inhibitor (TFPI) inhibit initiation, while antithrombin III (ATIII) directly inhibits amplification. 
+All initiation and inhibition processes, as well as the dependence of amplification upon other coagulation factors, was approximated using our rule-based approach (Fig. \ref{fig-control-schematic}). 
+Individual regulatory contributions to the activity of pathway enzymes were integrated into control coefficients using an integration rule.  
+These control coefficients then modified the rates of model processes at each time step.
+Hill-like transfer functions $0 \leq f\left(\mathcal{Z}\right) \leq 1$ quantified the influence of components upon a target process. 
+Components were either individual inhibitor or activator levels or some function of levels, e.g., the product of factor levels. 
+In this study, $\mathcal{Z}$ corresponded to the abundance of individual inhibitors or activators, 
+with the exception of the dependence of amplification upon specific coagulation factors (modeled as the product of factors). 
+When a process was potentially sensitive to multiple inputs, logical integration rules were used to select which transfer functions influenced the process at any given time. 
+In our proof of concept model, we used a winner takes all strategy; the maximum or minimum transfer function was selected at any given time step. 
+However, other integration rules are certainly possible. 
+Taken together, while the reduced order coagulation model encodes significant biological complexity, it is highly compact, consisting of only five differential equations. 
+Thus, it will serve as an excellent proof of principle example to study the reduction of a highly complex human subsystem.
 
-\subsection*{Identification of reduced order coagulation model parameters using particle swarm optimization.}
+\subsection*{Identification of model parameters using particle swarm optimization.}
 A critical challenge for any dynamic model is the estimation of kinetic parameters. 
@@
 \subsection*{Sensitivity analysis of the ensemble of reduced order coagulation models.}
-We conducted a global sensitivity analysis to determine which parameters controlled the performance of the reduced order coagulation model. 
-We calculated the sensitivity of two performance metrics, the time to maximum thrombin and the thrombin exposure (area under the thrombin curve) with respect to
-changes in parameter values (Fig. \ref{fig-sensitivity}).
+We conducted a global sensitivity analysis to estimate which parameters controlled the performance of the reduced order coagulation model. 
+We calculated the sensitivity of the time to maximum thrombin (peak time) and the thrombin exposure (area under the thrombin curve) 
+for different levels of prothrombin, and protein C (Fig. \ref{fig-sensitivity}). 
+Globally, 41\% of the model parameters shifted in importance between the $(\%fII,\%PC) = (50\%, 0\%)$ and $(150\%,100\%)$ cases for the peak thrombin time (Fig. \ref{fig-sensitivity}A). 
+The majority of these shifts were involved increased prothrombin and the protein C pathway, while only
+5\% were directly associated with increased prothrombin alone. The rate constant for thrombin amplification was the most important parameter controlling the
+peak thrombin time. While this parameter was differentially important for different prothrombin levels, and in the presence or absence of 
+the activated protein C pathway, it's sensitivity was consistently above all other parameters in the model. The saturation constant governing thrombin amplification was
+the second most important parameter, followed by the initiation control gain parameter. Other important parameters influencing the thrombin peak time included the control gain for
+activated protein C formation, as well as the rate constant controlling ATIII inhibition of thrombin activity. 
+On the other hand, only 27\% of the model parameters were differentially sensitive between the (\%fII,\%PC) (50\%, 0\%) and (150\%,100\%) cases 
+for the thrombin exposure (Fig. \ref{fig-sensitivity}B). Of these parameters, all of the shifts were associated the interplay between thrombin formation and the protein C pathway. 
+The rate constant controlling ATIII inhibition was the most important parameter controlling the thrombin exposure.
+While this parameter was less important in the presence of the protein C pathway for 150\% prothrombin levels, it was significantly above all other parameters. 
+Similar to the peak time, for 150\% prothrombin, the control gain for activated protein C formation was differentially important along with the rate constant controlling
+amplification. However, the influence of the amplification parameter was much less in the thrombin exposure case versus peak time.
 
@@
 Thus, while the results presented here may be a valuable first step, more studies are required.
+
+While the performance of the proof of principle reduced order model was impressive, there are several questions that should be explored further.
+[FINISH]
 
@@
 \includegraphics[width=1.0\textwidth]{./figs/Figure-1-Network.pdf}
-\caption{Proof of concept cell-free metabolic networks considered in this study. Substrate $S$ is converted to products $P_{1}$ and $P_{2}$ through a series of chemical conversions
-catalyzed by enzyme(s) $E_{j}$. The activity of the pathway enzymes is subject to both positive and negative allosteric regulation. }\label{fig-networks}
+\caption{Schematic of the connectivity of the reduced order coagulation model. A trigger compound e.g., TF/FVIIa initiates thrombin production (FIIa) from prothrombin (fII).
+Once activated, thrombin catalyzes its own activation (amplification step), as well as its own inhibition via the conversion of protein C to activated protein C (APC).
+APC and tissue factor pathway inhibitor (TFPI) inhibit initiation, while antithrombin III (ATIII) directly inhibits thrombin. 
+All inhibition steps, and trigger-induced initiation were modeled using a rule-based approach. 
+Likewise, the dependence of amplification on other coagulation factors was also modeled using a rule-based approach. }\label{fig-network}
 \end{figure}
@@
 Control variables are functions of many possible regulatory factors encoded by arbitrary functions of the form $0\leq f_{j}\left(\mathcal{Z}\right)\leq 1$.
-At each simulation time step, the $v_{j}$ variables are calculated by evaluating integration rules such as the max or min of the set of factors $f_{1},\hdots$ 
+At each simulation time step, the $v_{j}$ variables are calculated by evaluating integration rules such as the max or min of the set of factors $f_{1},\hdots,f_{n}$ 
 influencing the activity of enzyme $E_{j}$. }\label{fig-control-schematic}
@@
 Reaction flux was calculated for each particle at $T = 0,4,6,8,10,12,14$ min after the initiation of coagulation. 
-Blue colors denote low flux values red colors denote high flux values. }\label{fig-flux-analysis}
+Reaction fluxes were calculated for each particle in the parameter ensemble (N = 20).
+Blue colors denote low flux values while red colors denote high flux values.}\label{fig-flux-analysis}
 \end{figure}
@@
 \includegraphics[width=1.0\textwidth]{./figs/Figure-9-Sensitivity.pdf}
-\caption{Global sensitivity of thrombin peak time and thrombin exposure. }\label{fig-sensitivity}
+\caption{Global sensitivity analysis of the reduced order coagulation model with with respect to the model parameters.
+A: Sensitivity analysis of the thrombin peak time for different prothrombin levels (150\%,100\% and 50\% of the physiological value) as a function of activated protein C. 
+B: Sensitivity analysis of the thrombin exposure for different prothrombin levels (150\%,100\% and 50\% of the physiological value) as a function of activated protein C. 
+Points denote the mean total sensitivity value, while the area around each point denotes the uncertainty in the sensitivity value. 
+The gray dashed line denotes the 45$^{o}$ degree diagonal, if sensitivity values are equal for different conditions they will lie on the diagonal.  
+Sensitivity values significantly above or below the diagonal indicate differentially important model parameters.
+}\label{fig-sensitivity}
 \end{figure}

### 2014-09-30 Paper_v1.tex
Updated introduction

@@
 \section*{Introduction}
-Developing mathematical models of the biochemical networks important in human disease is a significant facet of systems biology.
-Modeling approaches differ in their degree of mechanistic detail, and the choice of approach is often determined by prior system knowledge \citep{2012_kholodenko_kolch_SciSig}. 
-The ability of ordinary differential equation (ODE) modeling to capture dynamics has made it one of the most common tools for studying signaling systems. 
-However, ODEs come with difficult (and sometimes impossible) to solve parameter identification problems. 
+Developing mathematical models of biochemical networks is a significant facet of systems biology.
+Modeling approaches differ in their degree of detail, and the choice of approach is often determined by prior system knowledge \citep{2012_kholodenko_kolch_SciSig}. 
+The ability of ordinary differential equations (ODEs) to capture dynamics has made it one of the most common tools for studying signaling systems. 
+However, ODEs come with difficult (and sometimes impossible) parameter identification problems. 
 For example, Gadkar et al., showed that even with near-perfect information and high frequency sampling, 
 it was often impossible to identify (to within a specified uncertainty) all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
-This results highlights the perhaps under-appreciated role that experimental design plays in generating the best training and validation data sets for model identification \citep{Apgar:2010fk}.
-Detailed ODE models also require significant mechanistic knowledge of the underlying biology. 
-However, as Bailey suggested more than a decade ago, achieving qualitative and quantitative understanding of complex biological systems should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}.
-Since Bailey's complex biology with no parameters hypothesis, Sethna and coworkers showed that model behavior and predictive ability were sensitive to only a few parameter combinations, a characteristic seemingly universal to multi-parameter signaling models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
-Thus, reasonable model predictions could be possible with limited parameter information. 
-Taking advantage of this property of model behavior, we developed sloppy techniques for parameter identification using ensembles of deterministic models. 
-Pareto optimal ensemble techniques (POETs) incorporate principles of competing objectives into a multi-objective optimization framework, enabling the exploration of kinetic parameter space while accounting for uncertainty and potential conflicts in the experimental training data \citep{2010_song_varner_BiotechJ}.
-We have proposed that the sloppy behavior of biological networks may also be a source of cell-to-cell \citep{2011_lequieu_chakrabarti_nayak_varner_PLoSCompBio} or even patient-to-patient heterogeneity \citep{2010_luan_varner_MolBioSys}.
-Recently, Bayesian techniques of parameter identification have also been used to explore cell-to-cell heterogeneity  \citep{2011_kalita_brasier_JBC,2011_hasenauer_allgower_BMCBioinfo}. 
-Thus, a population of cells could be viewed as a dynamic ensemble of networks as the operational biochemical pathways are often context-specific \citep{2012_creixell_linding_NatBiotech}.
+This result highlights the perhaps under-appreciated role that experimental design could play in generating the best training and validation data sets for model identification \citep{Apgar:2010fk}.
+ODE models also require significant mechanistic knowledge of the underlying biology. 
+However, Bailey suggested more than a decade ago, that achieving qualitative or even quantitative understanding of biological systems 
+should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}. 
+Since Bailey's complex biology with no parameters hypothesis, Sethna and coworkers showed that model behavior and predictive ability are typically 
+sensitive to only a few parameter combinations, a characteristic seemingly universal to multi-parameter signaling models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
+Alternatively, data-driven approaches \citep{Cirit:2012kx} or logical model formulations \citep{Morris:2011ys} 
+are emerging paradigms that constrain model structure (and hence the identifiability challenge) by the availability of training and validation data. 
+These are interesting techniques because the data naturally informs model structure, which can be expanded as more data becomes available. 
+Thus, while there are many possible strategies to simulate biochemical networks, ODE models remain a common choice despite
+significant identification challenges.
 
@@
 resulting in the formation of the \emph{tenase} and \emph{prothrombinase} complexes on the surface of activated platelets.
-These complexes amplify the early coagulation signal by activating downstream and upstream coagulation factors.  
-Termination occurs after prothrombin is consumed or activated thrombin is neutralized by inhibitors such as activated protein C (APC) or antithrombin III (ATIII).
-
-Malfunctions in coagulation can have serious or potentially even fatal consequences.
-For example, aggressive clotting is involved with Coronary Artery Diseases (CADs), which collectively account for 38\% of all deaths in North America \cite{HANSSON2005}.
-Coagulation management during surgery can also be challenging, particularly because of the increasing clinical use of antithrombotic drugs \cite{Tanaka:2009wo}.   
-Insufficient coagulation due to genetic disorders such as hemophilia can also result in recurrent bleeding. 
-The coagulation factors VIII (fVIII) and IX (fIX) are deficient in Hemophilia A and B, respectively \cite{COOPER1994,MANNUCCI2001,MITCHELL2008}.
-People with mild hemophilia have 5-40\% of the normal clotting factor levels while severe hemophiliacs have $<$1\% \cite{MITCHELL2008}.
-Hemophilia can be controlled with regular infusions of the deficient clotting factors. 
-However, clotting factor replacement sometimes leads to the formation of \emph{in-vivo} fVIII and fIX inhibitors \cite{TOMOKIYO2003}.
-Activated Prothrombin Complex Concentrates (aPCCs) from pooled plasma have been used to treat hemophilia. 
-However, aPCCs can have safety issues \cite{CHAVIN1988,ALLEN2006,SCHNEIDERMAN2007}. 
-Alternatively, recombinant factor VIIa (rFVIIa) has been used to treat bleeding disorders \cite{Hedner:2008rm,Talbot:2009kk} including
-hemophilia with and without factor VIII/IX inhibitors \cite{Lusher:1998bs,Shapiro:1998sh,Shapiro:2000pr,Shapiro:2008px}.
-However, rFVIIa requires frequent administration (every 2-3 hr) because of its short half-life in the circulation. 
-Many questions also remain about its mechanism \cite{KOMIYAMA1990,PAYNE1996,MONROE1997,HOFFMAN1998,BUTENAS2002a,ROBERTS2004,ROBERTS2006,ALLEN2007}
-and effective dose range \cite{TOMOKIYO2003}.
-Thus, despite its wide adoption, the utility of rFVIIa in trauma-associated hemorrhage remains controversial \cite{Duchesne:2008hw}.
+Tenese amplifies the early coagulation signal by activating FXa, while prothrombinase, a complex of FVa-FXa, 
+directly converts prothrombin to thrombin. There are several control points in the cascade that promote 
+or inhibit thrombin formation. Tissue Factor Pathway Inhibitor (TFPI) downregulates FXa formation catalyzed by TF/FVIIa, while antithrombin III (ATIII) neutralizes 
+several of the serine proteases generated during coagulation.  
+Thrombin itself also inadvertently plays a role in its own inhibition; thrombin, through interaction with thrombomodulin and endothelial cell protein C receptor (EPCR),
+converts protein C to activated protein C (APC) which attenuates the coagulation response by the proteolytic cleavage of fV/FVa and fVIII/FVIIIa. 
+Thus, termination occurs after prothrombin is consumed or activated thrombin is neutralized by inhibitors such as activated protein C (APC) or antithrombin III (ATIII).
 
@@
 Once activated, thrombin catalyzes its own formation (amplification step), and inhibition via the conversion of protein C to activated protein C (APC).
-APC and tissue factor pathway inhibitor (TFPI) inhibit initiation, while antithrombin III (ATIII) directly inhibits amplification. 
+APC and tissue factor pathway inhibitor (TFPI) inhibit initiation, while antithrombin III (ATIII) inhibits amplification. 
 All initiation and inhibition processes, as well as the dependence of amplification upon other coagulation factors, was approximated using our rule-based approach (Fig. \ref{fig-control-schematic}). 
@@
 
-\subsection*{Sensitivity analysis of the ensemble of reduced order coagulation models.}
-We conducted a global sensitivity analysis to estimate which parameters controlled the performance of the reduced order coagulation model. 
+\clearpage
+
+\subsection*{Sensitivity analysis of the reduced order coagulation model.}
+We conducted a global sensitivity analysis to estimate which parameters controlled the performance of the reduced order model. 
 We calculated the sensitivity of the time to maximum thrombin (peak time) and the thrombin exposure (area under the thrombin curve) 
 for different levels of prothrombin, and protein C (Fig. \ref{fig-sensitivity}). 
-Globally, 41\% of the model parameters shifted in importance between the $(\%fII,\%PC) = (50\%, 0\%)$ and $(150\%,100\%)$ cases for the peak thrombin time (Fig. \ref{fig-sensitivity}A). 
+Globally, 41\% of the parameters shifted in importance between the (fII,PC) = (50\%, 0\%) and (150\%,100\%) cases for the peak thrombin time (Fig. \ref{fig-sensitivity}A). 
 The majority of these shifts were involved increased prothrombin and the protein C pathway, while only
@@
 The rate constant controlling ATIII inhibition was the most important parameter controlling the thrombin exposure.
-While this parameter was less important in the presence of the protein C pathway for 150\% prothrombin levels, it was significantly above all other parameters. 
+While this parameter was less important in the presence of protein C for 150\% prothrombin levels, it was significantly above all other parameters. 
 Similar to the peak time, for 150\% prothrombin, the control gain for activated protein C formation was differentially important along with the rate constant controlling
@@
 Thus, while the results presented here may be a valuable first step, more studies are required.
+
+Malfunctions in coagulation can have serious or potentially even fatal consequences.
+For example, aggressive clotting is involved with Coronary Artery Diseases (CADs), which collectively account for 38\% of all deaths in North America \cite{HANSSON2005}.
+Coagulation management during surgery can also be challenging, particularly because of the increasing clinical use of antithrombotic drugs \cite{Tanaka:2009wo}.   
+Insufficient coagulation due to genetic disorders such as hemophilia can also result in recurrent bleeding. 
+The coagulation factors VIII (fVIII) and IX (fIX) are deficient in Hemophilia A and B, respectively \cite{COOPER1994,MANNUCCI2001,MITCHELL2008}.
+People with mild hemophilia have 5-40\% of the normal clotting factor levels while severe hemophiliacs have $<$1\% \cite{MITCHELL2008}.
+Hemophilia can be controlled with regular infusions of the deficient clotting factors. 
+However, clotting factor replacement sometimes leads to the formation of \emph{in-vivo} fVIII and fIX inhibitors \cite{TOMOKIYO2003}.
+Activated Prothrombin Complex Concentrates (aPCCs) from pooled plasma have been used to treat hemophilia. 
+However, aPCCs can have safety issues \cite{CHAVIN1988,ALLEN2006,SCHNEIDERMAN2007}. 
+Alternatively, recombinant factor VIIa (rFVIIa) has been used to treat bleeding disorders \cite{Hedner:2008rm,Talbot:2009kk} including
+hemophilia with and without factor VIII/IX inhibitors \cite{Lusher:1998bs,Shapiro:1998sh,Shapiro:2000pr,Shapiro:2008px}.
+However, rFVIIa requires frequent administration (every 2-3 hr) because of its short half-life in the circulation. 
+Many questions also remain about its mechanism \cite{KOMIYAMA1990,PAYNE1996,MONROE1997,HOFFMAN1998,BUTENAS2002a,ROBERTS2004,ROBERTS2006,ALLEN2007}
+and effective dose range \cite{TOMOKIYO2003}.
+Thus, despite its wide adoption, the utility of rFVIIa in trauma-associated hemorrhage remains controversial \cite{Duchesne:2008hw}.
 

### 2014-10-01 Paper_v1.tex
Updated material and methods

@@
 The quantities $j^{+}$ and $j^{-}$ denoted the sets of activating and inhibitory factors for enzyme $j$. 
-If an enzyme had no allosteric factors, we set $v_{j} = 1$.
+If a process has no modifying factors, we set $v_{j} = 1$.
 There are many possible functional forms for $0\leq f_{ij}\left(\mathcal{Z}\right)\leq 1$. 
@@
 The $\kappa_{ij}$ parameter was species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).
+Applying the general framework to the reduced coagulation network resulted in five ordinary differential equations:
+\begin{eqnarray}
+	\frac{dx_{1}}{dt} & =& -\left(r_{init}v_{init} + r_{amp}v_{amp}\right)\\
+	\frac{dx_{2}}{dt} &=& r_{amp}v_{amp} + r_{init}v_{init} - \left(r_{inh,ATIII}v_{inh,ATIII} + r_{inh,APC}v_{inh,APC}\right)\\
+	\frac{dx_{3}}{dt} &=& -r_{apc}v_{apc} \\
+	\frac{dx_{4}}{dt} &=& r_{apc}v_{apc} \\
+	\frac{dx_{5}}{dt} & = & -r_{inh,ATIII}v_{inh,ATIII}
+\end{eqnarray}where $\mathbf{x} = \left(fII,FIIa,PC,APC,ATIII\right)^{T}$.
 The model equations were encoded using the Python programming language and solved using the ODEINT routine of the SciPy module \citep{SCIPY}.

### 2014-10-02 Paper_v1.tex
New network fig, mm equations and intro

@@
 
+%
+%Thus, while there are many possible strategies to simulate biochemical networks, ODE models remain a common choice despite
+%significant identification challenges.
+%This result highlights the perhaps under-appreciated role that experimental design could play in generating the best training and validation data sets for model identification \citep{Apgar:2010fk}.
+
 \section*{Introduction}
 Developing mathematical models of biochemical networks is a significant facet of systems biology.
-Modeling approaches differ in their degree of detail, and the choice of approach is often determined by prior system knowledge \citep{2012_kholodenko_kolch_SciSig}. 
-The ability of ordinary differential equations (ODEs) to capture dynamics has made it one of the most common tools for studying signaling systems. 
-However, ODEs come with difficult (and sometimes impossible) parameter identification problems. 
+Modeling approaches differ in their degree of detail. 
+The choice of approach is often determined by prior system knowledge, and model requirements \citep{2012_kholodenko_kolch_SciSig}. 
+Ordinary differential equation (ODE) models are a common tool for modeling biochemical systems because of their ability to capture dynamics. 
+However, ODEs often come with difficult (and sometimes impossible) parameter identification problems. 
 For example, Gadkar et al., showed that even with near-perfect information and high frequency sampling, 
 it was often impossible to identify (to within a specified uncertainty) all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
-This result highlights the perhaps under-appreciated role that experimental design could play in generating the best training and validation data sets for model identification \citep{Apgar:2010fk}.
-ODE models also require significant mechanistic knowledge of the underlying biology. 
+ODE models can also require significant mechanistic knowledge of the underlying biology. 
 However, Bailey suggested more than a decade ago, that achieving qualitative or even quantitative understanding of biological systems 
 should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}. 
-Since Bailey's complex biology with no parameters hypothesis, Sethna and coworkers showed that model behavior and predictive ability are typically 
+Since Bailey's complex biology with no parameters hypothesis, Sethna and coworkers showed that model behavior and predictive ability is typically 
 sensitive to only a few parameter combinations, a characteristic seemingly universal to multi-parameter signaling models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
+Thus, accurate model predictions may be possible even if parameters or mechanistic structure is not well understood.
 Alternatively, data-driven approaches \citep{Cirit:2012kx} or logical model formulations \citep{Morris:2011ys} 
@@
 These are interesting techniques because the data naturally informs model structure, which can be expanded as more data becomes available. 
-Thus, while there are many possible strategies to simulate biochemical networks, ODE models remain a common choice despite
-significant identification challenges.
+Thus, while not mechanistic, these models can simulate surprisingly rich system responses in the absence of mechanism. 
+A potential third approach could be a combination of ODEs and logical models, where mechanistic information could be used when available, 
+and missing or incomplete knowledge encoded using causal local rules.
 
-The human coagulation system is an archetype proteolytic cascade regulated by both positive and negative feedback loops.
-The biology of coagulation is well studied \cite{ROBERTS1998,MANN1999,BUTENAS2002,FURIE2004}. 
-Coagulation is mediated by a family proteases, called factors and a key group of blood cells, called platelets. 
+In this study, we explored a hybrid modeling approach involving both ODEs and logical rules on a well studied biochemical network, 
+the human coagulation system. The human coagulation system is an archetype proteolytic cascade involving both positive 
+and negative feedback \cite{Butenas:2002aa,Schenone:2004aa,Adams:2009aa}. 
+Coagulation is mediated by a family proteases in the circulation, called factors and a key group of blood cells, called platelets. 
 The activation of coagulation factors and platelets, which are normally inactive in the circulation, requires a trigger event.
@@
 These materials drive platelet localization and activation, coagulation factor activation and ultimately clot formation \cite{GIESEN1999}.  
-Once coagulation has been activated, two converging pathways activate the master protease thrombin. 
+Once coagulation has been activated, two converging pathways activate the master protease thrombin (FIIa) from prothrombin (fII). 
 The extrinsic cascade is generally believed to be the main mechanism of thrombinogenesis in the blood \cite{MANN1990,ROBERTS1998,MANN1999}.
-Thrombin generation consists of three phases, initiation, propagation and termination \cite{GOLDHABER2006, BRUMMEL2002}.  
-Initially thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa) \cite{BUTENAS2002a}. 
-Picomolar amounts of thrombin can then activate the cofactors factors V and VIII (fV and fVIII) and platelets, 
-resulting in the formation of the \emph{tenase} and \emph{prothrombinase} complexes on the surface of activated platelets.
+Thrombin generation consists of three phases, initiation, amplification and termination \cite{GOLDHABER2006, BRUMMEL2002}.  
+Initially, thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa) \cite{BUTENAS2002a}. 
+Picomolar amounts of thrombin then activate the cofactors factors V and VIII (fV and fVIII) and platelets, 
+resulting in the formation of the tenase and prothrombinase complexes on activated platelets.
 Tenese amplifies the early coagulation signal by activating FXa, while prothrombinase, a complex of FVa-FXa, 
-directly converts prothrombin to thrombin. There are several control points in the cascade that promote 
-or inhibit thrombin formation. Tissue Factor Pathway Inhibitor (TFPI) downregulates FXa formation catalyzed by TF/FVIIa, while antithrombin III (ATIII) neutralizes 
-several of the serine proteases generated during coagulation.  
-Thrombin itself also inadvertently plays a role in its own inhibition; thrombin, through interaction with thrombomodulin and endothelial cell protein C receptor (EPCR),
-converts protein C to activated protein C (APC) which attenuates the coagulation response by the proteolytic cleavage of fV/FVa and fVIII/FVIIIa. 
-Thus, termination occurs after prothrombin is consumed or activated thrombin is neutralized by inhibitors such as activated protein C (APC) or antithrombin III (ATIII).
+directly converts prothrombin to thrombin. There are several control points in the cascade that inhibit thrombin formation. 
+Tissue Factor Pathway Inhibitor (TFPI) downregulates FXa formation catalyzed by TF/FVIIa, while antithrombin III (ATIII) \
+neutralizes several of the serine proteases generated during coagulation.  
+Thrombin itself also inadvertently plays a role in its own inhibition; thrombin, through interaction with thrombomodulin, protein C and endothelial cell protein C receptor (EPCR),
+converts protein C to activated protein C (APC) which attenuates the coagulation response by proteolytic cleavage of fV/FVa and fVIII/FVIIIa. 
+Termination occurs after prothrombin is consumed or activated thrombin formation is neutralized by inhibitors such as activated protein C (APC) or antithrombin III (ATIII).
 
@@
 A critical challenge for any dynamic model is the estimation of kinetic parameters. 
-We estimated kinetic and control parameters simultaneously from \emph{in vitro} coagulation data sets with and without the protein C pathway. 
+We estimated kinetic and control parameters simultaneously from eight \emph{in vitro} time-series coagulation data sets with and without the protein C pathway. 
 The residual between model simulations and experimental measurements was minimized using particle swarm optimization (PSO). 
@@
 The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$. 
-If $\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$. 
-Conversely, if $\sigma_{ij}>0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates metabolite $i$ is not connected with reaction $j$. 
-Lastly, $\lambda_{i}$ denotes the scaled enzyme degradation constant. 
-The system material balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$.
+If $\sigma_{ij}>0$, species $i$ is produced by reaction $j$. 
+Conversely, if $\sigma_{ij}<0$, species $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates species $i$ is not connected with reaction $j$. 
+The system material balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$, which were specified by the experimental setup.
 
@@
 	\frac{dx_{1}}{dt} & =& -\left(r_{init}v_{init} + r_{amp}v_{amp}\right)\\
-	\frac{dx_{2}}{dt} &=& r_{amp}v_{amp} + r_{init}v_{init} - \left(r_{inh,ATIII}v_{inh,ATIII} + r_{inh,APC}v_{inh,APC}\right)\\
+	\frac{dx_{2}}{dt} &=& r_{amp}v_{amp} + r_{init}v_{init} - r_{inh,ATIII}v_{inh,ATIII}\\
 	\frac{dx_{3}}{dt} &=& -r_{apc}v_{apc} \\
@@
 \end{eqnarray}where $\mathbf{x} = \left(fII,FIIa,PC,APC,ATIII\right)^{T}$. 
+The terms $r_{*}v_{*}$ in the balance equations denote corrected kinetic expressions for initiation, amplification and inhibition processes. 
+The rate of initiation $\bar{r}_{init}$ was modeled as:
+\begin{equation}
+	\bar{r}_{init} = k_{init}\left(trigger\right)\frac{x_{1}}{K_{init,fII} + x_{1}}
+\end{equation}where $k_{init},~K_{init,fII}$ are the rate and saturation constants governing initiation, respectively. 
+The rate of initiation was modified by $v_{init}$, the control parameter governing initiation. 
+Initiation was sensitive to level of trigger (activator), and TFPI (inhibitor):
+\begin{equation}
+	v_{init} = \min\left(f^{-}_{init}\left(TFPI\right),f^{+}_{init}\left(trigger\right)\right)
+\end{equation}where the transfer functions $f$ took the form of Eqn \eqref{eqn:control-factor}.
+The rate of thrombin amplification was given by:
+\begin{equation}
+	\bar{r}_{amp} = k_{amp}\left(x_{2}\right)\frac{x_{1}}{K_{amp,fII}+x_{1}}
+\end{equation}where $k_{amp},~K_{amp,fII}$ denote the rate and saturation constants governing amplification, respectively. 
+The amplification control term, which modified amplification rate, was modeled as a combination of multiple inhibition terms and one activation term:
+\begin{equation}
+	v_{amp} = \min\left(f^{-}_{amp}\left(TFPI\right),f^{-}_{amp}\left(x_{4}\right),f^{+}_{amp}\left(\mathcal{Z}_{amp}\right)\right)
+\end{equation}where $\mathcal{Z}_{amp} = fV\times fX\times fVIII\times fIX$. Although, $f^{+}_{amp}\left(\mathcal{Z}_{amp}\right)$ is an activating term, 
+we included it in the min integration rule; the factors in $\mathcal{Z}_{amp}$ were essential for amplification (if any of these factors was missing
+the amplification reaction would not occur). Thus, the factors in $\mathcal{Z}_{amp}$ were required components, a classification
+that we implemented by the min selection rule.
+The rate activated protein C formation was given by:
+\begin{equation}
+	\bar{r}_{apc} = k_{APC,formation}\left(TM\right)\frac{x_{3}}{K_{formation,PC} + x_{3}}
+\end{equation}where $k_{APC,formation},~K_{formation,PC}$ denote the rate and saturation constants governing activated protein C formation, respectively and $TM$ denotes
+the thrombomodulin abundance. We modeled the control term which governed APC formation as a single thrombin dependent activation term:
+\begin{equation}
+	v_{apc} = \max\left(f_{apc}^{+}\left(x_{2}\right)\right)
+\end{equation}
+Lastly, we included direct irreversible inhibition of FIIa by ATIII:
+\begin{equation}
+	\bar{r}_{inh,ATIII} = k_{ATIII,inhibition}\left(x_{5}x_{2}^{\gamma}\right)
+\end{equation}where $\gamma$ was estimated to be $\gamma = 1.26$. For ATIII inhibition of FIIa, the control variables $v_{inh,ATIII}$ was taken to be unity.
 The model equations were encoded using the Python programming language and solved using the ODEINT routine of the SciPy module \citep{SCIPY}.
+The model files can be downloaded from http://www.varnerlab.org.
 
@@
 parameters ($\mathcal{K}\times{1}$). In thus study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$. The quality of parameter
-estimates was measured using two criteria, goodness of fit (model residual) and angle between the estimated parameter vector $\mathbf{k}_{j}$ and the best parameter set $\mathbf{k}^{*}$:
-\begin{equation}
-	\alpha_{j} = cos^{-1}\left(\frac{\mathbf{k}_{j}\cdot{\mathbf{k^{*}}}}{\norm{\mathbf{k}_{j}}\norm{\mathbf{k^{*}}}}\right)
-\end{equation}If the candidate parameter set $\mathbf{k}_{j}$ were perfect, the residual between the model and synthetic data 
-and the angle between $\mathbf{k}_{j}$ and the true parameter set $\mathbf{k}^{*}$ would be equal to zero.
-The particle swarm optimization routine was implemented in the Python programming language. 
+estimates was measured using goodness of fit (model residual). The particle swarm optimization routine was implemented in the Python programming language. 
 All plots were made using the Matplotlib module of Python \citep{Matplotlib}.
+The optimization code can be downloaded from http://www.varnerlab.org.
+
+\subsection*{Global sensitivity analysis of model performance.}
+We conducted a global sensitivity analysis to estimate which parameters controlled the performance of the reduced order model. 
+We computed the total sensitivity index of each parameter relative to two performance objectives, the peak thrombin time and the area under the thrombin curve 
+(thrombin exposure). [FINISH~ME].
 
@@
 Once activated, thrombin catalyzes its own activation (amplification step), as well as its own inhibition via the conversion of protein C to activated protein C (APC).
-APC and tissue factor pathway inhibitor (TFPI) inhibit initiation, while antithrombin III (ATIII) directly inhibits thrombin. 
+APC and tissue factor pathway inhibitor (TFPI) inhibit initiation and amplification, while antithrombin III (ATIII) directly inhibits thrombin. 
 All inhibition steps, and trigger-induced initiation were modeled using a rule-based approach. 
-Likewise, the dependence of amplification on other coagulation factors was also modeled using a rule-based approach. }\label{fig-network}
+Likewise, the dependence of amplification on other coagulation factors was also modeled using a rule-based approach. 
+The abundance of the highlighted species (in the dashed boxes) was governed by an ordinary differential equation. 
+All other species were assumed to be constant.}\label{fig-network}
 \end{figure}

### 2014-10-07 Paper_v1.tex
First complete draft of the introduction

@@
 Developing mathematical models of biochemical networks is a significant facet of systems biology.
-Modeling approaches differ in their degree of detail. 
-The choice of approach is often determined by prior system knowledge, and model requirements \citep{2012_kholodenko_kolch_SciSig}. 
+Modeling approaches differ in their degree of detail, and the choice of approach is often determined by prior knowledge, or model requirements \citep{2012_kholodenko_kolch_SciSig}. 
 Ordinary differential equation (ODE) models are a common tool for modeling biochemical systems because of their ability to capture dynamics. 
@@
 should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}. 
-Since Bailey's complex biology with no parameters hypothesis, Sethna and coworkers showed that model behavior and predictive ability is typically 
+Since Bailey's complex biology with no parameters hypothesis, Sethna and coworkers have shown that model behavior and predictive ability is typically 
 sensitive to only a few parameter combinations, a characteristic seemingly universal to multi-parameter signaling models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
-Thus, accurate model predictions may be possible even if parameters or mechanistic structure is not well understood.
+Thus, accurate model predictions from a sloppy model may be possible, even if the parameters or mechanistic structure is not well understood, 
+if the critical directions in parameter space are properly constrained. 
 Alternatively, data-driven approaches \citep{Cirit:2012kx} or logical model formulations \citep{Morris:2011ys} 
@@
 These are interesting techniques because the data naturally informs model structure, which can be expanded as more data becomes available. 
-Thus, while not mechanistic, these models can simulate surprisingly rich system responses in the absence of mechanism. 
-A potential third approach could be a combination of ODEs and logical models, where mechanistic information could be used when available, 
+While not mechanistic, these models can simulate surprisingly rich system responses in the absence of mechanism [REFHERE]. 
+However, modeling dynamics using logical models is not straightforward. 
+Thus, a potential third approach could be a combination of ODEs and logical models, where mechanistic information was used when available, 
 and missing or incomplete knowledge encoded using causal local rules.
 
-In this study, we explored a hybrid modeling approach involving both ODEs and logical rules on a well studied biochemical network, 
-the human coagulation system. The human coagulation system is an archetype proteolytic cascade involving both positive 
+In this study, we developed a hybrid approach which combined ODEs and logical rules to model a well studied biochemical network, 
+the human coagulation system. Coagulation is an archetype proteolytic cascade involving both positive 
 and negative feedback \cite{Butenas:2002aa,Schenone:2004aa,Adams:2009aa}. 
 Coagulation is mediated by a family proteases in the circulation, called factors and a key group of blood cells, called platelets. 
-The activation of coagulation factors and platelets, which are normally inactive in the circulation, requires a trigger event.
-Trigger events such as vessel injury expose procoagulant materials like collagen, fibronectin, von Willebrand factor (vWF) or tissue factor (TF).
-These materials drive platelet localization and activation, coagulation factor activation and ultimately clot formation \cite{GIESEN1999}.  
-Once coagulation has been activated, two converging pathways activate the master protease thrombin (FIIa) from prothrombin (fII). 
+The central process in coagulation is the conversion of prothrombin (fII), an inactive coagulation factor, to the master protease thrombin (FIIa).
+Thrombin generation involves three phases, initiation, amplification and termination \cite{GOLDHABER2006, BRUMMEL2002}.
+Initiation requires a trigger event, for example vessel injury, which leads to the activation of factor VII (FVIIa).
+Two converging pathways, the extrinsic and intrinsic cascades, then process and amplify this initial coagulation signal. 
 The extrinsic cascade is generally believed to be the main mechanism of thrombinogenesis in the blood \cite{MANN1990,ROBERTS1998,MANN1999}.
-Thrombin generation consists of three phases, initiation, amplification and termination \cite{GOLDHABER2006, BRUMMEL2002}.  
-Initially, thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa) \cite{BUTENAS2002a}. 
+Initially, thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa), which itself has been activated by FVIIa \cite{BUTENAS2002a}. 
 Picomolar amounts of thrombin then activate the cofactors factors V and VIII (fV and fVIII) and platelets, 
-resulting in the formation of the tenase and prothrombinase complexes on activated platelets.
-Tenese amplifies the early coagulation signal by activating FXa, while prothrombinase, a complex of FVa-FXa, 
-directly converts prothrombin to thrombin. There are several control points in the cascade that inhibit thrombin formation. 
-Tissue Factor Pathway Inhibitor (TFPI) downregulates FXa formation catalyzed by TF/FVIIa, while antithrombin III (ATIII) \
-neutralizes several of the serine proteases generated during coagulation.  
+leading to the formation of the tenase and prothrombinase complexes on activated platelets.
+These complexes amplify the early coagulation signal by further activating FXa, and directly converting prothrombin to thrombin. 
+There are several control points in the cascade that inhibit thrombin formation, and eventually terminate thrombin generation. 
+Tissue Factor Pathway Inhibitor (TFPI) inhibits FXa formation catalyzed by TF/FVIIa, while antithrombin III (ATIII)
+neutralizes several of the proteases generated during coagulation, including thrombin.  
 Thrombin itself also inadvertently plays a role in its own inhibition; thrombin, through interaction with thrombomodulin, protein C and endothelial cell protein C receptor (EPCR),
 converts protein C to activated protein C (APC) which attenuates the coagulation response by proteolytic cleavage of fV/FVa and fVIII/FVIIIa. 
-Termination occurs after prothrombin is consumed or activated thrombin formation is neutralized by inhibitors such as activated protein C (APC) or antithrombin III (ATIII).
+Termination occurs after either prothrombin is consumed, or thrombin formation is neutralized by inhibitors such as activated protein C (APC) or antithrombin III (ATIII).
 
-In this study, we constructed a reduced order effective model of thrombin generation in normal and hemophilic blood.
-We used this model to understand how physiological coagulation was altered by prothrombin levels, the activated protein C pathway and the level of ATIII.  
-Previous models of coagulation have been formulated as systems of nonlinear ordinary differential equations, often using mass-action kinetics to describe the rates of biochemical conversions [REFHERE].
-While this approach has proven highly effective, it does require significant experimental data for model identification and validation.
-For example, a previous coagulation model from our laboratory, which described 193 proteins and protein complexes interconnected by 301 biochemical interactions, 
-contained 467 unknown parameters (301 kinetic parameters and 166 initial conditions) [REFHERE]. 
-The key innovation of our approach is the seamless integration of simple effective rules encoding complex regulatory motifs with traditional kinetic pathway modeling. 
-This integration allows the description of complex regulatory interactions, such as time-dependent allosteric regulation of enzyme activity, in the absence of specific mechanistic information. 
-The regulatory rules are easy to understand, easy to formulate and do not rely on overarching theoretical abstractions or restrictive assumptions. 
-We tested our approach by modeling the time evolution of the human coagulation cascade.
+There are many mathematical models of the human coagulation cascade in the literature.
+Previous coagulation models have typically been formulated as systems of nonlinear ordinary differential equations, 
+often using mass-action kinetics to describe the rates of biochemical conversions [REFHERE].
+Previous ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys}, 
+were built upon the pioneering studies of Jones and Mann \citep{Jones:1994aa}, 
+and later Hockin \emph{et al.} \citep{Hockin:2002aa}, who developed and subsequently refined a highly mechanistic model of the extrinsic coagulation cascade. 
+Other laboratories have also expanded upon Hockin \emph{et al.}, for example Chatterjee \emph{et al.,} augmented the Hockin model to include 
+the intrinsic pathway, and the role of thrombin in clot formation \citep{Chatterjee:2010aa}. 
+More recently, modeling studies have explored other biochemical or biophysical aspects of coagulation, such as the biochemical networks inside platelets \citep{Stalker:2013aa}, 
+multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and the transport of material inside growing clots \citep{Voronov:2013aa}.
+However, most of these previous studies share the common feature of a mechanism based description of coagulation. 
+Such a description is possible because blood, while enormously complex, can be systematically interrogated. 
+However, other biochemical networks, such as the intracellular networks which mediate cancer, are more difficult to probe and model.  
+Towards this need, we have proposed a hybrid approach which combined ODEs and logical rules that can be used to model 
+biochemical processes where mechanistic information is difficult or impossible to obtain. 
+We have used this approach to model the coagulation cascade, because of the wealth of training and validation data available for this system.
+Our hypothesis is that coagulation represents a best case, if our approach fails in this proof-of-concept study, it will never be acceptable in model difficult cases.
+We proposed a reduced order coagulation model that consisted of five differential equations augmented with several logical rules.
+We identified the reduced order coagulation model parameters from in vitro extrinsic data sets of thrombin generation initiated by TF/FVIIa in the presence of ATIII, with and without the protein C pathway.
+We then validated the model using thrombin data sets from multiple laboratories, for both normal and hemophilia coagulation.
+The reduced order hybrid approach produced a surprisingly predictive coagulation model, suggesting this framework could potentially be used to model other biochemical networks
+important to human health.
 

### 2014-10-07 Paper_v1.tex
First complete draft of the introduction

@@
 However, ODEs often come with difficult (and sometimes impossible) parameter identification problems. 
-For example, Gadkar et al., showed that even with near-perfect information and high frequency sampling, 
-it was often impossible to identify (to within a specified uncertainty) all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
-ODE models can also require significant mechanistic knowledge of the underlying biology. 
+For example, Gadkar et al., showed that even with near-perfect information, 
+it was often impossible to identify all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
 However, Bailey suggested more than a decade ago, that achieving qualitative or even quantitative understanding of biological systems 
@@
 Since Bailey's complex biology with no parameters hypothesis, Sethna and coworkers have shown that model behavior and predictive ability is typically 
-sensitive to only a few parameter combinations, a characteristic seemingly universal to multi-parameter signaling models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
-Thus, accurate model predictions from a sloppy model may be possible, even if the parameters or mechanistic structure is not well understood, 
-if the critical directions in parameter space are properly constrained. 
+sensitive to only a few parameter combinations, a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
+Thus, reasonable predictions from sloppy models may be possible, even if the parameters or structure is uncertain, 
+if a few critical directions in parameter space are properly constrained. 
+Typical ODE models can also require significant mechanistic information.
 Alternatively, data-driven approaches \citep{Cirit:2012kx} or logical model formulations \citep{Morris:2011ys} 
-are emerging paradigms that constrain model structure (and hence the identifiability challenge) by the availability of training and validation data. 
-These are interesting techniques because the data naturally informs model structure, which can be expanded as more data becomes available. 
-While not mechanistic, these models can simulate surprisingly rich system responses in the absence of mechanism [REFHERE]. 
+are emerging paradigms that constrain model structure by the availability of training and validation data. 
+These are interesting techniques because the data naturally informs model structure, which can be expanded as more data becomes available, and limits the 
+identification challenge. Although not mechanistic, these models can simulate surprisingly rich system responses in the absence of mechanism [REFHERE]. 
 However, modeling dynamics using logical models is not straightforward. 
@@
 
-There are many mathematical models of the human coagulation cascade in the literature.
-Previous coagulation models have typically been formulated as systems of nonlinear ordinary differential equations, 
+Coagulation models in the literature have typically been formulated as systems of nonlinear ordinary differential equations, 
 often using mass-action kinetics to describe the rates of biochemical conversions [REFHERE].
 Previous ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys}, 
-were built upon the pioneering studies of Jones and Mann \citep{Jones:1994aa}, 
-and later Hockin \emph{et al.} \citep{Hockin:2002aa}, who developed and subsequently refined a highly mechanistic model of the extrinsic coagulation cascade. 
-Other laboratories have also expanded upon Hockin \emph{et al.}, for example Chatterjee \emph{et al.,} augmented the Hockin model to include 
-the intrinsic pathway, and the role of thrombin in clot formation \citep{Chatterjee:2010aa}. 
-More recently, modeling studies have explored other biochemical or biophysical aspects of coagulation, such as the biochemical networks inside platelets \citep{Stalker:2013aa}, 
-multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and the transport of material inside growing clots \citep{Voronov:2013aa}.
-However, most of these previous studies share the common feature of a mechanism based description of coagulation. 
+were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, 
+and later Hockin \emph{et al.} \citep{Hockin:2002aa}, who developed and then refined a highly mechanistic model of extrinsic coagulation. 
+Other laboratories have also expanded upon Hockin \emph{et al.}, for example including the intrinsic pathway, and thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
+Other biochemical or biophysical aspects of coagulation have also been modeled, such as platelet biochemistry \citep{Stalker:2013aa}, 
+multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and transport inside growing clots \citep{Voronov:2013aa}.
+However, most of these previous studies were based upon detailed mechanistic knowledge. 
 Such a description is possible because blood, while enormously complex, can be systematically interrogated. 
-However, other biochemical networks, such as the intracellular networks which mediate cancer, are more difficult to probe and model.  
-Towards this need, we have proposed a hybrid approach which combined ODEs and logical rules that can be used to model 
-biochemical processes where mechanistic information is difficult or impossible to obtain. 
-We have used this approach to model the coagulation cascade, because of the wealth of training and validation data available for this system.
-Our hypothesis is that coagulation represents a best case, if our approach fails in this proof-of-concept study, it will never be acceptable in model difficult cases.
+However, other systems, such as the intracellular networks which mediate cancer, are more difficult to experimentally probe and model.  
+Towards this unmet need, we have formulated a hybrid modeling approach which combines ODEs and logical rules to model 
+biochemical processes for which mechanistic information is difficult or impossible to obtain. 
+We have used this approach to model the coagulation cascade because of the wealth of training and validation data available for this system. 
 We proposed a reduced order coagulation model that consisted of five differential equations augmented with several logical rules.
-We identified the reduced order coagulation model parameters from in vitro extrinsic data sets of thrombin generation initiated by TF/FVIIa in the presence of ATIII, with and without the protein C pathway.
-We then validated the model using thrombin data sets from multiple laboratories, for both normal and hemophilia coagulation.
-The reduced order hybrid approach produced a surprisingly predictive coagulation model, suggesting this framework could potentially be used to model other biochemical networks
-important to human health.
+We identified the reduced order coagulation model parameters from in vitro extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
+We then validated the model using thrombin data sets from multiple laboratories, for both normal and hemophilic coagulation.
+Flux and sensitivity analysis identified which parameter or parameters combinations were critical to model performance in several conditions. 
+The reduced order hybrid approach produced a surprisingly predictive coagulation model, suggesting this framework could potentially be used to model other biochemical networks important to human health.
 

### 2014-10-15 Paper_v1.tex
Working on discussion

@@
 Developing mathematical models of biochemical networks is a significant facet of systems biology.
-Modeling approaches differ in their degree of detail, and the choice of approach is often determined by prior knowledge, or model requirements \citep{2012_kholodenko_kolch_SciSig}. 
+Modeling approaches differ in their degree of detail, but the choice of approach is often determined by prior knowledge, or model requirements \citep{2012_kholodenko_kolch_SciSig}. 
 Ordinary differential equation (ODE) models are a common tool for modeling biochemical systems because of their ability to capture dynamics. 
-However, ODEs often come with difficult (and sometimes impossible) parameter identification problems. 
+However, ODE models typically come with difficult (or sometimes impossible) parameter identification problems. 
 For example, Gadkar et al., showed that even with near-perfect information, 
@@
 should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}. 
-Since Bailey's complex biology with no parameters hypothesis, Sethna and coworkers have shown that model behavior and predictive ability is typically 
+Since Bailey's complex biology with no parameters hypothesis, Sethna and coworkers showed that model performance is typically 
 sensitive to only a few parameter combinations, a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
-Thus, reasonable predictions from sloppy models may be possible, even if the parameters or structure is uncertain, 
-if a few critical directions in parameter space are properly constrained. 
-Typical ODE models can also require significant mechanistic information.
-Alternatively, data-driven approaches \citep{Cirit:2012kx} or logical model formulations \citep{Morris:2011ys} 
-are emerging paradigms that constrain model structure by the availability of training and validation data. 
-These are interesting techniques because the data naturally informs model structure, which can be expanded as more data becomes available, and limits the 
-identification challenge. Although not mechanistic, these models can simulate surprisingly rich system responses in the absence of mechanism [REFHERE]. 
+Thus, reasonable predictions from sloppy models may be possible, despite parametric uncertainty, if a few critical components are well understood. 
+Perhaps a more important issue is that ODE models also require significant mechanistic information, thereby limiting their utility in poorly understood systems.
+Data-driven \citep{Cirit:2012kx} or logical model approaches \citep{Morris:2011ys} 
+are emerging modeling paradigms that constrain model structure by the availability of experimental data. 
+These are interesting techniques because the data naturally informs model structure, thereby limiting the 
+identification challenge. Although not always mechanistic, these models can simulate surprisingly rich system responses in the absence of mechanism [REFHERE]. 
 However, modeling dynamics using logical models is not straightforward. 
@@
 were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, 
-and later Hockin \emph{et al.} \citep{Hockin:2002aa}, who developed and then refined a highly mechanistic model of extrinsic coagulation. 
+and later Hockin \emph{et al.} \citep{Hockin:2002aa}, who developed and then refined highly mechanistic models of extrinsic coagulation. 
 Other laboratories have also expanded upon Hockin \emph{et al.}, for example including the intrinsic pathway, and thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
@@
 \subsection*{Formulation of reduced order coagulation models.}
-We developed a reduced order coagulation model to test our hybrid kinetic modeling approach (Fig. \ref{fig-network}).
+We developed a reduced order coagulation model to test our hybrid kinetic rule based modeling approach (Fig. \ref{fig-network}).
 A trigger event initiates thrombin formation (FIIa) from prothrombin (fII) through a lumped initiation step.
@@
 (i) estimation of parameters associated with thrombin formation in the absence of the protein C pathway and (ii) estimation of parameters associated with the protein C pathway.
-Only those parameters associated with each subproblem were varied during the optimization procedure for that subproblem e.g., thrombin parameters were \textit{not} varied during the protein C subproblem.
-The PSO procedure was run for 20 generations for each subproblem, where each generation was 1200 iterations. The best particle from each generation was used to generate the particle population
-for the next generation. We rotated the subproblems, starting with subproblem 1 in the first generation.
+Only those parameters associated with each subproblem were varied during the optimization procedure for that subproblem e.g., thrombin parameters were \textit{not} varied during the protein C subproblem.The PSO procedure was run for 20 generations for each subproblem, where each generation was 1200 iterations.
+The best particle from each generation was used to generate the particle population for the next generation.
+We rotated the subproblems, starting with subproblem 1 in the first generation.
 
@@
 For example, for the highest trigger strength simulated (50$\times$trigger), 95\% of the thrombin was gone by 20 min after initiation. 
-We performed flux analysis to understand how the reduced order coagulation model balanced initiation, amplification and inhibition of thrombin formation for normal coagulation and hemophilia.util..
+We performed flux analysis to understand how the reduced order coagulation model balanced initiation, amplification and inhibition of thrombin formation for normal coagulation and hemophilia.
 Analysis of the reaction flux through the reduced order network for thrombin generation in normal, hemophilia and rFVIIa-treated hemophilia identified three distinct operational modes (Fig.
@@
 
+%Controlling hemorrhage has been a driving force to understand the molecular basis of blood coagulation.
+%Replacement therapy improves bleeding times for most hemophilia patients. However, repeated fVIII/fIX administration can lead to neutralizing inhibitors. 
+%Alternatively, rFVIIa has been used as a universal hemostatic agent to initiate clotting with and without fVIII or fIX inhibitors \cite{Hedner:2008rm}.
+%Despite its wide adoption, the mechanism and the utility of rFVIIa in trauma-associated hemorrhage remains controversial \cite{Duchesne:2008hw}. 
+%The model described the extrinsic activation of thrombin, in the presence of ATIII, with and without the protein C pathway.
+%The coagulation network architecture was based on literature and a previous mathematical model from our laboratory \cite{VARNER2007}.
+%Simulations of thrombin dynamics were independently validated using blood drawn from patients with coronary artery disease (reported here) and previous TF/FVIIa and FVIIa 
+%\emph{in-vitro} studies \cite{BUTENAS2002}.
+
+%Additionally, using sensitivity analysis on the ensemble of models, we identified 
+%structurally sensitive components of the coagulation architecture as a function of condition.
+%We also demonstrated that the ensemble robustly constrained model predictions of independent validation sets, despite having many poorly constrained parameters.
+%Thus, the model ensemble displayed \emph{sloppy} behavior similar to that observed by Sethna and coworkers for other signal transduction networks \cite{Gutenkunst2007}.  
+%By studying a family of models, we perhaps partially addressed the uncertainty stemming from the many poorly characterized model parameters.
+%However, many other factors could influence our results. For example, missing structural interactions or biophysical factors such as blood flow could play a large role.
+%Thus, while the results presented here may be a valuable first step, more studies are required.
+% \cite{ALLEN2007}.
+
+%We used this ensemble of parameters in all subsequent simulations to estimate the uncertainty in model simulations following from uncertainty in the model parameters.
+
 \section*{Discussion}
-Controlling hemorrhage has been a driving force to understand the molecular basis of blood coagulation.
-Replacement therapy improves bleeding times for most hemophilia patients. However, repeated fVIII/fIX administration can lead to neutralizing inhibitors. 
-Alternatively, rFVIIa has been used as a universal hemostatic agent to initiate clotting with and without fVIII or fIX inhibitors \cite{Hedner:2008rm}.
-Despite its wide adoption, the mechanism and the utility of rFVIIa in trauma-associated hemorrhage remains controversial \cite{Duchesne:2008hw}. 
-In this study, we formulated a model of the human coagulation cascade and studied thrombin formation in normal and rFVIIa treated and untreated
-hemophilic plasma. The model described 193 proteins or protein complexes connected by 301 interactions. 
-The coagulation network architecture was based on literature and a previous mathematical model from our laboratory \cite{VARNER2007}.
-Simulations of thrombin dynamics were independently validated using blood drawn from patients with coronary artery disease (reported here) and previous TF/FVIIa and FVIIa 
-\emph{in-vitro} studies \cite{BUTENAS2002}.
-While the human coagulation cascade is perhaps an ideal model system to develop network analysis tools, not all the model parameters were identifiable given the current training data.
-Instead, an \emph{ensemble} of models (N = 437) was estimated using nine cell-based coagulation training sets \cite{ALLEN2007}.
-We used this family of consistent coagulation models in all simulations instead of a single best-fit but uncertain model.
-Additionally, using sensitivity analysis on the ensemble of models, we identified 
-structurally sensitive components of the coagulation architecture as a function of condition.
-We also demonstrated that the ensemble robustly constrained model predictions of independent validation sets, despite having many poorly constrained parameters.
-Thus, the model ensemble displayed \emph{sloppy} behavior similar to that observed by Sethna and coworkers for other signal transduction networks \cite{Gutenkunst2007}.  
-By studying a family of models, we perhaps partially addressed the uncertainty stemming from the many poorly characterized model parameters.
-However, many other factors could influence our results. For example, missing structural interactions or biophysical factors such as blood flow could play a large role.
-Thus, while the results presented here may be a valuable first step, more studies are required.
+In this study, we developed a reduced order model of the human coagulation cascade.
+We chose coagulation because it is well studied, has a complex architecture, 
+and has an abundance of experimental data available for model identification and validation. 
+However, coagulation was just a proof-of concept; 
+the proposed framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
+or potentially even cell free metabolic systems. 
+The model consisted of five differential equations augmented with several logical rules describing 
+regulatory connections between model components, and unmodeled interactions in the network. 
+We estimated model parameters from in vitro extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
+To estimate parameters, the residual between model simulations and experimental measurements was minimized using particle swarm optimization (PSO). 
+However, not all of the model parameters were uniquely identifiable, given the training data.
+Instead, we estimated an ensemble of likely parameter sets (N = 20)
+from eight \emph{in vitro} time-series coagulation data sets with and without the protein C pathway.
+We validated the ensemble using thrombin data sets taken from multiple laboratories for a variety of experimental conditions not used during training. 
+The ensemble predicted thrombin trajectories for conditions not used for model training, including thrombin generation for normal and hemophilic coagulation 
+in the presence of platelets (a significant unmodeled component).
+We then used flux analysis to understand how the network operated in a variety of conditions, and global sensitivity analysis to identify which parameters controlled the 
+performance of the network. Taken together, the reduced order hybrid approach produced a surprisingly predictive model, suggesting this approach could potentially 
+be used to model other biochemical networks important to human health.
 
@@
 
-While the performance of the proof of principle reduced order model was impressive, there are several questions that should be explored further.
-[FINISH]
+Although the performance of the proof of principle coagulation model was impressive given its dimension, there are several mathematical and biological questions that can be explored further.
+First, the choice of max/min integration rules, or the particular form of the transfer functions could be generalized to include other rule types and functions. 
+Theoretically, an integration rule is simply a function whose domain is a set of transfer function inputs, and whose range is $[0,1]$.
+Thus, integration rules other than max/min could be used, such as the mean or the product, assuming the range of the transfer functions is always $[0,1]$.
+Alternative integration rules such the mean would have different properties which could influence model identification or performance. 
+For example, a mean or product integration rule would be differentiable, which allows derivative based optimization approaches to be used. 
+The particular form of the transfer function could also be explored. We choose a hill-like function because of its 
+prominence in the systems and synthetic biology community. 
+However the only mathematical requirement for a transfer function is to map a non-negative continuous or categorical variable into the 
+range $[0,1]$. Thus, many other function types are possible.
 

### 2014-10-15 Paper_v1.tex
Working on discussion

@@
 
-Although the performance of the proof of principle coagulation model was impressive given its dimension, there are several mathematical and biological questions that can be explored further.
+Although the performance of the proof of principle coagulation model was impressive given its size, there were several mathematical and biological questions that can be explored further.
 First, the choice of max/min integration rules, or the particular form of the transfer functions could be generalized to include other rule types and functions. 
@@
 prominence in the systems and synthetic biology community. 
-However the only mathematical requirement for a transfer function is to map a non-negative continuous or categorical variable into the 
-range $[0,1]$. Thus, many other function types are possible.
+However, the only mathematical requirement for a transfer function is to map a non-negative continuous or categorical variable into the 
+range $[0,1]$. Thus, many other function types e.g., tanh are also possible.
+There are several additional biological modules that can be added to the core model presented here. 
+First, we need to update the model to included thrombin induced platelet activation. 
+While we were able to capture thrombin generation data in the presence of platelets, 
+the initial shape of the activation curve and the time-scale of activation was not always consistent with the data.
+Platelets are activated by thrombin through the cleavage of the extracellular domain of protease-activated receptors (PARs) on the platelet surface.
+Next, we should add the intrinsic or contact thrombin activation pathway to the model. 
+The intrinsic coagulation pathway is triggered by contact activation of the plasma protease factor XII (fXII) by negatively charged surfaces.
+The activation of fXII is followed by proteolytic activation of factor XI (fXI) and the activation of downstream factors like fIX \cite{NAITO1991}.
+It has also been suggested that activated platelets can release polyphosphate which directly activates fXII \cite{Smith:2006zl}. 
+Thus, platelet and fXI activation may be integrated.
+Lastly, to make the model more clinically relevant, we should include the biochemical processes driving clot formation, and clot dissolution (fibrinolysis).
+Clot formation is driven by thrombin activity, while fibrinolysis is driven by plasmin activity, a key enzyme that cleaves fibrin (one of the main materials in a clot).
+Similar to coagulation, fibrinolysis is managed by several activating and inhibitory factors which control the balance between clot formation and dissolution.
 

### 2014-10-15 Paper_v1.tex
Working on discussion

@@
 The central process in coagulation is the conversion of prothrombin (fII), an inactive coagulation factor, to the master protease thrombin (FIIa).
-Thrombin generation involves three phases, initiation, amplification and termination \cite{GOLDHABER2006, BRUMMEL2002}.
+Thrombin generation involves three phases, initiation, amplification and termination \cite{GOLDHABER2006, Brummel:2002aa}.
 Initiation requires a trigger event, for example vessel injury, which leads to the activation of factor VII (FVIIa).
@@
 The extrinsic cascade is generally believed to be the main mechanism of thrombinogenesis in the blood \cite{MANN1990,ROBERTS1998,MANN1999}.
-Initially, thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa), which itself has been activated by FVIIa \cite{BUTENAS2002a}. 
+Initially, thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa), which itself has been activated by FVIIa \cite{Butenas:2002aa}. 
 Picomolar amounts of thrombin then activate the cofactors factors V and VIII (fV and fVIII) and platelets, 
@@
 
-\subsection*{Sensitivity analysis of the reduced order coagulation model.}
+\subsection*{Global sensitivity analysis of the reduced order coagulation model.}
 We conducted a global sensitivity analysis to estimate which parameters controlled the performance of the reduced order model. 
@@
 
-Malfunctions in coagulation can have serious or potentially even fatal consequences.
-For example, aggressive clotting is involved with Coronary Artery Diseases (CADs), which collectively account for 38\% of all deaths in North America \cite{HANSSON2005}.
+Malfunctions in coagulation can have potentially fatal consequences.
+For example, aggressive clotting involved with Coronary Artery Diseases (CADs), collectively accounts for 38\% of all deaths in North America \cite{HANSSON2005}.
 Coagulation management during surgery can also be challenging, particularly because of the increasing clinical use of antithrombotic drugs \cite{Tanaka:2009wo}.   
@@
 Hemophilia can be controlled with regular infusions of the deficient clotting factors. 
-However, clotting factor replacement sometimes leads to the formation of \emph{in-vivo} fVIII and fIX inhibitors \cite{TOMOKIYO2003}.
-Activated Prothrombin Complex Concentrates (aPCCs) from pooled plasma have been used to treat hemophilia. 
-However, aPCCs can have safety issues \cite{CHAVIN1988,ALLEN2006,SCHNEIDERMAN2007}. 
+However, clotting factor replacement sometimes leads to the formation of fVIII and fIX inhibitors \emph{in-vivo} \cite{TOMOKIYO2003}.
 Alternatively, recombinant factor VIIa (rFVIIa) has been used to treat bleeding disorders \cite{Hedner:2008rm,Talbot:2009kk} including
-hemophilia with and without factor VIII/IX inhibitors \cite{Lusher:1998bs,Shapiro:1998sh,Shapiro:2000pr,Shapiro:2008px}.
+hemophilia with and without factor VIII/IX inhibitors \cite{Shapiro:2008px}.
 However, rFVIIa requires frequent administration (every 2-3 hr) because of its short half-life in the circulation. 
-Many questions also remain about its mechanism \cite{KOMIYAMA1990,PAYNE1996,MONROE1997,HOFFMAN1998,BUTENAS2002a,ROBERTS2004,ROBERTS2006,ALLEN2007}
-and effective dose range \cite{TOMOKIYO2003}.
-Thus, despite its wide adoption, the utility of rFVIIa in trauma-associated hemorrhage remains controversial \cite{Duchesne:2008hw}.
+Many questions also remain about the mechanism of action of rFVIIa, its effective dose range \cite{TOMOKIYO2003}, and its overall utility 
+for the treatment trauma-associated hemorrhage \cite{Duchesne:2008hw}. In the current proof of concept model, we did not model rFVIIa directly. 
+Rather we modeled a broad class of trigger materials which initiated the extrinsic coagulation cascade. Since we identified the model using TF/FVIIa, inherent to
+our rFVIIa simulations was the presence of TF. However, even with this complication, 
+the model generated potentially useful insight into the rFVIIa mechanism of action, and its possible shortcomings especially for the treatment of hemophilia. 
+Addition of rFVIIa directly activated thrombin through the fluid phase initiation pathway. 
+However, in the presence of ATIII and the protein C pathway but without fVIII or fIX, 
+this initial thrombin signal was quickly extinguished by inhibition, as thrombin amplification could not occur. 
+As the dose of rFVIIa increased, the peak thrombin time decreased, eventually saturating.
+Butenas \emph{et al.,} performed an extensive study of rFVIIa induced thrombin generation in hemophilia \citep{Butenas:2002ab}.
 
@@
 First, the choice of max/min integration rules, or the particular form of the transfer functions could be generalized to include other rule types and functions. 
-Theoretically, an integration rule is simply a function whose domain is a set of transfer function inputs, and whose range is $[0,1]$.
-Thus, integration rules other than max/min could be used, such as the mean or the product, assuming the range of the transfer functions is always $[0,1]$.
+Theoretically, an integration rule is simply a function whose domain is a set of transfer function inputs, and whose range is $v\in[0,1]$.
+Thus, integration rules other than max/min could be used, such as the mean or the product, assuming the range of the transfer functions is always $f\in[0,1]$.
 Alternative integration rules such the mean would have different properties which could influence model identification or performance. 
@@
 prominence in the systems and synthetic biology community. 
-However, the only mathematical requirement for a transfer function is to map a non-negative continuous or categorical variable into the 
-range $[0,1]$. Thus, many other function types e.g., tanh are also possible.
+However, the only mathematical requirement for a transfer function is that it must map a non-negative continuous or categorical variable into the 
+range $f\in[0,1]$. Thus, many function types are possible.
 There are several additional biological modules that can be added to the core model presented here. 
@@
 It has also been suggested that activated platelets can release polyphosphate which directly activates fXII \cite{Smith:2006zl}. 
-Thus, platelet and fXI activation may be integrated.
-Lastly, to make the model more clinically relevant, we should include the biochemical processes driving clot formation, and clot dissolution (fibrinolysis).
+Lastly, to make the model more clinically relevant, we should include the biochemical processes responsible for clot formation, and clot dissolution (fibrinolysis).
 Clot formation is driven by thrombin activity, while fibrinolysis is driven by plasmin activity, a key enzyme that cleaves fibrin (one of the main materials in a clot).
 Similar to coagulation, fibrinolysis is managed by several activating and inhibitory factors which control the balance between clot formation and dissolution.
+Tissue plasminogen activator (t-PA) and urokinase activate plasmin, along with contact pathway factors such fXIa. On the other hand, thrombin activatable fibrinolysis inhibitor (TAFI)
+inhibits the degradation of fibrin by plasmin. Also similar to coagulation, there is considerable fibrinolysis and contact pathway data sets that can be used to train the model.
 

### 2014-10-21 Paper_v1.tex
Abstract and introduction

@@
 \section*{Abstract}
+In this study, we present a novel hybrid modeling approach which combines ordinary differential equation (ODE) modeling with logical rules to simulate an archetype biochemical network, 
+the human coagulation cascade.  
+The model consisted of five differential equations augmented with several logical rules describing 
+regulatory connections between model components, and unmodeled interactions in the network. 
+This formulation was more than an order of magnitude smaller than current coagulation models, 
+because many of the mechanistic details of coagulation were encoded as logical rules.
+We estimated an ensemble of likely model parameters (N = 20) from \textit{in~vitro} extrinsic coagulation data sets, with and without inhibitors, 
+by minimizing the residual between model simulations and experimental measurements using particle swarm optimization (PSO). 
+We then validated the model ensemble using thrombin data sets that were not used during training. 
+The ensemble predicted thrombin trajectories for conditions not used for model training, including thrombin generation for normal and hemophilic coagulation 
+in the presence of platelets (a significant unmodeled component).
+We then used flux analysis to understand how the network operated in a variety of conditions, and global sensitivity analysis to identify which parameters controlled the 
+performance of the network. Taken together, the hybrid approach produced a surprisingly predictive model given its small size, 
+suggesting the proposed hybrid framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
+or potentially even cell free metabolic systems.
 
-{\noindent \textbf{Keywords:}~Coagulation, Mathematical modeling, Systems Biology}
+\vspace{0.25in}
+{\noindent \textbf{Keywords:}~Blood coagulation, Mathematical modeling, Systems biology}
 
@@
 
+%We chose coagulation because it is well studied, has a complex architecture, 
+%and has an abundance of experimental data available for model identification and validation. 
+%However, coagulation was just a proof-of concept to validate our approach; 
+%
+
+%Developing mathematical modeling tools to interrogate the molecular basis of human disease is one of the grand challenges facing systems biology.
 %Developing computational tools to understand the molecular basis of human disease is a grand challenge facing systems biology.
@@
 Developing mathematical models of biochemical networks is a significant facet of systems biology.
-Modeling approaches differ in their degree of detail, but the choice of approach is often determined by prior knowledge, or model requirements \citep{2012_kholodenko_kolch_SciSig}. 
-Ordinary differential equation (ODE) models are a common tool for modeling biochemical systems because of their ability to capture dynamics. 
+Modeling approaches differ in their degree of detail, where the choice of approach is often determined by prior knowledge, or model requirements \citep{2012_kholodenko_kolch_SciSig}. 
+Ordinary differential equation (ODE) models are a common tool for modeling biochemical systems because of their ability to capture dynamics, and encode mechanism.
 However, ODE models typically come with difficult (or sometimes impossible) parameter identification problems. 
@@
 it was often impossible to identify all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
-However, Bailey suggested more than a decade ago, that achieving qualitative or even quantitative understanding of biological systems 
+However, it is not clear whether we actually need precise estimates for all model parameters. 
+Bailey suggested more than a decade ago, that achieving qualitative or even quantitative understanding of biological systems 
 should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}. 
-Since Bailey's complex biology with no parameters hypothesis, Sethna and coworkers showed that model performance is typically 
-sensitive to only a few parameter combinations, a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
-Thus, reasonable predictions from sloppy models may be possible, despite parametric uncertainty, if a few critical components are well understood. 
-Perhaps a more important issue is that ODE models also require significant mechanistic information, thereby limiting their utility in poorly understood systems.
-Data-driven \citep{Cirit:2012kx} or logical model approaches \citep{Morris:2011ys} 
-are emerging modeling paradigms that constrain model structure by the availability of experimental data. 
-These are interesting techniques because the data naturally informs model structure, thereby limiting the 
-identification challenge. Although not always mechanistic, these models can simulate surprisingly rich system responses in the absence of mechanism [REFHERE]. 
-However, modeling dynamics using logical models is not straightforward. 
-Thus, a potential third approach could be a combination of ODEs and logical models, where mechanistic information was used when available, 
-and missing or incomplete knowledge encoded using causal local rules.
+Since Bailey's complex biology with no parameters hypothesis, Sethna showed that model performance is typically 
+sensitive to only a few parameters, a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
+Thus, reasonable predictions may be possible, despite parametric uncertainty, if a few critical parameters are well defined. 
+Perhaps more importantly, ODE models also require significant mechanistic information, thereby limiting their utility in poorly understood systems.
+Toward this challenge, logical modeling is an emerging paradigm that encodes causal relationships using non-linear transfer functions \citep{Morris:2010aa}.
+These functions encode interactions between model components, without requiring mechanism.
+Despite their simplicity, logical models have captured rich system behaviors in the absence of mechanistic information \citep{Saez-Rodriguez:2011aa,Morris:2011ys,Morris:2012aa}. 
+However, modeling complex dynamics remains difficult with this class of model.
+Thus, there is an unmet need for a third approach which combines ODEs and logical models, where ODEs could encode mechanistic information, 
+while missing or incomplete mechanistic knowledge can be approximated using a logical approach.
 
-In this study, we developed a hybrid approach which combined ODEs and logical rules to model a well studied biochemical network, 
+In this study, we developed a hybrid approach which combines ODE modeling with logical rules to model a well studied biochemical network, 
 the human coagulation system. Coagulation is an archetype proteolytic cascade involving both positive 
@@
 
-Coagulation models in the literature have typically been formulated as systems of nonlinear ordinary differential equations, 
-often using mass-action kinetics to describe the rates of biochemical conversions [REFHERE].
+Coagulation models in the literature are typically formulated as systems of nonlinear ordinary differential equations, 
+using mass-action or other kinetics, to describe the rates of biochemical conversions \citep{Kuharsky:2001aa}.
 Previous ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys}, 
@@
 and has an abundance of experimental data available for model identification and validation. 
-However, coagulation was just a proof-of concept; 
-the proposed framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
+However, coagulation was just a proof-of concept validation of our approach; 
+the proposed hybrid framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
 or potentially even cell free metabolic systems. 
@@
 Malfunctions in coagulation can have potentially fatal consequences.
-For example, aggressive clotting involved with Coronary Artery Diseases (CADs), collectively accounts for 38\% of all deaths in North America \cite{HANSSON2005}.
-Coagulation management during surgery can also be challenging, particularly because of the increasing clinical use of antithrombotic drugs \cite{Tanaka:2009wo}.   
+Aggressive clotting involved with Coronary Artery Diseases (CADs), collectively accounts for 38\% of all deaths in North America \cite{HANSSON2005}.
+Coagulation management during surgery can also be challenging, particularly with the increasing clinical use of antithrombotic drugs \cite{Tanaka:2009wo}.   
 Insufficient coagulation due to genetic disorders such as hemophilia can also result in recurrent bleeding. 
@@
 hemophilia with and without factor VIII/IX inhibitors \cite{Shapiro:2008px}.
-However, rFVIIa requires frequent administration (every 2-3 hr) because of its short half-life in the circulation. 
-Many questions also remain about the mechanism of action of rFVIIa, its effective dose range \cite{TOMOKIYO2003}, and its overall utility 
-for the treatment trauma-associated hemorrhage \cite{Duchesne:2008hw}. In the current proof of concept model, we did not model rFVIIa directly. 
-Rather we modeled a broad class of trigger materials which initiated the extrinsic coagulation cascade. Since we identified the model using TF/FVIIa, inherent to
-our rFVIIa simulations was the presence of TF. However, even with this complication, 
+However, rFVIIa requires frequent administration (every 2-3 hr), and many questions remain about its mechanism of action, 
+its effective dose range \cite{TOMOKIYO2003}, and its overall utility for the treatment trauma-associated hemorrhage \cite{Duchesne:2008hw}. 
+In this study, we did not model rFVIIa induced coagulation directly. 
+Rather, we modeled a general trigger which initiated the extrinsic coagulation cascade. 
+Since we identified the model using TF/FVIIa, inherent to our rFVIIa simulations (and the rate constant governing initiation) 
+was the presence of TF. However, even with this complication, 
 the model generated potentially useful insight into the rFVIIa mechanism of action, and its possible shortcomings especially for the treatment of hemophilia. 
-Addition of rFVIIa directly activated thrombin through the fluid phase initiation pathway. 
-However, in the presence of ATIII and the protein C pathway but without fVIII or fIX, 
-this initial thrombin signal was quickly extinguished by inhibition, as thrombin amplification could not occur. 
-As the dose of rFVIIa increased, the peak thrombin time decreased, eventually saturating.
-Butenas \emph{et al.,} performed an extensive study of rFVIIa induced thrombin generation in hemophilia \citep{Butenas:2002ab}.
+The addition of rFVIIa directly activated thrombin through the initiation pathway. 
+However, no amplification of the thrombin signal occurred without fVIII or fIX. 
+Thus, the peak thrombin signal was lower than normal coagulation, the peak thrombin time was longer, and the thrombin signal was eventually inhibited by ATIII and the protein C pathway.
+However, as the dose of rFVIIa increased, the peak thrombin time decreased (eventually saturating around 100$\times$nominal trigger), and the peak thrombin value increased.
+For 100$\times$nominal trigger, the thrombin profile resembled normal coagulation.
+Butenas \emph{et al.,} performed an extensive in vitro study of rFVIIa induced thrombin generation under normal and hemophilic conditions \citep{Butenas:2002ab}.
+They found qualitatively similar trends, namely rFVIIa could restore normal coagulation (even in the absence of TF) for large enough rFVIIa doses, although rFVIIa induced coagulation 
+in hemophilia (even for large rFVIIa doses) lagged the normal profile. 
+These results suggest that rFVIIa administration alone
+might not be able to initiate normal coagulation in recurrent bleeding, unless the dosage is above a critical threshold. 
+Unfortunately, this threshold is likely patient specific, thus limiting potentially limiting the utility of rFVIIa administration.
 
@@
 \caption{Reduced order coagulation model predictions of rFVIIa administration. 
-Simulations of thrombin formation in the presence of ATIII and the protein C pathway were conducted for a range of trigger values (1x - 50x) in the absence of fVIII and fIX.
+A: Simulations of thrombin formation in the presence of ATIII and the protein C pathway were conducted for a range of trigger values (1x - 200x nominal) in the absence of fVIII and fIX.
+B: Comparison of thrombin generation for normal versus hemophilia for 10x nominal trigger.
 Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles.

### 2014-10-23 Paper_v1.tex
Near final draft

@@
 Perhaps more importantly, ODE models also require significant mechanistic information, thereby limiting their utility in poorly understood systems.
-Toward this challenge, logical modeling is an emerging paradigm that encodes causal relationships using non-linear transfer functions \citep{Morris:2010aa}.
-These functions encode interactions between model components, without requiring mechanism.
-Despite their simplicity, logical models have captured rich system behaviors in the absence of mechanistic information \citep{Saez-Rodriguez:2011aa,Morris:2011ys,Morris:2012aa}. 
-However, modeling complex dynamics remains difficult with this class of model.
+Toward this challenge, logical modeling is an emerging paradigm that encodes causal relationships between model components using non-linear transfer functions, 
+without requiring mechanism. \citep{Morris:2010aa}. 
+Despite their simplicity, logical models have captured rich behaviors in a variety of systems important to human health \citep{Saez-Rodriguez:2011aa,Morris:2011ys,Morris:2012aa}. 
+However, modeling complex dynamics with logical models is challenging. 
 Thus, there is an unmet need for a third approach which combines ODEs and logical models, where ODEs could encode mechanistic information, 
@@
 The extrinsic cascade is generally believed to be the main mechanism of thrombinogenesis in the blood \cite{MANN1990,ROBERTS1998,MANN1999}.
-Initially, thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa), which itself has been activated by FVIIa \cite{Butenas:2002aa}. 
+Initially, thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa), which itself has been activated by TF/FVIIa \cite{Butenas:2002aa}. 
 Picomolar amounts of thrombin then activate the cofactors factors V and VIII (fV and fVIII) and platelets, 
@@
 \subsection*{Formulation of reduced order coagulation models.}
-We developed a reduced order coagulation model to test our hybrid kinetic rule based modeling approach (Fig. \ref{fig-network}).
+We developed a reduced order extrinsic coagulation model to test our hybrid modeling approach (Fig. \ref{fig-network}).
 A trigger event initiates thrombin formation (FIIa) from prothrombin (fII) through a lumped initiation step.
-This step loosely represents the initial fluid phase activation of thrombin by activated FXa.
+This step loosely represents the initial activation of thrombin by activated FXa.
 Once activated, thrombin catalyzes its own formation (amplification step), and inhibition via the conversion of protein C to activated protein C (APC).
-APC and tissue factor pathway inhibitor (TFPI) inhibit initiation, while antithrombin III (ATIII) inhibits amplification. 
+Antithrombin III (ATIII) inhibits amplification, while APC and tissue factor pathway inhibitor (TFPI) potentially inhibit both initiation and amplification. 
 All initiation and inhibition processes, as well as the dependence of amplification upon other coagulation factors, was approximated using our rule-based approach (Fig. \ref{fig-control-schematic}). 
-Individual regulatory contributions to the activity of pathway enzymes were integrated into control coefficients using an integration rule.  
+Individual regulatory contributions to the activity of pathway enzymes were integrated into control coefficients (v's) using an integration rule (min/max).  
 These control coefficients then modified the rates of model processes at each time step.
-Hill-like transfer functions $0 \leq f\left(\mathcal{Z}\right) \leq 1$ quantified the influence of components upon a target process. 
+Hill-like transfer functions $0 \leq f\left(\mathcal{Z}\right) \leq 1$ quantified the contribution of components upon a target process. 
 Components were either individual inhibitor or activator levels or some function of levels, e.g., the product of factor levels. 
@@
 However, other integration rules are certainly possible. 
-Taken together, while the reduced order coagulation model encodes significant biological complexity, it is highly compact, consisting of only five differential equations. 
+Taken together, while the reduced order coagulation model encodes significant biological complexity, it is highly compact (consisting of only five differential equations). 
 Thus, it will serve as an excellent proof of principle example to study the reduction of a highly complex human subsystem.
@@
 All plots were made using the Matplotlib module of Python \citep{Matplotlib}.
-The optimization code can be downloaded from http://www.varnerlab.org.
+%The optimization and plotting code can be downloaded from http://www.varnerlab.org.
 
 \subsection*{Global sensitivity analysis of model performance.}
-We conducted a global sensitivity analysis to estimate which parameters controlled the performance of the reduced order model. 
-We computed the total sensitivity index of each parameter relative to two performance objectives, the peak thrombin time and the area under the thrombin curve 
-(thrombin exposure). [FINISH~ME].
+We conducted a global sensitivity analysis, using the variance based method of Sobol, to estimate which parameters controlled the performance of the reduced order model \citep{SOBOL_METHOD}. 
+We computed the total sensitivity index of each parameter relative to two performance objectives, the peak thrombin time and the area under the thrombin curve (thrombin exposure).
+We established the sampling bounds for each parameter from the minimum and maximum value of that parameter in the parameter set ensemble. 
+We used the sampling method of Saltelli et al., \citep{Saltelli:2010} to compute a family of $N\left(2d+2\right)$ parameter sets which obeyed our parameter ranges, 
+where $N$ was the number of trials, and $d$ was the number of parameters in the model. In our case, N = 10,000 and d = 22, so the total sensitivity indices were computed from
+460,000 model evaluations. The variance based sensitivity analysis was conducted using the SALib module encoded the Python programming language \citep{SALIB}.
 
@@
 Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles.
-The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation). 
+The shaded region denotes the 99\% confidence estimate of the mean thrombin value (uncertainty in the model simulation). 
 All factors and control proteins were at their physiological concentration unless others noted.}\label{fig-rFVII}

### 2014-10-23 Paper_v1.tex
Final draft version 1

@@
 \vspace{0.1in}
-{\par \centering \textbf{Running Title:}~Modeling blood coagulation}
+{\par \centering \textbf{Running Title:}~Reduced order modeling of blood coagulation}
 \vspace{0.1in}
@@
 Toward this challenge, logical modeling is an emerging paradigm that encodes causal relationships between model components using non-linear transfer functions, 
-without requiring mechanism. \citep{Morris:2010aa}. 
+without requiring mechanism \citep{Morris:2010aa}. 
 Despite their simplicity, logical models have captured rich behaviors in a variety of systems important to human health \citep{Saez-Rodriguez:2011aa,Morris:2011ys,Morris:2012aa}. 
@@
 
-In this study, we developed a hybrid approach which combines ODE modeling with logical rules to model a well studied biochemical network, 
+In this study, we developed a hybrid approach which combined ODE modeling with logical rules to model a well studied biochemical network, 
 the human coagulation system. Coagulation is an archetype proteolytic cascade involving both positive 
@@
 
-Coagulation models in the literature are typically formulated as systems of nonlinear ordinary differential equations, 
-using mass-action or other kinetics, to describe the rates of biochemical conversions \citep{Kuharsky:2001aa}.
+Previous coagulation models have typically been formulated as systems of nonlinear ordinary differential equations, 
+using mass-action or more complex kinetics, to describe the rates of biochemical conversions \citep{Kuharsky:2001aa}.
 Previous ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys}, 
@@
 Other laboratories have also expanded upon Hockin \emph{et al.}, for example including the intrinsic pathway, and thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
-Other biochemical or biophysical aspects of coagulation have also been modeled, such as platelet biochemistry \citep{Stalker:2013aa}, 
-multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and transport inside growing clots \citep{Voronov:2013aa}.
-However, most of these previous studies were based upon detailed mechanistic knowledge. 
-Such a description is possible because blood, while enormously complex, can be systematically interrogated. 
-However, other systems, such as the intracellular networks which mediate cancer, are more difficult to experimentally probe and model.  
-Towards this unmet need, we have formulated a hybrid modeling approach which combines ODEs and logical rules to model 
-biochemical processes for which mechanistic information is difficult or impossible to obtain. 
-We have used this approach to model the coagulation cascade because of the wealth of training and validation data available for this system. 
-We proposed a reduced order coagulation model that consisted of five differential equations augmented with several logical rules.
-We identified the reduced order coagulation model parameters from in vitro extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
-We then validated the model using thrombin data sets from multiple laboratories, for both normal and hemophilic coagulation.
-Flux and sensitivity analysis identified which parameter or parameters combinations were critical to model performance in several conditions. 
-The reduced order hybrid approach produced a surprisingly predictive coagulation model, suggesting this framework could potentially be used to model other biochemical networks important to human health.
+Other aspects of coagulation have also been modeled, such as platelet biochemistry \citep{Stalker:2013aa}, 
+multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and transport inside clots \citep{Voronov:2013aa}.
+However, these previous studies were largely based upon extensive mechanistic knowledge. 
+This is possible because blood, while enormously complex, can be systematically interrogated.
+Other systems, such as intracellular signaling networks, are much more difficult to experimentally interrogate.  
+Towards this unmet need, we formulated a hybrid modeling approach which combines ODEs and logical rules to model 
+biochemical processes for which a complete mechanistic understanding is missing. 
+We tested this approach by modeling the human coagulation cascade. 
+The hybrid model consisted of only five differential equations augmented with several logical rules.
+Thus, the model was more than an order of magnitude smaller than comparable purely ODE models in the literature.
+We estimated the model parameters from \textit{in~vitro} extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
+We then compared the model predictions with thrombin data sets, for both normal and hemophilic coagulation, that were not used for model training.
+Once validated, we performed flux and sensitivity analysis on the model to estimate which parameters were critical to model performance in several conditions. 
+The reduced order hybrid approach produced a surprisingly predictive coagulation model, 
+suggesting this framework could potentially be used to model other biochemical networks important to human health.
 
@@
 %However, components could also correspond to non-modeled influences, categorial variables or other abstract quantities.
+%In the current proof-of-principle model, 
+%APC can inhibit amplification, and upstream of amplification thus APC activity will directly slow the rate of thrombin formation from the initiation trigger, and amplification itself.
+%On the other hand, ATIII inhibits thrombin directly.
 
@@
 of both ATIII and the protein C pathway (Fig. \ref{fig-diff_fII_APC}). Lastly, we were unable to capture global differences in initiation time \textit{across} separate data 
-sets with a single ensemble of model parameters. These differences likely result from normal experimental variability.
-For example different thrombin generation experiments within our 
-training data (at the same physiological factor levels) have significantly different initiation times (supplemental results). 
-However, this also highlights a potential shortcoming of the initiation module within the model. 
+sets with a single ensemble of model parameters. These differences likely resulted from normal experimental variability.
+For example, different thrombin generation experiments within the 
+training data (at the same physiological factor levels) had significantly different initiation times (data not shown). 
+However, the inability to globally capture initiation time also highlighted a potential shortcoming of the initiation module within the model. 
 To capture the variability in initiation time \textit{across} training data sets, we included a constant time-delay parameter ($T_{D}$) for each data group.
-The delay parameter was constant within a data set, but allowed to vary \textit{across} training data sets. Introduction of the delay parameter allowed the model
-to simulate multiple training data sets using a single ensemble of model parameters.
+The delay parameter was constant within a data set, but allowed to vary \textit{across} training data sets. 
+Introduction of the delay parameter allowed the model to simulate multiple training data sets using a single ensemble of model parameters.
 Taken together, the model identification results suggested 
-that our kinetic-rules based approach could reproduce a panel of thrombin generation data sets conducted at physiological factor and inhibitors concentrations.
+that our hybrid approach could reproduce a panel of thrombin generation data sets in the neighborhood of physiological factor and inhibitors concentrations.
 However, it was unclear whether the reduced order model could predict new data, without updating the model parameters.
@@
 Taken together, the reduced order model performed well in the physiological range of factors, even with
-unmodeled components such as platelet activation in the hemophilia data set. Thus, our kinetic-rules framework predicted 
-the output of a physiologically important cascades such as coagulation, despite significant unmodeled components.
+unmodeled components such as platelet activation in the hemophilia data set.
 
 The model ensemble predicted a direct correlation between thrombin generation and rFVIIa addition in hemophilia (Fig. \ref{fig-rFVII}).
-In the current model, we cannot distinguish between different initiation sources, e.g., TF/FVIIa and rFVIIa, as we have only a single lumped initiation source (trigger).
+In the current model, we cannot distinguish between different initiation sources, e.g., TF/FVIIa versus rFVIIa, as we have only a single lumped initiation source (trigger).
 Thus, we simulated the addition of rFVIIa in hemophilia by removing fVIII and fIX from the model, and modulating the initial level of trigger.
-Simulations with a baseline level of trigger were consistent with the previous hemophilia simulations, where the only thrombin being produced was from initiation (Fig.
-\ref{fig-rFVII}, 1$\times$~trigger).
-As we increased the strength of the trigger event, the thrombin peak time and the maximum value of thrombin increased (Fig. \ref{fig-rFVII}, 50$\times$~trigger). 
-However, as the trigger strength increased, the thrombin generated quickly decayed.
-For example, for the highest trigger strength simulated (50$\times$trigger), 95\% of the thrombin was gone by 20 min after initiation. 
+Simulations with a baseline level of trigger were consistent with the previous hemophilia simulations, where the only thrombin produced was from initiation (Fig. \ref{fig-rFVII}A, 1$\times$~trigger).
+However, as we increased the trigger strength, the thrombin profile began to approximate normal coagulation, showing a pronounced peak albeit with a slower peak time (Fig. \ref{fig-rFVII}B).
+Further increases in trigger strength resulted in decreased thrombin peak time and increased maximum thrombin values (Fig. \ref{fig-rFVII}A, 50$\times$~trigger). 
+Thus, for large trigger values (200$\times$trigger) the hemophilic thrombin profile approximated normal coagulation, where peak thrombin was achieved shortly 
+after administration and 95\% of the thrombin was gone by 20 min after initiation. 
 We performed flux analysis to understand how the reduced order coagulation model balanced initiation, amplification and inhibition of thrombin formation for normal coagulation and hemophilia.
@@
 \ref{fig-flux-analysis}).
-The reduced order network includes four lumped reactions, initiation, amplification, thrombin-induced APC generation and total thrombin inhibition (including both APC and ATIII action).
+We calculated the flux through four lumped reactions, initiation, amplification, thrombin-induced APC generation and total thrombin inhibition (including both APC and ATIII action).
 Directly after the addition of a trigger (e.g., TF/FVIIa or rFVIIa), the lumped initiation flux was the largest for all three cases.
@@
 During amplification, thrombin catalyzes its own formation and inhibition by generating activated protein C (APC), a potent inhibitor of the coagulation cascade.
-For normal coagulation, amplification and thrombin inhibition are the dominate reactions by 6 min after initiation (Fig. \ref{fig-flux-analysis}, left). 
-After 10 min, the dominate reaction has shifted to thrombin inhibition (both ATIII and APC action). In the current proof-of-principle model, 
-APC inhibits upstream of amplification thus APC activity will slow the rate of thrombin formation directly from the initiation trigger. 
-On the other hand, ATIII inhibits thrombin directly, as well as upstream coagulation factors.
-In hemophilia (missing both fVIII and fIX), the amplification reaction does not occur and the only thrombin produced is from initiation (Fig. \ref{fig-flux-analysis}, center).
-Initiation is quickly inhibited by APC, and the thrombin level stabilizes and eventually decays because of ATIII activity. 
-Lastly, when 50$\times$rFVIIa is used to induce thrombin formation in hemophilia (absence of fVIII/fIX), 
-initiation mechanisms dominate for up to 6 min following initiation (Fig. \ref{fig-flux-analysis}, right). 
-However, similar to hemophilia alone, without amplification the thrombin signal is quickly extinguished by the combined action of ATIII and APC generated by thrombin.
+For normal coagulation, amplification and thrombin inhibition were the dominate reactions by 6 min after initiation (Fig. \ref{fig-flux-analysis}, left). 
+After 10 min, the dominate reaction had shifted to thrombin inhibition (both ATIII and APC action). 
+In hemophilia (missing both fVIII and fIX), the amplification reaction did not occur, and thrombin was produced only by initiation (Fig. \ref{fig-flux-analysis}, center).
+Initiation was quickly inhibited by APC, and the thrombin level stabilized (eventually decaying at longer times because of ATIII activity). 
+Lastly, when 50$\times$trigger was used to induce thrombin formation in hemophilia (absence of fVIII/fIX), 
+initiation mechanisms dominated for up to 6 min following initiation (Fig. \ref{fig-flux-analysis}, right). 
+Similar to hemophilia alone, no amplification occurred in the 50$\times$trigger+hemophilia case, 
+and the rate of thrombin generation was extinguished by the combined action of ATIII and APC. 
+Taken together, the hybrid modeling approach captured the transition between the modes of thrombin generation, as well as the role that inhibitors play 
+in attenuating the thrombin generation rate. Thus, the transfer function approach encoded the inhibitory logic of this cascade in the absence of specific mechanism.
 
@@
 Globally, 41\% of the parameters shifted in importance between the (fII,PC) = (50\%, 0\%) and (150\%,100\%) cases for the peak thrombin time (Fig. \ref{fig-sensitivity}A). 
-The majority of these shifts were involved increased prothrombin and the protein C pathway, while only
+The majority of these shifts involved the interaction between increased prothrombin and the protein C pathway, while only
 5\% were directly associated with increased prothrombin alone. The rate constant for thrombin amplification was the most important parameter controlling the
 peak thrombin time. While this parameter was differentially important for different prothrombin levels, and in the presence or absence of 
-the activated protein C pathway, it's sensitivity was consistently above all other parameters in the model. The saturation constant governing thrombin amplification was
+the activated protein C pathway, it's was consistently the most sensitive parameter in the model. The saturation constant governing thrombin amplification was
 the second most important parameter, followed by the initiation control gain parameter. Other important parameters influencing the thrombin peak time included the control gain for
-activated protein C formation, as well as the rate constant controlling ATIII inhibition of thrombin activity. 
-On the other hand, only 27\% of the model parameters were differentially sensitive between the (\%fII,\%PC) (50\%, 0\%) and (150\%,100\%) cases 
-for the thrombin exposure (Fig. \ref{fig-sensitivity}B). Of these parameters, all of the shifts were associated the interplay between thrombin formation and the protein C pathway. 
+activated protein C formation, and the rate constant controlling ATIII inhibition of thrombin activity. 
+On the other hand, only 27\% of the model parameters were differentially sensitive between the (fII,PC) = (50\%, 0\%) and (150\%,100\%) cases 
+for thrombin exposure (Fig. \ref{fig-sensitivity}B). Of these parameters, all of the shifts were associated with the interplay between thrombin formation and the protein C pathway. 
 The rate constant controlling ATIII inhibition was the most important parameter controlling the thrombin exposure.
@@
 Similar to the peak time, for 150\% prothrombin, the control gain for activated protein C formation was differentially important along with the rate constant controlling
-amplification. However, the influence of the amplification parameter was much less in the thrombin exposure case versus peak time.
+amplification. However, the amplification parameter was much less important for thrombin exposure versus peak time.
 
@@
 and has an abundance of experimental data available for model identification and validation. 
-However, coagulation was just a proof-of concept validation of our approach; 
+However, coagulation was just a proof-of concept test of our approach; 
 the proposed hybrid framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
@@
 We then used flux analysis to understand how the network operated in a variety of conditions, and global sensitivity analysis to identify which parameters controlled the 
-performance of the network. Taken together, the reduced order hybrid approach produced a surprisingly predictive model, suggesting this approach could potentially 
+performance of the network. Taken together, the proposed hybrid framework produced a surprisingly predictive model, suggesting this approach could 
 be used to model other biochemical networks important to human health.
@@
 However, rFVIIa requires frequent administration (every 2-3 hr), and many questions remain about its mechanism of action, 
-its effective dose range \cite{TOMOKIYO2003}, and its overall utility for the treatment trauma-associated hemorrhage \cite{Duchesne:2008hw}. 
+its effective dosage \cite{TOMOKIYO2003}, and its overall utility for the treatment trauma-associated hemorrhage \cite{Duchesne:2008hw}. 
 In this study, we did not model rFVIIa induced coagulation directly. 
@@
 However, no amplification of the thrombin signal occurred without fVIII or fIX. 
-Thus, the peak thrombin signal was lower than normal coagulation, the peak thrombin time was longer, and the thrombin signal was eventually inhibited by ATIII and the protein C pathway.
-However, as the dose of rFVIIa increased, the peak thrombin time decreased (eventually saturating around 100$\times$nominal trigger), and the peak thrombin value increased.
-For 100$\times$nominal trigger, the thrombin profile resembled normal coagulation.
+Thus, the peak thrombin signal was lower than normal coagulation, the peak thrombin time was longer, and thrombin generation was eventually inhibited by the combined action of ATIII and the protein C pathway.
+However, as the dose of rFVIIa increased, the peak thrombin time decreased (eventually saturating around 200$\times$nominal trigger), and the peak thrombin value increased such that
+the thrombin profile resembled normal coagulation.
 Butenas \emph{et al.,} performed an extensive in vitro study of rFVIIa induced thrombin generation under normal and hemophilic conditions \citep{Butenas:2002ab}.
@@
 
-Although the performance of the proof of principle coagulation model was impressive given its size, there were several mathematical and biological questions that can be explored further.
-First, the choice of max/min integration rules, or the particular form of the transfer functions could be generalized to include other rule types and functions. 
-Theoretically, an integration rule is simply a function whose domain is a set of transfer function inputs, and whose range is $v\in[0,1]$.
-Thus, integration rules other than max/min could be used, such as the mean or the product, assuming the range of the transfer functions is always $f\in[0,1]$.
-Alternative integration rules such the mean would have different properties which could influence model identification or performance. 
-For example, a mean or product integration rule would be differentiable, which allows derivative based optimization approaches to be used. 
-The particular form of the transfer function could also be explored. We choose a hill-like function because of its 
-prominence in the systems and synthetic biology community. 
-However, the only mathematical requirement for a transfer function is that it must map a non-negative continuous or categorical variable into the 
-range $f\in[0,1]$. Thus, many function types are possible.
-There are several additional biological modules that can be added to the core model presented here. 
-First, we need to update the model to included thrombin induced platelet activation. 
-While we were able to capture thrombin generation data in the presence of platelets, 
+Although the performance of the proof of principle coagulation model was impressive given its limited size, there were several issues that could be further explored.
+First, the prediction of initiation time should be investigated. 
+We were able to estimate initiation time within a data set, but unable to predict initiation time \textit{across} independent data sets. 
+This suggested that we should update the initiation module to distinguish between different triggers e.g., TF/FVIIa versus rFVIIa alone, 
+and to include key biological milestones such as FXa activation (a prerequisite to thrombin formation).
+Next, there are several additional biological modules that could be added to the core model presented here. 
+First, we could include thrombin induced platelet activation, and the role of activated platelets in amplification.
+We captured thrombin generation data in the presence of platelets, however,
 the initial shape of the activation curve and the time-scale of activation was not always consistent with the data.
 Platelets are activated by thrombin through the cleavage of the extracellular domain of protease-activated receptors (PARs) on the platelet surface.
-Next, we should add the intrinsic or contact thrombin activation pathway to the model. 
-The intrinsic coagulation pathway is triggered by contact activation of the plasma protease factor XII (fXII) by negatively charged surfaces.
-The activation of fXII is followed by proteolytic activation of factor XI (fXI) and the activation of downstream factors like fIX \cite{NAITO1991}.
-It has also been suggested that activated platelets can release polyphosphate which directly activates fXII \cite{Smith:2006zl}. 
-Lastly, to make the model more clinically relevant, we should include the biochemical processes responsible for clot formation, and clot dissolution (fibrinolysis).
+Once activated, platelets play an important role in amplification, and are key mediators of the positive feedback driving amplification.
+Thus, this biology is a potentially important component of an expanded model.
+We should also add the intrinsic pathway to the model. 
+The intrinsic pathway is triggered by contact activation of the plasma protease factor XI (fXI) by negatively charged surfaces, and by thrombin and upstream factors such as activated
+plasma protease factor XII (FXIIa) \citep{NAITO1991,Gailani:1991aa}.
+Activated platelets may also release polyphosphate which directly activates fXII \cite{Smith:2006zl}. 
+Arguably a minor player in acute bleeding, contact activation could also be important in other wound healing contexts. 
+Finally, to make the model more clinically relevant, we should include the biochemical processes responsible for clot formation, and clot dissolution (fibrinolysis).
 Clot formation is driven by thrombin activity, while fibrinolysis is driven by plasmin activity, a key enzyme that cleaves fibrin (one of the main materials in a clot).
@@
 inhibits the degradation of fibrin by plasmin. Also similar to coagulation, there is considerable fibrinolysis and contact pathway data sets that can be used to train the model.
+Lastly, the choice of max/min integration rules, or the particular form of the transfer functions could be generalized to include other rule types and functions. 
+Theoretically, an integration rule is a function whose domain is a set of transfer function inputs, and whose range is $v\in[0,1]$.
+Thus, integration rules other than max/min could be used, such as the mean or the product, assuming the range of the transfer functions is always $f\in[0,1]$.
+Alternative integration rules such the mean might have different properties which could influence model identification or performance. 
+For example, a mean integration rule would be differentiable, which allows derivative based optimization approaches to be used. 
+The particular form of the transfer function could also be explored. We choose a hill-like function because of its 
+prominence in the systems and synthetic biology community. 
+However, the only mathematical requirement for a transfer function is that it map a non-negative continuous or categorical variable into the 
+range $f\in[0,1]$. Thus, many transfer function types are possible.
 
@@
 Sensitivity values significantly above or below the diagonal indicate differentially important model parameters. 
+The radius of the shaded region around each total sensitivity value was the maximum uncertainty in that value estimated by the Sobol method. 
 }\label{fig-sensitivity}

### 2014-10-24 Paper_v2.tex
final edits Paper_v2

--- 2014-10-23 Paper_v1.tex
@@
 Modeling approaches differ in their degree of detail, where the choice of approach is often determined by prior knowledge, or model requirements \citep{2012_kholodenko_kolch_SciSig}. 
-Ordinary differential equation (ODE) models are a common tool for modeling biochemical systems because of their ability to capture dynamics, and encode mechanism.
+Ordinary differential equation (ODE) models are common tools for modeling biochemical systems because of their ability to capture dynamics and encode mechanism.
 However, ODE models typically come with difficult (or sometimes impossible) parameter identification problems. 
-For example, Gadkar et al., showed that even with near-perfect information, 
+For example, Gadkar \textit{et al.}, showed that even with near-perfect information, 
 it was often impossible to identify all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
@@
 sensitive to only a few parameters, a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
-Thus, reasonable predictions may be possible, despite parametric uncertainty, if a few critical parameters are well defined. 
+Thus, reasonable predictions may be possible, despite parametric uncertainty, if a few critical parameters are well-defined. 
 Perhaps more importantly, ODE models also require significant mechanistic information, thereby limiting their utility in poorly understood systems.
@@
 converts protein C to activated protein C (APC) which attenuates the coagulation response by proteolytic cleavage of fV/FVa and fVIII/FVIIIa. 
-Termination occurs after either prothrombin is consumed, or thrombin formation is neutralized by inhibitors such as activated protein C (APC) or antithrombin III (ATIII).
+Termination occurs after either prothrombin is consumed, or thrombin formation is neutralized by inhibitors such as APC or ATIII.
 
 Previous coagulation models have typically been formulated as systems of nonlinear ordinary differential equations, 
-using mass-action or more complex kinetics, to describe the rates of biochemical conversions \citep{Kuharsky:2001aa}.
-Previous ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys}, 
+using mass action or more complex kinetics, to describe the rates of biochemical conversions \citep{Kuharsky:2001aa}.
+Previous ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys} 
 were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, 
@@
 All initiation and inhibition processes, as well as the dependence of amplification upon other coagulation factors, was approximated using our rule-based approach (Fig. \ref{fig-control-schematic}). 
-Individual regulatory contributions to the activity of pathway enzymes were integrated into control coefficients (v's) using an integration rule (min/max).  
+Individual regulatory contributions to the activity of pathway enzymes were integrated into control coefficients ($v$'s) using an integration rule (min/max).  
 These control coefficients then modified the rates of model processes at each time step.
@@
 (i) estimation of parameters associated with thrombin formation in the absence of the protein C pathway and (ii) estimation of parameters associated with the protein C pathway.
-Only those parameters associated with each subproblem were varied during the optimization procedure for that subproblem e.g., thrombin parameters were \textit{not} varied during the protein C subproblem.The PSO procedure was run for 20 generations for each subproblem, where each generation was 1200 iterations.
+Only those parameters associated with each subproblem were varied during the optimization procedure for that subproblem, e.g., thrombin parameters were \textit{not} varied during the protein C subproblem.The PSO procedure was run for 20 generations for each subproblem, where each generation was 1200 iterations.
 The best particle from each generation was used to generate the particle population for the next generation.
@@
 Taken together, the model identification results suggested 
-that our hybrid approach could reproduce a panel of thrombin generation data sets in the neighborhood of physiological factor and inhibitors concentrations.
+that our hybrid approach could reproduce a panel of thrombin generation data sets in the neighborhood of physiological factor and inhibitor concentrations.
 However, it was unclear whether the reduced order model could predict new data, without updating the model parameters.
@@
 Simulations near the physiological range (fII,ATIII) = (100\%, 100\%) or (125\%,75\%) tracked the measured thrombin values (Fig. \ref{fig-fII_ATIII}B and C).
-On the other hand, predictions for factors levels outside of the physiological range (fII,ATIII) = (50\%, 150\%) or (150\%, 50\%), while qualitatively consistent with measured thrombin values, did show significant
+On the other hand, predictions for factor levels outside of the physiological range (fII,ATIII) = (50\%, 150\%) or (150\%, 50\%), while qualitatively consistent with measured thrombin values, did show significant
 deviation from the measurements (Fig. \ref{fig-fII_ATIII}A and D).
@@
 Further increases in trigger strength resulted in decreased thrombin peak time and increased maximum thrombin values (Fig. \ref{fig-rFVII}A, 50$\times$~trigger). 
-Thus, for large trigger values (200$\times$trigger) the hemophilic thrombin profile approximated normal coagulation, where peak thrombin was achieved shortly 
+Thus, for large trigger values (200$\times$trigger), the hemophilic thrombin profile approximated normal coagulation, where peak thrombin was achieved shortly 
 after administration and 95\% of the thrombin was gone by 20 min after initiation. 
@@
 peak thrombin time. While this parameter was differentially important for different prothrombin levels, and in the presence or absence of 
-the activated protein C pathway, it's was consistently the most sensitive parameter in the model. The saturation constant governing thrombin amplification was
+the activated protein C pathway, it was consistently the most sensitive parameter in the model. The saturation constant governing thrombin amplification was
 the second most important parameter, followed by the initiation control gain parameter. Other important parameters influencing the thrombin peak time included the control gain for
@@
 %Simulations of thrombin dynamics were independently validated using blood drawn from patients with coronary artery disease (reported here) and previous TF/FVIIa and FVIIa 
-%\emph{in-vitro} studies \cite{BUTENAS2002}.
+%\emph{in vitro} studies \cite{BUTENAS2002}.
 
@@
 and has an abundance of experimental data available for model identification and validation. 
-However, coagulation was just a proof-of concept test of our approach; 
-the proposed hybrid framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
+However, coagulation was just a proof of concept test of our approach. 
+The proposed hybrid framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
 or potentially even cell free metabolic systems. 
 The model consisted of five differential equations augmented with several logical rules describing 
-regulatory connections between model components, and unmodeled interactions in the network. 
-We estimated model parameters from in vitro extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
+regulatory connections between model components and unmodeled interactions in the network. 
+We estimated model parameters from \textit{in vitro} extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
 To estimate parameters, the residual between model simulations and experimental measurements was minimized using particle swarm optimization (PSO). 
@@
 Aggressive clotting involved with Coronary Artery Diseases (CADs), collectively accounts for 38\% of all deaths in North America \cite{HANSSON2005}.
-Coagulation management during surgery can also be challenging, particularly with the increasing clinical use of antithrombotic drugs \cite{Tanaka:2009wo}.   
+Coagulation management during surgery can also be challenging, particularly with the increase in clinical use of antithrombotic drugs \cite{Tanaka:2009wo}.   
 Insufficient coagulation due to genetic disorders such as hemophilia can also result in recurrent bleeding. 
@@
 Hemophilia can be controlled with regular infusions of the deficient clotting factors. 
-However, clotting factor replacement sometimes leads to the formation of fVIII and fIX inhibitors \emph{in-vivo} \cite{TOMOKIYO2003}.
+However, clotting factor replacement sometimes leads to the formation of fVIII and fIX inhibitors \emph{in vivo} \cite{TOMOKIYO2003}.
 Alternatively, recombinant factor VIIa (rFVIIa) has been used to treat bleeding disorders \cite{Hedner:2008rm,Talbot:2009kk} including
@@
 its effective dosage \cite{TOMOKIYO2003}, and its overall utility for the treatment trauma-associated hemorrhage \cite{Duchesne:2008hw}. 
-In this study, we did not model rFVIIa induced coagulation directly. 
+In this study, we did not model rFVIIa-induced coagulation directly. 
 Rather, we modeled a general trigger which initiated the extrinsic coagulation cascade. 
@@
 the thrombin profile resembled normal coagulation.
-Butenas \emph{et al.,} performed an extensive in vitro study of rFVIIa induced thrombin generation under normal and hemophilic conditions \citep{Butenas:2002ab}.
-They found qualitatively similar trends, namely rFVIIa could restore normal coagulation (even in the absence of TF) for large enough rFVIIa doses, although rFVIIa induced coagulation 
+Butenas \emph{et al.} performed an extensive \textit{in vitro} study of rFVIIa-induced thrombin generation under normal and hemophilic conditions \citep{Butenas:2002ab}.
+They found qualitatively similar trends, namely rFVIIa could restore normal coagulation (even in the absence of TF) for large enough rFVIIa doses, although rFVIIa-induced coagulation 
 in hemophilia (even for large rFVIIa doses) lagged the normal profile. 
@@
 might not be able to initiate normal coagulation in recurrent bleeding, unless the dosage is above a critical threshold. 
-Unfortunately, this threshold is likely patient specific, thus limiting potentially limiting the utility of rFVIIa administration.
+Unfortunately, this threshold is likely patient specific, thus potentially limiting the utility of rFVIIa administration.
 
@@
 We were able to estimate initiation time within a data set, but unable to predict initiation time \textit{across} independent data sets. 
-This suggested that we should update the initiation module to distinguish between different triggers e.g., TF/FVIIa versus rFVIIa alone, 
+This suggested that we should update the initiation module to distinguish between different triggers, e.g., TF/FVIIa versus rFVIIa alone, 
 and to include key biological milestones such as FXa activation (a prerequisite to thrombin formation).
 Next, there are several additional biological modules that could be added to the core model presented here. 
-First, we could include thrombin induced platelet activation, and the role of activated platelets in amplification.
+First, we could include thrombin-induced platelet activation and the role of activated platelets in amplification.
 We captured thrombin generation data in the presence of platelets, however,
@@
 We should also add the intrinsic pathway to the model. 
-The intrinsic pathway is triggered by contact activation of the plasma protease factor XI (fXI) by negatively charged surfaces, and by thrombin and upstream factors such as activated
+The intrinsic pathway is triggered by contact activation of the plasma protease factor XI (fXI) by negatively charged surfaces and by thrombin and upstream factors such as activated
 plasma protease factor XII (FXIIa) \citep{NAITO1991,Gailani:1991aa}.
@@
 Arguably a minor player in acute bleeding, contact activation could also be important in other wound healing contexts. 
-Finally, to make the model more clinically relevant, we should include the biochemical processes responsible for clot formation, and clot dissolution (fibrinolysis).
+Finally, to make the model more clinically relevant, we should include the biochemical processes responsible for clot formation and clot dissolution (fibrinolysis).
 Clot formation is driven by thrombin activity, while fibrinolysis is driven by plasmin activity, a key enzyme that cleaves fibrin (one of the main materials in a clot).
 Similar to coagulation, fibrinolysis is managed by several activating and inhibitory factors which control the balance between clot formation and dissolution.
-Tissue plasminogen activator (t-PA) and urokinase activate plasmin, along with contact pathway factors such fXIa. On the other hand, thrombin activatable fibrinolysis inhibitor (TAFI)
-inhibits the degradation of fibrin by plasmin. Also similar to coagulation, there is considerable fibrinolysis and contact pathway data sets that can be used to train the model.
-Lastly, the choice of max/min integration rules, or the particular form of the transfer functions could be generalized to include other rule types and functions. 
+Tissue plasminogen activator (t-PA) and urokinase activate plasmin, along with contact pathway factors such as fXIa. On the other hand, thrombin activatable fibrinolysis inhibitor (TAFI)
+inhibits the degradation of fibrin by plasmin. Also, similar to coagulation, there is considerable fibrinolysis and contact pathway data sets that can be used to train the model.
+Lastly, the choice of max/min integration rules or the particular form of the transfer functions could be generalized to include other rule types and functions. 
 Theoretically, an integration rule is a function whose domain is a set of transfer function inputs, and whose range is $v\in[0,1]$.
 Thus, integration rules other than max/min could be used, such as the mean or the product, assuming the range of the transfer functions is always $f\in[0,1]$.
-Alternative integration rules such the mean might have different properties which could influence model identification or performance. 
-For example, a mean integration rule would be differentiable, which allows derivative based optimization approaches to be used. 
-The particular form of the transfer function could also be explored. We choose a hill-like function because of its 
+Alternative integration rules such as the mean might have different properties which could influence model identification or performance. 
+For example, a mean integration rule would be differentiable, which allows derivative-based optimization approaches to be used. 
+The particular form of the transfer function could also be explored. We choose a Hill-like function because of its 
 prominence in the systems and synthetic biology community. 
@@
 \end{equation}where $0\leq f_{ij}\left(\mathcal{Z}\right)\leq 1$ was a regulatory transfer function that calculated the influence of metabolite $i$ on the activity of enzyme $j$. 
-Conversely, if enzyme activity was inhibited by a $m$ metabolites, we modeling this inhibition as:
+Conversely, if enzyme activity was inhibited by $m$ metabolites, we modeled this inhibition as:
 \begin{equation}
@@
 \end{equation}where $\mathcal{Z}_{j}$ denotes the abundance of the $j$ factor (e.g., metabolite abundance), and $\kappa_{ij}$ and $\eta$ are control parameters. 
-The $\kappa_{ij}$ parameter was species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).
+$\kappa_{ij}$ was the species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).
 Applying the general framework to the reduced coagulation network resulted in five ordinary differential equations:
@@
 The rate of initiation was modified by $v_{init}$, the control parameter governing initiation. 
-Initiation was sensitive to level of trigger (activator), and TFPI (inhibitor):
+Initiation was sensitive to the level of trigger (activator) and TFPI (inhibitor):
 \begin{equation}
@@
 	v_{amp} = \min\left(f^{-}_{amp}\left(TFPI\right),f^{-}_{amp}\left(x_{4}\right),f^{+}_{amp}\left(\mathcal{Z}_{amp}\right)\right)
-\end{equation}where $\mathcal{Z}_{amp} = fV\times fX\times fVIII\times fIX$. Although, $f^{+}_{amp}\left(\mathcal{Z}_{amp}\right)$ is an activating term, 
+\end{equation}where $\mathcal{Z}_{amp} = fV\times fX\times fVIII\times fIX$. Although $f^{+}_{amp}\left(\mathcal{Z}_{amp}\right)$ is an activating term, 
 we included it in the min integration rule; the factors in $\mathcal{Z}_{amp}$ were essential for amplification (if any of these factors was missing
@@
 	\bar{r}_{apc} = k_{APC,formation}\left(TM\right)\frac{x_{3}}{K_{formation,PC} + x_{3}}
-\end{equation}where $k_{APC,formation},~K_{formation,PC}$ denote the rate and saturation constants governing activated protein C formation, respectively and $TM$ denotes
-the thrombomodulin abundance. We modeled the control term which governed APC formation as a single thrombin dependent activation term:
+\end{equation}where $k_{APC,formation}$ and $K_{formation,PC}$ denote the rate and saturation constants governing activated protein C formation, respectively and $TM$ denotes
+the thrombomodulin abundance. We modeled the control term which governed APC formation as a single thrombin-dependent activation term:
 \begin{equation}
@@
 \end{equation}where $\hat{x}_{j}\left(\tau\right)$ denotes the measured value of species $j$ at time $\tau$, $x_{j}\left(\tau,\mathbf{k}\right)$ denotes the simulated 
-value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$. The outer summation is respect to
+value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$. The outer summation is with respect to
 time, while the inner summation is with respect to state. We minimized the model residual using Particle swarm optimization (PSO) \citep{PSO}.
@@
 	\mathbf{k}_{i} &=& \mathbf{k}_{i} + \mathbf{\Delta}_{i}
-\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes local best solution found by particle $i$, and
+\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes the local best solution found by particle $i$, and
 $\mathcal{G}$ denotes the best solution found over the entire population of particles.
@@
 \subsection*{Global sensitivity analysis of model performance.}
-We conducted a global sensitivity analysis, using the variance based method of Sobol, to estimate which parameters controlled the performance of the reduced order model \citep{SOBOL_METHOD}. 
+We conducted a global sensitivity analysis, using the variance-based method of Sobol, to estimate which parameters controlled the performance of the reduced order model \citep{SOBOL_METHOD}. 
 We computed the total sensitivity index of each parameter relative to two performance objectives, the peak thrombin time and the area under the thrombin curve (thrombin exposure).
 We established the sampling bounds for each parameter from the minimum and maximum value of that parameter in the parameter set ensemble. 
-We used the sampling method of Saltelli et al., \citep{Saltelli:2010} to compute a family of $N\left(2d+2\right)$ parameter sets which obeyed our parameter ranges, 
-where $N$ was the number of trials, and $d$ was the number of parameters in the model. In our case, N = 10,000 and d = 22, so the total sensitivity indices were computed from
-460,000 model evaluations. The variance based sensitivity analysis was conducted using the SALib module encoded the Python programming language \citep{SALIB}.
+We used the sampling method of Saltelli \textit{et al.} \citep{Saltelli:2010} to compute a family of $N\left(2d+2\right)$ parameter sets which obeyed our parameter ranges, 
+where $N$ was the number of trials, and $d$ was the number of parameters in the model. In our case, $N$ = 10,000 and $d$ = 22, so the total sensitivity indices were computed from
+460,000 model evaluations. The variance-based sensitivity analysis was conducted using the SALib module encoded in the Python programming language \citep{SALIB}.
 
@@
 \includegraphics[width=1.0\textwidth]{./figs/Figure-1-Network.pdf}
-\caption{Schematic of the connectivity of the reduced order coagulation model. A trigger compound e.g., TF/FVIIa initiates thrombin production (FIIa) from prothrombin (fII).
+\caption{Schematic of the connectivity of the reduced order coagulation model. A trigger compound, e.g., TF/FVIIa initiates thrombin production (FIIa) from prothrombin (fII).
 Once activated, thrombin catalyzes its own activation (amplification step), as well as its own inhibition via the conversion of protein C to activated protein C (APC).
 APC and tissue factor pathway inhibitor (TFPI) inhibit initiation and amplification, while antithrombin III (ATIII) directly inhibits thrombin. 
-All inhibition steps, and trigger-induced initiation were modeled using a rule-based approach. 
+All inhibition steps and trigger-induced initiation were modeled using a rule-based approach. 
 Likewise, the dependence of amplification on other coagulation factors was also modeled using a rule-based approach. 
@@
 \includegraphics[width=1.0\textwidth]{./figs/Figure-2-ControlSchematic.pdf}
-\caption{Schematic of rule based effective control laws. 
-Traditional enzyme kinetic expressions e.g., Michaelis–Menten or multiple saturation kinetics are multiplied by an enzyme activity control variable $0 \leq v_{j} \leq 1 $. 
+\caption{Schematic of rule-based effective control laws. 
+Traditional enzyme kinetic expressions, e.g., Michaelis–Menten or multiple saturation kinetics are multiplied by an enzyme activity control variable $0 \leq v_{j} \leq 1 $. 
 Control variables are functions of many possible regulatory factors encoded by arbitrary functions of the form $0\leq f_{j}\left(\mathcal{Z}\right)\leq 1$.
@@
 All factors and control proteins were at their physiological concentration unless others denoted. 
-The experimental training data was reproduced from the study of Butenas et al. \citep{Butenas:1999aa}.}\label{fig-diff_fII_NO_APC}
+The experimental training data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}.}\label{fig-diff_fII_NO_APC}
 \end{figure}
@@
 All factors and control proteins were at their physiological concentration unless others denoted.
-The experimental training data was reproduced from the study of Butenas et al. \citep{Butenas:1999aa}.}\label{fig-diff_fII_APC}
+The experimental training data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}.}\label{fig-diff_fII_APC}
 \end{figure}
@@
 All factors and control proteins were at their physiological concentration unless others denoted.
-The experimental validation data was reproduced from the study of Butenas et al. \citep{Butenas:1999aa}}\label{fig-fII_ATIII}
+The experimental validation data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}}\label{fig-fII_ATIII}
 \end{figure}
@@
 All factors and control proteins were at their physiological concentration unless others noted.
-The experimental validation data was reproduced from the study of Allen et al. \citep{ALLEN2006}.}\label{fig-hemophilia}
+The experimental validation data was reproduced from the study of Allen \textit{et al.} \citep{ALLEN2006}.}\label{fig-hemophilia}
 \end{figure}
@@
 \includegraphics[width=1.0\textwidth]{./figs/Figure-9-Sensitivity.pdf}
-\caption{Global sensitivity analysis of the reduced order coagulation model with with respect to the model parameters.
+\caption{Global sensitivity analysis of the reduced order coagulation model with respect to the model parameters.
 A: Sensitivity analysis of the thrombin peak time for different prothrombin levels (150\%,100\% and 50\% of the physiological value) as a function of activated protein C. 
@@
 Points denote the mean total sensitivity value, while the area around each point denotes the uncertainty in the sensitivity value. 
-The gray dashed line denotes the 45$^{o}$ degree diagonal, if sensitivity values are equal for different conditions they will lie on the diagonal.  
+The gray dashed line denotes the 45$^{\circ}$ degree diagonal, if sensitivity values are equal for different conditions they will lie on the diagonal.  
 Sensitivity values significantly above or below the diagonal indicate differentially important model parameters. 

### 2014-10-28 Paper_v2.tex
Update after reviews

@@
 \section*{Abstract}
-In this study, we present a novel hybrid modeling approach which combines ordinary differential equation (ODE) modeling with logical rules to simulate an archetype biochemical network, 
-the human coagulation cascade.  
+In this study, we present a novel modeling approach which combines ordinary differential equation (ODE) 
+modeling with logical rules to simulate an archetype biochemical network, the human coagulation cascade.  
 The model consisted of five differential equations augmented with several logical rules describing 
@@
 using mass action or more complex kinetics, to describe the rates of biochemical conversions \citep{Kuharsky:2001aa}.
-Previous ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys} 
-were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, 
-and later Hockin \emph{et al.} \citep{Hockin:2002aa}, who developed and then refined highly mechanistic models of extrinsic coagulation. 
+Mechanistic ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys} 
+were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, Hockin \emph{et al.} \citep{Hockin:2002aa}, and later Butenas et al., \citep{Butenas:2004aa}
+who developed and then subsequently refined highly mechanistic models of coagulation. 
 Other laboratories have also expanded upon Hockin \emph{et al.}, for example including the intrinsic pathway, and thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
@@
 We developed a reduced order extrinsic coagulation model to test our hybrid modeling approach (Fig. \ref{fig-network}).
+The core of our model was based upon the earlier work of Ismagilov and coworkers \citep{Runyon:2004aa,Kastrup:2006aa,Runyon:2007aa,Runyon:2008aa}, 
+where we added initiation, factor dependence, and specific inhibition terms to the earlier simplified models.
 A trigger event initiates thrombin formation (FIIa) from prothrombin (fII) through a lumped initiation step.
@@
 In this study, we developed a reduced order model of the human coagulation cascade.
-We chose coagulation because it is well studied, has a complex architecture, 
+We modeled coagulation because it is well studied, has a complex architecture, 
 and has an abundance of experimental data available for model identification and validation. 
@@
 from eight \emph{in vitro} time-series coagulation data sets with and without the protein C pathway.
+Ensemble approaches have been used previously for other signal transduction models \citep{Kuepfer:2007aa,Song:2009aa,2010_song_varner_BiotechJ,Tasseff:2010aa,Lequieu:2011aa}, 
+and for metabolic models \citep{Tran:2008aa} to 
+estimate the impact of poorly constrained parameter values or poorly understood network structure on simulation performance. 
+Thus, ensemble approaches are common in the dynamic modeling community. 
+However, perhaps a unique feature of the current study is the direct connection between our particle swarm approach, 
+and the parameter ensemble; each particle in our swarm uniquely corresponded to a parameter set in our ensemble. 
+Thus, by constraining particles to operate in different parameter regions, giving each particle a different parameter combination to explore, 
+or perhaps even suppling a different model formulation to each particle we can effectively 
+traverse through complex parameter and model spaces. 
 We validated the ensemble using thrombin data sets taken from multiple laboratories for a variety of experimental conditions not used during training. 
@@
 performance of the network. Taken together, the proposed hybrid framework produced a surprisingly predictive model, suggesting this approach could 
-be used to model other biochemical networks important to human health.
+be used to effectively model other biochemical networks important to human health.
 
@@
 Butenas \emph{et al.} performed an extensive \textit{in vitro} study of rFVIIa-induced thrombin generation under normal and hemophilic conditions \citep{Butenas:2002ab}.
-They found qualitatively similar trends, namely rFVIIa could restore normal coagulation (even in the absence of TF) for large enough rFVIIa doses, although rFVIIa-induced coagulation 
+They found qualitatively similar trends, namely rFVIIa restored normal coagulation (even in the absence of TF) for large enough rFVIIa doses, although rFVIIa-induced coagulation 
 in hemophilia (even for large rFVIIa doses) lagged the normal profile. 
 These results suggest that rFVIIa administration alone
-might not be able to initiate normal coagulation in recurrent bleeding, unless the dosage is above a critical threshold. 
-Unfortunately, this threshold is likely patient specific, thus potentially limiting the utility of rFVIIa administration.
+might not be able to initiate normal coagulation in recurrent bleeding, unless the dosage is well above a critical threshold.
+However, defining this threshold, which is likely patient specific, is difficult as there is tremendous patient to patient variability even with a normal
+coagulation phenotype \citep{Danforth:2012aa}.
 

### 2014-10-28 Paper_v2.tex
Almost ready for submission

@@
 Thus, reasonable predictions may be possible, despite parametric uncertainty, if a few critical parameters are well-defined. 
-Perhaps more importantly, ODE models also require significant mechanistic information, thereby limiting their utility in poorly understood systems.
-Toward this challenge, logical modeling is an emerging paradigm that encodes causal relationships between model components using non-linear transfer functions, 
-without requiring mechanism \citep{Morris:2010aa}. 
-Despite their simplicity, logical models have captured rich behaviors in a variety of systems important to human health \citep{Saez-Rodriguez:2011aa,Morris:2011ys,Morris:2012aa}. 
+For example, Tasseff et al., showed in a model of Retinoic acid (RA) induced differentiation of HL-60 cells, that correct 
+predictions were possible even when 75\% of the parameters were known only to an order of magnitude \citep{Tasseff:2011aa}. 
+Perhaps more importantly, ODE models require significant mechanistic information, thereby limiting their utility in poorly understood systems, 
+or conversely explode in size when considering multiple pathways or subsystems. 
+Toward this challenge, logical modeling is an emerging paradigm that encodes causal relationships between model components using quasi-mechanistic non-linear transfer functions \citep{Morris:2010aa}. 
+Logical models are highly flexible, and despite their simplicity, they have captured rich behaviors in a variety of systems important to human health \citep{Saez-Rodriguez:2011aa,Morris:2011ys,Morris:2012aa}.
 However, modeling complex dynamics with logical models is challenging. 
@@
 Previous coagulation models have typically been formulated as systems of nonlinear ordinary differential equations, 
-using mass action or more complex kinetics, to describe the rates of biochemical conversions \citep{Kuharsky:2001aa}.
+using mass action or more complex kinetics, to describe the rates of biochemical conversions \citep{Khanin:1989aa,Willems:1991aa,Baldwin:1994aa,Leipold:1995aa,Kuharsky:2001aa}.
 Mechanistic ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys} 
@@
 who developed and then subsequently refined highly mechanistic models of coagulation. 
-Other laboratories have also expanded upon Hockin \emph{et al.}, for example including the intrinsic pathway, and thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
+Other laboratories have also expanded upon Hockin \emph{et al.}, for example by exploring the intrinsic pathway, the role of stochastic fluctuations in coagulation \citep{Lo:2005aa},
+and the dynamics of thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
 Other aspects of coagulation have also been modeled, such as platelet biochemistry \citep{Stalker:2013aa}, 
@@
 The core of our model was based upon the earlier work of Ismagilov and coworkers \citep{Runyon:2004aa,Kastrup:2006aa,Runyon:2007aa,Runyon:2008aa}, 
-where we added initiation, factor dependence, and specific inhibition terms to the earlier simplified models.
+where we added initiation, factor dependence, and specific inhibition terms to the earlier simplified model.
 A trigger event initiates thrombin formation (FIIa) from prothrombin (fII) through a lumped initiation step.
@@
 Simulations with a baseline level of trigger were consistent with the previous hemophilia simulations, where the only thrombin produced was from initiation (Fig. \ref{fig-rFVII}A, 1$\times$~trigger).
-However, as we increased the trigger strength, the thrombin profile began to approximate normal coagulation, showing a pronounced peak albeit with a slower peak time (Fig. \ref{fig-rFVII}B).
+However, as we increased the trigger strength, the thrombin profile began to approximate normal coagulation, showing a pronounced peak albeit with a slower peak time (Fig.
+\ref{fig-rFVII}B, $t^{**}>t^{*}$).
 Further increases in trigger strength resulted in decreased thrombin peak time and increased maximum thrombin values (Fig. \ref{fig-rFVII}A, 50$\times$~trigger). 
@@
 in attenuating the thrombin generation rate. Thus, the transfer function approach encoded the inhibitory logic of this cascade in the absence of specific mechanism.
-
-\clearpage
 
@@
 Traditional enzyme kinetic expressions, e.g., Michaelis–Menten or multiple saturation kinetics are multiplied by an enzyme activity control variable $0 \leq v_{j} \leq 1 $. 
-Control variables are functions of many possible regulatory factors encoded by arbitrary functions of the form $0\leq f_{j}\left(\mathcal{Z}\right)\leq 1$.
-At each simulation time step, the $v_{j}$ variables are calculated by evaluating integration rules such as the max or min of the set of factors $f_{1},\hdots,f_{n}$ 
+Control variables are functions of many possible regulatory factors encoded by arbitrary transfer functions of the form $0\leq f_{j}\left(\mathcal{Z}\right)\leq 1$.
+At each simulation time step, the $v_{j}$ variables are calculated by evaluating integration rules such as the max or min of the set of transfer functions $f_{1},\hdots,f_{n}$ 
 influencing the activity of enzyme $E_{j}$. }\label{fig-control-schematic}
@@
 Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles. 
+The peak thrombin time for normal coagulation ($t^{*}$) is less than rFVIIa induced coagulation in hemophilia ($t^{**}$), while the peak thrombin value was greater in normal coagulation.
 The shaded region denotes the 99\% confidence estimate of the mean thrombin value (uncertainty in the model simulation). 

### 2014-10-30 submission_v1-10-29-14/Paper_v2.tex
Submission on 10-29-14

--- 2014-10-28 Paper_v2.tex
@@
 by minimizing the residual between model simulations and experimental measurements using particle swarm optimization (PSO). 
+Each parameter set in our ensemble corresponded to a unique particle in the PSO.
 We then validated the model ensemble using thrombin data sets that were not used during training. 
@@
 performance of the network. Taken together, the hybrid approach produced a surprisingly predictive model given its small size, 
-suggesting the proposed hybrid framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
+suggesting the proposed framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
 or potentially even cell free metabolic systems.
@@
 were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, Hockin \emph{et al.} \citep{Hockin:2002aa}, and later Butenas et al., \citep{Butenas:2004aa}
-who developed and then subsequently refined highly mechanistic models of coagulation. 
+who developed and then subsequently refined highly mechanistic coagulation models. 
 Other laboratories have also expanded upon Hockin \emph{et al.}, for example by exploring the intrinsic pathway, the role of stochastic fluctuations in coagulation \citep{Lo:2005aa},
@@
 after administration and 95\% of the thrombin was gone by 20 min after initiation. 
-We performed flux analysis to understand how the reduced order coagulation model balanced initiation, amplification and inhibition of thrombin formation for normal coagulation and hemophilia.
+We performed flux analysis to understand how the reduced order coagulation model balanced initiation, amplification and termination of thrombin generation for normal and hemophilic coagulation.
 Analysis of the reaction flux through the reduced order network for thrombin generation in normal, hemophilia and rFVIIa-treated hemophilia identified three distinct operational modes (Fig.
@@
 Thus, ensemble approaches are common in the dynamic modeling community. 
-However, perhaps a unique feature of the current study is the direct connection between our particle swarm approach, 
+However, a unique feature of the current study is the direct connection between our particle swarm approach, 
 and the parameter ensemble; each particle in our swarm uniquely corresponded to a parameter set in our ensemble. 
@@
 We then used flux analysis to understand how the network operated in a variety of conditions, and global sensitivity analysis to identify which parameters controlled the 
-performance of the network. Taken together, the proposed hybrid framework produced a surprisingly predictive model, suggesting this approach could 
+performance of the network. Flux analysis showed the logical rules formulation encoded the transitions between initiation, amplification and termination of thrombin generation.
+Sensitivity analysis suggested that the amplification rate constant was more important to the time to peak thrombin, while the ATIII inhibition constant controlled thrombin exposure.
+Taken together, the proposed hybrid framework produced a surprisingly predictive model, suggesting this approach could 
 be used to effectively model other biochemical networks important to human health.
@@
 
-Although the performance of the proof of principle coagulation model was impressive given its limited size, there were several issues that could be further explored.
+The performance of the proof of principle coagulation model was impressive given its limited size, however, there are several issues that could be further explored.
 First, the prediction of initiation time should be investigated. 
@@
 However, the only mathematical requirement for a transfer function is that it map a non-negative continuous or categorical variable into the 
-range $f\in[0,1]$. Thus, many transfer function types are possible.
+range $f\in[0,1]$. Thus, many types of transfer functions are possible.
 

### 2014-10-30 Paper_v2.tex
Starting final version edits

--- 2014-10-24 Paper_v2.tex
@@
 %\usepackage[authoryear,round,comma,sort&compress]{natbib}
-\usepackage[square,sort&compress,comma,numbers]{natbib}
+\usepackage[square,sort,comma,numbers]{natbib}
 %\usepackage[authoryear,round]{natbib}
@@
 \begin{titlepage}
-{\par\centering\textbf{\Large Dynamic Modeling of the Human Coagulation Cascade using Reduced Order Effective Kinetic Models}}
+{\par\centering\textbf{\Large Modeling and Analysis of Hormone and Mitogenic Signal Integration in Prostate Cancer}}
 \vspace{0.05in}
-{\par \centering \large{Adithya Sagar and Jeffrey D. Varner$^{*}$}}
+{\par \centering \large{ Katharine V. Rogers, Joseph A. Wayman, Ryan Tasseff, Timothy Chu, Kyriakos Tsahalis, Spencer Davis, Rohit Yallamraju, Gongcheng Lu, Matthew P. DeLisa, 
+and Jeffrey D. Varner$^{*}$}}
 \vspace{0.10in}
@@
 \vspace{0.1in}
-{\par \centering \textbf{Running Title:}~Reduced order modeling of blood coagulation}
+{\par \centering \textbf{Running Title:}~Modeling signal transduction in prostate cancer}
 \vspace{0.1in}
-{\par \centering \textbf{To be submitted:}~\emph{Processes}}
+{\par \centering \textbf{To be submitted:}~\emph{Molecular Systems Biology}}
 \vspace{0.5in}
@@
 \section*{Abstract}
-In this study, we present a novel hybrid modeling approach which combines ordinary differential equation (ODE) modeling with logical rules to simulate an archetype biochemical network, 
-the human coagulation cascade.  
-The model consisted of five differential equations augmented with several logical rules describing 
-regulatory connections between model components, and unmodeled interactions in the network. 
-This formulation was more than an order of magnitude smaller than current coagulation models, 
-because many of the mechanistic details of coagulation were encoded as logical rules.
-We estimated an ensemble of likely model parameters (N = 20) from \textit{in~vitro} extrinsic coagulation data sets, with and without inhibitors, 
-by minimizing the residual between model simulations and experimental measurements using particle swarm optimization (PSO). 
-We then validated the model ensemble using thrombin data sets that were not used during training. 
-The ensemble predicted thrombin trajectories for conditions not used for model training, including thrombin generation for normal and hemophilic coagulation 
-in the presence of platelets (a significant unmodeled component).
-We then used flux analysis to understand how the network operated in a variety of conditions, and global sensitivity analysis to identify which parameters controlled the 
-performance of the network. Taken together, the hybrid approach produced a surprisingly predictive model given its small size, 
-suggesting the proposed hybrid framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
-or potentially even cell free metabolic systems.
+Prostate cancer is the most common cancer in men and the second leading cause of cancer related death in the United States. 
+Androgens, such as testosterone, are required for prostate cancer growth. 
+Androgen ablation in combination with radiation or chemotherapy remains the primary non-surgical treatment for androgen-dependent prostate cancer. 
+However, androgen ablation typically fails to permanently arrest cancer progression, often resulting in castration resistant prostate cancer (CRPC). 
+CRPC is closely related to metastasis and decreased survival. 
+In this study, we developed and analyzed a population of mathematical models describing growth factor and hormone signal integration in androgen dependent, intermediate and resistant prostate cancer cells.
+The model describes the integration of two simultaneous extracellular signaling inputs, namely, androgen and growth factors into a G1/S cell cycle checkpoint decision. 
+Model parameters were identified from 43 studies in androgen dependent and resistant LNCaP cell lines. 
+The model was validated by comparing simulations with an additional 29 data sets from LNCaP cell lines that were not used during training. 
+Additionally, data from four drug trials was also used to evaluate the model’s performance.
+Sensitivity analysis, conducted over an ensemble of prostate signaling models, suggested that in an androgen free environment general translation and transcription was more sensitive in androgen dependent cells, while in androgen independent cells the PI3K and MAPK pathway species were more sensitive.
+In a constant DHT environment sensitive species were conserved between the cell lines. 
+These results suggest targeting the PI3K and MAPK pathways in addition to anti-androgen therapies as a treatment for AIPC.
 
-\vspace{0.25in}
-{\noindent \textbf{Keywords:}~Blood coagulation, Mathematical modeling, Systems biology}
+\vspace{0.5in}
+{\noindent \textbf{Keywords:}~Prostate cancer, signal transduction, mathematical modeling}
 
@@
 
-%We chose coagulation because it is well studied, has a complex architecture, 
-%and has an abundance of experimental data available for model identification and validation. 
-%However, coagulation was just a proof-of concept to validate our approach; 
-%
-
-%Developing mathematical modeling tools to interrogate the molecular basis of human disease is one of the grand challenges facing systems biology.
-%Developing computational tools to understand the molecular basis of human disease is a grand challenge facing systems biology.
-%Many have suggested that the integration of experimental and computational research is required to unravel
-%critical questions facing molecular medicine \cite{KITANO2002,ASSMUS2006,ARNAUD2006}. 
-%Despite this, mechanistic network modeling has not played a significant role in the development of new therapies for cancer, cardiovascular diseases or for acute events  
-%like thrombosis during surgery.
-%A critical limiting issue often cited has been model uncertainty \cite{BAILEY2001}. 
-%Development and validation of detailed network models requires a significant investment of time and material resources. 
-%Alternatively, development of tools using toy models often fails to demonstrate relevance to biologists or physicians.  
-%What is needed is a model system, that is simultaneously tractable and directly applicable to the study of disease.
-%Human blood is perhaps such an ideal model system \cite{Diamond:2009yf}. 
-%Blood is a complex mixture of proteins and cells which collectively carry out many functions.
-%However, one well studied facet of blood, which could serve as a valuable model system, is the coagulation cascade. 
-%Understanding the mechanism of rFVIIa might provide insight into the therapeutic dose under different physiological conditions.
-%More generally, rFVIIa treatment could also serve an ideal test case to develop strategies which retrospectively estimate operational perturbations associated with drugs.
-%For example, the effect of rFVIIa is directly observable in coagulation outputs such as thrombin concentration.
-%Thus, by analyzing these outputs in a normal versus a perturbed network, it might be possible to reverse engineer the molecular influence associated with a therapeutic intervention.
+\section*{Introduction}
 
 %
-%Thus, while there are many possible strategies to simulate biochemical networks, ODE models remain a common choice despite
-%significant identification challenges.
-%This result highlights the perhaps under-appreciated role that experimental design could play in generating the best training and validation data sets for model identification \citep{Apgar:2010fk}.
+%Sipuleucel-T is a first generation cancer vaccine \cite{Kantoff2010}, while radium-223 is a targeted alpha emitter for bone metastasis \cite{Parker2013}. 
+%Abiraterone is an inhibitor of cytochrome P450 17A1 (CYP17), a critical enzyme in androgen biosynthesis \cite{Sartor2013} while enzalutamide inhibits androgen receptor (AR) translocation to the nucleus, as well as DNA binding and coactivator recruitment \cite{Scher2012}.
+%Thus, understanding the operation of the biological pathways which ultimately facilitate the transition to CRPC is vital to the discovery of new therapeutic strategies against this disease.
+%The model focuses on the outlaw pathway, in which constitutively or overly-activated receptor tyrosine kinases (RTKs) stimulate AKT and mitogen-activated protein kinases (MAPKs), which in turn activate cytosolic androgen receptor (AR) in the absence of androgen.
+Interestingly, among the few genes activated AR represses is cellular prostatic acid phosphatase (cPAcP), itself a key regulatory of RTK activation.
+Thus, the outlaw pathway encodes crosstalk between RTK-dependent kinase and AR activation inside of a positive feedback loop.
+The model was validated with multiple experiments in LNCaP cell lines.
+In addition  For example, in response to enzalutamide, an AR inhibitor, the model predicted a decline in PSA expression similar to patient response.
+%and there is limited data on combination therapies or the use of biomarkers to develop personalized treatment strategies.
 
-\section*{Introduction}
-Developing mathematical models of biochemical networks is a significant facet of systems biology.
-Modeling approaches differ in their degree of detail, where the choice of approach is often determined by prior knowledge, or model requirements \citep{2012_kholodenko_kolch_SciSig}. 
-Ordinary differential equation (ODE) models are common tools for modeling biochemical systems because of their ability to capture dynamics and encode mechanism.
-However, ODE models typically come with difficult (or sometimes impossible) parameter identification problems. 
-For example, Gadkar \textit{et al.}, showed that even with near-perfect information, 
-it was often impossible to identify all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
-However, it is not clear whether we actually need precise estimates for all model parameters. 
-Bailey suggested more than a decade ago, that achieving qualitative or even quantitative understanding of biological systems 
-should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}. 
-Since Bailey's complex biology with no parameters hypothesis, Sethna showed that model performance is typically 
-sensitive to only a few parameters, a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
-Thus, reasonable predictions may be possible, despite parametric uncertainty, if a few critical parameters are well-defined. 
-Perhaps more importantly, ODE models also require significant mechanistic information, thereby limiting their utility in poorly understood systems.
-Toward this challenge, logical modeling is an emerging paradigm that encodes causal relationships between model components using non-linear transfer functions, 
-without requiring mechanism \citep{Morris:2010aa}. 
-Despite their simplicity, logical models have captured rich behaviors in a variety of systems important to human health \citep{Saez-Rodriguez:2011aa,Morris:2011ys,Morris:2012aa}. 
-However, modeling complex dynamics with logical models is challenging. 
-Thus, there is an unmet need for a third approach which combines ODEs and logical models, where ODEs could encode mechanistic information, 
-while missing or incomplete mechanistic knowledge can be approximated using a logical approach.
+Prostate cancer (PCa) is the most commonly diagnosed cancer and the second leading cause of cancer-related death in men in the United States \cite{Siegel2013}. 
+Initially, PCa cells depend upon the activation of cytosolic androgen receptors (AR) by androgen hormones, such as testosterone, for survival and growth. 
+Androgen ablation in combination with radiation or chemotherapy remains the primary non-surgical treatment for androgen-dependent prostate cancer (ADPC) \cite{Huggins1967}. 
+However, androgen ablation typically fails to permanently arrest cancer progression as malfunctioning cells eventually lose androgen sensitivity and proliferate without hormone. 
+The loss of androgen sensitivity results in castration resistant prostate cancer (CRPC), a phenotype closely linked with metastasis and greatly reduced survival \cite{Harris2009}.
+Currently, there are six approved treatments that demonstrate a survival advantage in patients with metastatic CRPC, each of these target diverse aspects of the disease \cite{Sartor2013}.
+The taxane family members docetaxel and cabazitaxel interact with microtubule stability \cite{Tannock2004,Bono2010}, while abiraterone \cite{Sartor2013} or enzalutamide \cite{Scher2012} interfere with androgen signaling by blocking androgen formation or nuclear translocation, respectively.
+Other approved treatments are non-specific to PCa. 
+For example, general treatments such as sipuleucel-T, a first generation cancer vaccine \cite{Kantoff2010}, 
+and radium-223, an alpha emitter which targets bone metastasis \cite{Parker2013}, are both approved to treat CRPC.
+Unfortunately, regardless of the therapeutic approach, the survival advantage of these treatments is typically only a few months. 
+Thus, understanding the molecular basis of the loss of androgen sensitivity in CRPC could be an important step for the development of the next generation of therapies with a prolonged survival
+advantage.
 
-In this study, we developed a hybrid approach which combined ODE modeling with logical rules to model a well studied biochemical network, 
-the human coagulation system. Coagulation is an archetype proteolytic cascade involving both positive 
-and negative feedback \cite{Butenas:2002aa,Schenone:2004aa,Adams:2009aa}. 
-Coagulation is mediated by a family proteases in the circulation, called factors and a key group of blood cells, called platelets. 
-The central process in coagulation is the conversion of prothrombin (fII), an inactive coagulation factor, to the master protease thrombin (FIIa).
-Thrombin generation involves three phases, initiation, amplification and termination \cite{GOLDHABER2006, Brummel:2002aa}.
-Initiation requires a trigger event, for example vessel injury, which leads to the activation of factor VII (FVIIa).
-Two converging pathways, the extrinsic and intrinsic cascades, then process and amplify this initial coagulation signal. 
-The extrinsic cascade is generally believed to be the main mechanism of thrombinogenesis in the blood \cite{MANN1990,ROBERTS1998,MANN1999}.
-Initially, thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa), which itself has been activated by TF/FVIIa \cite{Butenas:2002aa}. 
-Picomolar amounts of thrombin then activate the cofactors factors V and VIII (fV and fVIII) and platelets, 
-leading to the formation of the tenase and prothrombinase complexes on activated platelets.
-These complexes amplify the early coagulation signal by further activating FXa, and directly converting prothrombin to thrombin. 
-There are several control points in the cascade that inhibit thrombin formation, and eventually terminate thrombin generation. 
-Tissue Factor Pathway Inhibitor (TFPI) inhibits FXa formation catalyzed by TF/FVIIa, while antithrombin III (ATIII)
-neutralizes several of the proteases generated during coagulation, including thrombin.  
-Thrombin itself also inadvertently plays a role in its own inhibition; thrombin, through interaction with thrombomodulin, protein C and endothelial cell protein C receptor (EPCR),
-converts protein C to activated protein C (APC) which attenuates the coagulation response by proteolytic cleavage of fV/FVa and fVIII/FVIIIa. 
-Termination occurs after either prothrombin is consumed, or thrombin formation is neutralized by inhibitors such as APC or ATIII.
+Androgen-induced proliferation and survival depends upon many coordinated signal transduction and gene expression events. 
+Androgen Receptor (AR) is part of the nuclear hormone receptor superfamily, which includes other important cancer targets such as progesterone receptor (PR) and estrogen receptor (ER) in breast cancer \cite{Aranda2001}.
+Nuclear hormone receptors act as ligand dependent transcription factors interacting with specific DNA sequences of target genes as either monomers, heterodimers, or homodimers; AR, PR, and ER act as homodimers.
+In the case of AR these specific DNA sequences are known as androgen response elements (ARE) \cite{Mangelsdorf1995}.
+In the absence of androgen, AR is predominately found in the cytoplasm bound to heat shock proteins (HSP) \cite{Prescott2006}.
+%Nuclear Hormone Receptors (Reviewed \cite{Aranda2001}) and AR in disease reviewed in \cite{Matsumoto2013}
+Androgen, either testosterone or testosterone metabolites such as 5$\alpha$-dihydrotestosterone (DHT), enter prostate cells and interact with the cytosolic androgen receptor (AR). 
+The interaction of DHT with AR promotes the dissociation of AR from chaperones such as HSP \citep{Pratt:1997kx} and its subsequent dimerization, phosphorylation and translocation to the nucleus (reviewed by Brinkmann \textit{et al.} \citep{Brinkmann1999}).
+Activated nuclear AR drives a gene expression program broadly referred to as androgen action, that promotes both proliferation and survival.
+In addition to many genes including itself, activated nuclear AR promotes the expression and secretion of prostate specific antigen (PSA), arguably the best known PCa biomarker \citep{Feldman2001}.
+PSA is commonly used as a prostate cancer indicator, although its prognostic ability is controversial \citep{Attard2009,Hoffman2011,Moyer2012}. 
+In CRPC, AR signals in the absence of androgens. 
+Androgen dependent (AD) prostate cells can become castration resistant (CR) through several possible mechanisms, including constitutively amplified AR expression and altered AR sensitivity to testosterone or other non-androgenic molecules \citep{Feldman2001}.
+In this study, we focused on the aberrant activation of AR by kinase signaling cascades, sometimes called the outlaw pathway. 
+Outlaw pathway activation is driven by over-activated receptor tyrosine kinases (RTKs), a common pathology in many cancer types including PCa \citep{Slamon1989,Craft1999}. 
+RTKs stimulate downstream kinases, including the AKT and mitogen-activated protein kinase (MAPK) pathways, which promote AR phosphorylation and dimerization in the absence of an androgen signal \citep{Craft1999,Yeh1999}.
+Interestingly, among the few genes activated AR represses is cellular prostatic acid phosphatase (cPAcP), itself a key regulatory of RTK activation \cite{Veeramani2005}. 
+Thus, in CRPC the androgen program is initiated without the corresponding extracellular hormone cue, potentially from crosstalk between growth factor and hormone receptor pathways.
 
-Previous coagulation models have typically been formulated as systems of nonlinear ordinary differential equations, 
-using mass action or more complex kinetics, to describe the rates of biochemical conversions \citep{Kuharsky:2001aa}.
-Previous ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys} 
-were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, 
-and later Hockin \emph{et al.} \citep{Hockin:2002aa}, who developed and then refined highly mechanistic models of extrinsic coagulation. 
-Other laboratories have also expanded upon Hockin \emph{et al.}, for example including the intrinsic pathway, and thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
-Other aspects of coagulation have also been modeled, such as platelet biochemistry \citep{Stalker:2013aa}, 
-multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and transport inside clots \citep{Voronov:2013aa}.
-However, these previous studies were largely based upon extensive mechanistic knowledge. 
-This is possible because blood, while enormously complex, can be systematically interrogated.
-Other systems, such as intracellular signaling networks, are much more difficult to experimentally interrogate.  
-Towards this unmet need, we formulated a hybrid modeling approach which combines ODEs and logical rules to model 
-biochemical processes for which a complete mechanistic understanding is missing. 
-We tested this approach by modeling the human coagulation cascade. 
-The hybrid model consisted of only five differential equations augmented with several logical rules.
-Thus, the model was more than an order of magnitude smaller than comparable purely ODE models in the literature.
-We estimated the model parameters from \textit{in~vitro} extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
-We then compared the model predictions with thrombin data sets, for both normal and hemophilic coagulation, that were not used for model training.
-Once validated, we performed flux and sensitivity analysis on the model to estimate which parameters were critical to model performance in several conditions. 
-The reduced order hybrid approach produced a surprisingly predictive coagulation model, 
-suggesting this framework could potentially be used to model other biochemical networks important to human health.
+In this study, we developed a mathematical model of growth factor and hormone signal integration in androgen dependent, intermediate and resistant prostate cancer cells. 
+We used this model to better understand which components and processes were differentially important in AD versus CR cells.
+The new model architecture was a significant advance over our previous prostate signaling model \citep{Tasseff2010}. 
+We added the regulated expression of ten additional proteins, including the cell cycle restriction point proteins cyclin D (and the differentially spliced variants cyclin D1a and cyclin D1b), cyclin E, cyclin-dependent kinase inhibitor 1A (p21Cip1), and cyclin-dependent kinase inhibitor 1B (p27Kip1).
+Also, we included the Rb/E2F pathway, expanded our description of the activation of the mammalian target of rapamycin (mTOR) protein and its role in translation initiation, and included the regulation of AR action by cyclin D1a and E2F.
+However, this upgraded architecture, while increasing the biological scope of the model, also expanded the number of unknown model parameters. 
+To estimate these parameters, we used multiobjective optimization in combination with dynamic and steady-state data sets generated in AD, intermediate and CR LNCaP cell lines. 
+We identified a population of approximately N = 5000 models (from well over a million candidate models) which described both AD and CR data sets using a single model structure.
+Furthermore, we tested the model using an additional 29 LNCaP data sets not used for model training, along with four drug studies.
+We analyzed the model population using sensitivity and robustness analysis to uncover differentially important mechanisms in AD versus CR cell lines. 
+In the presence of androgen, the sensitivity profile was similar between AD and CR cells. 
+Components of the MAPK and PI3K pathways were highly fragile, irrespective of the level of androgen dependence. 
+However, in the absence of androgen, there were 609 statistically significant shifts in species sensitivity between AD and CR cells.
+Of these, 108 were larger than one standard deviation above the mean.
+In CR cells, HER2 activation of the MAPK and PI3K pathways was significantly more important, as was AR activation through the MAPK pathway. 
+On the other hand, components of the translation and transcription infrastructure were differentially more important in AD cells in the absence of androgen.  
+Taken together, our analysis suggested that independently targeting the PI3K or MAPK pathways in combination with anti-androgen therapies could perhaps be an effective treatment strategy for CRPC.
+
+%(323 more and 287 less sensitive)
+%In this study, we rank-ordered the importance of the molecular components controlling the transition from ADPC to CRPC.
+%Toward this, we developed and analyzed an ensemble of mechanistic mathematical models of the signal transduction processes governing the 
+%transition from androgen dependent to castration resistant prostate cancer.
+%The objective of this study was to rank-order the importance of components in CRPC and ADPC using computational tools. 
+%We formulated and analyzed an ensemble of mechanistic mathematical models of the signal transduction processes in CRPC and ADPC 
+%that was an extension of our previous work .
+%In particular, we focused on the outlaw pathway and the possible crosstalk between RTK-dependent kinases and AR activation. 
+%First, we expanded the original model by adding ten  
+%Third, we added additional co-activators, such as cyclin E, and inhibitors of AR, such as
+%ODE models typically contain a large number of unknown model parameters, which must be estimated  
+%Due to uncertainty in parameter estimation and to account for differences between individual cells and cell sublines, an ensemble of parameter sets was used to model the system. 
+%Proteins that were not modeled at a gene regulatory level were generated using zeroth order reactions. 
+%All protein and RNA species were degraded with first order reaction rates. 
+%To obtain an ensemble of parameter sets, we used multiobjective optimization. 
+%Relative concentration values were quantified 
+%Training data came from experiments on LNCaP sublines. 
+%We used , previously developed in the lab \cite{Song2010}, to determine a diverse ensemble of models. 
+%C-33 cells, having undergone fewer than 33 passages, are considered AD while C-51 cells, having between 34 and 80 passages, are considered an intermediate prostate cancer cell line. 
+%C-81 (passage numbers beyond 80) display a CR phenotype \cite{Lin2000}. 
+%http://endo.endojournals.org/content/140/11/5054.long or http://endo.endojournals.org/content/140/11/5054.long#ref-19
 
@@
 
-%However, components could also correspond to non-modeled influences, categorial variables or other abstract quantities.
-%In the current proof-of-principle model, 
-%APC can inhibit amplification, and upstream of amplification thus APC activity will directly slow the rate of thrombin formation from the initiation trigger, and amplification itself.
-%On the other hand, ATIII inhibits thrombin directly.
+\subsection*{Estimating an ensemble of prostate signaling models.} 
+We modeled the integration of growth factor, cell cycle and hormone signaling pathways in AD and CR LNCaP cells (Fig. \ref{fg:NetworkSchematic}). 
+The signaling architecture was hand curated from over 80 primary literature sources in combination with biological databases.
+The model equations were formulated as a system of ordinary differential equations (ODEs), where biochemical reaction rates were modeled using mass action kinetics.
+ODEs and mass action kinetics are common modeling tools \citep{Wayman2013}, however, ODEs have the disadvantage of requiring estimates for unknown model parameters.
+Many techniques have been developed to estimate ODE model parameters, often from noisy and sparse experimental data \citep{Moles:2003dn}.
+Typically these identification problems are underdetermined, hence no unique parameter values can be estimated \citep{Villaverde:2014kl}.
+Thus, instead of estimating a single yet highly uncertain parameter set, we estimated an ensemble of possible parameter sets using the Pareto Optimal Ensemble Techniques (POETs) algorithm \citep{Song2010}.
+POETs uses a combination of simulated annealing and local optimization techniques coupled with Pareto optimality-based ranking to simultaneously optimize multiple objective functions.
+Starting from an initial best fit set, we estimated the 1687 unknown model parameters (1674 kinetic parameters and 13 non-zero initial conditions) using 43 \textit{in vitro} data sets taken from AD, intermediate and CR LNCaP cells (Table~\ref{objective_table}).
+Each of the 43 training data sets was a separate objective in the multiobjective calculation. 
+The training data were steady-state or dynamic immunoblots from which we extracted relative species abundance from their optical density profiles. 
+POETs sampled well over a million possible parameter sets, from which we selected N = 5000 sets for further analysis. 
+Over the ensemble, the coefficient of variation (CV) of the kinetic parameters spanned 0.5 - 5.8, with 33\% of the parameters having a CV of less than one (Fig. ZZ). 
+As a control, we also performed simulations for R = 100 random parameter sets to compare against the ensemble generated by POETs.
 
-\subsection*{Formulation of reduced order coagulation models.}
-We developed a reduced order extrinsic coagulation model to test our hybrid modeling approach (Fig. \ref{fig-network}).
-A trigger event initiates thrombin formation (FIIa) from prothrombin (fII) through a lumped initiation step.
-This step loosely represents the initial activation of thrombin by activated FXa.
-Once activated, thrombin catalyzes its own formation (amplification step), and inhibition via the conversion of protein C to activated protein C (APC).
-Antithrombin III (ATIII) inhibits amplification, while APC and tissue factor pathway inhibitor (TFPI) potentially inhibit both initiation and amplification. 
-All initiation and inhibition processes, as well as the dependence of amplification upon other coagulation factors, was approximated using our rule-based approach (Fig. \ref{fig-control-schematic}). 
-Individual regulatory contributions to the activity of pathway enzymes were integrated into control coefficients ($v$'s) using an integration rule (min/max).  
-These control coefficients then modified the rates of model processes at each time step.
-Hill-like transfer functions $0 \leq f\left(\mathcal{Z}\right) \leq 1$ quantified the contribution of components upon a target process. 
-Components were either individual inhibitor or activator levels or some function of levels, e.g., the product of factor levels. 
-In this study, $\mathcal{Z}$ corresponded to the abundance of individual inhibitors or activators, 
-with the exception of the dependence of amplification upon specific coagulation factors (modeled as the product of factors). 
-When a process was potentially sensitive to multiple inputs, logical integration rules were used to select which transfer functions influenced the process at any given time. 
-In our proof of concept model, we used a winner takes all strategy; the maximum or minimum transfer function was selected at any given time step. 
-However, other integration rules are certainly possible. 
-Taken together, while the reduced order coagulation model encodes significant biological complexity, it is highly compact (consisting of only five differential equations). 
-Thus, it will serve as an excellent proof of principle example to study the reduction of a highly complex human subsystem.
+The ensemble of PCa models recapitulated training data in both AD and CR cell lines with only two experimentally justified parameter changes (Fig. \ref{fg:ColorPlot} and Fig. \ref{fg:Training}). 
+Data from the LNCaP clones C-33 (dependent), C-51 (intermediate), and C-81 (resistant) \cite{Lin2000,Igawa2002,Horoszewicz1983} along with the CR LNCaP cell lines LNCaP-Rf \cite{Murillo2001}, LNCaP-AI \cite{Chen2011} and LNAI \cite{Graff2009} were used for model identification.
+To simulate the effective difference between LNCaP cell lines, the parameter controlling the maximum rate of PAcP gene expression was scaled by 0.1 and 0.5, respectively, for the C-81 and C-51 cell-lines compared to C-33.
+This modification was based upon steady-state PAcP data from the three LNCaP clones \cite{Lee2003}. 
+Similarly, the expression of p16INK4 was adjusted in accordance with the study of Lu \textit{et al.} \citep{Lu1997}. 
+These two parameters were the only adjustable parameter differences between AD and CR cells. 
+To simulate an increased mTOR activation in the presence of a DHT stimulus, we added a first order activation term for mTOR activation with a DHT stimulus.
+Androgens have been shown to increase expression of proteins involved in cellular metabolism, which may lead to an increase in mTOR activation \cite{Xu2006}.
+The model fit 36 of the 43 training objectives for greater than 40\% of the ensemble members (Fig. \ref{fg:ColorPlot}A). 
+Conversely, only 10 of the 43 training objectives were captured with the random parameter control (Fig. \ref{fg:ColorPlot}B). 
+The model captured the crosstalk between RTK activation and androgen action (Fig. \ref{fg:Training}). 
+The model described DHT-induced PSA expression in both C-33 (Fig. \ref{fg:Training}A) and C-81 (Fig. \ref{fg:Training}B) cells. 
+Interestingly, simulations with the HER2 inhibitor AG879 recapitulated decreased PSA expression in C-81 cells (Fig. \ref{fg:Training}C) in the absence of androgen stimulation.
+AR action decreased the PAcP mRNA message (Fig. \ref{fg:Training}D), presumably leading to increased HER2 activity. 
+The model also recapitulated the integration of androgen action with AR expression, G1/S cell cycle protein expression and AKT phosphorylation. 
+For example, the model captured AR-induced AR expression following a DHT stimulus (Fig. \ref{fg:Training}H).
+Conversely, the transcription factor E2F inhibits AR transcription in LNCaP cells (Fig. \ref{fg:Training}I). 
+Other cell cycle proteins were also integrated with androgen action.
+For example, the cyclin D1 abundance increased in CR compared to AD cells in the absence of androgen (Fig.
+\ref{fg:Training}E), while DHT induced p21Cip1 expression in C-33 cells (Fig.
+\ref{fg:Training}F).
+The level of phosphorylated AKT was also increased in higher passage number cells (Fig. \ref{fg:Training}G). Taken together, [FINISH ME].
 
-\subsection*{Identification of model parameters using particle swarm optimization.}
-A critical challenge for any dynamic model is the estimation of kinetic parameters. 
-We estimated kinetic and control parameters simultaneously from eight \emph{in vitro} time-series coagulation data sets with and without the protein C pathway. 
-The residual between model simulations and experimental measurements was minimized using particle swarm optimization (PSO). 
-A population of particles (N = 20) was initialized with randomized kinetic and control parameters and allowed to search for parameter vectors that minimized
-the residual. However, not all parameters were varied simultaneously. 
-We partitioned the parameter estimation problem into two subproblems based upon the biological organization of the training data; 
-(i) estimation of parameters associated with thrombin formation in the absence of the protein C pathway and (ii) estimation of parameters associated with the protein C pathway.
-Only those parameters associated with each subproblem were varied during the optimization procedure for that subproblem, e.g., thrombin parameters were \textit{not} varied during the protein C subproblem.The PSO procedure was run for 20 generations for each subproblem, where each generation was 1200 iterations.
-The best particle from each generation was used to generate the particle population for the next generation.
-We rotated the subproblems, starting with subproblem 1 in the first generation.
+%We used a single model architecture to describe the behavior of both AD and CR LNCaP cells, 
+%Two new additions to the model were the cell cycle protein p21Cip1 and transcriptional regulation of AR. 
+%We were able to capture training data for both mechanisms. 
+%AR transcriptionally upregulates its own expression \cite{Lin2003}\cite{Grad1999}. 
+%We captured the increase in AR concentration due to AR activation, and thus increased transcriptional activity, upon  
+%Higher basal AKT phosphorylation leads to increased activation of the PI3K/AKT pathway in PCa cells with less androgen dependence \cite{Lin2003}. 
+%\cite{Davis2006}
 
-The reduced order coagulation model captured the role of initial prothrombin abundance, and the decay of the thrombin signal following from ATIII activity (Fig. \ref{fig-diff_fII_NO_APC}).
-However, we systematically under-predicted the thrombin peak and the strength of ATIII inhibition in this training data set.
-On the other hand, with fixed thrombin parameters, we captured peak thrombin values and the decay of the thrombin signal (at least for the 150\% fII case) in the presence
-of both ATIII and the protein C pathway (Fig. \ref{fig-diff_fII_APC}). Lastly, we were unable to capture global differences in initiation time \textit{across} separate data 
-sets with a single ensemble of model parameters. These differences likely resulted from normal experimental variability.
-For example, different thrombin generation experiments within the 
-training data (at the same physiological factor levels) had significantly different initiation times (data not shown). 
-However, the inability to globally capture initiation time also highlighted a potential shortcoming of the initiation module within the model. 
-To capture the variability in initiation time \textit{across} training data sets, we included a constant time-delay parameter ($T_{D}$) for each data group.
-The delay parameter was constant within a data set, but allowed to vary \textit{across} training data sets. 
-Introduction of the delay parameter allowed the model to simulate multiple training data sets using a single ensemble of model parameters.
-Taken together, the model identification results suggested 
-that our hybrid approach could reproduce a panel of thrombin generation data sets in the neighborhood of physiological factor and inhibitor concentrations.
-However, it was unclear whether the reduced order model could predict new data, without updating the model parameters.
+%An additional protein we considered in validation that was present in training was phosphorylated SHC-transforming protein 1 (Shc).
+%Simulations predicted a similar increase in Shc phosphorylation after DHT stimulus (Fig. \ref{fg:Validation}D) \cite{Veeramani2005Onc}. 
+%The Prostate Cancer Working Group (PCWG) updated outcome measurements for castration resistant prostate cancer studies in 2008 \cite{Scher2008}. 
+%The primary endpoint in prostate cancer studies is overall survival, while PSA response is often a secondary endpoint. 
+%Figure \ref{fg:Validation}I shows fold change of PSA of ensemble members due to the application of sorafenib.
 
-\subsection*{Validation of the reduced order coagulation model.}
-We tested the predictive power of the reduced order coagulation model with validation data sets not used during model training. 
-Two validation data sets were used, thrombin generation for various prothrombin and ATIII concentrations with the protein C pathway, 
-and thrombin generation in normal versus hemophilic plasma in the presence of the protein C pathway. 
-Lastly, we compared the qualitative output of the model to rFVIIa addition in the presence of hemophilia.
-The hemophilia case was an especially difficult test as it was taken from a different study which used a plasma-based
-\textit{in~vitro} assay involving platelets instead of phospholipid vesicles (PCPS). 
-All kinetic and control parameters were fixed for the validation simulations. 
-The only globally adjustable parameter $T_{D}$, was fixed within each validation data set but allowed to vary between data sets.
-The reduced order model predicted the thrombin generation profile for ratios of prothrombin and ATIII in the absence of the protein C pathway (Fig. \ref{fig-fII_ATIII}).
-Simulations near the physiological range (fII,ATIII) = (100\%, 100\%) or (125\%,75\%) tracked the measured thrombin values (Fig. \ref{fig-fII_ATIII}B and C).
-On the other hand, predictions for factor levels outside of the physiological range (fII,ATIII) = (50\%, 150\%) or (150\%, 50\%), while qualitatively consistent with measured thrombin values, did show significant
-deviation from the measurements (Fig. \ref{fig-fII_ATIII}A and D).
-Likewise, simulations of thrombin generation in normal versus hemophilia (missing both fVIII and fIX) were consistent with measured thrombin values (Fig. \ref{fig-hemophilia}).
-We modeled the dependence of thrombin amplification on factor levels using a product rule ($\mathcal{Z} = fV\times fX\times fVIII\times fIX$), 
-which was then was integrated using a $\min$ integration rule into the control variable governing amplification. 
-Thus, in the absence of fVIII or fIX, the amplification control variable evaluated to zero, and the only thrombin produced was from initiation (Fig. \ref{fig-hemophilia}B).
-However, the decay of the thrombin signal was underpredicted in the normal case (Fig. \ref{fig-hemophilia}A), while the activated thrombin level was overpredicted 
-in hemophilia simulations, although thrombin generation was far less than normal (Fig. \ref{fig-hemophilia}B). 
-Taken together, the reduced order model performed well in the physiological range of factors, even with
-unmodeled components such as platelet activation in the hemophilia data set.
+%(Fig. \ref{fg:Validation}H)
 
-The model ensemble predicted a direct correlation between thrombin generation and rFVIIa addition in hemophilia (Fig. \ref{fig-rFVII}).
-In the current model, we cannot distinguish between different initiation sources, e.g., TF/FVIIa versus rFVIIa, as we have only a single lumped initiation source (trigger).
-Thus, we simulated the addition of rFVIIa in hemophilia by removing fVIII and fIX from the model, and modulating the initial level of trigger.
-Simulations with a baseline level of trigger were consistent with the previous hemophilia simulations, where the only thrombin produced was from initiation (Fig. \ref{fig-rFVII}A, 1$\times$~trigger).
-However, as we increased the trigger strength, the thrombin profile began to approximate normal coagulation, showing a pronounced peak albeit with a slower peak time (Fig. \ref{fig-rFVII}B).
-Further increases in trigger strength resulted in decreased thrombin peak time and increased maximum thrombin values (Fig. \ref{fig-rFVII}A, 50$\times$~trigger). 
-Thus, for large trigger values (200$\times$trigger), the hemophilic thrombin profile approximated normal coagulation, where peak thrombin was achieved shortly 
-after administration and 95\% of the thrombin was gone by 20 min after initiation. 
-We performed flux analysis to understand how the reduced order coagulation model balanced initiation, amplification and inhibition of thrombin formation for normal coagulation and hemophilia.
-Analysis of the reaction flux through the reduced order network for thrombin generation in normal, hemophilia and rFVIIa-treated hemophilia identified three distinct operational modes (Fig.
-\ref{fig-flux-analysis}).
-We calculated the flux through four lumped reactions, initiation, amplification, thrombin-induced APC generation and total thrombin inhibition (including both APC and ATIII action).
-Directly after the addition of a trigger (e.g., TF/FVIIa or rFVIIa), the lumped initiation flux was the largest for all three cases.
-However, within a few minutes enough thrombin was generated by the initiation mechanism to induce the amplification stage. 
-During amplification, thrombin catalyzes its own formation and inhibition by generating activated protein C (APC), a potent inhibitor of the coagulation cascade.
-For normal coagulation, amplification and thrombin inhibition were the dominate reactions by 6 min after initiation (Fig. \ref{fig-flux-analysis}, left). 
-After 10 min, the dominate reaction had shifted to thrombin inhibition (both ATIII and APC action). 
-In hemophilia (missing both fVIII and fIX), the amplification reaction did not occur, and thrombin was produced only by initiation (Fig. \ref{fig-flux-analysis}, center).
-Initiation was quickly inhibited by APC, and the thrombin level stabilized (eventually decaying at longer times because of ATIII activity). 
-Lastly, when 50$\times$trigger was used to induce thrombin formation in hemophilia (absence of fVIII/fIX), 
-initiation mechanisms dominated for up to 6 min following initiation (Fig. \ref{fig-flux-analysis}, right). 
-Similar to hemophilia alone, no amplification occurred in the 50$\times$trigger+hemophilia case, 
-and the rate of thrombin generation was extinguished by the combined action of ATIII and APC. 
-Taken together, the hybrid modeling approach captured the transition between the modes of thrombin generation, as well as the role that inhibitors play 
-in attenuating the thrombin generation rate. Thus, the transfer function approach encoded the inhibitory logic of this cascade in the absence of specific mechanism.
+\subsection*{Validation simulations revealed missing network structure.} 
+The model was validated against 29 \textit{in vitro} and four \textit{in vivo} studies (Table~\ref{prediction_table}). 
+For 15 of the 29 cases, greater than 40\% of the ensemble was qualitatively consistent with the experimental data (Fig. \ref{fg:ColorPlot}C). 
+However, for the random parameter control, only 7 of the 29 cases were satisfied (Fig. \ref{fg:ColorPlot}D). 
+We correctly predicted positive feedback between HER2 auto-activation and androgen action (Fig. \ref{fg:Validation}A and Fig. \ref{fg:Validation}B). 
+We also captured the dose-dependence of AR abundance on DHT (Fig. \ref{fg:Validation}C). 
+In addition to the cell line studies, we simulated the outcome of enzalutamide, lapatinib, and sorafenib clinical trials in AD and CRPC patients. 
+The trial end points were the reduction in PSA expression relative to an untreated baseline.  
+Enzalutamide acts on AR by inhibiting its nuclear translocation, DNA binding, and coactivator recruitment \cite{Scher2012}. 
+In the enzalutamide trial, 54\% of the patients that received the drug showed a PSA decline of $\geq$ 50\% while 25\% showed a decline $\geq$ 90\%. 
+We simulated enzalutamide exposure by reducing the rate constants governing activated AR binding to nuclear importer, cyclin E, and CDK6 to 1\% of their initial values. 
+Consistent with the trial, 62\% of ensemble members showed a $\geq$ 50\% decline in PSA abundance, while 14\% showed a $\geq$ 90\% decline (Fig. \ref{fg:Validation}G). 
+The second trial we simulated involved exposure of CRPC patients to sorafenib. 
+Sorafenib is a kinase inhibitor with activity against Raf, vascular endothelial growth factor receptor (VEGFR), platelet-derived growth factor receptor (PDGFR), c-kit and c-Ret \cite{Dahut2008}. 
+We considered only the effects of sorafenib on the protein kinase Raf, as VEGFR, PDGFR, c-kit and c-Ret were not included in the model. 
+None of the 22 patients in the sorafenib study showed a PSA decline of $>$ 50\%. 
+However, our simulations showed that approximately 55\% of the ensemble members had a PSA decline of $\geq$ 50\%. 
+The last drug we considered was lapatinib, an inhibitor of epidermal growth factor receptor (EGFR) and HER2 tyrosine kinase activity \cite{Liu2013}. 
+Two lapatinib drug trials were considered: one in which patients had CRPC and one in which patients had biochemically relapsed ADPC \cite{Whang2013,Liu2013}. 
+In the CRPC lapatinib drug trial, two of the 21 enrolled patients had a PSA response $\geq$ 47\%  \cite{Whang2013}. 
+For the CRPC case, our model showed 26.5\% of ensemble members with a PSA response $\geq$ 47\%. 
+Of the 35 patients enrolled in the ADPC lapatinib study, no PSA decreases was observed \cite{Liu2013}. 
+In this case, our model showed 9.2\% of ensemble members with a PSA response $\geq$ 50\%.
+Although no response to lapatinib was seen in ADPC clinical trials, \textit{in vitro} AD LNCaP experiments showed decreased PSA expression in response to lapatinib, most notably with the addition of DHT \cite{Liu2005}.
 
-\clearpage
+Validation and training failures suggested the original signaling architecture was missing critical components. 
+Several of the failed training and validation simulations involved the response of the network to epidermal growth factor (EGF) stimulation. 
+For example, Chen \textit{et al.} showed that HER2 phosphorylation increased within five minutes following EGF stimulation of LNCaP-AI cells \citep{Chen2011}. 
+We predicted no connection between HER2 phosphorylation and EGF stimulation on this short timescale (Fig. \ref{fg:Validation}E). 
+Interestingly, we initially neglected the heterodimerization of HER2 with other ErbB family members in order to simplify the model.
+However, Chen \textit{et al.} suggested that HER2-EGFR heterodimerization could be an important factor in EGF-driven activation of HER2 \cite{Chen2011}. 
+We tested this hypothesis by developing a new model that included HER2 and EGFR heterodimerization. 
+We set the rate constants governing the assembly of HER2/EGFR heterodimers equal to EGFR homodimer assembly; all other parameters were unchanged.
+This was a reasonable first approximation, as the affinity of HER2/EGFR heterodimerization and EGRF homodimerization is thought to be similar \cite{Hendriks2003}. 
+With the inclusion of HER2-EGFR heterodimerization, we qualitatively fit the EGF-induced HER2 activation case and improved our training for experiments that involved an EGF stimulus, e.g., cyclin D mRNA and protein abundance following an EGF stimulus in C-33 cells (Fig \ref{fg:ColorPlot}A and C, white pixels).
 
-\subsection*{Global sensitivity analysis of the reduced order coagulation model.}
-We conducted a global sensitivity analysis to estimate which parameters controlled the performance of the reduced order model. 
-We calculated the sensitivity of the time to maximum thrombin (peak time) and the thrombin exposure (area under the thrombin curve) 
-for different levels of prothrombin, and protein C (Fig. \ref{fig-sensitivity}). 
-Globally, 41\% of the parameters shifted in importance between the (fII,PC) = (50\%, 0\%) and (150\%,100\%) cases for the peak thrombin time (Fig. \ref{fig-sensitivity}A). 
-The majority of these shifts involved the interaction between increased prothrombin and the protein C pathway, while only
-5\% were directly associated with increased prothrombin alone. The rate constant for thrombin amplification was the most important parameter controlling the
-peak thrombin time. While this parameter was differentially important for different prothrombin levels, and in the presence or absence of 
-the activated protein C pathway, it was consistently the most sensitive parameter in the model. The saturation constant governing thrombin amplification was
-the second most important parameter, followed by the initiation control gain parameter. Other important parameters influencing the thrombin peak time included the control gain for
-activated protein C formation, and the rate constant controlling ATIII inhibition of thrombin activity. 
-On the other hand, only 27\% of the model parameters were differentially sensitive between the (fII,PC) = (50\%, 0\%) and (150\%,100\%) cases 
-for thrombin exposure (Fig. \ref{fig-sensitivity}B). Of these parameters, all of the shifts were associated with the interplay between thrombin formation and the protein C pathway. 
-The rate constant controlling ATIII inhibition was the most important parameter controlling the thrombin exposure.
-While this parameter was less important in the presence of protein C for 150\% prothrombin levels, it was significantly above all other parameters. 
-Similar to the peak time, for 150\% prothrombin, the control gain for activated protein C formation was differentially important along with the rate constant controlling
-amplification. However, the amplification parameter was much less important for thrombin exposure versus peak time.
+%in diverse environments (i.e., DHT, no DHT, drug) 
+%Stimuli included the presence or absence of DHT, presence of therapeutics, and also EGF. 
+%PIP3 recruits AKT and PDK1 to the membrane where PDK1 activates AKT \cite{Vivanco2002}. 
+%Thus, the PIP3 and AKT interaction is required in the model for AKT phosphorylation to occur.
 
-\clearpage
+\subsection*{Sensitivity analysis identified differentially important features of the prostate architecture.} 
+We used sensitivity analysis to identify important signaling components in AD versus CR cells (Fig. \ref{fg:Sensitivity}). 
+We calculated first order steady-state sensitivity coefficients under different stimuli for 500 parameter sets selected from the ensemble. 
+Signaling components were rank-ordered based upon analysis of their sensitivity coefficient values. 
+In the presence of DHT, the sensitivity profile was similar for AD versus CR cells, with only a few differences (Fig. \ref{fg:Sensitivity}B). 
+The top 2\% of sensitive species, regardless of androgen dependence, involved components from the MAPK and PI3K pathways. 
+In particular, activated Ras, Raf, phosphorylated MEK, as well as PIP3 localized AKT, phosphorylated AKT, and PI3K were sensitive in both AD and CR cells. 
+Species involving PAcP and p16INK4 were more sensitive in AD cells, which was expected since the expression of these two proteins were the only parameters changed between AD and CR cells. 
+Other species such as E2F, cyclin E, and DHT-activated AR were also more sensitive in AD cells. 
+On the other hand, HER2-Grb2-Gab activation of PI3K and AKT inhibition of RAF were more sensitive in CR cells.
 
-%Controlling hemorrhage has been a driving force to understand the molecular basis of blood coagulation.
-%Replacement therapy improves bleeding times for most hemophilia patients. However, repeated fVIII/fIX administration can lead to neutralizing inhibitors. 
-%Alternatively, rFVIIa has been used as a universal hemostatic agent to initiate clotting with and without fVIII or fIX inhibitors \cite{Hedner:2008rm}.
-%Despite its wide adoption, the mechanism and the utility of rFVIIa in trauma-associated hemorrhage remains controversial \cite{Duchesne:2008hw}. 
-%The model described the extrinsic activation of thrombin, in the presence of ATIII, with and without the protein C pathway.
-%The coagulation network architecture was based on literature and a previous mathematical model from our laboratory \cite{VARNER2007}.
-%Simulations of thrombin dynamics were independently validated using blood drawn from patients with coronary artery disease (reported here) and previous TF/FVIIa and FVIIa 
-%\emph{in vitro} studies \cite{BUTENAS2002}.
+The importance of signaling components varied with androgen dependence in the absence of DHT (Fig. \ref{fg:Sensitivity}A). 
+There were 609 statistically significant shifts in species sensitivity (318 more and 291 less sensitive) between CR and AD cells in a non-androgen environment. 
+However, only 108 of these shifts were greater than one standard deviation above the mean. 
+In CR cells, HER2 activation of ERK and PI3K was more sensitive, as was AR activation through the MAPK pathway. 
+This was expected, as outlaw pathway activity was elevated in castration resistant cells. 
+Species in the MAPK pathway were in general more sensitive in CR cells (128 out of 140 significant), with all forms of sPAcP more robust in CR cells.
+On the other hand, infrastructure pathways encoding transcription and translation were more sensitive in AD cells.
+PSA  and cyclin D1b (mRNA and mRNA complexes) were the only species involved in translation that were more robust in AD cells (14 out of 116).
+The transcription factor, E2F was more fragile in AD cells, while the transcription factors ETS and AP1 were more robust. 
+ETS and AP1 are activated by phosphorylated ERK, and ETS is also activated by active PKC \cite{Wilkinson2000,Lindemann2003}. 
+E2F is deactivated through binding to Rb, which is deactivated by cyclin D1 and CDK phosphorylation \cite{Lapenna2009}. 
+The model also included AP1 suppression of AR transcriptional activity (more sensitive in CR) \cite{Sato1997}, as well as inhibition of transcription of the AR gene by E2F (more sensitive in AD) \cite{Davis2006}.
+Species in the PI3K pathway that were more fragile in AD cells included Rheb and TOR complexes. 
+Interestingly, these species were included as the last step in the PI3K pathway prior to translation, with the phosphorylation of 4E-BP1 by TOR being considered the beginning of translation in this model.
+This again indicates that in the absence of DHT general translation is more fragile in AD cells.
 
-%Additionally, using sensitivity analysis on the ensemble of models, we identified 
-%structurally sensitive components of the coagulation architecture as a function of condition.
-%We also demonstrated that the ensemble robustly constrained model predictions of independent validation sets, despite having many poorly constrained parameters.
-%Thus, the model ensemble displayed \emph{sloppy} behavior similar to that observed by Sethna and coworkers for other signal transduction networks \cite{Gutenkunst2007}.  
-%By studying a family of models, we perhaps partially addressed the uncertainty stemming from the many poorly characterized model parameters.
-%However, many other factors could influence our results. For example, missing structural interactions or biophysical factors such as blood flow could play a large role.
-%Thus, while the results presented here may be a valuable first step, more studies are required.
-% \cite{ALLEN2007}.
+There was a large shift in sensitive species between an androgen and a non-androgen environment in both AD and CR cell lines (Fig. \ref{fg:Sensitivity}C and Fig. \ref{fg:Supp_Sensitivity}). 
+Of the 664 statistically significant shifts in AD cells, 288 were more sensitive between androgen versus non-androgen environments. 
+However, only 119 shifts were larger than one standard deviation above the mean. 
+Unsurprisingly, AR activation through DHT binding, with and without coactivators, in a DHT environment was more sensitive, as was AR inhibition of PAcP transcription (repressed by AR in the model). 
+Species further upstream, such as HER2 activation of the MAPK and PI3K/AKT pathways, were also more sensitive in a DHT environment. 
+Cell cycle species that were more fragile in the presence of DHT, included complexes involving p21Cip1 and CDC25A.
+In a non-androgen environment, basal transcription (68 out of 72) and translation (114 out of 120) were more fragile. 
+Other fragile species in the absence of DHT included Rb, E2F, Sam68, cyclin D1a complexes, phosphatases in the MAPK pathway, Rheb complexes, and TOR complexes.
 
-%We used this ensemble of parameters in all subsequent simulations to estimate the uncertainty in model simulations following from uncertainty in the model parameters.
+We also considered the sensitivity of CR cells following the application of the AR inhibitor enzalutamide in the presence of DHT (Fig. \ref{fg:Sensitivity}D). 
+In an androgen environment with enzalutamide, species which were more fragile included cytosolic AR, cPAcP, and p21Cip1. 
+As we would expect, AR species found in the nucleus and/or bound to coactivators, were more robust in the absence of enzalutamide. 
+The top two percent of sensitive species with and without enzalutamide were conserved. 
+In a CR cell, enzalutamide had no effect on the sensitivity of PI3K/AKT species as well as many MAPK species (ERK, Raf, and MEK). 
+Next, we looked at the effect of enzalutamide on a CR cell in both a non-androgen and DHT environment (Fig. \ref{fg:Supp_Sensitivity}). 
+More sensitive species in a non-androgen environment included dimerized HER2, ERK, and PAcP. 
+Species which were more robust in the non-androgen environment included, AR activated by DHT, AKT, p70, and AR bound to HSP.
+The results of our sensitivity analysis indicate that instead of inhibiting solely the AR pathway (enzalutamide), a combination therapy targeting the PI3K or MAPK pathways in addition to AR may be more effective.
+
+\subsection*{Robustness analysis identified key regulators of prostate cancer.} 
+Robustness analysis was conducted for 80 proteins to quantify the effects of amplifying or knocking down key model components in both AD and CR cells. 
+Gene expression parameters were altered by a factor 10, 0.5, and 0 for knock-in, knock-down, or knock-out perturbations, respectively.
+We calculated the effect of these perturbations on different protein markers, such as PSA, AR, and cyclin D. 
+The results of the model suggest that an inhibition of either the PI3K pathway or the MAPK pathway in combination with an AR inhibitor as a possible therapy for CRPC. 
+Sensitivity analysis revealed no change in the top sensitive species in the presence or absence of the AR inhibitor, enzalutamide.
+PI3K/AKT and MAPK species continued to fall in the top two percent of sensitive species.
+A study by Carver \textit{et al.} looked at dual inhibition of AR and PI3K signaling in LNCaP cells and in a Pten-deficient murine prostate cancer model \cite{Parker2013}. 
+Using both the PI3K inhibitor, BEZ235, and the AR inhibitor, MDV3100 (enzalutamide), the group saw a drastic decrease in the total number of cells. 
+Each inhibitor on it’s own had a much smaller effect on total cell number. 
+They saw an increase in the cell death marker, c-PARP, in the dual inhibition case. 
+The group hypothesized that AKT inhibition leads to increased AR signaling activity through increased protein concentrations of HER3. 
+On the other hand, AR inhibition leads to increased AKT activity due to the down regulation of PHLPP, a protein phosphatase that regulates AKT. 
+For simplicity, the HER3 pathway and also cell death were not included in the model. 
+Dual knock-out studies of PI3K and AR in our model show no additive effect on any cell cycle proteins through the dual knock-out. 
+The pathways in our model appear to be uncoupled and therefore no synergy is shown in the dual knock-out case. 
+This could indicate that the combined decrease in cell population is entirely due to cell death. 
+The Carver \textit{et al.} study did not consider cell cycle proteins or cell growth. 
+Our model does show a decrease in cell cycle proteins in the PI3K knock-out as well as in the dual knock-out case. 
+This result seems to be consistent with the decreased cell growth in the PI3K knock-out case which is not dependent on cell death, as c-PARP levels are low. 
+The decrease in cell cycle proteins is due to a decrease in general translation, including free eIF4E levels and activated 40S ribosome subunit. 
+The decrease in p70 (S6) activation due to inhibition of PI3K is shown in both the model and in the Carver \textit{et al.} study (Supplementary Figure), indicating this result is due to the PI3K pathway.
+It would be interesting to repeat the experiment looking at cell cycle proteins and also performing the experiment in CR LNCaP cells, instead of AD cells.
+
+A knock-out of Raf, MEK or ERK showed an overall increase in cyclin D levels in CR cells (Fig. \ref{fg:Supp_Robustness}). 
+This was unexpected and we saw a similar increase in cyclin D due to the knock-in of Raf, MEK or ERK. 
+We found that individual ensemble members showed different response to a Raf knock-out, in both cyclin D concentration and PSA concentration. 
+Of the 500 ensemble members, 126 members saw an increase in PSA concentration and 62 members saw an increase in cyclin D concentration due to the knock-out of Raf (Fig. \ref{fg:Robustness}). 
+We saw three distinct regions: (1) increase in PSA concentration, (2) increase in cyclin D concentration, and (3) a decrease in both PSA and cyclin D. 
+We explored the flux vectors of the outlying parameter sets to understand the mechanistic effect of Raf knock-out on PSA and cyclin D. 
+Outlying parameter sets in region 1 displayed high activation of PI3K through HER2 signaling as well as high association of AP1 with AR. 
+AP1 binds and suppresses AR transcriptional activity in LNCaP cells \cite{Sato1997}. 
+Knocking out Raf lowered AP1 levels and, therefore, freed AR for increased transcription of PSA.
+In region 2, parameter sets also had high activation of PI3K through HER2. 
+They also had higher association of E2F with Rb and cyclin D1a with AR. 
+Cyclin D levels in this region increase due to an increase in E2F levels caused by the Raf knock-out. 
+Parameters in region 3 have high association of TOR. 
+Interestingly, the drug sorafenib, a multi-kinase inhibitor that has activity against Raf, showed no measurable PSA decline in prostate cancer patients in clinical trials \cite{Dahut2008}. 
+The robustness analysis showed that network perturbation can result in unexpected responses due to cell-to-cell heterogeneity in gene expression. 
+These outlying cell types could be critical for understanding when designing drug targets and combination therapies.
 
 \section*{Discussion}
-In this study, we developed a reduced order model of the human coagulation cascade.
-We chose coagulation because it is well studied, has a complex architecture, 
-and has an abundance of experimental data available for model identification and validation. 
-However, coagulation was just a proof of concept test of our approach. 
-The proposed hybrid framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
-or potentially even cell free metabolic systems. 
-The model consisted of five differential equations augmented with several logical rules describing 
-regulatory connections between model components and unmodeled interactions in the network. 
-We estimated model parameters from \textit{in vitro} extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
-To estimate parameters, the residual between model simulations and experimental measurements was minimized using particle swarm optimization (PSO). 
-However, not all of the model parameters were uniquely identifiable, given the training data.
-Instead, we estimated an ensemble of likely parameter sets (N = 20)
-from eight \emph{in vitro} time-series coagulation data sets with and without the protein C pathway.
-We validated the ensemble using thrombin data sets taken from multiple laboratories for a variety of experimental conditions not used during training. 
-The ensemble predicted thrombin trajectories for conditions not used for model training, including thrombin generation for normal and hemophilic coagulation 
-in the presence of platelets (a significant unmodeled component).
-We then used flux analysis to understand how the network operated in a variety of conditions, and global sensitivity analysis to identify which parameters controlled the 
-performance of the network. Taken together, the proposed hybrid framework produced a surprisingly predictive model, suggesting this approach could 
-be used to model other biochemical networks important to human health.
+In this study, we developed a population of mathematical models describing growth factor and hormone signal integration in androgen dependent, intermediate and resistant prostate cancer cells. 
+These models described the regulation of androgen receptor expression and activation through androgen binding as well as a ligand-independent, MAPK-driven mechanism referred to as the outlaw pathway. 
+An ensemble of model parameters was estimated using 43 steady-state and dynamic data sets taken from androgen dependent, intermediate and independent LNCaP cell lines using multiobjective optimization.
+Further, we tested the predictive power of the model by comparing model predictions against 33 novel data sets (including four \textit{in vivo} drug studies) not used during model training. 
+The model ensemble captured 84\% of the training data and 52\% of the validation data relative to 23\% and 24\% for a random control population. 
+Interestingly, during the the initial round of parameter estimation, we identified several potentially missing structural components not present in the original connectivity. 
+One such component, EGF-induced HER2/EGFR heterodimerization, was added to the current generation model. 
+Inclusion of this structural component significantly improved both training and validation performance using the same rate constants as the EGFR-homodimer case (no additional parameter fitting). 
+We then analyzed the population of signaling models, using both sensitivity and robustness analysis, to identify the critical components controlling network performance in a variety of conditions.
 
-Malfunctions in coagulation can have potentially fatal consequences.
-Aggressive clotting involved with Coronary Artery Diseases (CADs), collectively accounts for 38\% of all deaths in North America \cite{HANSSON2005}.
-Coagulation management during surgery can also be challenging, particularly with the increase in clinical use of antithrombotic drugs \cite{Tanaka:2009wo}.   
-Insufficient coagulation due to genetic disorders such as hemophilia can also result in recurrent bleeding. 
-The coagulation factors VIII (fVIII) and IX (fIX) are deficient in Hemophilia A and B, respectively \cite{COOPER1994,MANNUCCI2001,MITCHELL2008}.
-People with mild hemophilia have 5-40\% of the normal clotting factor levels while severe hemophiliacs have $<$1\% \cite{MITCHELL2008}.
-Hemophilia can be controlled with regular infusions of the deficient clotting factors. 
-However, clotting factor replacement sometimes leads to the formation of fVIII and fIX inhibitors \emph{in vivo} \cite{TOMOKIYO2003}.
-Alternatively, recombinant factor VIIa (rFVIIa) has been used to treat bleeding disorders \cite{Hedner:2008rm,Talbot:2009kk} including
-hemophilia with and without factor VIII/IX inhibitors \cite{Shapiro:2008px}.
-However, rFVIIa requires frequent administration (every 2-3 hr), and many questions remain about its mechanism of action, 
-its effective dosage \cite{TOMOKIYO2003}, and its overall utility for the treatment trauma-associated hemorrhage \cite{Duchesne:2008hw}. 
-In this study, we did not model rFVIIa-induced coagulation directly. 
-Rather, we modeled a general trigger which initiated the extrinsic coagulation cascade. 
-Since we identified the model using TF/FVIIa, inherent to our rFVIIa simulations (and the rate constant governing initiation) 
-was the presence of TF. However, even with this complication, 
-the model generated potentially useful insight into the rFVIIa mechanism of action, and its possible shortcomings especially for the treatment of hemophilia. 
-The addition of rFVIIa directly activated thrombin through the initiation pathway. 
-However, no amplification of the thrombin signal occurred without fVIII or fIX. 
-Thus, the peak thrombin signal was lower than normal coagulation, the peak thrombin time was longer, and thrombin generation was eventually inhibited by the combined action of ATIII and the protein C pathway.
-However, as the dose of rFVIIa increased, the peak thrombin time decreased (eventually saturating around 200$\times$nominal trigger), and the peak thrombin value increased such that
-the thrombin profile resembled normal coagulation.
-Butenas \emph{et al.} performed an extensive \textit{in vitro} study of rFVIIa-induced thrombin generation under normal and hemophilic conditions \citep{Butenas:2002ab}.
-They found qualitatively similar trends, namely rFVIIa could restore normal coagulation (even in the absence of TF) for large enough rFVIIa doses, although rFVIIa-induced coagulation 
-in hemophilia (even for large rFVIIa doses) lagged the normal profile. 
-These results suggest that rFVIIa administration alone
-might not be able to initiate normal coagulation in recurrent bleeding, unless the dosage is above a critical threshold. 
-Unfortunately, this threshold is likely patient specific, thus potentially limiting the utility of rFVIIa administration.
+%Hendrick's \textit{et al.} hypothesized that HER2 heterodimerization amplified signals by allowing just one molecule of EGF to activate two receptors \cite{Hendriks2003}. 
+%In our case, we see an amplified cyclin D concentration due to an equal amount of EGF stimulus. %
+%Model connectivity involving EGF and the downstream effect on AR and AR-regulated genes also came up as an open question. 
+%Whether EGF downregulates or upregulates AR activity is debatable and the biology is still unknown \cite{Cai2009,Cinar2005,Lin2003,Lin2001PNAS, Hakariya2006,Gregory2004,Ponguta2008,Guo2006,Weber2004}.
+%The current model assumed no direct effect of EGF on AR, and instead we saw increased AR activity through increased HER2 activation. 
+%Further studies will need to be done to update this section of the model.
 
-Although the performance of the proof of principle coagulation model was impressive given its limited size, there were several issues that could be further explored.
-First, the prediction of initiation time should be investigated. 
-We were able to estimate initiation time within a data set, but unable to predict initiation time \textit{across} independent data sets. 
-This suggested that we should update the initiation module to distinguish between different triggers, e.g., TF/FVIIa versus rFVIIa alone, 
-and to include key biological milestones such as FXa activation (a prerequisite to thrombin formation).
-Next, there are several additional biological modules that could be added to the core model presented here. 
-First, we could include thrombin-induced platelet activation and the role of activated platelets in amplification.
-We captured thrombin generation data in the presence of platelets, however,
-the initial shape of the activation curve and the time-scale of activation was not always consistent with the data.
-Platelets are activated by thrombin through the cleavage of the extracellular domain of protease-activated receptors (PARs) on the platelet surface.
-Once activated, platelets play an important role in amplification, and are key mediators of the positive feedback driving amplification.
-Thus, this biology is a potentially important component of an expanded model.
-We should also add the intrinsic pathway to the model. 
-The intrinsic pathway is triggered by contact activation of the plasma protease factor XI (fXI) by negatively charged surfaces and by thrombin and upstream factors such as activated
-plasma protease factor XII (FXIIa) \citep{NAITO1991,Gailani:1991aa}.
-Activated platelets may also release polyphosphate which directly activates fXII \cite{Smith:2006zl}. 
-Arguably a minor player in acute bleeding, contact activation could also be important in other wound healing contexts. 
-Finally, to make the model more clinically relevant, we should include the biochemical processes responsible for clot formation and clot dissolution (fibrinolysis).
-Clot formation is driven by thrombin activity, while fibrinolysis is driven by plasmin activity, a key enzyme that cleaves fibrin (one of the main materials in a clot).
-Similar to coagulation, fibrinolysis is managed by several activating and inhibitory factors which control the balance between clot formation and dissolution.
-Tissue plasminogen activator (t-PA) and urokinase activate plasmin, along with contact pathway factors such as fXIa. On the other hand, thrombin activatable fibrinolysis inhibitor (TAFI)
-inhibits the degradation of fibrin by plasmin. Also, similar to coagulation, there is considerable fibrinolysis and contact pathway data sets that can be used to train the model.
-Lastly, the choice of max/min integration rules or the particular form of the transfer functions could be generalized to include other rule types and functions. 
-Theoretically, an integration rule is a function whose domain is a set of transfer function inputs, and whose range is $v\in[0,1]$.
-Thus, integration rules other than max/min could be used, such as the mean or the product, assuming the range of the transfer functions is always $f\in[0,1]$.
-Alternative integration rules such as the mean might have different properties which could influence model identification or performance. 
-For example, a mean integration rule would be differentiable, which allows derivative-based optimization approaches to be used. 
-The particular form of the transfer function could also be explored. We choose a Hill-like function because of its 
-prominence in the systems and synthetic biology community. 
-However, the only mathematical requirement for a transfer function is that it map a non-negative continuous or categorical variable into the 
-range $f\in[0,1]$. Thus, many transfer function types are possible.
+In addition, three of the validation cases involved the effect of EGF on AR and AR-activated genes, i.e., PSA. 
+Cai \textit{et al.} showed decreased expression of endogenous AR as well as androgen-regulated PSA in AD LNCaP cells due to an EGF stimulus \cite{Cai2009}. 
+Cinar \textit{et al.} also showed decreased AR protein expression due to EGF, an effect reversed by the mTOR inhibitor, Rapamycin \cite{Cinar2005}. 
+Model simulations show either the opposite trend or no effect due to EGF stimulus (Fig. \ref{fg:Validation}F) \cite{Cinar2005}. 
+These results suggest missing network structure. 
+From additional literature searches, the inhibition of AR activation through EGF is still an open question, with many groups debating the biology involved, predominately in the PI3K/AKT pathway. 
+Lin \textit{et al.} found that in low passage number LNCaP cells (C-33), AKT negatively regulates AR by destabilizing it and marking it for ubiquitylation. 
+In high passage number LNCaP (C-81), AKT levels are high which contribute to AR stability and less degradation \cite{Lin2003}. 
+Wen \textit{et al.} showed that HER2 could induce AKT activation and LNCaP cell growth in the presence and absence of androgen \cite{Wen2000}. 
+Another study shows AKT phosphorylation of AR at S213 and S790 suppresses AR transactivation and AR-mediated apoptosis of LNCaP \cite{Lin2001PNAS}. 
+The study from Cai \textit{et al.} showed the reduction in AR was not due to degradation or PI3K/AKT signaling, but instead was due to decreased AR mRNA levels \cite{Cai2009}. 
+They found that AR protein levels in CR cells were not affected by EGF. 
+Others though have found that PSA expression, even in C-81 cells, is decreased by EGF \cite{Hakariya2006}. 
+In other prostate cell lines, EGF has been shown to increase AR transactivation \cite{Gregory2004,Ponguta2008}. 
+The MAPK pathway, which is downstream of EGFR, may also enhance AR responses to low levels of androgen \cite{Guo2006,Weber2004}. 
+Due to the discrepancies in the literature, experiments should be performed before adding additional network connectivity to the model.
+
+The population of PCa models was analyzed using sensitivity analysis to identify key signaling components and processes in both AD and CR cells. 
+There was very little difference between sensitive and robust components in AD versus CR cells in the presence of androgen. 
+MAPK and PI3K pathway components were consistently ranked in the top 2\% of sensitive species in the presence of androgen for both AD and CR cells.
+On the other hand, cell cycle species, such as cyclin D-CDK4/6 complexes bound to cell cycle inhibitors (p27Kip1, p21Cip1, p16INK4), were consistently robust.
+However, this profile changed considerably in the absence of androgen.
+The activation of PI3K and ERK by HER2 dimerization and autophosphorylation was significantly more important in CR versus AD cells.
+Interestingly, AR activation by ERK was also more sensitive in CR versus AD cells in the absence of androgen.
+Lastly, although AR-regulated transcriptional processes were equally sensitive between the cell types, general translational and transcriptional components were more robust in CR versus AD cells.
+This evidence supports the current theory that CR cells will still respond to androgen and, thus, AR is still an active target in therapeutic against CRPC \cite{Karantanos2013}. 
+Supporting the argument that AR can be activated in the absence of androgens by MAPK activation \cite{Feldman2001}. 
+Advanced prostate cancers often have higher levels of E2F and other transcription factors \cite{Davis2006}.
+Interestingly, E2F was more sensitive in AD cells, while other transcription factors (ETS and AP1) were more robust. 
+The drug enzalutamide had no effect on the top 2\% of sensitive species. 
+Species in the PI3K/AKT and MAPK pathways in the presence of enzalutamide were still highly sensitive. 
+The application of enzalutamide, increased sensitivity of AR species found outside of the nucleus as well as PAcP species.
+Robustness analysis indicated diverse effects of Raf knock-out on PSA and cyclin D concentrations. 
+Clinical studies of sorafenib, a multi-kinase inhibitor that has activity against Raf, showed increase PSA levels in patients \cite{Dahut2008}. 
+Our results indicate that cell-to cell heterogeneity in gene expression can play a significant role in determining cell response. 
+Thus, combination therapies need to be considered even in the case of a Raf knock-out.
+
+The results of the model suggest that an inhibition of either the PI3K pathway or the MAPK pathway in combination with an AR inhibitor as a possible therapy for CRPC. 
+Sensitivity analysis revealed no change in the top sensitive species in the presence or absence of the AR inhibitor, enzalutamide. 
+PI3K/AKT and MAPK species continued to fall in the top two percent of sensitive species. 
+A study by Carver \textit{et al.} looked at dual inhibition of AR and PI3K signaling in LNCaP cells and in a Pten-deficient murine prostate cancer model \cite{Parker2013}. 
+Using both the PI3K inhibitor, BEZ235, and the AR inhibitor, MDV3100 (enzalutamide), the group saw a drastic decrease in the total number of cells. 
+Each inhibitor on it’s own had a much smaller effect on total cell number. They saw an increase in the cell death marker, c-PARP, in the dual inhibition case. 
+The group hypothesized that AKT inhibition leads to increased AR signaling activity through increased protein concentrations of HER3. 
+On the other hand AR inhibition leads to increased AKT activity due to the down regulation of PHLPP, a protein phosphatase that regulates AKT. 
+For the simplicity of this model, the HER3 pathway and also cell death were not included in the model. 
+Dual knock-out studies of PI3K and AR in our model show no additive effect on any cell cycle proteins through the dual knock-out. 
+The pathways in our model appear to be uncoupled and therefore no synergy is shown in the dual knock-out case. 
+This could indicate that the combined decrease in cell population is entirely due to cell death. 
+The Carver \textit{et al.} study did not consider cell cycle proteins or cell growth. 
+Our model does show a decrease in cell cycle proteins in the PI3K knock-out as well as in the dual knock-out case. 
+This result seems to be consistent with the decreased cell growth in the PI3K knock-out case which is not dependent on cell death, as c-PARP levels are low. 
+The decrease in cell cycle proteins is due to a decrease in general translation, including free eIF4E levels and activated 40S ribosome subunit. 
+The decrease in p70 (S6) activation due to inhibition of PI3K is shown in both the model and in the Carver \textit{et al.} study (Supplementary Figure), indicating this result is due to the PI3K pathway.
+It would be interesting to repeat the experiment looking at cell cycle proteins and also performing the experiment in CR LNCaP cells, instead of AD cells.
+
+%and  in the CR CWR-R1 prostate cell line.
+%a steroid receptor coactivator family member along with other important AR coregulators such as 
+%transcriptional intermediary factor 2 (TIF-2)
+
+The PCa signaling architecture was assembled after extensive literature review and hand curation of the biochemical interactions.
+However, there are a number of areas where model connectivity could be refined, e.g., the regulation of AR phosphorylation.
+We assumed a single canonical activating AR phosphorylation site (S515), with ERK being the major kinase and PP2A or PP1 being the major phosphatases responsible for regulating this site.  
+MAPK activation following EGF treatment increases AR transcription and cell growth, partially through AR phosphorylation on MAPK consensus site S515 \cite{Ponguta2008}.
+However, there are at least 13 phosphorylation sites identified on AR, with phosphorylation at six of these being androgen induced \cite{Gioeli2012}. 
+Moreover, other kinases such as AKT, protein kinase C (PKC) family members, as well as Src-family kinases can all phosphorylate AR in prostate cells \cite{Guo2006,Ponguta2008}. 
+For example, AKT activation leads to AR  phosphorylation at both S213 and S791 (however, the role of these sites remains unclear) \cite{Wen2000,Lin2001PNAS,Taneja2005,Lin2003}.  
+AKT effects on AR may also be passage number dependent, with AKT repressing AR transcription in low passage number cells and enhancing transcription in higher passage numbers \cite{Lin2003}. 
+Androgen independent phosphorylation of AR by Src family kinases (not currently in model) at Y534 \cite{Guo2006} or by protein kinase C (PKC) family members at the consensus site S578 could also be important for understanding the regulation of AR activity.
+A second area we will revisit is the gene expression program associated with androgen action, and particularly the role of AR coregulators.
+Currently, we included only two AR coactivators, cyclin E and CDK6 \cite{Yamamoto2000,Lim2005} and three corepressors AP1, Cdc25A, and cyclin D1a in the model \cite{Sato1997,Chiu2009,Petre-Draviam2003}.
+However, there are at least 169 proteins classified as potential AR coregulators \cite{Heinlein:2002qy,Heemers2007} with many of these being differentially expressed in malignant cells.
+For example, the expression of steroid receptor coactivator-1 (Src-1) and transcriptional intermediary factor 2 (Tif-2), both members of the steroid receptor coactivator family, are elevated in prostate cancer \cite{Gregory2001,Gregory2004}.
+Src-1 is phosphorylated by MAPK and interacts directly with AR to enhance AR-mediated transcription \cite{Heemers2007}. 
+Another class of potentially important AR coregulators are the cell cycle proteins Cdc25 and Rb.
+Unlike Cdc25A, Cdc25B (not in the model) can act as an AR coactivator leading to enhanced AR transcription activity \cite{Ngan2003}. 
+The Rb protein, in addition to being a key cell cycle regulator, has been shown to be an AR coactivator in an androgen-independent manner in DU145 cells \cite{Yeh1998}.
+However, there is some uncertainty about the role of Rb as Sharma \textit{et al.} showed that Rb decreased AR activation in multiple prostate cancer cell lines and xenografts \cite{Sharma2010}. 
+Forkhead proteins have also been shown to activate as well as repress AR function. 
+In prostate cancer, AKT suppresses AFX/Forkhead proteins, which diminishes expression of AFX target genes, such as p27Kip1 \cite{Graff2000, Brunet1999, Medema2000, Takaishi1999}. 
+Lastly, undoubtedly there are several other signaling axes important in PCa, such as cytokine or insulin- and insulin-like growth factor signaling \citep{Cardillo2003,Heinlein2004,Tam2007,Seaton2008}.
+Understanding the pathways associated with these signals and how they relate to the current model, may give us a more complete picture of CR prostate cancer.
+
+%Lastly, with an extended model, it would be beneficial to include possible biomarkers for drug outcomes, as currently there is no biomarker that is used to select treatments for patients  \cite{Sartor2013}.
 
@@
 
+\subsection*{Prostate model signaling architecture.}
+We modeled the expression, translation and post-translational modifications of key components of the signaling architecture.  
+The model, which consisted of 780 protein, lipid or mRNA species interconnected by 1674 interactions, was a significant extension to our previous model \cite{Tasseff2010} in several important areas. 
+First, we included well-mixed nuclear, cytosolic, membrane and extracellular compartments (including transfer terms between compartments).
+Next, we expanded the description of growth factor receptor signaling, considering both homo- and heterodimer formation between ErbB family members and the role of cellular and secreted prostatic acid phosphatase (cPAcP and sPAcP, respectively).
+Both forms of PAcP were included because cPAcP downregulates HER2 activity, while sPAcP promotes modest HER2 activation \cite{Veeramani2005}. 
+Third, we expanded the description of the G1/S transition of the cell cycle (restriction point). 
+The previous model used the abundance of cyclin D as a proliferation marker, but did not include other proteins or interactions potentially important to the restriction point.
+Toward this shortcoming, we included cyclin E expression (and its role as a coregulator of androgen receptor expression), enhanced the description of cyclin D expression and the alternative splicing of cyclin D mRNA (including the role of the splice variants in androgen action), included the Rb/E2F pathway as well as E2F inhibition of androgen receptor expression \cite{Davis2006}, and the cyclin-dependent kinases cyclin-dependent kinase 4 (CDK4) and cyclin-dependent kinase 6 (CDK6).
+We also included key inhibitors of the restriction point including cyclin-dependent kinase inhibitor 1 (p21Cip1), cyclin-dependent kinase inhibitor 1B (p27Kip1), and cyclin-dependent kinase inhibitor 2A (p16INK4) \cite{Sherr1999}.
+Fourth, we enhanced the description of growth factor induced translation initiation.
+One of the key findings of the previous model was that growth factor induced translation initiation was globally sensitive (important in both androgen dependent and independent conditions). 
+However, the description of this important subsystem was simplified in the previous model. 
+Here, we expanded this subsystem, using connectivity similar to previous study of Lequieu \textit{et al.} \cite{Lequieu:2011eu}, and re-examined the importance of key components of this axis, such as mammalian target of rapamycin (mTOR), phosphatidylinositide 3-kinase (PI3K) and AKT.
+Lastly, we significantly expanded the description of the role of androgen receptor. 
+The previous model assumed constant AR expression, consistent with studies in androgen dependent and independent LNCaP sublines \cite{Lee2003}.
+However, other prostate cancer cell lines vary in their AR expression \cite{Sobel2005}. 
+Thus, to capture androgen signaling in a variety of prostate cancer cells, we included the transcriptional regulation governing androgen receptor expression, updated our description of the regulation of androgen receptor activity and androgen action (gene expression program driven by activated androgen receptor).
+At the expression level, we included AR auto-regulation in combination with the co-activators cyclin E and CDK6 \cite{Yamamoto2000,Lim2005}. 
+We also assumed androgen receptor could be activated through androgen binding or a ligand-independent, MAPK-driven mechanism referred to as the outlaw pathway \cite{Feldman2001,Yeh1999}. 
+We assumed a single canonical activating AR phosphorylation site (S515), with phosphorylated extracellular-signal-regulated kinase 1/2 (ppERK1/2) being the major kinase and protein phosphatase 2 (PP2A) or phosphoprotein phosphatase 1 (PP1) being the major phosphatases responsible for regulating this site.
+Finally, we modeled androgen receptor induced gene expression, including prostate specific antigen (PSA), cPAcP and p21Cip1.
+
+%This description includes  were added to the model because of their importance in co-regulating AR transcription regulation activity in prostate cells . 
+%With the addition of cell cycle proteins, we also included 
+%only considered cyclin D as a cell marker and 
+%In both models we considered nuclear hormone (AR) and transmembrane growth factor receptor (homo- and heterodimers of ErbB1/2) activation through both ligand-dependent and independent mechanisms. 
+%The model investigated the activation of AR through androgen binding or a ligand-independent, MAPK-driven mechanism. 
+%In addition to MAPK-dependent outlaw activation of AR, the model also included transcriptional activation via MAPK signaling \cite{Feldman2001,Yeh1999}. 
+%The PI3K/AKT/mTOR pathway for translation initiation was also included in the model, as well as the regulated expression of PSA, cyclin D,  
+%Thus, 
+%The model of Tasseff \textit{et al.}  
+%In order 
+%Additionally, expression of transcription factors AP-1 and ETS were also added to the model. 
+%Another significant addition to the model was the inclusion of a nuclear compartment separate from the cytosol, membrane and extracellular environments. 
+%The prostate model was developed to determine the shift in important signaling components between CR and AD prostate cancer cells.
+
 \subsection*{Formulation and solution of the model equations.}
-We used ordinary differential equations (ODEs) to model the time evolution of proteins ($x_{i}$) in our reduced order coagulation model:
+The prostate model was formulated as a coupled set of non-linear ordinary differential equations (ODEs):
+\begin{equation}\label{asseassertain_massbalance}
+\frac{d\mathbf{x}}{dt}=\mathbf{S}\cdot\mathbf{r}\left(\mathbf{x},\mathbf{k}\right)\qquad\mathbf{x}\left(t_{o}\right)=\mathbf{x}_{o}
+\end{equation}
+The quantity $\mathbf{x}$ denotes the vector describing the abundance of protein, mRNA, and other species in the model ($780\times 1$).
+The stoichiometric matrix $\mathbf{S}$ encodes the signaling architecture considered in the model ($780\times 1674$).  
+Each row of $\mathbf{S}$ describes a signaling component while each column describes a particular interaction.
+The $(i,j)$ element of $\mathbf{S}$, denoted by $\sigma_{ij}$, describes how species $i$ is involved with interaction $j$. 
+If $\sigma_{ij}>0$, species $i$ is produced by interaction $j$. Conversely, If $\sigma_{ij}<0$, then species $i$ is consumed in interaction $j$. 
+Lastly, if $\sigma_{ij}=0$, then species $i$ is not involved in interaction $j$.
+The term  $\mathbf{r}\left(\mathbf{x},\mathbf{k}\right)$ denotes the vector of interactions rates ($1674\times{1}$).
+Gene expression and translation processes as well as all biochemical transformations were decomposed into simple elementary steps, 
+where all reversible interactions were split into two irreversible steps (supplemental materials). 
+We modeled each network interaction using elementary rate laws where all reversible interactions were split into two irreversible steps. 
+Thus, the rate expression for interaction $q$ was given by:
+\begin{equation}\label{eq-mass-action}
+r_{q}\left(\mathbf{x},k_{q}\right)=k_{q}\prod_{j\in\left\{\mathbf{R}_{q}\right\}}x_{j}^{-\sigma_{jq}}
+\end{equation}
+The set $\left\{\mathbf{R}_{q}\right\}$ denotes reactants for reaction $q$, while $\sigma_{jq}$ denotes the stoichiometric coefficient (element of the matrix $\mathbf{S}$) governing species $j$ in reaction $q$.
+The quantity $k_{q}$ denotes the rate constant (unknown) governing reaction $q$.
+Model equations were generated in the C-programming language using the UNIVERSAL code generator, starting from an text-based input file (available in supplemental materials). 
+UNIVERSAL, an open source Objective-C/Java code generator, is freely available as a Google Code project (http://code.google.com/p/universal-code-generator/). 
+Model equations were solved using the CVODE solver in the SUNDIALS library \cite{Hindmarsh2005} on an Apple workstation (Apple, Cupertino, CA; OS X v10.6.8).
+
+We ran the model to steady-state before calculating the response to DHT or growth factor inputs.
+The steady-state was estimated numerically by repeatedly solving the model equations and estimating the difference between subsequent time points:
 \begin{equation}
-	\frac{dx_{i}}{dt}  =  \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)\qquad{i=1,2,\hdots,\mathcal{M}}\\
+\Vert\mathbf{x}\left(t+\Delta{t}\right) - \mathbf{x}\left(t\right)\Vert_{2}\leq\gamma
 \end{equation}
-where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of protein species in the model. 
-The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of reaction $j$. 
-Typically, reaction $j$ is a non-linear function of biochemical species abundance, as well as unknown kinetic parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$). 
-The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$. 
-If $\sigma_{ij}>0$, species $i$ is produced by reaction $j$. 
-Conversely, if $\sigma_{ij}<0$, species $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates species $i$ is not connected with reaction $j$. 
-The system material balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$, which were specified by the experimental setup.
+The quantities $\mathbf{x}\left(t\right)$ and $\mathbf{x}\left(t+\Delta{t}\right)$ denote the simulated abundance vector at time $t$ and $t + \Delta{t}$, respectively.
+The $L_{2}$ vector-norm was used as the distance metric, where $\Delta{t} = 100$ hr of simulated time and $\gamma$ = 0.001 for all simulations.
 
-Each reaction rate was written as the product of two terms, a kinetic term ($\bar{r}_{j}$) and a regulatory term ($v_{j}$):
-\begin{equation}\label{eqn:rate-factor}
-	r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right) = \bar{r}_{j}v_{j}
+We estimated an ensemble of model parameter sets using the Pareto Optimal Ensemble Techniques (POETs) multiobjective optimization routine \cite{Song:2009dq,Song2010,Lequieu:2011eu}. 
+POETs minimized the residual between model simulations and 43 separate training objectives taken from protein and mRNA signaling data generated in androgen dependent, intermediate and independent LNCaP cell lines (Table \ref{objective_table}).
+From these training objectives, POETs generated $>10^6$ candidate parameter vectors from which we selected N = 5000 Pareto rank-zero vectors for further analysis. 
+The set-to-set correlation between selected sets was approximately 0.60, suggesting only modest similarity between ensemble members.
+Approximately 33\%, or 560 of the 1674 parameters had a coefficient of variation (CV) of less than 1.0, where the CV ranged from 0.59 to 5.8 over the ensemble.
+Details of the parameter estimation problem and POETs are given in the supplemental materials.
+
+\subsection*{Sensitivity and robustness analysis.}
+Steady-state sensitivity coefficients were calculated for N = 500 parameter sets selected from the ensemble by solving the augmented kinetic-sensitivity equations \cite{Dickinson1976}:
+\begin{equation}
+	\begin{bmatrix}
+		\mathbf{S}\cdot\mathbf{r}(\mathbf{x},\mathbf{k}) \\
+		\mathbf{A}(t_s)\mathbf{s}_j + \mathbf{b}_j(t_s)
+	\end{bmatrix} = \begin{pmatrix}
+		\mathbf{0} \\
+		\mathbf{0}
+	\end{pmatrix}
+	   \qquad j = 1,2,\dots,\mathcal{P}
 \end{equation}
-We used multiple saturation kinetics to model the reaction term $\bar{r}_{j}$:
-\begin{equation}\label{eqn:rate-bar}
-	\bar{r}_{j} = k_{j}^{max}\epsilon_{i}\left(\prod_{s\in{m_{j}^{-}}}\frac{x_{s}}{K_{js} + x_{s}}\right)
-\end{equation}where $k_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the scaled enzyme activity which catalyzes reaction $j$, and
-$K_{js}$ denotes the saturation constant for species $s$ in reaction $j$. 
-The product in Eqn. \eqref{eqn:rate-bar} was carried out over the set of \textit{reactants} for reaction $j$ (denoted as $m_{j}^{-}$).
+where
+\begin{equation}
+s_{ij}(t_s)= \left. \frac{\partial{x_i}}{\partial{k_j}} \right|_{t_s}
+\end{equation} for each parameter set.
+Steady-state was calculated as described previously.
+The quantity $j$ denotes the parameter index, $\mathbf{A}$ denotes the Jacobian matrix, and $\mathcal{P}$ denotes the number of parameters in the model. 
+The vector $\mathbf{b}_j$ denotes the $j$th column of the matrix of first-derivatives of the mass balances with respect to the parameters. 
+Steady-state sensitivity coefficients were used because of the computational burden associated with sampling several hundred parameters sets for each of the 1674 parameters. 
+The steady-state sensitivity coefficients $\mathcal{N}_{ij} \equiv s_{ij}$ were organized into an array for each parameter set in the ensemble: 
+\begin{equation}
+	\mathcal{N}^{\left(\epsilon\right)} = 
+	\begin{pmatrix}
+		\mathcal{N}_{11}^{\left(\epsilon\right)} & \mathcal{N}_{12}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{1j}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{1P}^{\left(\epsilon\right)} \\
+		\mathcal{N}_{21}^{\left(\epsilon\right)} & \mathcal{N}_{22}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{2j}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{2P}^{\left(\epsilon\right)} \\
+		\vdots & \vdots & & \vdots & & \vdots \\
+		\mathcal{N}_{M1}^{\left(\epsilon\right)} & \mathcal{N}_{M2}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{Mj}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{MP}^{\left(\epsilon\right)} \\
+	\end{pmatrix}\qquad\epsilon = 1,2,\hdots,N_{\epsilon}
+\end{equation}
+where $\epsilon$ denotes the index of the ensemble member, $P$ denotes the number of parameters, $N_{\epsilon}$ denotes the number of parameter sets sampled (N = 500) and $M$ denotes the number of model species.
+To estimate the relative fragility or robustness of species and reactions in the network, we decomposed $\mathcal{N}^{\left(\epsilon\right)}$ 
+using Singular Value Decomposition (SVD):
 
-The control term $v_{j}$ depended upon the combination of factors which influenced the activity of enzyme $i$.
-For each enzyme, we used a rule-based approach to select from competing control factors (Fig. \ref{fig-control-schematic}). 
-If an enzyme was activated by $m$ metabolites, we modeled this activation as:
 \begin{equation}
-	v_{j} = \max\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{mj}\left(\mathcal{Z}\right)\right)
-\end{equation}where $0\leq f_{ij}\left(\mathcal{Z}\right)\leq 1$ was a regulatory transfer function that calculated the influence of metabolite $i$ on the activity of enzyme $j$. 
-Conversely, if enzyme activity was inhibited by $m$ metabolites, we modeled this inhibition as:
+	\mathcal{N}^{\left(\epsilon\right)} = \mathbf{U}^{\left(\epsilon\right)}\Sigma^{\left(\epsilon\right)}\mathbf{V}^{T,{\left(\epsilon\right)}}
+\end{equation}
+Coefficients of the left singular vectors corresponding to largest $\theta$ $\leq$ 15 singular values of $\mathcal{N}^{\left(\epsilon\right)}$ were rank-ordered to estimate important species combinations, while coefficients of the right singular vectors were used to rank important reaction combinations.
+Only coefficients with magnitude greater than a threshold ($\delta$ = 0.001) were considered. 
+The fraction of the $\theta$ vectors in which a reaction or species index occurred was used to quantify its importance (sensitivity ranking). 
+We compared the sensitivity ranking between different conditions to understand how control in the network shifted in different cellular environments.
+
+Robustness coefficients were calculated as shown previously \cite{Tasseff2011}. 
+Robustness coefficients denoted by $\alpha\left(i,j,t_{o},t_{f}\right)$ are defined as:
 \begin{equation}
-	v_{j} = 1 - \max\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{mj}\left(\mathcal{Z}\right)\right)
-\end{equation}Lastly, if an enzyme had both $m$ activating and $n$ inhibitory factors, we modeled the regulatory term as:
-\begin{equation}
-	v_{j} = \min\left(u_{j},d_{j}\right)
-\end{equation}where:
-\begin{eqnarray}
-	u_{j} &=& \max_{j^{+}}\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{mj}\left(\mathcal{Z}\right)\right) \\
-	d_{j} &=& 1 - \max_{j^{-}}\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{nj}\left(\mathcal{Z}\right)\right)
-\end{eqnarray}
-The quantities $j^{+}$ and $j^{-}$ denoted the sets of activating and inhibitory factors for enzyme $j$. 
-If a process has no modifying factors, we set $v_{j} = 1$.
-There are many possible functional forms for $0\leq f_{ij}\left(\mathcal{Z}\right)\leq 1$. 
-However, in this study, each individual transfer function took the form:
-\begin{equation}\label{eqn:control-factor}
-	f_{i}\left(\mathbf{x}\right) = \frac{\kappa_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}{1 + \kappa_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}
-\end{equation}where $\mathcal{Z}_{j}$ denotes the abundance of the $j$ factor (e.g., metabolite abundance), and $\kappa_{ij}$ and $\eta$ are control parameters. 
-$\kappa_{ij}$ was the species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).
-Applying the general framework to the reduced coagulation network resulted in five ordinary differential equations:
-\begin{eqnarray}
-	\frac{dx_{1}}{dt} & =& -\left(r_{init}v_{init} + r_{amp}v_{amp}\right)\\
-	\frac{dx_{2}}{dt} &=& r_{amp}v_{amp} + r_{init}v_{init} - r_{inh,ATIII}v_{inh,ATIII}\\
-	\frac{dx_{3}}{dt} &=& -r_{apc}v_{apc} \\
-	\frac{dx_{4}}{dt} &=& r_{apc}v_{apc} \\
-	\frac{dx_{5}}{dt} & = & -r_{inh,ATIII}v_{inh,ATIII}
-\end{eqnarray}where $\mathbf{x} = \left(fII,FIIa,PC,APC,ATIII\right)^{T}$. 
-The terms $r_{*}v_{*}$ in the balance equations denote corrected kinetic expressions for initiation, amplification and inhibition processes. 
-The rate of initiation $\bar{r}_{init}$ was modeled as:
-\begin{equation}
-	\bar{r}_{init} = k_{init}\left(trigger\right)\frac{x_{1}}{K_{init,fII} + x_{1}}
-\end{equation}where $k_{init},~K_{init,fII}$ are the rate and saturation constants governing initiation, respectively. 
-The rate of initiation was modified by $v_{init}$, the control parameter governing initiation. 
-Initiation was sensitive to the level of trigger (activator) and TFPI (inhibitor):
-\begin{equation}
-	v_{init} = \min\left(f^{-}_{init}\left(TFPI\right),f^{+}_{init}\left(trigger\right)\right)
-\end{equation}where the transfer functions $f$ took the form of Eqn \eqref{eqn:control-factor}.
-The rate of thrombin amplification was given by:
-\begin{equation}
-	\bar{r}_{amp} = k_{amp}\left(x_{2}\right)\frac{x_{1}}{K_{amp,fII}+x_{1}}
-\end{equation}where $k_{amp},~K_{amp,fII}$ denote the rate and saturation constants governing amplification, respectively. 
-The amplification control term, which modified amplification rate, was modeled as a combination of multiple inhibition terms and one activation term:
-\begin{equation}
-	v_{amp} = \min\left(f^{-}_{amp}\left(TFPI\right),f^{-}_{amp}\left(x_{4}\right),f^{+}_{amp}\left(\mathcal{Z}_{amp}\right)\right)
-\end{equation}where $\mathcal{Z}_{amp} = fV\times fX\times fVIII\times fIX$. Although $f^{+}_{amp}\left(\mathcal{Z}_{amp}\right)$ is an activating term, 
-we included it in the min integration rule; the factors in $\mathcal{Z}_{amp}$ were essential for amplification (if any of these factors was missing
-the amplification reaction would not occur). Thus, the factors in $\mathcal{Z}_{amp}$ were required components, a classification
-that we implemented by the min selection rule.
-The rate activated protein C formation was given by:
-\begin{equation}
-	\bar{r}_{apc} = k_{APC,formation}\left(TM\right)\frac{x_{3}}{K_{formation,PC} + x_{3}}
-\end{equation}where $k_{APC,formation}$ and $K_{formation,PC}$ denote the rate and saturation constants governing activated protein C formation, respectively and $TM$ denotes
-the thrombomodulin abundance. We modeled the control term which governed APC formation as a single thrombin-dependent activation term:
-\begin{equation}
-	v_{apc} = \max\left(f_{apc}^{+}\left(x_{2}\right)\right)
+\alpha\left(i,j,t_{o},t_{f}\right)=\left(\displaystyle\int_{t_{o}}^{t_{f}}x_{i}\left(t\right)dt\right)^{-1}\left(\displaystyle\int_{t_{o}}^{t_{f}}x^{(j)}_{i}\left(t\right)dt\right)
 \end{equation}
-Lastly, we included direct irreversible inhibition of FIIa by ATIII:
-\begin{equation}
-	\bar{r}_{inh,ATIII} = k_{ATIII,inhibition}\left(x_{5}x_{2}^{\gamma}\right)
-\end{equation}where $\gamma$ was estimated to be $\gamma = 1.26$. For ATIII inhibition of FIIa, the control variables $v_{inh,ATIII}$ was taken to be unity.
-The model equations were encoded using the Python programming language and solved using the ODEINT routine of the SciPy module \citep{SCIPY}.
-The model files can be downloaded from http://www.varnerlab.org.
-
-\subsection*{Estimation of model parameters from experimental data.}
-Model parameters were estimated by minimizing the difference between simulations and experimental thrombin measurements (squared residual):
-\begin{equation}\label{eqn:objective-function}
-	\min_{\mathbf{k}} \sum_{\tau=1}^{\mathcal{T}}\sum_{j=1}^{\mathcal{S}}\left(\frac{\hat{x}_{j}\left(\tau\right) - x_{j}\left(\tau,\mathbf{k}\right)}{\omega_{j}\left(\tau\right)}\right)^{2}
-\end{equation}where $\hat{x}_{j}\left(\tau\right)$ denotes the measured value of species $j$ at time $\tau$, $x_{j}\left(\tau,\mathbf{k}\right)$ denotes the simulated 
-value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$. The outer summation is with respect to
-time, while the inner summation is with respect to state. We minimized the model residual using Particle swarm optimization (PSO) \citep{PSO}.
-PSO uses a \textit{swarming} metaheuristic to explore parameter spaces. 
-A strength of PSO is its ability to find the global minimum, even in the presence of potentially many local minima, by communicating the local
-error landscape experienced by each particle collectively to the swarm. Thus, PSO acts both as a local and a global search algorithm. 
-For each iteration, particles in the swarm compute their local error by evaluating the model equations using their specific parameter vector realization.
-From each of these local points, a globally best error is identified. Both the local and global error 
-are then used to update the parameter estimates of each particle using the rules:
-\begin{eqnarray}
-	\mathbf{\Delta}_{i} &=&\theta_{1}\mathbf{\Delta}_{i} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{k}_{i}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G} - \mathbf{k}_{i}\right) \\
-	\mathbf{k}_{i} &=& \mathbf{k}_{i} + \mathbf{\Delta}_{i}
-\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes the local best solution found by particle $i$, and
-$\mathcal{G}$ denotes the best solution found over the entire population of particles.
-The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
-parameters ($\mathcal{K}\times{1}$). In thus study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$. The quality of parameter
-estimates was measured using goodness of fit (model residual). The particle swarm optimization routine was implemented in the Python programming language. 
-All plots were made using the Matplotlib module of Python \citep{Matplotlib}.
-%The optimization and plotting code can be downloaded from http://www.varnerlab.org.
-
-\subsection*{Global sensitivity analysis of model performance.}
-We conducted a global sensitivity analysis, using the variance-based method of Sobol, to estimate which parameters controlled the performance of the reduced order model \citep{SOBOL_METHOD}. 
-We computed the total sensitivity index of each parameter relative to two performance objectives, the peak thrombin time and the area under the thrombin curve (thrombin exposure).
-We established the sampling bounds for each parameter from the minimum and maximum value of that parameter in the parameter set ensemble. 
-We used the sampling method of Saltelli \textit{et al.} \citep{Saltelli:2010} to compute a family of $N\left(2d+2\right)$ parameter sets which obeyed our parameter ranges, 
-where $N$ was the number of trials, and $d$ was the number of parameters in the model. In our case, $N$ = 10,000 and $d$ = 22, so the total sensitivity indices were computed from
-460,000 model evaluations. The variance-based sensitivity analysis was conducted using the SALib module encoded in the Python programming language \citep{SALIB}.
+Robustness coefficients quantify the response of a marker to a structural or operational perturbation to the network architecture.
+Here $t_o$ and $t_f$ denote the initial and final simulation time respectively, while $i$ and $j$ denote the indices for the marker and the perturbation respectively. 
+A value of $\alpha\left(i,j,t_{o},t_{f}\right)>1$, indicates increased marker abundance, while $\alpha\left(i,j,t_{o},t_{f}\right)<1$ indicates decreased marker abundance following perturbation $j$. 
+If $\alpha\left(i,j,t_{o},t_{f}\right)\sim{1}$ the $j$th perturbation does not influence the abundance of marker $i$.
+Robustness coefficients were calculated (starting from steady-state) from $t_{o} = 0$ hr to $t_{f} = 72$ hr following the addition of 10nM DHT at $t_{o}$.
+Robustness coefficients were calculated for the same N = 500 models selected for sensitivity analysis.
 
 \section*{Acknowledgements}
-This study was supported by an award from the Army Research Office (ARO \#59155-LS).
+This study was supported by award number U54CA143876 from the National Cancer Institute.
+The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Cancer Institute or the National Institutes of Health.
+This study was also supported by the National Science Foundation GK12 award number DGE-1045513.
 
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-1-Network.pdf}
-\caption{Schematic of the connectivity of the reduced order coagulation model. A trigger compound, e.g., TF/FVIIa initiates thrombin production (FIIa) from prothrombin (fII).
-Once activated, thrombin catalyzes its own activation (amplification step), as well as its own inhibition via the conversion of protein C to activated protein C (APC).
-APC and tissue factor pathway inhibitor (TFPI) inhibit initiation and amplification, while antithrombin III (ATIII) directly inhibits thrombin. 
-All inhibition steps and trigger-induced initiation were modeled using a rule-based approach. 
-Likewise, the dependence of amplification on other coagulation factors was also modeled using a rule-based approach. 
-The abundance of the highlighted species (in the dashed boxes) was governed by an ordinary differential equation. 
-All other species were assumed to be constant.}\label{fig-network}
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Fig_1_LNCAPModel.pdf}
+\caption{Schematic overview of the prostate signaling network.
+The model describes hormone and growth factor induced expression of several proteins, including PSA.
+In the absence of outside hormones/growth factors, overactive HER2 can stimulate the MAPK and AKT pathways.
+AR can be activated directly by the MAPK pathway.
+}
+\label{fg:NetworkSchematic}
 \end{figure}
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-2-ControlSchematic.pdf}
-\caption{Schematic of rule-based effective control laws. 
-Traditional enzyme kinetic expressions, e.g., Michaelis–Menten or multiple saturation kinetics are multiplied by an enzyme activity control variable $0 \leq v_{j} \leq 1 $. 
-Control variables are functions of many possible regulatory factors encoded by arbitrary functions of the form $0\leq f_{j}\left(\mathcal{Z}\right)\leq 1$.
-At each simulation time step, the $v_{j}$ variables are calculated by evaluating integration rules such as the max or min of the set of factors $f_{1},\hdots,f_{n}$ 
-influencing the activity of enzyme $E_{j}$. }\label{fig-control-schematic}
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Fig_3_ColorPlot.pdf}
+\caption{Simulation results versus experimental results for training and validation data.
+Experiment numbers 1 through 43 were used for training, while experiments 44 through 72 were validation.
+Gray means the ensemble member qualitatively fit experimental data in both models.
+White means the the ensemble member only fit the data using the new model that included HER2 heterodimerization.
+Red means the ensemble member fit using only the old model.
+Black corresponds to an incorrect cellular response in both models.
+$\bf{A.,C.}$ Training and validation results, respectively, for entire ensemble population using both the original model and an updated model including HER2 heterodimerization (N = 5000).
+$\bf{B.,D.}$ Simulation results for training and validation of a random set of 100 members using both models.
+ }
+\label{fg:ColorPlot}
 \end{figure}
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-3-DIFF_FII_NO_APC.pdf}
-\caption{Reduced order coagulation model training simulations.
-Reduced order coagulation model parameters were estimated using particle swarm optimization (PSO) with and without the protein C pathway as a function of prothrombin. 
-Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data. 
-The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation).
-(A,B,C) training results for 150\%, 100\% and 50\% of physiological prothrombin levels in the absence of the protein C pathway.
-Thrombin generation was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
-All factors and control proteins were at their physiological concentration unless others denoted. 
-The experimental training data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}.}\label{fig-diff_fII_NO_APC}
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Fig_2_Training.pdf}
+\caption{Ensemble performance against selected training objectives (N = 5000).
+A, B.
+Time course data for PSA concentration due to a stimulus of 10 nM DHT in LNCaP C33 cells and LNCaP C81 cells, respectively (O2, O3).
+C.
+PSA levels in the presence and absence of a HER2 inhibitor (LNCaP C81 cells, O7).
+D.
+PAcP mRNA levels at 72 hours in the presence and absence of DHT (LNCaP C51 cells, O14).
+E.
+Steady-state cyclin D levels in LNCaP C33 vs.
+C81 (O17).
+F.
+p21Cip1 levels at 48 hrs in the presence and absence of DHT (LNCaP C33, O25).
+G.
+Steady-state AKT phosphorylation levels in LNCaP C33 vs.
+C51 (O30).
+H.
+AR levels at 24 hours in the presence and absence of DHT (LNCaP C51, O31).
+I.
+AR mRNA levels in the presence and absence of E2F over expression (LNCaP C33, O34).}
+\label{fg:Training}
 \end{figure}
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-4-DIFF_FII_APC.pdf}
-\caption{Reduced order coagulation model training simulations.
-Reduced order coagulation model parameters were estimated using particle swarm optimization (PSO) with and without the protein C pathway as a function of prothrombin. 
-Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data.
-The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation).
-(A,B,C) training results for 150\%, 100\% and 50\% of physiological prothrombin levels in the presence of the protein C pathway. 
-Only APC pathway parameters were allowed to vary in the simulations on the right.
-Thrombin generation was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
-All factors and control proteins were at their physiological concentration unless others denoted.
-The experimental training data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}.}\label{fig-diff_fII_APC}
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Fig_4_Validation.pdf}
+\caption{Blind model predictions for the ensemble (N = 5000).
+The model ensemble’s predictive ability was assessed by comparing simulation versus experimental data not used for training.
+A.
+Time course data for HER phosphorylation due to a stimulus of 10 nM DHT (LNCaP C33, P1).
+B.
+ERK phosphorylation levels in the presence and absence of a PAcP inhibitor (LNCaP C33 cells, P3).
+C.
+AR levels at 24 hrs in varying levels of DHT (LNCaP C33, P17).
+D.
+Shc phosphorylation levels at 24 hrs in the presence and absence of DHT (LNCaP C33, P22).
+E.
+Time course data for HER phosphorylation due to a stimulus of 1.6 nM EGF (LNCaP C33, P7).
+F.
+AR levels in varying levels of EGF (LNCaP C33, P14).
+G, H, I.
+Fold change in PSA concentration due to drug stimulus: enzalutamide, lapatinib, and sorafenib.}
+\label{fg:Validation}
 \end{figure}
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-5-FII_ATIII.pdf}
-\caption{Reduced order coagulation model predictions versus experimental data for normal coagulation. 
-The reduced order coagulation model parameter estimates were tested against data not used during model training.
-Simulations of different levels of prothrombin and ATIII were compared with experimental data in the absence of the protein C pathway. 
-Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data.
-The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation). 
-(A,B,C,D) prediction results for (FII,ATIII): (50\%,150\%), (100\%, 100\%), (125\%, 75\%) and (150\%, 50\%) of physiological prothrombin and ATIII levels in the absence of the protein C pathway. 
-Thrombin generation was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
-All factors and control proteins were at their physiological concentration unless others denoted.
-The experimental validation data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}}\label{fig-fII_ATIII}
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Fig_5_Sensitivity.pdf}
+\caption{Sensitivity analysis of a population of prostate models (N = 500).
+Species with a low sensitivity are considered robust, while species with a high sensitivity ranking are considered fragile.
+$\bf{A,B.}$ Sensitivity ranking of network species in AD versus CR cells in the absence (presence) of DHT.
+$\bf{C.}$ Sensitivity ranking of network species in AD cells in the absence and presence of DHT.
+$\bf{D.}$ Sensitivity ranking of network species in CR cells in the presence and absence of enzalutamide with a DHT stimulus.}
+\label{fg:Sensitivity}
 \end{figure}
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-6-Hemophilia.pdf}
-\caption{Reduced order coagulation model predictions versus experimental data with and without FVIII and FIX.
-The reduced order coagulation model parameter estimates were tested against data not used during model training. 
-Simulations of normal thrombin formation with ATIII and the protein C pathway were compared with thrombin formation in the absence of fVIII and fIX.
-Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data.
-The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation). 
-(A,B) prediction results for normal thrombin generation and thrombin generation in hemophilia. 
-All factors and control proteins were at their physiological concentration unless others noted.
-The experimental validation data was reproduced from the study of Allen \textit{et al.} \citep{ALLEN2006}.}\label{fig-hemophilia}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-7-rFVIIa-Trigger.pdf}
-\caption{Reduced order coagulation model predictions of rFVIIa administration. 
-A: Simulations of thrombin formation in the presence of ATIII and the protein C pathway were conducted for a range of trigger values (1x - 200x nominal) in the absence of fVIII and fIX.
-B: Comparison of thrombin generation for normal versus hemophilia for 10x nominal trigger.
-Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles.
-The shaded region denotes the 99\% confidence estimate of the mean thrombin value (uncertainty in the model simulation). 
-All factors and control proteins were at their physiological concentration unless others noted.}\label{fig-rFVII}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-8-Flux-rFVIIa-Trigger.pdf}
-\caption{Reaction flux distribution as a function of time for thrombin generation under normal (left), hemophilia (center) and rFVIIa treated hemophilia (right).
-Reaction flux was calculated for each particle at $T = 0,4,6,8,10,12,14$ min after the initiation of coagulation. 
-Reaction fluxes were calculated for each particle in the parameter ensemble (N = 20).
-Blue colors denote low flux values while red colors denote high flux values.}\label{fig-flux-analysis}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-9-Sensitivity.pdf}
-\caption{Global sensitivity analysis of the reduced order coagulation model with respect to the model parameters.
-A: Sensitivity analysis of the thrombin peak time for different prothrombin levels (150\%,100\% and 50\% of the physiological value) as a function of activated protein C. 
-B: Sensitivity analysis of the thrombin exposure for different prothrombin levels (150\%,100\% and 50\% of the physiological value) as a function of activated protein C. 
-Points denote the mean total sensitivity value, while the area around each point denotes the uncertainty in the sensitivity value. 
-The gray dashed line denotes the 45$^{\circ}$ degree diagonal, if sensitivity values are equal for different conditions they will lie on the diagonal.  
-Sensitivity values significantly above or below the diagonal indicate differentially important model parameters. 
-The radius of the shaded region around each total sensitivity value was the maximum uncertainty in that value estimated by the Sobol method. 
-}\label{fig-sensitivity}
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure6_RAF_K0_C81.pdf}
+\caption{Robustness analysis of a population of CR prostate models with Raf knock-out (N = 500).
+A log fold change of greater then zero implies that the concentration of the protein increased with the knock-out of Raf, while a log fold change of less then zero indicates that the concentration of protein decreased.
+A log of fold change equal to 0, shows no response due to Raf knock-out.
+Three distinct regions emerge in Raf knock-out case: (1) PSA increases, (2) cyclin D concentration increases, and (3) PSA and cyclin D concentration decrease.
+}
+\label{fg:Robustness}
 \end{figure}
@@
 
+\section*{Supplementary materials}
+
+\subsection*{Estimation of a population of models using Pareto Optimal Ensemble Techniques (POETs).}
+
+We used multiobjective optimization to estimate an ensemble of prostate models.
+Although computationally more complex than single-objective formulations, multiobjective optimization can be used to address qualitative conflicts in training data arising from experimental error or cell-line artifacts \cite{Handl2007}.
+In this study we used the Pareto Optimal Ensemble Technique (POETs) to perform the optimization.
+POETs integrates standard search strategies, e.g., Simulated Annealing (SA) or Local Pattern Search (PS) with a Pareto-rank fitness assignment \cite{Song2010}.
+The mean squared error, $\eta$, of parameter set $\mathbf{k}$ for training objective $j$ was defined as:
+
+\begin{equation}
+\eta_j(\mathbf{p}_k) = \frac{1}{N}\sum_{i}^{N}\frac{(\hat{x}_{i,j} - \beta_j x(\mathbf{p_k})_{i,j})^2}{\hat{\sigma}^2_{i,j}}
+\end{equation}
+The symbol $\hat{x}_{i,j}$ denotes scaled experimental observations (from training objective j) while $x(\mathbf{p_k})_{i,j}$ denotes the simulation output (from training objective j).
+The quantity $i$ denotes the sampled time-index or condition, and $N$ denotes the number of time points or conditions for experiment j.
+The standard deviation, $\hat{\sigma}_{i,j}$, was assumed to be equal to 10\% of the reported observation, if no experimental error was reported.
+$\beta_j$ is a scaling factor which is required when considering experimental data that is accurate only to a multiplicative constant.
+In this study, the experimental data used for training and validation was typically band intensity from immunoblots, where intensity was estimated using the ImageJ software package \cite{Abramoff2004}.
+The scaling factor used was chosen to minimize the normalized squared error \cite{Brown2003}:
+
+\begin{equation}
+\beta_j = \frac{\sum_{i}(\hat{x}_{i,j}x_{i,j}/\hat{\sigma}_{i,j}^2)}{\sum_{i}(x_{i,j}/\hat{\sigma}_{i,j})^2}
+\end{equation} 
+By using the scaling factor, the concentration units on simulation results were arbitrary, which was consistent with the arbitrary units on the experimental training data.
+All simulation data was scaled by the corresponding $\beta_j$.
+
+We computed the Pareto rank of parameter set $\mathbf{k}_{i+1}$ by comparing the simulation error at iteration $i+1$ against the simulation archive, denoted as $\mathbf{K}_i$.
+We used the Fonseca and Fleming ranking scheme \cite{Fonseca1993} to estimate the rank of the parameter set $\mathbf{k}_{i+1}$.
+Parameter sets with increasing rank are progressively further away from the optimal trade-off surface.
+The parameter set $\mathbf{k}_{i+1}$ was accepted or rejected by the SA with probability $\mathcal{P}\left(\mathbf{k}_{i+1}\right)$:
+\begin{equation}\label{eqn_costMOSA}
+\mathcal{P}(\mathbf{k}_{i+1}) \equiv \exp{\{-rank\left(\mathbf{k}_{i+1} \mid \mathbf{K}_{i} \right)/T\}}
+\end{equation}
+where $T$ is the computational annealing temperature.
+The Pareto rank for $\mathbf{k}_{i+1}$ is denoted by $rank\left(\mathbf{k}_{i+1}\mid \mathbf{K}_{i}\right)$.
+The annealing temperature was  adjusted according to the schedule
+$T_k = \beta^k T_0$
+where $\beta$ was defined as $\beta = \left(\frac{T_{f}}{T_{o}}\right)^{1/10}$.
+The initial temperature was given by $T_0 = n/log(2)$, with $n=4$ and the final temperature $T_f = 0.1$ used in this study.
+The epoch-counter $k$ was incremented after the addition of 50 members to the ensemble.
+As the ensemble grew, the likelihood of accepting a high rank set decreased.
+Parameter sets were generated by applying a random perturbation in log space:
+
+\begin{equation}
+\log\mathbf{k}_{i+1} = \log\mathbf{k}_i + \mathcal{N}(0,\nu)
+\end{equation}
+where $\mathcal{N}(0,\nu)$ is a normally distributed random number with zero mean and variance $\nu$, set as 0.1 in this model.
+The perturbation was applied in log space to account for large variation in parameter scales and to ensure positive parameter values.
+We used a local pattern search every $q$ steps, in our case 20, to minimize error for a single randomly selected objective.
+The local pattern-search algorithm used has been described previously \cite{Gadkar2003}.
+
+\input{Prostate_Obj_Table}
+\clearpage
+
+\input{Prostate_Pred_Table}
+\clearpage
+
+\begin{figure}\centering
+%\includegraphics[width = 7in]{Supp_Figure_Sensitivity.pdf}
+\caption{Sensitivity analysis of a population of prostate models (N = 500).
+Species with a low sensitivity are considered robust, while species with a high sensitivity ranking are considered fragile.
+$\bf{A}$ Sensitivity ranking of network species in CR cells in the absence and presence of DHT.
+$\bf{B.}$ Sensitivity ranking of network species in CR cells in the presence and absence of enzalutamide in the absence of a DHT stimulus.
+$\bf{C.,D.}$ Sensitivity ranking of network species in CR cells in the presence and absence of sorafenib and lapatinib, respectively, with a DHT stimulus.}
+\label{fg:Supp_Sensitivity}
+\end{figure}
+
+\begin{figure}\centering
+%\includegraphics[width = 7in]{Supp_Figure_Robustness.pdf}
+\caption{Robustness analysis of protein markers.
+Expression level of key proteins was altered by a factor of 2, 0.1, or 0 (knock-in, knock-down, or knock-out) and robustness coefficients were calculated for five key protein markers.
+Simulations shown were from CR cells, with indicated perturbation.
+Mean of 500 ensemble members is shown.
+}
+\label{fg:Supp_Robustness}
+\end{figure}
+
+%\end{comment}
+
 \end{document}

### 2014-10-30 Build.sh

@@ -1,10 +1,10 @@
 #!/bin/sh
 
 # Tex this mofo -
-latex $1.tex
+pdflatex $1.tex
 bibtex $1
-latex $1.tex
-latex $1.tex
+pdflatex $1.tex
+pdflatex $1.tex
 
 # Make the pdf -
-dvipdfm -p "letter" $1.dvi
+#dvipdfm -p "letter" $1.dvi

### 2014-11-10 Paper_v1.tex
Submitted version

--- 2014-10-21 Paper_v1.tex
@@
 \begin{titlepage}
-{\par\centering\textbf{\Large Dynamic Modeling of the Human Coagulation Cascade using Reduced Order Effective Kinetic Models}}
+{\par\centering\textbf{\Large Dynamic Modeling of Cell-Free Biochemical Networks using Effective Kinetic Models}}
 \vspace{0.05in}
-{\par \centering \large{Adithya Sagar and Jeffrey D. Varner$^{*}$}}
+{\par \centering \large{Joseph A. Wayman, Adithya Sagar and Jeffrey D. Varner$^{*}$}}
 \vspace{0.10in}
@@
 \vspace{0.1in}
-{\par \centering \textbf{Running Title:}~Modeling blood coagulation}
+{\par \centering \textbf{Running Title:}~Effective models of metabolism}
 \vspace{0.1in}
-{\par \centering \textbf{To be submitted:}~\emph{Processes}}
+{\par \centering \textbf{To be submitted:}~\textit{Processes}}
 \vspace{0.5in}
@@
 \section*{Abstract}
-In this study, we present a novel hybrid modeling approach which combines ordinary differential equation (ODE) modeling with logical rules to simulate an archetype biochemical network, 
-the human coagulation cascade.  
-The model consisted of five differential equations augmented with several logical rules describing 
-regulatory connections between model components, and unmodeled interactions in the network. 
-This formulation was more than an order of magnitude smaller than current coagulation models, 
-because many of the mechanistic details of coagulation were encoded as logical rules.
-We estimated an ensemble of likely model parameters (N = 20) from \textit{in~vitro} extrinsic coagulation data sets, with and without inhibitors, 
-by minimizing the residual between model simulations and experimental measurements using particle swarm optimization (PSO). 
-We then validated the model ensemble using thrombin data sets that were not used during training. 
-The ensemble predicted thrombin trajectories for conditions not used for model training, including thrombin generation for normal and hemophilic coagulation 
-in the presence of platelets (a significant unmodeled component).
-We then used flux analysis to understand how the network operated in a variety of conditions, and global sensitivity analysis to identify which parameters controlled the 
-performance of the network. Taken together, the hybrid approach produced a surprisingly predictive model given its small size, 
-suggesting the proposed hybrid framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
-or potentially even cell free metabolic systems.
+Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
+Many of the challenges confronting genome-scale kinetic modeling can potentially be overcome in a cell-free system. 
+For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth. 
+Thus, cell-free operation holds several significant advantages for model development, identification and validation. 
+Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms, such as \textit{E.
+coli}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.
+Toward this unmet need, we present an effective biochemical network modeling framework for building dynamic cell-free metabolic models.
+The key innovation of our approach is the integration of simple effective rules encoding complex allosteric regulation with traditional kinetic pathway modeling. 
+We tested our approach by modeling the time evolution of several hypothetical cell-free metabolic networks. 
+We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, captured complex allosteric patterns such as ultrasensitivity or non-competitive inhibition in the absence of mechanistic information.
+Second, when integrated into network models, these rules captured classic regulatory patterns such as product-induced feedback inhibition. 
+Lastly, we showed, at least for the network architectures considered here, that we could simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data. 
+While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.
 
-\vspace{0.25in}
-{\noindent \textbf{Keywords:}~Blood coagulation, Mathematical modeling, Systems biology}
+{\noindent \textbf{Keywords:}~Cell-free metabolism, Mathematical modeling}
 
@@
 
-%We chose coagulation because it is well studied, has a complex architecture, 
-%and has an abundance of experimental data available for model identification and validation. 
-%However, coagulation was just a proof-of concept to validate our approach; 
-%
+\section*{Introduction}
+Mathematical modeling has long contributed to our understanding of metabolism. 
+Decades before the genomics revolution, mechanistically, structured metabolic models arose from the desire to predict microbial phenotypes resulting from changes in intracellular or extracellular states \citep{1976_fredrickson_BiotechBioeng}.
+The single cell \textit{E.
+coli} models of Shuler and coworkers pioneered the construction of large-scale, dynamic metabolic models that incorporated multiple, regulated catabolic and anabolic pathways constrained by experimentally determined kinetic parameters \citep{1984_domach_shuler_BiotechBioeng_01}.
+Shuler and coworkers generated many single cell kinetic models, including single cell models of eukaryotes \citep{1989_steinmeyer_shuler_ChemEngSci,1992_wu_shuler_AnnNYAcadSci}, minimal cell architectures \citep{2004_castellanos_shuler_PNAS}, as well as DNA sequence based whole-cell models of \textit{E.
+coli} \citep{2008_atlas_shuler_IETSysBio}.
+Conversely, highly abstracted kinetic frameworks, such as the cybernetic framework, represented a paradigm shift, viewing cells as growth-optimizing strategists \citep{1985_dhurjati_ramkrishna_tsao_BiotechBioeng}.
+Cybernetic models have been highly successful at predicting metabolic choice behavior, e.g., diauxie behavior \citep{1986_kompala_ramkrishna_tsao_BiotechBioeng}, steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, as well as the cellular response to metabolic engineering modifications \citep{1999_varner_ramkrishna_MetaEng}.
+Unfortunately, cybernetic models also suffer from an identifiability challenge, as both the kinetic parameters and an abstracted model of cellular objectives must be estimated simultaneously.
 
-%Developing mathematical modeling tools to interrogate the molecular basis of human disease is one of the grand challenges facing systems biology.
-%Developing computational tools to understand the molecular basis of human disease is a grand challenge facing systems biology.
-%Many have suggested that the integration of experimental and computational research is required to unravel
-%critical questions facing molecular medicine \cite{KITANO2002,ASSMUS2006,ARNAUD2006}. 
-%Despite this, mechanistic network modeling has not played a significant role in the development of new therapies for cancer, cardiovascular diseases or for acute events  
-%like thrombosis during surgery.
-%A critical limiting issue often cited has been model uncertainty \cite{BAILEY2001}. 
-%Development and validation of detailed network models requires a significant investment of time and material resources. 
-%Alternatively, development of tools using toy models often fails to demonstrate relevance to biologists or physicians.  
-%What is needed is a model system, that is simultaneously tractable and directly applicable to the study of disease.
-%Human blood is perhaps such an ideal model system \cite{Diamond:2009yf}. 
-%Blood is a complex mixture of proteins and cells which collectively carry out many functions.
-%However, one well studied facet of blood, which could serve as a valuable model system, is the coagulation cascade. 
-%Understanding the mechanism of rFVIIa might provide insight into the therapeutic dose under different physiological conditions.
-%More generally, rFVIIa treatment could also serve an ideal test case to develop strategies which retrospectively estimate operational perturbations associated with drugs.
-%For example, the effect of rFVIIa is directly observable in coagulation outputs such as thrombin concentration.
-%Thus, by analyzing these outputs in a normal versus a perturbed network, it might be possible to reverse engineer the molecular influence associated with a therapeutic intervention.
+In the post genomics world, large-scale stoichiometric reconstructions of microbial metabolism popularized by static, constraint-based modeling techniques such as flux balance analysis (FBA) have become standard tools \citep{2012_lewis_palsson_NatRevMicrobio}.
+Since the first genome-scale stoichiometric model of \textit{E.
+coli}, developed by Edwards and Palsson \citep{2000_edwards_palsson_PNAS}, well over 100 organisms, including industrially important prokaryotes such as \textit{E.
+coli} \citep{Feist:2007aa} or \textit{B.
+subtilis} \citep{Oh:2007aa}, are now available \citep{2009_feist_palsson_NatRevMicrobio}.
+Stoichiometric models rely on a pseudo-steady-state assumption to reduce unidentifiable genome-scale kinetic models to an underdetermined linear algebraic system, which can be solved efficiently even for large systems.
+Traditionally, stoichiometric models have also neglected explicit descriptions of metabolic regulation and control mechanisms, instead opting to describe the choice of pathways by prescribing an objective function on metabolism.
+Interestingly, similar to early cybernetic models, the most common metabolic objective function has been the optimization of biomass formation \citep{2002_ibarra_edwards_palsson_Nat}, although other metabolic objectives have also been estimated \citep{2007_schuetz_sauer_MolSysBio}.
+Recent advances in constraint-based modeling have overcome the early shortcomings of the platform, including capturing metabolic regulation and control \citep{2013_hyduke_lewis_palsson_MolBioSys}. 
+Thus, modern constraint-based approaches have proven extremely useful in the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \citep{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}.
+However, genome-scale kinetic models of industrial important organisms such as \textit{E.
+coli} have yet to be constructed.
 
-%
-%Thus, while there are many possible strategies to simulate biochemical networks, ODE models remain a common choice despite
-%significant identification challenges.
-%This result highlights the perhaps under-appreciated role that experimental design could play in generating the best training and validation data sets for model identification \citep{Apgar:2010fk}.
+Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
+Central amongst these advantages is direct access to metabolites and the microbial biosynthetic machinery without the interference of a cell wall. 
+This allows us to control as well as interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
+Second, cell-free systems also allow us to study biological processes without the complications associated with cell growth.
+Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \citep{Jewett:2008aa}. 
+However, CFPS is not new; CFPS in crude \textit{E. coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \citep{MATTHAEI:1961aa,NIRENBERG:1961aa}. 
+Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \citep{Lu:2014aa} to synthetic biology \citep{Hodgman:2012aa}. 
+Interestingly, many of the challenges confronting genome-scale kinetic modeling can potentially be overcome in a cell-free system. 
+For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth. 
+Thus, cell-free operation holds several significant advantages for model development, identification and validation.
+Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms, such as \textit{E.
+coli} or \textit{B.
+subtilis}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.
 
-\section*{Introduction}
-Developing mathematical models of biochemical networks is a significant facet of systems biology.
-Modeling approaches differ in their degree of detail, where the choice of approach is often determined by prior knowledge, or model requirements \citep{2012_kholodenko_kolch_SciSig}. 
-Ordinary differential equation (ODE) models are a common tool for modeling biochemical systems because of their ability to capture dynamics, and encode mechanism.
-However, ODE models typically come with difficult (or sometimes impossible) parameter identification problems. 
-For example, Gadkar et al., showed that even with near-perfect information, 
-it was often impossible to identify all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
-However, it is not clear whether we actually need precise estimates for all model parameters. 
-Bailey suggested more than a decade ago, that achieving qualitative or even quantitative understanding of biological systems 
-should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}. 
-Since Bailey's complex biology with no parameters hypothesis, Sethna showed that model performance is typically 
-sensitive to only a few parameters, a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
-Thus, reasonable predictions may be possible, despite parametric uncertainty, if a few critical parameters are well defined. 
-Perhaps more importantly, ODE models also require significant mechanistic information, thereby limiting their utility in poorly understood systems.
-Toward this challenge, logical modeling is an emerging paradigm that encodes causal relationships using non-linear transfer functions \citep{Morris:2010aa}.
-These functions encode interactions between model components, without requiring mechanism.
-Despite their simplicity, logical models have captured rich system behaviors in the absence of mechanistic information \citep{Saez-Rodriguez:2011aa,Morris:2011ys,Morris:2012aa}. 
-However, modeling complex dynamics remains difficult with this class of model.
-Thus, there is an unmet need for a third approach which combines ODEs and logical models, where ODEs could encode mechanistic information, 
-while missing or incomplete mechanistic knowledge can be approximated using a logical approach.
-
-In this study, we developed a hybrid approach which combines ODE modeling with logical rules to model a well studied biochemical network, 
-the human coagulation system. Coagulation is an archetype proteolytic cascade involving both positive 
-and negative feedback \cite{Butenas:2002aa,Schenone:2004aa,Adams:2009aa}. 
-Coagulation is mediated by a family proteases in the circulation, called factors and a key group of blood cells, called platelets. 
-The central process in coagulation is the conversion of prothrombin (fII), an inactive coagulation factor, to the master protease thrombin (FIIa).
-Thrombin generation involves three phases, initiation, amplification and termination \cite{GOLDHABER2006, Brummel:2002aa}.
-Initiation requires a trigger event, for example vessel injury, which leads to the activation of factor VII (FVIIa).
-Two converging pathways, the extrinsic and intrinsic cascades, then process and amplify this initial coagulation signal. 
-The extrinsic cascade is generally believed to be the main mechanism of thrombinogenesis in the blood \cite{MANN1990,ROBERTS1998,MANN1999}.
-Initially, thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa), which itself has been activated by FVIIa \cite{Butenas:2002aa}. 
-Picomolar amounts of thrombin then activate the cofactors factors V and VIII (fV and fVIII) and platelets, 
-leading to the formation of the tenase and prothrombinase complexes on activated platelets.
-These complexes amplify the early coagulation signal by further activating FXa, and directly converting prothrombin to thrombin. 
-There are several control points in the cascade that inhibit thrombin formation, and eventually terminate thrombin generation. 
-Tissue Factor Pathway Inhibitor (TFPI) inhibits FXa formation catalyzed by TF/FVIIa, while antithrombin III (ATIII)
-neutralizes several of the proteases generated during coagulation, including thrombin.  
-Thrombin itself also inadvertently plays a role in its own inhibition; thrombin, through interaction with thrombomodulin, protein C and endothelial cell protein C receptor (EPCR),
-converts protein C to activated protein C (APC) which attenuates the coagulation response by proteolytic cleavage of fV/FVa and fVIII/FVIIIa. 
-Termination occurs after either prothrombin is consumed, or thrombin formation is neutralized by inhibitors such as activated protein C (APC) or antithrombin III (ATIII).
-
-Coagulation models in the literature are typically formulated as systems of nonlinear ordinary differential equations, 
-using mass-action or other kinetics, to describe the rates of biochemical conversions \citep{Kuharsky:2001aa}.
-Previous ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys}, 
-were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, 
-and later Hockin \emph{et al.} \citep{Hockin:2002aa}, who developed and then refined highly mechanistic models of extrinsic coagulation. 
-Other laboratories have also expanded upon Hockin \emph{et al.}, for example including the intrinsic pathway, and thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
-Other biochemical or biophysical aspects of coagulation have also been modeled, such as platelet biochemistry \citep{Stalker:2013aa}, 
-multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and transport inside growing clots \citep{Voronov:2013aa}.
-However, most of these previous studies were based upon detailed mechanistic knowledge. 
-Such a description is possible because blood, while enormously complex, can be systematically interrogated. 
-However, other systems, such as the intracellular networks which mediate cancer, are more difficult to experimentally probe and model.  
-Towards this unmet need, we have formulated a hybrid modeling approach which combines ODEs and logical rules to model 
-biochemical processes for which mechanistic information is difficult or impossible to obtain. 
-We have used this approach to model the coagulation cascade because of the wealth of training and validation data available for this system. 
-We proposed a reduced order coagulation model that consisted of five differential equations augmented with several logical rules.
-We identified the reduced order coagulation model parameters from in vitro extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
-We then validated the model using thrombin data sets from multiple laboratories, for both normal and hemophilic coagulation.
-Flux and sensitivity analysis identified which parameter or parameters combinations were critical to model performance in several conditions. 
-The reduced order hybrid approach produced a surprisingly predictive coagulation model, suggesting this framework could potentially be used to model other biochemical networks important to human health.
+In this study, we present an effective biochemical network modeling framework for building dynamic cell-free metabolic models.
+The key innovation of our approach is the seamless integration of simple effective rules encoding complex regulation with traditional kinetic pathway modeling. 
+This integration allows the description of complex regulatory interactions, such as time-dependent allosteric regulation of enzyme activity, in the absence of specific mechanistic information. 
+The regulatory rules are easy to understand, easy to formulate and do not rely on overarching theoretical abstractions or restrictive assumptions. 
+We tested our approach by modeling the time evolution of several hypothetical cell-free metabolic networks. 
+In particular, we tested whether our effective modeling approach could describe classically expected enzyme kinetic behavior, and second whether we could simultaneously estimate kinetic parameters and regulatory connectivity, in the absence of specific mechanistic knowledge, from synthetic experimental data.
+Toward these questions, we explored five hypothetical cell-free networks. 
+Each network shared the same enzymatic connectivity, but had different allosteric regulatory connectivity. 
+We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, captured complex allosteric patterns such as ultrasensitivity or non-competitive inhibition in the absence of mechanistic information.
+Second, when integrated into network models, these rules captured classical regulatory patterns such as product-induced feedback inhibition. 
+Lastly, we showed, at least for the network architectures considered here, that we could simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data. 
+While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.
 
@@
 
-%However, components could also correspond to non-modeled influences, categorial variables or other abstract quantities.
+\subsection*{Formulation and properties of effective cell-free metabolic models.}
+We developed two proof-of-concept metabolic networks to investigate the features of our effective biochemical network modeling approach (Fig. \ref{fig-networks}). 
+In both examples, substrate S was converted to the end products P$_{1}$ and P$_{2}$ through a series of enzymatically catalyzed reactions, including a branch point at hypothetical metabolite M$_{2}$. 
+Several of these reactions involved cofactor dependence (AH or A), and various allosteric regulatory mechanisms modified the activity of pathway enzymes. 
+Network A included feedback inhibition of the initial pathway enzyme ($E_{1}$) by pathway end products P$_{1}$ and P$_{2}$ (Fig. \ref{fig-networks}A). 
+On the other hand, network B involved feedback inhibition of $E_{1}$ by P$_{2}$ and $E_{6}$ by P$_{1}$ (Fig. \ref{fig-networks}B). 
+In both networks, branch point enzymes $E_{3}$ and $E_{6}$ were subject to feed-forward activation by reduced cofactor AH. 
+Lastly, it is known experimentally that cell-free systems have a finite operational lifespan. 
+Loss of biosynthetic capability could be a function of many factors, e.g., cofactor or metabolite limitations. 
+We modeled the loss of biosynthetic capability as a non-specific first-order decay of enzyme activity.
 
-\subsection*{Formulation of reduced order coagulation models.}
-We developed a reduced order coagulation model to test our hybrid kinetic rule based modeling approach (Fig. \ref{fig-network}).
-A trigger event initiates thrombin formation (FIIa) from prothrombin (fII) through a lumped initiation step.
-This step loosely represents the initial fluid phase activation of thrombin by activated FXa.
-Once activated, thrombin catalyzes its own formation (amplification step), and inhibition via the conversion of protein C to activated protein C (APC).
-APC and tissue factor pathway inhibitor (TFPI) inhibit initiation, while antithrombin III (ATIII) inhibits amplification. 
-All initiation and inhibition processes, as well as the dependence of amplification upon other coagulation factors, was approximated using our rule-based approach (Fig. \ref{fig-control-schematic}). 
-Individual regulatory contributions to the activity of pathway enzymes were integrated into control coefficients using an integration rule.  
-These control coefficients then modified the rates of model processes at each time step.
-Hill-like transfer functions $0 \leq f\left(\mathcal{Z}\right) \leq 1$ quantified the influence of components upon a target process. 
-Components were either individual inhibitor or activator levels or some function of levels, e.g., the product of factor levels. 
-In this study, $\mathcal{Z}$ corresponded to the abundance of individual inhibitors or activators, 
-with the exception of the dependence of amplification upon specific coagulation factors (modeled as the product of factors). 
-When a process was potentially sensitive to multiple inputs, logical integration rules were used to select which transfer functions influenced the process at any given time. 
-In our proof of concept model, we used a winner takes all strategy; the maximum or minimum transfer function was selected at any given time step. 
-However, other integration rules are certainly possible. 
-Taken together, while the reduced order coagulation model encodes significant biological complexity, it is highly compact, consisting of only five differential equations. 
-Thus, it will serve as an excellent proof of principle example to study the reduction of a highly complex human subsystem.
+Allosteric regulation of enzyme activity was modeled by combining individual regulatory contributions to the activity of pathway enzymes into a control coefficient using an integration rule (Fig.
+\ref{fig-control-schematic}).
+This strategy is similar in spirit to the Constrained Fuzzy Logic (cFL) approach of Lauffenburger and coworkers which has been used to effectively model signal transduction pathways important in human health \citep{Morris:2011aa}.
+In our formulation, Hill-like transfer functions $0 \leq f\left(\mathcal{Z}\right) \leq 1$ were used to calculate the influence of factor abundance upon target enzyme activity. 
+In this context, factors can be individual metabolite levels or some function, e.g., the product of metabolite levels. 
+However, more generally, factors can also correspond to non-modeled influences, categorial variables or other abstract quantities. 
+In the current study, we simply let $\mathcal{Z}$ correspond to the abundance of individual metabolites, however in general this can be a complex function of both modeled and unmodeled factors. 
+When an enzyme was potentially sensitive to more than one regulatory input, logical integration rules were used to select which regulatory transfer function influenced enzyme activity at any given time.
+Thus, our test networks involved important features such as cofactor recycling, enzyme activity and metabolite dynamics, as well as multiple overlapping allosteric regulatory mechanisms.
 
-\subsection*{Identification of model parameters using particle swarm optimization.}
+The rule-based regulatory strategy approximated the behavior of classical allosteric activation and inhibition mechanisms (Fig. \ref{fig-kinetics-simulations}). 
+We considered the enzyme catalyzed conversion of substrate S to a product P, where the overall reaction rate was modeled as the product of a Michaelis-Menten term and an effective allosteric control variable reflecting the particular regulatory interaction.
+We first explored feed-forward substrate activation of enzyme activity (for both positive and negative cooperativity). 
+Consistent with classical data, the rule-based strategy predicted a sigmoidal relationship between substrate abundance and reaction rate as a function of the cooperativity parameter (Fig.
+\ref{fig-kinetics-simulations}A).
+For cooperativity parameters less than unity, increased substrate abundance \textit{decreased} the reaction rate. 
+This was consistent with the idea that substrate binding \textit{decreased} at regulatory sites, which negatively impacted substrate binding at the active site. 
+On the other hand, as the cooperativity parameter increased past unity, the rate of conversion of substrate S to product P by enzyme $E$ approached a step function. 
+In the presence of an inhibitor, the rule-based strategy predicted non-competitive like behavior as a function of the cooperativity parameter (Fig. \ref{fig-kinetics-simulations}B). 
+When the control gain parameter, $\kappa_{ij}$ in Eqn.
+\eqref{eqn:control-factor}, was greater than unity, the inhibitory force was directly proportional to the cooperativity parameter, $\eta$ in Eqn.
+\eqref{eqn:control-factor}.
+Thus, as the cooperativity parameter increased, the maximum reaction rate decreased (Fig. \ref{fig-kinetics-simulations}B). 
+Interestingly, our rule-based approach was unable to directly simulate competitive inhibition of enzyme activity. 
+Taken together, the rule-based strategy captured classical regulatory patterns for both enzyme activation and inhibition. 
+Thus, we are able to model complex kinetic phenomena such as ultrasensitivity, despite an effective description of reaction kinetics.
+
+End product yield was controlled by feedback inhibition, while product selectivity was controlled by branch point enzyme inhibition (Fig. \ref{fig-onoff-simulations}). 
+A critical test of our modeling approach was to simulate networks with known behavior. 
+If we cannot reproduce the expected behavior of simple networks, then our effective modeling strategy, and particularly the rule-based approximation of allosteric regulation, will not be feasible for genome-scale cell-free problems.
+We considered two cases, control ON/OFF, for each network configuration.
+Each of these cases had identical kinetic parameters and initial conditions; the \textit{only} differences between the cases were the allosteric regulation rules and the control parameters associated with these rules.
+As expected, end product accumulation was larger for network A when the control was OFF (no feedback inhibition of $E_{1}$ by P$_{1}$ and P$_{2}$),
+as compared to the ON case (Fig. \ref{fig-onoff-simulations}A). 
+We found this behavior was robust to the choice of underlying kinetic parameters, as we observed that same qualitative response across an ensemble of randomized parameter sets (N = 100), for fixed control parameters.
+The control ON/OFF response of network B was more subtle. 
+In the OFF case, the behavior was qualitatively similar to network A. 
+However, for the ON case, flux was diverted away from P$_{2}$ formation by feedback inhibition of $E_{6}$ activity at the M$_{2}$ branch point by P$_{1}$ (Fig. \ref{fig-onoff-simulations}B). 
+Lower $E_{6}$ activity at the M$_{2}$ branch point allowed more flux toward P$_{1}$ formation, hence the yield of P$_{1}$ also increased (Fig. \ref{fig-onoff-simulations}C). 
+Again, the control ON/OFF behavior of network B was robust to changes in kinetic parameters, as the same qualitative trend was conserved across an ensemble of randomized parameters (N = 100), for fixed control parameters.
+Taken together, these simulations suggested that the rule-based allosteric control concept could robustly capture expected feedback behavior for networks with uncertain kinetic parameters.
+
+\subsection*{Estimating parameters and effective allosteric regulatory structures.}
 A critical challenge for any dynamic model is the estimation of kinetic parameters. 
-We estimated kinetic and control parameters simultaneously from eight \emph{in vitro} time-series coagulation data sets with and without the protein C pathway. 
-The residual between model simulations and experimental measurements was minimized using particle swarm optimization (PSO). 
-A population of particles (N = 20) was initialized with randomized kinetic and control parameters and allowed to search for parameter vectors that minimized
-the residual. However, not all parameters were varied simultaneously. 
-We partitioned the parameter estimation problem into two subproblems based upon the biological organization of the training data; 
-(i) estimation of parameters associated with thrombin formation in the absence of the protein C pathway and (ii) estimation of parameters associated with the protein C pathway.
-Only those parameters associated with each subproblem were varied during the optimization procedure for that subproblem e.g., thrombin parameters were \textit{not} varied during the protein C subproblem.The PSO procedure was run for 20 generations for each subproblem, where each generation was 1200 iterations.
-The best particle from each generation was used to generate the particle population for the next generation.
-We rotated the subproblems, starting with subproblem 1 in the first generation.
+For metabolic processes, there is also the added challenge of identifying the regulation and control structures that manage metabolism. 
+Of course, these issues are not independent; any description of enzyme activity regulation will be a function of system state, which in turn depends upon the kinetic parameters. 
+For cell-free systems, regulated gene expression has been removed, however, enzyme activity regulation is still operational. 
+We explored this linkage by estimating model parameters from synthetic data using both network structures. 
+We generated noise corrupted synthetic measurements of the substrate S, intermediate M$_{5}$ and end product P$_1$ approximately every 20 min using
+network A.
+We then generated an ensemble of model parameter estimates by minimizing the difference between model simulations and the synthetic data using particle swarm optimization (PSO), starting from random initial parameter guesses.
+The estimation of kinetic parameters was sensitive to the choice of regulatory structure (Fig.
+\ref{fig-parameter-fit}).
+PSO identified an ensemble of parameters that bracketed the mean of the synthetic measurements in less than 1000 iterations when the control structure was correct (Fig.
+\ref{fig-parameter-fit}A and B).
+However, with control mismatch (network B simulated with network A parameters), model simulations were not consistent with the synthetic data  (Fig. \ref{fig-parameter-fit}C and D). 
+Taken together, these results suggested that we could perhaps simultaneously estimate both parameters and network control architectures, as incorrect control structures would be manifest as poor model fits.
 
-The reduced order coagulation model captured the role of initial prothrombin abundance, and the decay of the thrombin signal following from ATIII activity (Fig. \ref{fig-diff_fII_NO_APC}).
-However, we systematically under-predicted the thrombin peak and the strength of ATIII inhibition in this training data set.
-On the other hand, with fixed thrombin parameters, we captured peak thrombin values and the decay of the thrombin signal (at least for the 150\% fII case) in the presence
-of both ATIII and the protein C pathway (Fig. \ref{fig-diff_fII_APC}). Lastly, we were unable to capture global differences in initiation time \textit{across} separate data 
-sets with a single ensemble of model parameters. These differences likely result from normal experimental variability.
-For example different thrombin generation experiments within our 
-training data (at the same physiological factor levels) have significantly different initiation times (supplemental results). 
-However, this also highlights a potential shortcoming of the initiation module within the model. 
-To capture the variability in initiation time \textit{across} training data sets, we included a constant time-delay parameter ($T_{D}$) for each data group.
-The delay parameter was constant within a data set, but allowed to vary \textit{across} training data sets. Introduction of the delay parameter allowed the model
-to simulate multiple training data sets using a single ensemble of model parameters.
-Taken together, the model identification results suggested 
-that our kinetic-rules based approach could reproduce a panel of thrombin generation data sets conducted at physiological factor and inhibitors concentrations.
-However, it was unclear whether the reduced order model could predict new data, without updating the model parameters.
+We modified our particle swarm identification strategy to simultaneously search over both kinetic parameters and putative control structures. 
+In addition to our initial networks, we constructed three additional presumptive network models, each with the same enzymatic connectivity but different allosteric regulation of the pathway enzymes (Fig.
+\ref{fig-alternative-networks}).
+We then initialized a population of particles, each with one of the five potential regulatory programs and randomized kinetic parameters. 
+Thus, we generated an initial population of particles that had \textit{both} different kinetic parameters as well as different control structures. 
+We biased the distribution of the particle population according to our \textit{a prior} belief of the correct regulatory program. 
+To this end, we considered three different priors, a uniform distribution where each putative regulatory structure represented 20\% of the population and two mixed distributions that were either positively or negatively biased towards the correct structure (network A).
+In both the positively biased and uniform cases the PSO clearly differentiated between the true or closely related structures and those that were materially different (Fig. \ref{fig-control-search}). 
+As expected, the positively biased population (40\% of the initial particle population seeded with network A) gave the best results, where the correct structure was preferentially identified (Fig.
+\ref{fig-control-search}A).
+On the other hand, when given a uniform distribution, the PSO approach identified a combination of network A and network C as the most likely control structures (Fig. \ref{fig-control-search}B). 
+Network A and C differ by the regulatory connection between the end product P$_2$ and enzyme $E_{1}$; in network A, end product P$_{2}$ was assumed to inhibit $E_{1}$, while in network C, end product P$_{2}$ activated $E_{1}$.
+Lastly, when the initial population was biased towards incorrect structures (initial population seeded with 90\% incorrect structures), the particle swarm \textit{misidentified} the correct allosteric structure (Fig.
+\ref{fig-control-search}C).
+Interestingly, while each particle swarm identified parameter sets that minimized the simulation error, the estimated parameter values were not necessarily similar to the true parameters. 
+The angle between the estimated and true parameters was not consistently small across the swarms (identical parameters would give an angle of zero). 
+This suggested that our particle swarm approach identified a \textit{sloppy} ensemble, i.e., parameter estimates that were individually incorrect but collectively exhibited the correct model behavior.
 
-\subsection*{Validation of the reduced order coagulation model.}
-We tested the predictive power of the reduced order coagulation model with validation data sets not used during model training. 
-Two validation data sets were used, thrombin generation for various prothrombin and ATIII concentrations with the protein C pathway, 
-and thrombin generation in normal versus hemophilic plasma in the presence of the protein C pathway. 
-Lastly, we compared the qualitative output of the model to rFVIIa addition in the presence of hemophilia.
-The hemophilia case was an especially difficult test as it was taken from a different study which used a plasma-based
-\textit{in~vitro} assay involving platelets instead of phospholipid vesicles (PCPS). 
-All kinetic and control parameters were fixed for the validation simulations. 
-The only globally adjustable parameter $T_{D}$, was fixed within each validation data set but allowed to vary between data sets.
-The reduced order model predicted the thrombin generation profile for ratios of prothrombin and ATIII in the absence of the protein C pathway (Fig. \ref{fig-fII_ATIII}).
-Simulations near the physiological range (fII,ATIII) = (100\%, 100\%) or (125\%,75\%) tracked the measured thrombin values (Fig. \ref{fig-fII_ATIII}B and C).
-On the other hand, predictions for factors levels outside of the physiological range (fII,ATIII) = (50\%, 150\%) or (150\%, 50\%), while qualitatively consistent with measured thrombin values, did show significant
-deviation from the measurements (Fig. \ref{fig-fII_ATIII}A and D).
-Likewise, simulations of thrombin generation in normal versus hemophilia (missing both fVIII and fIX) were consistent with measured thrombin values (Fig. \ref{fig-hemophilia}).
-We modeled the dependence of thrombin amplification on factor levels using a product rule ($\mathcal{Z} = fV\times fX\times fVIII\times fIX$), 
-which was then was integrated using a $\min$ integration rule into the control variable governing amplification. 
-Thus, in the absence of fVIII or fIX, the amplification control variable evaluated to zero, and the only thrombin produced was from initiation (Fig. \ref{fig-hemophilia}B).
-However, the decay of the thrombin signal was underpredicted in the normal case (Fig. \ref{fig-hemophilia}A), while the activated thrombin level was overpredicted 
-in hemophilia simulations, although thrombin generation was far less than normal (Fig. \ref{fig-hemophilia}B). 
-Taken together, the reduced order model performed well in the physiological range of factors, even with
-unmodeled components such as platelet activation in the hemophilia data set. Thus, our kinetic-rules framework predicted 
-the output of a physiologically important cascades such as coagulation, despite significant unmodeled components.
-
-The model ensemble predicted a direct correlation between thrombin generation and rFVIIa addition in hemophilia (Fig. \ref{fig-rFVII}).
-In the current model, we cannot distinguish between different initiation sources, e.g., TF/FVIIa and rFVIIa, as we have only a single lumped initiation source (trigger).
-Thus, we simulated the addition of rFVIIa in hemophilia by removing fVIII and fIX from the model, and modulating the initial level of trigger.
-Simulations with a baseline level of trigger were consistent with the previous hemophilia simulations, where the only thrombin being produced was from initiation (Fig.
-\ref{fig-rFVII}, 1$\times$~trigger).
-As we increased the strength of the trigger event, the thrombin peak time and the maximum value of thrombin increased (Fig. \ref{fig-rFVII}, 50$\times$~trigger). 
-However, as the trigger strength increased, the thrombin generated quickly decayed.
-For example, for the highest trigger strength simulated (50$\times$trigger), 95\% of the thrombin was gone by 20 min after initiation. 
-We performed flux analysis to understand how the reduced order coagulation model balanced initiation, amplification and inhibition of thrombin formation for normal coagulation and hemophilia.
-Analysis of the reaction flux through the reduced order network for thrombin generation in normal, hemophilia and rFVIIa-treated hemophilia identified three distinct operational modes (Fig.
-\ref{fig-flux-analysis}).
-The reduced order network includes four lumped reactions, initiation, amplification, thrombin-induced APC generation and total thrombin inhibition (including both APC and ATIII action).
-Directly after the addition of a trigger (e.g., TF/FVIIa or rFVIIa), the lumped initiation flux was the largest for all three cases.
-However, within a few minutes enough thrombin was generated by the initiation mechanism to induce the amplification stage. 
-During amplification, thrombin catalyzes its own formation and inhibition by generating activated protein C (APC), a potent inhibitor of the coagulation cascade.
-For normal coagulation, amplification and thrombin inhibition are the dominate reactions by 6 min after initiation (Fig. \ref{fig-flux-analysis}, left). 
-After 10 min, the dominate reaction has shifted to thrombin inhibition (both ATIII and APC action). In the current proof-of-principle model, 
-APC inhibits upstream of amplification thus APC activity will slow the rate of thrombin formation directly from the initiation trigger. 
-On the other hand, ATIII inhibits thrombin directly, as well as upstream coagulation factors.
-In hemophilia (missing both fVIII and fIX), the amplification reaction does not occur and the only thrombin produced is from initiation (Fig. \ref{fig-flux-analysis}, center).
-Initiation is quickly inhibited by APC, and the thrombin level stabilizes and eventually decays because of ATIII activity. 
-Lastly, when 50$\times$rFVIIa is used to induce thrombin formation in hemophilia (absence of fVIII/fIX), 
-initiation mechanisms dominate for up to 6 min following initiation (Fig. \ref{fig-flux-analysis}, right). 
-However, similar to hemophilia alone, without amplification the thrombin signal is quickly extinguished by the combined action of ATIII and APC generated by thrombin.
+We calculated control program output and scaled metabolic flux for the positively, uniformly and negatively biased particle swarms (Fig. \ref{fig-flux-pattern}). 
+Network A and network C models from the positively (Fig.
+\ref{fig-flux-pattern}A) and uniformly (Fig.
+\ref{fig-flux-pattern}B) biased particle swarms showed similar operational patterns, despite differences in kinetic parameters and control structures.
+While models from the negatively biased population had error values similar to the correct structures in the previous swarms, they have different flux and control profiles (Fig.
+\ref{fig-flux-pattern}C).
+In all cases, regardless of network configuration or parameter values, the rate of enzyme decay was small compared to the other fluxes, and all networks had qualitatively similar trends for $E_{3}$ and $E_{6}$ control.
+Moreover, consistent with the correct model structure, production of end product P$_1$ was the preferred branch for all model configurations. 
+However, there was variability in P$_{2}$ production flux across the population of models, especially for the uniform swarm when compared with the other cases. 
+High P$_{1}$ branch flux resulted in end product inhibition of $E_{1}$ in both network A and network C, however in network D and E, high P$_1$ flux induced $E_1$ activation. 
+These trends were manifested in different flux profiles, where the negatively biased population appeared more uniform across the population compared with the other swarms, and had higher $E_{1}$ specific activity.
+Interestingly, the behavior of network A and network C highlighted an artifact of our integration rule; both a positive or negative feedback connection from P$_2$ to $E_{1}$ were ignored because the P$_{1}$ inhibition of $E_{1}$ was dominate.
+Thus, while theoretically distinct, network A and network C appeared operationally to the PSO algorithm to be that same network.
+On the other hand, networks B, D and E showed distinct behavior that was not consistent with the true network. 
+These architectures exhibited either limited inhibition (network B) or activation (network D and E) of $E_{1}$ activity, resulting in significantly different metabolic flux profiles. 
+However, the PSO was able to find low error parameter solutions, despite the mismatch in the control structures (error values similar, but not better than the best network A and network C estimates). 
+Taken together, these results suggested that a uniform sampling approach could potentially yield an unbiassed estimate of both kinetic parameters and control structures. 
+However, the negatively biased particle swarm results illustrated a potential shortcoming of the approach, namely convergence to a local error minimum despite a significantly incorrect control structure.
+This suggested that estimated model structures will need to be further evaluated, for example by generating falsifiable experimental designs which could distinguish between low error solutions.
 
@@
 
-\subsection*{Global sensitivity analysis of the reduced order coagulation model.}
-We conducted a global sensitivity analysis to estimate which parameters controlled the performance of the reduced order model. 
-We calculated the sensitivity of the time to maximum thrombin (peak time) and the thrombin exposure (area under the thrombin curve) 
-for different levels of prothrombin, and protein C (Fig. \ref{fig-sensitivity}). 
-Globally, 41\% of the parameters shifted in importance between the (fII,PC) = (50\%, 0\%) and (150\%,100\%) cases for the peak thrombin time (Fig. \ref{fig-sensitivity}A). 
-The majority of these shifts were involved increased prothrombin and the protein C pathway, while only
-5\% were directly associated with increased prothrombin alone. The rate constant for thrombin amplification was the most important parameter controlling the
-peak thrombin time. While this parameter was differentially important for different prothrombin levels, and in the presence or absence of 
-the activated protein C pathway, it's sensitivity was consistently above all other parameters in the model. The saturation constant governing thrombin amplification was
-the second most important parameter, followed by the initiation control gain parameter. Other important parameters influencing the thrombin peak time included the control gain for
-activated protein C formation, as well as the rate constant controlling ATIII inhibition of thrombin activity. 
-On the other hand, only 27\% of the model parameters were differentially sensitive between the (\%fII,\%PC) (50\%, 0\%) and (150\%,100\%) cases 
-for the thrombin exposure (Fig. \ref{fig-sensitivity}B). Of these parameters, all of the shifts were associated the interplay between thrombin formation and the protein C pathway. 
-The rate constant controlling ATIII inhibition was the most important parameter controlling the thrombin exposure.
-While this parameter was less important in the presence of protein C for 150\% prothrombin levels, it was significantly above all other parameters. 
-Similar to the peak time, for 150\% prothrombin, the control gain for activated protein C formation was differentially important along with the rate constant controlling
-amplification. However, the influence of the amplification parameter was much less in the thrombin exposure case versus peak time.
+\section*{Discussion}
+In this study, we presented an effective kinetic modeling strategy to dynamically simulate cell-free biochemical networks. 
+Our proposed strategy integrated traditional kinetic modeling with an effective rules based approach to dynamically describe metabolic regulation and control. 
+We tested this approach by developing kinetic models of hypothetical cell-free metabolic networks. 
+In particular, we tested whether our effective modeling approach could describe classically expected behavior, and second whether we could simultaneously estimate kinetic parameters and regulatory connectivity, in the absence of specific mechanistic knowledge, from synthetic experimental data.
+Toward these questions, we explored five hypothetical cell-free networks. 
+In each network, a substrate S was converted to the end products P$_{1}$ and P$_{2}$ through a series of enzymatically catalyzed reactions, including a branch point at a hypothetical metabolite M$_{2}$.
+Each network also included the same cofactors and cofactor recycle architecture. 
+However, while all five networks shared the same enzymatic connectivity, each had different allosteric regulatory connectivity. 
+We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, could capture complex allosteric patterns such as ultrasensitivity, or non-competitive inhibition in the absence of specific mechanistic information.
+Moreover, when integrated into network models, these rules captured classical regulatory patterns such as product-induced feedback inhibition. 
+Lastly, we simultaneously estimated kinetic parameters and discriminated between competing regulatory structures, using synthetic data in combination with a modified particle swarm approach.
+If we considered all putative regulatory architectures to be equally likely, we were able to estimate a \textit{sloppy} ensemble of models with the correct architecture and kinetic parameters.
 
-\clearpage
+The proposed modeling strategy shares features with other popular techniques, but has also has several key differences. 
+At its core, our effective modeling approach is similar to regulatory constraints based methods, and to the cybernetic modeling paradigm developed by Ramkrishna and colleagues.
+Covert, Palsson and coworkers drastically improved the predictability of constraints based approaches by integrating boolean rules into the calculation of metabolic fluxes \citep{Covert:2001aa}.
+If the regulated intracellular flux problem is coupled with time-dependent extracellular balances, these models can predict complex behavior such as diauxie growth or the switch between 
+aerobic and anaerobic metabolism. Another important feature of this approach is that it scales with biological complexity. 
+For example, Covert \emph{et al.,} showed that a genome scale model of \emph{E.~coli} augmented with a boolean rule layer, 
+correctly predicted approximately 80\% of the outcomes of a high-throughput growth phenotyping experiment in \emph{E. coli}.
+Further, they showed that they could learn new biology by iteratively refining the model and its associated rules \citep{Covert:2004aa}. 
+However, while regulated flux balance analysis is a powerful technique, it does not easily allow the calculation of time-resolved metabolite abundance. 
+Additionally, the boolean rules which populate the regulatory layer are limited to ON/OFF decisions; for qualitative predictions of gene expression this is a reasonable limitation. 
+However, boolean rules will likely be less effective at capturing dynamic allosteric regulation in a cell free metabolic system. 
+On the other hand, the strength of cybernetic models is the integration of optimal metabolic control heuristics with traditional kinetic pathway modeling. 
+Cybernetic models are highly predictive; they have successfully predicted mutant behavior from limited wild-type data \citep{1999_varner_ramkrishna_MetaEng,Varner:2000aa,Song:2012aa}, 
+steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, strain specific metabolic function \citep{Song:2011aa} and have been used in bioprocess control applications \citep{Gadkar:2003aa}.
+However, cybernetic control heuristics are not mechanistic, instead they are the output of an optimal decision with respect to a set of hypothetical physiological objectives. 
+Thus, they are abstractions which are difficult to translate into a specific biological mechanism. 
+Our approach addresses the shortcomings of both regulatory constraints based models and cybernetic models. 
+First, similar to cybernetic models, the core of our approach is a kinetic model. Thus, we are able to directly calculate the time evolution of metabolism, for example the dynamic abundance of 
+network metabolites. Second, similar to regulatory flux balance analysis, our control laws describe specific mechanistic motifs such a activation or inhibition of enzyme activity. 
+However, our rules are continuous, thus they potentially allow a finer grained description metabolic regulation and control mechanisms.
+Lastly, we can naturally incorporate non modeled factors and categorical factors or factor combinations into our control law formulations.
 
-%Controlling hemorrhage has been a driving force to understand the molecular basis of blood coagulation.
-%Replacement therapy improves bleeding times for most hemophilia patients. However, repeated fVIII/fIX administration can lead to neutralizing inhibitors. 
-%Alternatively, rFVIIa has been used as a universal hemostatic agent to initiate clotting with and without fVIII or fIX inhibitors \cite{Hedner:2008rm}.
-%Despite its wide adoption, the mechanism and the utility of rFVIIa in trauma-associated hemorrhage remains controversial \cite{Duchesne:2008hw}. 
-%The model described the extrinsic activation of thrombin, in the presence of ATIII, with and without the protein C pathway.
-%The coagulation network architecture was based on literature and a previous mathematical model from our laboratory \cite{VARNER2007}.
-%Simulations of thrombin dynamics were independently validated using blood drawn from patients with coronary artery disease (reported here) and previous TF/FVIIa and FVIIa 
-%\emph{in-vitro} studies \cite{BUTENAS2002}.
-
-%Additionally, using sensitivity analysis on the ensemble of models, we identified 
-%structurally sensitive components of the coagulation architecture as a function of condition.
-%We also demonstrated that the ensemble robustly constrained model predictions of independent validation sets, despite having many poorly constrained parameters.
-%Thus, the model ensemble displayed \emph{sloppy} behavior similar to that observed by Sethna and coworkers for other signal transduction networks \cite{Gutenkunst2007}.  
-%By studying a family of models, we perhaps partially addressed the uncertainty stemming from the many poorly characterized model parameters.
-%However, many other factors could influence our results. For example, missing structural interactions or biophysical factors such as blood flow could play a large role.
-%Thus, while the results presented here may be a valuable first step, more studies are required.
-% \cite{ALLEN2007}.
-
-%We used this ensemble of parameters in all subsequent simulations to estimate the uncertainty in model simulations following from uncertainty in the model parameters.
-
-\section*{Discussion}
-In this study, we developed a reduced order model of the human coagulation cascade.
-We chose coagulation because it is well studied, has a complex architecture, 
-and has an abundance of experimental data available for model identification and validation. 
-However, coagulation was just a proof-of concept validation of our approach; 
-the proposed hybrid framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
-or potentially even cell free metabolic systems. 
-The model consisted of five differential equations augmented with several logical rules describing 
-regulatory connections between model components, and unmodeled interactions in the network. 
-We estimated model parameters from in vitro extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
-To estimate parameters, the residual between model simulations and experimental measurements was minimized using particle swarm optimization (PSO). 
-However, not all of the model parameters were uniquely identifiable, given the training data.
-Instead, we estimated an ensemble of likely parameter sets (N = 20)
-from eight \emph{in vitro} time-series coagulation data sets with and without the protein C pathway.
-We validated the ensemble using thrombin data sets taken from multiple laboratories for a variety of experimental conditions not used during training. 
-The ensemble predicted thrombin trajectories for conditions not used for model training, including thrombin generation for normal and hemophilic coagulation 
-in the presence of platelets (a significant unmodeled component).
-We then used flux analysis to understand how the network operated in a variety of conditions, and global sensitivity analysis to identify which parameters controlled the 
-performance of the network. Taken together, the reduced order hybrid approach produced a surprisingly predictive model, suggesting this approach could potentially 
-be used to model other biochemical networks important to human health.
-
-Malfunctions in coagulation can have potentially fatal consequences.
-Aggressive clotting involved with Coronary Artery Diseases (CADs), collectively accounts for 38\% of all deaths in North America \cite{HANSSON2005}.
-Coagulation management during surgery can also be challenging, particularly with the increasing clinical use of antithrombotic drugs \cite{Tanaka:2009wo}.   
-Insufficient coagulation due to genetic disorders such as hemophilia can also result in recurrent bleeding. 
-The coagulation factors VIII (fVIII) and IX (fIX) are deficient in Hemophilia A and B, respectively \cite{COOPER1994,MANNUCCI2001,MITCHELL2008}.
-People with mild hemophilia have 5-40\% of the normal clotting factor levels while severe hemophiliacs have $<$1\% \cite{MITCHELL2008}.
-Hemophilia can be controlled with regular infusions of the deficient clotting factors. 
-However, clotting factor replacement sometimes leads to the formation of fVIII and fIX inhibitors \emph{in-vivo} \cite{TOMOKIYO2003}.
-Alternatively, recombinant factor VIIa (rFVIIa) has been used to treat bleeding disorders \cite{Hedner:2008rm,Talbot:2009kk} including
-hemophilia with and without factor VIII/IX inhibitors \cite{Shapiro:2008px}.
-However, rFVIIa requires frequent administration (every 2-3 hr), and many questions remain about its mechanism of action, 
-its effective dose range \cite{TOMOKIYO2003}, and its overall utility for the treatment trauma-associated hemorrhage \cite{Duchesne:2008hw}. 
-In this study, we did not model rFVIIa induced coagulation directly. 
-Rather, we modeled a general trigger which initiated the extrinsic coagulation cascade. 
-Since we identified the model using TF/FVIIa, inherent to our rFVIIa simulations (and the rate constant governing initiation) 
-was the presence of TF. However, even with this complication, 
-the model generated potentially useful insight into the rFVIIa mechanism of action, and its possible shortcomings especially for the treatment of hemophilia. 
-The addition of rFVIIa directly activated thrombin through the initiation pathway. 
-However, no amplification of the thrombin signal occurred without fVIII or fIX. 
-Thus, the peak thrombin signal was lower than normal coagulation, the peak thrombin time was longer, and the thrombin signal was eventually inhibited by ATIII and the protein C pathway.
-However, as the dose of rFVIIa increased, the peak thrombin time decreased (eventually saturating around 100$\times$nominal trigger), and the peak thrombin value increased.
-For 100$\times$nominal trigger, the thrombin profile resembled normal coagulation.
-Butenas \emph{et al.,} performed an extensive in vitro study of rFVIIa induced thrombin generation under normal and hemophilic conditions \citep{Butenas:2002ab}.
-They found qualitatively similar trends, namely rFVIIa could restore normal coagulation (even in the absence of TF) for large enough rFVIIa doses, although rFVIIa induced coagulation 
-in hemophilia (even for large rFVIIa doses) lagged the normal profile. 
-These results suggest that rFVIIa administration alone
-might not be able to initiate normal coagulation in recurrent bleeding, unless the dosage is above a critical threshold. 
-Unfortunately, this threshold is likely patient specific, thus limiting potentially limiting the utility of rFVIIa administration.
-
-Although the performance of the proof of principle coagulation model was impressive given its size, there were several mathematical and biological questions that can be explored further.
-First, the choice of max/min integration rules, or the particular form of the transfer functions could be generalized to include other rule types and functions. 
-Theoretically, an integration rule is simply a function whose domain is a set of transfer function inputs, and whose range is $v\in[0,1]$.
-Thus, integration rules other than max/min could be used, such as the mean or the product, assuming the range of the transfer functions is always $f\in[0,1]$.
-Alternative integration rules such the mean would have different properties which could influence model identification or performance. 
-For example, a mean or product integration rule would be differentiable, which allows derivative based optimization approaches to be used. 
-The particular form of the transfer function could also be explored. We choose a hill-like function because of its 
-prominence in the systems and synthetic biology community. 
-However, the only mathematical requirement for a transfer function is that it must map a non-negative continuous or categorical variable into the 
-range $f\in[0,1]$. Thus, many function types are possible.
-There are several additional biological modules that can be added to the core model presented here. 
-First, we need to update the model to included thrombin induced platelet activation. 
-While we were able to capture thrombin generation data in the presence of platelets, 
-the initial shape of the activation curve and the time-scale of activation was not always consistent with the data.
-Platelets are activated by thrombin through the cleavage of the extracellular domain of protease-activated receptors (PARs) on the platelet surface.
-Next, we should add the intrinsic or contact thrombin activation pathway to the model. 
-The intrinsic coagulation pathway is triggered by contact activation of the plasma protease factor XII (fXII) by negatively charged surfaces.
-The activation of fXII is followed by proteolytic activation of factor XI (fXI) and the activation of downstream factors like fIX \cite{NAITO1991}.
-It has also been suggested that activated platelets can release polyphosphate which directly activates fXII \cite{Smith:2006zl}. 
-Lastly, to make the model more clinically relevant, we should include the biochemical processes responsible for clot formation, and clot dissolution (fibrinolysis).
-Clot formation is driven by thrombin activity, while fibrinolysis is driven by plasmin activity, a key enzyme that cleaves fibrin (one of the main materials in a clot).
-Similar to coagulation, fibrinolysis is managed by several activating and inhibitory factors which control the balance between clot formation and dissolution.
-Tissue plasminogen activator (t-PA) and urokinase activate plasmin, along with contact pathway factors such fXIa. On the other hand, thrombin activatable fibrinolysis inhibitor (TAFI)
-inhibits the degradation of fibrin by plasmin. Also similar to coagulation, there is considerable fibrinolysis and contact pathway data sets that can be used to train the model.
+There are several critical questions that should be explored following this proof of concept study.
+It is unclear how parameter identification will scale to genome scale networks, and second it is unclear how we will identify allosteric connectivity at a genome scale. 
+The enzymatic connectivity for genome scale cell free networks can easily be established by stripping away the growth and cell wall machinery from whole cell genome reconstructions.
+Then metabolic fluxes can transformed into kinetic expressions using heuristics such multiple saturation kinetics, which are then modified by our rule based control variables.  
+This leaves a large number of unknown kinetic constants that must be estimated from time-resolved metabolite measurements.
+We showed that particle swarm optimization quickly identified an ensemble of model parameters, at least for proof of concept metabolic networks using synthetic data.
+This suggested that we can expect reasonable model predictions, despite only partial parameter knowledge, as network size grows if we have properly designed experiments. 
+Brown and Sethna showed in a model of signal transduction that good predictions were possible despite only order of magnitude estimates of parameter values \citep{Brown:2003aa}.
+Sethna and coworkers later showed that model performance is often controlled by only a few parameter combinations, 
+a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
+We have also demonstrated \textit{sloppy} behavior in a wide variety of signal transduction processes \citep{Luan:2007aa,Song:2009aa,Tasseff:2010aa,Tasseff:2011aa,Nayak:2011aa,Lequieu:2011aa}.
+Thus, given our previous experience with models with hundreds of unknown parameters, we expect parameter estimation to be a manageable challenge. 
+On the other hand, a critical challenge will be the estimation of allosteric connectivity at a genome scale. 
+The regulation of glycolytic enzymes, such as phosphofructokinase I, has been studied for many years \citep{BERG_BIOCHEMISTRY,Peskov:2008aa}. 
+The allosteric regulation of metabolic enzymes can also be established from organism specific databases, such as EcoCyc \citep{Keseler:2013aa} 
+or more general allosteric databases such as the AlloSteric Database \citep{Huang:2014aa}. 
+However, for those enzymes that have not been well studied, we will need to infer allosteric interactions from experimental data.
+In general, the reverse engineering of regulatory network structure from data is a very difficult problem.
+There are many different approaches from the reverse engineering of gene regulatory networks that perhaps could be adopted to this problem, however this remains an open question.
 
@@
 \subsection*{Formulation and solution of the model equations.}
-We used ordinary differential equations (ODEs) to model the time evolution of proteins ($x_{i}$) in our reduced order coagulation model:
-\begin{equation}
-	\frac{dx_{i}}{dt}  =  \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)\qquad{i=1,2,\hdots,\mathcal{M}}\\
-\end{equation}
-where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of protein species in the model. 
+We used ordinary differential equations (ODEs) to model the time evolution of metabolite ($x_{i}$) and scaled enzyme abundance ($\epsilon_{i}$) in hypothetical cell-free metabolic networks:
+\begin{eqnarray}
+	\frac{dx_{i}}{dt} & = & \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)\qquad{i=1,2,\hdots,\mathcal{M}}\\
+	\frac{d\epsilon_{i}}{dt} & = & -\lambda_{i}\epsilon_{i}\qquad{i = 1,2,\hdots,\mathcal{E}}
+\end{eqnarray}where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of metabolites and $\mathcal{E}$ denotes the number of enzymes in the model. 
 The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of reaction $j$. 
-Typically, reaction $j$ is a non-linear function of biochemical species abundance, as well as unknown kinetic parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$). 
+Typically, reaction $j$ is a non-linear function of metabolite and enzyme abundance, as well as unknown kinetic parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$). 
 The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$. 
-If $\sigma_{ij}>0$, species $i$ is produced by reaction $j$. 
-Conversely, if $\sigma_{ij}<0$, species $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates species $i$ is not connected with reaction $j$. 
-The system material balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$, which were specified by the experimental setup.
+If $\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$. 
+Conversely, if $\sigma_{ij}>0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates metabolite $i$ is not connected with reaction $j$. 
+Lastly, $\lambda_{i}$ denotes the scaled enzyme degradation constant. 
+The system material balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$ and $\mathbf{\epsilon}\left(t_{o}\right) = \mathbf{1}$ (initially we have 100\% cell-free enzyme abundance).
 
@@
 
-The control term $v_{j}$ depended upon the combination of factors which influenced the activity of enzyme $i$.
+The allosteric regulation term $v_{j}$ depended upon the combination of factors which influenced the activity of enzyme $i$.
 For each enzyme, we used a rule-based approach to select from competing control factors (Fig. \ref{fig-control-schematic}). 
@@
 The quantities $j^{+}$ and $j^{-}$ denoted the sets of activating and inhibitory factors for enzyme $j$. 
-If a process has no modifying factors, we set $v_{j} = 1$.
+If an enzyme had no allosteric factors, we set $v_{j} = 1$.
 There are many possible functional forms for $0\leq f_{ij}\left(\mathcal{Z}\right)\leq 1$. 
@@
 The $\kappa_{ij}$ parameter was species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).
-Applying the general framework to the reduced coagulation network resulted in five ordinary differential equations:
-\begin{eqnarray}
-	\frac{dx_{1}}{dt} & =& -\left(r_{init}v_{init} + r_{amp}v_{amp}\right)\\
-	\frac{dx_{2}}{dt} &=& r_{amp}v_{amp} + r_{init}v_{init} - r_{inh,ATIII}v_{inh,ATIII}\\
-	\frac{dx_{3}}{dt} &=& -r_{apc}v_{apc} \\
-	\frac{dx_{4}}{dt} &=& r_{apc}v_{apc} \\
-	\frac{dx_{5}}{dt} & = & -r_{inh,ATIII}v_{inh,ATIII}
-\end{eqnarray}where $\mathbf{x} = \left(fII,FIIa,PC,APC,ATIII\right)^{T}$. 
-The terms $r_{*}v_{*}$ in the balance equations denote corrected kinetic expressions for initiation, amplification and inhibition processes. 
-The rate of initiation $\bar{r}_{init}$ was modeled as:
-\begin{equation}
-	\bar{r}_{init} = k_{init}\left(trigger\right)\frac{x_{1}}{K_{init,fII} + x_{1}}
-\end{equation}where $k_{init},~K_{init,fII}$ are the rate and saturation constants governing initiation, respectively. 
-The rate of initiation was modified by $v_{init}$, the control parameter governing initiation. 
-Initiation was sensitive to level of trigger (activator), and TFPI (inhibitor):
-\begin{equation}
-	v_{init} = \min\left(f^{-}_{init}\left(TFPI\right),f^{+}_{init}\left(trigger\right)\right)
-\end{equation}where the transfer functions $f$ took the form of Eqn \eqref{eqn:control-factor}.
-The rate of thrombin amplification was given by:
-\begin{equation}
-	\bar{r}_{amp} = k_{amp}\left(x_{2}\right)\frac{x_{1}}{K_{amp,fII}+x_{1}}
-\end{equation}where $k_{amp},~K_{amp,fII}$ denote the rate and saturation constants governing amplification, respectively. 
-The amplification control term, which modified amplification rate, was modeled as a combination of multiple inhibition terms and one activation term:
-\begin{equation}
-	v_{amp} = \min\left(f^{-}_{amp}\left(TFPI\right),f^{-}_{amp}\left(x_{4}\right),f^{+}_{amp}\left(\mathcal{Z}_{amp}\right)\right)
-\end{equation}where $\mathcal{Z}_{amp} = fV\times fX\times fVIII\times fIX$. Although, $f^{+}_{amp}\left(\mathcal{Z}_{amp}\right)$ is an activating term, 
-we included it in the min integration rule; the factors in $\mathcal{Z}_{amp}$ were essential for amplification (if any of these factors was missing
-the amplification reaction would not occur). Thus, the factors in $\mathcal{Z}_{amp}$ were required components, a classification
-that we implemented by the min selection rule.
-The rate activated protein C formation was given by:
-\begin{equation}
-	\bar{r}_{apc} = k_{APC,formation}\left(TM\right)\frac{x_{3}}{K_{formation,PC} + x_{3}}
-\end{equation}where $k_{APC,formation},~K_{formation,PC}$ denote the rate and saturation constants governing activated protein C formation, respectively and $TM$ denotes
-the thrombomodulin abundance. We modeled the control term which governed APC formation as a single thrombin dependent activation term:
-\begin{equation}
-	v_{apc} = \max\left(f_{apc}^{+}\left(x_{2}\right)\right)
-\end{equation}
-Lastly, we included direct irreversible inhibition of FIIa by ATIII:
-\begin{equation}
-	\bar{r}_{inh,ATIII} = k_{ATIII,inhibition}\left(x_{5}x_{2}^{\gamma}\right)
-\end{equation}where $\gamma$ was estimated to be $\gamma = 1.26$. For ATIII inhibition of FIIa, the control variables $v_{inh,ATIII}$ was taken to be unity.
-The model equations were encoded using the Python programming language and solved using the ODEINT routine of the SciPy module \citep{SCIPY}.
-The model files can be downloaded from http://www.varnerlab.org.
+The model equations were encoded using the Octave programming language and solved using the LSODE routine in Octave \citep{Octave:2014}.
 
-\subsection*{Estimation of model parameters from experimental data.}
-Model parameters were estimated by minimizing the difference between simulations and experimental thrombin measurements (squared residual):
+\subsection*{Estimation of model parameters and structures from synthetic experimental data.}
+Model parameters were estimated by minimizing the difference between simulations and synthetic experimental data (squared residual):
 \begin{equation}\label{eqn:objective-function}
@@
 value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$. The outer summation is respect to
-time, while the inner summation is with respect to state. We minimized the model residual using Particle swarm optimization (PSO) \citep{PSO}.
+time, while the inner summation is with respect to state. We approximated a realistic model identification scenario, assuming noisy experimental data, 
+limited sampling resolution (approximately 20 minutes per sample) and a limited number of measurable metabolites.
+
+We minimized the model residual using particle swarm optimization (PSO) \citep{PSO}.
 PSO uses a \textit{swarming} metaheuristic to explore parameter spaces. 
@@
 parameters ($\mathcal{K}\times{1}$). In thus study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$. The quality of parameter
-estimates was measured using goodness of fit (model residual). The particle swarm optimization routine was implemented in the Python programming language. 
-All plots were made using the Matplotlib module of Python \citep{Matplotlib}.
-The optimization code can be downloaded from http://www.varnerlab.org.
+estimates was measured using two criteria, goodness of fit (model residual) and angle between the estimated parameter vector $\mathbf{k}_{j}$ and the true parameter set $\mathbf{k}^{*}$:
+\begin{equation}
+	\alpha_{j} = cos^{-1}\left(\frac{\mathbf{k}_{j}\cdot{\mathbf{k^{*}}}}{\norm{\mathbf{k}_{j}}\norm{\mathbf{k^{*}}}}\right)
+\end{equation}If the candidate parameter set $\mathbf{k}_{j}$ were perfect, the residual between the model and synthetic data 
+and the angle between $\mathbf{k}_{j}$ and the true parameter set $\mathbf{k}^{*}$ would be equal to zero.
 
-\subsection*{Global sensitivity analysis of model performance.}
-We conducted a global sensitivity analysis to estimate which parameters controlled the performance of the reduced order model. 
-We computed the total sensitivity index of each parameter relative to two performance objectives, the peak thrombin time and the area under the thrombin curve 
-(thrombin exposure). [FINISH~ME].
+We modified our PSO implementation to simultaneously search over kinetic parameters and putative model control structures.
+In the combined case, each particle potentially carried a different model realization in addition to a different kinetic parameter vector. 
+We kept the update rules the same (along with the update parameters). 
+Thus, each particle competed on the basis of goodness of fit, which allowed different model structures to contribute to the overall behavior of the swarm. 
+We considered five possible model structures (A through E), where network A was the correct formulation (used to generate the synthetic data). 
+We considered a population N = 100 particles, where each particle in the swarm was assigned a model structure, and a random parameter vector.
+The PSO algorithm, model equations, and the objective function were encoded and solved in the Octave programming language \citep{Octave:2014}.
 
 \section*{Acknowledgements}
-This study was supported by an award from the Army Research Office (ARO \#59155-LS).
+This study was supported by an award from the National Science Foundation (MCB \#1411715) 
+and the Army Research Office (ARO \#59155-LS).
 
@@
 \centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-1-Network.pdf}
-\caption{Schematic of the connectivity of the reduced order coagulation model. A trigger compound e.g., TF/FVIIa initiates thrombin production (FIIa) from prothrombin (fII).
-Once activated, thrombin catalyzes its own activation (amplification step), as well as its own inhibition via the conversion of protein C to activated protein C (APC).
-APC and tissue factor pathway inhibitor (TFPI) inhibit initiation and amplification, while antithrombin III (ATIII) directly inhibits thrombin. 
-All inhibition steps, and trigger-induced initiation were modeled using a rule-based approach. 
-Likewise, the dependence of amplification on other coagulation factors was also modeled using a rule-based approach. 
-The abundance of the highlighted species (in the dashed boxes) was governed by an ordinary differential equation. 
-All other species were assumed to be constant.}\label{fig-network}
+\includegraphics[width=1.0\textwidth]{./figs/Figure-1-Networks.pdf}
+\caption{Proof of concept cell-free metabolic networks considered in this study. Substrate $S$ is converted to products P$_{1}$ and P$_{2}$ through a series of chemical conversions
+catalyzed by enzyme(s) $E_{j}$. The activity of the pathway enzymes is subject to both positive and negative allosteric regulation. }\label{fig-networks}
 \end{figure}
@@
 \includegraphics[width=1.0\textwidth]{./figs/Figure-2-ControlSchematic.pdf}
-\caption{Schematic of rule based effective control laws. 
-Traditional enzyme kinetic expressions e.g., Michaelis–Menten or multiple saturation kinetics are multiplied by an enzyme activity control variable $0 \leq v_{j} \leq 1 $. 
+\caption{Schematic of rule-based allosteric enzyme activity control laws. 
+Traditional enzyme kinetic expressions, e.g., Michaelis–Menten or multiple saturation kinetics, are multiplied by an enzyme activity control variable $0 \leq v_{j} \leq 1 $. 
 Control variables are functions of many possible regulatory factors encoded by arbitrary functions of the form $0\leq f_{j}\left(\mathcal{Z}\right)\leq 1$.
-At each simulation time step, the $v_{j}$ variables are calculated by evaluating integration rules such as the max or min of the set of factors $f_{1},\hdots,f_{n}$ 
+At each simulation time step, the $v_{j}$ variables are calculated by evaluating integration rules such as the max or min of the set of factors $f_{1},\hdots$ 
 influencing the activity of enzyme $E_{j}$. }\label{fig-control-schematic}
@@
 \centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-3-DIFF_FII_NO_APC.pdf}
-\caption{Reduced order coagulation model training simulations.
-Reduced order coagulation model parameters were estimated using particle swarm optimization (PSO) with and without the protein C pathway as a function of prothrombin. 
-Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data. 
-The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation).
-(A,B,C) training results for 150\%, 100\% and 50\% of physiological prothrombin levels in the absence of the protein C pathway.
-Thrombin generation was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
-All factors and control proteins were at their physiological concentration unless others denoted. 
-The experimental training data was reproduced from the study of Butenas et al. \citep{Butenas:1999aa}.}\label{fig-diff_fII_NO_APC}
+\includegraphics[width=1.0\textwidth]{./figs/Figure-3-EnzymeKinetics.pdf}
+\caption{Kinetics of simple transformations in the presence of activation and inhibition. 
+\textbf{A}:The conversion of substrate $S$ to product $P$ by enzyme $E$ was activated by $S$. 
+For a fixed control gain parameter $k_{control}$, the reaction rate approached a step for increasing control order $N$. 
+\textbf{B}:The conversion of substrate $S$ to product $P$ by enzyme $E$ with inhibitor $I$. 
+For a fixed control gain parameter $k_{control}$, the reaction rate approximated non-competitive inhibition for increasing control order $N$. 
+}\label{fig-kinetics-simulations}
 \end{figure}
@@
 \centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-4-DIFF_FII_APC.pdf}
-\caption{Reduced order coagulation model training simulations.
-Reduced order coagulation model parameters were estimated using particle swarm optimization (PSO) with and without the protein C pathway as a function of prothrombin. 
-Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data.
-The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation).
-(A,B,C) training results for 150\%, 100\% and 50\% of physiological prothrombin levels in the presence of the protein C pathway. 
-Only APC pathway parameters were allowed to vary in the simulations on the right.
-Thrombin generation was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
-All factors and control proteins were at their physiological concentration unless others denoted.
-The experimental training data was reproduced from the study of Butenas et al. \citep{Butenas:1999aa}.}\label{fig-diff_fII_APC}
+\includegraphics[width=0.50\textwidth]{./figs/Figure-4-OnOffSimulations.pdf}
+\caption{ON/OFF control simulations for network A and network B for an ensemble of kinetic parameter sets versus time (N = 100). 
+For each case, N = 100 simulations were conducted using kinetic and initial conditions generated randomly from a hypothetical true parameter set. 
+The gray area represents $\pm$ one standard deviation surrounding the mean. 
+Control parameters were fixed during the ensemble calculations.
+\textbf{A}: End product P$_{1}$ abundance versus time for Network A. 
+The abundance of P$_{1}$ decreased with end product inhibition of $E_{1}$ activity (Control-ON) versus the no inhibition case (Control-OFF). 
+\textbf{B}: End product P$_{2}$ abundance versus time for Network B. Inhibition of branch point $E_{6}$ by end product P$_{1}$ decreased P$_{2}$ abundance (Control-ON) versus the
+no inhibition case (Control-OFF).
+\textbf{C}: End product P$_{1}$ abundance versus time for Network A. 
+Inhibition of branch point $E_{6}$ by end product P$_{1}$ decreased P$_{1}$ abundance (Control-ON) versus the no inhibition case (Control-OFF).
+}\label{fig-onoff-simulations}
 \end{figure}
@@
 \centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-5-FII_ATIII.pdf}
-\caption{Reduced order coagulation model predictions versus experimental data for normal coagulation. 
-The reduced order coagulation model parameter estimates were tested against data not used during model training.
-Simulations of different levels of prothrombin and ATIII were compared with experimental data in the absence of the protein C pathway. 
-Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data.
-The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation). 
-(A,B,C,D) prediction results for (FII,ATIII): (50\%,150\%), (100\%, 100\%), (125\%, 75\%) and (150\%, 50\%) of physiological prothrombin and ATIII levels in the absence of the protein C pathway. 
-Thrombin generation was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
-All factors and control proteins were at their physiological concentration unless others denoted.
-The experimental validation data was reproduced from the study of Butenas et al. \citep{Butenas:1999aa}}\label{fig-fII_ATIII}
+\includegraphics[width=1.0\textwidth]{./figs/Figure-5-ParameterFit.pdf}
+\caption{Parameter estimation from synthetic data for the same and mismatched allosteric control logic using particle swarm optimization (PSO). 
+Synthetic experimental data was generated from a hypothetical parameter set using Network A, 
+where substrate $S$, end product P$_{1}$ and intermediate M$_5$ were sampled approximately every 20 minutes. 
+For cases $\textbf{A,B}$ 20 particles were initialized with randomized parameters and allowed to search for 300 iterations. 
+\textbf{A,B}: PSO estimated an ensemble of parameters sets (N = 20) consistent with the synthetic experimental data assuming the correct
+enzymatic and control connectivity starting from randomized initial parameters.
+\textbf{C,D}: In the presence of control mismatch (Network B control policy simulated with Network A kinetic parameters) 
+the ensemble of models did not describe the synthetic data. 
+}\label{fig-parameter-fit}
 \end{figure}
@@
 \centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-6-Hemophilia.pdf}
-\caption{Reduced order coagulation model predictions versus experimental data with and without FVIII and FIX.
-The reduced order coagulation model parameter estimates were tested against data not used during model training. 
-Simulations of normal thrombin formation with ATIII and the protein C pathway were compared with thrombin formation in the absence of fVIII and fIX.
-Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data.
-The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation). 
-(A,B) prediction results for normal thrombin generation and thrombin generation in hemophilia. 
-All factors and control proteins were at their physiological concentration unless others noted.
-The experimental validation data was reproduced from the study of Allen et al. \citep{ALLEN2006}.}\label{fig-hemophilia}
+\includegraphics[width=1.0\textwidth]{./figs/Figure-6-AlternativeNetworks.pdf}
+\caption{Schematic of the alternative allosteric control programs used in the structural particle swarm computation. 
+Each network had the same enzymatic connectivity, initial conditions and kinetic parameters, 
+but alternative feedback control structures for the first enzyme in the pathway.}\label{fig-alternative-networks}
 \end{figure}
@@
 \centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-7-rFVIIa-Trigger.pdf}
-\caption{Reduced order coagulation model predictions of rFVIIa administration. 
-A: Simulations of thrombin formation in the presence of ATIII and the protein C pathway were conducted for a range of trigger values (1x - 200x nominal) in the absence of fVIII and fIX.
-B: Comparison of thrombin generation for normal versus hemophilia for 10x nominal trigger.
-Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles.
-The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation). 
-All factors and control proteins were at their physiological concentration unless others noted.}\label{fig-rFVII}
+\includegraphics[width=0.5\textwidth]{./figs/Figure-7-ControlSearch.pdf}
+\caption{Combined control and kinetic parameter search using modified particle swarm optimization (PSO). 
+A population of N = 100 particles was initialized with randomized kinetic parameters and one of five possible control configurations (Network A - E).
+Simulation error was minimized for a synthetic data set 
+($S$, end product P$_{1}$ and intermediate M$_5$ sampled approximately every 20 min) generated using Network A.
+\textbf{A}: Simulation error versus parameter set angle for N = 100 particles biased toward the correct regulatory program (A,B,C,D,E) = (40\%, 10\%, 20\%, 20\% and 10\%).
+\textbf{B}: Simulation error versus parameter set angle for N = 100 uniformly distributed particles (A,B,C,D,E) = (20\%, 20\%, 20\%, 20\% and 20\%).
+\textbf{C}: Simulation error versus parameter set angle for N = 100 negatively biased particles (A,B,C,D,E) = (10\%, 40\%, 10\%, 20\% and 20\%).
+Network A (the correct structure) was preferentially identified for positively and uniform biased particle distributions, but misidentified in the presence of
+a large incorrect bias. 
+}\label{fig-control-search}
 \end{figure}
@@
 \centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-8-Flux-rFVIIa-Trigger.pdf}
-\caption{Reaction flux distribution as a function of time for thrombin generation under normal (left), hemophilia (center) and rFVIIa treated hemophilia (right).
-Reaction flux was calculated for each particle at $T = 0,4,6,8,10,12,14$ min after the initiation of coagulation. 
-Reaction fluxes were calculated for each particle in the parameter ensemble (N = 20).
-Blue colors denote low flux values while red colors denote high flux values.}\label{fig-flux-analysis}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-9-Sensitivity.pdf}
-\caption{Global sensitivity analysis of the reduced order coagulation model with with respect to the model parameters.
-A: Sensitivity analysis of the thrombin peak time for different prothrombin levels (150\%,100\% and 50\% of the physiological value) as a function of activated protein C. 
-B: Sensitivity analysis of the thrombin exposure for different prothrombin levels (150\%,100\% and 50\% of the physiological value) as a function of activated protein C. 
-Points denote the mean total sensitivity value, while the area around each point denotes the uncertainty in the sensitivity value. 
-The gray dashed line denotes the 45$^{o}$ degree diagonal, if sensitivity values are equal for different conditions they will lie on the diagonal.  
-Sensitivity values significantly above or below the diagonal indicate differentially important model parameters.
-}\label{fig-sensitivity}
+\includegraphics[width=1.0\textwidth,height=0.6\textheight]{./figs/Figure-8-Flux.pdf}
+\caption{Metabolic flux and control variables as a function of network type and particle index at t = 100 min.
+The control variables governing $E_{1},E_{3}$ and $E_{6}$ activity and the scaled metabolic flux and were calculated for the positively, 
+uniformly and negatively biased particle swarms (N = 100). The particles from each swarm were sorted based upon simulation error (low to high error). 
+\textbf{A}: Model performance for the positively biased particle swarm as a function of particle index.
+\textbf{B}: Model performance for the uniformly biased particle swarm as a function of particle index.
+\textbf{C}: Model performance for the negatively biased particle swarm as a function of particle index. 
+Models with significant control mismatch showed distinct control and flux patterns versus those models with the correct or closely related control policies.
+In particular, models with the correct control policy showed stronger inhibition of $E_{1}$ activity, leading to decreased flux from S$\rightarrow$P$_{1}$.
+Conversely, models with significant mismatch had increased $E_{1}$ activity, leading to an altered flux distribution. 
+This is especially apparent in the negatively biased particle swarm. 
+}\label{fig-flux-pattern}
 \end{figure}

### 2014-11-11 Paper_v2.tex
response/edits for Reviewer 2

--- 2014-10-28 Paper_v2.tex
@@
 \vspace{0.1in}
-{\par \centering \textbf{Running Title:}~Reduced order modeling of blood coagulation}
+{\par \centering \textbf{Running Title:}~Effective models of metabolism}
 \vspace{0.1in}
-{\par \centering \textbf{To be submitted:}~\emph{Processes}}
+{\par \centering \textbf{To be submitted:}~\textit{Processes}}
 \vspace{0.5in}
@@
 \section*{Abstract}
-In this study, we present a novel modeling approach which combines ordinary differential equation (ODE) 
-modeling with logical rules to simulate an archetype biochemical network, the human coagulation cascade.  
-The model consisted of five differential equations augmented with several logical rules describing 
-regulatory connections between model components, and unmodeled interactions in the network. 
-This formulation was more than an order of magnitude smaller than current coagulation models, 
-because many of the mechanistic details of coagulation were encoded as logical rules.
-We estimated an ensemble of likely model parameters (N = 20) from \textit{in~vitro} extrinsic coagulation data sets, with and without inhibitors, 
-by minimizing the residual between model simulations and experimental measurements using particle swarm optimization (PSO). 
-We then validated the model ensemble using thrombin data sets that were not used during training. 
-The ensemble predicted thrombin trajectories for conditions not used for model training, including thrombin generation for normal and hemophilic coagulation 
-in the presence of platelets (a significant unmodeled component).
-We then used flux analysis to understand how the network operated in a variety of conditions, and global sensitivity analysis to identify which parameters controlled the 
-performance of the network. Taken together, the hybrid approach produced a surprisingly predictive model given its small size, 
-suggesting the proposed hybrid framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
-or potentially even cell free metabolic systems.
+Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
+Many of the challenges confronting genome-scale kinetic modeling can potentially be overcome in a cell-free system. 
+For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth. 
+Thus, cell-free operation holds several significant advantages for model development, identification and validation. 
+Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms, such as \textit{E.
+coli}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.
+Toward this unmet need, we present an effective biochemical network modeling framework for building dynamic cell-free metabolic models.
+The key innovation of our approach is the integration of simple effective rules encoding complex allosteric regulation with traditional kinetic pathway modeling. 
+We tested our approach by modeling the time evolution of several hypothetical cell-free metabolic networks. 
+We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, captured complex allosteric patterns such as ultrasensitivity or non-competitive inhibition in the absence of mechanistic information.
+Second, when integrated into network models, these rules captured classic regulatory patterns such as product-induced feedback inhibition. 
+Lastly, we showed, at least for the network architectures considered here, that we could simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data. 
+While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.
 
-\vspace{0.25in}
-{\noindent \textbf{Keywords:}~Blood coagulation, Mathematical modeling, Systems biology}
+{\noindent \textbf{Keywords:}~Cell-free metabolism, Mathematical modeling}
 
@@
 
-%We chose coagulation because it is well studied, has a complex architecture, 
-%and has an abundance of experimental data available for model identification and validation. 
-%However, coagulation was just a proof-of concept to validate our approach; 
-%
+\section*{Introduction}
+Mathematical modeling has long contributed to our understanding of metabolism. 
+Decades before the genomics revolution, mechanistically, structured metabolic models arose from the desire to predict microbial phenotypes resulting from changes in intracellular or extracellular states \citep{1976_fredrickson_BiotechBioeng}.
+The single cell \textit{E.
+coli} models of Shuler and coworkers pioneered the construction of large-scale, dynamic metabolic models that incorporated multiple, regulated catabolic and anabolic pathways constrained by experimentally determined kinetic parameters \citep{1984_domach_shuler_BiotechBioeng_01}.
+Shuler and coworkers generated many single cell kinetic models, including single cell models of eukaryotes \citep{1989_steinmeyer_shuler_ChemEngSci,1992_wu_shuler_AnnNYAcadSci}, minimal cell architectures \citep{2004_castellanos_shuler_PNAS}, as well as DNA sequence based whole-cell models of \textit{E.
+coli} \citep{2008_atlas_shuler_IETSysBio}.
+Conversely, highly abstracted kinetic frameworks, such as the cybernetic framework, represented a paradigm shift, viewing cells as growth-optimizing strategists \citep{1985_dhurjati_ramkrishna_tsao_BiotechBioeng}.
+Cybernetic models have been highly successful at predicting metabolic choice behavior, e.g., diauxie behavior \citep{1986_kompala_ramkrishna_tsao_BiotechBioeng}, steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, as well as the cellular response to metabolic engineering modifications \citep{1999_varner_ramkrishna_MetaEng}.
+Unfortunately, traditional, fully structured cybernetic models also suffer from an identifiability challenge, as both the kinetic parameters and an abstracted model of cellular objectives must be estimated simultaneously.
+Recent cybernetic formulations from Ramkrishna and colleagues have successfully treated this identifiability challenge through elementary mode reduction, though the techniques replace detailed biological mechanism with an optimization heuristic \cite{2009_song_ramkrishna_BiotechBioeng,Song:2011aa}.
 
-%Developing mathematical modeling tools to interrogate the molecular basis of human disease is one of the grand challenges facing systems biology.
-%Developing computational tools to understand the molecular basis of human disease is a grand challenge facing systems biology.
-%Many have suggested that the integration of experimental and computational research is required to unravel
-%critical questions facing molecular medicine \cite{KITANO2002,ASSMUS2006,ARNAUD2006}. 
-%Despite this, mechanistic network modeling has not played a significant role in the development of new therapies for cancer, cardiovascular diseases or for acute events  
-%like thrombosis during surgery.
-%A critical limiting issue often cited has been model uncertainty \cite{BAILEY2001}. 
-%Development and validation of detailed network models requires a significant investment of time and material resources. 
-%Alternatively, development of tools using toy models often fails to demonstrate relevance to biologists or physicians.  
-%What is needed is a model system, that is simultaneously tractable and directly applicable to the study of disease.
-%Human blood is perhaps such an ideal model system \cite{Diamond:2009yf}. 
-%Blood is a complex mixture of proteins and cells which collectively carry out many functions.
-%However, one well studied facet of blood, which could serve as a valuable model system, is the coagulation cascade. 
-%Understanding the mechanism of rFVIIa might provide insight into the therapeutic dose under different physiological conditions.
-%More generally, rFVIIa treatment could also serve an ideal test case to develop strategies which retrospectively estimate operational perturbations associated with drugs.
-%For example, the effect of rFVIIa is directly observable in coagulation outputs such as thrombin concentration.
-%Thus, by analyzing these outputs in a normal versus a perturbed network, it might be possible to reverse engineer the molecular influence associated with a therapeutic intervention.
+In the post genomics world, large-scale stoichiometric reconstructions of microbial metabolism popularized by static, constraint-based modeling techniques such as flux balance analysis (FBA) have become standard tools \citep{2012_lewis_palsson_NatRevMicrobio}.
+Since the first genome-scale stoichiometric model of \textit{E.
+coli}, developed by Edwards and Palsson \citep{2000_edwards_palsson_PNAS}, well over 100 organisms, including industrially important prokaryotes such as \textit{E.
+coli} \citep{Feist:2007aa} or \textit{B.
+subtilis} \citep{Oh:2007aa}, are now available \citep{2009_feist_palsson_NatRevMicrobio}.
+Stoichiometric models rely on a pseudo-steady-state assumption to reduce unidentifiable genome-scale kinetic models to an underdetermined linear algebraic system, which can be solved efficiently even for large systems.
+Traditionally, stoichiometric models have also neglected explicit descriptions of metabolic regulation and control mechanisms, instead opting to describe the choice of pathways by prescribing an objective function on metabolism.
+Interestingly, similar to early cybernetic models, the most common metabolic objective function has been the optimization of biomass formation \citep{2002_ibarra_edwards_palsson_Nat}, although other metabolic objectives have also been estimated \citep{2007_schuetz_sauer_MolSysBio}.
+Recent advances in constraint-based modeling have overcome the early shortcomings of the platform, including capturing metabolic regulation and control \citep{2013_hyduke_lewis_palsson_MolBioSys}. 
+Thus, modern constraint-based approaches have proven extremely useful in the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \citep{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}.
+However, genome-scale kinetic models of industrial important organisms such as \textit{E.
+coli} have yet to be constructed.
 
-%
-%Thus, while there are many possible strategies to simulate biochemical networks, ODE models remain a common choice despite
-%significant identification challenges.
-%This result highlights the perhaps under-appreciated role that experimental design could play in generating the best training and validation data sets for model identification \citep{Apgar:2010fk}.
+Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
+Central amongst these advantages is direct access to metabolites and the microbial biosynthetic machinery without the interference of a cell wall. 
+This allows us to control as well as interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
+Second, cell-free systems also allow us to study biological processes without the complications associated with cell growth.
+Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \citep{Jewett:2008aa}. 
+However, CFPS is not new; CFPS in crude \textit{E. coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \citep{MATTHAEI:1961aa,NIRENBERG:1961aa}. 
+Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \citep{Lu:2014aa} to synthetic biology \citep{Hodgman:2012aa}. 
+Interestingly, many of the challenges confronting genome-scale kinetic modeling can potentially be overcome in a cell-free system. 
+For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth. 
+Thus, cell-free operation holds several significant advantages for model development, identification and validation.
+Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms, such as \textit{E.
+coli} or \textit{B.
+subtilis}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.
 
-\section*{Introduction}
-Developing mathematical models of biochemical networks is a significant facet of systems biology.
-Modeling approaches differ in their degree of detail, where the choice of approach is often determined by prior knowledge, or model requirements \citep{2012_kholodenko_kolch_SciSig}. 
-Ordinary differential equation (ODE) models are common tools for modeling biochemical systems because of their ability to capture dynamics and encode mechanism.
-However, ODE models typically come with difficult (or sometimes impossible) parameter identification problems. 
-For example, Gadkar \textit{et al.}, showed that even with near-perfect information, 
-it was often impossible to identify all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
-However, it is not clear whether we actually need precise estimates for all model parameters. 
-Bailey suggested more than a decade ago, that achieving qualitative or even quantitative understanding of biological systems 
-should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}. 
-Since Bailey's complex biology with no parameters hypothesis, Sethna showed that model performance is typically 
-sensitive to only a few parameters, a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
-Thus, reasonable predictions may be possible, despite parametric uncertainty, if a few critical parameters are well-defined. 
-Perhaps more importantly, ODE models also require significant mechanistic information, thereby limiting their utility in poorly understood systems.
-Toward this challenge, logical modeling is an emerging paradigm that encodes causal relationships between model components using non-linear transfer functions, 
-without requiring mechanism \citep{Morris:2010aa}. 
-Despite their simplicity, logical models have captured rich behaviors in a variety of systems important to human health \citep{Saez-Rodriguez:2011aa,Morris:2011ys,Morris:2012aa}. 
-However, modeling complex dynamics with logical models is challenging. 
-Thus, there is an unmet need for a third approach which combines ODEs and logical models, where ODEs could encode mechanistic information, 
-while missing or incomplete mechanistic knowledge can be approximated using a logical approach.
-
-In this study, we developed a hybrid approach which combined ODE modeling with logical rules to model a well studied biochemical network, 
-the human coagulation system. Coagulation is an archetype proteolytic cascade involving both positive 
-and negative feedback \cite{Butenas:2002aa,Schenone:2004aa,Adams:2009aa}. 
-Coagulation is mediated by a family proteases in the circulation, called factors and a key group of blood cells, called platelets. 
-The central process in coagulation is the conversion of prothrombin (fII), an inactive coagulation factor, to the master protease thrombin (FIIa).
-Thrombin generation involves three phases, initiation, amplification and termination \cite{GOLDHABER2006, Brummel:2002aa}.
-Initiation requires a trigger event, for example vessel injury, which leads to the activation of factor VII (FVIIa).
-Two converging pathways, the extrinsic and intrinsic cascades, then process and amplify this initial coagulation signal. 
-The extrinsic cascade is generally believed to be the main mechanism of thrombinogenesis in the blood \cite{MANN1990,ROBERTS1998,MANN1999}.
-Initially, thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa), which itself has been activated by TF/FVIIa \cite{Butenas:2002aa}. 
-Picomolar amounts of thrombin then activate the cofactors factors V and VIII (fV and fVIII) and platelets, 
-leading to the formation of the tenase and prothrombinase complexes on activated platelets.
-These complexes amplify the early coagulation signal by further activating FXa, and directly converting prothrombin to thrombin. 
-There are several control points in the cascade that inhibit thrombin formation, and eventually terminate thrombin generation. 
-Tissue Factor Pathway Inhibitor (TFPI) inhibits FXa formation catalyzed by TF/FVIIa, while antithrombin III (ATIII)
-neutralizes several of the proteases generated during coagulation, including thrombin.  
-Thrombin itself also inadvertently plays a role in its own inhibition; thrombin, through interaction with thrombomodulin, protein C and endothelial cell protein C receptor (EPCR),
-converts protein C to activated protein C (APC) which attenuates the coagulation response by proteolytic cleavage of fV/FVa and fVIII/FVIIIa. 
-Termination occurs after either prothrombin is consumed, or thrombin formation is neutralized by inhibitors such as APC or ATIII.
-
-Previous coagulation models have typically been formulated as systems of nonlinear ordinary differential equations, 
-using mass action or more complex kinetics, to describe the rates of biochemical conversions \citep{Kuharsky:2001aa}.
-Mechanistic ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys} 
-were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, Hockin \emph{et al.} \citep{Hockin:2002aa}, and later Butenas et al., \citep{Butenas:2004aa}
-who developed and then subsequently refined highly mechanistic models of coagulation. 
-Other laboratories have also expanded upon Hockin \emph{et al.}, for example including the intrinsic pathway, and thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
-Other aspects of coagulation have also been modeled, such as platelet biochemistry \citep{Stalker:2013aa}, 
-multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and transport inside clots \citep{Voronov:2013aa}.
-However, these previous studies were largely based upon extensive mechanistic knowledge. 
-This is possible because blood, while enormously complex, can be systematically interrogated.
-Other systems, such as intracellular signaling networks, are much more difficult to experimentally interrogate.  
-Towards this unmet need, we formulated a hybrid modeling approach which combines ODEs and logical rules to model 
-biochemical processes for which a complete mechanistic understanding is missing. 
-We tested this approach by modeling the human coagulation cascade. 
-The hybrid model consisted of only five differential equations augmented with several logical rules.
-Thus, the model was more than an order of magnitude smaller than comparable purely ODE models in the literature.
-We estimated the model parameters from \textit{in~vitro} extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
-We then compared the model predictions with thrombin data sets, for both normal and hemophilic coagulation, that were not used for model training.
-Once validated, we performed flux and sensitivity analysis on the model to estimate which parameters were critical to model performance in several conditions. 
-The reduced order hybrid approach produced a surprisingly predictive coagulation model, 
-suggesting this framework could potentially be used to model other biochemical networks important to human health.
+In this study, we present an effective biochemical network modeling framework for building dynamic cell-free metabolic models.
+The key innovation of our approach is the seamless integration of simple effective rules encoding complex regulation with traditional kinetic pathway modeling. 
+This integration allows the description of complex regulatory interactions, such as time-dependent allosteric regulation of enzyme activity, in the absence of specific mechanistic information. 
+The regulatory rules are easy to understand, easy to formulate and do not rely on overarching theoretical abstractions or restrictive assumptions. 
+We tested our approach by modeling the time evolution of several hypothetical cell-free metabolic networks. 
+In particular, we tested whether our effective modeling approach could describe classically expected enzyme kinetic behavior, and second whether we could simultaneously estimate kinetic parameters and regulatory connectivity, in the absence of specific mechanistic knowledge, from synthetic experimental data.
+Toward these questions, we explored five hypothetical cell-free networks. 
+Each network shared the same enzymatic connectivity, but had different allosteric regulatory connectivity. 
+We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, captured complex allosteric patterns such as ultrasensitivity or non-competitive inhibition in the absence of mechanistic information.
+Second, when integrated into network models, these rules captured classical regulatory patterns such as product-induced feedback inhibition. 
+Lastly, we showed, at least for the network architectures considered here, that we could simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data. 
+While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.
 
@@
 
-%However, components could also correspond to non-modeled influences, categorial variables or other abstract quantities.
-%In the current proof-of-principle model, 
-%APC can inhibit amplification, and upstream of amplification thus APC activity will directly slow the rate of thrombin formation from the initiation trigger, and amplification itself.
-%On the other hand, ATIII inhibits thrombin directly.
+\subsection*{Formulation and properties of effective cell-free metabolic models.}
+We developed two proof-of-concept metabolic networks to investigate the features of our effective biochemical network modeling approach (Fig. \ref{fig-networks}). 
+In both examples, substrate S was converted to the end products P$_{1}$ and P$_{2}$ through a series of enzymatically catalyzed reactions, including a branch point at hypothetical metabolite M$_{2}$. 
+Several of these reactions involved cofactor dependence (AH or A), and various allosteric regulatory mechanisms modified the activity of pathway enzymes. 
+Network A included feedback inhibition of the initial pathway enzyme ($E_{1}$) by pathway end products P$_{1}$ and P$_{2}$ (Fig. \ref{fig-networks}A). 
+On the other hand, network B involved feedback inhibition of $E_{1}$ by P$_{2}$ and $E_{6}$ by P$_{1}$ (Fig. \ref{fig-networks}B). 
+In both networks, branch point enzymes $E_{3}$ and $E_{6}$ were subject to feed-forward activation by reduced cofactor AH. 
+Lastly, it is known experimentally that cell-free systems have a finite operational lifespan. 
+Loss of biosynthetic capability could be a function of many factors, e.g., cofactor or metabolite limitations. 
+We modeled the loss of biosynthetic capability as a non-specific first-order decay of enzyme activity.
 
-\subsection*{Formulation of reduced order coagulation models.}
-We developed a reduced order extrinsic coagulation model to test our hybrid modeling approach (Fig. \ref{fig-network}).
-The core of our model was based upon the earlier work of Ismagilov and coworkers \citep{Runyon:2004aa,Kastrup:2006aa,Runyon:2007aa,Runyon:2008aa}, 
-where we added initiation, factor dependence, and specific inhibition terms to the earlier simplified models.
-A trigger event initiates thrombin formation (FIIa) from prothrombin (fII) through a lumped initiation step.
-This step loosely represents the initial activation of thrombin by activated FXa.
-Once activated, thrombin catalyzes its own formation (amplification step), and inhibition via the conversion of protein C to activated protein C (APC).
-Antithrombin III (ATIII) inhibits amplification, while APC and tissue factor pathway inhibitor (TFPI) potentially inhibit both initiation and amplification. 
-All initiation and inhibition processes, as well as the dependence of amplification upon other coagulation factors, was approximated using our rule-based approach (Fig. \ref{fig-control-schematic}). 
-Individual regulatory contributions to the activity of pathway enzymes were integrated into control coefficients ($v$'s) using an integration rule (min/max).  
-These control coefficients then modified the rates of model processes at each time step.
-Hill-like transfer functions $0 \leq f\left(\mathcal{Z}\right) \leq 1$ quantified the contribution of components upon a target process. 
-Components were either individual inhibitor or activator levels or some function of levels, e.g., the product of factor levels. 
-In this study, $\mathcal{Z}$ corresponded to the abundance of individual inhibitors or activators, 
-with the exception of the dependence of amplification upon specific coagulation factors (modeled as the product of factors). 
-When a process was potentially sensitive to multiple inputs, logical integration rules were used to select which transfer functions influenced the process at any given time. 
-In our proof of concept model, we used a winner takes all strategy; the maximum or minimum transfer function was selected at any given time step. 
-However, other integration rules are certainly possible. 
-Taken together, while the reduced order coagulation model encodes significant biological complexity, it is highly compact (consisting of only five differential equations). 
-Thus, it will serve as an excellent proof of principle example to study the reduction of a highly complex human subsystem.
+Allosteric regulation of enzyme activity was modeled by combining individual regulatory contributions to the activity of pathway enzymes into a control coefficient using an integration rule (Fig.
+\ref{fig-control-schematic}).
+This strategy is similar in spirit to the Constrained Fuzzy Logic (cFL) approach of Lauffenburger and coworkers which has been used to effectively model signal transduction pathways important in human health \citep{Morris:2011aa}.
+In our formulation, Hill-like transfer functions $0 \leq f\left(\mathcal{Z}\right) \leq 1$ were used to calculate the influence of factor abundance upon target enzyme activity. 
+In this context, factors can be individual metabolite levels or some function, e.g., the product of metabolite levels. 
+However, more generally, factors can also correspond to non-modeled influences, categorial variables or other abstract quantities. 
+In the current study, we simply let $\mathcal{Z}$ correspond to the abundance of individual metabolites, however in general this can be a complex function of both modeled and unmodeled factors. 
+When an enzyme was potentially sensitive to more than one regulatory input, logical integration rules were used to select which regulatory transfer function influenced enzyme activity at any given time.
+Thus, our test networks involved important features such as cofactor recycling, enzyme activity and metabolite dynamics, as well as multiple overlapping allosteric regulatory mechanisms.
 
-\subsection*{Identification of model parameters using particle swarm optimization.}
+The rule-based regulatory strategy approximated the behavior of classical allosteric activation and inhibition mechanisms (Fig. \ref{fig-kinetics-simulations}). 
+We considered the enzyme catalyzed conversion of substrate S to a product P, where the overall reaction rate was modeled as the product of a Michaelis-Menten term and an effective allosteric control variable reflecting the particular regulatory interaction.
+We first explored feed-forward substrate activation of enzyme activity (for both positive and negative cooperativity). 
+Consistent with classical data, the rule-based strategy predicted a sigmoidal relationship between substrate abundance and reaction rate as a function of the cooperativity parameter (Fig.
+\ref{fig-kinetics-simulations}A).
+For cooperativity parameters less than unity, increased substrate abundance \textit{decreased} the reaction rate. 
+This was consistent with the idea that substrate binding \textit{decreased} at regulatory sites, which negatively impacted substrate binding at the active site. 
+On the other hand, as the cooperativity parameter increased past unity, the rate of conversion of substrate S to product P by enzyme $E$ approached a step function. 
+In the presence of an inhibitor, the rule-based strategy predicted non-competitive like behavior as a function of the cooperativity parameter (Fig. \ref{fig-kinetics-simulations}B). 
+When the control gain parameter, $\kappa_{ij}$ in Eqn.
+\eqref{eqn:control-factor}, was greater than unity, the inhibitory force was directly proportional to the cooperativity parameter, $\eta$ in Eqn.
+\eqref{eqn:control-factor}.
+Thus, as the cooperativity parameter increased, the maximum reaction rate decreased (Fig. \ref{fig-kinetics-simulations}B). 
+Interestingly, our rule-based approach was unable to directly simulate competitive inhibition of enzyme activity. 
+Taken together, the rule-based strategy captured classical regulatory patterns for both enzyme activation and inhibition. 
+Thus, we are able to model complex kinetic phenomena such as ultrasensitivity, despite an effective description of reaction kinetics.
+
+End product yield was controlled by feedback inhibition, while product selectivity was controlled by branch point enzyme inhibition (Fig. \ref{fig-onoff-simulations}). 
+A critical test of our modeling approach was to simulate networks with known behavior. 
+If we cannot reproduce the expected behavior of simple networks, then our effective modeling strategy, and particularly the rule-based approximation of allosteric regulation, will not be feasible for genome-scale cell-free problems.
+We considered two cases, control ON/OFF, for each network configuration.
+Each of these cases had identical kinetic parameters and initial conditions; the \textit{only} differences between the cases were the allosteric regulation rules and the control parameters associated with these rules.
+As expected, end product accumulation was larger for network A when the control was OFF (no feedback inhibition of $E_{1}$ by P$_{1}$ and P$_{2}$),
+as compared to the ON case (Fig. \ref{fig-onoff-simulations}A). 
+We found this behavior was robust to the choice of underlying kinetic parameters, as we observed that same qualitative response across an ensemble of 100 randomized parameter sets, for fixed control parameters.
+The control ON/OFF response of network B was more subtle. 
+In the OFF case, the behavior was qualitatively similar to network A. 
+However, for the ON case, flux was diverted away from P$_{2}$ formation by feedback inhibition of $E_{6}$ activity at the M$_{2}$ branch point by P$_{1}$ (Fig. \ref{fig-onoff-simulations}B). 
+Lower $E_{6}$ activity at the M$_{2}$ branch point allowed more flux toward P$_{1}$ formation, hence the yield of P$_{1}$ also increased (Fig. \ref{fig-onoff-simulations}C). 
+Again, the control ON/OFF behavior of network B was robust to changes in kinetic parameters, as the same qualitative trend was conserved across an ensemble of 100 randomized parameters, for fixed control parameters.
+Taken together, these simulations suggested that the rule-based allosteric control concept could robustly capture expected feedback behavior for networks with uncertain kinetic parameters.
+
+\subsection*{Estimating parameters and effective allosteric regulatory structures.}
 A critical challenge for any dynamic model is the estimation of kinetic parameters. 
-We estimated kinetic and control parameters simultaneously from eight \emph{in vitro} time-series coagulation data sets with and without the protein C pathway. 
-The residual between model simulations and experimental measurements was minimized using particle swarm optimization (PSO). 
-A population of particles (N = 20) was initialized with randomized kinetic and control parameters and allowed to search for parameter vectors that minimized
-the residual. However, not all parameters were varied simultaneously. 
-We partitioned the parameter estimation problem into two subproblems based upon the biological organization of the training data; 
-(i) estimation of parameters associated with thrombin formation in the absence of the protein C pathway and (ii) estimation of parameters associated with the protein C pathway.
-Only those parameters associated with each subproblem were varied during the optimization procedure for that subproblem, e.g., thrombin parameters were \textit{not} varied during the protein C subproblem.The PSO procedure was run for 20 generations for each subproblem, where each generation was 1200 iterations.
-The best particle from each generation was used to generate the particle population for the next generation.
-We rotated the subproblems, starting with subproblem 1 in the first generation.
+For metabolic processes, there is also the added challenge of identifying the regulation and control structures that manage metabolism. 
+Of course, these issues are not independent; any description of enzyme activity regulation will be a function of system state, which in turn depends upon the kinetic parameters. 
+For cell-free systems, regulated gene expression has been removed, however, enzyme activity regulation is still operational. 
+We explored this linkage by estimating model parameters from synthetic data using both network structures. 
+We generated noise corrupted synthetic measurements of the substrate S, intermediate M$_{5}$ and end product P$_1$ approximately every 20 min using
+network A.
+We then generated an ensemble of model parameter estimates by minimizing the difference between model simulations and the synthetic data using particle swarm optimization (PSO), starting from random initial parameter guesses.
+The estimation of kinetic parameters was sensitive to the choice of regulatory structure (Fig.
+\ref{fig-parameter-fit}).
+PSO identified an ensemble of parameters that bracketed the mean of the synthetic measurements in less than 1000 iterations when the control structure was correct (Fig.
+\ref{fig-parameter-fit}A and B).
+However, with control mismatch (network B simulated with network A parameters), model simulations were not consistent with the synthetic data  (Fig. \ref{fig-parameter-fit}C and D). 
+Taken together, these results suggested that we could perhaps simultaneously estimate both parameters and network control architectures, as incorrect control structures would be manifest as poor model fits.
 
-The reduced order coagulation model captured the role of initial prothrombin abundance, and the decay of the thrombin signal following from ATIII activity (Fig. \ref{fig-diff_fII_NO_APC}).
-However, we systematically under-predicted the thrombin peak and the strength of ATIII inhibition in this training data set.
-On the other hand, with fixed thrombin parameters, we captured peak thrombin values and the decay of the thrombin signal (at least for the 150\% fII case) in the presence
-of both ATIII and the protein C pathway (Fig. \ref{fig-diff_fII_APC}). Lastly, we were unable to capture global differences in initiation time \textit{across} separate data 
-sets with a single ensemble of model parameters. These differences likely resulted from normal experimental variability.
-For example, different thrombin generation experiments within the 
-training data (at the same physiological factor levels) had significantly different initiation times (data not shown). 
-However, the inability to globally capture initiation time also highlighted a potential shortcoming of the initiation module within the model. 
-To capture the variability in initiation time \textit{across} training data sets, we included a constant time-delay parameter ($T_{D}$) for each data group.
-The delay parameter was constant within a data set, but allowed to vary \textit{across} training data sets. 
-Introduction of the delay parameter allowed the model to simulate multiple training data sets using a single ensemble of model parameters.
-Taken together, the model identification results suggested 
-that our hybrid approach could reproduce a panel of thrombin generation data sets in the neighborhood of physiological factor and inhibitor concentrations.
-However, it was unclear whether the reduced order model could predict new data, without updating the model parameters.
+We modified our particle swarm identification strategy to simultaneously search over both kinetic parameters and putative control structures. 
+In addition to our initial networks, we constructed three additional presumptive network models, each with the same enzymatic connectivity but different allosteric regulation of the pathway enzymes (Fig.
+\ref{fig-alternative-networks}).
+We then initialized a population of particles, each with one of the five potential regulatory programs and randomized kinetic parameters. 
+Thus, we generated an initial population of particles that had \textit{both} different kinetic parameters as well as different control structures. 
+We biased the distribution of the particle population according to our \textit{a prior} belief of the correct regulatory program. 
+To this end, we considered three different priors, a uniform distribution where each putative regulatory structure represented 20\% of the population and two mixed distributions that were either positively or negatively biased towards the correct structure (network A).
+In both the positively biased and uniform cases the PSO clearly differentiated between the true or closely related structures and those that were materially different (Fig. \ref{fig-control-search}). 
+As expected, the positively biased population (40\% of the initial particle population seeded with network A) gave the best results, where the correct structure was preferentially identified (Fig.
+\ref{fig-control-search}A).
+On the other hand, when given a uniform distribution, the PSO approach identified a combination of network A and network C as the most likely control structures (Fig. \ref{fig-control-search}B). 
+Network A and C differ by the regulatory connection between the end product P$_2$ and enzyme $E_{1}$; in network A, end product P$_{2}$ was assumed to inhibit $E_{1}$, while in network C, end product P$_{2}$ activated $E_{1}$.
+Lastly, when the initial population was biased towards incorrect structures (initial population seeded with 90\% incorrect structures), the particle swarm \textit{misidentified} the correct allosteric structure (Fig.
+\ref{fig-control-search}C).
+Interestingly, while each particle swarm identified parameter sets that minimized the simulation error, the estimated parameter values were not necessarily similar to the true parameters. 
+The angle between the estimated and true parameters was not consistently small across the swarms (identical parameters would give an angle of zero). 
+This suggested that our particle swarm approach identified a \textit{sloppy} ensemble, i.e., parameter estimates that were individually incorrect but collectively exhibited the correct model behavior.
 
-\subsection*{Validation of the reduced order coagulation model.}
-We tested the predictive power of the reduced order coagulation model with validation data sets not used during model training. 
-Two validation data sets were used, thrombin generation for various prothrombin and ATIII concentrations with the protein C pathway, 
-and thrombin generation in normal versus hemophilic plasma in the presence of the protein C pathway. 
-Lastly, we compared the qualitative output of the model to rFVIIa addition in the presence of hemophilia.
-The hemophilia case was an especially difficult test as it was taken from a different study which used a plasma-based
-\textit{in~vitro} assay involving platelets instead of phospholipid vesicles (PCPS). 
-All kinetic and control parameters were fixed for the validation simulations. 
-The only globally adjustable parameter $T_{D}$, was fixed within each validation data set but allowed to vary between data sets.
-The reduced order model predicted the thrombin generation profile for ratios of prothrombin and ATIII in the absence of the protein C pathway (Fig. \ref{fig-fII_ATIII}).
-Simulations near the physiological range (fII,ATIII) = (100\%, 100\%) or (125\%,75\%) tracked the measured thrombin values (Fig. \ref{fig-fII_ATIII}B and C).
-On the other hand, predictions for factor levels outside of the physiological range (fII,ATIII) = (50\%, 150\%) or (150\%, 50\%), while qualitatively consistent with measured thrombin values, did show significant
-deviation from the measurements (Fig. \ref{fig-fII_ATIII}A and D).
-Likewise, simulations of thrombin generation in normal versus hemophilia (missing both fVIII and fIX) were consistent with measured thrombin values (Fig. \ref{fig-hemophilia}).
-We modeled the dependence of thrombin amplification on factor levels using a product rule ($\mathcal{Z} = fV\times fX\times fVIII\times fIX$), 
-which was then was integrated using a $\min$ integration rule into the control variable governing amplification. 
-Thus, in the absence of fVIII or fIX, the amplification control variable evaluated to zero, and the only thrombin produced was from initiation (Fig. \ref{fig-hemophilia}B).
-However, the decay of the thrombin signal was underpredicted in the normal case (Fig. \ref{fig-hemophilia}A), while the activated thrombin level was overpredicted 
-in hemophilia simulations, although thrombin generation was far less than normal (Fig. \ref{fig-hemophilia}B). 
-Taken together, the reduced order model performed well in the physiological range of factors, even with
-unmodeled components such as platelet activation in the hemophilia data set.
-
-The model ensemble predicted a direct correlation between thrombin generation and rFVIIa addition in hemophilia (Fig. \ref{fig-rFVII}).
-In the current model, we cannot distinguish between different initiation sources, e.g., TF/FVIIa versus rFVIIa, as we have only a single lumped initiation source (trigger).
-Thus, we simulated the addition of rFVIIa in hemophilia by removing fVIII and fIX from the model, and modulating the initial level of trigger.
-Simulations with a baseline level of trigger were consistent with the previous hemophilia simulations, where the only thrombin produced was from initiation (Fig. \ref{fig-rFVII}A, 1$\times$~trigger).
-However, as we increased the trigger strength, the thrombin profile began to approximate normal coagulation, showing a pronounced peak albeit with a slower peak time (Fig. \ref{fig-rFVII}B).
-Further increases in trigger strength resulted in decreased thrombin peak time and increased maximum thrombin values (Fig. \ref{fig-rFVII}A, 50$\times$~trigger). 
-Thus, for large trigger values (200$\times$trigger), the hemophilic thrombin profile approximated normal coagulation, where peak thrombin was achieved shortly 
-after administration and 95\% of the thrombin was gone by 20 min after initiation. 
-We performed flux analysis to understand how the reduced order coagulation model balanced initiation, amplification and inhibition of thrombin formation for normal coagulation and hemophilia.
-Analysis of the reaction flux through the reduced order network for thrombin generation in normal, hemophilia and rFVIIa-treated hemophilia identified three distinct operational modes (Fig.
-\ref{fig-flux-analysis}).
-We calculated the flux through four lumped reactions, initiation, amplification, thrombin-induced APC generation and total thrombin inhibition (including both APC and ATIII action).
-Directly after the addition of a trigger (e.g., TF/FVIIa or rFVIIa), the lumped initiation flux was the largest for all three cases.
-However, within a few minutes enough thrombin was generated by the initiation mechanism to induce the amplification stage. 
-During amplification, thrombin catalyzes its own formation and inhibition by generating activated protein C (APC), a potent inhibitor of the coagulation cascade.
-For normal coagulation, amplification and thrombin inhibition were the dominate reactions by 6 min after initiation (Fig. \ref{fig-flux-analysis}, left). 
-After 10 min, the dominate reaction had shifted to thrombin inhibition (both ATIII and APC action). 
-In hemophilia (missing both fVIII and fIX), the amplification reaction did not occur, and thrombin was produced only by initiation (Fig. \ref{fig-flux-analysis}, center).
-Initiation was quickly inhibited by APC, and the thrombin level stabilized (eventually decaying at longer times because of ATIII activity). 
-Lastly, when 50$\times$trigger was used to induce thrombin formation in hemophilia (absence of fVIII/fIX), 
-initiation mechanisms dominated for up to 6 min following initiation (Fig. \ref{fig-flux-analysis}, right). 
-Similar to hemophilia alone, no amplification occurred in the 50$\times$trigger+hemophilia case, 
-and the rate of thrombin generation was extinguished by the combined action of ATIII and APC. 
-Taken together, the hybrid modeling approach captured the transition between the modes of thrombin generation, as well as the role that inhibitors play 
-in attenuating the thrombin generation rate. Thus, the transfer function approach encoded the inhibitory logic of this cascade in the absence of specific mechanism.
+We calculated control program output and scaled metabolic flux for the positively, uniformly and negatively biased particle swarms (Fig. \ref{fig-flux-pattern}). 
+Network A and network C models from the positively (Fig.
+\ref{fig-flux-pattern}A) and uniformly (Fig.
+\ref{fig-flux-pattern}B) biased particle swarms showed similar operational patterns, despite differences in kinetic parameters and control structures.
+While models from the negatively biased population had error values similar to the correct structures in the previous swarms, they have different flux and control profiles (Fig.
+\ref{fig-flux-pattern}C).
+In all cases, regardless of network configuration or parameter values, the rate of enzyme decay was small compared to the other fluxes, and all networks had qualitatively similar trends for $E_{3}$ and $E_{6}$ control.
+Moreover, consistent with the correct model structure, production of end product P$_1$ was the preferred branch for all model configurations. 
+However, there was variability in P$_{2}$ production flux across the population of models, especially for the uniform swarm when compared with the other cases. 
+High P$_{1}$ branch flux resulted in end product inhibition of $E_{1}$ in both network A and network C, however in network D and E, high P$_1$ flux induced $E_1$ activation. 
+These trends were manifested in different flux profiles, where the negatively biased population appeared more uniform across the population compared with the other swarms, and had higher $E_{1}$ specific activity.
+Interestingly, the behavior of network A and network C highlighted an artifact of our integration rule; both a positive or negative feedback connection from P$_2$ to $E_{1}$ were ignored because the P$_{1}$ inhibition of $E_{1}$ was dominate.
+Thus, while theoretically distinct, network A and network C appeared operationally to the PSO algorithm to be that same network.
+On the other hand, networks B, D and E showed distinct behavior that was not consistent with the true network. 
+These architectures exhibited either limited inhibition (network B) or activation (network D and E) of $E_{1}$ activity, resulting in significantly different metabolic flux profiles. 
+However, the PSO was able to find low error parameter solutions, despite the mismatch in the control structures (error values similar, but not better than the best network A and network C estimates). 
+Taken together, these results suggested that a uniform sampling approach could potentially yield an unbiassed estimate of both kinetic parameters and control structures. 
+However, the negatively biased particle swarm results illustrated a potential shortcoming of the approach, namely convergence to a local error minimum despite a significantly incorrect control structure.
+This suggested that estimated model structures will need to be further evaluated, for example by generating falsifiable experimental designs which could distinguish between low error solutions.
 
@@
 
-\subsection*{Global sensitivity analysis of the reduced order coagulation model.}
-We conducted a global sensitivity analysis to estimate which parameters controlled the performance of the reduced order model. 
-We calculated the sensitivity of the time to maximum thrombin (peak time) and the thrombin exposure (area under the thrombin curve) 
-for different levels of prothrombin, and protein C (Fig. \ref{fig-sensitivity}). 
-Globally, 41\% of the parameters shifted in importance between the (fII,PC) = (50\%, 0\%) and (150\%,100\%) cases for the peak thrombin time (Fig. \ref{fig-sensitivity}A). 
-The majority of these shifts involved the interaction between increased prothrombin and the protein C pathway, while only
-5\% were directly associated with increased prothrombin alone. The rate constant for thrombin amplification was the most important parameter controlling the
-peak thrombin time. While this parameter was differentially important for different prothrombin levels, and in the presence or absence of 
-the activated protein C pathway, it was consistently the most sensitive parameter in the model. The saturation constant governing thrombin amplification was
-the second most important parameter, followed by the initiation control gain parameter. Other important parameters influencing the thrombin peak time included the control gain for
-activated protein C formation, and the rate constant controlling ATIII inhibition of thrombin activity. 
-On the other hand, only 27\% of the model parameters were differentially sensitive between the (fII,PC) = (50\%, 0\%) and (150\%,100\%) cases 
-for thrombin exposure (Fig. \ref{fig-sensitivity}B). Of these parameters, all of the shifts were associated with the interplay between thrombin formation and the protein C pathway. 
-The rate constant controlling ATIII inhibition was the most important parameter controlling the thrombin exposure.
-While this parameter was less important in the presence of protein C for 150\% prothrombin levels, it was significantly above all other parameters. 
-Similar to the peak time, for 150\% prothrombin, the control gain for activated protein C formation was differentially important along with the rate constant controlling
-amplification. However, the amplification parameter was much less important for thrombin exposure versus peak time.
+\section*{Discussion}
+In this study, we presented an effective kinetic modeling strategy to dynamically simulate cell-free biochemical networks. 
+Our proposed strategy integrated traditional kinetic modeling with an effective rules based approach to dynamically describe metabolic regulation and control. 
+We tested this approach by developing kinetic models of hypothetical cell-free metabolic networks. 
+In particular, we tested whether our effective modeling approach could describe classically expected behavior, and second whether we could simultaneously estimate kinetic parameters and regulatory connectivity, in the absence of specific mechanistic knowledge, from synthetic experimental data.
+Toward these questions, we explored five hypothetical cell-free networks. 
+In each network, a substrate S was converted to the end products P$_{1}$ and P$_{2}$ through a series of enzymatically catalyzed reactions, including a branch point at a hypothetical metabolite M$_{2}$.
+Each network also included the same cofactors and cofactor recycle architecture. 
+However, while all five networks shared the same enzymatic connectivity, each had different allosteric regulatory connectivity. 
+We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, could capture complex allosteric patterns such as ultrasensitivity, or non-competitive inhibition in the absence of specific mechanistic information.
+Moreover, when integrated into network models, these rules captured classical regulatory patterns such as product-induced feedback inhibition. 
+Lastly, we simultaneously estimated kinetic parameters and discriminated between competing regulatory structures, using synthetic data in combination with a modified particle swarm approach.
+If we considered all putative regulatory architectures to be equally likely, we were able to estimate a \textit{sloppy} ensemble of models with the correct architecture and kinetic parameters.
 
-\clearpage
+The proposed modeling strategy shares features with other popular techniques, but also has several key differences. 
+At its core, our effective modeling approach is similar to regulatory constraint-based methods, and to the cybernetic modeling paradigm developed by Ramkrishna and colleagues.
+Covert, Palsson and coworkers drastically improved the predictability of constraint-based approaches by integrating Boolean rules into the calculation of metabolic fluxes \citep{Covert:2001aa}.
+If the regulated intracellular flux problem is coupled with time-dependent extracellular balances, these models can predict complex behavior such as diauxie growth or the switch between 
+aerobic and anaerobic metabolism. Another important feature of this approach is that it scales with biological complexity. 
+For example, Covert \emph{et al.} showed that a genome-scale model of \emph{E.~coli} augmented with a Boolean rule layer, 
+correctly predicted approximately 80\% of the outcomes of a high-throughput growth phenotyping experiment in \emph{E. coli}.
+Further, they showed that they could learn new biology by iteratively refining the model and its associated rules \citep{Covert:2004aa}. 
+However, while regulated flux balance analysis is a powerful technique, it does not easily allow the calculation of time-resolved metabolite abundance. 
+Additionally, the Boolean rules which populate the regulatory layer are limited to ON/OFF decisions; for qualitative predictions of gene expression this is a reasonable limitation. 
+However, Boolean rules will likely be less effective at capturing dynamic allosteric regulation in a cell-free metabolic system. 
+On the other hand, the strength of cybernetic models is the integration of optimal metabolic control heuristics with traditional kinetic pathway modeling. 
+Cybernetic models are highly predictive; they have successfully predicted mutant behavior from limited wild-type data \citep{1999_varner_ramkrishna_MetaEng,Varner:2000aa,Song:2012aa}, 
+steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, strain specific metabolic function \citep{Song:2011aa} and have been used in bioprocess control applications \citep{Gadkar:2003aa}.
+However, cybernetic control heuristics are not mechanistic, instead they are the output of an optimal decision with respect to a set of hypothetical physiological objectives. 
+Thus, they are abstractions which are difficult to translate into a specific biological mechanism. 
+Our approach addresses the shortcomings of both regulatory constraint-based models and cybernetic models. 
+First, similar to cybernetic models, the core of our approach is a kinetic model.
+Thus, we are able to directly calculate the time evolution of metabolism, for example the dynamic abundance of network metabolites.
+Second, similar to regulatory flux balance analysis, our control laws describe specific mechanistic motifs, such as activation or inhibition of enzyme activity.
+However, our rules are continuous, thus they potentially allow a finer grained description of metabolic regulation and control mechanisms.
+Lastly, we can naturally incorporate unmodeled factors and categorical factors or combinations thereof into our control law formulations. 
+Though requiring a more complex description of cellular metabolism, our approach may even be extended to simulate cell-based systems by incorporating the same control laws into transcription factor activation and gene expression regulation.
 
-%Controlling hemorrhage has been a driving force to understand the molecular basis of blood coagulation.
-%Replacement therapy improves bleeding times for most hemophilia patients. However, repeated fVIII/fIX administration can lead to neutralizing inhibitors. 
-%Alternatively, rFVIIa has been used as a universal hemostatic agent to initiate clotting with and without fVIII or fIX inhibitors \cite{Hedner:2008rm}.
-%Despite its wide adoption, the mechanism and the utility of rFVIIa in trauma-associated hemorrhage remains controversial \cite{Duchesne:2008hw}. 
-%The model described the extrinsic activation of thrombin, in the presence of ATIII, with and without the protein C pathway.
-%The coagulation network architecture was based on literature and a previous mathematical model from our laboratory \cite{VARNER2007}.
-%Simulations of thrombin dynamics were independently validated using blood drawn from patients with coronary artery disease (reported here) and previous TF/FVIIa and FVIIa 
-%\emph{in vitro} studies \cite{BUTENAS2002}.
-
-%Additionally, using sensitivity analysis on the ensemble of models, we identified 
-%structurally sensitive components of the coagulation architecture as a function of condition.
-%We also demonstrated that the ensemble robustly constrained model predictions of independent validation sets, despite having many poorly constrained parameters.
-%Thus, the model ensemble displayed \emph{sloppy} behavior similar to that observed by Sethna and coworkers for other signal transduction networks \cite{Gutenkunst2007}.  
-%By studying a family of models, we perhaps partially addressed the uncertainty stemming from the many poorly characterized model parameters.
-%However, many other factors could influence our results. For example, missing structural interactions or biophysical factors such as blood flow could play a large role.
-%Thus, while the results presented here may be a valuable first step, more studies are required.
-% \cite{ALLEN2007}.
-
-%We used this ensemble of parameters in all subsequent simulations to estimate the uncertainty in model simulations following from uncertainty in the model parameters.
-
-\section*{Discussion}
-In this study, we developed a reduced order model of the human coagulation cascade.
-We modeled coagulation because it is well studied, has a complex architecture, 
-and has an abundance of experimental data available for model identification and validation. 
-However, coagulation was just a proof of concept test of our approach. 
-The proposed hybrid framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
-or potentially even cell free metabolic systems. 
-The model consisted of five differential equations augmented with several logical rules describing 
-regulatory connections between model components and unmodeled interactions in the network. 
-We estimated model parameters from \textit{in vitro} extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
-To estimate parameters, the residual between model simulations and experimental measurements was minimized using particle swarm optimization (PSO). 
-However, not all of the model parameters were uniquely identifiable, given the training data.
-Instead, we estimated an ensemble of likely parameter sets (N = 20)
-from eight \emph{in vitro} time-series coagulation data sets with and without the protein C pathway.
-Ensemble approaches have been used previously for other signal transduction models \citep{Kuepfer:2007aa,Song:2009aa,2010_song_varner_BiotechJ,Tasseff:2010aa,Lequieu:2011aa}, 
-and for metabolic models \citep{Tran:2008aa} to 
-estimate the impact of poorly constrained parameter values or poorly understood network structure on simulation performance. 
-Thus, ensemble approaches are common in the dynamic modeling community. 
-However, perhaps a unique feature of the current study is the direct connection between our particle swarm approach, 
-and the parameter ensemble; each particle in our swarm uniquely corresponded to a parameter set in our ensemble. 
-Thus, by constraining particles to operate in different parameter regions, giving each particle a different parameter combination to explore, 
-or perhaps even suppling a different model formulation to each particle we can effectively 
-traverse through complex parameter and model spaces. 
-We validated the ensemble using thrombin data sets taken from multiple laboratories for a variety of experimental conditions not used during training. 
-The ensemble predicted thrombin trajectories for conditions not used for model training, including thrombin generation for normal and hemophilic coagulation 
-in the presence of platelets (a significant unmodeled component).
-We then used flux analysis to understand how the network operated in a variety of conditions, and global sensitivity analysis to identify which parameters controlled the 
-performance of the network. Taken together, the proposed hybrid framework produced a surprisingly predictive model, suggesting this approach could 
-be used to effectively model other biochemical networks important to human health.
-
-Malfunctions in coagulation can have potentially fatal consequences.
-Aggressive clotting involved with Coronary Artery Diseases (CADs), collectively accounts for 38\% of all deaths in North America \cite{HANSSON2005}.
-Coagulation management during surgery can also be challenging, particularly with the increase in clinical use of antithrombotic drugs \cite{Tanaka:2009wo}.   
-Insufficient coagulation due to genetic disorders such as hemophilia can also result in recurrent bleeding. 
-The coagulation factors VIII (fVIII) and IX (fIX) are deficient in Hemophilia A and B, respectively \cite{COOPER1994,MANNUCCI2001,MITCHELL2008}.
-People with mild hemophilia have 5-40\% of the normal clotting factor levels while severe hemophiliacs have $<$1\% \cite{MITCHELL2008}.
-Hemophilia can be controlled with regular infusions of the deficient clotting factors. 
-However, clotting factor replacement sometimes leads to the formation of fVIII and fIX inhibitors \emph{in vivo} \cite{TOMOKIYO2003}.
-Alternatively, recombinant factor VIIa (rFVIIa) has been used to treat bleeding disorders \cite{Hedner:2008rm,Talbot:2009kk} including
-hemophilia with and without factor VIII/IX inhibitors \cite{Shapiro:2008px}.
-However, rFVIIa requires frequent administration (every 2-3 hr), and many questions remain about its mechanism of action, 
-its effective dosage \cite{TOMOKIYO2003}, and its overall utility for the treatment trauma-associated hemorrhage \cite{Duchesne:2008hw}. 
-In this study, we did not model rFVIIa-induced coagulation directly. 
-Rather, we modeled a general trigger which initiated the extrinsic coagulation cascade. 
-Since we identified the model using TF/FVIIa, inherent to our rFVIIa simulations (and the rate constant governing initiation) 
-was the presence of TF. However, even with this complication, 
-the model generated potentially useful insight into the rFVIIa mechanism of action, and its possible shortcomings especially for the treatment of hemophilia. 
-The addition of rFVIIa directly activated thrombin through the initiation pathway. 
-However, no amplification of the thrombin signal occurred without fVIII or fIX. 
-Thus, the peak thrombin signal was lower than normal coagulation, the peak thrombin time was longer, and thrombin generation was eventually inhibited by the combined action of ATIII and the protein C pathway.
-However, as the dose of rFVIIa increased, the peak thrombin time decreased (eventually saturating around 200$\times$nominal trigger), and the peak thrombin value increased such that
-the thrombin profile resembled normal coagulation.
-Butenas \emph{et al.} performed an extensive \textit{in vitro} study of rFVIIa-induced thrombin generation under normal and hemophilic conditions \citep{Butenas:2002ab}.
-They found qualitatively similar trends, namely rFVIIa restored normal coagulation (even in the absence of TF) for large enough rFVIIa doses, although rFVIIa-induced coagulation 
-in hemophilia (even for large rFVIIa doses) lagged the normal profile. 
-These results suggest that rFVIIa administration alone
-might not be able to initiate normal coagulation in recurrent bleeding, unless the dosage is well above a critical threshold.
-However, defining this threshold, which is likely patient specific, is difficult as there is tremendous patient to patient variability even with a normal
-coagulation phenotype \citep{Danforth:2012aa}.
-
-Although the performance of the proof of principle coagulation model was impressive given its limited size, there were several issues that could be further explored.
-First, the prediction of initiation time should be investigated. 
-We were able to estimate initiation time within a data set, but unable to predict initiation time \textit{across} independent data sets. 
-This suggested that we should update the initiation module to distinguish between different triggers, e.g., TF/FVIIa versus rFVIIa alone, 
-and to include key biological milestones such as FXa activation (a prerequisite to thrombin formation).
-Next, there are several additional biological modules that could be added to the core model presented here. 
-First, we could include thrombin-induced platelet activation and the role of activated platelets in amplification.
-We captured thrombin generation data in the presence of platelets, however,
-the initial shape of the activation curve and the time-scale of activation was not always consistent with the data.
-Platelets are activated by thrombin through the cleavage of the extracellular domain of protease-activated receptors (PARs) on the platelet surface.
-Once activated, platelets play an important role in amplification, and are key mediators of the positive feedback driving amplification.
-Thus, this biology is a potentially important component of an expanded model.
-We should also add the intrinsic pathway to the model. 
-The intrinsic pathway is triggered by contact activation of the plasma protease factor XI (fXI) by negatively charged surfaces and by thrombin and upstream factors such as activated
-plasma protease factor XII (FXIIa) \citep{NAITO1991,Gailani:1991aa}.
-Activated platelets may also release polyphosphate which directly activates fXII \cite{Smith:2006zl}. 
-Arguably a minor player in acute bleeding, contact activation could also be important in other wound healing contexts. 
-Finally, to make the model more clinically relevant, we should include the biochemical processes responsible for clot formation and clot dissolution (fibrinolysis).
-Clot formation is driven by thrombin activity, while fibrinolysis is driven by plasmin activity, a key enzyme that cleaves fibrin (one of the main materials in a clot).
-Similar to coagulation, fibrinolysis is managed by several activating and inhibitory factors which control the balance between clot formation and dissolution.
-Tissue plasminogen activator (t-PA) and urokinase activate plasmin, along with contact pathway factors such as fXIa. On the other hand, thrombin activatable fibrinolysis inhibitor (TAFI)
-inhibits the degradation of fibrin by plasmin. Also, similar to coagulation, there is considerable fibrinolysis and contact pathway data sets that can be used to train the model.
-Lastly, the choice of max/min integration rules or the particular form of the transfer functions could be generalized to include other rule types and functions. 
-Theoretically, an integration rule is a function whose domain is a set of transfer function inputs, and whose range is $v\in[0,1]$.
-Thus, integration rules other than max/min could be used, such as the mean or the product, assuming the range of the transfer functions is always $f\in[0,1]$.
-Alternative integration rules such as the mean might have different properties which could influence model identification or performance. 
-For example, a mean integration rule would be differentiable, which allows derivative-based optimization approaches to be used. 
-The particular form of the transfer function could also be explored. We choose a Hill-like function because of its 
-prominence in the systems and synthetic biology community. 
-However, the only mathematical requirement for a transfer function is that it map a non-negative continuous or categorical variable into the 
-range $f\in[0,1]$. Thus, many transfer function types are possible.
+There are several critical questions that should be explored following this proof-of-concept study.
+It is unclear how parameter identification will scale to genome-scale networks, and second it is unclear how we will identify allosteric connectivity at a genome-scale. 
+The enzymatic connectivity for genome-scale cell-free networks can easily be established by stripping away the growth and cell wall machinery from whole cell genome reconstructions.
+Then metabolic fluxes can be transformed into kinetic expressions using heuristics such multiple saturation kinetics, which are then modified by our rule-based control variables.  
+This leaves a large number of unknown kinetic constants that must be estimated from time-resolved metabolite measurements. 
+Ensemble modeling is a well-established approach for parameter identification in large-scale deterministic models. 
+Liao and coworkers developed a method that generates an ensemble of kinetic models that all approach the same steady-state, one determined by fluxomics measurements \cite{Tran:2008aa}. 
+The best subpopulation of candidate models are selected based on their agreement with further measurements of genetically perturbed systems. 
+Our work relies on heuristic search optimization to identify kinetic models consistent with steady-state and dynamic time-series measurements of cellular species \citep{Luan:2007aa,Song:2009aa,Tasseff:2010aa,Tasseff:2011aa,Nayak:2011aa,Lequieu:2011aa}.
+Instead of estimating a single yet highly uncertain parameter set, both approaches estimate an ensemble of parameter sets whose model behavior recapitulates experimental measurements. 
+Here, we showed that particle swarm optimization quickly identified an ensemble of model parameters, at least for proof-of-concept metabolic networks using synthetic data. 
+This suggested that we can expect reasonable model predictions, despite only partial parameter knowledge, as network size grows if we have properly designed experiments. 
+Brown and Sethna showed in a model of signal transduction that good predictions were possible despite only order of magnitude estimates of parameter values \citep{Brown:2003aa}.
+Sethna and coworkers later showed that model performance is often controlled by only a few parameter combinations, 
+a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
+We have also demonstrated \textit{sloppy} behavior in a wide variety of signal transduction processes \citep{Luan:2007aa,Song:2009aa,Tasseff:2010aa,Tasseff:2011aa,Nayak:2011aa,Lequieu:2011aa}.
+Thus, given our previous experience with models containing hundreds of unknown parameters, we expect parameter estimation to be a manageable challenge. 
+On the other hand, a critical challenge will be the estimation of allosteric connectivity at a genome scale. 
+The regulation of glycolytic enzymes, such as phosphofructokinase I, has been studied for many years \citep{BERG_BIOCHEMISTRY,Peskov:2008aa}. 
+The allosteric regulation of metabolic enzymes can also be established from organism specific databases, such as EcoCyc \citep{Keseler:2013aa}, 
+or more general allosteric databases, such as the AlloSteric Database \citep{Huang:2014aa}. 
+However, for those enzymes that have not been well studied, we will need to infer allosteric interactions from experimental data.
+In general, the reverse engineering of regulatory network structure from data is a very difficult problem. 
+Recently, Sauer and colleagues have developed a systematic, model-based approach for the identification of allosteric regulation \textit{in vivo} \cite{Link:2013aa}. 
+They tested the effects of many putative allosteric protein-metabolite interactions on the performance of a kinetic model of glycolysis against dynamic metabolomic and fluxomic measurements. 
+A method similar to this may be easily applied to cell-free systems in order to identify relevant \textit{in vitro} allosteric interactions. 
+Because omics measurements of cell-free environments are easy to obtain, identification of large-scale allosteric control structures may be possible. 
+Also, there are many different approaches from the reverse engineering of gene regulatory networks that perhaps could be adopted to this problem, however this remains an open question.
 
@@
 \end{equation}where $0\leq f_{ij}\left(\mathcal{Z}\right)\leq 1$ was a regulatory transfer function that calculated the influence of metabolite $i$ on the activity of enzyme $j$. 
-Conversely, if enzyme activity was inhibited by $m$ metabolites, we modeled this inhibition as:
+Conversely, if enzyme activity was inhibited by a $m$ metabolites, we modeling this inhibition as:
 \begin{equation}
@@
 \end{equation}where $\mathcal{Z}_{j}$ denotes the abundance of the $j$ factor (e.g., metabolite abundance), and $\kappa_{ij}$ and $\eta$ are control parameters. 
-$\kappa_{ij}$ was the species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).
-Applying the general framework to the reduced coagulation network resulted in five ordinary differential equations:
-\begin{eqnarray}
-	\frac{dx_{1}}{dt} & =& -\left(r_{init}v_{init} + r_{amp}v_{amp}\right)\\
-	\frac{dx_{2}}{dt} &=& r_{amp}v_{amp} + r_{init}v_{init} - r_{inh,ATIII}v_{inh,ATIII}\\
-	\frac{dx_{3}}{dt} &=& -r_{apc}v_{apc} \\
-	\frac{dx_{4}}{dt} &=& r_{apc}v_{apc} \\
-	\frac{dx_{5}}{dt} & = & -r_{inh,ATIII}v_{inh,ATIII}
-\end{eqnarray}where $\mathbf{x} = \left(fII,FIIa,PC,APC,ATIII\right)^{T}$. 
-The terms $r_{*}v_{*}$ in the balance equations denote corrected kinetic expressions for initiation, amplification and inhibition processes. 
-The rate of initiation $\bar{r}_{init}$ was modeled as:
-\begin{equation}
-	\bar{r}_{init} = k_{init}\left(trigger\right)\frac{x_{1}}{K_{init,fII} + x_{1}}
-\end{equation}where $k_{init},~K_{init,fII}$ are the rate and saturation constants governing initiation, respectively. 
-The rate of initiation was modified by $v_{init}$, the control parameter governing initiation. 
-Initiation was sensitive to the level of trigger (activator) and TFPI (inhibitor):
-\begin{equation}
-	v_{init} = \min\left(f^{-}_{init}\left(TFPI\right),f^{+}_{init}\left(trigger\right)\right)
-\end{equation}where the transfer functions $f$ took the form of Eqn \eqref{eqn:control-factor}.
-The rate of thrombin amplification was given by:
-\begin{equation}
-	\bar{r}_{amp} = k_{amp}\left(x_{2}\right)\frac{x_{1}}{K_{amp,fII}+x_{1}}
-\end{equation}where $k_{amp},~K_{amp,fII}$ denote the rate and saturation constants governing amplification, respectively. 
-The amplification control term, which modified amplification rate, was modeled as a combination of multiple inhibition terms and one activation term:
-\begin{equation}
-	v_{amp} = \min\left(f^{-}_{amp}\left(TFPI\right),f^{-}_{amp}\left(x_{4}\right),f^{+}_{amp}\left(\mathcal{Z}_{amp}\right)\right)
-\end{equation}where $\mathcal{Z}_{amp} = fV\times fX\times fVIII\times fIX$. Although $f^{+}_{amp}\left(\mathcal{Z}_{amp}\right)$ is an activating term, 
-we included it in the min integration rule; the factors in $\mathcal{Z}_{amp}$ were essential for amplification (if any of these factors was missing
-the amplification reaction would not occur). Thus, the factors in $\mathcal{Z}_{amp}$ were required components, a classification
-that we implemented by the min selection rule.
-The rate activated protein C formation was given by:
-\begin{equation}
-	\bar{r}_{apc} = k_{APC,formation}\left(TM\right)\frac{x_{3}}{K_{formation,PC} + x_{3}}
-\end{equation}where $k_{APC,formation}$ and $K_{formation,PC}$ denote the rate and saturation constants governing activated protein C formation, respectively and $TM$ denotes
-the thrombomodulin abundance. We modeled the control term which governed APC formation as a single thrombin-dependent activation term:
-\begin{equation}
-	v_{apc} = \max\left(f_{apc}^{+}\left(x_{2}\right)\right)
-\end{equation}
-Lastly, we included direct irreversible inhibition of FIIa by ATIII:
-\begin{equation}
-	\bar{r}_{inh,ATIII} = k_{ATIII,inhibition}\left(x_{5}x_{2}^{\gamma}\right)
-\end{equation}where $\gamma$ was estimated to be $\gamma = 1.26$. For ATIII inhibition of FIIa, the control variables $v_{inh,ATIII}$ was taken to be unity.
-The model equations were encoded using the Python programming language and solved using the ODEINT routine of the SciPy module \citep{SCIPY}.
-The model files can be downloaded from http://www.varnerlab.org.
+The $\kappa_{ij}$ parameter represents a species gain parameter, while $\eta$ is a cooperativity parameter (similar to a Hill coefficient). 
+In the case $\eta>1$, the allosteric interaction displays positive cooperativity.
+For $\eta<1$, the interaction is negatively cooperative.
+Finally, if $\eta=1$, the interaction displays no cooperativity.
+The effect of different values of $\eta$ on reaction rate can be seen in Figure \ref{fig-kinetics-simulations}.
+The model equations were encoded using the Octave programming language and solved using the LSODE routine in Octave (v 3.8.1; www.octave.org).
 
-\subsection*{Estimation of model parameters from experimental data.}
-Model parameters were estimated by minimizing the difference between simulations and experimental thrombin measurements (squared residual):
+\subsection*{Estimation of model parameters and structures from synthetic experimental data.}
+Model parameters were estimated by minimizing the difference between simulations and synthetic experimental data (squared residual):
 \begin{equation}\label{eqn:objective-function}
@@
 \end{equation}where $\hat{x}_{j}\left(\tau\right)$ denotes the measured value of species $j$ at time $\tau$, $x_{j}\left(\tau,\mathbf{k}\right)$ denotes the simulated 
-value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$. The outer summation is with respect to
-time, while the inner summation is with respect to state. We minimized the model residual using Particle swarm optimization (PSO) \citep{PSO}.
+value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$. The outer summation is respect to
+time, while the inner summation is with respect to state. We approximated a realistic model identification scenario, assuming noisy experimental data, 
+limited sampling resolution (approximately 20 minutes per sample) and a limited number of measurable metabolites.
+
+We minimized the model residual using particle swarm optimization (PSO) \citep{PSO}.
 PSO uses a \textit{swarming} metaheuristic to explore parameter spaces. 
@@
 	\mathbf{k}_{i} &=& \mathbf{k}_{i} + \mathbf{\Delta}_{i}
-\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes the local best solution found by particle $i$, and
+\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes local best solution found by particle $i$, and
 $\mathcal{G}$ denotes the best solution found over the entire population of particles.
@@
 parameters ($\mathcal{K}\times{1}$). In thus study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$. The quality of parameter
-estimates was measured using goodness of fit (model residual). The particle swarm optimization routine was implemented in the Python programming language. 
-All plots were made using the Matplotlib module of Python \citep{Matplotlib}.
-%The optimization and plotting code can be downloaded from http://www.varnerlab.org.
+estimates was measured using two criteria, goodness of fit (model residual) and angle between the estimated parameter vector $\mathbf{k}_{j}$ and the true parameter set $\mathbf{k}^{*}$:
+\begin{equation}
+	\alpha_{j} = cos^{-1}\left(\frac{\mathbf{k}_{j}\cdot{\mathbf{k^{*}}}}{\norm{\mathbf{k}_{j}}\norm{\mathbf{k^{*}}}}\right)
+\end{equation}If the candidate parameter set $\mathbf{k}_{j}$ were perfect, the residual between the model and synthetic data 
+and the angle between $\mathbf{k}_{j}$ and the true parameter set $\mathbf{k}^{*}$ would be equal to zero.
 
-\subsection*{Global sensitivity analysis of model performance.}
-We conducted a global sensitivity analysis, using the variance-based method of Sobol, to estimate which parameters controlled the performance of the reduced order model \citep{SOBOL_METHOD}. 
-We computed the total sensitivity index of each parameter relative to two performance objectives, the peak thrombin time and the area under the thrombin curve (thrombin exposure).
-We established the sampling bounds for each parameter from the minimum and maximum value of that parameter in the parameter set ensemble. 
-We used the sampling method of Saltelli \textit{et al.} \citep{Saltelli:2010} to compute a family of $N\left(2d+2\right)$ parameter sets which obeyed our parameter ranges, 
-where $N$ was the number of trials, and $d$ was the number of parameters in the model. In our case, $N$ = 10,000 and $d$ = 22, so the total sensitivity indices were computed from
-460,000 model evaluations. The variance-based sensitivity analysis was conducted using the SALib module encoded in the Python programming language \citep{SALIB}.
+We modified our PSO implementation to simultaneously search over kinetic parameters and putative model control structures.
+In the combined case, each particle potentially carried a different model realization in addition to a different kinetic parameter vector. 
+We kept the update rules the same (along with the update parameters). 
+Thus, each particle competed on the basis of goodness of fit, which allowed different model structures to contribute to the overall behavior of the swarm. 
+We considered five possible model structures (A through E), where network A was the correct formulation (used to generate the synthetic data). 
+We considered a population of 100 particles, where each particle in the swarm was assigned a model structure, and a random parameter vector.
+The PSO algorithm, model equations, and the objective function were encoded and solved in the Octave programming language (v 3.8.1; www.octave.org).
 
 \section*{Acknowledgements}
-This study was supported by an award from the Army Research Office (ARO \#59155-LS).
+This study was supported by an award from the National Science Foundation (MCB \#1411715) 
+and the Army Research Office (ARO \#59155-LS).
 
@@
 
-\bibliography{References_v1}
+\bibliography{References_v2}
 
@@
 \centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-1-Network.pdf}
-\caption{Schematic of the connectivity of the reduced order coagulation model. A trigger compound, e.g., TF/FVIIa initiates thrombin production (FIIa) from prothrombin (fII).
-Once activated, thrombin catalyzes its own activation (amplification step), as well as its own inhibition via the conversion of protein C to activated protein C (APC).
-APC and tissue factor pathway inhibitor (TFPI) inhibit initiation and amplification, while antithrombin III (ATIII) directly inhibits thrombin. 
-All inhibition steps and trigger-induced initiation were modeled using a rule-based approach. 
-Likewise, the dependence of amplification on other coagulation factors was also modeled using a rule-based approach. 
-The abundance of the highlighted species (in the dashed boxes) was governed by an ordinary differential equation. 
-All other species were assumed to be constant.}\label{fig-network}
+\includegraphics[width=1.0\textwidth]{./figs/Figure-1-Networks.pdf}
+\caption{Proof-of-concept cell-free metabolic networks considered in this study. Substrate $S$ is converted to products P$_{1}$ and P$_{2}$ through a series of chemical conversions
+catalyzed by enzyme(s) $E_{j}$. The activity of the pathway enzymes is subject to both positive and negative allosteric regulation. }\label{fig-networks}
 \end{figure}
@@
 \includegraphics[width=1.0\textwidth]{./figs/Figure-2-ControlSchematic.pdf}
-\caption{Schematic of rule-based effective control laws. 
-Traditional enzyme kinetic expressions, e.g., Michaelis–Menten or multiple saturation kinetics are multiplied by an enzyme activity control variable $0 \leq v_{j} \leq 1 $. 
+\caption{Schematic of rule-based allosteric enzyme activity control laws. 
+Traditional enzyme kinetic expressions, e.g., Michaelis–Menten or multiple saturation kinetics, are multiplied by an enzyme activity control variable $0 \leq v_{j} \leq 1 $. 
 Control variables are functions of many possible regulatory factors encoded by arbitrary functions of the form $0\leq f_{j}\left(\mathcal{Z}\right)\leq 1$.
-At each simulation time step, the $v_{j}$ variables are calculated by evaluating integration rules such as the max or min of the set of factors $f_{1},\hdots,f_{n}$ 
+At each simulation time step, the $v_{j}$ variables are calculated by evaluating integration rules such as the max or min of the set of factors $f_{1},\hdots$ 
 influencing the activity of enzyme $E_{j}$. }\label{fig-control-schematic}
@@
 \centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-3-DIFF_FII_NO_APC.pdf}
-\caption{Reduced order coagulation model training simulations.
-Reduced order coagulation model parameters were estimated using particle swarm optimization (PSO) with and without the protein C pathway as a function of prothrombin. 
-Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data. 
-The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation).
-(A,B,C) training results for 150\%, 100\% and 50\% of physiological prothrombin levels in the absence of the protein C pathway.
-Thrombin generation was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
-All factors and control proteins were at their physiological concentration unless others denoted. 
-The experimental training data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}.}\label{fig-diff_fII_NO_APC}
+\includegraphics[width=1.0\textwidth]{./figs/Figure-3-EnzymeKinetics.pdf}
+\caption{Kinetics of simple transformations in the presence of activation and inhibition. 
+\textbf{A}:The conversion of substrate $S$ to product $P$ by enzyme $E$ was activated by $S$. 
+For a fixed control gain parameter $\kappa_{control}$, the reaction rate approached a step for increasing cooperativity control parameter $\eta$. 
+\textbf{B}:The conversion of substrate $S$ to product $P$ by enzyme $E$ with inhibitor $I$. 
+For a fixed control gain parameter $\kappa_{control}$, the reaction rate approximated non-competitive inhibition for increasing cooperativity control parameter $\eta$. 
+}\label{fig-kinetics-simulations}
 \end{figure}
@@
 \centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-4-DIFF_FII_APC.pdf}
-\caption{Reduced order coagulation model training simulations.
-Reduced order coagulation model parameters were estimated using particle swarm optimization (PSO) with and without the protein C pathway as a function of prothrombin. 
-Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data.
-The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation).
-(A,B,C) training results for 150\%, 100\% and 50\% of physiological prothrombin levels in the presence of the protein C pathway. 
-Only APC pathway parameters were allowed to vary in the simulations on the right.
-Thrombin generation was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
-All factors and control proteins were at their physiological concentration unless others denoted.
-The experimental training data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}.}\label{fig-diff_fII_APC}
+\includegraphics[width=0.50\textwidth]{./figs/Figure-4-OnOffSimulations.pdf}
+\caption{ON/OFF control simulations for network A and network B for an ensemble of kinetic parameter sets versus time (N = 100). 
+For each case, N = 100 simulations were conducted using kinetic and initial conditions generated randomly from a hypothetical true parameter set. 
+The gray area represents $\pm$ one standard deviation surrounding the mean. 
+Control parameters were fixed during the ensemble calculations.
+\textbf{A}: End product P$_{1}$ abundance versus time for Network A. 
+The abundance of P$_{1}$ decreased with end product inhibition of $E_{1}$ activity (Control-ON) versus the no inhibition case (Control-OFF). 
+\textbf{B}: End product P$_{2}$ abundance versus time for Network B. Inhibition of branch point $E_{6}$ by end product P$_{1}$ decreased P$_{2}$ abundance (Control-ON) versus the
+no inhibition case (Control-OFF).
+\textbf{C}: End product P$_{1}$ abundance versus time for Network A. 
+Inhibition of branch point $E_{6}$ by end product P$_{1}$ decreased P$_{1}$ abundance (Control-ON) versus the no inhibition case (Control-OFF).
+}\label{fig-onoff-simulations}
 \end{figure}
@@
 \centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-5-FII_ATIII.pdf}
-\caption{Reduced order coagulation model predictions versus experimental data for normal coagulation. 
-The reduced order coagulation model parameter estimates were tested against data not used during model training.
-Simulations of different levels of prothrombin and ATIII were compared with experimental data in the absence of the protein C pathway. 
-Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data.
-The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation). 
-(A,B,C,D) prediction results for (FII,ATIII): (50\%,150\%), (100\%, 100\%), (125\%, 75\%) and (150\%, 50\%) of physiological prothrombin and ATIII levels in the absence of the protein C pathway. 
-Thrombin generation was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
-All factors and control proteins were at their physiological concentration unless others denoted.
-The experimental validation data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}}\label{fig-fII_ATIII}
+\includegraphics[width=1.0\textwidth]{./figs/Figure-5-ParameterFit.pdf}
+\caption{Parameter estimation from synthetic data for the same and mismatched allosteric control logic using particle swarm optimization (PSO). 
+Synthetic experimental data was generated from a hypothetical parameter set using Network A, 
+where substrate $S$, end product P$_{1}$ and intermediate M$_5$ were sampled approximately every 20 minutes. 
+For cases $\textbf{A,B}$ 20 particles were initialized with randomized parameters and allowed to search for 300 iterations. 
+\textbf{A,B}: PSO estimated an ensemble of parameters sets (N = 20) consistent with the synthetic experimental data assuming the correct
+enzymatic and control connectivity starting from randomized initial parameters.
+\textbf{C,D}: In the presence of control mismatch (Network B control policy simulated with Network A kinetic parameters) 
+the ensemble of models did not describe the synthetic data. 
+}\label{fig-parameter-fit}
 \end{figure}
@@
 \centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-6-Hemophilia.pdf}
-\caption{Reduced order coagulation model predictions versus experimental data with and without FVIII and FIX.
-The reduced order coagulation model parameter estimates were tested against data not used during model training. 
-Simulations of normal thrombin formation with ATIII and the protein C pathway were compared with thrombin formation in the absence of fVIII and fIX.
-Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data.
-The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation). 
-(A,B) prediction results for normal thrombin generation and thrombin generation in hemophilia. 
-All factors and control proteins were at their physiological concentration unless others noted.
-The experimental validation data was reproduced from the study of Allen \textit{et al.} \citep{ALLEN2006}.}\label{fig-hemophilia}
+\includegraphics[width=1.0\textwidth]{./figs/Figure-6-AlternativeNetworks.pdf}
+\caption{Schematic of the alternative allosteric control programs used in the structural particle swarm computation. 
+Each network had the same enzymatic connectivity, initial conditions and kinetic parameters, 
+but alternative feedback control structures for the first enzyme in the pathway.}\label{fig-alternative-networks}
 \end{figure}
@@
 \centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-7-rFVIIa-Trigger.pdf}
-\caption{Reduced order coagulation model predictions of rFVIIa administration. 
-A: Simulations of thrombin formation in the presence of ATIII and the protein C pathway were conducted for a range of trigger values (1x - 200x nominal) in the absence of fVIII and fIX.
-B: Comparison of thrombin generation for normal versus hemophilia for 10x nominal trigger.
-Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles.
-The shaded region denotes the 99\% confidence estimate of the mean thrombin value (uncertainty in the model simulation). 
-All factors and control proteins were at their physiological concentration unless others noted.}\label{fig-rFVII}
+\includegraphics[width=0.5\textwidth]{./figs/Figure-7-ControlSearch.pdf}
+\caption{Combined control and kinetic parameter search using modified particle swarm optimization (PSO). 
+A population of N = 100 particles was initialized with randomized kinetic parameters and one of five possible control configurations (Network A - E).
+Simulation error was minimized for a synthetic data set 
+($S$, end product P$_{1}$ and intermediate M$_5$ sampled approximately every 20 min) generated using Network A.
+\textbf{A}: Simulation error versus parameter set angle for N = 100 particles biased toward the correct regulatory program (A,B,C,D,E) = (40\%, 10\%, 20\%, 20\% and 10\%).
+\textbf{B}: Simulation error versus parameter set angle for N = 100 uniformly distributed particles (A,B,C,D,E) = (20\%, 20\%, 20\%, 20\% and 20\%).
+\textbf{C}: Simulation error versus parameter set angle for N = 100 negatively biased particles (A,B,C,D,E) = (10\%, 40\%, 10\%, 20\% and 20\%).
+Network A (the correct structure) was preferentially identified for positively and uniform biased particle distributions, but misidentified in the presence of
+a large incorrect bias. 
+}\label{fig-control-search}
 \end{figure}
@@
 \centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-8-Flux-rFVIIa-Trigger.pdf}
-\caption{Reaction flux distribution as a function of time for thrombin generation under normal (left), hemophilia (center) and rFVIIa treated hemophilia (right).
-Reaction flux was calculated for each particle at $T = 0,4,6,8,10,12,14$ min after the initiation of coagulation. 
-Reaction fluxes were calculated for each particle in the parameter ensemble (N = 20).
-Blue colors denote low flux values while red colors denote high flux values.}\label{fig-flux-analysis}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-9-Sensitivity.pdf}
-\caption{Global sensitivity analysis of the reduced order coagulation model with respect to the model parameters.
-A: Sensitivity analysis of the thrombin peak time for different prothrombin levels (150\%,100\% and 50\% of the physiological value) as a function of activated protein C. 
-B: Sensitivity analysis of the thrombin exposure for different prothrombin levels (150\%,100\% and 50\% of the physiological value) as a function of activated protein C. 
-Points denote the mean total sensitivity value, while the area around each point denotes the uncertainty in the sensitivity value. 
-The gray dashed line denotes the 45$^{\circ}$ degree diagonal, if sensitivity values are equal for different conditions they will lie on the diagonal.  
-Sensitivity values significantly above or below the diagonal indicate differentially important model parameters. 
-The radius of the shaded region around each total sensitivity value was the maximum uncertainty in that value estimated by the Sobol method. 
-}\label{fig-sensitivity}
+\includegraphics[width=1.0\textwidth,height=0.6\textheight]{./figs/Figure-8-Flux.pdf}
+\caption{Metabolic flux and control variables as a function of network type and particle index at t = 100 min.
+The control variables governing $E_{1},E_{3}$ and $E_{6}$ activity and the scaled metabolic flux and were calculated for the positively, 
+uniformly and negatively biased particle swarms (N = 100). The particles from each swarm were sorted based upon simulation error (low to high error). 
+\textbf{A}: Model performance for the positively biased particle swarm as a function of particle index.
+\textbf{B}: Model performance for the uniformly biased particle swarm as a function of particle index.
+\textbf{C}: Model performance for the negatively biased particle swarm as a function of particle index. 
+Models with significant control mismatch showed distinct control and flux patterns versus those models with the correct or closely related control policies.
+In particular, models with the correct control policy showed stronger inhibition of $E_{1}$ activity, leading to decreased flux from S$\rightarrow$P$_{1}$.
+Conversely, models with significant mismatch had increased $E_{1}$ activity, leading to an altered flux distribution. 
+This is especially apparent in the negatively biased particle swarm. 
+}\label{fig-flux-pattern}
 \end{figure}

### 2014-11-24 Paper_v3.tex
reviewer 3 updates

--- 2014-11-11 Paper_v2.tex
@@
 These trends were manifested in different flux profiles, where the negatively biased population appeared more uniform across the population compared with the other swarms, and had higher $E_{1}$ specific activity.
-Interestingly, the behavior of network A and network C highlighted an artifact of our integration rule; both a positive or negative feedback connection from P$_2$ to $E_{1}$ were ignored because the P$_{1}$ inhibition of $E_{1}$ was dominate.
-Thus, while theoretically distinct, network A and network C appeared operationally to the PSO algorithm to be that same network.
+Interestingly, the behavior of network A and network C highlighted an artifact of our integration rule; both a positive or negative feedback connection from P$_2$ to $E_{1}$ were ignored because the P$_{1}$ inhibition of $E_{1}$ dominated.
+Thus, while theoretically distinct, network A and network C appeared operationally to the PSO algorithm to be the same network.
 On the other hand, networks B, D and E showed distinct behavior that was not consistent with the true network. 
@@
 If $\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$. 
-Conversely, if $\sigma_{ij}>0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates metabolite $i$ is not connected with reaction $j$. 
+Conversely, if $\sigma_{ij}<0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates metabolite $i$ is not connected with reaction $j$. 
 Lastly, $\lambda_{i}$ denotes the scaled enzyme degradation constant. 
@@
 	\bar{r}_{j} = k_{j}^{max}\epsilon_{i}\left(\prod_{s\in{m_{j}^{-}}}\frac{x_{s}}{K_{js} + x_{s}}\right)
-\end{equation}where $k_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the scaled enzyme activity which catalyzes reaction $j$, and
+\end{equation}
+where $k_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the scaled enzyme activity which catalyzes reaction $j$, and
 $K_{js}$ denotes the saturation constant for species $s$ in reaction $j$. 
@@
 	\mathbf{k}_{i} &=& \mathbf{k}_{i} + \mathbf{\Delta}_{i}
-\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes local best solution found by particle $i$, and
+\end{eqnarray}
+where $\mathbf{\Delta_i}$ denotes the perturbation to the vector of parameters $\mathbf{k_i}$ for particle $i$. 
+$\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes the best local solution found by particle $i$, and
 $\mathcal{G}$ denotes the best solution found over the entire population of particles.
 The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
-parameters ($\mathcal{K}\times{1}$). In thus study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$. The quality of parameter
+parameters ($\mathcal{K}\times{1}$). In this study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$. The quality of parameter
 estimates was measured using two criteria, goodness of fit (model residual) and angle between the estimated parameter vector $\mathbf{k}_{j}$ and the true parameter set $\mathbf{k}^{*}$:
@@
 
-\bibliography{References_v2}
+\bibliography{References_v3}
 
@@
 \includegraphics[width=0.50\textwidth]{./figs/Figure-4-OnOffSimulations.pdf}
-\caption{ON/OFF control simulations for network A and network B for an ensemble of kinetic parameter sets versus time (N = 100). 
-For each case, N = 100 simulations were conducted using kinetic and initial conditions generated randomly from a hypothetical true parameter set. 
+\caption{ON/OFF control simulations for Network A and Network B for an ensemble of 100 kinetic parameter sets versus time. 
+For each case, simulations were conducted using kinetic and initial conditions generated randomly from a hypothetical true parameter set. 
 The gray area represents $\pm$ one standard deviation surrounding the mean. 
@@
 For cases $\textbf{A,B}$ 20 particles were initialized with randomized parameters and allowed to search for 300 iterations. 
-\textbf{A,B}: PSO estimated an ensemble of parameters sets (N = 20) consistent with the synthetic experimental data assuming the correct
+\textbf{A,B}: PSO estimated an ensemble of 20 parameters sets consistent with the synthetic experimental data assuming the correct
 enzymatic and control connectivity starting from randomized initial parameters.
@@
 \caption{Combined control and kinetic parameter search using modified particle swarm optimization (PSO). 
-A population of N = 100 particles was initialized with randomized kinetic parameters and one of five possible control configurations (Network A - E).
+A population of 100 particles was initialized with randomized kinetic parameters and one of five possible control configurations (Network A - E).
 Simulation error was minimized for a synthetic data set 
 ($S$, end product P$_{1}$ and intermediate M$_5$ sampled approximately every 20 min) generated using Network A.
-\textbf{A}: Simulation error versus parameter set angle for N = 100 particles biased toward the correct regulatory program (A,B,C,D,E) = (40\%, 10\%, 20\%, 20\% and 10\%).
-\textbf{B}: Simulation error versus parameter set angle for N = 100 uniformly distributed particles (A,B,C,D,E) = (20\%, 20\%, 20\%, 20\% and 20\%).
-\textbf{C}: Simulation error versus parameter set angle for N = 100 negatively biased particles (A,B,C,D,E) = (10\%, 40\%, 10\%, 20\% and 20\%).
+\textbf{A}: Simulation error versus parameter set angle for 100 particles biased toward the correct regulatory program (A,B,C,D,E) = (40\%, 10\%, 20\%, 20\% and 10\%).
+\textbf{B}: Simulation error versus parameter set angle for 100 uniformly distributed particles (A,B,C,D,E) = (20\%, 20\%, 20\%, 20\% and 20\%).
+\textbf{C}: Simulation error versus parameter set angle for 100 negatively biased particles (A,B,C,D,E) = (10\%, 40\%, 10\%, 20\% and 20\%).
 Network A (the correct structure) was preferentially identified for positively and uniform biased particle distributions, but misidentified in the presence of

### 2014-12-01 revision/Paper_v3.tex
JVs changes for reviewers
JVs changes for revised manuscript

--- 2014-11-24 Paper_v3.tex
@@
 \ref{fig-kinetics-simulations}A).
-For cooperativity parameters less than unity, increased substrate abundance \textit{decreased} the reaction rate. 
+For cooperativity parameters less than unity, increased substrate abundance \textit{decreased} the maximum reaction rate. 
 This was consistent with the idea that substrate binding \textit{decreased} at regulatory sites, which negatively impacted substrate binding at the active site. 
@@
 Of course, these issues are not independent; any description of enzyme activity regulation will be a function of system state, which in turn depends upon the kinetic parameters. 
-For cell-free systems, regulated gene expression has been removed, however, enzyme activity regulation is still operational. 
+For cell free systems, regulated gene expression has been removed, however, enzyme activity regulation is still operational. 
 We explored this linkage by estimating model parameters from synthetic data using both network structures. 
-We generated noise corrupted synthetic measurements of the substrate S, intermediate M$_{5}$ and end product P$_1$ approximately every 20 min using
-network A.
+We generated synthetic measurements of the substrate S, intermediate M$_{5}$ and end product P$_1$ approximately every 20 min using
+network A. This data set is similar to published cell free studies both in terms of network coverage, and sampling frequency \citep{Jewett:2008aa}. 
 We then generated an ensemble of model parameter estimates by minimizing the difference between model simulations and the synthetic data using particle swarm optimization (PSO), starting from random initial parameter guesses.
@@
 Also, there are many different approaches from the reverse engineering of gene regulatory networks that perhaps could be adopted to this problem, however this remains an open question.
+Lastly, the choice of max/min integration rules or the particular form of the transfer functions could be generalized to include other rule types and functions. 
+Theoretically, an integration rule is a function whose domain is a set of transfer function inputs, and whose range is $v\in[0,1]$.
+Thus, integration rules other than max/min could be used, such as the mean or the product, assuming the range of the transfer functions is always $f\in[0,1]$.
+Alternative integration rules such as the mean might have different properties which could influence model identification or performance. 
+For example, a mean integration rule would be differentiable, which allows derivative-based optimization approaches to be used. 
+The particular form of the transfer function could also be explored. We choose a Hill-like function because of its 
+prominence in the systems and synthetic biology community. However, the only mathematical requirement for a transfer function is that it map a non-negative continuous or categorical variable into the 
+range $f\in[0,1]$. Thus, many types of transfer functions are possible.
 
@@
 The model equations were encoded using the Octave programming language and solved using the LSODE routine in Octave (v 3.8.1; www.octave.org). 
+In some cases, metabolic fluxes (or other quantities) were scaled according to:
+\begin{equation}
+	\hat{r}_{j}\left(t = \tau\right) = \left(\frac{r_{j} - \min\mathbf{r}}{\max\mathbf{r} - \min\mathbf{r}}\right)\Bigr|_{t = \tau}
+\end{equation}where $0\leq\hat{r}_{j}\left(t = \tau\right)\leq{1}$ denotes the scaled value for flux j evaluated at time $\tau$. 
+We have used this scaling in a variety of other contexts \citep{Song:2010ab,Lequieu:2011aa}.
 
@@
 time, while the inner summation is with respect to state. We approximated a realistic model identification scenario, assuming noisy experimental data, 
-limited sampling resolution (approximately 20 minutes per sample) and a limited number of measurable metabolites.
+limited sampling resolution (approximately 20 minutes per sample) and a limited number of measurable metabolites. We assumed a constant coefficient of variation of 10\% for the synthetic data set.
 
@@
 \textbf{C,D}: In the presence of control mismatch (Network B control policy simulated with Network A kinetic parameters) 
-the ensemble of models did not describe the synthetic data. 
+the ensemble of models did not describe the synthetic data. The synthetic data plotted here was unperturbed by noise. 
+However, we assumed a coefficient of variation of 10\% for the synthetic data during parameter estimation.
 }\label{fig-parameter-fit}
@@
 \caption{Metabolic flux and control variables as a function of network type and particle index at t = 100 min.
-The control variables governing $E_{1},E_{3}$ and $E_{6}$ activity and the scaled metabolic flux and were calculated for the positively, 
-uniformly and negatively biased particle swarms (N = 100). The particles from each swarm were sorted based upon simulation error (low to high error). 
+The control variables governing $E_{1},E_{3}$ and $E_{6}$ activity and the scaled metabolic flux were calculated for the positively (top), 
+uniformly (middle) and negatively (bottom) biased particle swarms (N = 100). 
+Blue denotes a low value, while red denotes a high value for the respective quantity being plotted.
+The particles from each swarm were sorted based upon simulation error (low to high error). 
 \textbf{A}: Model performance for the positively biased particle swarm as a function of particle index.

### 2014-12-15 processes-69103_review_response/processes-69103.tex
Folder with the new tex file,rebuttal,original pdf and new pdf marked with changes

--- 2014-10-30 submission_v1-10-29-14/Paper_v2.tex
@@
-\documentclass[12pt]{article}
+\documentclass[processes,article,received,moreauthors,pdftex,12pt,a4paper]{mdpi} 
 % Load packages
-\usepackage{url}  % Formatting web addresses  
-\usepackage{ifthen}  % Conditional 
-\usepackage{multicol}   %Columns
-\usepackage[utf8]{inputenc} %unicode support
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{epsfig}
-\usepackage{epstopdf}
+\usepackage[mediumspace,mediumqspace,Grey,squaren]{SIunits}
+\usepackage{soul}
+\usepackage{amssymb,amsmath}
 \usepackage{graphicx}
-\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
-\usepackage{setspace}
-%\usepackage{longtable}
-\usepackage{colortbl}
-%\usepackage{palatino,lettrine}
-%\usepackage{times}
-%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
-%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
-\usepackage[wide]{sidecap}
-%\usepackage[authoryear,round,comma,sort&compress]{natbib}
-\usepackage[square,sort&compress,comma,numbers]{natbib}
-%\usepackage[authoryear,round]{natbib}
-\usepackage{supertabular}
-\usepackage{simplemargins}
-\usepackage{comment}
-\usepackage{lineno}
-
-\urlstyle{rm}
+\usepackage{subfigure}
+\makeatletter
+\renewcommand{\@thesubfigure}{\normalsize(\textbf{\alph{subfigure}})}
+\makeatother
+\usepackage{array, dcolumn, multirow}
+\usepackage{hyphenat}
+\hyphenpenalty=2000 \exhyphenpenalty=2000 \sloppy
+\usepackage{enumitem,caption,booktabs}
+\setitemize{parsep=0pt,itemsep=0pt,leftmargin=.4in}
+\setenumerate{parsep=0pt,itemsep=0pt,leftmargin=.4in}
+
+  % Formatting web addresses  
+%\usepackage{ifthen}  % Conditional 
+%\usepackage{multicol}   %Columns
+%\usepackage[utf8]{inputenc} %unicode support
+%\usepackage{amsmath}
+%\usepackage{amssymb}
+%\usepackage{epsfig}
+%\usepackage{epstopdf}
+%\usepackage{graphicx}
+%\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
+%\usepackage{setspace}
+%%\usepackage{longtable}
+%\usepackage{colortbl}
+%%\usepackage{palatino,lettrine}
+%%\usepackage{times}
+%%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
+%%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
+%\usepackage[wide]{sidecap}
+%%\usepackage[authoryear,round,comma,sort&compress]{natbib}
+%\usepackage[square,sort&compress,comma,numbers]{natbib}
+%%\usepackage[authoryear,round]{natbib}
+%\usepackage{supertabular}
+%\usepackage{simplemargins}
+%\usepackage{comment}
+%\usepackage{lineno}
+%
+%\urlstyle{rm}
 
@@
 %\linespread{1.75}
-\doublespace
 
 %\usepackage{geometry}
-\usepackage{fullpage}
+%\usepackage{fullpage}
 
 %\bibliographystyle{plain}
-\bibliographystyle{plos2009}
-
-\makeatletter
-\renewcommand\subsection{\@startsection
-	{subsection}{2}{0mm}
-	{-0.05in}
-	{-0.5\baselineskip}
-	{\normalfont\normalsize\bfseries}}
-\renewcommand\subsubsection{\@startsection
-	{subsubsection}{2}{0mm}
-	{-0.05in}
-	{-0.5\baselineskip}
-	{\normalfont\normalsize\itshape}}
-\renewcommand\section{\@startsection
-	{subsection}{2}{0mm}
-	{-0.2in}
-	{0.05\baselineskip}
-	{\normalfont\large\bfseries}}	
-\renewcommand\paragraph{\@startsection
-	{paragraph}{2}{0mm}
-	{-0.05in}
-	{-0.5\baselineskip}
-	{\normalfont\normalsize\itshape}}
-\makeatother
+%\bibliographystyle{plos2009}
+%
+%\makeatletter
+%\renewcommand\subsection{\@startsection
+%	{subsection}{2}{0mm}
+%	{-0.05in}
+%	{-0.5\baselineskip}
+%	{\normalfont\normalsize\bfseries}}
+%\renewcommand\subsubsection{\@startsection
+%	{subsubsection}{2}{0mm}
+%	{-0.05in}
+%	{-0.5\baselineskip}
+%	{\normalfont\normalsize\itshape}}
+%\renewcommand\section{\@startsection
+%	{subsection}{2}{0mm}
+%	{-0.2in}
+%	{0.05\baselineskip}
+%	{\normalfont\large\bfseries}}	
+%\renewcommand\paragraph{\@startsection
+%	{paragraph}{2}{0mm}
+%	{-0.05in}
+%	{-0.5\baselineskip}
+%	{\normalfont\normalsize\itshape}}
+%\makeatother
 
@@
 % Single space'd bib -
-\setlength\bibsep{0pt}
-
-\renewcommand{\rmdefault}{phv}\renewcommand{\sfdefault}{phv}
-\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
-
-% Change the number format in the ref list -
-\renewcommand{\bibnumfmt}[1]{#1.}
-
-% Change Figure to Fig.
-\renewcommand{\figurename}{Fig.}
-
+%\setlength\bibsep{0pt}
+%
+%\renewcommand{\rmdefault}{phv}\renewcommand{\sfdefault}{phv}
+%\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
+%
+%% Change the number format in the ref list -
+%\renewcommand{\bibnumfmt}[1]{#1.}
+%
+%% Change Figure to Fig.
+%\renewcommand{\figurename}{Fig.}
+
+\setcounter{page}{1}
+\lastpage{x}
+\doinum{10.3390/------}
+\pubvolume{xx}
+\pubyear{2014}
+\externaleditor{\hl{External Editor:}}
+\history{Received: xx / Accepted: xx /\\ Published: xx}
 % Begin ...
-\begin{document}
-\begin{titlepage}
-{\par\centering\textbf{\Large Dynamic Modeling of the Human Coagulation Cascade using Reduced Order Effective Kinetic Models}}
-\vspace{0.05in}
-{\par \centering \large{Adithya Sagar and Jeffrey D. Varner$^{*}$}}
-\vspace{0.10in}
-{\par \centering \large{School of Chemical and Biomolecular Engineering}}
-{\par \centering \large{Cornell University, Ithaca NY 14853}}
-\vspace{0.1in}
-{\par \centering \textbf{Running Title:}~Reduced order modeling of blood coagulation}
-\vspace{0.1in}
-{\par \centering \textbf{To be submitted:}~\emph{Processes}}
-\vspace{0.5in}
-{\par \centering $^{*}$Corresponding author:}
-{\par \centering Jeffrey D. Varner,}
-{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
-{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
-{\par \centering Email: jdv27@cornell.edu} 
-{\par \centering Phone: (607) 255 - 4258}
-{\par \centering Fax: (607) 255 - 9166}
-\end{titlepage}
-\date{}
-\thispagestyle{empty}
-\pagebreak
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Abstract}
-In this study, we present a novel modeling approach which combines ordinary differential equation (ODE) 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\Title{Dynamic Modeling of the Human Coagulation Cascade using Reduced Order Effective Kinetic Models}
+
+\Author{Adithya Sagar and Jeffrey D. Varner $^{*}$}
+
+\address[1]{%
+School of Chemical and Biomolecular Engineering, 244 Olin Hall,  Cornell University, Ithaca, \linebreak NY 14853, USA; \hl{E-Mail: }
+}
+
+\corres{E-Mail:  jdv27@cornell.edu; \linebreak Tel.: +607-255-4258;  Fax: +607-255-9166.}
+
+\abstract{In this study, we present a novel modeling approach which combines ordinary differential equation (ODE) 
 modeling with logical rules to simulate an archetype biochemical network, the human coagulation cascade.  
@@
 suggesting the proposed framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
-or potentially even cell free metabolic systems.
-
-\vspace{0.25in}
-{\noindent \textbf{Keywords:}~Blood coagulation, Mathematical modeling, Systems biology}
-
-\pagebreak
-
-\setcounter{page}{1}
-
-\linenumbers
+or potentially even cell free metabolic systems.}
+
+% Keywords: add 3 to 10 keywords
+\keyword{blood coagulation; mathematical modeling; systems biology}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{document}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\section*{Abstract}
+% 
+%
+%\vspace{0.25in}
+%{\noindent \textbf{Keywords:}~}
+%
+%\pagebreak
+%
+%\setcounter{page}{1}
+%
+%\linenumbers
 
@@
 %This result highlights the perhaps under-appreciated role that experimental design could play in generating the best training and validation data sets for model identification \citep{Apgar:2010fk}.
-
-\section*{Introduction}
+\vspace{-12pt}
+
+\section{Introduction}
+
 Developing mathematical models of biochemical networks is a significant facet of systems biology.
@@
 The extrinsic cascade is generally believed to be the main mechanism of thrombinogenesis in the blood \cite{MANN1990,ROBERTS1998,MANN1999}.
-Initially, thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa), which itself has been activated by TF/FVIIa \cite{Butenas:2002aa}. 
+Initially, thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa), which itself has been activated by Tissue Factor/factor VII (TF/FVIIa) \cite{Butenas:2002aa}. 
 Picomolar amounts of thrombin then activate the cofactors factors V and VIII (fV and fVIII) and platelets, 
@@
 
-\clearpage
-
-\section*{Results}
+\section{Results}
 
@@
 
-\subsection*{Formulation of reduced order coagulation models.}
+\subsection{Formulation of Reduced Order Coagulation Models}
+
 We developed a reduced order extrinsic coagulation model to test our hybrid modeling approach (Fig. \ref{fig-network}).
@@
 
-\subsection*{Identification of model parameters using particle swarm optimization.}
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure-1-Network.pdf}
+\caption{Schematic of the connectivity of the reduced order coagulation model. A trigger compound, e.g., TF/FVIIa initiates thrombin production (FIIa) from prothrombin (fII).
+Once activated, thrombin catalyzes its own activation (amplification step), as well as its own inhibition via the conversion of protein C to activated protein C (APC).
+APC and tissue factor pathway inhibitor (TFPI) inhibit initiation and amplification, while antithrombin III (ATIII) directly inhibits thrombin. 
+All inhibition steps and trigger-induced initiation were modeled using a rule-based approach. 
+Likewise, the dependence of amplification on other coagulation factors was also modeled using a rule-based approach. 
+The abundance of the highlighted species (in the dashed boxes) was governed by an ordinary differential equation. 
+All other species were assumed to be constant.}\label{fig-network}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure-2-ControlSchematic.pdf}
+\caption{Schematic of rule-based effective control laws. 
+Traditional enzyme kinetic expressions, e.g., Michaelis-Menten or multiple saturation kinetics are multiplied by an enzyme activity control variable $0 \leq v_{j} \leq 1 $. 
+Control variables are functions of many possible regulatory factors encoded by arbitrary transfer functions of the form $0\leq f_{j}\left(\mathcal{Z}\right)\leq 1$.
+At each simulation time step, the $v_{j}$ variables are calculated by evaluating integration rules such as the max or min of the set of transfer functions $f_{1},\hdots,f_{n}$ 
+influencing the activity of enzyme $E_{j}$. }\label{fig-control-schematic}
+\end{figure}
+
+\subsection{Identification of Model Parameters Using Particle Swarm Optimization}
+
 A critical challenge for any dynamic model is the estimation of kinetic parameters. 
@@
 We rotated the subproblems, starting with subproblem 1 in the first generation.
+
+The experimental training data for parameter estimation was reproduced from the experiments of Butenas and co-workers \citep{Butenas:1999aa}.
+In these experiments thrombin generation was initiated by FVIIa-TF using mean plasma concentrations of coagulation proteins and inhibitors.
+To prepare FVIIa-TF, TF (0.5 nmol/L) was relipidated into 400 $\mu$mol/L of phospholipid vesicles (PCPS) by incubation in 20 mmol/L HEPES, 150 mmol/L NaCl, and 2 mmol/L CaCl$_{2}$ pH 7.4 (HBS/Ca$^{2+}$) for 30 minutes at $37\,^{\circ}\mathrm{C}$.
+The relipidated TF was incubated with 10 pmol/L factor VIIa for 20 minutes to allow the formation of FVIIa-TF.
+Factors V, VIII and thrombomodulin (Tm) (when protein C activation is required) were added to FVIIa-TF complex.
+Thrombin generation was then initiated by adding equal volumes of this mixture with a mixture containing  prothrombin, factor IX and factor X, TFPI, AT-III and protein C (added when required), protein S (added when required) and factor XI (added when required).
+In the experimental training data sets that we used for parameter estimation 5 pmol/L FVIIa-TF was used along with 200 $\mu$mol/L of phospholipid vesicles (PCPS) to initiate thrombin generation.
+When protein C pathway was involved, protein C and protein S were at mean plasma concentrations and 0.1 nmol/L Tm was used.
+All the other coagulation proteins and inhibitors i.e.
+factors X, IX, V, and VIII, prothrombin, TFPI and AT-III were at their mean plasma concentration levels.
 
@@
 
-\subsection*{Validation of the reduced order coagulation model.}
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure-3-DIFF_FII_NO_APC.pdf}
+\caption{Reduced order coagulation model training simulations.
+Reduced order coagulation model parameters were estimated using particle swarm optimization (PSO) without the protein C pathway as a function of prothrombin. 
+Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data. 
+The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation).
+(A,B,C) depict training data and results for 150\%, 100\% and 50\% of physiological prothrombin levels in the absence of protein C pathway.
+The experimental training data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}.
+Thrombin generation in these experiments was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
+As depicted in (A,B,C) the prothrombin levels were at 150\%, 100\% and 50\% of their physiological concentration in the absence of protein C pathway. 
+All factors and control proteins in these experiments were at their physiological concentration unless otherwise denoted.}\label{fig-diff_fII_NO_APC} 
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure-4-DIFF_FII_APC.pdf}
+\caption{Reduced order coagulation model training simulations.
+Reduced order coagulation model parameters were estimated using particle swarm optimization (PSO) with the protein C pathway as a function of prothrombin.
+Only APC pathway parameters were allowed to vary in these simulations keeping the parameters estimated without protein C pathways constant. 
+Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data.
+The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation).
+(A,B,C) depict training data and results for 150\%, 100\% and 50\% of physiological prothrombin levels in the presence of the protein C pathway. 
+The experimental training data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}.
+Thrombin generation in these experiments was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
+As depicted in (A,B,C) the prothrombin levels were at 150\%, 100\% and 50\% of their physiological concentration in the presence of protein C pathway. 
+All factors and control proteins in these experiments were at their physiological concentration unless otherwise denoted.}\label{fig-diff_fII_APC}
+\end{figure}
+
+\subsection{Validation of the Reduced Order Coagulation Model}
+
 We tested the predictive power of the reduced order coagulation model with validation data sets not used during model training. 
@@
 unmodeled components such as platelet activation in the hemophilia data set.
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure-5-FII_ATIII.pdf}
+\caption{Reduced order coagulation model predictions versus experimental data for normal coagulation. 
+The reduced order coagulation model parameter estimates were tested against data not used during model training.
+Simulations of different levels of prothrombin and ATIII were compared with experimental data in the absence of the protein C pathway. 
+Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data.
+The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation). 
+(A,B,C,D) prediction results for (FII,ATIII): (50\%,150\%), (100\%, 100\%), (125\%, 75\%) and (150\%, 50\%) of physiological prothrombin and ATIII levels in the absence of the protein C pathway.
+The experimental validation data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}.
+Thrombin generation in these experiments was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
+As depicted in (A,B,C,D) the prothrombin and ATIII levels were at (50\%,150\%), (100\%, 100\%), (125\%, 75\%) and (150\%, 50\%) of their physiological concentrations in the absence of protein C pathway.
+All factors and control proteins were at their physiological concentration unless otherswise denoted.}\label{fig-fII_ATIII}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure-6-Hemophilia.pdf}
+\caption{Reduced order coagulation model predictions versus experimental data with and without FVIII and FIX.
+The reduced order coagulation model parameter estimates were tested against data not used during model training. 
+Simulations of normal thrombin formation with ATIII and the protein C pathway were compared with thrombin formation in the absence of fVIII and fIX.
+Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data.
+The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation). 
+(A,B) prediction results for normal thrombin generation and thrombin generation in hemophilia. 
+All factors and control proteins were at their physiological concentration unless others noted.
+The experimental validation data was reproduced from the study of Allen \textit{et al.} \citep{ALLEN2006}.}\label{fig-hemophilia}
+\end{figure}
 
@@
 
-\subsection*{Global sensitivity analysis of the reduced order coagulation model.}
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure-7-rFVIIa-Trigger.pdf}
+\caption{Reduced order coagulation model predictions of rFVIIa administration. 
+A: Simulations of thrombin formation in the presence of ATIII and the protein C pathway were conducted for a range of trigger values (1x - 200x nominal) in the absence of fVIII and fIX.
+B: Comparison of thrombin generation for normal versus hemophilia for 10x nominal trigger.
+Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles. 
+The peak thrombin time for normal coagulation ($t^{*}$) is less than rFVIIa induced coagulation in hemophilia ($t^{**}$), while the peak thrombin value was greater in normal coagulation.
+The shaded region denotes the 99\% confidence estimate of the mean thrombin value (uncertainty in the model simulation). 
+All factors and control proteins were at their physiological concentration unless others noted.}\label{fig-rFVII}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure-8-Flux-rFVIIa-Trigger.pdf}
+\caption{Reaction flux distribution as a function of time for thrombin generation under normal (left), hemophilia (center) and rFVIIa treated hemophilia (right).
+Reaction flux was calculated for each particle at $T = 0,4,6,8,10,12,14$ min after the initiation of coagulation. 
+Reaction fluxes were calculated for each particle in the parameter ensemble (N = 20).
+Blue colors denote low flux values while red colors denote high flux values.}\label{fig-flux-analysis}
+\end{figure}
+
+\subsection{Global Sensitivity Analysis of the Reduced Order Coagulation Model}
+
 We conducted a global sensitivity analysis to estimate which parameters controlled the performance of the reduced order model. 
@@
 
-\clearpage
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure-9-Sensitivity.pdf}
+\caption{Global sensitivity analysis of the reduced order coagulation model with respect to the model parameters.
+A: Sensitivity analysis of the thrombin peak time for different prothrombin levels (150\%,100\% and 50\% of the physiological value) as a function of activated protein C. 
+B: Sensitivity analysis of the thrombin exposure for different prothrombin levels (150\%,100\% and 50\% of the physiological value) as a function of activated protein C. 
+Points denote the mean total sensitivity value, while the area around each point denotes the uncertainty in the sensitivity value. 
+The gray dashed line denotes the 45$^{\circ}$ degree diagonal, if sensitivity values are equal for different conditions they will lie on the diagonal.  
+Sensitivity values significantly above or below the diagonal indicate differentially important model parameters. 
+The radius of the shaded region around each total sensitivity value was the maximum uncertainty in that value estimated by the Sobol method. 
+}\label{fig-sensitivity}
+\end{figure}
 
@@
 
-\section*{Discussion}
+\section{Discussion}
+
 In this study, we developed a reduced order model of the human coagulation cascade.
@@
 
-\clearpage
-
-\section*{Materials and Methods}
-
-\subsection*{Formulation and solution of the model equations.}
+\section{Materials and Methods}
+\vspace{-12pt}
+
+\subsection{Formulation and Solution of the Model Equations
+}
+
 We used ordinary differential equations (ODEs) to model the time evolution of proteins ($x_{i}$) in our reduced order coagulation model:
@@
 \begin{equation}\label{eqn:control-factor}
-	f_{i}\left(\mathbf{x}\right) = \frac{\kappa_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}{1 + \kappa_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}
+	f_{i}\left(\mathcal{Z}_{j}\right) = \frac{\kappa_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}{1 + \kappa_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}
 \end{equation}where $\mathcal{Z}_{j}$ denotes the abundance of the $j$ factor (e.g., metabolite abundance), and $\kappa_{ij}$ and $\eta$ are control parameters. 
@@
 
-\subsection*{Estimation of model parameters from experimental data.}
+\subsection{Estimation of Model Parameters From Experimental Data}
+
 Model parameters were estimated by minimizing the difference between simulations and experimental thrombin measurements (squared residual):
@@
 
-\subsection*{Global sensitivity analysis of model performance.}
+\subsection{Global Sensitivity Analysis of Model Performance}
+
 We conducted a global sensitivity analysis, using the variance-based method of Sobol, to estimate which parameters controlled the performance of the reduced order model \citep{SOBOL_METHOD}. 
@@
 
-\section*{Acknowledgements}
+\section*{\noindent  Acknowledgements}
+\vspace{12pt}
+
 This study was supported by an award from the Army Research Office (ARO \#59155-LS).
 
-\clearpage
-%\bibliographystyle{plain}
-%\bibliographystyle{IEEEbib}
-
-\bibliography{References_v1}
+\section*{\noindent Author Contributions}
+\vspace{12pt}
+
+\hl{Please provide ``Author Contributions'' here. }
+
+\section*{\noindent Conflicts of Interest}
+\vspace{12pt}
+
+The authors declare no conflict of interest.
+
+\begin{thebibliography}{9999}
+%\providecommand{\url}[1]{\texttt{#1}}
+%\providecommand{\urlprefix}{URL }
+%\expandafter\ifx\csname urlstyle\endcsname\relax
+%  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
+%  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
+%  \urlstyle{rm}\Url}\fi
+%\providecommand{\bibAnnoteFile}[1]{%
+%  \IfFileExists{#1}{\begin{quotation}\noindent\textsc{Key:} #1\\
+%  \textsc{Annotation:}\ \input{#1}\end{quotation}}{}}
+%\providecommand{\bibAnnote}[2]{%
+%  \begin{quotation}\noindent\textsc{Key:} #1\\
+%  \textsc{Annotation:}\ #2\end{quotation}}
+%\providecommand{\eprint}[2][]{\url{#2}}
+
+\bibitem{2012_kholodenko_kolch_SciSig}
+Kholodenko B, Yaffe MB, Kolch W (2012) Computational approaches for analyzing
+  information flow in biological networks.
+\newblock Sci Signal 5: re1.
+%\bibAnnoteFile{2012_kholodenko_kolch_SciSig}
+
+\bibitem{Gadkar:2005ad}
+Gadkar KG, Varner J, Doyle FJ (2005) {Model identification of signal
+  transduction networks from data using a state regulator problem.}
+\newblock Syst Biol (Stevenage) 2: 17--30.
+%\bibAnnoteFile{Gadkar:2005ad}
+
+\bibitem{2001_bailey_NatBiotech}
+Bailey JE (2001) Complex biology with no parameters.
+\newblock Nat Biotechnol 19: 503-4.
+%\bibAnnoteFile{2001_bailey_NatBiotech}
+
+\bibitem{Machta:2013aa}
+Machta BB, Chachra R, Transtrum MK, Sethna JP (2013) Parameter space
+  compression underlies emergent theories and predictive models.
+\newblock Science 342: 604-7.
+%\bibAnnoteFile{Machta:2013aa}
+
+\bibitem{Tasseff:2011aa}
+Tasseff R, Nayak S, Song SO, Yen A, Varner JD (2011) Modeling and analysis of
+  retinoic acid induced differentiation of uncommitted precursor cells.
+\newblock Integr Biol (Camb) 3: 578-91.
+%\bibAnnoteFile{Tasseff:2011aa}
+
+\bibitem{Morris:2010aa}
+Morris MK, Saez-Rodriguez J, Sorger PK, Lauffenburger DA (2010) Logic-based
+  models for the analysis of cell signaling networks.
+\newblock Biochemistry 49: 3216-24.
+%\bibAnnoteFile{Morris:2010aa}
+
+\bibitem{Saez-Rodriguez:2011aa}
+Saez-Rodriguez J, Alexopoulos LG, Zhang M, Morris MK, Lauffenburger DA, et~al.
+  (2011) Comparing signaling networks between normal and transformed
+  hepatocytes using discrete logical models.
+\newblock Cancer Res 71: 5400-11.
+%\bibAnnoteFile{Saez-Rodriguez:2011aa}
+
+\bibitem{Morris:2011ys}
+Morris MK, Saez-Rodriguez J, Clarke DC, Sorger PK, Lauffenburger DA (2011)
+  Training signaling pathway maps to biochemical data with constrained fuzzy
+  logic: quantitative analysis of liver cell responses to inflammatory stimuli.
+\newblock PLoS Comput Biol 7: e1001099.
+%\bibAnnoteFile{Morris:2011ys}
+
+\bibitem{Morris:2012aa}
+Morris MK, Shriver Z, Sasisekharan R, Lauffenburger DA (2012) Querying
+  quantitative logic models (q2lm) to study intracellular signaling networks
+  and cell-cytokine interactions.
+\newblock Biotechnol J 7: 374-86.
+%\bibAnnoteFile{Morris:2012aa}
+
+\bibitem{Butenas:2002aa}
+Butenas S, Mann KG (2002) Blood coagulation.
+\newblock Biochemistry (Mosc) 67: 3-12.
+%\bibAnnoteFile{Butenas:2002aa}
+
+\bibitem{Schenone:2004aa}
+Schenone M, Furie BC, Furie B (2004) The blood coagulation cascade.
+\newblock Curr Opin Hematol 11: 272-7.
+%\bibAnnoteFile{Schenone:2004aa}
+
+\bibitem{Adams:2009aa}
+Adams RLC, Bird RJ (2009) Review article: Coagulation cascade and therapeutics
+  update: relevance to nephrology. part 1: Overview of coagulation,
+  thrombophilias and history of anticoagulants.
+\newblock Nephrology (Carlton) 14: 462-70.
+%\bibAnnoteFile{Adams:2009aa}
+
+\bibitem{GOLDHABER2006}
+Goldhaber SZ, Colman RW, Clowes AW, editors (2006) {Hemostasis and Thrombosis:
+  Basic Principles and Clinical Practice}.
+\newblock Lippincott Williams and Wilkins.
+%\bibAnnoteFile{GOLDHABER2006}
+
+\bibitem{Brummel:2002aa}
+Brummel KE, Paradis SG, Butenas S, Mann KG (2002) Thrombin functions during
+  tissue factor-induced blood coagulation.
+\newblock Blood 100: 148-52.
+%\bibAnnoteFile{Brummel:2002aa}
+
+\bibitem{MANN1990}
+Mann K, Nesheim M, Church W, Haley P, Krishnaswamy S (1990) Surface-dependent
+  reactions of vitamin k-dependent enzyme complexes.
+\newblock Blood 76: 1-16.
+%\bibAnnoteFile{MANN1990}
+
+\bibitem{ROBERTS1998}
+Roberts H, Monroe D, Oliver J, Chang J, Hoffman M (1998) Newer concepts of
+  blood coagulation.
+\newblock Haemophilia 4: 331-334.
+%\bibAnnoteFile{ROBERTS1998}
+
+\bibitem{MANN1999}
+Mann K (1999) Biochemistry and physiology of blood coagulation.
+\newblock Thromb Haemost 82: 165-174.
+%\bibAnnoteFile{MANN1999}
+
+\bibitem{Khanin:1989aa}
+Khanin MA, Semenov VV (1989) A mathematical model of the kinetics of blood
+  coagulation.
+\newblock J Theor Biol 136: 127-34.
+%\bibAnnoteFile{Khanin:1989aa}
+
+\bibitem{Willems:1991aa}
+Willems GM, Lindhout T, Hermens WT, Hemker HC (1991) Simulation model for
+  thrombin generation in plasma.
+\newblock Haemostasis 21: 197-207.
+%\bibAnnoteFile{Willems:1991aa}
+
+\bibitem{Baldwin:1994aa}
+Baldwin SA, Basmadjian D (1994) A mathematical model of thrombin production in
+  blood coagulation, part i: The sparsely covered membrane case.
+\newblock Ann Biomed Eng 22: 357-70.
+%\bibAnnoteFile{Baldwin:1994aa}
+
+\bibitem{Leipold:1995aa}
+Leipold RJ, Bozarth TA, Racanelli AL, Dicker IB (1995) Mathematical model of
+  serine protease inhibition in the tissue factor pathway to thrombin.
+\newblock J Biol Chem 270: 25383-7.
+%\bibAnnoteFile{Leipold:1995aa}
+
+\bibitem{Kuharsky:2001aa}
+Kuharsky AL, Fogelson AL (2001) Surface-mediated control of blood coagulation:
+  the role of binding site densities and platelet deposition.
+\newblock Biophys J 80: 1050-74.
+%\bibAnnoteFile{Kuharsky:2001aa}
+
+\bibitem{Luan:2007aa}
+Luan D, Zai M, Varner JD (2007) Computationally derived points of fragility of
+  a human cascade are consistent with current therapeutic strategies.
+\newblock PLoS Comput Biol 3: e142.
+%\bibAnnoteFile{Luan:2007aa}
+
+\bibitem{2010_luan_varner_MolBioSys}
+Luan D, Szlam F, Tanaka KA, Barie PS, Varner JD (2010) Ensembles of uncertain
+  mathematical models can identify network response to therapeutic
+  interventions.
+\newblock Mol Biosyst 6: 2272-86.
+%\bibAnnoteFile{2010_luan_varner_MolBioSys}
+
+\bibitem{Jones:1994aa}
+Jones KC, Mann KG (1994) A model for the tissue factor pathway to thrombin. ii.
+  a mathematical simulation.
+\newblock J Biol Chem 269: 23367-73.
+%\bibAnnoteFile{Jones:1994aa}
+
+\bibitem{Hockin:2002aa}
+Hockin MF, Jones KC, Everse SJ, Mann KG (2002) A model for the stoichiometric
+  regulation of blood coagulation.
+\newblock J Biol Chem 277: 18322-33.
+%\bibAnnoteFile{Hockin:2002aa}
+
+\bibitem{Butenas:2004aa}
+Butenas S, Orfeo T, Gissel MT, Brummel KE, Mann KG (2004) The significance of
+  circulating factor ixa in blood.
+\newblock J Biol Chem 279: 22875-82.
+%\bibAnnoteFile{Butenas:2004aa}
+
+\bibitem{Lo:2005aa}
+Lo K, Denney WS, Diamond SL (2005) Stochastic modeling of blood coagulation
+  initiation.
+\newblock Pathophysiol Haemost Thromb 34: 80-90.
+%\bibAnnoteFile{Lo:2005aa}
+
+\bibitem{Chatterjee:2010aa}
+Chatterjee MS, Denney WS, Jing H, Diamond SL (2010) Systems biology of
+  coagulation initiation: kinetics of thrombin generation in resting and
+  activated human blood.
+\newblock PLoS Comput Biol 6.
+%\bibAnnoteFile{Chatterjee:2010aa}
+
+\bibitem{Stalker:2013aa}
+Stalker TJ, Traxler EA, Wu J, Wannemacher KM, Cermignano SL, et~al. (2013)
+  Hierarchical organization in the hemostatic response and its relationship to
+  the platelet-signaling network.
+\newblock Blood 121: 1875-85.
+%\bibAnnoteFile{Stalker:2013aa}
+
+\bibitem{Leiderman:2014aa}
+Leiderman K, Fogelson A (2014) An overview of mathematical modeling of thrombus
+  formation under flow.
+\newblock Thromb Res 133 Suppl 1: S12-4.
+%\bibAnnoteFile{Leiderman:2014aa}
+
+\bibitem{Bannish:2014ab}
+Bannish BE, Keener JP, Fogelson AL (2014) Modelling fibrinolysis: a 3d
+  stochastic multiscale model.
+\newblock Math Med Biol 31: 17-44.
+%\bibAnnoteFile{Bannish:2014ab}
+
+\bibitem{Voronov:2013aa}
+Voronov RS, Stalker TJ, Brass LF, Diamond SL (2013) Simulation of intrathrombus
+  fluid and solute transport using in vivo clot structures with single platelet
+  resolution.
+\newblock Ann Biomed Eng 41: 1297-307.
+%\bibAnnoteFile{Voronov:2013aa}
+
+\bibitem{Runyon:2004aa}
+Runyon MK, Johnson-Kerner BL, Ismagilov RF (2004) Minimal functional model of
+  hemostasis in a biomimetic microfluidic system.
+\newblock Angew Chem Int Ed Engl 43: 1531-6.
+%\bibAnnoteFile{Runyon:2004aa}
+
+\bibitem{Kastrup:2006aa}
+Kastrup CJ, Runyon MK, Shen F, Ismagilov RF (2006) Modular chemical mechanism
+  predicts spatiotemporal dynamics of initiation in the complex network of
+  hemostasis.
+\newblock Proc Natl Acad Sci U S A 103: 15747-52.
+%\bibAnnoteFile{Kastrup:2006aa}
+
+\bibitem{Runyon:2007aa}
+Runyon MK, Johnson-Kerner BL, Kastrup CJ, Van~Ha TG, Ismagilov RF (2007)
+  Propagation of blood clotting in the complex biochemical network of
+  hemostasis is described by a simple mechanism.
+\newblock J Am Chem Soc 129: 7014-5.
+%\bibAnnoteFile{Runyon:2007aa}
+
+\bibitem{Runyon:2008aa}
+Runyon MK, Kastrup CJ, Johnson-Kerner BL, Ha TGV, Ismagilov RF (2008) Effects
+  of shear rate on propagation of blood clotting determined using microfluidics
+  and numerical simulations.
+\newblock J Am Chem Soc 130: 3458-64.
+%\bibAnnoteFile{Runyon:2008aa}
+
+\bibitem{Kuepfer:2007aa}
+Kuepfer L, Peter M, Sauer U, Stelling J (2007) Ensemble modeling for analysis
+  of cell signaling dynamics.
+\newblock Nat Biotechnol 25: 1001-6.
+%\bibAnnoteFile{Kuepfer:2007aa}
+
+\bibitem{Song:2009aa}
+Song SO, Varner J (2009) Modeling and analysis of the molecular basis of pain
+  in sensory neurons.
+\newblock PLoS One 4: e6758.
+%\bibAnnoteFile{Song:2009aa}
+
+\bibitem{2010_song_varner_BiotechJ}
+Song SO, Chakrabarti A, Varner JD (2010) Ensembles of signal transduction
+  models using pareto optimal ensemble techniques (poets).
+\newblock Biotechnol J 5: 768-80.
+%\bibAnnoteFile{2010_song_varner_BiotechJ}
+
+\bibitem{Tasseff:2010aa}
+Tasseff R, Nayak S, Salim S, Kaushik P, Rizvi N, et~al. (2010) Analysis of the
+  molecular networks in androgen dependent and independent prostate cancer
+  revealed fragile and robust subsystems.
+\newblock PLoS One 5: e8864.
+%\bibAnnoteFile{Tasseff:2010aa}
+
+\bibitem{Lequieu:2011aa}
+Lequieu J, Chakrabarti A, Nayak S, Varner JD (2011) Computational modeling and
+  analysis of insulin induced eukaryotic translation initiation.
+\newblock PLoS Comput Biol 7: e1002263.
+%\bibAnnoteFile{Lequieu:2011aa}
+
+\bibitem{Tran:2008aa}
+Tran LM, Rizk ML, Liao JC (2008) Ensemble modeling of metabolic networks.
+\newblock Biophys J 95: 5606-17.
+%\bibAnnoteFile{Tran:2008aa}
+
+\bibitem{HANSSON2005}
+GKHansson (2005) {Inflammation, Atherosclerosis and Coronary Artery Disease}.
+\newblock N Engl J Med 352: 1685 - 1695.
+%\bibAnnoteFile{HANSSON2005}
+
+\bibitem{Tanaka:2009wo}
+Tanaka KA, Key NS, Levy JH (2009) Blood coagulation: hemostasis and thrombin
+  regulation.
+\newblock Anesth Analg 108: 1433-46.
+%\bibAnnoteFile{Tanaka:2009wo}
+
+\bibitem{COOPER1994}
+Tuddenham E, Cooper D (1994) {The molecular genetics of haemostasis and its
+  inherited disorders.}, volume~25 of \emph{Oxford monographs in medical
+  genetics}.
+\newblock Oxford University Press.
+%\bibAnnoteFile{COOPER1994}
+
+\bibitem{MANNUCCI2001}
+Mannucci MP, Tuddenham EGD (2001) The hemophilias - from royal genes to gene
+  therapy.
+\newblock N Engl J Med 344: 1773 - 1780.
+%\bibAnnoteFile{MANNUCCI2001}
+
+\bibitem{MITCHELL2008}
+Mitchell J, Phillott A (2008) Haemophilia and inhibitors 1: diagnosis and
+  treatment.
+\newblock Nursing Times 104: 26--27.
+%\bibAnnoteFile{MITCHELL2008}
+
+\bibitem{TOMOKIYO2003}
+Tomokiyo K, Nakatomi Y, Araki T, Teshima K, Nakano H, et~al. (2003) A novel
+  therapeutic approach combining human plasma-derived factors viia and x for
+  haemophiliacs with inhibitors: evidence of a higher thrombin generation rate
+  in vitro and more sustained haemostatic activity in vivo than obtained with
+  factor viia alone.
+\newblock Vox Sanguinis 85: 290-299.
+%\bibAnnoteFile{TOMOKIYO2003}
+
+\bibitem{Hedner:2008rm}
+Hedner U (2008) Factor viia and its potential therapeutic use in
+  bleeding-associated pathologies.
+\newblock Thromb Haemost 100: 557--562.
+%\bibAnnoteFile{Hedner:2008rm}
+
+\bibitem{Talbot:2009kk}
+Talbot M, Tien HC (2009) The use of recombinant factor viia in trauma patients.
+\newblock J Am Acad Orthop Surg 17: 477-81.
+%\bibAnnoteFile{Talbot:2009kk}
+
+\bibitem{Shapiro:2008px}
+Shapiro AD (2008) Single-dose recombinant activated factor vii for the
+  treatment of joint bleeds in hemophilia patients with inhibitors.
+\newblock Clin Adv Hematol Oncol 6: 579--586.
+%\bibAnnoteFile{Shapiro:2008px}
+
+\bibitem{Duchesne:2008hw}
+Duchesne JC, Mathew KA, Marr AB, Pinsky MR, Barbeau JM, et~al. (2008) Current
+  evidence based guidelines for factor viia use in trauma: the good, the bad,
+  and the ugly.
+\newblock Am Surg 74: 1159-65.
+%\bibAnnoteFile{Duchesne:2008hw}
+
+\bibitem{Butenas:2002ab}
+Butenas S, Brummel KE, Branda RF, Paradis SG, Mann KG (2002) Mechanism of
+  factor viia-dependent coagulation in hemophilia blood.
+\newblock Blood 99: 923-30.
+%\bibAnnoteFile{Butenas:2002ab}
+
+\bibitem{Danforth:2012aa}
+Danforth CM, Orfeo T, Everse SJ, Mann KG, Brummel-Ziedins KE (2012) Defining
+  the boundaries of normal thrombin generation: investigations into hemostasis.
+\newblock PLoS One 7: e30385.
+%\bibAnnoteFile{Danforth:2012aa}
+
+\bibitem{NAITO1991}
+Naito K, Fujikawa K (1991) Activation of human blood coagulation factor {XI}
+  independent of factor {XII}.
+\newblock J Biol Chem 266: 7353--7358.
+%\bibAnnoteFile{NAITO1991}
+
+\bibitem{Gailani:1991aa}
+Gailani D, Broze GJ Jr (1991) Factor xi activation in a revised model of blood
+  coagulation.
+\newblock Science 253: 909-12.
+%\bibAnnoteFile{Gailani:1991aa}
+
+\bibitem{Smith:2006zl}
+Smith SA, Mutch NJ, Baskar D, Rohloff P, Docampo R, et~al. (2006) Polyphosphate
+  modulates blood coagulation and fibrinolysis.
+\newblock Proc Natl Acad Sci U S A 103: 903-8.
+%\bibAnnoteFile{Smith:2006zl}
+
+\bibitem{SCIPY}
+Jones E, Oliphant T, Peterson P (2001--).
+\newblock {SciPy}: Open source scientific tools for {Python}.
+\newblock http://www.scipy.org/.
+%\bibAnnoteFile{SCIPY}
+
+\bibitem{PSO}
+Kennedy J, Eberhart R (1995) Particle swarm optimization.
+\newblock In: Proceedings of the International Conference on Neural Networks.
+  pp. 1942--1948.
+%\bibAnnoteFile{PSO}
+
+\bibitem{Matplotlib}
+Hunter JD (2007) Matplotlib: A 2d graphics environment.
+\newblock Computing in Science and Engineering 9: 90 - 95.
+%\bibAnnoteFile{Matplotlib}
+
+\bibitem{SOBOL_METHOD}
+Sobol I (2001) Global sensitivity indices for nonlinear mathematical models and
+  their monte carlo estimates.
+\newblock Mathematics and Computers in Simulation 55: 271 - 280.
+%\bibAnnoteFile{SOBOL_METHOD}
+
+\bibitem{Saltelli:2010}
+Saltelli A, Annoni P, Azzini I, Campolongo F, Ratto M, et~al. (2010) Variance
+  based sensitivity analysis of model output. design and estimator for the
+  total sensitivity index.
+\newblock Computer Physics Communications 181: 259 - 270.
+%\bibAnnoteFile{Saltelli:2010}
+
+\bibitem{SALIB}
+Herman JD.
+\newblock http://jdherman.github.io/salib/.
+%\bibAnnoteFile{SALIB}
+
+\bibitem{Butenas:1999aa}
+Butenas S, van't Veer C, Mann KG (1999) "normal" thrombin generation.
+\newblock Blood 94: 2169-78.
+%\bibAnnoteFile{Butenas:1999aa}
+
+\bibitem{ALLEN2006}
+Allen GA, Hoffman M, Roberts HR, Monroe DM (2006) Manipulation of prothrombin
+  concentration improves response to high-dose factor {VIIa} in a cell-based
+  model of haemophilia.
+\newblock Br J Haematology 134: 314 - 319.
+%\bibAnnoteFile{ALLEN2006}
+
+\end{thebibliography}
+
+%\begin{thebibliography}{9999}
+%\providecommand{\url}[1]{\texttt{#1}}
+%\providecommand{\urlprefix}{URL }
+%\expandafter\ifx\csname urlstyle\endcsname\relax
+%  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
+%  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
+%  \urlstyle{rm}\Url}\fi
+%\providecommand{\bibAnnoteFile}[1]{%
+%  \IfFileExists{#1}{\begin{quotation}\noindent\textsc{Key:} #1\\
+%  \textsc{Annotation:}\ \input{#1}\end{quotation}}{}}
+%\providecommand{\bibAnnote}[2]{%
+%  \begin{quotation}\noindent\textsc{Key:} #1\\
+%  \textsc{Annotation:}\ #2\end{quotation}}
+%\providecommand{\eprint}[2][]{\url{#2}}
+
+%\bibitem{2012_kholodenko_kolch_SciSig}
+%Kholodenko B, Yaffe MB, Kolch W (2012) Computational approaches for analyzing
+%  information flow in biological networks.
+%\newblock Sci Signal 5: re1.
+%%\bibAnnoteFile{2012_kholodenko_kolch_SciSig}
+%
+%\bibitem{Gadkar:2005ad}
+%Gadkar KG, Varner J, Doyle FJ (2005) {Model identification of signal
+%  transduction networks from data using a state regulator problem.}
+%\newblock Syst Biol (Stevenage) 2: 17--30.
+%\bibAnnoteFile{Gadkar:2005ad}
+%
+%\bibitem{2001_bailey_NatBiotech}
+%Bailey JE (2001) Complex biology with no parameters.
+%\newblock Nat Biotechnol 19: 503-4.
+%\bibAnnoteFile{2001_bailey_NatBiotech}
+%
+%\bibitem{Machta:2013aa}
+%Machta BB, Chachra R, Transtrum MK, Sethna JP (2013) Parameter space
+%  compression underlies emergent theories and predictive models.
+%\newblock Science 342: 604-7.
+%\bibAnnoteFile{Machta:2013aa}
+%
+%\bibitem{Tasseff:2011aa}
+%Tasseff R, Nayak S, Song SO, Yen A, Varner JD (2011) Modeling and analysis of
+%  retinoic acid induced differentiation of uncommitted precursor cells.
+%\newblock Integr Biol (Camb) 3: 578-91.
+%\bibAnnoteFile{Tasseff:2011aa}
+%
+%\bibitem{Morris:2010aa}
+%Morris MK, Saez-Rodriguez J, Sorger PK, Lauffenburger DA (2010) Logic-based
+%  models for the analysis of cell signaling networks.
+%\newblock Biochemistry 49: 3216-24.
+%\bibAnnoteFile{Morris:2010aa}
+%
+%\bibitem{Saez-Rodriguez:2011aa}
+%Saez-Rodriguez J, Alexopoulos LG, Zhang M, Morris MK, Lauffenburger DA, et~al.
+%  (2011) Comparing signaling networks between normal and transformed
+%  hepatocytes using discrete logical models.
+%\newblock Cancer Res 71: 5400-11.
+%\bibAnnoteFile{Saez-Rodriguez:2011aa}
+%
+%\bibitem{Morris:2011ys}
+%Morris MK, Saez-Rodriguez J, Clarke DC, Sorger PK, Lauffenburger DA (2011)
+%  Training signaling pathway maps to biochemical data with constrained fuzzy
+%  logic: quantitative analysis of liver cell responses to inflammatory stimuli.
+%\newblock PLoS Comput Biol 7: e1001099.
+%\bibAnnoteFile{Morris:2011ys}
+%
+%\bibitem{Morris:2012aa}
+%Morris MK, Shriver Z, Sasisekharan R, Lauffenburger DA (2012) Querying
+%  quantitative logic models (q2lm) to study intracellular signaling networks
+%  and cell-cytokine interactions.
+%\newblock Biotechnol J 7: 374-86.
+%\bibAnnoteFile{Morris:2012aa}
+%
+%\bibitem{Butenas:2002aa}
+%Butenas S, Mann KG (2002) Blood coagulation.
+%\newblock Biochemistry (Mosc) 67: 3-12.
+%\bibAnnoteFile{Butenas:2002aa}
+%
+%\bibitem{Schenone:2004aa}
+%Schenone M, Furie BC, Furie B (2004) The blood coagulation cascade.
+%\newblock Curr Opin Hematol 11: 272-7.
+%\bibAnnoteFile{Schenone:2004aa}
+%
+%\bibitem{Adams:2009aa}
+%Adams RLC, Bird RJ (2009) Review article: Coagulation cascade and therapeutics
+%  update: relevance to nephrology. part 1: Overview of coagulation,
+%  thrombophilias and history of anticoagulants.
+%\newblock Nephrology (Carlton) 14: 462-70.
+%\bibAnnoteFile{Adams:2009aa}
+%
+%\bibitem{GOLDHABER2006}
+%Goldhaber SZ, Colman RW, Clowes AW, editors (2006) {Hemostasis and Thrombosis:
+%  Basic Principles and Clinical Practice}.
+%\newblock Lippincott Williams and Wilkins.
+%\bibAnnoteFile{GOLDHABER2006}
+%
+%\bibitem{Brummel:2002aa}
+%Brummel KE, Paradis SG, Butenas S, Mann KG (2002) Thrombin functions during
+%  tissue factor-induced blood coagulation.
+%\newblock Blood 100: 148-52.
+%\bibAnnoteFile{Brummel:2002aa}
+%
+%\bibitem{MANN1990}
+%Mann K, Nesheim M, Church W, Haley P, Krishnaswamy S (1990) Surface-dependent
+%  reactions of vitamin k-dependent enzyme complexes.
+%\newblock Blood 76: 1-16.
+%\bibAnnoteFile{MANN1990}
+%
+%\bibitem{ROBERTS1998}
+%Roberts H, Monroe D, Oliver J, Chang J, Hoffman M (1998) Newer concepts of
+%  blood coagulation.
+%\newblock Haemophilia 4: 331-334.
+%\bibAnnoteFile{ROBERTS1998}
+%
+%\bibitem{MANN1999}
+%Mann K (1999) Biochemistry and physiology of blood coagulation.
+%\newblock Thromb Haemost 82: 165-174.
+%\bibAnnoteFile{MANN1999}
+%
+%\bibitem{Khanin:1989aa}
+%Khanin MA, Semenov VV (1989) A mathematical model of the kinetics of blood
+%  coagulation.
+%\newblock J Theor Biol 136: 127-34.
+%\bibAnnoteFile{Khanin:1989aa}
+%
+%\bibitem{Willems:1991aa}
+%Willems GM, Lindhout T, Hermens WT, Hemker HC (1991) Simulation model for
+%  thrombin generation in plasma.
+%\newblock Haemostasis 21: 197-207.
+%\bibAnnoteFile{Willems:1991aa}
+%
+%\bibitem{Baldwin:1994aa}
+%Baldwin SA, Basmadjian D (1994) A mathematical model of thrombin production in
+%  blood coagulation, part i: The sparsely covered membrane case.
+%\newblock Ann Biomed Eng 22: 357-70.
+%\bibAnnoteFile{Baldwin:1994aa}
+%
+%\bibitem{Leipold:1995aa}
+%Leipold RJ, Bozarth TA, Racanelli AL, Dicker IB (1995) Mathematical model of
+%  serine protease inhibition in the tissue factor pathway to thrombin.
+%\newblock J Biol Chem 270: 25383-7.
+%\bibAnnoteFile{Leipold:1995aa}
+%
+%\bibitem{Kuharsky:2001aa}
+%Kuharsky AL, Fogelson AL (2001) Surface-mediated control of blood coagulation:
+%  the role of binding site densities and platelet deposition.
+%\newblock Biophys J 80: 1050-74.
+%\bibAnnoteFile{Kuharsky:2001aa}
+%
+%\bibitem{Luan:2007aa}
+%Luan D, Zai M, Varner JD (2007) Computationally derived points of fragility of
+%  a human cascade are consistent with current therapeutic strategies.
+%\newblock PLoS Comput Biol 3: e142.
+%\bibAnnoteFile{Luan:2007aa}
+%
+%\bibitem{2010_luan_varner_MolBioSys}
+%Luan D, Szlam F, Tanaka KA, Barie PS, Varner JD (2010) Ensembles of uncertain
+%  mathematical models can identify network response to therapeutic
+%  interventions.
+%\newblock Mol Biosyst 6: 2272-86.
+%\bibAnnoteFile{2010_luan_varner_MolBioSys}
+%
+%\bibitem{Jones:1994aa}
+%Jones KC, Mann KG (1994) A model for the tissue factor pathway to thrombin. ii.
+%  a mathematical simulation.
+%\newblock J Biol Chem 269: 23367-73.
+%\bibAnnoteFile{Jones:1994aa}
+%
+%\bibitem{Hockin:2002aa}
+%Hockin MF, Jones KC, Everse SJ, Mann KG (2002) A model for the stoichiometric
+%  regulation of blood coagulation.
+%\newblock J Biol Chem 277: 18322-33.
+%\bibAnnoteFile{Hockin:2002aa}
+%
+%\bibitem{Butenas:2004aa}
+%Butenas S, Orfeo T, Gissel MT, Brummel KE, Mann KG (2004) The significance of
+%  circulating factor ixa in blood.
+%\newblock J Biol Chem 279: 22875-82.
+%\bibAnnoteFile{Butenas:2004aa}
+%
+%\bibitem{Lo:2005aa}
+%Lo K, Denney WS, Diamond SL (2005) Stochastic modeling of blood coagulation
+%  initiation.
+%\newblock Pathophysiol Haemost Thromb 34: 80-90.
+%\bibAnnoteFile{Lo:2005aa}
+%
+%\bibitem{Chatterjee:2010aa}
+%Chatterjee MS, Denney WS, Jing H, Diamond SL (2010) Systems biology of
+%  coagulation initiation: kinetics of thrombin generation in resting and
+%  activated human blood.
+%\newblock PLoS Comput Biol 6.
+%\bibAnnoteFile{Chatterjee:2010aa}
+%
+%\bibitem{Stalker:2013aa}
+%Stalker TJ, Traxler EA, Wu J, Wannemacher KM, Cermignano SL, et~al. (2013)
+%  Hierarchical organization in the hemostatic response and its relationship to
+%  the platelet-signaling network.
+%\newblock Blood 121: 1875-85.
+%\bibAnnoteFile{Stalker:2013aa}
+%
+%\bibitem{Leiderman:2014aa}
+%Leiderman K, Fogelson A (2014) An overview of mathematical modeling of thrombus
+%  formation under flow.
+%\newblock Thromb Res 133 Suppl 1: S12-4.
+%\bibAnnoteFile{Leiderman:2014aa}
+%
+%\bibitem{Bannish:2014ab}
+%Bannish BE, Keener JP, Fogelson AL (2014) Modelling fibrinolysis: a 3d
+%  stochastic multiscale model.
+%\newblock Math Med Biol 31: 17-44.
+%\bibAnnoteFile{Bannish:2014ab}
+%
+%\bibitem{Voronov:2013aa}
+%Voronov RS, Stalker TJ, Brass LF, Diamond SL (2013) Simulation of intrathrombus
+%  fluid and solute transport using in vivo clot structures with single platelet
+%  resolution.
+%\newblock Ann Biomed Eng 41: 1297-307.
+%\bibAnnoteFile{Voronov:2013aa}
+%
+%\bibitem{Runyon:2004aa}
+%Runyon MK, Johnson-Kerner BL, Ismagilov RF (2004) Minimal functional model of
+%  hemostasis in a biomimetic microfluidic system.
+%\newblock Angew Chem Int Ed Engl 43: 1531-6.
+%\bibAnnoteFile{Runyon:2004aa}
+%
+%\bibitem{Kastrup:2006aa}
+%Kastrup CJ, Runyon MK, Shen F, Ismagilov RF (2006) Modular chemical mechanism
+%  predicts spatiotemporal dynamics of initiation in the complex network of
+%  hemostasis.
+%\newblock Proc Natl Acad Sci U S A 103: 15747-52.
+%\bibAnnoteFile{Kastrup:2006aa}
+%
+%\bibitem{Runyon:2007aa}
+%Runyon MK, Johnson-Kerner BL, Kastrup CJ, Van~Ha TG, Ismagilov RF (2007)
+%  Propagation of blood clotting in the complex biochemical network of
+%  hemostasis is described by a simple mechanism.
+%\newblock J Am Chem Soc 129: 7014-5.
+%\bibAnnoteFile{Runyon:2007aa}
+%
+%\bibitem{Runyon:2008aa}
+%Runyon MK, Kastrup CJ, Johnson-Kerner BL, Ha TGV, Ismagilov RF (2008) Effects
+%  of shear rate on propagation of blood clotting determined using microfluidics
+%  and numerical simulations.
+%\newblock J Am Chem Soc 130: 3458-64.
+%\bibAnnoteFile{Runyon:2008aa}
+%
+%\bibitem{Kuepfer:2007aa}
+%Kuepfer L, Peter M, Sauer U, Stelling J (2007) Ensemble modeling for analysis
+%  of cell signaling dynamics.
+%\newblock Nat Biotechnol 25: 1001-6.
+%\bibAnnoteFile{Kuepfer:2007aa}
+%
+%\bibitem{Song:2009aa}
+%Song SO, Varner J (2009) Modeling and analysis of the molecular basis of pain
+%  in sensory neurons.
+%\newblock PLoS One 4: e6758.
+%\bibAnnoteFile{Song:2009aa}
+%
+%\bibitem{2010_song_varner_BiotechJ}
+%Song SO, Chakrabarti A, Varner JD (2010) Ensembles of signal transduction
+%  models using pareto optimal ensemble techniques (poets).
+%\newblock Biotechnol J 5: 768-80.
+%\bibAnnoteFile{2010_song_varner_BiotechJ}
+%
+%\bibitem{Tasseff:2010aa}
+%Tasseff R, Nayak S, Salim S, Kaushik P, Rizvi N, et~al. (2010) Analysis of the
+%  molecular networks in androgen dependent and independent prostate cancer
+%  revealed fragile and robust subsystems.
+%\newblock PLoS One 5: e8864.
+%\bibAnnoteFile{Tasseff:2010aa}
+%
+%\bibitem{Lequieu:2011aa}
+%Lequieu J, Chakrabarti A, Nayak S, Varner JD (2011) Computational modeling and
+%  analysis of insulin induced eukaryotic translation initiation.
+%\newblock PLoS Comput Biol 7: e1002263.
+%\bibAnnoteFile{Lequieu:2011aa}
+%
+%\bibitem{Tran:2008aa}
+%Tran LM, Rizk ML, Liao JC (2008) Ensemble modeling of metabolic networks.
+%\newblock Biophys J 95: 5606-17.
+%\bibAnnoteFile{Tran:2008aa}
+%
+%\bibitem{HANSSON2005}
+%GKHansson (2005) {Inflammation, Atherosclerosis and Coronary Artery Disease}.
+%\newblock N Engl J Med 352: 1685 - 1695.
+%\bibAnnoteFile{HANSSON2005}
+%
+%\bibitem{Tanaka:2009wo}
+%Tanaka KA, Key NS, Levy JH (2009) Blood coagulation: hemostasis and thrombin
+%  regulation.
+%\newblock Anesth Analg 108: 1433-46.
+%\bibAnnoteFile{Tanaka:2009wo}
+%
+%\bibitem{COOPER1994}
+%Tuddenham E, Cooper D (1994) {The molecular genetics of haemostasis and its
+%  inherited disorders.}, volume~25 of \emph{Oxford monographs in medical
+%  genetics}.
+%\newblock Oxford University Press.
+%\bibAnnoteFile{COOPER1994}
+%
+%\bibitem{MANNUCCI2001}
+%Mannucci MP, Tuddenham EGD (2001) The hemophilias - from royal genes to gene
+%  therapy.
+%\newblock N Engl J Med 344: 1773 - 1780.
+%\bibAnnoteFile{MANNUCCI2001}
+%
+%\bibitem{MITCHELL2008}
+%Mitchell J, Phillott A (2008) Haemophilia and inhibitors 1: diagnosis and
+%  treatment.
+%\newblock Nursing Times 104: 26--27.
+%\bibAnnoteFile{MITCHELL2008}
+%
+%\bibitem{TOMOKIYO2003}
+%Tomokiyo K, Nakatomi Y, Araki T, Teshima K, Nakano H, et~al. (2003) A novel
+%  therapeutic approach combining human plasma-derived factors viia and x for
+%  haemophiliacs with inhibitors: evidence of a higher thrombin generation rate
+%  in vitro and more sustained haemostatic activity in vivo than obtained with
+%  factor viia alone.
+%\newblock Vox Sanguinis 85: 290-299.
+%\bibAnnoteFile{TOMOKIYO2003}
+%
+%\bibitem{Hedner:2008rm}
+%Hedner U (2008) Factor viia and its potential therapeutic use in
+%  bleeding-associated pathologies.
+%\newblock Thromb Haemost 100: 557--562.
+%\bibAnnoteFile{Hedner:2008rm}
+%
+%\bibitem{Talbot:2009kk}
+%Talbot M, Tien HC (2009) The use of recombinant factor viia in trauma patients.
+%\newblock J Am Acad Orthop Surg 17: 477-81.
+%\bibAnnoteFile{Talbot:2009kk}
+%
+%\bibitem{Shapiro:2008px}
+%Shapiro AD (2008) Single-dose recombinant activated factor vii for the
+%  treatment of joint bleeds in hemophilia patients with inhibitors.
+%\newblock Clin Adv Hematol Oncol 6: 579--586.
+%\bibAnnoteFile{Shapiro:2008px}
+%
+%\bibitem{Duchesne:2008hw}
+%Duchesne JC, Mathew KA, Marr AB, Pinsky MR, Barbeau JM, et~al. (2008) Current
+%  evidence based guidelines for factor viia use in trauma: the good, the bad,
+%  and the ugly.
+%\newblock Am Surg 74: 1159-65.
+%\bibAnnoteFile{Duchesne:2008hw}
+%
+%\bibitem{Butenas:2002ab}
+%Butenas S, Brummel KE, Branda RF, Paradis SG, Mann KG (2002) Mechanism of
+%  factor viia-dependent coagulation in hemophilia blood.
+%\newblock Blood 99: 923-30.
+%\bibAnnoteFile{Butenas:2002ab}
+%
+%\bibitem{Danforth:2012aa}
+%Danforth CM, Orfeo T, Everse SJ, Mann KG, Brummel-Ziedins KE (2012) Defining
+%  the boundaries of normal thrombin generation: investigations into hemostasis.
+%\newblock PLoS One 7: e30385.
+%\bibAnnoteFile{Danforth:2012aa}
+%
+%\bibitem{NAITO1991}
+%Naito K, Fujikawa K (1991) Activation of human blood coagulation factor {XI}
+%  independent of factor {XII}.
+%\newblock J Biol Chem 266: 7353-7358.
+%\bibAnnoteFile{NAITO1991}
+%
+%\bibitem{Gailani:1991aa}
+%Gailani D, Broze GJ Jr (1991) Factor xi activation in a revised model of blood
+%  coagulation.
+%\newblock Science 253: 909-12.
+%\bibAnnoteFile{Gailani:1991aa}
+%
+%\bibitem{Smith:2006zl}
+%Smith SA, Mutch NJ, Baskar D, Rohloff P, Docampo R, et~al. (2006) Polyphosphate
+%  modulates blood coagulation and fibrinolysis.
+%\newblock Proc Natl Acad Sci U S A 103: 903-8.
+%\bibAnnoteFile{Smith:2006zl}
+%
+%\bibitem{SCIPY}
+%Jones E, Oliphant T, Peterson P (2001--).
+%\newblock {SciPy}: Open source scientific tools for {Python}.
+%\newblock http://www.scipy.org/.
+%\bibAnnoteFile{SCIPY}
+%
+%\bibitem{PSO}
+%Kennedy J, Eberhart R (1995) Particle swarm optimization.
+%\newblock In: Proceedings of the International Conference on Neural Networks.
+%  pp. 1942 - 1948.
+%\bibAnnoteFile{PSO}
+%
+%\bibitem{Matplotlib}
+%Hunter JD (2007) Matplotlib: A 2d graphics environment.
+%\newblock Computing in Science and Engineering 9: 90 - 95.
+%\bibAnnoteFile{Matplotlib}
+%
+%\bibitem{SOBOL_METHOD}
+%Sobol I (2001) Global sensitivity indices for nonlinear mathematical models and
+%  their monte carlo estimates.
+%\newblock Mathematics and Computers in Simulation 55: 271 - 280.
+%\bibAnnoteFile{SOBOL_METHOD}
+%
+%\bibitem{Saltelli:2010}
+%Saltelli A, Annoni P, Azzini I, Campolongo F, Ratto M, et~al. (2010) Variance
+%  based sensitivity analysis of model output. design and estimator for the
+%  total sensitivity index.
+%\newblock Computer Physics Communications 181: 259 - 270.
+%\bibAnnoteFile{Saltelli:2010}
+%
+%\bibitem{SALIB}
+%Herman JD.
+%\newblock http://jdherman.github.io/salib/.
+%\bibAnnoteFile{SALIB}
+%
+%\bibitem{Butenas:1999aa}
+%Butenas S, van't Veer C, Mann KG (1999) "normal" thrombin generation.
+%\newblock Blood 94: 2169-78.
+%\bibAnnoteFile{Butenas:1999aa}
+%
+%\bibitem{ALLEN2006}
+%Allen GA, Hoffman M, Roberts HR, Monroe DM (2006) Manipulation of prothrombin
+%  concentration improves response to high-dose factor {VIIa} in a cell-based
+%  model of haemophilia.
+%\newblock Br J Haematology 134: 314 - 319.
+%\bibAnnoteFile{ALLEN2006}
+
+%\end{thebibliography}
 
@@
 
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-1-Network.pdf}
-\caption{Schematic of the connectivity of the reduced order coagulation model. A trigger compound, e.g., TF/FVIIa initiates thrombin production (FIIa) from prothrombin (fII).
-Once activated, thrombin catalyzes its own activation (amplification step), as well as its own inhibition via the conversion of protein C to activated protein C (APC).
-APC and tissue factor pathway inhibitor (TFPI) inhibit initiation and amplification, while antithrombin III (ATIII) directly inhibits thrombin. 
-All inhibition steps and trigger-induced initiation were modeled using a rule-based approach. 
-Likewise, the dependence of amplification on other coagulation factors was also modeled using a rule-based approach. 
-The abundance of the highlighted species (in the dashed boxes) was governed by an ordinary differential equation. 
-All other species were assumed to be constant.}\label{fig-network}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-2-ControlSchematic.pdf}
-\caption{Schematic of rule-based effective control laws. 
-Traditional enzyme kinetic expressions, e.g., Michaelis–Menten or multiple saturation kinetics are multiplied by an enzyme activity control variable $0 \leq v_{j} \leq 1 $. 
-Control variables are functions of many possible regulatory factors encoded by arbitrary transfer functions of the form $0\leq f_{j}\left(\mathcal{Z}\right)\leq 1$.
-At each simulation time step, the $v_{j}$ variables are calculated by evaluating integration rules such as the max or min of the set of transfer functions $f_{1},\hdots,f_{n}$ 
-influencing the activity of enzyme $E_{j}$. }\label{fig-control-schematic}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-3-DIFF_FII_NO_APC.pdf}
-\caption{Reduced order coagulation model training simulations.
-Reduced order coagulation model parameters were estimated using particle swarm optimization (PSO) with and without the protein C pathway as a function of prothrombin. 
-Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data. 
-The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation).
-(A,B,C) training results for 150\%, 100\% and 50\% of physiological prothrombin levels in the absence of the protein C pathway.
-Thrombin generation was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
-All factors and control proteins were at their physiological concentration unless others denoted. 
-The experimental training data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}.}\label{fig-diff_fII_NO_APC}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-4-DIFF_FII_APC.pdf}
-\caption{Reduced order coagulation model training simulations.
-Reduced order coagulation model parameters were estimated using particle swarm optimization (PSO) with and without the protein C pathway as a function of prothrombin. 
-Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data.
-The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation).
-(A,B,C) training results for 150\%, 100\% and 50\% of physiological prothrombin levels in the presence of the protein C pathway. 
-Only APC pathway parameters were allowed to vary in the simulations on the right.
-Thrombin generation was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
-All factors and control proteins were at their physiological concentration unless others denoted.
-The experimental training data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}.}\label{fig-diff_fII_APC}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-5-FII_ATIII.pdf}
-\caption{Reduced order coagulation model predictions versus experimental data for normal coagulation. 
-The reduced order coagulation model parameter estimates were tested against data not used during model training.
-Simulations of different levels of prothrombin and ATIII were compared with experimental data in the absence of the protein C pathway. 
-Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data.
-The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation). 
-(A,B,C,D) prediction results for (FII,ATIII): (50\%,150\%), (100\%, 100\%), (125\%, 75\%) and (150\%, 50\%) of physiological prothrombin and ATIII levels in the absence of the protein C pathway. 
-Thrombin generation was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
-All factors and control proteins were at their physiological concentration unless others denoted.
-The experimental validation data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}}\label{fig-fII_ATIII}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-6-Hemophilia.pdf}
-\caption{Reduced order coagulation model predictions versus experimental data with and without FVIII and FIX.
-The reduced order coagulation model parameter estimates were tested against data not used during model training. 
-Simulations of normal thrombin formation with ATIII and the protein C pathway were compared with thrombin formation in the absence of fVIII and fIX.
-Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data.
-The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation). 
-(A,B) prediction results for normal thrombin generation and thrombin generation in hemophilia. 
-All factors and control proteins were at their physiological concentration unless others noted.
-The experimental validation data was reproduced from the study of Allen \textit{et al.} \citep{ALLEN2006}.}\label{fig-hemophilia}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-7-rFVIIa-Trigger.pdf}
-\caption{Reduced order coagulation model predictions of rFVIIa administration. 
-A: Simulations of thrombin formation in the presence of ATIII and the protein C pathway were conducted for a range of trigger values (1x - 200x nominal) in the absence of fVIII and fIX.
-B: Comparison of thrombin generation for normal versus hemophilia for 10x nominal trigger.
-Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles. 
-The peak thrombin time for normal coagulation ($t^{*}$) is less than rFVIIa induced coagulation in hemophilia ($t^{**}$), while the peak thrombin value was greater in normal coagulation.
-The shaded region denotes the 99\% confidence estimate of the mean thrombin value (uncertainty in the model simulation). 
-All factors and control proteins were at their physiological concentration unless others noted.}\label{fig-rFVII}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-8-Flux-rFVIIa-Trigger.pdf}
-\caption{Reaction flux distribution as a function of time for thrombin generation under normal (left), hemophilia (center) and rFVIIa treated hemophilia (right).
-Reaction flux was calculated for each particle at $T = 0,4,6,8,10,12,14$ min after the initiation of coagulation. 
-Reaction fluxes were calculated for each particle in the parameter ensemble (N = 20).
-Blue colors denote low flux values while red colors denote high flux values.}\label{fig-flux-analysis}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-9-Sensitivity.pdf}
-\caption{Global sensitivity analysis of the reduced order coagulation model with respect to the model parameters.
-A: Sensitivity analysis of the thrombin peak time for different prothrombin levels (150\%,100\% and 50\% of the physiological value) as a function of activated protein C. 
-B: Sensitivity analysis of the thrombin exposure for different prothrombin levels (150\%,100\% and 50\% of the physiological value) as a function of activated protein C. 
-Points denote the mean total sensitivity value, while the area around each point denotes the uncertainty in the sensitivity value. 
-The gray dashed line denotes the 45$^{\circ}$ degree diagonal, if sensitivity values are equal for different conditions they will lie on the diagonal.  
-Sensitivity values significantly above or below the diagonal indicate differentially important model parameters. 
-The radius of the shaded region around each total sensitivity value was the maximum uncertainty in that value estimated by the Sobol method. 
-}\label{fig-sensitivity}
-\end{figure}
-
-\clearpage
-
-% Supplemental figures -
-% Set the S- 
-\renewcommand\thefigure{S\arabic{figure}}
-\renewcommand\thetable{T\arabic{table}}
-\renewcommand\thepage{S-\arabic{page}}
-\renewcommand\theequation{S\arabic{equation}}
-
-% Reset the counters -
-\setcounter{equation}{0}
-\setcounter{table}{0}
-\setcounter{figure}{0}
-\setcounter{page}{1}
+%\clearpage
+%
+%% Supplemental figures -
+%% Set the S- 
+%\renewcommand\thefigure{S\arabic{figure}}
+%\renewcommand\thetable{T\arabic{table}}
+%\renewcommand\thepage{S-\arabic{page}}
+%\renewcommand\theequation{S\arabic{equation}}
+%
+%% Reset the counters -
+%\setcounter{equation}{0}
+%\setcounter{table}{0}
+%\setcounter{figure}{0}
+%\setcounter{page}{1}
 

### 2014-12-16 revision/Paper_v3.tex
Final revesion from JV

@@
 Second, when integrated into network models, these rules captured classic regulatory patterns such as product-induced feedback inhibition.
-Lastly, we showed, at least for the network architectures considered here, that we could simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data. 
+Lastly, we showed, at least for the network architectures considered here, that we could simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data 
+starting from an unbiased collection of possible allosteric structures using particle swarm optimization.
+However, when starting with an initial population that was heavily enriched with incorrect structures, our particle swarm approach could converge to an incorrect structure.
 While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.
@@
 Unfortunately, traditional, fully structured cybernetic models also suffer from an identifiability challenge, as both the kinetic parameters and an abstracted model of cellular objectives must be estimated simultaneously.
-Recent cybernetic formulations from Ramkrishna and colleagues have successfully treated this identifiability challenge through elementary mode reduction, though the techniques replace detailed biological mechanism with an optimization heuristic \cite{2009_song_ramkrishna_BiotechBioeng,Song:2011aa}.
+However, recent cybernetic formulations from Ramkrishna and colleagues have successfully treated this identifiability challenge through elementary mode reduction, though the techniques replace detailed biological mechanism with an optimization heuristic \cite{2009_song_ramkrishna_BiotechBioeng,Song:2011aa}.
 
@@
 Thus, cell-free operation holds several significant advantages for model development, identification and validation.
-Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms, such as \textit{E.
-coli} or \textit{B.
-subtilis}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.
+Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms,
+such as \textit{E. coli} or \textit{B. subtilis}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.
 
@@
 Second, when integrated into network models, these rules captured classical regulatory patterns such as product-induced feedback inhibition.
-Lastly, we showed, at least for the network architectures considered here, that we could simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data. 
+Lastly, we showed, at least for the network architectures considered here, that we could simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data 
+starting from an unbiased collection of possible allosteric structures using particle swarm optimization.
+However, when starting with an initial population that was heavily enriched with incorrect structures, our particle swarm approach could converge to an incorrect structure.
 While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.
@@
 Network A and C differ by the regulatory connection between the end product P$_2$ and enzyme $E_{1}$; in network A, end product P$_{2}$ was assumed to inhibit $E_{1}$, while in network C, end product P$_{2}$ activated $E_{1}$.
-Lastly, when the initial population was biased towards incorrect structures (initial population seeded with 90\% incorrect structures), the particle swarm \textit{misidentified} the correct allosteric structure (Fig.
+Lastly, when the initial population was heavily biased towards incorrect structures (initial population seeded with 90\% incorrect structures), the particle swarm \textit{misidentified} the correct allosteric structure (Fig.
 \ref{fig-control-search}C).
@@
 If we considered all putative regulatory architectures to be equally likely, we were able to estimate a \textit{sloppy} ensemble of models with the correct architecture and kinetic parameters.
+Thus, we identified parameter values that were different from their true values, but nonetheless produced reasonable model performance (low error).
+This suggested that we captured important parameter combinations (stiff combinations), while simultaneously missing other parameter combinations (sloppy combinations).
+This was similar to the earlier study of Brown and Sethna \citep{Brown:2003aa},
+which showed that reasonable model predictions were possible, despite sometimes only order of magnitude parameter estimates, if the stiff parameter combinations were well constrained.
 
@@
 Liao and coworkers developed a method that generates an ensemble of kinetic models that all approach the same steady-state, one determined by fluxomics measurements \cite{Tran:2008aa}. 
-The best subpopulation of candidate models are selected based on their agreement with further measurements of genetically perturbed systems. 
+The best subpopulation of candidate models were selected based on their agreement with further measurements of genetically perturbed systems. 
 Our work relies on heuristic search optimization to identify kinetic models consistent with steady-state and dynamic time-series measurements of cellular species \citep{Luan:2007aa,Song:2009aa,Tasseff:2010aa,Tasseff:2011aa,Nayak:2011aa,Lequieu:2011aa}.
@@
 Thus, given our previous experience with models containing hundreds of unknown parameters, we expect parameter estimation to be a manageable challenge.
-On the other hand, a critical challenge will be the estimation of allosteric connectivity at a genome scale. 
+
+A critical challenge will be the estimation of allosteric connectivity at a genome scale. 
 The regulation of glycolytic enzymes, such as phosphofructokinase I, has been studied for many years \citep{BERG_BIOCHEMISTRY,Peskov:2008aa}. 
@@
 Also, there are many different approaches from the reverse engineering of gene regulatory networks that perhaps could be adopted to this problem, however this remains an open question.
+For example, one could imagine designing pulse chase experiments which maximally distinguish between competing allosteric models, similar to the earlier work of Kremling et al \citep{Kremling:2004aa}, or
+iteratively estimate model structures similar to Doyle and coworkers \citep{Gadkar:2005aa}.
 Lastly, the choice of max/min integration rules or the particular form of the transfer functions could be generalized to include other rule types and functions. 
@@
 For a fixed control gain parameter $\kappa_{control}$, the reaction rate approached a step for increasing cooperativity control parameter $\eta$. 
+For activation simulations $\kappa_{control}=$0.05 and $\eta=\left\{0.01,0.1,1,2,4,6,8,10\right\}$.
 \textbf{B}:The conversion of substrate $S$ to product $P$ by enzyme $E$ with inhibitor $I$. 
 For a fixed control gain parameter $\kappa_{control}$, the reaction rate approximated non-competitive inhibition for increasing cooperativity control parameter $\eta$. 
-}\label{fig-kinetics-simulations}
+For the inhibition simulations $\kappa_{control}=$1.5 and $\eta=\left\{0.01,0.1,1,2,4,6,8,10\right\}$.}\label{fig-kinetics-simulations}
 \end{figure}
@@
 \caption{Metabolic flux and control variables as a function of network type and particle index at t = 100 min.
-The control variables governing $E_{1},E_{3}$ and $E_{6}$ activity and the scaled metabolic flux were calculated for the positively (top), 
+The particle error, the control variables governing $E_{1},E_{3}$ and $E_{6}$ activity ($v_{1},v_{3}$ and $v_{3}$) and the scaled metabolic flux were calculated for the positively (top), 
 uniformly (middle) and negatively (bottom) biased particle swarms (N = 100). 

### 2014-12-17 revision/Paper_v3.tex
final revisions

@@
 Of course, these issues are not independent; any description of enzyme activity regulation will be a function of system state, which in turn depends upon the kinetic parameters. 
-For cell free systems, regulated gene expression has been removed, however, enzyme activity regulation is still operational. 
+For cell-free systems, regulated gene expression has been removed, however, enzyme activity regulation is still operational. 
 We explored this linkage by estimating model parameters from synthetic data using both network structures. 
 We generated synthetic measurements of the substrate S, intermediate M$_{5}$ and end product P$_1$ approximately every 20 min using
-network A. This data set is similar to published cell free studies both in terms of network coverage, and sampling frequency \citep{Jewett:2008aa}. 
+network A. This data set is similar to published cell-free studies, both in terms of network coverage and sampling frequency \citep{Jewett:2008aa}. 
 We then generated an ensemble of model parameter estimates by minimizing the difference between model simulations and the synthetic data using particle swarm optimization (PSO), starting from random initial parameter guesses.
@@
 For example, one could imagine designing pulse chase experiments which maximally distinguish between competing allosteric models, similar to the earlier work of Kremling et al \citep{Kremling:2004aa}, or
-iteratively estimate model structures similar to Doyle and coworkers \citep{Gadkar:2005aa}.
+iteratively estimating model structures similar to Doyle and coworkers \citep{Gadkar:2005aa}.
 Lastly, the choice of max/min integration rules or the particular form of the transfer functions could be generalized to include other rule types and functions. 
@@
 \end{eqnarray}
-where $\mathbf{\Delta_i}$ denotes the perturbation to the vector of parameters $\mathbf{k_i}$ for particle $i$. 
+where $\mathbf{\Delta}_i$ denotes the perturbation to the vector of parameters $\mathbf{k}_i$ for particle $i$. 
 $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes the best local solution found by particle $i$, and

### 2015-01-22 Paper_2015_Ec_glyco_v1.tex
rough draft figures + captions

\documentclass[12pt]{article}

% Load packages
\usepackage{cite}              % Make references as [1-4], not [1,2,3,4]
\usepackage{url}              	% Formatting web addresses
\usepackage{ifthen}           	% Conditional
\usepackage{multicol}			% Multi-column pages
\usepackage[utf8]{inputenc}   	% Unicode support
\usepackage{amsmath}           % Support for writing math formulas
\usepackage{amssymb}           % Support for writing math formulas
\usepackage{epsfig}            % Support separate PostScript files for figs
\usepackage{epstopdf}			% Converts eps figs to pdf
\usepackage{graphicx}			% Graphics functions
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
% \usepackage{caption}			% Styles figure captions
\usepackage{setspace}			% Provides line spacing environments
\usepackage{colortbl}
\usepackage[wide]{sidecap}		% Can typeset caption aside the figure, allows use of margin for figs wider than \textwidth.
\usepackage[square,sort,comma,numbers]{natbib}
% \usepackage[authoryear,square,comma,sort&compress]{natbib}
\usepackage{supertabular}		% Tables spanning multiple pages
\usepackage{comment}
\usepackage{lineno}
\urlstyle{rm}					% URL style
\usepackage{wrapfig}			% Wrap text around figures
\usepackage[FIGTOPCAP,nooneline]{subfigure}
\captionsetup{font={scriptsize},aboveskip=0pt}	% Styles figure captions
\usepackage{simplemargins}
\setallmargins{1.0in}         % Set all margins
% \setlength\bibsep{0pt}     % Single spaced bib
\date{} 
\doublespacing
% \singlespacing
\frenchspacing     % Eliminates double spaces between sentences

\makeatletter
\renewcommand\subsection{\@startsection
	{subsection}{2}{0mm}
	{-0.05in}
	{-0.5\baselineskip}
	{\normalfont\normalsize\bfseries}}
\renewcommand\subsubsection{\@startsection
	{subsubsection}{2}{0mm}
	{-0.05in}
	{-0.5\baselineskip}
	{\normalfont\normalsize\itshape}}
\renewcommand\section{\@startsection
	{subsection}{2}{0mm}
	{-0.2in}
	{0.05\baselineskip}
	{\normalfont\large\bfseries}}	
\renewcommand\paragraph{\@startsection
	{paragraph}{2}{0mm}
	{-0.05in}
	{-0.5\baselineskip}
	{\normalfont\normalsize\itshape}}
\makeatother

\newboolean{publ}

%Publication style settings
%\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}
\renewcommand{\rmdefault}{phv}\renewcommand{\sfdefault}{phv}
\renewcommand{\bibnumfmt}[1]{#1.}    % Change the number format in the ref list
\renewcommand{\figurename}{Fig.}     % Change Figure to Fig.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TITLE PAGE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{titlepage}

{\par\centering\textbf{\Large Improving Glycosylation Efficiency in \textit{Escherichia coli} through Model-Guided Metabolic Engineering}}
\vspace{0.05in}
{\par \centering \large{Joseph A. Wayman$^{1}$, Thomas J. Mansell$^{3}$, Matthew P. DeLisa$^{2}$, and Jeffrey D. Varner$^{2*}$}}
\vspace{0.10in}
{\par \centering \large{$^{1*}$School of Applied and Engineering Physics, Cornell University, Ithaca NY 14853}}
{\par \centering \large{$^{2*}$School of Chemical and Biomolecular Engineering, Cornell University, Ithaca NY 14853}}
{\par \centering \large{$^{3*}$School of Chemical and Biological Engineering, University of Colorado Boulder, Boulder CO 80309}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Model-guided glycoengineering in \textit{E. coli}}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Metabolic Engineering}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
{\par \centering Email: jdv27@cornell.edu} 
{\par \centering Phone: (607) 255 - 4258} 
{\par \centering Fax: (607) 255 - 9166} 
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ABSTRACT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}

Abstract goes here ...

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% INTRODUCTION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\setcounter{page}{1}
\linenumbers
\section*{Introduction}

Intro goes here ...

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% RESULTS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section*{Results}

Results go here ...

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DISCUSSION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section*{Discussion}

Discussion go here ...

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MATERIALS & METHODS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section*{Materials and Methods}

Reactions encoding \textit{C. jejuni} glycan formation (See Table) were added to the genome-scale metabolic model of \textit{E. coli} iAF1260 \cite{2007_feist_reed_hatzimanikatis_palsson_MolSysBio}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ACKNOWLEDGEMENTS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section*{Acknowledgements}

The authors thank the anonymous reviewers for their helpful suggestions. 
We also acknowledge the gracious financial support to J.V. by the National Science Foundation CAREER (CBET-0846876) for the support of J.W.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% AUTHOR CONTRIBUTIONS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Author Contributions}

Author contributions go here ...

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% CONFLICT OF INTEREST
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Conflict of Interest}

The authors declare no conflicts of interest.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% BIBLIOGRAPHY
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\bibliographystyle{nature}
\bibliography{Paper_2015_Ec_glyco_ref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% FIGURES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage

% Figure 1
\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{./figures/fig1_glyco_pathway.pdf}
\caption{Glycosylation pathway in \textit{C. jejuni} and \textit{E. coli}. 
Glycan assembly, facilitated by \textit{pgl} locus enzymes, takes place on a lipid carrier, undecaprenyl pyrophosphate (UDCP), from cytoplasmic pools of nucleotide-activated sugars N-acetylglucosamine (GlcNAc), N-acetylgalactosamine (GalNAc), and glucose (Glc). 
The glycan is then flipped onto the periplasmic side of the inner membrane, where it is transferred to an asparagine residue on a glycoprotein acceptor motif.}
\label{fig_pathway_cjejuni}
\end{figure}

\clearpage

% Figure 2
\begin{figure}\centering
\includegraphics[width=1.0\textwidth]{./figures/fig2_workflow.pdf}
\caption{Heuristic optimization approach used to identify strains coupling growth to glycan production. 
The chromosome is defined as two separate binary arrays, one defining the state of metabolic enzyme expression and another defining the state of transcriptional regulator activation. 
Gene repression and knockouts are designated by zeros. 
Nutrient conditions define the boundary constraints within the stoichiometric model which in turn affect the state of the metabolic enzyme chromosome. 
Gene repression and knockouts determine the constraints placed on fluxes in the stoichiometric model. 
Nutrients are mapped to the state of transcriptional regulators and genes are mapped to the state of flux constraints using Boolean rules as defined in \cite{2004_covert_reed_herrgard_palsson_Nat,2007_feist_reed_hatzimanikatis_palsson_MolSysBio}. 
FBA is used to maximize growth rate under the constraints imposed by the mutant strain and transcriptional regulation and the fitness objective is calculated. 
Here, we use shadow price. 
The strain is accepted or rejected based on the change in fitness and a Boltzmann criterion. 
New mutant strains are randomly generated from accepted ones. 
The search continues until a positive shadow price is achieved.}
\label{fig_workflow_heuristic}
\end{figure}

\clearpage

% Figure 3
\begin{figure}\centering
\includegraphics[width=1.0\textwidth]{./figures/fig3_FACS_hist.pdf}
\caption{\textbf{(A)} Fluorescence from outer membrane labeled \textit{C. jejuni} glycans of single knockout $\Delta$\textit{gnd} and double knockout $\Delta$\textit{sdhC} $\Delta$\textit{gnd}, detecting by flow cytometry. 
$\dagger$ indicates a strain predicted to eliminate glycan flux. 
Stars indicate statistically significant increases in fluorescence according to a t-test ($p<0.05$). 
\textbf{(B)} Geometric mean fluorescence, normalized to the wild-type value, from gene knockout strains appearing in growth-coupled strains identified by the constraint-based model. 
Error bars indicate the average of three replicates.}
\label{fig_FACS_all}
\end{figure}

\clearpage

% Figure 4
\begin{figure}\centering
\includegraphics[width=1.0\textwidth]{./figures/fig4_western_gnd.pdf}
\caption{Western blot analysis of glycosylation efficiency in \textit{gnd} mutant. 
scFvR4 protein expression was induced during exponential growth phase. 
Time points indicate time after induction. 
\textbf{(A)} Relative protein expression over time. 
Striped bars indicate the portion of glycosylated protein. 
\textbf{(B)} Glycosylation efficiencies for each strain over time. 
Blot intensities were determined using image analysis software called ImageJ \cite{2012_schneider_eliceiri_NatMeth}.}
\label{fig_western_CLM24_gnd}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TABLES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage

% Table 1
\begin{center}
\tablefirsthead{
\hline
\rowcolor[gray]{0.8} Strain & Strain & Growth Rate & Glycan Flux \\
\rowcolor[gray]{0.8} Name & Genotype & (/hr) & (mmol/gDCW/hr) \\
\hline}
\topcaption{Wild-type along with growth-coupled strains producing \textit{C. jejuni} glycan identified by FBA and heuristic optimization. 
Simulations were performed with and without the transcription factor regulatory network. 
Knockouts listing multiple genes indicate that knockout of any one of those genes produces the same phenotype in the model. \\}
\label{tbl_sims_strains_KO}
\begin{scriptsize}
\begin{supertabular}{l l l l}
\\
Wild-type  &  & 0.78 & 0.0 \\
EcM1  & $\Delta$sdh $\Delta$zwf/pgl/gnd & 0.65 & 0.012 \\
EcM2  & $\Delta$sdh $\Delta$zwf/pgl/gnd $\Delta$pta $\Delta$eutD & 0.53 & 0.098 \\ 
EcM3  & $\Delta$sdh $\Delta$zwf/pgl/gnd $\Delta$pykAF $\Delta$mdh & 0.64 & 0.016 \\
EcM4$^\dagger$  & $\Delta$sdh $\Delta$zwf/pgl/gnd $\Delta$purU $\Delta$xdh/allABC & 0.65 & 0.007 \\
EcM5$^\dagger$  & $\Delta$sdh $\Delta$zwf/pgl/gnd $\Delta$purU $\Delta$xdh/allABC $\Delta$pta $\Delta$eutD & 0.58 & 0.055 \\
\\ \hline
  & $\dagger$ Strains without transcription factor regulation &  &  \\
\end{supertabular}
\end{scriptsize}
\end{center}

\end{document}

### 2015-02-03 revision_v2/Paper_v3.tex
second rebuttal - new paragraph

--- 2014-12-17 revision/Paper_v3.tex
@@
-\documentclass[12pt]{article}
-% Load packages
+%  LaTeX support: latex@mdpi.com
+%  In case you need support, please attach any log files that you could have, and specify the details of your LaTeX setup (which operating system and LaTeX version / tools you are using).
+
+%=================================================================
+
+% LaTeX Class File and Rendering Mode (choose one)
+% You will need to save the "mdpi.cls" and "mdpi.bst" files into the same folder as this template file.
+
+%=================================================================
+
+\documentclass[processes,article,accept,moreauthors,pdftex,12pt,a4paper]{mdpi}
+
 \usepackage{url}  % Formatting web addresses  
-\usepackage{ifthen}  % Conditional 
+%\usepackage{ifthen}  % Conditional 
 \usepackage{multicol}   %Columns
-\usepackage[utf8]{inputenc} %unicode support
+%\usepackage[utf8]{inputenc} %unicode support
 \usepackage{amsmath}
@@
 \usepackage{graphicx}
-\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
-\usepackage{setspace}
+\usepackage{soul}
+
+%\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
+%\usepackage{setspace}
 %\usepackage{longtable}
-\usepackage{colortbl}
+%\usepackage{colortbl}
 %\usepackage{palatino,lettrine}
-%\usepackage{times}
+\usepackage{times}
 %\usepackage[applemac]{inputenc} %applemac support if unicode package fails
 %\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
-\usepackage[wide]{sidecap}
+%\usepackage[wide]{sidecap}
 %\usepackage[authoryear,round,comma,sort&compress]{natbib}
-\usepackage[square,sort&compress,comma,numbers]{natbib}
+%\usepackage[square,sort&compress,comma,numbers]{natbib}
 %\usepackage[authoryear,round]{natbib}
@@
 \usepackage{comment}
-\usepackage{lineno}
+%\usepackage{lineno}
+%
+%\urlstyle{rm}
 
-\urlstyle{rm}
-
+\setcounter{page}{1}
+\lastpage{x}
+\doinum{10.3390/------}
+\pubvolume{xx}
+\pubyear{2014}
+\history{Received: xx / Accepted: xx / \\
+Published: xx}
 %\textwidth = 6.50 in
@@
 %\linespread{1.75}
-\doublespace
+%\doublespace
 
 %\usepackage{geometry}
-\usepackage{fullpage}
+%\usepackage{fullpage}
 
 %\bibliographystyle{plain}
-\bibliographystyle{plos2009}
+%\bibliographystyle{plos2009}
 
-\makeatletter
-\renewcommand\subsection{\@startsection
-	{subsection}{2}{0mm}
-	{-0.05in}
-	{-0.5\baselineskip}
-	{\normalfont\normalsize\bfseries}}
-\renewcommand\subsubsection{\@startsection
-	{subsubsection}{2}{0mm}
-	{-0.05in}
-	{-0.5\baselineskip}
-	{\normalfont\normalsize\itshape}}
-\renewcommand\section{\@startsection
-	{subsection}{2}{0mm}
-	{-0.2in}
-	{0.05\baselineskip}
-	{\normalfont\large\bfseries}}	
-\renewcommand\paragraph{\@startsection
-	{paragraph}{2}{0mm}
-	{-0.05in}
-	{-0.5\baselineskip}
-	{\normalfont\normalsize\itshape}}
-\makeatother
+%\makeatletter
+%\renewcommand\subsection{\@startsection
+%	{subsection}{2}{0mm}
+%	{-0.05in}
+%	{-0.5\baselineskip}
+%	{\normalfont\normalsize\bfseries}}
+%\renewcommand\subsubsection{\@startsection
+%	{subsubsection}{2}{0mm}
+%	{-0.05in}
+%	{-0.5\baselineskip}
+%	{\normalfont\normalsize\itshape}}
+%\renewcommand\section{\@startsection
+%	{subsection}{2}{0mm}
+%	{-0.2in}
+%	{0.05\baselineskip}
+%	{\normalfont\large\bfseries}}	
+%\renewcommand\paragraph{\@startsection
+%	{paragraph}{2}{0mm}
+%	{-0.05in}
+%	{-0.5\baselineskip}
+%	{\normalfont\normalsize\itshape}}
+%\makeatother
 
@@
 % Single space'd bib -
-\setlength\bibsep{0pt}
+%\setlength\bibsep{0pt}
 
-\renewcommand{\rmdefault}{phv}\renewcommand{\sfdefault}{phv}
+%\renewcommand{\rmdefault}{phv}\renewcommand{\sfdefault}{phv}
 \newcommand{\norm}[1]{\left\lVert#1\right\rVert}
@@
 % Change the number format in the ref list -
-\renewcommand{\bibnumfmt}[1]{#1.}
+%\renewcommand{\bibnumfmt}[1]{#1.}
 
-% Change Figure to Fig.
-\renewcommand{\figurename}{Fig.}
+% Change Figure to Figure
+%\renewcommand{\figurename}{Figure}
+%--------------------
+% Class Options:
+%--------------------
+% journal
+%----------
+% Choose between the following MDPI journals:
+% actuators, administrativesciences, aerospace, agriculture, agronomy, algorithms, animals, antibiotics, antibodies, antioxidants, appliedsciences, arts, atmosphere, atoms, axioms, batteries, behavioralsciences, bioengineering, biology, biomedicines, biomolecules, biosensors, brainsciences, buildings, cancers, catalysts, cells, challenges, chemosensors, children, chromatography, climate, coatings, computation, computers, cosmetics, crystals, dentistryjournal, diagnostics, diseases, diversity, econometrics, economies, education, electronics, energies, entropy, environmentalsciences, environments, epigenomes, fibers, foods, forests, futureinternet, galaxies, games, gels, genes, geosciences, healthcare, horticulturae, humanities, hydrology, informatics, information, inorganics, insects, ijerph, ijfs, ijms, ijgi, jcdd, jcm, jdb, jfb, jof, joi, jlpea, jmse, jpm, jrfm, jsan, land, laws, life, lubricants, machines, marinedrugs, materials, mathematics, medicalsciences, membranes, metabolites, metals, microarrays, micromachines, microorganisms, minerals, molbank, molecules, nanomaterials, ncrna, nutrients, pathogens, pharmaceuticals, pharmaceutics, pharmacy, photonics, plants, polymers, processes, proteomes, publications, religions, remotesensing, resources, risks, robotics, safety, sensors, sinusitis, socialsciences, societies, sports, standards, sustainability, symmetry, systems, technologies, toxics, toxins, universe, vaccines, veterinarysciences, viruses, water
+%---------
+% article
+%---------
+% The default type of manuscript is article, but could be replaced by using one of the class options: 
+% article, review, communication, commentary, bookreview, correction, addendum, editorial, changes, supfile, casereport, comment, conceptpaper, conferencereport, meetingreport, discussion, essay, letter, newbookreceived, opinion, projectreport, reply, retraction, shortnote, technicalnote, creative
+%----------
+% submit
+%----------
+% The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted.
+This will only make changes to the frontpage (e.g.
+the logo of the journal will get visible), the headings, and the copyright information.
+Journal info and pagination for accepted papers will also be assigned by the Editorial Office.
+% Please insert a blank line is before and after all equation and eqnarray environments to ensure proper line numbering when option submit is chosen
+%------------------
+% moreauthors
+%------------------
+% If there is only one author the class option oneauthor should be used. Otherwise use the class option moreauthors.
+%---------
+% pdftex
+%---------
+% The option "pdftex" is for use with pdfLaTeX only. If eps figure are used, use the optioin "dvipdfm", with LaTeX and dvi2pdf only.
 
-% Begin ...
-\begin{document}
-\begin{titlepage}
-{\par\centering\textbf{\Large Dynamic Modeling of Cell-Free Biochemical Networks using Effective Kinetic Models}}
-\vspace{0.05in}
-{\par \centering \large{Joseph A. Wayman, Adithya Sagar and Jeffrey D. Varner$^{*}$}}
-\vspace{0.10in}
-{\par \centering \large{School of Chemical and Biomolecular Engineering}}
-{\par \centering \large{Cornell University, Ithaca NY 14853}}
-\vspace{0.1in}
-{\par \centering \textbf{Running Title:}~Effective models of metabolism}
-\vspace{0.1in}
-{\par \centering \textbf{To be submitted:}~\textit{Processes}}
-\vspace{0.5in}
-{\par \centering $^{*}$Corresponding author:}
-{\par \centering Jeffrey D. Varner,}
-{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
-{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
-{\par \centering Email: jdv27@cornell.edu} 
-{\par \centering Phone: (607) 255 - 4258}
-{\par \centering Fax: (607) 255 - 9166}
-\end{titlepage}
-\date{}
-\thispagestyle{empty}
-\pagebreak
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Abstract}
+%=================================================================
+
+%------------------------------------------------------------------
+% The following line should be uncommented if the LaTeX file is uploaded to arXiv.org
+%\pdfoutput=1
+
+%=================================================================
+
+% Add packages and commands to include here
+% The amsmath, amsthm, amssymb, hyperref, caption, float and color packages are loaded by the MDPI class.
+%\usepackage{graphicx}
+%\usepackage{subfigure,psfig}
+
+%=================================================================
+%% Please use the following mathematics environments:
+%\theoremstyle{mdpi}
+%\newcounter{thm}
+%\setcounter{thm}{0}
+%\newcounter{ex}
+%\setcounter{ex}{0}
+%\newcounter{re}
+%\setcounter{re}{0}
+%\newtheorem{Theorem}[thm]{Theorem}
+%\newtheorem{Lemma}[thm]{Lemma}
+%\newtheorem{Characterization}[thm]{Characterization}
+%\newtheorem{Proposition}[thm]{Proposition}
+%\newtheorem{Property}[thm]{Property}
+%\newtheorem{Problem}[thm]{Problem}
+%\newtheorem{Example}[ex]{Example}
+%\newtheorem{Remark}[re]{Remark}
+%\newtheorem{Corollary}[thm]{Corollary}
+%\newtheorem{Definition}[thm]{Definition}
+%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).
+
+%=================================================================
+
+% Full title of the paper (Capitalized)
+\Title{Dynamic Modeling of Cell-Free Biochemical Networks using Effective Kinetic Models}
+
+% Authors (Add full first names)
+\Author{Joseph A. Wayman, Adithya Sagar and Jeffrey D. Varner {*}}
+
+% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
+\address[1]{%
+School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY 14853, USA\\ \vspace{12pt}
+
+Running Title:~Effective models of metabolism}
+
+% Contact information of the corresponding author (Add [2] after \corres if there are more than one corresponding author.)
+\corres{E-Mail: jdv27@cornell.edu; \mbox{Tel.: +607-255-4258;} Fax: +607-255-9166.}
+
+\abstract{
 Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
@@
 However, when starting with an initial population that was heavily enriched with incorrect structures, our particle swarm approach could converge to an incorrect structure.
-While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.
+While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.}
 
-{\noindent \textbf{Keywords:}~Cell-free metabolism, Mathematical modeling}
+% Keywords: add 3 to 10 keywords
+\keyword{Cell-free metabolism, Mathematical modeling}
 
-\pagebreak
+% The fields PACS, MSC, and JEL may be left empty or commented out if not applicable
+%\PACS{}
+%\MSC{}
+%\JEL{}
 
-\setcounter{page}{1}
+% If this is an expanded version of a conference paper, please cite it here: enter the full citation of your conference paper, and add $^\dagger$ in the end of the title of this article.
+%\conference{}
 
-\linenumbers
+\begin{document}
 
-\section*{Introduction}
+\section{Introduction}
+
 Mathematical modeling has long contributed to our understanding of metabolism.
@@
 
-\clearpage
+\newpage
 
-\section*{Results}
+\section{Results}\vspace{-12pt}
 
-\subsection*{Formulation and properties of effective cell-free metabolic models.}
+\subsection{Formulation and properties of effective cell-free metabolic models.}
+
 We developed two proof-of-concept metabolic networks to investigate the features of our effective biochemical network modeling approach (Fig. \ref{fig-networks}). 
@@
 
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{./figs/Figure-1-Networks.pdf}
+\caption{Proof-of-concept cell-free metabolic networks considered in this study. Substrate $S$ is converted to products P$_{1}$ and P$_{2}$ through a series of chemical conversions
+catalyzed by enzyme(s) $E_{j}$. The activity of the pathway enzymes is subject to both positive and negative allosteric regulation.}\label{fig-networks}
+\end{figure}
+
 Allosteric regulation of enzyme activity was modeled by combining individual regulatory contributions to the activity of pathway enzymes into a control coefficient using an integration rule (Fig.
@@
 Thus, our test networks involved important features such as cofactor recycling, enzyme activity and metabolite dynamics, as well as multiple overlapping allosteric regulatory mechanisms.  
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{./figs/Figure-2-ControlSchematic.pdf}
+\caption{Schematic of rule-based allosteric enzyme activity control laws. 
+Traditional enzyme kinetic expressions, e.g., Michaelis–Menten or multiple saturation kinetics, are multiplied by an enzyme activity control variable $0 \leq v_{j} \leq 1 $. 
+Control variables are functions of many possible regulatory factors encoded by arbitrary functions of the form $0\leq f_{j}\left(\mathcal{Z}\right)\leq 1$.
+At each simulation time step, the $v_{j}$ variables are calculated by evaluating integration rules such as the max or min of the set of factors $f_{1},\hdots$ 
+influencing the activity of enzyme $E_{j}$. }\label{fig-control-schematic}
+\end{figure}
 
@@
 
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure-3-EnzymeKinetics.pdf}
+\caption{Kinetics of simple transformations in the presence of activation and inhibition. 
+\textbf{A}:The conversion of substrate $S$ to product $P$ by enzyme $E$ was activated by $S$. 
+For a fixed control gain parameter $\kappa_{control}$, the reaction rate approached a step for increasing cooperativity control parameter $\eta$. 
+For activation simulations $\kappa_{control}=$0.05 and $\eta=\left\{0.01,0.1,1,2,4,6,8,10\right\}$.
+\textbf{B}:The conversion of substrate $S$ to product $P$ by enzyme $E$ with inhibitor $I$. 
+For a fixed control gain parameter $\kappa_{control}$, the reaction rate approximated non-competitive inhibition for increasing cooperativity control parameter $\eta$. 
+For the inhibition simulations $\kappa_{control}=$1.5 and $\eta=\left\{0.01,0.1,1,2,4,6,8,10\right\}$.}\label{fig-kinetics-simulations}
+\end{figure}
+
 End product yield was controlled by feedback inhibition, while product selectivity was controlled by branch point enzyme inhibition (Fig. \ref{fig-onoff-simulations}). 
@@
 
-\subsection*{Estimating parameters and effective allosteric regulatory structures.}
+\subsection{Estimating parameters and effective allosteric regulatory structures.}
+
 A critical challenge for any dynamic model is the estimation of kinetic parameters. 
@@
 Taken together, these results suggested that we could perhaps simultaneously estimate both parameters and network control architectures, as incorrect control structures would be manifest as poor model fits.
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.450\textwidth]{./figs/Figure-4-OnOffSimulations.pdf}
+\caption{ON/OFF control simulations for Network A and Network B for an ensemble of 100 kinetic parameter sets versus time. 
+For each case, simulations were conducted using kinetic and initial conditions generated randomly from a hypothetical true parameter set. 
+The gray area represents $\pm$ one standard deviation surrounding the mean. 
+Control parameters were fixed during the ensemble calculations.
+\textbf{A}: End product P$_{1}$ abundance versus time for Network A. 
+The abundance of P$_{1}$ decreased with end product inhibition of $E_{1}$ activity (Control-ON) versus the no inhibition case (Control-OFF). 
+\textbf{B}: End product P$_{2}$ abundance versus time for Network B. Inhibition of branch point $E_{6}$ by end product P$_{1}$ decreased P$_{2}$ abundance (Control-ON) versus the
+no inhibition case (Control-OFF).
+\textbf{C}: End product P$_{1}$ abundance versus time for Network A. 
+Inhibition of branch point $E_{6}$ by end product P$_{1}$ decreased P$_{1}$ abundance (Control-ON) versus the no inhibition case (Control-OFF).}\label{fig-onoff-simulations}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure-5-ParameterFit.pdf}
+\caption{Parameter estimation from synthetic data for the same and mismatched allosteric control logic using particle swarm optimization (PSO). 
+Synthetic experimental data was generated from a hypothetical parameter set using Network A, 
+where substrate $S$, end product P$_{1}$ and intermediate M$_5$ were sampled approximately every 20 minutes. 
+For cases $\textbf{A,B}$ 20 particles were initialized with randomized parameters and allowed to search for 300 iterations. 
+\textbf{A,B}: PSO estimated an ensemble of 20 parameters sets consistent with the synthetic experimental data assuming the correct
+enzymatic and control connectivity starting from randomized initial parameters.
+\textbf{C,D}: In the presence of control mismatch (Network B control policy simulated with Network A kinetic parameters) 
+the ensemble of models did not describe the synthetic data. The synthetic data plotted here was unperturbed by noise. 
+However, we assumed a constant coefficient of variation of 10\% for the synthetic data during parameter estimation.}\label{fig-parameter-fit}
+\end{figure}
 
@@
 This suggested that our particle swarm approach identified a \textit{sloppy} ensemble, i.e., parameter estimates that were individually incorrect but collectively exhibited the correct model behavior.
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure-6-AlternativeNetworks.pdf}
+\caption{Schematic of the alternative allosteric control programs used in the structural particle swarm computation. 
+Each network had the same enzymatic connectivity, initial conditions and kinetic parameters, 
+but alternative feedback control structures for the first enzyme in the pathway.}\label{fig-alternative-networks}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.47\textwidth]{./figs/Figure-7-ControlSearch.pdf}
+\caption{Combined control and kinetic parameter search using modified particle swarm optimization (PSO). 
+A population of 100 particles was initialized with randomized kinetic parameters and one of five possible control configurations (Network A - E).
+Simulation error was minimized for a synthetic data set 
+($S$, end product P$_{1}$ and intermediate M$_5$ sampled approximately every 20 min) generated using Network A.
+\textbf{A}: Simulation error versus parameter set angle for 100 particles biased toward the correct regulatory program (A,B,C,D,E) = (40\%, 10\%, 20\%, 20\% and 10\%).
+\textbf{B}: Simulation error versus parameter set angle for 100 uniformly distributed particles (A,B,C,D,E) = (20\%, 20\%, 20\%, 20\% and 20\%).
+\textbf{C}: Simulation error versus parameter set angle for 100 negatively biased particles (A,B,C,D,E) = (10\%, 40\%, 10\%, 20\% and 20\%).
+Network A (the correct structure) was preferentially identified for positively and uniform biased particle distributions, but misidentified in the presence of
+a large incorrect bias. }\label{fig-control-search}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.0\textwidth,height=0.6\textheight]{./figs/Figure-8-Flux.pdf}
+\caption{Metabolic flux and control variables as a function of network type and particle index at t = 100 min.
+The particle error, the control variables governing $E_{1},E_{3}$ and $E_{6}$ activity ($v_{1},v_{3}$ and $v_{3}$) and the scaled metabolic flux were calculated for the positively (top), 
+uniformly (middle) and negatively (bottom) biased particle swarms (N = 100). 
+Blue denotes a low value, while red denotes a high value for the respective quantity being plotted.
+The particles from each swarm were sorted based upon simulation error (low to high error). 
+\textbf{A}: Model performance for the positively biased particle swarm as a function of particle index.
+\textbf{B}: Model performance for the uniformly biased particle swarm as a function of particle index.
+\textbf{C}: Model performance for the negatively biased particle swarm as a function of particle index. 
+Models with significant control mismatch showed distinct control and flux patterns versus those models with the correct or closely related control policies.
+In particular, models with the correct control policy showed stronger inhibition of $E_{1}$ activity, leading to decreased flux from S$\rightarrow$P$_{1}$.
+Conversely, models with significant mismatch had increased $E_{1}$ activity, leading to an altered flux distribution. 
+This is especially apparent in the negatively biased particle swarm. }\label{fig-flux-pattern}
+\end{figure}
 
@@
 
-\clearpage
+\section{Discussion}
 
-\section*{Discussion}
 In this study, we presented an effective kinetic modeling strategy to dynamically simulate cell-free biochemical networks. 
@@
 
+The proposed modeling framework also differs appreciably from previously established non-mechanistic, kinetic approximations of complex biochemical network behavior. 
+Such frameworks replace parameter dense mechanistic kinetic expressions with heuristics quantifying the relationship between metabolic rate and metabolite effectors. 
+These approaches arose in response to uncertainties associated with obtaining correct mechanistic kinetic expressions and parameters of \textit{in vivo} systems. 
+Similarly, available kinetic parameters measured \textit{in vitro} may differ in a specialized cell-free \textit{in vitro} environments. 
+Factors affecting kinetics, such as enzyme channeling, macromolecular crowding, and pH, are likely dramatically different in cell-free environments than in both \textit{in vivo} systems as well as typical \textit{in vitro} conditions used for parameter measurements.
+Thus, a more generalized, approximate biochemical reaction network formulation may be desirable in the case of cell-free systems. 
+Our approach is similar to generalized mass action-based power law formulations of Savageau and colleagues \cite{1991_savageau_JTheorBio} and linlog kinetics of Visser \textit{et al.} \cite{2003_visser_heijnen_MetabEng} in that metabolic rates are proportional to corresponding enzyme levels modified by metabolite effectors.
+Power law and linlog approaches suffer from several known limitations \cite{2005_heijnen_BiotechBioeng}. 
+Power law reaction rates do not capture saturation effects and become infinite for small concentrations of inhibitory effectors. 
+Linlog kinetics also become ill-defined when effector concentrations go to zero. 
+Our framework does not suffer from such drawbacks. 
+Also, models employing linlog kinetics typically rely on an experimentally determined reference state to describe dynamics taking place after a perturbation to a steady-state. 
+Our framework is not restricted to such a requirement. 
+Cell-free systems are unlikely to satisfy such a steady-state approximation after extract preparation and prior to culture initiation. 
+Our framework is similar to the generic kinetic formulation from Hadlich \textit{et al.} \cite{2009_hadlich_wiechert_MetabEng}, but differs in its inclusion of cooperative effects as well as proposes a simplified integration of competition amongst allosteric effectors using max/min rules.
+In summary, our proposed framework offers an effective kinetic approximation that captures saturation effects and allosteric competition within cell-free systems that may also be extensible to \textit{in vivo} metabolic and gene regulatory networks.
+
 There are several critical questions that should be explored following this proof-of-concept study.
@@
 We have also demonstrated \textit{sloppy} behavior in a wide variety of signal transduction processes \citep{Luan:2007aa,Song:2009aa,Tasseff:2010aa,Tasseff:2011aa,Nayak:2011aa,Lequieu:2011aa}.
-Thus, given our previous experience with models containing hundreds of unknown parameters, we expect parameter estimation to be a manageable challenge.
+Thus, given our previous experience with models containing hundreds of unknown parameters, we expect parameter estimation to be a manageable challenge assuming 
+we have good quality experimental data.
 
-A critical challenge will be the estimation of allosteric connectivity at a genome scale. 
+A second critical challenge will be the estimation of allosteric connectivity at a genome scale. 
 The regulation of glycolytic enzymes, such as phosphofructokinase I, has been studied for many years \citep{BERG_BIOCHEMISTRY,Peskov:2008aa}. 
@@
 However, for those enzymes that have not been well studied, we will need to infer allosteric interactions from experimental data.
-In general, the reverse engineering of regulatory network structure from data is a very difficult problem. 
+In general, the reverse engineering of regulatory network structure from data is a difficult problem. 
 Recently, Sauer and colleagues have developed a systematic, model-based approach for the identification of allosteric regulation \textit{in vivo} \cite{Link:2013aa}. 
@@
 The particular form of the transfer function could also be explored. We choose a Hill-like function because of its 
-prominence in the systems and synthetic biology community. However, the only mathematical requirement for a transfer function is that it map a non-negative continuous or categorical variable into the 
-range $f\in[0,1]$. Thus, many types of transfer functions are possible.
+prominence in the systems and synthetic biology community.
+However, the only mathematical requirement for a transfer function is that it map a non-negative continuous or categorical variable into the range $f\in[0,1]$.
+Thus, many types of transfer functions are possible.
 
-\clearpage
+\section{Materials and Methods}\vspace{-12pt}
 
-\section*{Materials and Methods}
+\subsection{Formulation and solution of the model equations.}
 
-\subsection*{Formulation and solution of the model equations.}
 We used ordinary differential equations (ODEs) to model the time evolution of metabolite ($x_{i}$) and scaled enzyme abundance ($\epsilon_{i}$) in hypothetical cell-free metabolic networks:
@@
 
-\section*{Acknowledgements}
+\section*{\noindent Acknowledgments}\vspace{12pt}
+
 This study was supported by an award from the National Science Foundation (MCB \#1411715) 
@@
 %\bibliographystyle{IEEEbib}
+\bibliographystyle{mdpi}
 
@@
 %\begin{comment}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-1-Networks.pdf}
-\caption{Proof-of-concept cell-free metabolic networks considered in this study. Substrate $S$ is converted to products P$_{1}$ and P$_{2}$ through a series of chemical conversions
-catalyzed by enzyme(s) $E_{j}$. The activity of the pathway enzymes is subject to both positive and negative allosteric regulation. }\label{fig-networks}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-2-ControlSchematic.pdf}
-\caption{Schematic of rule-based allosteric enzyme activity control laws. 
-Traditional enzyme kinetic expressions, e.g., Michaelis–Menten or multiple saturation kinetics, are multiplied by an enzyme activity control variable $0 \leq v_{j} \leq 1 $. 
-Control variables are functions of many possible regulatory factors encoded by arbitrary functions of the form $0\leq f_{j}\left(\mathcal{Z}\right)\leq 1$.
-At each simulation time step, the $v_{j}$ variables are calculated by evaluating integration rules such as the max or min of the set of factors $f_{1},\hdots$ 
-influencing the activity of enzyme $E_{j}$. }\label{fig-control-schematic}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-3-EnzymeKinetics.pdf}
-\caption{Kinetics of simple transformations in the presence of activation and inhibition. 
-\textbf{A}:The conversion of substrate $S$ to product $P$ by enzyme $E$ was activated by $S$. 
-For a fixed control gain parameter $\kappa_{control}$, the reaction rate approached a step for increasing cooperativity control parameter $\eta$. 
-For activation simulations $\kappa_{control}=$0.05 and $\eta=\left\{0.01,0.1,1,2,4,6,8,10\right\}$.
-\textbf{B}:The conversion of substrate $S$ to product $P$ by enzyme $E$ with inhibitor $I$. 
-For a fixed control gain parameter $\kappa_{control}$, the reaction rate approximated non-competitive inhibition for increasing cooperativity control parameter $\eta$. 
-For the inhibition simulations $\kappa_{control}=$1.5 and $\eta=\left\{0.01,0.1,1,2,4,6,8,10\right\}$.}\label{fig-kinetics-simulations}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=0.50\textwidth]{./figs/Figure-4-OnOffSimulations.pdf}
-\caption{ON/OFF control simulations for Network A and Network B for an ensemble of 100 kinetic parameter sets versus time. 
-For each case, simulations were conducted using kinetic and initial conditions generated randomly from a hypothetical true parameter set. 
-The gray area represents $\pm$ one standard deviation surrounding the mean. 
-Control parameters were fixed during the ensemble calculations.
-\textbf{A}: End product P$_{1}$ abundance versus time for Network A. 
-The abundance of P$_{1}$ decreased with end product inhibition of $E_{1}$ activity (Control-ON) versus the no inhibition case (Control-OFF). 
-\textbf{B}: End product P$_{2}$ abundance versus time for Network B. Inhibition of branch point $E_{6}$ by end product P$_{1}$ decreased P$_{2}$ abundance (Control-ON) versus the
-no inhibition case (Control-OFF).
-\textbf{C}: End product P$_{1}$ abundance versus time for Network A. 
-Inhibition of branch point $E_{6}$ by end product P$_{1}$ decreased P$_{1}$ abundance (Control-ON) versus the no inhibition case (Control-OFF).
-}\label{fig-onoff-simulations}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-5-ParameterFit.pdf}
-\caption{Parameter estimation from synthetic data for the same and mismatched allosteric control logic using particle swarm optimization (PSO). 
-Synthetic experimental data was generated from a hypothetical parameter set using Network A, 
-where substrate $S$, end product P$_{1}$ and intermediate M$_5$ were sampled approximately every 20 minutes. 
-For cases $\textbf{A,B}$ 20 particles were initialized with randomized parameters and allowed to search for 300 iterations. 
-\textbf{A,B}: PSO estimated an ensemble of 20 parameters sets consistent with the synthetic experimental data assuming the correct
-enzymatic and control connectivity starting from randomized initial parameters.
-\textbf{C,D}: In the presence of control mismatch (Network B control policy simulated with Network A kinetic parameters) 
-the ensemble of models did not describe the synthetic data. The synthetic data plotted here was unperturbed by noise. 
-However, we assumed a coefficient of variation of 10\% for the synthetic data during parameter estimation.
-}\label{fig-parameter-fit}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-6-AlternativeNetworks.pdf}
-\caption{Schematic of the alternative allosteric control programs used in the structural particle swarm computation. 
-Each network had the same enzymatic connectivity, initial conditions and kinetic parameters, 
-but alternative feedback control structures for the first enzyme in the pathway.}\label{fig-alternative-networks}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=0.5\textwidth]{./figs/Figure-7-ControlSearch.pdf}
-\caption{Combined control and kinetic parameter search using modified particle swarm optimization (PSO). 
-A population of 100 particles was initialized with randomized kinetic parameters and one of five possible control configurations (Network A - E).
-Simulation error was minimized for a synthetic data set 
-($S$, end product P$_{1}$ and intermediate M$_5$ sampled approximately every 20 min) generated using Network A.
-\textbf{A}: Simulation error versus parameter set angle for 100 particles biased toward the correct regulatory program (A,B,C,D,E) = (40\%, 10\%, 20\%, 20\% and 10\%).
-\textbf{B}: Simulation error versus parameter set angle for 100 uniformly distributed particles (A,B,C,D,E) = (20\%, 20\%, 20\%, 20\% and 20\%).
-\textbf{C}: Simulation error versus parameter set angle for 100 negatively biased particles (A,B,C,D,E) = (10\%, 40\%, 10\%, 20\% and 20\%).
-Network A (the correct structure) was preferentially identified for positively and uniform biased particle distributions, but misidentified in the presence of
-a large incorrect bias. 
-}\label{fig-control-search}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth,height=0.6\textheight]{./figs/Figure-8-Flux.pdf}
-\caption{Metabolic flux and control variables as a function of network type and particle index at t = 100 min.
-The particle error, the control variables governing $E_{1},E_{3}$ and $E_{6}$ activity ($v_{1},v_{3}$ and $v_{3}$) and the scaled metabolic flux were calculated for the positively (top), 
-uniformly (middle) and negatively (bottom) biased particle swarms (N = 100). 
-Blue denotes a low value, while red denotes a high value for the respective quantity being plotted.
-The particles from each swarm were sorted based upon simulation error (low to high error). 
-\textbf{A}: Model performance for the positively biased particle swarm as a function of particle index.
-\textbf{B}: Model performance for the uniformly biased particle swarm as a function of particle index.
-\textbf{C}: Model performance for the negatively biased particle swarm as a function of particle index. 
-Models with significant control mismatch showed distinct control and flux patterns versus those models with the correct or closely related control policies.
-In particular, models with the correct control policy showed stronger inhibition of $E_{1}$ activity, leading to decreased flux from S$\rightarrow$P$_{1}$.
-Conversely, models with significant mismatch had increased $E_{1}$ activity, leading to an altered flux distribution. 
-This is especially apparent in the negatively biased particle swarm. 
-}\label{fig-flux-pattern}
-\end{figure}
-
-\clearpage
 

### 2015-02-06 processes-69103_review_response/processes-69103.tex
commit changes

@@
 
-Each reaction rate was written as the product of two terms, a kinetic term ($\bar{r}_{j}$) and a regulatory term ($v_{j}$):
+Each reaction rate was written as the product of two terms, a kinetic term ($\bar{r}_{j}$) and a control term ($v_{j}$) that depends on multiple regulators:
 \begin{equation}\label{eqn:rate-factor}
@@
 	v_{j} = 1 - \max\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{mj}\left(\mathcal{Z}\right)\right)
-\end{equation}Lastly, if an enzyme had both $m$ activating and $n$ inhibitory factors, we modeled the regulatory term as:
+\end{equation}Lastly, if an enzyme had both $m$ activating and $n$ inhibitory factors, we modeled the control term as:
 \begin{equation}

### 2015-02-07 processes-69103_review_response/processes-69103.tex
Changes to tex
Changes

@@
 who developed and then subsequently refined highly mechanistic coagulation models.
-Other laboratories have also expanded upon Hockin \emph{et al.}, for example by exploring the intrinsic pathway, the role of stochastic fluctuations in coagulation \citep{Lo:2005aa},
-and the dynamics of thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
+Recently Papadopoulos and co-workers used a phenomenological  mathematical model for thrombin generation \cite{Atkin:2014}.
+Using a set of four ordinary differential equations they were able to derive an equation for temporal evolution of thrombin generation.
+However the model focuses more on thrombin generation rather than dynamics of coagulation.
+It lacks any detail regarding regulatory activity and is unable to capture the roles of primary inhibitors like ATIII or protein C as the hybrid approach captures.The model also requires that the parameters are adjusted or fine tuned for each data set.
+This makes the model unfeasible in cases where we need to understand the dynamics of coagulation without complex modeling approaches or when there is incomplete mechanistic knowledge.
+Other laboratories have also expanded upon Hockin \emph{et al.}, for example by exploring the intrinsic pathway, the role of stochastic fluctuations in coagulation \citep{Lo:2005aa}, and the dynamics of thrombin mediated clot formation \citep{Chatterjee:2010aa}.
 Other aspects of coagulation have also been modeled, such as platelet biochemistry \citep{Stalker:2013aa}, 
@@
 (A,B,C) depict training data and results for 150\%, 100\% and 50\% of physiological prothrombin levels in the absence of protein C pathway.
-The experimental training data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}.
+The experimental training data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}\citep{Veer:1997}.
 Thrombin generation in these experiments was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
@@
 (A,B,C) depict training data and results for 150\%, 100\% and 50\% of physiological prothrombin levels in the presence of the protein C pathway. 
-The experimental training data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}.
+The experimental training data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}\citep{Veer:1997}.
 Thrombin generation in these experiments was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
@@
 (A,B,C,D) prediction results for (FII,ATIII): (50\%,150\%), (100\%, 100\%), (125\%, 75\%) and (150\%, 50\%) of physiological prothrombin and ATIII levels in the absence of the protein C pathway.
-The experimental validation data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}.
+The experimental validation data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}\citep{Veer:1997}.
 Thrombin generation in these experiments was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
@@
 \begin{equation}\label{eqn:control-factor}
-	f_{i}\left(\mathcal{Z}_{j}\right) = \frac{\kappa_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}{1 + \kappa_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}
-\end{equation}where $\mathcal{Z}_{j}$ denotes the abundance of the $j$ factor (e.g., metabolite abundance), and $\kappa_{ij}$ and $\eta$ are control parameters. 
-$\kappa_{ij}$ was the species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).
+	f_{i}\left(\mathcal{Z}_{j},k_{ij}\right) = \frac{k_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}{1 + k_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}
+\end{equation}where $\mathcal{Z}_{j}$ denotes the abundance of the $j$ factor (e.g., metabolite abundance), and $k_{ij}$ and $\eta$ are control parameters. 
+$k_{ij}$ was the species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).
 Applying the general framework to the reduced coagulation network resulted in five ordinary differential equations:
@@
 estimates was measured using goodness of fit (model residual). The particle swarm optimization routine was implemented in the Python programming language. 
-All plots were made using the Matplotlib module of Python \citep{Matplotlib}.
+All plots were made using the Matplotlib module of Python \citep{Matplotlib}.\newline
 %The optimization and plotting code can be downloaded from http://www.varnerlab.org.
+%Experimental data,simulation code,parameter ensembles and initial conditions for each experimental simulation can be viewed and downloaded from https://github.com/jeffreyvarner/HybridCoagulationModel\_v1.
 
@@
 
+\bibitem{Atkin:2014}
+Papadopoulos, Konstantinos and Gavaises, Manolis and Atkin,Chris (2014) A simplified mathematical model for thrombin generation.
+\newblock Medical Engineering \& Physics 36: 196 - 204.
+
+\bibitem{Veer:1997}
+Cornelis van't Veer, Neal J. Golden, Michael Kalafatis and Kenneth G. Mann (1997) Inhibitory Mechanism of the Protein C Pathway on Tissue Factor-induced Thrombin Generation.
+\newblock The Journal of Biological Chemistry 272: 7963-7994.
+
 \end{thebibliography}
@@
 
+\bibliography{References_v1.bib}
 \end{document}

### 2015-02-07 processes-69103_review_response/processes-69103.tex
Final changes

@@
 However the model focuses more on thrombin generation rather than dynamics of coagulation.
-It lacks any detail regarding regulatory activity and is unable to capture the roles of primary inhibitors like ATIII or protein C as the hybrid approach captures.The model also requires that the parameters are adjusted or fine tuned for each data set.
-This makes the model unfeasible in cases where we need to understand the dynamics of coagulation without complex modeling approaches or when there is incomplete mechanistic knowledge.
+Unlike our hybrid approach it does not consider regulatory detail and thus is unable to capture the roles of primary inhibitors like ATIII or protein C.
+Thus in cases where we need to understand the dynamics of coagulation without complex modeling or when there is incomplete mechanistic knowledge, the hybrid approach is a better candidate.
+The model also requires that the parameters are adjusted or fine tuned for each data set unlike the hybrid approach where we use a single ensemble across all experiments.
+However platelet modeling is a significant advantage of this model over the hybrid approach.
+Future versions of the hybrid model can involve modeling platelets as done by Papdopoulos et al.
+in their phenomenological model.
 Other laboratories have also expanded upon Hockin \emph{et al.}, for example by exploring the intrinsic pathway, the role of stochastic fluctuations in coagulation \citep{Lo:2005aa}, and the dynamics of thrombin mediated clot formation \citep{Chatterjee:2010aa}.
@@
 
-The performance of the proof of principle coagulation model was impressive given its limited size, however, there are several issues that could be further explored.
+The performance of the proof of principle coagulation model was impressive given its limited size.
+The most detailed mechanistic model of coagulation is the one by Luan and co-workers built  with 193 proteins and protein complexes that are interconnected by 301 reactions \citep{2010_luan_varner_MolBioSys}.
+Akin to the hybrid model this mechanistic used the normal thrombin data and hemophilia data from Allen \emph{et al.} \citep{ALLEN2006}.
+However unlike the mechanistic model we used this data for validation rather than training.
+Results from our model are surprisingly comparable to the training simulations of the mechanistic model.
+The initiation time and amplification of thrombin signal were accurately predicted in the normal case whereas in the hemophilia case we correctly predict the initiation time but slightly over predict the amplification signal.
+This performance is nearly similar to that of the detailed mechanistic model with 301 reactions.
+Despite the impressive performance, there are several issues that could be further explored.
 First, the prediction of initiation time should be investigated.

### 2015-02-08 revision_v2/Paper_v3.tex
jw - final revisions second rebuttal

@@
 Unfortunately, traditional, fully structured cybernetic models also suffer from an identifiability challenge, as both the kinetic parameters and an abstracted model of cellular objectives must be estimated simultaneously.
-However, recent cybernetic formulations from Ramkrishna and colleagues have successfully treated this identifiability challenge through elementary mode reduction, though the techniques replace detailed biological mechanism with an optimization heuristic \cite{2009_song_ramkrishna_BiotechBioeng,Song:2011aa}.
+However, recent cybernetic formulations from Ramkrishna and colleagues have successfully treated this identifiability challenge through elementary mode reduction, though the techniques replace detailed biological mechanism with a more approximative treatment of metabolism \cite{2009_song_ramkrishna_BiotechBioeng,Song:2011aa}.
 
@@
 However, Boolean rules will likely be less effective at capturing dynamic allosteric regulation in a cell-free metabolic system. 
-On the other hand, the strength of cybernetic models is the integration of optimal metabolic control heuristics with traditional kinetic pathway modeling. 
-Cybernetic models are highly predictive; they have successfully predicted mutant behavior from limited wild-type data \citep{1999_varner_ramkrishna_MetaEng,Varner:2000aa,Song:2012aa}, 
-steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, strain specific metabolic function \citep{Song:2011aa} and have been used in bioprocess control applications \citep{Gadkar:2003aa}.
-However, cybernetic control heuristics are not mechanistic, instead they are the output of an optimal decision with respect to a set of hypothetical physiological objectives. 
+On the other hand, the strength of cybernetic models is the integration of optimal metabolic control strategies with traditional kinetic pathway modeling. 
+Cybernetic models are highly predictive; they have successfully predicted mutant behavior from limited wild-type data \citep{1999_varner_ramkrishna_MetaEng,Varner:2000aa,Song:2012aa}, steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, strain specific metabolic function \citep{Song:2011aa} and have been used in bioprocess control applications \citep{Gadkar:2003aa}.
+However, cybernetic control strategies are not mechanistic, instead they are the output of an optimal decision with respect to a set of hypothetical physiological objectives. 
 Thus, they are abstractions which are difficult to translate into a specific biological mechanism. 
@@
 Such frameworks replace parameter dense mechanistic kinetic expressions with heuristics quantifying the relationship between metabolic rate and metabolite effectors. 
+A review of approximative kinetic formats can be found in \cite{2005_heijnen_BiotechBioeng}.  
 These approaches arose in response to uncertainties associated with obtaining correct mechanistic kinetic expressions and parameters of \textit{in vivo} systems. 
@@
 Power law and linlog approaches suffer from several known limitations \cite{2005_heijnen_BiotechBioeng}. 
-Power law reaction rates do not capture saturation effects and become infinite for small concentrations of inhibitory effectors. 
+Power law reaction rates do not capture saturation effects and become infinite for small concentrations of inhibitory regulators. 
 Linlog kinetics also become ill-defined when effector concentrations go to zero. 
@@
 This suggested that we can expect reasonable model predictions, despite only partial parameter knowledge, as network size grows if we have properly designed experiments. 
+Though we expect that a model's computational complexity will scale poorly with network size, we are optimistic that the construction of large-scale, predictive models of metabolism are possible. 
+There is evidence to suggest that achieving a quantitative understanding of complex biological systems should not require complete parametric knowledge. 
 Brown and Sethna showed in a model of signal transduction that good predictions were possible despite only order of magnitude estimates of parameter values \citep{Brown:2003aa}.
-Sethna and coworkers later showed that model performance is often controlled by only a few parameter combinations, 
-a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
+Sethna and coworkers later showed that model performance is often controlled by only a few parameter combinations, a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
 We have also demonstrated \textit{sloppy} behavior in a wide variety of signal transduction processes \citep{Luan:2007aa,Song:2009aa,Tasseff:2010aa,Tasseff:2011aa,Nayak:2011aa,Lequieu:2011aa}.
-Thus, given our previous experience with models containing hundreds of unknown parameters, we expect parameter estimation to be a manageable challenge assuming 
-we have good quality experimental data.
+Thus, given our previous experience with models containing hundreds of unknown parameters, we expect parameter estimation to be a manageable challenge assuming we have good quality experimental data.
 
@@
 We considered a population of 100 particles, where each particle in the swarm was assigned a model structure, and a random parameter vector. 
+The optimization simulations of Figure \ref{fig-control-search}, running on a single CPU on an Apple workstation (Apple, Cupertino, CA; OS X v10.10), required on the order of several hours for completion.
 The PSO algorithm, model equations, and the objective function were encoded and solved in the Octave programming language (v 3.8.1; www.octave.org).

### 2015-02-16 revision_v2/Paper_v3.tex
Submission of revised v2

@@
 Unfortunately, traditional, fully structured cybernetic models also suffer from an identifiability challenge, as both the kinetic parameters and an abstracted model of cellular objectives must be estimated simultaneously.
-However, recent cybernetic formulations from Ramkrishna and colleagues have successfully treated this identifiability challenge through elementary mode reduction, though the techniques replace detailed biological mechanism with a more approximative treatment of metabolism \cite{2009_song_ramkrishna_BiotechBioeng,Song:2011aa}.
+However, recent cybernetic formulations from Ramkrishna and colleagues have successfully treated this identifiability challenge through elementary mode reduction \cite{2009_song_ramkrishna_BiotechBioeng,Song:2011aa}.
 
@@
 
-The proposed modeling framework also differs appreciably from previously established non-mechanistic, kinetic approximations of complex biochemical network behavior. 
+The proposed modeling framework also differs appreciably from previously established kinetic approximations of complex biochemical network behavior. 
 Such frameworks replace parameter dense mechanistic kinetic expressions with heuristics quantifying the relationship between metabolic rate and metabolite effectors. 
@@
 Our approach is similar to generalized mass action-based power law formulations of Savageau and colleagues \cite{1991_savageau_JTheorBio} and linlog kinetics of Visser \textit{et al.} \cite{2003_visser_heijnen_MetabEng} in that metabolic rates are proportional to corresponding enzyme levels modified by metabolite effectors.
-Power law and linlog approaches suffer from several known limitations \cite{2005_heijnen_BiotechBioeng}. 
+Power law and linlog approaches suffer from several limitations \cite{2005_heijnen_BiotechBioeng}. 
 Power law reaction rates do not capture saturation effects and become infinite for small concentrations of inhibitory regulators. 
 Linlog kinetics also become ill-defined when effector concentrations go to zero. 
+Also, models employing linlog kinetics typically rely on an experimentally determined reference state to describe dynamics taking place after a perturbation to a steady-state. 
 Our framework does not suffer from such drawbacks.  
-Also, models employing linlog kinetics typically rely on an experimentally determined reference state to describe dynamics taking place after a perturbation to a steady-state. 
-Our framework is not restricted to such a requirement. 
-Cell-free systems are unlikely to satisfy such a steady-state approximation after extract preparation and prior to culture initiation. 
-Our framework is similar to the generic kinetic formulation from Hadlich \textit{et al.} \cite{2009_hadlich_wiechert_MetabEng}, but differs in its inclusion of cooperative effects as well as proposes a simplified integration of competition amongst allosteric effectors using max/min rules.
+Moreover, cell-free systems are unlikely to satisfy such a steady-state approximation after extract preparation and prior to culture initiation. 
+Our framework is similar to the generic kinetic formulation from Hadlich \textit{et al.} \cite{2009_hadlich_wiechert_MetabEng}, 
+but differs in its inclusion of cooperative effects as well as proposes a simplified integration of competition amongst allosteric effectors using max/min rules. 
 In summary, our proposed framework offers an effective kinetic approximation that captures saturation effects and allosteric competition within cell-free systems that may also be extensible to \textit{in vivo} metabolic and gene regulatory networks.
@@
 This suggested that we can expect reasonable model predictions, despite only partial parameter knowledge, as network size grows if we have properly designed experiments. 
-Though we expect that a model's computational complexity will scale poorly with network size, we are optimistic that the construction of large-scale, predictive models of metabolism are possible. 
+Though we expect computational complexity will scale poorly with network size, we are optimistic that large-scale, predictive models of metabolism are possible.
 There is evidence to suggest that achieving a quantitative understanding of complex biological systems should not require complete parametric knowledge.
 Brown and Sethna showed in a model of signal transduction that good predictions were possible despite only order of magnitude estimates of parameter values \citep{Brown:2003aa}.
-Sethna and coworkers later showed that model performance is often controlled by only a few parameter combinations, a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
+Sethna and coworkers later showed that model performance is often controlled by only a few parameter combinations, a characteristic seemingly universal to multi-parameter 
+models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
 We have also demonstrated \textit{sloppy} behavior in a wide variety of signal transduction processes \citep{Luan:2007aa,Song:2009aa,Tasseff:2010aa,Tasseff:2011aa,Nayak:2011aa,Lequieu:2011aa}.
@@
 We considered a population of 100 particles, where each particle in the swarm was assigned a model structure, and a random parameter vector. 
-The optimization simulations of Figure \ref{fig-control-search}, running on a single CPU on an Apple workstation (Apple, Cupertino, CA; OS X v10.10), required on the order of several hours for completion.
+The optimization simulations shown in Figure \ref{fig-control-search} required several hours to complete on a single CPU Apple workstation (Apple, Cupertino, CA; OS X v10.10).  
 The PSO algorithm, model equations, and the objective function were encoded and solved in the Octave programming language (v 3.8.1; www.octave.org).

### 2015-02-16 revision_v2/processes-65165_layout/Paper_v1.tex
Submission of revised v2

--- 2014-11-10 Paper_v1.tex
@@
 % Change the number format in the ref list -
-\renewcommand{\bibnumfmt}[1]{#1.}
+%\renewcommand{\bibnumfmt}[1]{#1.}
 
-% Change Figure to Fig.
-\renewcommand{\figurename}{Fig.}
+% Change Figure to Figure
+%\renewcommand{\figurename}{Figure}
+%--------------------
+% Class Options:
+%--------------------
+% journal
+%----------
+% Choose between the following MDPI journals:
+% actuators, administrativesciences, aerospace, agriculture, agronomy, algorithms, animals, antibiotics, antibodies, antioxidants, appliedsciences, arts, atmosphere, atoms, axioms, batteries, behavioralsciences, bioengineering, biology, biomedicines, biomolecules, biosensors, brainsciences, buildings, cancers, catalysts, cells, challenges, chemosensors, children, chromatography, climate, coatings, computation, computers, cosmetics, crystals, dentistryjournal, diagnostics, diseases, diversity, econometrics, economies, education, electronics, energies, entropy, environmentalsciences, environments, epigenomes, fibers, foods, forests, futureinternet, galaxies, games, gels, genes, geosciences, healthcare, horticulturae, humanities, hydrology, informatics, information, inorganics, insects, ijerph, ijfs, ijms, ijgi, jcdd, jcm, jdb, jfb, jof, joi, jlpea, jmse, jpm, jrfm, jsan, land, laws, life, lubricants, machines, marinedrugs, materials, mathematics, medicalsciences, membranes, metabolites, metals, microarrays, micromachines, microorganisms, minerals, molbank, molecules, nanomaterials, ncrna, nutrients, pathogens, pharmaceuticals, pharmaceutics, pharmacy, photonics, plants, polymers, processes, proteomes, publications, religions, remotesensing, resources, risks, robotics, safety, sensors, sinusitis, socialsciences, societies, sports, standards, sustainability, symmetry, systems, technologies, toxics, toxins, universe, vaccines, veterinarysciences, viruses, water
+%---------
+% article
+%---------
+% The default type of manuscript is article, but could be replaced by using one of the class options: 
+% article, review, communication, commentary, bookreview, correction, addendum, editorial, changes, supfile, casereport, comment, conceptpaper, conferencereport, meetingreport, discussion, essay, letter, newbookreceived, opinion, projectreport, reply, retraction, shortnote, technicalnote, creative
+%----------
+% submit
+%----------
+% The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted.
+This will only make changes to the frontpage (e.g.
+the logo of the journal will get visible), the headings, and the copyright information.
+Journal info and pagination for accepted papers will also be assigned by the Editorial Office.
+% Please insert a blank line is before and after all equation and eqnarray environments to ensure proper line numbering when option submit is chosen
+%------------------
+% moreauthors
+%------------------
+% If there is only one author the class option oneauthor should be used. Otherwise use the class option moreauthors.
+%---------
+% pdftex
+%---------
+% The option "pdftex" is for use with pdfLaTeX only. If eps figure are used, use the optioin "dvipdfm", with LaTeX and dvi2pdf only.
 
-% Begin ...
-\begin{document}
-\begin{titlepage}
-{\par\centering\textbf{\Large Dynamic Modeling of Cell-Free Biochemical Networks using Effective Kinetic Models}}
-\vspace{0.05in}
-{\par \centering \large{Joseph A. Wayman, Adithya Sagar and Jeffrey D. Varner$^{*}$}}
-\vspace{0.10in}
-{\par \centering \large{School of Chemical and Biomolecular Engineering}}
-{\par \centering \large{Cornell University, Ithaca NY 14853}}
-\vspace{0.1in}
-{\par \centering \textbf{Running Title:}~Effective models of metabolism}
-\vspace{0.1in}
-{\par \centering \textbf{To be submitted:}~\textit{Processes}}
-\vspace{0.5in}
-{\par \centering $^{*}$Corresponding author:}
-{\par \centering Jeffrey D. Varner,}
-{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
-{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
-{\par \centering Email: jdv27@cornell.edu} 
-{\par \centering Phone: (607) 255 - 4258}
-{\par \centering Fax: (607) 255 - 9166}
-\end{titlepage}
-\date{}
-\thispagestyle{empty}
-\pagebreak
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Abstract}
-Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
+%=================================================================
+
+%------------------------------------------------------------------
+% The following line should be uncommented if the LaTeX file is uploaded to arXiv.org
+%\pdfoutput=1
+
+%=================================================================
+
+% Add packages and commands to include here
+% The amsmath, amsthm, amssymb, hyperref, caption, float and color packages are loaded by the MDPI class.
+%\usepackage{graphicx}
+%\usepackage{subfigure,psfig}
+
+%=================================================================
+%% Please use the following mathematics environments:
+%\theoremstyle{mdpi}
+%\newcounter{thm}
+%\setcounter{thm}{0}
+%\newcounter{ex}
+%\setcounter{ex}{0}
+%\newcounter{re}
+%\setcounter{re}{0}
+%\newtheorem{Theorem}[thm]{Theorem}
+%\newtheorem{Lemma}[thm]{Lemma}
+%\newtheorem{Characterization}[thm]{Characterization}
+%\newtheorem{Proposition}[thm]{Proposition}
+%\newtheorem{Property}[thm]{Property}
+%\newtheorem{Problem}[thm]{Problem}
+%\newtheorem{Example}[ex]{Example}
+%\newtheorem{Remark}[re]{Remark}
+%\newtheorem{Corollary}[thm]{Corollary}
+%\newtheorem{Definition}[thm]{Definition}
+%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).
+
+%=================================================================
+
+% Full title of the paper (Capitalized)
+\Title{Dynamic Modeling of Cell-Free Biochemical Networks using Effective Kinetic Models}
+
+% Authors (Add full first names)
+\Author{Joseph A. Wayman, Adithya Sagar and Jeffrey D. Varner {*}}
+
+% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
+\address[1]{%
+School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY 14853, USA\\ \vspace{12pt}
+
+Running Title:~Effective models of metabolism}
+
+% Contact information of the corresponding author (Add [2] after \corres if there are more than one corresponding author.)
+\corres{E-Mail: jdv27@cornell.edu; \mbox{Tel.: +607-255-4258;} Fax: +607-255-9166.}
+
+% Abstract (Do not use inserted blank lines, i.e. \\) 
+\abstract{Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
 Many of the challenges confronting genome-scale kinetic modeling can potentially be overcome in a cell-free system. 
@@
 Lastly, we showed, at least for the network architectures considered here, that we could simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data. 
-While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.
+While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.}
 
-{\noindent \textbf{Keywords:}~Cell-free metabolism, Mathematical modeling}
+% Keywords: add 3 to 10 keywords
+\keyword{Cell-free metabolism, Mathematical modeling}
 
-\pagebreak
+% The fields PACS, MSC, and JEL may be left empty or commented out if not applicable
+%\PACS{}
+%\MSC{}
+%\JEL{}
 
-\setcounter{page}{1}
+% If this is an expanded version of a conference paper, please cite it here: enter the full citation of your conference paper, and add $^\dagger$ in the end of the title of this article.
+%\conference{}
 
-\linenumbers
+\begin{document}
 
-\section*{Introduction}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Introduction}
+
 Mathematical modeling has long contributed to our understanding of metabolism. 
@@
 
-\clearpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
 
-\section*{Results}
+\section{Results}\vspace{-12pt}
 
-\subsection*{Formulation and properties of effective cell-free metabolic models.}
-We developed two proof-of-concept metabolic networks to investigate the features of our effective biochemical network modeling approach (Fig. \ref{fig-networks}). 
+%% Only for the Journal of Physical and Chemical Gels: Please place the Experimental Section after the Conclusions
+
+\subsection{Formulation and Properties of Effective Cell-Free Metabolic Models}
+
+We developed two proof-of-concept metabolic networks to investigate the features of our effective biochemical network modeling approach (Figure \ref{fig-networks}). 
 In both examples, substrate S was converted to the end products P$_{1}$ and P$_{2}$ through a series of enzymatically catalyzed reactions, including a branch point at hypothetical metabolite M$_{2}$. 
 Several of these reactions involved cofactor dependence (AH or A), and various allosteric regulatory mechanisms modified the activity of pathway enzymes. 
-Network A included feedback inhibition of the initial pathway enzyme ($E_{1}$) by pathway end products P$_{1}$ and P$_{2}$ (Fig. \ref{fig-networks}A). 
-On the other hand, network B involved feedback inhibition of $E_{1}$ by P$_{2}$ and $E_{6}$ by P$_{1}$ (Fig. \ref{fig-networks}B). 
+Network A included feedback inhibition of the initial pathway enzyme ($E_{1}$) by pathway end products P$_{1}$ and P$_{2}$ (Figure \ref{fig-networks}A). 
+On the other hand, network B involved feedback inhibition of $E_{1}$ by P$_{2}$ and $E_{6}$ by P$_{1}$ (Figure \ref{fig-networks}B). 
 In both networks, branch point enzymes $E_{3}$ and $E_{6}$ were subject to feed-forward activation by reduced cofactor AH. 
@@
 
-Allosteric regulation of enzyme activity was modeled by combining individual regulatory contributions to the activity of pathway enzymes into a control coefficient using an integration rule (Fig.
-\ref{fig-control-schematic}).
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{./figs/Figure-1-Networks.pdf}
+\caption{Proof of concept cell-free metabolic networks considered in this study. Substrate $S$ is converted to products P$_{1}$ and P$_{2}$ through a series of chemical conversions
+catalyzed by enzyme(s) $E_{j}$. The activity of the pathway enzymes is subject to both positive and negative allosteric regulation. }\label{fig-networks}
+\end{figure}
+
+Allosteric regulation of enzyme activity was modeled by combining individual regulatory \linebreak contributions to the activity of pathway enzymes into a control coefficient using an integration rule (Figure \ref{fig-control-schematic}).
 This strategy is similar in spirit to the Constrained Fuzzy Logic (cFL) approach of Lauffenburger and coworkers which has been used to effectively model signal transduction pathways important in human health \citep{Morris:2011aa}.
@@
 Thus, our test networks involved important features such as cofactor recycling, enzyme activity and metabolite dynamics, as well as multiple overlapping allosteric regulatory mechanisms.  
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{./figs/Figure-2-ControlSchematic.pdf}
+\caption{Schematic of rule-based allosteric enzyme activity control laws. 
+Traditional enzyme kinetic expressions, e.g., Michaelis–Menten or multiple saturation kinetics, are multiplied by an enzyme activity control variable $0 \leq v_{j} \leq 1 $. 
+Control variables are functions of many possible regulatory factors encoded by arbitrary functions of the form $0\leq f_{j}\left(\mathcal{Z}\right)\leq 1$.
+At each simulation time step, the $v_{j}$ variables are calculated by evaluating integration rules such as the max or min of the set of factors $f_{1},\hdots$ 
+influencing the activity of enzyme $E_{j}$. }\label{fig-control-schematic}
+\end{figure}
 
-The rule-based regulatory strategy approximated the behavior of classical allosteric activation and inhibition mechanisms (Fig. \ref{fig-kinetics-simulations}). 
+The rule-based regulatory strategy approximated the behavior of classical allosteric activation and inhibition mechanisms (Figure \ref{fig-kinetics-simulations}). 
 We considered the enzyme catalyzed conversion of substrate S to a product P, where the overall reaction rate was modeled as the product of a Michaelis-Menten term and an effective allosteric control variable reflecting the particular regulatory interaction.
 We first explored feed-forward substrate activation of enzyme activity (for both positive and negative cooperativity). 
-Consistent with classical data, the rule-based strategy predicted a sigmoidal relationship between substrate abundance and reaction rate as a function of the cooperativity parameter (Fig.
-\ref{fig-kinetics-simulations}A).
+Consistent with classical data, the rule-based strategy predicted a sigmoidal relationship between substrate abundance and reaction rate as a function of the cooperativity parameter (Figure \ref{fig-kinetics-simulations}A).
 For cooperativity parameters less than unity, increased substrate abundance \textit{decreased} the reaction rate. 
@@
 On the other hand, as the cooperativity parameter increased past unity, the rate of conversion of substrate S to product P by enzyme $E$ approached a step function. 
-In the presence of an inhibitor, the rule-based strategy predicted non-competitive like behavior as a function of the cooperativity parameter (Fig. \ref{fig-kinetics-simulations}B). 
-When the control gain parameter, $\kappa_{ij}$ in Eqn.
-\eqref{eqn:control-factor}, was greater than unity, the inhibitory force was directly proportional to the cooperativity parameter, $\eta$ in Eqn.
-\eqref{eqn:control-factor}.
-Thus, as the cooperativity parameter increased, the maximum reaction rate decreased (Fig. \ref{fig-kinetics-simulations}B). 
+In the presence of an inhibitor, the rule-based strategy predicted non-competitive like behavior as a function of the cooperativity parameter (Figure \ref{fig-kinetics-simulations}B). 
+When the control gain parameter, $\kappa_{ij}$ in Equation~\eqref{eqn:control-factor}, was greater than unity, the inhibitory force was directly proportional to the cooperativity parameter, $\eta$ in Equation~\eqref{eqn:control-factor}.
+Thus, as the cooperativity parameter increased, the maximum reaction rate decreased (Figure \ref{fig-kinetics-simulations}B). 
 Interestingly, our rule-based approach was unable to directly simulate competitive inhibition of enzyme activity. 
@@
 
-End product yield was controlled by feedback inhibition, while product selectivity was controlled by branch point enzyme inhibition (Fig. \ref{fig-onoff-simulations}). 
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure-3-EnzymeKinetics.pdf}
+\caption{Kinetics of simple transformations in the presence of activation and inhibition. 
+\textbf{A}:The conversion of substrate $S$ to product $P$ by enzyme $E$ was activated by $S$. 
+For a fixed control gain parameter $k_{control}$, the reaction rate approached a step for increasing control order $N$. 
+\textbf{B}:The conversion of substrate $S$ to product $P$ by enzyme $E$ with inhibitor $I$. 
+For a fixed control gain parameter $k_{control}$, the reaction rate approximated non-competitive inhibition for increasing control order $N$. 
+}\label{fig-kinetics-simulations}
+\end{figure}
+
+End product yield was controlled by feedback inhibition, while product selectivity was controlled by branch point enzyme inhibition (Figure \ref{fig-onoff-simulations}). 
 A critical test of our modeling approach was to simulate networks with known behavior. 
@@
 As expected, end product accumulation was larger for network A when the control was OFF (no feedback inhibition of $E_{1}$ by P$_{1}$ and P$_{2}$),
-as compared to the ON case (Fig. \ref{fig-onoff-simulations}A). 
+as compared to the ON case (Figure~\ref{fig-onoff-simulations}A). 
 We found this behavior was robust to the choice of underlying kinetic parameters, as we observed that same qualitative response across an ensemble of randomized parameter sets (N = 100), for fixed control parameters.
@@
 In the OFF case, the behavior was qualitatively similar to network A. 
-However, for the ON case, flux was diverted away from P$_{2}$ formation by feedback inhibition of $E_{6}$ activity at the M$_{2}$ branch point by P$_{1}$ (Fig. \ref{fig-onoff-simulations}B). 
-Lower $E_{6}$ activity at the M$_{2}$ branch point allowed more flux toward P$_{1}$ formation, hence the yield of P$_{1}$ also increased (Fig. \ref{fig-onoff-simulations}C). 
+However, for the ON case, flux was diverted away from P$_{2}$ formation by feedback inhibition of $E_{6}$ activity at the M$_{2}$ branch point by P$_{1}$ (Figure \ref{fig-onoff-simulations}B). 
+Lower $E_{6}$ activity at the M$_{2}$ branch point allowed more flux toward P$_{1}$ formation, hence the yield of P$_{1}$ also increased (Figure \ref{fig-onoff-simulations}C). 
 Again, the control ON/OFF behavior of network B was robust to changes in kinetic parameters, as the same qualitative trend was conserved across an ensemble of randomized parameters (N = 100), for fixed control parameters.
@@
 
-\subsection*{Estimating parameters and effective allosteric regulatory structures.}
+\subsection{Estimating Parameters and Effective Allosteric Regulatory Structures}
+
 A critical challenge for any dynamic model is the estimation of kinetic parameters. 
@@
 We then generated an ensemble of model parameter estimates by minimizing the difference between model simulations and the synthetic data using particle swarm optimization (PSO), starting from random initial parameter guesses.
-The estimation of kinetic parameters was sensitive to the choice of regulatory structure (Fig.
-\ref{fig-parameter-fit}).
-PSO identified an ensemble of parameters that bracketed the mean of the synthetic measurements in less than 1000 iterations when the control structure was correct (Fig.
-\ref{fig-parameter-fit}A and B).
-However, with control mismatch (network B simulated with network A parameters), model simulations were not consistent with the synthetic data  (Fig. \ref{fig-parameter-fit}C and D). 
+The estimation of kinetic parameters was sensitive to the choice of regulatory structure (Figure~\ref{fig-parameter-fit}).
+PSO identified an ensemble of parameters that bracketed the mean of the synthetic measurements in less than 1000 iterations when the control structure was correct (Figure \ref{fig-parameter-fit}A,B). 
+However, with control mismatch (network B simulated with network A parameters), model simulations were not consistent with the synthetic data  (Figure \ref{fig-parameter-fit}C,D). 
 Taken together, these results suggested that we could perhaps simultaneously estimate both parameters and network control architectures, as incorrect control structures would be manifest as poor model fits.
 
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.450\textwidth]{./figs/Figure-4-OnOffSimulations.pdf}
+\caption{ON/OFF control simulations for network A and network B for an ensemble of kinetic parameter sets versus time (N = 100). 
+For each case, N = 100 simulations were conducted using kinetic and initial conditions generated randomly from a hypothetical true parameter set. 
+The gray area represents $\pm$ one standard deviation surrounding the mean. 
+Control parameters were fixed during the ensemble calculations.
+\textbf{A}: End product P$_{1}$ abundance versus time for Network A. 
+The abundance of P$_{1}$ decreased with end product inhibition of $E_{1}$ activity (Control-ON) versus the no inhibition case (Control-OFF). 
+\textbf{B}: End product P$_{2}$ abundance versus time for Network B. Inhibition of branch point $E_{6}$ by end product P$_{1}$ decreased P$_{2}$ abundance (Control-ON) versus the
+no inhibition case (Control-OFF).
+\textbf{C}: End product P$_{1}$ abundance versus time for Network A. 
+Inhibition of branch point $E_{6}$ by end product P$_{1}$ decreased P$_{1}$ abundance (Control-ON) versus the no inhibition case (Control-OFF).
+}\label{fig-onoff-simulations}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure-5-ParameterFit.pdf}
+\caption{Parameter estimation from synthetic data for the same and mismatched allosteric control logic using particle swarm optimization (PSO). 
+Synthetic experimental data was generated from a hypothetical parameter set using Network A, 
+where substrate $S$, end product P$_{1}$ and intermediate M$_5$ were sampled approximately every 20 minutes. 
+For cases $\textbf{A,B}$ 20 particles were initialized with randomized parameters and allowed to search for 300 iterations. 
+\textbf{A,B}: PSO estimated an ensemble of parameters sets (N = 20) consistent with the synthetic experimental data assuming the correct
+enzymatic and control connectivity starting from randomized initial parameters.
+\textbf{C,D}: In the presence of control mismatch (Network B control policy simulated with Network A kinetic parameters) 
+the ensemble of models did not describe the synthetic data. 
+}\label{fig-parameter-fit}
+\end{figure}
+
 We modified our particle swarm identification strategy to simultaneously search over both kinetic parameters and putative control structures. 
-In addition to our initial networks, we constructed three additional presumptive network models, each with the same enzymatic connectivity but different allosteric regulation of the pathway enzymes (Fig.
-\ref{fig-alternative-networks}).
+In addition to our initial networks, we constructed three additional presumptive network models, each with the same enzymatic connectivity but different allosteric regulation of the pathway enzymes (Figure \ref{fig-alternative-networks}).
 We then initialized a population of particles, each with one of the five potential regulatory programs and randomized kinetic parameters. 
@@
 To this end, we considered three different priors, a uniform distribution where each putative regulatory structure represented 20\% of the population and two mixed distributions that were either positively or negatively biased towards the correct structure (network A).
-In both the positively biased and uniform cases the PSO clearly differentiated between the true or closely related structures and those that were materially different (Fig. \ref{fig-control-search}). 
-As expected, the positively biased population (40\% of the initial particle population seeded with network A) gave the best results, where the correct structure was preferentially identified (Fig.
-\ref{fig-control-search}A).
-On the other hand, when given a uniform distribution, the PSO approach identified a combination of network A and network C as the most likely control structures (Fig. \ref{fig-control-search}B). 
+In both the positively biased and uniform cases the PSO clearly differentiated between the true or closely related structures and those that were materially different (Figure \ref{fig-control-search}).
+As expected, the positively biased population (40\% of the initial particle population seeded with network A) gave the best results, where the correct structure was preferentially identified (Figure \ref{fig-control-search}A).
+On the other hand, when given a uniform distribution, the PSO approach identified a combination of network A and network C as the most likely control structures (Figure \ref{fig-control-search}B). 
 Network A and C differ by the regulatory connection between the end product P$_2$ and enzyme $E_{1}$; in network A, end product P$_{2}$ was assumed to inhibit $E_{1}$, while in network C, end product P$_{2}$ activated $E_{1}$.
-Lastly, when the initial population was biased towards incorrect structures (initial population seeded with 90\% incorrect structures), the particle swarm \textit{misidentified} the correct allosteric structure (Fig.
-\ref{fig-control-search}C).
+Lastly, when the initial population was biased towards incorrect structures (initial population seeded with 90\% incorrect structures), the particle swarm \textit{misidentified} the correct allosteric structure (Figure \ref{fig-control-search}C).
 Interestingly, while each particle swarm identified parameter sets that minimized the simulation error, the estimated parameter values were not necessarily similar to the true parameters. 
@@
 
-We calculated control program output and scaled metabolic flux for the positively, uniformly and negatively biased particle swarms (Fig. \ref{fig-flux-pattern}). 
-Network A and network C models from the positively (Fig.
-\ref{fig-flux-pattern}A) and uniformly (Fig.
-\ref{fig-flux-pattern}B) biased particle swarms showed similar operational patterns, despite differences in kinetic parameters and control structures.
-While models from the negatively biased population had error values similar to the correct structures in the previous swarms, they have different flux and control profiles (Fig.
-\ref{fig-flux-pattern}C).
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure-6-AlternativeNetworks.pdf}
+\caption{Schematic of the alternative allosteric control programs used in the structural particle swarm computation. 
+Each network had the same enzymatic connectivity, initial conditions and kinetic parameters, 
+but alternative feedback control structures for the first enzyme in the pathway.}\label{fig-alternative-networks}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.47\textwidth]{./figs/Figure-7-ControlSearch.pdf}
+\caption{Combined control and kinetic parameter search using modified particle swarm optimization (PSO). 
+A population of N = 100 particles was initialized with randomized kinetic parameters and one of five possible control configurations (Network A - E).
+Simulation error was minimized for a synthetic data set 
+($S$, end product P$_{1}$ and intermediate M$_5$ sampled approximately every 20 min) generated using Network A.
+\textbf{A}: Simulation error versus parameter set angle for N = 100 particles biased toward the correct regulatory program (A,B,C,D,E) = (40\%, 10\%, 20\%, 20\% and 10\%).
+\textbf{B}: Simulation error versus parameter set angle for N = 100 uniformly distributed particles (A,B,C,D,E) = (20\%, 20\%, 20\%, 20\% and 20\%).
+\textbf{C}: Simulation error versus parameter set angle for N = 100 negatively biased particles (A,B,C,D,E) = (10\%, 40\%, 10\%, 20\% and 20\%).
+Network A (the correct structure) was preferentially identified for positively and uniform biased particle distributions, but misidentified in the presence of
+a large incorrect bias. 
+}\label{fig-control-search}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.0\textwidth,height=0.6\textheight]{./figs/Figure-8-Flux.pdf}
+\caption{Metabolic flux and control variables as a function of network type and particle index at t = 100 min.
+The control variables governing $E_{1},E_{3}$ and $E_{6}$ activity and the scaled metabolic flux and were calculated for the positively, 
+uniformly and negatively biased particle swarms (N = 100). The particles from each swarm were sorted based upon simulation error (low to high error). 
+\textbf{A}: Model performance for the positively biased particle swarm as a function of particle index.
+\textbf{B}: Model performance for the uniformly biased particle swarm as a function of particle index.
+\textbf{C}: Model performance for the negatively biased particle swarm as a function of particle index. 
+Models with significant control mismatch showed distinct control and flux patterns versus those models with the correct or closely related control policies.
+In particular, models with the correct control policy showed stronger inhibition of $E_{1}$ activity, leading to decreased flux from S$\rightarrow$P$_{1}$.
+Conversely, models with significant mismatch had increased $E_{1}$ activity, leading to an altered flux distribution. 
+This is especially apparent in the negatively biased particle swarm. 
+}\label{fig-flux-pattern}
+\end{figure}
+
+We calculated control program output and scaled metabolic flux for the positively, uniformly and negatively biased particle swarms (Figure \ref{fig-flux-pattern}). 
+Network A and network C models from the positively (Figure \ref{fig-flux-pattern}A) and uniformly (Figure \ref{fig-flux-pattern}B) biased particle swarms showed similar operational patterns, despite differences in kinetic parameters and control structures.
+While models from the negatively biased population had error values similar to the correct structures in the previous swarms, they have different flux and control profiles (Figure \ref{fig-flux-pattern}C).
 In all cases, regardless of network configuration or parameter values, the rate of enzyme decay was small compared to the other fluxes, and all networks had qualitatively similar trends for $E_{3}$ and $E_{6}$ control.
@@
 
-\clearpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section*{Discussion}
+\section{Discussion}
+
 In this study, we presented an effective kinetic modeling strategy to dynamically simulate cell-free biochemical networks. 
@@
 Lastly, we simultaneously estimated kinetic parameters and discriminated between competing regulatory structures, using synthetic data in combination with a modified particle swarm approach.
-If we considered all putative regulatory architectures to be equally likely, we were able to estimate a \textit{sloppy} ensemble of models with the correct architecture and kinetic parameters.
+If we considered all putative regulatory architectures to be equally likely, we were able to estimate a \textit{sloppy} ensemble of models with the correct architecture and \mbox{kinetic parameters.}
 
@@
 If the regulated intracellular flux problem is coupled with time-dependent extracellular balances, these models can predict complex behavior such as diauxie growth or the switch between 
-aerobic and anaerobic metabolism. Another important feature of this approach is that it scales with biological complexity. 
+aerobic and anaerobic metabolism. Another important features of this approach is that it scales with biological complexity. 
 For example, Covert \emph{et al.,} showed that a genome scale model of \emph{E.~coli} augmented with a boolean rule layer, 
@@
 
-\clearpage
+\section{Materials and Methods}\vspace{-12pt}
 
-\section*{Materials and Methods}
+\subsection{Formulation and Solution of the Model Equations}
 
-\subsection*{Formulation and solution of the model equations.}
 We used ordinary differential equations (ODEs) to model the time evolution of metabolite ($x_{i}$) and scaled enzyme abundance ($\epsilon_{i}$) in hypothetical cell-free metabolic networks:
@@
 
-Each reaction rate was written as the product of two terms, a kinetic term ($\bar{r}_{j}$) and a regulatory term ($v_{j}$):
+Each reaction rate was written as the product of two terms, a kinetic term ($\bar{r}_{j}$) and a regulatory \mbox{term ($v_{j}$):}
 \begin{equation}\label{eqn:rate-factor}
@@
 $K_{js}$ denotes the saturation constant for species $s$ in reaction $j$. 
-The product in Eqn. \eqref{eqn:rate-bar} was carried out over the set of \textit{reactants} for reaction $j$ (denoted as $m_{j}^{-}$).
+The product in Equation~\eqref{eqn:rate-bar} was carried out over the set of \textit{reactants} for reaction $j$ (denoted as $m_{j}^{-}$).
 
 The allosteric regulation term $v_{j}$ depended upon the combination of factors which influenced the activity of enzyme $i$.
-For each enzyme, we used a rule-based approach to select from competing control factors (Fig. \ref{fig-control-schematic}). 
+For each enzyme, we used a rule-based approach to select from competing control factors (Figure \ref{fig-control-schematic}). 
 If an enzyme was activated by $m$ metabolites, we modeled this activation as:
@@
 The $\kappa_{ij}$ parameter was species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).
-The model equations were encoded using the Octave programming language and solved using the LSODE routine in Octave \citep{Octave:2014}.
+The model equations were encoded using the Octave programming language and solved using the LSODE routine in {Octave} \citep{Octave:2014}.
 
-\subsection*{Estimation of model parameters and structures from synthetic experimental data.}
+\subsection{Estimation of Model Parameters and Structures From Synthetic Experimental Data}
+
 Model parameters were estimated by minimizing the difference between simulations and synthetic experimental data (squared residual):
@@
 We considered a population N = 100 particles, where each particle in the swarm was assigned a model structure, and a random parameter vector.
-The PSO algorithm, model equations, and the objective function were encoded and solved in the Octave programming language \citep{Octave:2014}.
+The PSO algorithm, model equations, and the objective function were encoded and solved in the Octave programming {language} \citep{Octave:2014}.
 
-\section*{Acknowledgements}
+%\section{Conclusions}
+%
+%\hl{Main text paragraph.}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section*{\noindent Acknowledgments}\vspace{12pt}
+
 This study was supported by an award from the National Science Foundation (MCB \#1411715) 
@@
 
-\clearpage
-%\bibliographystyle{plain}
-%\bibliographystyle{IEEEbib}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\bibliography{References_v1}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-%\begin{comment}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\clearpage
+%\section*{\noindent Author Contributions}\vspace{12pt}
+%
+%\hl{Main text.}
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\section*{\noindent Conflicts of Interest}\vspace{12pt}
+%
+%\hl{The authors declare no conflict of interest.}
 
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-1-Networks.pdf}
-\caption{Proof of concept cell-free metabolic networks considered in this study. Substrate $S$ is converted to products P$_{1}$ and P$_{2}$ through a series of chemical conversions
-catalyzed by enzyme(s) $E_{j}$. The activity of the pathway enzymes is subject to both positive and negative allosteric regulation. }\label{fig-networks}
-\end{figure}
+%=================================================================
+% References: Variant A
+%=================================================================
+% Back Matter (References and Notes)
+%----------------------------------------------------------
+% Style and layout of the references
+\bibliographystyle{mdpi}
 
-\clearpage
+%\bibliography{References_v1}
 
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-2-ControlSchematic.pdf}
-\caption{Schematic of rule-based allosteric enzyme activity control laws. 
-Traditional enzyme kinetic expressions, e.g., Michaelis–Menten or multiple saturation kinetics, are multiplied by an enzyme activity control variable $0 \leq v_{j} \leq 1 $. 
-Control variables are functions of many possible regulatory factors encoded by arbitrary functions of the form $0\leq f_{j}\left(\mathcal{Z}\right)\leq 1$.
-At each simulation time step, the $v_{j}$ variables are calculated by evaluating integration rules such as the max or min of the set of factors $f_{1},\hdots$ 
-influencing the activity of enzyme $E_{j}$. }\label{fig-control-schematic}
-\end{figure}
+\begin{thebibliography}{----}
+\providecommand{\natexlab}[1]{#1}
 
-\clearpage
+\bibitem[Fredrickson(1976)]{1976_fredrickson_BiotechBioeng}
+Fredrickson, A.G.
+\newblock Formulation of structured growth models.
+\newblock {\em Biotechnol Bioeng} {\bf 1976},
+\newblock {\em 18},~1481--6.
 
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-3-EnzymeKinetics.pdf}
-\caption{Kinetics of simple transformations in the presence of activation and inhibition. 
-\textbf{A}:The conversion of substrate $S$ to product $P$ by enzyme $E$ was activated by $S$. 
-For a fixed control gain parameter $k_{control}$, the reaction rate approached a step for increasing control order $N$. 
-\textbf{B}:The conversion of substrate $S$ to product $P$ by enzyme $E$ with inhibitor $I$. 
-For a fixed control gain parameter $k_{control}$, the reaction rate approximated non-competitive inhibition for increasing control order $N$. 
-}\label{fig-kinetics-simulations}
-\end{figure}
+\bibitem[Domach \em{et~al.}(1984)Domach, Leung, Cahn, Cocks, and
+  Shuler]{1984_domach_shuler_BiotechBioeng_01}
+Domach, M.M.; Leung, S.K.; Cahn, R.E.; Cocks, G.G.; Shuler, M.L.
+\newblock Computer model for glucose-limited growth of a single cell of
+  Escherichia coli B/r-A.
+\newblock {\em Biotechnol Bioeng} {\bf 1984},
+\newblock {\em 26},~203--16.
 
-\clearpage
+\bibitem[Steinmeyer and Shuler(1989)]{1989_steinmeyer_shuler_ChemEngSci}
+Steinmeyer, D.; Shuler, M.
+\newblock {Structured model for Saccharomyces cerevisiae.}
+\newblock {\em Chem. Eng. Sci.} {\bf 1989},
+\newblock {\em 44},~2017--2030.
 
-\begin{figure}
-\centering
-\includegraphics[width=0.50\textwidth]{./figs/Figure-4-OnOffSimulations.pdf}
-\caption{ON/OFF control simulations for network A and network B for an ensemble of kinetic parameter sets versus time (N = 100). 
-For each case, N = 100 simulations were conducted using kinetic and initial conditions generated randomly from a hypothetical true parameter set. 
-The gray area represents $\pm$ one standard deviation surrounding the mean. 
-Control parameters were fixed during the ensemble calculations.
-\textbf{A}: End product P$_{1}$ abundance versus time for Network A. 
-The abundance of P$_{1}$ decreased with end product inhibition of $E_{1}$ activity (Control-ON) versus the no inhibition case (Control-OFF). 
-\textbf{B}: End product P$_{2}$ abundance versus time for Network B. Inhibition of branch point $E_{6}$ by end product P$_{1}$ decreased P$_{2}$ abundance (Control-ON) versus the
-no inhibition case (Control-OFF).
-\textbf{C}: End product P$_{1}$ abundance versus time for Network A. 
-Inhibition of branch point $E_{6}$ by end product P$_{1}$ decreased P$_{1}$ abundance (Control-ON) versus the no inhibition case (Control-OFF).
-}\label{fig-onoff-simulations}
-\end{figure}
+\bibitem[Wu \em{et~al.}(1992)Wu, Ray, and Shuler]{1992_wu_shuler_AnnNYAcadSci}
+Wu, P.; Ray, N.G.; Shuler, M.L.
+\newblock A single-cell model for CHO cells.
+\newblock {\em Ann N Y Acad Sci} {\bf 1992},
+\newblock {\em 665},~152--87.
 
-\clearpage
+\bibitem[Castellanos \em{et~al.}(2004)Castellanos, Wilson, and
+  Shuler]{2004_castellanos_shuler_PNAS}
+Castellanos, M.; Wilson, D.B.; Shuler, M.L.
+\newblock A modular minimal cell model: purine and pyrimidine transport and
+  metabolism.
+\newblock {\em Proc Natl Acad Sci U S A} {\bf 2004},
+\newblock {\em 101},~6681--6.
 
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-5-ParameterFit.pdf}
-\caption{Parameter estimation from synthetic data for the same and mismatched allosteric control logic using particle swarm optimization (PSO). 
-Synthetic experimental data was generated from a hypothetical parameter set using Network A, 
-where substrate $S$, end product P$_{1}$ and intermediate M$_5$ were sampled approximately every 20 minutes. 
-For cases $\textbf{A,B}$ 20 particles were initialized with randomized parameters and allowed to search for 300 iterations. 
-\textbf{A,B}: PSO estimated an ensemble of parameters sets (N = 20) consistent with the synthetic experimental data assuming the correct
-enzymatic and control connectivity starting from randomized initial parameters.
-\textbf{C,D}: In the presence of control mismatch (Network B control policy simulated with Network A kinetic parameters) 
-the ensemble of models did not describe the synthetic data. 
-}\label{fig-parameter-fit}
-\end{figure}
+\bibitem[Atlas \em{et~al.}(2008)Atlas, Nikolaev, Browning, and
+  Shuler]{2008_atlas_shuler_IETSysBio}
+Atlas, J.C.; Nikolaev, E.V.; Browning, S.T.; Shuler, M.L.
+\newblock Incorporating genome-wide DNA sequence information into a dynamic
+  whole-cell model of Escherichia coli: application to DNA replication.
+\newblock {\em IET Syst Biol} {\bf 2008},
+\newblock {\em 2},~369--82.
 
-\clearpage
+\bibitem[Dhurjati \em{et~al.}(1985)Dhurjati, Ramkrishna, Flickinger, and
+  Tsao]{1985_dhurjati_ramkrishna_tsao_BiotechBioeng}
+Dhurjati, P.; Ramkrishna, D.; Flickinger, M.C.; Tsao, G.T.
+\newblock A cybernetic view of microbial growth: modeling of cells as optimal
+  strategists.
+\newblock {\em Biotechnol Bioeng} {\bf 1985},
+\newblock {\em 27},~1--9.
 
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-6-AlternativeNetworks.pdf}
-\caption{Schematic of the alternative allosteric control programs used in the structural particle swarm computation. 
-Each network had the same enzymatic connectivity, initial conditions and kinetic parameters, 
-but alternative feedback control structures for the first enzyme in the pathway.}\label{fig-alternative-networks}
-\end{figure}
+\bibitem[Kompala \em{et~al.}(1986)Kompala, Ramkrishna, Jansen, and
+  Tsao]{1986_kompala_ramkrishna_tsao_BiotechBioeng}
+Kompala, D.S.; Ramkrishna, D.; Jansen, N.B.; Tsao, G.T.
+\newblock Investigation of bacterial growth on mixed substrates: experimental
+  evaluation of cybernetic models.
+\newblock {\em Biotechnol Bioeng} {\bf 1986},
+\newblock {\em 28},~1044--55.
 
-\clearpage
+\bibitem[Kim \em{et~al.}(2012)Kim, Song, Sunkara, Lali, and
+  Ramkrishna]{2012_kim_ramkrishna_BiotechProg}
+Kim, J.I.; Song, H.S.; Sunkara, S.R.; Lali, A.; Ramkrishna, D.
+\newblock Exacting predictions by cybernetic model confirmed experimentally:
+  steady state multiplicity in the chemostat.
+\newblock {\em Biotechnol Prog} {\bf 2012},
+\newblock {\em 28},~1160--6.
 
-\begin{figure}
-\centering
-\includegraphics[width=0.5\textwidth]{./figs/Figure-7-ControlSearch.pdf}
-\caption{Combined control and kinetic parameter search using modified particle swarm optimization (PSO). 
-A population of N = 100 particles was initialized with randomized kinetic parameters and one of five possible control configurations (Network A - E).
-Simulation error was minimized for a synthetic data set 
-($S$, end product P$_{1}$ and intermediate M$_5$ sampled approximately every 20 min) generated using Network A.
-\textbf{A}: Simulation error versus parameter set angle for N = 100 particles biased toward the correct regulatory program (A,B,C,D,E) = (40\%, 10\%, 20\%, 20\% and 10\%).
-\textbf{B}: Simulation error versus parameter set angle for N = 100 uniformly distributed particles (A,B,C,D,E) = (20\%, 20\%, 20\%, 20\% and 20\%).
-\textbf{C}: Simulation error versus parameter set angle for N = 100 negatively biased particles (A,B,C,D,E) = (10\%, 40\%, 10\%, 20\% and 20\%).
-Network A (the correct structure) was preferentially identified for positively and uniform biased particle distributions, but misidentified in the presence of
-a large incorrect bias. 
-}\label{fig-control-search}
-\end{figure}
+\bibitem[Varner and Ramkrishna(1999)]{1999_varner_ramkrishna_MetaEng}
+Varner, J.; Ramkrishna, D.
+\newblock Metabolic engineering from a cybernetic perspective: aspartate family
+  of amino acids.
+\newblock {\em Metab Eng} {\bf 1999},
+\newblock {\em 1},~88--116.
 
-\clearpage
+\bibitem[Lewis \em{et~al.}(2012)Lewis, Nagarajan, and
+  Palsson]{2012_lewis_palsson_NatRevMicrobio}
+Lewis, N.E.; Nagarajan, H.; Palsson, B.O.
+\newblock Constraining the metabolic genotype-phenotype relationship using a
+  phylogeny of in silico methods.
+\newblock {\em Nat Rev Microbiol} {\bf 2012},
+\newblock {\em 10},~291--305.
 
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth,height=0.6\textheight]{./figs/Figure-8-Flux.pdf}
-\caption{Metabolic flux and control variables as a function of network type and particle index at t = 100 min.
-The control variables governing $E_{1},E_{3}$ and $E_{6}$ activity and the scaled metabolic flux and were calculated for the positively, 
-uniformly and negatively biased particle swarms (N = 100). The particles from each swarm were sorted based upon simulation error (low to high error). 
-\textbf{A}: Model performance for the positively biased particle swarm as a function of particle index.
-\textbf{B}: Model performance for the uniformly biased particle swarm as a function of particle index.
-\textbf{C}: Model performance for the negatively biased particle swarm as a function of particle index. 
-Models with significant control mismatch showed distinct control and flux patterns versus those models with the correct or closely related control policies.
-In particular, models with the correct control policy showed stronger inhibition of $E_{1}$ activity, leading to decreased flux from S$\rightarrow$P$_{1}$.
-Conversely, models with significant mismatch had increased $E_{1}$ activity, leading to an altered flux distribution. 
-This is especially apparent in the negatively biased particle swarm. 
-}\label{fig-flux-pattern}
-\end{figure}
+\bibitem[Edwards and Palsson(2000)]{2000_edwards_palsson_PNAS}
+Edwards, J.S.; Palsson, B.O.
+\newblock The Escherichia coli MG1655 in silico metabolic genotype: its
+  definition, characteristics, and capabilities.
+\newblock {\em Proc Natl Acad Sci U S A} {\bf 2000},
+\newblock {\em 97},~5528--33.
 
-\clearpage
+\bibitem[Feist \em{et~al.}(2007)Feist, Henry, Reed, Krummenacker, Joyce, Karp,
+  Broadbelt, Hatzimanikatis, and Palsson]{Feist:2007aa}
+Feist, A.M.; Henry, C.S.; Reed, J.L.; Krummenacker, M.; Joyce, A.R.; Karp,
+  P.D.; Broadbelt, L.J.; Hatzimanikatis, V.; Palsson, B.{\O}.
+\newblock A genome-scale metabolic reconstruction for Escherichia coli K-12
+  MG1655 that accounts for 1260 ORFs and thermodynamic information.
+\newblock {\em Mol Syst Biol} {\bf 2007},
+\newblock {\em 3},~121.
 
-% Supplemental figures -
-% Set the S- 
-\renewcommand\thefigure{S\arabic{figure}}
-\renewcommand\thetable{T\arabic{table}}
-\renewcommand\thepage{S-\arabic{page}}
-\renewcommand\theequation{S\arabic{equation}}
+\bibitem[Oh \em{et~al.}(2007)Oh, Palsson, Park, Schilling, and
+  Mahadevan]{Oh:2007aa}
+Oh, Y.K.; Palsson, B.O.; Park, S.M.; Schilling, C.H.; Mahadevan, R.
+\newblock Genome-scale reconstruction of metabolic network in Bacillus subtilis
+  based on high-throughput phenotyping and gene essentiality data.
+\newblock {\em J Biol Chem} {\bf 2007},
+\newblock {\em 282},~28791--9.
 
-% Reset the counters -
-\setcounter{equation}{0}
-\setcounter{table}{0}
-\setcounter{figure}{0}
-\setcounter{page}{1}
+\bibitem[Feist \em{et~al.}(2009)Feist, Herrg{\aa}rd, Thiele, Reed, and
+  Palsson]{2009_feist_palsson_NatRevMicrobio}
+Feist, A.M.; Herrg{\aa}rd, M.J.; Thiele, I.; Reed, J.L.; Palsson, B.{\O}.
+\newblock Reconstruction of biochemical networks in microorganisms.
+\newblock {\em Nat Rev Microbiol} {\bf 2009},
+\newblock {\em 7},~129--43.
+
+\bibitem[Ibarra \em{et~al.}(2002)Ibarra, Edwards, and
+  Palsson]{2002_ibarra_edwards_palsson_Nat}
+Ibarra, R.U.; Edwards, J.S.; Palsson, B.O.
+\newblock Escherichia coli K-12 undergoes adaptive evolution to achieve in
+  silico predicted optimal growth.
+\newblock {\em Nature} {\bf 2002},
+\newblock {\em 420},~186--9.
+
+\bibitem[Schuetz \em{et~al.}(2007)Schuetz, Kuepfer, and
+  Sauer]{2007_schuetz_sauer_MolSysBio}
+Schuetz, R.; Kuepfer, L.; Sauer, U.
+\newblock Systematic evaluation of objective functions for predicting
+  intracellular fluxes in Escherichia coli.
+\newblock {\em Mol Syst Biol} {\bf 2007},
+\newblock {\em 3},~119.
+
+\bibitem[Hyduke \em{et~al.}(2013)Hyduke, Lewis, and
+  Palsson]{2013_hyduke_lewis_palsson_MolBioSys}
+Hyduke, D.R.; Lewis, N.E.; Palsson, B.{\O}.
+\newblock Analysis of omics data with genome-scale models of metabolism.
+\newblock {\em Mol Biosyst} {\bf 2013},
+\newblock {\em 9},~167--74.
+
+\bibitem[McCloskey \em{et~al.}(2013)McCloskey, Palsson, and
+  Feist]{2013_mccloskey_palsson_feist_MolSysBio}
+McCloskey, D.; Palsson, B.{\O}.; Feist, A.M.
+\newblock Basic and applied uses of genome-scale metabolic network
+  reconstructions of Escherichia coli.
+\newblock {\em Mol Syst Biol} {\bf 2013},
+\newblock {\em 9},~661.
+
+\bibitem[Zomorrodi \em{et~al.}(2012)Zomorrodi, Suthers, Ranganathan, and
+  Maranas]{2012_zomorrodi_maranas_MetaEng}
+Zomorrodi, A.R.; Suthers, P.F.; Ranganathan, S.; Maranas, C.D.
+\newblock Mathematical optimization applications in metabolic networks.
+\newblock {\em Metab Eng} {\bf 2012},
+\newblock {\em 14},~672--86.
+
+\bibitem[Jewett \em{et~al.}(2008)Jewett, Calhoun, Voloshin, Wuu, and
+  Swartz]{Jewett:2008aa}
+Jewett, M.C.; Calhoun, K.A.; Voloshin, A.; Wuu, J.J.; Swartz, J.R.
+\newblock An integrated cell-free metabolic platform for protein production and
+  synthetic biology.
+\newblock {\em Mol Syst Biol} {\bf 2008},
+\newblock {\em 4},~220.
+
+\bibitem[MATTHAEI and NIRENBERG(1961)]{MATTHAEI:1961aa}
+MATTHAEI, J.H.; NIRENBERG, M.W.
+\newblock Characteristics and stabilization of DNAase-sensitive protein
+  synthesis in E. coli extracts.
+\newblock {\em Proc Natl Acad Sci U S A} {\bf 1961},
+\newblock {\em 47},~1580--8.
+
+\bibitem[NIRENBERG and MATTHAEI(1961)]{NIRENBERG:1961aa}
+NIRENBERG, M.W.; MATTHAEI, J.H.
+\newblock The dependence of cell-free protein synthesis in E. coli upon
+  naturally occurring or synthetic polyribonucleotides.
+\newblock {\em Proc Natl Acad Sci U S A} {\bf 1961},
+\newblock {\em 47},~1588--602.
+
+\bibitem[Lu \em{et~al.}(2014)Lu, Welsh, and Swartz]{Lu:2014aa}
+Lu, Y.; Welsh, J.P.; Swartz, J.R.
+\newblock Production and stabilization of the trimeric influenza hemagglutinin
+  stem domain for potentially broadly protective influenza vaccines.
+\newblock {\em Proc Natl Acad Sci U S A} {\bf 2014},
+\newblock {\em 111},~125--30.
+
+\bibitem[Hodgman and Jewett(2012)]{Hodgman:2012aa}
+Hodgman, C.E.; Jewett, M.C.
+\newblock Cell-free synthetic biology: thinking outside the cell.
+\newblock {\em Metab Eng} {\bf 2012},
+\newblock {\em 14},~261--9.
+
+\bibitem[Morris \em{et~al.}(2011)Morris, Saez-Rodriguez, Clarke, Sorger, and
+  Lauffenburger]{Morris:2011aa}
+Morris, M.K.; Saez-Rodriguez, J.; Clarke, D.C.; Sorger, P.K.; Lauffenburger,
+  D.A.
+\newblock Training signaling pathway maps to biochemical data with constrained
+  fuzzy logic: quantitative analysis of liver cell responses to inflammatory
+  stimuli.
+\newblock {\em PLoS Comput Biol} {\bf 2011},
+\newblock {\em 7},~e1001099.
+
+\bibitem[Covert \em{et~al.}(2001)Covert, Schilling, and Palsson]{Covert:2001aa}
+Covert, M.W.; Schilling, C.H.; Palsson, B.
+\newblock Regulation of gene expression in flux balance models of metabolism.
+\newblock {\em J Theor Biol} {\bf 2001},
+\newblock {\em 213},~73--88.
+
+\bibitem[Covert \em{et~al.}(2004)Covert, Knight, Reed, Herrgard, and
+  Palsson]{Covert:2004aa}
+Covert, M.W.; Knight, E.M.; Reed, J.L.; Herrgard, M.J.; Palsson, B.O.
+\newblock Integrating high-throughput and computational data elucidates
+  bacterial networks.
+\newblock {\em Nature} {\bf 2004},
+\newblock {\em 429},~92--6.
+
+\bibitem[Varner(2000)]{Varner:2000aa}
+Varner, J.D.
+\newblock Large-scale prediction of phenotype: concept.
+\newblock {\em Biotechnol Bioeng} {\bf 2000},
+\newblock {\em 69},~664--78.
+
+\bibitem[Song and Ramkrishna(2012)]{Song:2012aa}
+Song, H.S.; Ramkrishna, D.
+\newblock Prediction of dynamic behavior of mutant strains from limited
+  wild-type data.
+\newblock {\em Metab Eng} {\bf 2012},
+\newblock {\em 14},~69--80.
+
+\bibitem[Song and Ramkrishna(2011)]{Song:2011aa}
+Song, H.S.; Ramkrishna, D.
+\newblock Cybernetic models based on lumped elementary modes accurately predict
+  strain-specific metabolic function.
+\newblock {\em Biotechnol Bioeng} {\bf 2011},
+\newblock {\em 108},~127--40.
+
+\bibitem[Gadkar \em{et~al.}(2003)Gadkar, Doyle, Crowley, and
+  Varner]{Gadkar:2003aa}
+Gadkar, K.G.; Doyle, 3rd, F.J.; Crowley, T.J.; Varner, J.D.
+\newblock Cybernetic model predictive control of a continuous bioreactor with
+  cell recycle.
+\newblock {\em Biotechnol Prog} {\bf 2003},
+\newblock {\em 19},~1487--97.
+
+\bibitem[Brown and Sethna(2003)]{Brown:2003aa}
+Brown, K.S.; Sethna, J.P.
+\newblock Statistical mechanical approaches to models with many poorly known
+  parameters.
+\newblock {\em Phys Rev E Stat Nonlin Soft Matter Phys} {\bf 2003},
+\newblock {\em 68},~021904.
+
+\bibitem[Machta \em{et~al.}(2013)Machta, Chachra, Transtrum, and
+  Sethna]{Machta:2013aa}
+Machta, B.B.; Chachra, R.; Transtrum, M.K.; Sethna, J.P.
+\newblock Parameter space compression underlies emergent theories and
+  predictive models.
+\newblock {\em Science} {\bf 2013},
+\newblock {\em 342},~604--7.
+
+\bibitem[Luan \em{et~al.}(2007)Luan, Zai, and Varner]{Luan:2007aa}
+Luan, D.; Zai, M.; Varner, J.D.
+\newblock Computationally derived points of fragility of a human cascade are
+  consistent with current therapeutic strategies.
+\newblock {\em PLoS Comput Biol} {\bf 2007},
+\newblock {\em 3},~e142.
+
+\bibitem[Song and Varner(2009)]{Song:2009aa}
+Song, S.O.; Varner, J.
+\newblock Modeling and analysis of the molecular basis of pain in sensory
+  neurons.
+\newblock {\em PLoS One} {\bf 2009},
+\newblock {\em 4},~e6758.
+
+\bibitem[Tasseff \em{et~al.}(2010)Tasseff, Nayak, Salim, Kaushik, Rizvi, and
+  Varner]{Tasseff:2010aa}
+Tasseff, R.; Nayak, S.; Salim, S.; Kaushik, P.; Rizvi, N.; Varner, J.D.
+\newblock Analysis of the molecular networks in androgen dependent and
+  independent prostate cancer revealed fragile and robust subsystems.
+\newblock {\em PLoS One} {\bf 2010},
+\newblock {\em 5},~e8864.
+
+\bibitem[Tasseff \em{et~al.}(2011)Tasseff, Nayak, Song, Yen, and
+  Varner]{Tasseff:2011aa}
+Tasseff, R.; Nayak, S.; Song, S.O.; Yen, A.; Varner, J.D.
+\newblock Modeling and analysis of retinoic acid induced differentiation of
+  uncommitted precursor cells.
+\newblock {\em Integr Biol (Camb)} {\bf 2011},
+\newblock {\em 3},~578--91.
+
+\bibitem[Nayak \em{et~al.}(2011)Nayak, Siddiqui, and Varner]{Nayak:2011aa}
+Nayak, S.; Siddiqui, J.K.; Varner, J.D.
+\newblock Modelling and analysis of an ensemble of eukaryotic translation
+  initiation models.
+\newblock {\em IET Syst Biol} {\bf 2011},
+\newblock {\em 5},~2.
+
+\bibitem[Lequieu \em{et~al.}(2011)Lequieu, Chakrabarti, Nayak, and
+  Varner]{Lequieu:2011aa}
+Lequieu, J.; Chakrabarti, A.; Nayak, S.; Varner, J.D.
+\newblock Computational modeling and analysis of insulin induced eukaryotic
+  translation initiation.
+\newblock {\em PLoS Comput Biol} {\bf 2011},
+\newblock {\em 7},~e1002263.
+
+\bibitem[Berg \em{et~al.}(2002)Berg, Tymoczko, and Stryer]{BERG_BIOCHEMISTRY}
+Berg, J.M.; Tymoczko, J.L.; Stryer, L.
+\newblock {\em Biochemistry}; W.H. Freeman,
+\newblock  2002.
+
+\bibitem[Peskov \em{et~al.}(2008)Peskov, Goryanin, and Demin]{Peskov:2008aa}
+Peskov, K.; Goryanin, I.; Demin, O.
+\newblock Kinetic model of phosphofructokinase-1 from Escherichia coli.
+\newblock {\em J Bioinform Comput Biol} {\bf 2008},
+\newblock {\em 6},~843--67.
+
+\bibitem[Keseler \em{et~al.}(2013)Keseler, Mackie, Peralta-Gil,
+  Santos-Zavaleta, Gama-Castro, Bonavides-Mart{\'\i}nez, Fulcher, Huerta,
+  Kothari, Krummenacker, Latendresse, Mu{\~n}iz-Rascado, Ong, Paley,
+  Schr{\"o}der, Shearer, Subhraveti, Travers, Weerasinghe, Weiss,
+  Collado-Vides, Gunsalus, Paulsen, and Karp]{Keseler:2013aa}
+Keseler, I.M.; Mackie, A.; Peralta-Gil, M.; Santos-Zavaleta, A.; Gama-Castro,
+  S.; Bonavides-Mart{\'\i}nez, C.; Fulcher, C.; Huerta, A.M.; Kothari, A.;
+  Krummenacker, M.; Latendresse, M.; Mu{\~n}iz-Rascado, L.; Ong, Q.; Paley, S.;
+  Schr{\"o}der, I.; Shearer, A.G.; Subhraveti, P.; Travers, M.; Weerasinghe,
+  D.; Weiss, V.; Collado-Vides, J.; Gunsalus, R.P.; Paulsen, I.; Karp, P.D.
+\newblock EcoCyc: fusing model organism databases with systems biology.
+\newblock {\em Nucleic Acids Res} {\bf 2013},
+\newblock {\em 41},~D605--12.
+
+\bibitem[Huang \em{et~al.}(2014)Huang, Mou, Shen, Lu, Li, Liu, Wang, Li, Geng,
+  Liu, Wu, Chen, and Zhang]{Huang:2014aa}
+Huang, Z.; Mou, L.; Shen, Q.; Lu, S.; Li, C.; Liu, X.; Wang, G.; Li, S.; Geng,
+  L.; Liu, Y.; Wu, J.; Chen, G.; Zhang, J.
+\newblock ASD v2.0: updated content and novel features focusing on allosteric
+  regulation.
+\newblock {\em Nucleic Acids Res} {\bf 2014},
+\newblock {\em 42},~D510--6.
+
+\bibitem[Kennedy and Eberhart(1995)]{PSO}
+Kennedy, J.; Eberhart, R.
+\newblock Particle swarm optimization.
+\newblock  Proceedings of the International Conference on Neural Networks,
+  1995,
+\newblock pp. 1942--1948.
+
+\end{thebibliography}
 

### 2015-02-16 revision_v3/Paper_v3.tex
Submission of revised v2

--- 2015-02-03 revision_v2/Paper_v3.tex
@@
 
-The proposed modeling framework also differs appreciably from previously established non-mechanistic, kinetic approximations of complex biochemical network behavior. 
-Such frameworks replace parameter dense mechanistic kinetic expressions with heuristics quantifying the relationship between metabolic rate and metabolite effectors. 
-These approaches arose in response to uncertainties associated with obtaining correct mechanistic kinetic expressions and parameters of \textit{in vivo} systems. 
-Similarly, available kinetic parameters measured \textit{in vitro} may differ in a specialized cell-free \textit{in vitro} environments. 
-Factors affecting kinetics, such as enzyme channeling, macromolecular crowding, and pH, are likely dramatically different in cell-free environments than in both \textit{in vivo} systems as well as typical \textit{in vitro} conditions used for parameter measurements.
-Thus, a more generalized, approximate biochemical reaction network formulation may be desirable in the case of cell-free systems. 
-Our approach is similar to generalized mass action-based power law formulations of Savageau and colleagues \cite{1991_savageau_JTheorBio} and linlog kinetics of Visser \textit{et al.} \cite{2003_visser_heijnen_MetabEng} in that metabolic rates are proportional to corresponding enzyme levels modified by metabolite effectors.
-Power law and linlog approaches suffer from several known limitations \cite{2005_heijnen_BiotechBioeng}. 
-Power law reaction rates do not capture saturation effects and become infinite for small concentrations of inhibitory effectors. 
-Linlog kinetics also become ill-defined when effector concentrations go to zero. 
-Our framework does not suffer from such drawbacks. 
-Also, models employing linlog kinetics typically rely on an experimentally determined reference state to describe dynamics taking place after a perturbation to a steady-state. 
-Our framework is not restricted to such a requirement. 
-Cell-free systems are unlikely to satisfy such a steady-state approximation after extract preparation and prior to culture initiation. 
-Our framework is similar to the generic kinetic formulation from Hadlich \textit{et al.} \cite{2009_hadlich_wiechert_MetabEng}, but differs in its inclusion of cooperative effects as well as proposes a simplified integration of competition amongst allosteric effectors using max/min rules.
-In summary, our proposed framework offers an effective kinetic approximation that captures saturation effects and allosteric competition within cell-free systems that may also be extensible to \textit{in vivo} metabolic and gene regulatory networks.
-
 There are several critical questions that should be explored following this proof-of-concept study.

### 2015-02-20 processes-69103_review_response/processes-69103.tex
Changes in discussion

@@
 
+A number of mathematical models and computational simulations concerning coagulation dynamics have been based on the Hockin-Mann model of coagulation.
+
 The performance of the proof of principle coagulation model was impressive given its limited size.
 The most detailed mechanistic model of coagulation is the one by Luan and co-workers built  with 193 proteins and protein complexes that are interconnected by 301 reactions \citep{2010_luan_varner_MolBioSys}.
-Akin to the hybrid model this mechanistic used the normal thrombin data and hemophilia data from Allen \emph{et al.} \citep{ALLEN2006}.
+Akin to the hybrid model this model used normal thrombin data and hemophilia data from Allen \emph{et al.} \citep{ALLEN2006}.
 However unlike the mechanistic model we used this data for validation rather than training.

### 2015-02-20 processes-69103_review_response/processes-69103.tex
Changes to discussion

@@
 A number of mathematical models and computational simulations concerning coagulation dynamics have been based on the Hockin-Mann model of coagulation.
+Recently, Brummel-Ziedins and co-workers built a model that incorporated the Protein C (PC) pathway into the Hockin-Mann network to investigate thrombin generation in cases of familial PC deficiency.
+Computational simulations using this model showed that PC mutation caused an elevation in peak thrombin levels without changing the initiation time or the slope of thrombin generation curve.
+Our hybrid model to a certain extent captured this trend qualitatively.
+We were able to show the decrease in peak thrombin concentration in the presence of PC pathway without any other inhibitors like ATIII.
+Mitrophanov et al.
+used kinetic modeling to study the effect of rfVIIa on thrombin generation.
+This study showed that increasing amounts of FVIIa accelerated the initiation phase and the propagation phase during thrombin generation.
+Analysis of thrombin generation curves showed that the FVIIa controlled the maximal slope of thrombin curve along with peak thrombin time and initiation time.
+Though we had issues in capturing the effect of FVIIa on initiation time we were able to show that change in concentration of FVIIa affected the maximal thrombin slope as well as propagation phase (Fig 7).
 

### 2015-02-21 processes-69103_review_response/processes-69103.tex
Some extra changes

--- 2015-02-20 processes-69103_review_response/processes-69103.tex
@@
 Using a set of four ordinary differential equations they were able to derive an equation for temporal evolution of thrombin generation.
-However the model focuses more on thrombin generation rather than dynamics of coagulation.
-Unlike our hybrid approach it does not consider regulatory detail and thus is unable to capture the roles of primary inhibitors like ATIII or protein C.
+However the model has a greater focus on thrombin generation and lesser emphasis on the influence of various regulatory elements on the dynamics of coagulation.
+Unlike our hybrid approach it does not model regulatory detail and thus may be unable to capture the roles of primary inhibitors like ATIII or protein C.
 Thus in cases where we need to understand the dynamics of coagulation without complex modeling or when there is incomplete mechanistic knowledge, the hybrid approach is a better candidate.
@@
 However platelet modeling is a significant advantage of this model over the hybrid approach.
-Future versions of the hybrid model can involve modeling platelets as done by Papdopoulos et al.
-in their phenomenological model.
+Future versions of the hybrid model can involve modeling platelets as done by Papdopoulos \emph{et al.}, in their phenomenological model.
 Other laboratories have also expanded upon Hockin \emph{et al.}, for example by exploring the intrinsic pathway, the role of stochastic fluctuations in coagulation \citep{Lo:2005aa}, and the dynamics of thrombin mediated clot formation \citep{Chatterjee:2010aa}.
@@
 However, defining this threshold, which is likely patient specific, is difficult as there is tremendous patient to patient variability even with a normal
-coagulation phenotype \citep{Danforth:2012aa}.
+co agulation phenotype \citep{Danforth:2012aa}.
 
 A number of mathematical models and computational simulations concerning coagulation dynamics have been based on the Hockin-Mann model of coagulation.
+Recently, Brummel-Ziedins and co-workers built a model that incorporated the Protein C (PC) pathway into the Hockin-Mann network to investigate thrombin generation in cases of familial PC deficiency \cite{BrummelPC:2012}.
+Computational simulations using this model showed that PC mutation caused an elevation in peak thrombin levels without changing the initiation time or the slope of thrombin generation curve.
+Our hybrid model to a certain extent captured this trend qualitatively.
+We were able to show the decrease in peak thrombin concentration in the presence of PC pathway without any other inhibitors like ATIII.
+Danforth \emph{et al.} computationally simulated normal thrombin generation using 5pM tissue factor with all other factors at their mean physiological level \citep{Danforth:2012aa}.
+The initiation time in this simulation was approximately 4.4 minutes.
+When predicting the normal thrombin generation curve using 0.2 nM of FVIIa we were able to show an approximate initiation time around 5 minutes with a similar thrombin generation profile (Fig.
+\ref{fig-hemophilia}A).
+In a kinetic modeling to study the effect of rfVIIa on thrombin generation \cite{Mitrophanov:2011} it was shown that increasing amounts of FVIIa accelerated the initiation phase and the propagation phase during thrombin generation.
+Analysis of thrombin generation curves showed that the FVIIa controlled the maximal slope of thrombin curve along with peak thrombin time and initiation time.
+Though we had issues in capturing the effect of FVIIa on initiation time we were able to show that change in concentration of FVIIa affected the maximal thrombin slope as well as propagation phase (Fig.
+\ref{fig-rFVII}A).
 
@@
 
+\bibitem{BrummelPC:2012}
+Kathleen E.
+Brummel-Ziedins,Thomas Orfeo,Peter W.
+Callas, Matthew Gissel, Kenneth G.
+Mann, Edwin G.
+Bovill (2012) The Prothrombotic Phenotypes in Familial Protein C Deficiency Are Differentiated by Computational Modeling of Thrombin Generation
+\newblock PLoS One 5: e44378.
+
+\bibitem{Mitrophanov:2011}
+Alexander Y. Mitrophanov, Jaques Reifman(2011) Kinetic modeling sheds light on the mode of action of recombinant factor VIIa on thrombin generation
+\newblock Thrombosis Research Volume 128, Issue 4, October 2011, Pages 381-390.
+
 \end{thebibliography}

### 2015-04-23 Paper_v1.tex
Updating

@@
 \begin{titlepage}
-{\par\centering\textbf{\Large Dynamic Modeling of Cell-Free Biochemical Networks using Effective Kinetic Models}}
+{\par\centering\textbf{\Large A novel metaheuristic approach for parameter estimation in high dimensional biochemical networks}}
 \vspace{0.05in}
-{\par \centering \large{Joseph A. Wayman, Adithya Sagar and Jeffrey D. Varner$^{*}$}}
+{\par \centering \large{Adithya Sagar, Christine Shoemaker and Jeffrey D. Varner$^{*}$}}
 \vspace{0.10in}
@@
 \vspace{0.1in}
-{\par \centering \textbf{Running Title:}~Effective models of metabolism}
+{\par \centering \textbf{Running Title:}~Parameter estimation in large scale biological systems}
 \vspace{0.1in}
-{\par \centering \textbf{To be submitted:}~\textit{Processes}}
+{\par \centering \textbf{To be submitted:}~\emph{Biotechnology Journal}}
 \vspace{0.5in}
@@
 \section*{Abstract}
-Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
-Many of the challenges confronting genome-scale kinetic modeling can potentially be overcome in a cell-free system. 
-For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth. 
-Thus, cell-free operation holds several significant advantages for model development, identification and validation. 
-Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms, such as \textit{E.
-coli}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.
-Toward this unmet need, we present an effective biochemical network modeling framework for building dynamic cell-free metabolic models.
-The key innovation of our approach is the integration of simple effective rules encoding complex allosteric regulation with traditional kinetic pathway modeling. 
-We tested our approach by modeling the time evolution of several hypothetical cell-free metabolic networks. 
-We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, captured complex allosteric patterns such as ultrasensitivity or non-competitive inhibition in the absence of mechanistic information.
-Second, when integrated into network models, these rules captured classic regulatory patterns such as product-induced feedback inhibition. 
-Lastly, we showed, at least for the network architectures considered here, that we could simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data. 
-While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.
+The problem of parameter estimation presents a significant challenge in the modeling of large complex systems.
+Mathematical modeling of biological systems is one such area where parameter estimation is a difficult non-linear optimization problem.
+This problem is further compounded when dealing with parameter vectors of high dimensions.
+Such problems generally involve expensive function evaluations and it is hard to obtain optimal or near optimal solutions within finite computational time.
+In this study we present a novel meta-heuristic approach that combines elements of multi-swarm optimization along with dynamically dimensioned search (DDS) to obtain optimal or near optimal solutions of high dimensional biochemical networks within a relatively few function evaluations.
+We use a multi-swarm optimization technique to generate candidate vectors which are then greedily updated using DDS by dynamically varying the perturbed parameter dimensions.
+We tested (25 trials with 4000 function evaluations in each trial) this algorithm on a biochemical network of coagulation (148 parameters and 92 species) and compared it's performance against other metaheuristics like Differential Evolution (DE), Particle Swarm Optimization (PSO), Simulated Annealing (SA) and also against DDS alone.
+The new algorithm outperforms all the other metaheuristics on the coagulation model.
+The parameter vectors obtained using this approach fit the experimental data well and also make accurate enough predictions on unseen experimental data.
+We also performed this comparison on commonly used test functions for global optimization and found the same behavior.
+Further we used two benchmark problems, a genome wide kinetic model with 1759 parameters and a metabolic model of Chinese Hamster Ovary cells with 117 parameters to evaluate the performance of our approach.
+We  surprisingly outperformed the enhanced scatter search algorithm on these benchmarks and obtained the nominal parameter vector with just 4000 function evaluations.
 
-{\noindent \textbf{Keywords:}~Cell-free metabolism, Mathematical modeling}
+{\noindent \textbf{Keywords:}~Parameter identification, Mathematical modeling}
 
@@
 \section*{Introduction}
-Mathematical modeling has long contributed to our understanding of metabolism. 
-Decades before the genomics revolution, mechanistically, structured metabolic models arose from the desire to predict microbial phenotypes resulting from changes in intracellular or extracellular states \citep{1976_fredrickson_BiotechBioeng}.
-The single cell \textit{E.
-coli} models of Shuler and coworkers pioneered the construction of large-scale, dynamic metabolic models that incorporated multiple, regulated catabolic and anabolic pathways constrained by experimentally determined kinetic parameters \citep{1984_domach_shuler_BiotechBioeng_01}.
-Shuler and coworkers generated many single cell kinetic models, including single cell models of eukaryotes \citep{1989_steinmeyer_shuler_ChemEngSci,1992_wu_shuler_AnnNYAcadSci}, minimal cell architectures \citep{2004_castellanos_shuler_PNAS}, as well as DNA sequence based whole-cell models of \textit{E.
-coli} \citep{2008_atlas_shuler_IETSysBio}.
-Conversely, highly abstracted kinetic frameworks, such as the cybernetic framework, represented a paradigm shift, viewing cells as growth-optimizing strategists \citep{1985_dhurjati_ramkrishna_tsao_BiotechBioeng}.
-Cybernetic models have been highly successful at predicting metabolic choice behavior, e.g., diauxie behavior \citep{1986_kompala_ramkrishna_tsao_BiotechBioeng}, steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, as well as the cellular response to metabolic engineering modifications \citep{1999_varner_ramkrishna_MetaEng}.
-Unfortunately, cybernetic models also suffer from an identifiability challenge, as both the kinetic parameters and an abstracted model of cellular objectives must be estimated simultaneously.
 
-In the post genomics world, large-scale stoichiometric reconstructions of microbial metabolism popularized by static, constraint-based modeling techniques such as flux balance analysis (FBA) have become standard tools \citep{2012_lewis_palsson_NatRevMicrobio}.
-Since the first genome-scale stoichiometric model of \textit{E.
-coli}, developed by Edwards and Palsson \citep{2000_edwards_palsson_PNAS}, well over 100 organisms, including industrially important prokaryotes such as \textit{E.
-coli} \citep{Feist:2007aa} or \textit{B.
-subtilis} \citep{Oh:2007aa}, are now available \citep{2009_feist_palsson_NatRevMicrobio}.
-Stoichiometric models rely on a pseudo-steady-state assumption to reduce unidentifiable genome-scale kinetic models to an underdetermined linear algebraic system, which can be solved efficiently even for large systems.
-Traditionally, stoichiometric models have also neglected explicit descriptions of metabolic regulation and control mechanisms, instead opting to describe the choice of pathways by prescribing an objective function on metabolism.
-Interestingly, similar to early cybernetic models, the most common metabolic objective function has been the optimization of biomass formation \citep{2002_ibarra_edwards_palsson_Nat}, although other metabolic objectives have also been estimated \citep{2007_schuetz_sauer_MolSysBio}.
-Recent advances in constraint-based modeling have overcome the early shortcomings of the platform, including capturing metabolic regulation and control \citep{2013_hyduke_lewis_palsson_MolBioSys}. 
-Thus, modern constraint-based approaches have proven extremely useful in the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \citep{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}.
-However, genome-scale kinetic models of industrial important organisms such as \textit{E.
-coli} have yet to be constructed.
+Biological systems interact with one another and with the external world through highly complex biochemical networks.
+The advent of  \textit {omics} era has provided researchers with a deluge of data about these networks.
+Over the last decade, with an increase in computational power, large scale mathematical modeling of biological systems has evolved as a powerful paradigm to understand this data deluge and to effectively characterize these complex biochemical interactions.
 
-Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
-Central amongst these advantages is direct access to metabolites and the microbial biosynthetic machinery without the interference of a cell wall. 
-This allows us to control as well as interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
-Second, cell-free systems also allow us to study biological processes without the complications associated with cell growth.
-Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \citep{Jewett:2008aa}. 
-However, CFPS is not new; CFPS in crude \textit{E. coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \citep{MATTHAEI:1961aa,NIRENBERG:1961aa}. 
-Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \citep{Lu:2014aa} to synthetic biology \citep{Hodgman:2012aa}. 
-Interestingly, many of the challenges confronting genome-scale kinetic modeling can potentially be overcome in a cell-free system. 
-For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth. 
-Thus, cell-free operation holds several significant advantages for model development, identification and validation.
-Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms, such as \textit{E.
-coli} or \textit{B.
-subtilis}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.
+The development of such mathematical models for biochemical networks can be broadly decomposed into two tasks a) Model construction b) Model calibration or Parameter Estimation.
+Model construction involves translating existing knowledge about the network (i.e.
+entities in the network, interactions amongst the entities, types of interactions) along with reasonable assumptions into an appropriate mathematical
+framework.  Some popular mathematical formulations have been Ordinary Differential Equations (ODEs), Boolean and fuzzy logic, probabilistic graphical models etc. [REFHERE]
 
-In this study, we present an effective biochemical network modeling framework for building dynamic cell-free metabolic models.
-The key innovation of our approach is the seamless integration of simple effective rules encoding complex regulation with traditional kinetic pathway modeling. 
-This integration allows the description of complex regulatory interactions, such as time-dependent allosteric regulation of enzyme activity, in the absence of specific mechanistic information. 
-The regulatory rules are easy to understand, easy to formulate and do not rely on overarching theoretical abstractions or restrictive assumptions. 
-We tested our approach by modeling the time evolution of several hypothetical cell-free metabolic networks. 
-In particular, we tested whether our effective modeling approach could describe classically expected enzyme kinetic behavior, and second whether we could simultaneously estimate kinetic parameters and regulatory connectivity, in the absence of specific mechanistic knowledge, from synthetic experimental data.
-Toward these questions, we explored five hypothetical cell-free networks. 
-Each network shared the same enzymatic connectivity, but had different allosteric regulatory connectivity. 
-We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, captured complex allosteric patterns such as ultrasensitivity or non-competitive inhibition in the absence of mechanistic information.
-Second, when integrated into network models, these rules captured classical regulatory patterns such as product-induced feedback inhibition. 
-Lastly, we showed, at least for the network architectures considered here, that we could simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data. 
-While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.
+Model calibration or parameter estimation follows the model construction step.
+In this step the model is trained against experimental data to determine the model parameters that give a good fit.
+After model design and calibration, validation is done wherein the model output is validated against unseen experimental data.
+If validation leads to unsatisfactory results model design and calibration are performed once again.
+This process is done iteratively till successful results are obtained from validation step.
+\\
+
+The objective function in Equation 5 is minimized using appropriate optimization technique to estimate the model parameters.
+These problems are generally multi-modal i.e.
+contain multiple local optima traditional.
+Hence we cannot use local optimization techniques cannot be used to solve these problems.
+In many of these problems derivative information is absent or the objective function may be discontinuous.
+This hinders the use of deterministic optimization techniques that require the problem to be convex and other derivative/gradient based optimization methods.
+Most of these methods are computationally expensive and for optimization of large-scale ODE systems, this becomes prohibitively expensive.
+Heuristic optimization techniques like genetic algorithm and differential evolution offer an attractive option in finding globally optimum solution or close to global optimum solution.
+However most of the population-based heuristics take a large number of function evaluations and have computationally complex operations, which reduces their effectiveness in finding feasible solutions with limited computational resources and time.
+
+This study tries to overcome the drawbacks mentioned above and presents a novel heuristics based search strategy/algorithm for parameter estimation in large-scale ODE models of biochemical networks with many unknown parameters.
+
+\subsection *{Parameter Estimation Literature}
+
+\subsection *{Modeling methodologies} 
+ Shuler and co-workers constructed large-scale dynamic metabolic models of E.coli that were based ordinary differential equations (ODEs).
+Various aspects of E.colis cellular processes like RNA synthesis, chromosome synthesis, cellular energy expenditure etc.
+were modeled using coupled non-linear ordinary differential equations.
+Post genomic era saw the rise of flux balance analysis (FBA), a static, constraint based modeling approach for microbial metabolism.
+FBA assumes a pseudo steady-state and reduces genome-scale kinetic models to a set of static mass balances.
+These mass balances are represented using an underdetermined set of linear algebraic equations.
+The advantage is such a system of equations can be solved efficiently for very large systems.
+
+The dynamics of signal transduction systems are commonly modeled using ODEs [Refs]-----.
+One of the advantages of using ODEs is that the dynamics of large-scale systems can be described with a great degree of mechanistic detail.
+Over the last half a decade we have seen the construction of increasingly large-scale ODE models of signal transduction systems.
+Chen et al.
+built a mass action model of ErbB signaling pathways with 499 ODEs and 229 parameters.
+Tasseff et al analyzed retinoic acid induced differentiation of uncommitted precursor cells using a dynamic ODE model with 729 proteins and protein complexes interconnected with 1356 interconnections.
+Luan et al captured the key aspects of coagulation using an ODE model with 193 proteins and protein complexes with 301 kinetic parameters.
+
+  Boolean logic, probabilistic modeling and fuzzy logic are other mathematical frameworks that are used for mechanistic modeling of signal transduction systems.
+Choi et al modeled the p53 signaling network that affects cellular response to DNA damage using Boolean logic.
+------.
+In Boolean modeling of signal transduction systems the network entities i.e.
+proteins assume a value of 0 or 1 based on their activation levels.
+Fuzzy logic models offer a greater detail by allowing protein activation to take on a range of discrete values between 0 and 1.
+Mitsos et al.
+used the constrained fuzzy logic approach (cFL) wherein protein activation takes real values and a transfer function (TF) is used to propagate the signal along the network.
+Boolean and fuzzy logic models offer a relatively simpler mathematical formulation as compared to ODE models and are not encumbered with a large number of parameters.
+However these models are static in nature and offer a more qualitative view of the system.
+In addition to these deterministic formulations, stochastic modeling is another way of representing biochemical networks.
+Stochastic models are primarily based on stochastic differential equations also called Langevin equations to describe the dynamics of biochemical network ----- [Refs].
+Stochastic models are limited to describing dynamics of small-scale biochemical networks and are computationally very expensive to characterize even in very small biological systems.
 
@@
 
-\section*{Results}
+\section*{Results}\vspace{-12pt}
 
-\subsection*{Formulation and properties of effective cell-free metabolic models.}
-We developed two proof-of-concept metabolic networks to investigate the features of our effective biochemical network modeling approach (Fig. \ref{fig-networks}). 
-In both examples, substrate S was converted to the end products P$_{1}$ and P$_{2}$ through a series of enzymatically catalyzed reactions, including a branch point at hypothetical metabolite M$_{2}$. 
-Several of these reactions involved cofactor dependence (AH or A), and various allosteric regulatory mechanisms modified the activity of pathway enzymes. 
-Network A included feedback inhibition of the initial pathway enzyme ($E_{1}$) by pathway end products P$_{1}$ and P$_{2}$ (Fig. \ref{fig-networks}A). 
-On the other hand, network B involved feedback inhibition of $E_{1}$ by P$_{2}$ and $E_{6}$ by P$_{1}$ (Fig. \ref{fig-networks}B). 
-In both networks, branch point enzymes $E_{3}$ and $E_{6}$ were subject to feed-forward activation by reduced cofactor AH. 
-Lastly, it is known experimentally that cell-free systems have a finite operational lifespan. 
-Loss of biosynthetic capability could be a function of many factors, e.g., cofactor or metabolite limitations. 
-We modeled the loss of biosynthetic capability as a non-specific first-order decay of enzyme activity.
+Coagulation is an archetype biochemical network that is highly interconnected and tightly regulated with multiple positive and negative feedback loops.
+The biochemistry underlying coagulation, though quite complex has been well studied [REFHERE], and reliable experimental coagulation models have been developed [REFHERE].
+This makes it an ideal system for mathematical modeling and parameter estimation.
+Coagulation is regulated by a set of serine proteases also known as coagulation factors and blood platelets.
+The coagulation factors are generally in an inactive state and are known as zymogens.
+These zymogens are activated through certain triggers.
+These trigger events like injury or trauma or sepsis expose factors like collagen, tissue factor and von Willebrand factor (vWF) to blood.
+The exposure of these factors to blood kick-starts a series of convergent cascades that lead to conversion of zymogen prothrombinase to thrombin.
+Luan et al.
+modeled coagulation using coupled non-linear ordinary differential equations with 148 reactions and 92 species [REFHERE].
+The model was validated using 21 published datasets.
+The model parameters used in the simulation were drawn from various literature sources.
 
-Allosteric regulation of enzyme activity was modeled by combining individual regulatory contributions to the activity of pathway enzymes into a control coefficient using an integration rule (Fig.
-\ref{fig-control-schematic}).
-This strategy is similar in spirit to the Constrained Fuzzy Logic (cFL) approach of Lauffenburger and coworkers which has been used to effectively model signal transduction pathways important in human health \citep{Morris:2011aa}.
-In our formulation, Hill-like transfer functions $0 \leq f\left(\mathcal{Z}\right) \leq 1$ were used to calculate the influence of factor abundance upon target enzyme activity. 
-In this context, factors can be individual metabolite levels or some function, e.g., the product of metabolite levels. 
-However, more generally, factors can also correspond to non-modeled influences, categorial variables or other abstract quantities. 
-In the current study, we simply let $\mathcal{Z}$ correspond to the abundance of individual metabolites, however in general this can be a complex function of both modeled and unmodeled factors. 
-When an enzyme was potentially sensitive to more than one regulatory input, logical integration rules were used to select which regulatory transfer function influenced enzyme activity at any given time.
-Thus, our test networks involved important features such as cofactor recycling, enzyme activity and metabolite dynamics, as well as multiple overlapping allosteric regulatory mechanisms.
+We compared the performance of DDSMLSPSO on this model of coagulation against commonly used meta heuristics like Simulated Annealing (SA), Differential Evolution (DE), Particle Swarm Optimization (PSO) and also against DDS.
+To train the model parameters we used data sets from TF-VIIa initiated coagulation with no anticoagulants.
+The objective error function is a linear combination of two different error functions that used data sets representing coagulation initiated with different concentrations of TF-VIIa (5pm, 5nm).
+Since this is an expensive error function we restricted the number of function evaluations to 4000 for each algorithm.
+We performed 25 trials of this experiment.
+DDSMLSPSO exhibits a much faster rate of error convergence and has a much lower final error as compared to the other algorithms (Figure 3).
+Within the first 1000 function evaluations (swarm phase) of DDSMLSPSO there is a very rapid drop in error.
+Subsequently, again after 2500 function evaluations (dynamically dimensioned phase) the error drops quickly.
+Overall at the end of 4000 function evaluations DDSMLSPSO minimizes the error to much a greater extent than any of the other algorithms.
+Amongst the rest of the algorithms DDS and SA have a faster rate of drop in error as compared to PSO or DE.
+However after 1500 function evaluations, the objective error remains nearly constant.
 
-The rule-based regulatory strategy approximated the behavior of classical allosteric activation and inhibition mechanisms (Fig. \ref{fig-kinetics-simulations}). 
-We considered the enzyme catalyzed conversion of substrate S to a product P, where the overall reaction rate was modeled as the product of a Michaelis-Menten term and an effective allosteric control variable reflecting the particular regulatory interaction.
-We first explored feed-forward substrate activation of enzyme activity (for both positive and negative cooperativity). 
-Consistent with classical data, the rule-based strategy predicted a sigmoidal relationship between substrate abundance and reaction rate as a function of the cooperativity parameter (Fig.
-\ref{fig-kinetics-simulations}A).
-For cooperativity parameters less than unity, increased substrate abundance \textit{decreased} the reaction rate. 
-This was consistent with the idea that substrate binding \textit{decreased} at regulatory sites, which negatively impacted substrate binding at the active site. 
-On the other hand, as the cooperativity parameter increased past unity, the rate of conversion of substrate S to product P by enzyme $E$ approached a step function. 
-In the presence of an inhibitor, the rule-based strategy predicted non-competitive like behavior as a function of the cooperativity parameter (Fig. \ref{fig-kinetics-simulations}B). 
-When the control gain parameter, $\kappa_{ij}$ in Eqn.
-\eqref{eqn:control-factor}, was greater than unity, the inhibitory force was directly proportional to the cooperativity parameter, $\eta$ in Eqn.
-\eqref{eqn:control-factor}.
-Thus, as the cooperativity parameter increased, the maximum reaction rate decreased (Fig. \ref{fig-kinetics-simulations}B). 
-Interestingly, our rule-based approach was unable to directly simulate competitive inhibition of enzyme activity. 
-Taken together, the rule-based strategy captured classical regulatory patterns for both enzyme activation and inhibition. 
-Thus, we are able to model complex kinetic phenomena such as ultrasensitivity, despite an effective description of reaction kinetics.
+Figure 2 shows the fit between model predictions and experimental data.
+The solid lines represent the mean value of prediction over 25 trials and the shaded region represents the 99\% confidence interval.
+Figure 3 shows model predictions on completely 'unseen'  or untrained data sets where coagulation was initiated with 500pm, 50pm, 10pm concentrations of TF-VIIa respectively.
+The parameters that were obtained by simultaneous training on the 2 data sets were able to 'fit' the experimental data well within a few number of function evaluations.
 
-End product yield was controlled by feedback inhibition, while product selectivity was controlled by branch point enzyme inhibition (Fig. \ref{fig-onoff-simulations}). 
-A critical test of our modeling approach was to simulate networks with known behavior. 
-If we cannot reproduce the expected behavior of simple networks, then our effective modeling strategy, and particularly the rule-based approximation of allosteric regulation, will not be feasible for genome-scale cell-free problems.
-We considered two cases, control ON/OFF, for each network configuration.
-Each of these cases had identical kinetic parameters and initial conditions; the \textit{only} differences between the cases were the allosteric regulation rules and the control parameters associated with these rules.
-As expected, end product accumulation was larger for network A when the control was OFF (no feedback inhibition of $E_{1}$ by P$_{1}$ and P$_{2}$),
-as compared to the ON case (Fig. \ref{fig-onoff-simulations}A). 
-We found this behavior was robust to the choice of underlying kinetic parameters, as we observed that same qualitative response across an ensemble of randomized parameter sets (N = 100), for fixed control parameters.
-The control ON/OFF response of network B was more subtle. 
-In the OFF case, the behavior was qualitatively similar to network A. 
-However, for the ON case, flux was diverted away from P$_{2}$ formation by feedback inhibition of $E_{6}$ activity at the M$_{2}$ branch point by P$_{1}$ (Fig. \ref{fig-onoff-simulations}B). 
-Lower $E_{6}$ activity at the M$_{2}$ branch point allowed more flux toward P$_{1}$ formation, hence the yield of P$_{1}$ also increased (Fig. \ref{fig-onoff-simulations}C). 
-Again, the control ON/OFF behavior of network B was robust to changes in kinetic parameters, as the same qualitative trend was conserved across an ensemble of randomized parameters (N = 100), for fixed control parameters.
-Taken together, these simulations suggested that the rule-based allosteric control concept could robustly capture expected feedback behavior for networks with uncertain kinetic parameters.
+\subsection*{Performance on Benchmark functions}
 
-\subsection*{Estimating parameters and effective allosteric regulatory structures.}
-A critical challenge for any dynamic model is the estimation of kinetic parameters. 
-For metabolic processes, there is also the added challenge of identifying the regulation and control structures that manage metabolism. 
-Of course, these issues are not independent; any description of enzyme activity regulation will be a function of system state, which in turn depends upon the kinetic parameters. 
-For cell-free systems, regulated gene expression has been removed, however, enzyme activity regulation is still operational. 
-We explored this linkage by estimating model parameters from synthetic data using both network structures. 
-We generated noise corrupted synthetic measurements of the substrate S, intermediate M$_{5}$ and end product P$_1$ approximately every 20 min using
-network A.
-We then generated an ensemble of model parameter estimates by minimizing the difference between model simulations and the synthetic data using particle swarm optimization (PSO), starting from random initial parameter guesses.
-The estimation of kinetic parameters was sensitive to the choice of regulatory structure (Fig.
-\ref{fig-parameter-fit}).
-PSO identified an ensemble of parameters that bracketed the mean of the synthetic measurements in less than 1000 iterations when the control structure was correct (Fig.
-\ref{fig-parameter-fit}A and B).
-However, with control mismatch (network B simulated with network A parameters), model simulations were not consistent with the synthetic data  (Fig. \ref{fig-parameter-fit}C and D). 
-Taken together, these results suggested that we could perhaps simultaneously estimate both parameters and network control architectures, as incorrect control structures would be manifest as poor model fits.
+We compared the performance of these algorithms on commonly used test functions for global optimization.
+Figure-xxx shows the rate of error convergence on a 300D rastrigin function where DDSMLSPSO clearly outperforms the other approaches.
+In Figure-yyyy, on a 300D ackley function we see the same trend wherein  DDSMLSPSO reaches the minimum much faster than other algorithms.
 
-We modified our particle swarm identification strategy to simultaneously search over both kinetic parameters and putative control structures. 
-In addition to our initial networks, we constructed three additional presumptive network models, each with the same enzymatic connectivity but different allosteric regulation of the pathway enzymes (Fig.
-\ref{fig-alternative-networks}).
-We then initialized a population of particles, each with one of the five potential regulatory programs and randomized kinetic parameters. 
-Thus, we generated an initial population of particles that had \textit{both} different kinetic parameters as well as different control structures. 
-We biased the distribution of the particle population according to our \textit{a prior} belief of the correct regulatory program. 
-To this end, we considered three different priors, a uniform distribution where each putative regulatory structure represented 20\% of the population and two mixed distributions that were either positively or negatively biased towards the correct structure (network A).
-In both the positively biased and uniform cases the PSO clearly differentiated between the true or closely related structures and those that were materially different (Fig. \ref{fig-control-search}). 
-As expected, the positively biased population (40\% of the initial particle population seeded with network A) gave the best results, where the correct structure was preferentially identified (Fig.
-\ref{fig-control-search}A).
-On the other hand, when given a uniform distribution, the PSO approach identified a combination of network A and network C as the most likely control structures (Fig. \ref{fig-control-search}B). 
-Network A and C differ by the regulatory connection between the end product P$_2$ and enzyme $E_{1}$; in network A, end product P$_{2}$ was assumed to inhibit $E_{1}$, while in network C, end product P$_{2}$ activated $E_{1}$.
-Lastly, when the initial population was biased towards incorrect structures (initial population seeded with 90\% incorrect structures), the particle swarm \textit{misidentified} the correct allosteric structure (Fig.
-\ref{fig-control-search}C).
-Interestingly, while each particle swarm identified parameter sets that minimized the simulation error, the estimated parameter values were not necessarily similar to the true parameters. 
-The angle between the estimated and true parameters was not consistently small across the swarms (identical parameters would give an angle of zero). 
-This suggested that our particle swarm approach identified a \textit{sloppy} ensemble, i.e., parameter estimates that were individually incorrect but collectively exhibited the correct model behavior.
+Villaverde and co-workers recently published a set of benchmark problems to evaluate parameter estimation methods [REFHERE].
+From a computational cost perspective problems they categorized the problems as most expensive, intermediate and least expensive.
+We evaluated the performance of our algorithms on a problem from each category.
+Table zzz shows that
 
-We calculated control program output and scaled metabolic flux for the positively, uniformly and negatively biased particle swarms (Fig. \ref{fig-flux-pattern}). 
-Network A and network C models from the positively (Fig.
-\ref{fig-flux-pattern}A) and uniformly (Fig.
-\ref{fig-flux-pattern}B) biased particle swarms showed similar operational patterns, despite differences in kinetic parameters and control structures.
-While models from the negatively biased population had error values similar to the correct structures in the previous swarms, they have different flux and control profiles (Fig.
-\ref{fig-flux-pattern}C).
-In all cases, regardless of network configuration or parameter values, the rate of enzyme decay was small compared to the other fluxes, and all networks had qualitatively similar trends for $E_{3}$ and $E_{6}$ control.
-Moreover, consistent with the correct model structure, production of end product P$_1$ was the preferred branch for all model configurations. 
-However, there was variability in P$_{2}$ production flux across the population of models, especially for the uniform swarm when compared with the other cases. 
-High P$_{1}$ branch flux resulted in end product inhibition of $E_{1}$ in both network A and network C, however in network D and E, high P$_1$ flux induced $E_1$ activation. 
-These trends were manifested in different flux profiles, where the negatively biased population appeared more uniform across the population compared with the other swarms, and had higher $E_{1}$ specific activity.
-Interestingly, the behavior of network A and network C highlighted an artifact of our integration rule; both a positive or negative feedback connection from P$_2$ to $E_{1}$ were ignored because the P$_{1}$ inhibition of $E_{1}$ was dominate.
-Thus, while theoretically distinct, network A and network C appeared operationally to the PSO algorithm to be that same network.
-On the other hand, networks B, D and E showed distinct behavior that was not consistent with the true network. 
-These architectures exhibited either limited inhibition (network B) or activation (network D and E) of $E_{1}$ activity, resulting in significantly different metabolic flux profiles. 
-However, the PSO was able to find low error parameter solutions, despite the mismatch in the control structures (error values similar, but not better than the best network A and network C estimates). 
-Taken together, these results suggested that a uniform sampling approach could potentially yield an unbiassed estimate of both kinetic parameters and control structures. 
-However, the negatively biased particle swarm results illustrated a potential shortcoming of the approach, namely convergence to a local error minimum despite a significantly incorrect control structure.
-This suggested that estimated model structures will need to be further evaluated, for example by generating falsifiable experimental designs which could distinguish between low error solutions.
+\begin{table}[h!]
+\centering
+\begin{tabular}{|c c c c|} 
+ \hline
+ ModelID & B1 & B4&\\ [0.5ex] 
+ \hline
+ Upper bound & ${5.pnom}$ & ${5.pnom}$ & \\  [0.5ex]
+ Lower bound & ${0.2.pnom}$ & ${0.2.pnom}$ &\\ [0.5ex]
+ CPU time & $38.308$ ${hours}$ & $6.2$ ${minutes}$& \\ [0.5ex]
+ Function Evaluations & $4000$ & $4000$&\\ [0.5ex]
+ Initial Objective Value & $1.0589.10 ^{10}$ & $1.4275.10 ^{7}$ &\\ [0.5ex] 
+ Final Objective Value & $4.96.10^{5}$ & $38.9375$&\\ [0.5ex]
+ Nominal Objective Value & $1.0986·10^{6}$ & $39.0676$&\\ [1ex]
+ \hline
+\end{tabular}
+\caption{Table to test captions and labels}
+\label{table:1}
+\end{table}
 
 \clearpage
+\section*{Discussion}
+\clearpage
+\section*{Materials and Methods}
+\subsection*{Formulation and solution of coagulation model equations.}
+\subsection*{Optimization Strategy}
+Dynamically dimensioned multi-swarm particle swarm optimization (DDSMLSPSO) is a novel optimization approach that combines elements from multi-swarm based PSO methods  with DDS.
+The goal of this approach is to obtain optimal or near optimal parameters for high-dimensional complex biological systems within a pre-specified number of function evaluations.
 
-\section*{Discussion}
-In this study, we presented an effective kinetic modeling strategy to dynamically simulate cell-free biochemical networks. 
-Our proposed strategy integrated traditional kinetic modeling with an effective rules based approach to dynamically describe metabolic regulation and control. 
-We tested this approach by developing kinetic models of hypothetical cell-free metabolic networks. 
-In particular, we tested whether our effective modeling approach could describe classically expected behavior, and second whether we could simultaneously estimate kinetic parameters and regulatory connectivity, in the absence of specific mechanistic knowledge, from synthetic experimental data.
-Toward these questions, we explored five hypothetical cell-free networks. 
-In each network, a substrate S was converted to the end products P$_{1}$ and P$_{2}$ through a series of enzymatically catalyzed reactions, including a branch point at a hypothetical metabolite M$_{2}$.
-Each network also included the same cofactors and cofactor recycle architecture. 
-However, while all five networks shared the same enzymatic connectivity, each had different allosteric regulatory connectivity. 
-We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, could capture complex allosteric patterns such as ultrasensitivity, or non-competitive inhibition in the absence of specific mechanistic information.
-Moreover, when integrated into network models, these rules captured classical regulatory patterns such as product-induced feedback inhibition. 
-Lastly, we simultaneously estimated kinetic parameters and discriminated between competing regulatory structures, using synthetic data in combination with a modified particle swarm approach.
-If we considered all putative regulatory architectures to be equally likely, we were able to estimate a \textit{sloppy} ensemble of models with the correct architecture and kinetic parameters.
+Using Latin Hyper Squares we randomly initialized a swarm of $\mathcal{K}$-dimensional particles (represented as ${x}_{i}$), wherein each of these particles corresponds to an $\mathcal{K}$-dimensional parameter vector.
+After initialization, the particles were grouped into different sub-swarms randomly.
+Thereafter within each sub-swarm ${S}_{k}$,  particles were updated according to the following rule.
+\begin{eqnarray}
+	\mathbf{x}_{i,j} &=&\theta_{1,j-1}\mathbf{x}_{i,j-1} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{x}_{i,j-1}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{GL}_{k} - \mathbf{x}_{i,j-1}\right)
+\end{eqnarray}
 
-The proposed modeling strategy shares features with other popular techniques, but has also has several key differences. 
-At its core, our effective modeling approach is similar to regulatory constraints based methods, and to the cybernetic modeling paradigm developed by Ramkrishna and colleagues.
-Covert, Palsson and coworkers drastically improved the predictability of constraints based approaches by integrating boolean rules into the calculation of metabolic fluxes \citep{Covert:2001aa}.
-If the regulated intracellular flux problem is coupled with time-dependent extracellular balances, these models can predict complex behavior such as diauxie growth or the switch between 
-aerobic and anaerobic metabolism. Another important feature of this approach is that it scales with biological complexity. 
-For example, Covert \emph{et al.,} showed that a genome scale model of \emph{E.~coli} augmented with a boolean rule layer, 
-correctly predicted approximately 80\% of the outcomes of a high-throughput growth phenotyping experiment in \emph{E. coli}.
-Further, they showed that they could learn new biology by iteratively refining the model and its associated rules \citep{Covert:2004aa}. 
-However, while regulated flux balance analysis is a powerful technique, it does not easily allow the calculation of time-resolved metabolite abundance. 
-Additionally, the boolean rules which populate the regulatory layer are limited to ON/OFF decisions; for qualitative predictions of gene expression this is a reasonable limitation. 
-However, boolean rules will likely be less effective at capturing dynamic allosteric regulation in a cell free metabolic system. 
-On the other hand, the strength of cybernetic models is the integration of optimal metabolic control heuristics with traditional kinetic pathway modeling. 
-Cybernetic models are highly predictive; they have successfully predicted mutant behavior from limited wild-type data \citep{1999_varner_ramkrishna_MetaEng,Varner:2000aa,Song:2012aa}, 
-steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, strain specific metabolic function \citep{Song:2011aa} and have been used in bioprocess control applications \citep{Gadkar:2003aa}.
-However, cybernetic control heuristics are not mechanistic, instead they are the output of an optimal decision with respect to a set of hypothetical physiological objectives. 
-Thus, they are abstractions which are difficult to translate into a specific biological mechanism. 
-Our approach addresses the shortcomings of both regulatory constraints based models and cybernetic models. 
-First, similar to cybernetic models, the core of our approach is a kinetic model. Thus, we are able to directly calculate the time evolution of metabolism, for example the dynamic abundance of 
-network metabolites. Second, similar to regulatory flux balance analysis, our control laws describe specific mechanistic motifs such a activation or inhibition of enzyme activity. 
-However, our rules are continuous, thus they potentially allow a finer grained description metabolic regulation and control mechanisms.
-Lastly, we can naturally incorporate non modeled factors and categorical factors or factor combinations into our control law formulations.
+where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes the best local solution found by particle $i$ till function evaluation $j$-$1$, and
+$\mathcal{GL}_{k}$ denotes the best local solution found over the population of all particles within the swarm ${S}_{k}$.
+The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
+parameters ($\mathcal{K}\times{1}$). In our algorithm the parameter $\theta_{1,j-1}$ depends on the function evaluations and is controlled according to the following equation
 
-There are several critical questions that should be explored following this proof of concept study.
-It is unclear how parameter identification will scale to genome scale networks, and second it is unclear how we will identify allosteric connectivity at a genome scale. 
-The enzymatic connectivity for genome scale cell free networks can easily be established by stripping away the growth and cell wall machinery from whole cell genome reconstructions.
-Then metabolic fluxes can transformed into kinetic expressions using heuristics such multiple saturation kinetics, which are then modified by our rule based control variables.  
-This leaves a large number of unknown kinetic constants that must be estimated from time-resolved metabolite measurements.
-We showed that particle swarm optimization quickly identified an ensemble of model parameters, at least for proof of concept metabolic networks using synthetic data.
-This suggested that we can expect reasonable model predictions, despite only partial parameter knowledge, as network size grows if we have properly designed experiments. 
-Brown and Sethna showed in a model of signal transduction that good predictions were possible despite only order of magnitude estimates of parameter values \citep{Brown:2003aa}.
-Sethna and coworkers later showed that model performance is often controlled by only a few parameter combinations, 
-a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
-We have also demonstrated \textit{sloppy} behavior in a wide variety of signal transduction processes \citep{Luan:2007aa,Song:2009aa,Tasseff:2010aa,Tasseff:2011aa,Nayak:2011aa,Lequieu:2011aa}.
-Thus, given our previous experience with models with hundreds of unknown parameters, we expect parameter estimation to be a manageable challenge. 
-On the other hand, a critical challenge will be the estimation of allosteric connectivity at a genome scale. 
-The regulation of glycolytic enzymes, such as phosphofructokinase I, has been studied for many years \citep{BERG_BIOCHEMISTRY,Peskov:2008aa}. 
-The allosteric regulation of metabolic enzymes can also be established from organism specific databases, such as EcoCyc \citep{Keseler:2013aa} 
-or more general allosteric databases such as the AlloSteric Database \citep{Huang:2014aa}. 
-However, for those enzymes that have not been well studied, we will need to infer allosteric interactions from experimental data.
-In general, the reverse engineering of regulatory network structure from data is a very difficult problem.
-There are many different approaches from the reverse engineering of gene regulatory networks that perhaps could be adopted to this problem, however this remains an open question.
+\begin{eqnarray}
+	\mathbf \theta_{1,j}&=&((\mathbf{N}-{j})*(\mathbf{w}_{max}-\mathbf{w}_{min}))/(\mathbf{N}-{1}) + \mathbf{w}_{min}
+\end{eqnarray}
+
+where $\mathbf{N}$ represents the total number of function evaluations, $\mathbf{w}_{max}$ and $\mathbf{w}_{min}$ are the maximum and minimum inertia weights respectively.
+While updating the particle, we made sure all dimensions of the solution represented by the particle was within bounds using a reflection boundary condition.
+After every $\mathbf{g}$  function evaluations, the particles within all sub-swarms were mixed and then randomly redistributed to a new sub-swarm.
+The particles were then again updated according Eq.1.
+This process continued till $\mathcal{FR}*\mathbf{N}$ number of functions evaluations, where $\mathcal{FR}$ represents the fraction of evaluations with the multi-swarms.
+At the end of these function evaluations, we froze all the solutions represented by various particles and chose the particle with best solution among $\mathcal{GL}_{1} \cdots \mathcal{GL}_{NS}$ as the initial candidate vector $\mathcal{G}$ for the remaining $({1}-\mathcal{FR})*\mathbf{N}$ number of function evaluations.
+
+This particle was then updated according to the following rule
+
+\begin{eqnarray}
+	\mathcal{G}_{new}(\mathbf{J})& = &\mathcal{G}(\mathbf{J})+\mathbf{r}_{normal}(\mathbf{J})\sigma(\mathbf{J})
+\end{eqnarray}
+
+where $\mathbf{J}$ represents the a vector containing the specific dimensions being perturbed, ${r}_{normal}$ denotes a normal random vector of the same dimensions as $\mathcal{G}$.
+$\sigma$ is the amplitude of perturbation given by following equation
+
+\begin{eqnarray}
+	\sigma& = &\mathbf{R}(\mathcal{MAX} -\mathcal{MIN})
+\end{eqnarray}
+
+where $\mathbf{R}$ is the scalar perturbation size parameter, $\mathcal{MAX}$ and $\mathcal{MIN}$ are ($\mathcal{K}\times{1}$) vectors that represent the maximum and minimum bounds on each dimension.
+The probability $\mathcal{P}$ that a specific dimension is perturbed is a  monotonically decreasing function that decreases with the number of function evaluations.
+$\mathcal{P}$ can be any monotonically decreasing function, in our approach we used the following function
+\begin{eqnarray}
+	\mathcal{P}_{j}&=&{1}-\log(j/ (\beta({1}-\mathcal{FR})*\mathbf{N}))
+\end{eqnarray}
+
+where $\beta$ is the perturbation frequency probability modulator.
+Thus the number of dimensions of the candidate vector that are updated or perturbed decreases with the as the number of function evaluations increase.
+These updates are greedy in nature that is $\mathcal{G}_{new}$ becomes the new solution vector only if it is better than the old one $\mathcal{G}$.
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.00\textwidth,height=0.7\textheight]{./figs/Figure_2_CoagulationNetwork.pdf}
+\caption{Schematic of the extrinsic and intrinsic coagulation cascade.Upstream coagulation factors are activated by materials exposed following vessel injury chief among these tissue factor (TF) [Ref] Tissue factor and activated factor VIIa (FVIIa) form a complex that activates factor X (fX) and IX (fIX).
+FXa activates downstream factors including factor VIII (fVIII) and fIX.
+Factor V (fV) is primarily activated by thrombin (FIIa).
+In addition, we included a secondary fV activation route involving FXa.
+FXa and FVa form a complex (prothrombinase) on activated platelets that converts
+prothrombin (fII) to FIIa. FIXa and FVIIIa can also form a complex (tenase) on activated platelets which catalyzes FXa formation. Localized
+platelets are activated by external signals such as adenosine diphosphate (ADP) and thromboxane A2 (TXA2) or thrombin through
+protease-activated receptors (PARs)[Ref] Thrombin also activates upstream coagulation factors, forming a strong positive feedback ensuring
+rapid activation. Tissue factor pathway inhibitor (TFPI) downregulates FXa formation and activity by sequestering free FXa and TF-FVIIa in a
+FXa-dependent manner. Antithrombin III (ATIII) neutralizes all proteases, making it perhaps the most powerful control element in the cascade.
+Thrombin itself plays an inadvertent role in its own inhibition by binding the surface protein thrombomodulin (TM), expressed on normal
+vasculature[Ref] The IIa-TM complex catalyzes the conversion of protein C (PC) to activated protein C (APC), which attenuates the coagulation response by the proteolytic cleavage of fV/FVa and fVIII/FVIIIa [Ref] }\label{fig-flux-pattern}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.0\textwidth,height=0.5\textheight]{./figs/Figure_1_Algorithm.pdf}
+\caption{Multi Swarm Particle Swarm Optimization with Dynamically Dimensioned Search.
+\textbf{A}: Each particle represents an N dimensional parameter vector.
+Particles are randomly initialized and grouped into various sub-swarms.
+The magnitude and direction of the movement a particle is influenced by the position of the best particle in its swarm and also by its own experience.
+After a certain number of function evaluations the particles are mixed and randomly assigned to different swarms.
+At the end of the evaluations assigned to swarm search, the global best particle amongst all sub-swarms is chosen as the candidate parameter vector for Dynamically Dimensioned Search \textbf{B}: The candidate vector does a greedy search in a dynamic neighborhood.
+This done by dynamically adjusting the number of parameter dimensions that are perturbed in each evaluation step.
+The number of dimensions that are perturbed generally decreases as the number of iterations increase.
+This preserves the optimality of the solution as the number of evaluations increases.
+}\label{fig-algorithm}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.0\textwidth,height=0.5\textheight]{./figs/Figure_4_Sim_Train_E1_E5.pdf}
+\caption{Model fits on experimental data using DDSMLSPSO.
+The model parameters were estimated using DDSMLSPSO.
+Solid black lines indicate the simulated mean thrombin concentration using parameter vectors from $N$= 25 trials.
+The grey shaded region represents the 99\% confidence estimate of the mean simulated thrombin concentration.
+The experimental data is reproduced from the synthetic plasma assays of Mann and co-workers [REFHERE].
+Thrombin generation is initiated by adding Factor VIIa-TF (5nM - Red and 5pM - Green) to synthetic plasma containing 200 $\mu$mol/L of phospholipid vesicles (PCPS) and a mixture of coagulation factors (II,V,VII,VIII,IX,X and XI) at their mean plasma concentrations.
+}\label{fig-train}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.0\textwidth,height=0.5\textheight]{./figs/Figure_5_Sim_Validate_E2_E4_E6.pdf}
+\caption{Model predictions on unseen experimental data using parameters obtained from DDSMLSPSO. 
+The parameter estimates that were obtained using DDSMLSPSO were tested against data that was not used in the model training.
+Solid black lines indicate the simulated mean thrombin concentration using parameter vectors from $N$= 25 trials.
+The grey shaded region represents the 99\% confidence estimate of the mean simulated thrombin concentration.
+The experimental data is reproduced from the synthetic plasma assays of Mann and co-workers [REFHERE].
+Thrombin generation is initiated by adding Factor VIIa-TF (500pM - Blue, 50pM - Pink and 10pM - Orange respectively) to synthetic plasma containing 200 $\mu$mol/L of phospholipid vesicles (PCPS) and a mixture of coagulation factors (II,V,VII,VIII,IX,X and XI) at their mean plasma concentrations.
+}\label{fig-validation}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.02\textwidth,height=0.5\textheight]{./figs/Figure6_Ackley_Rast.pdf}
+\caption{Error convergence rates on a 300 dimensional Rastrigin function and Ackley function.
+Ackley and Rastrigin are commonly used test functions for global optimization.
+DDSMLSPSO finds the minimum $ {0} $ much faster than other approaches in both cases within 4000 function evaluations \textbf {A} On a 300-D Rastrigin, DDSMLPSO, DDS and SA perform the best followed by PSO and DE  \textbf {B} On a 300-D Ackley, DDSMLSPSO clearly outperforms the rest of the algorithms.
+}\label{fig-Rast}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.02\textwidth,height=0.5\textheight]{./figs/Figure_7_Benchmarks_Parameters}
+\caption{Difference between optimal and nominal parameter vector values on benchmark problems [REFHERE].
+\textbf {(A)} Problem B1: Genome wide kinetic model of E.coli \textit{S.cerevisiae} with 1759 unknown parameters.
+\textbf {(B)} Problem B4: Metabolic model of Chinese  Hamster Ovary Cells (CHO) cells with 117 parameters.
+}\label{fig-benchmark}
+\end{figure}
+
+\section*{Acknowledgements}
+This study was supported by the National Science Foundation GK12 award (DGE-1045513) 
+and by the National Science Foundation CAREER award (FILLMEIN).
 
 \clearpage
-
-\section*{Materials and Methods}
-
-\subsection*{Formulation and solution of the model equations.}
-We used ordinary differential equations (ODEs) to model the time evolution of metabolite ($x_{i}$) and scaled enzyme abundance ($\epsilon_{i}$) in hypothetical cell-free metabolic networks:
-\begin{eqnarray}
-	\frac{dx_{i}}{dt} & = & \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)\qquad{i=1,2,\hdots,\mathcal{M}}\\
-	\frac{d\epsilon_{i}}{dt} & = & -\lambda_{i}\epsilon_{i}\qquad{i = 1,2,\hdots,\mathcal{E}}
-\end{eqnarray}where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of metabolites and $\mathcal{E}$ denotes the number of enzymes in the model. 
-The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of reaction $j$. 
-Typically, reaction $j$ is a non-linear function of metabolite and enzyme abundance, as well as unknown kinetic parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$). 
-The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$. 
-If $\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$. 
-Conversely, if $\sigma_{ij}>0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates metabolite $i$ is not connected with reaction $j$. 
-Lastly, $\lambda_{i}$ denotes the scaled enzyme degradation constant. 
-The system material balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$ and $\mathbf{\epsilon}\left(t_{o}\right) = \mathbf{1}$ (initially we have 100\% cell-free enzyme abundance).
-
-Each reaction rate was written as the product of two terms, a kinetic term ($\bar{r}_{j}$) and a regulatory term ($v_{j}$):
-\begin{equation}\label{eqn:rate-factor}
-	r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right) = \bar{r}_{j}v_{j}
-\end{equation}
-We used multiple saturation kinetics to model the reaction term $\bar{r}_{j}$:
-\begin{equation}\label{eqn:rate-bar}
-	\bar{r}_{j} = k_{j}^{max}\epsilon_{i}\left(\prod_{s\in{m_{j}^{-}}}\frac{x_{s}}{K_{js} + x_{s}}\right)
-\end{equation}where $k_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the scaled enzyme activity which catalyzes reaction $j$, and
-$K_{js}$ denotes the saturation constant for species $s$ in reaction $j$. 
-The product in Eqn. \eqref{eqn:rate-bar} was carried out over the set of \textit{reactants} for reaction $j$ (denoted as $m_{j}^{-}$).
-
-The allosteric regulation term $v_{j}$ depended upon the combination of factors which influenced the activity of enzyme $i$.
-For each enzyme, we used a rule-based approach to select from competing control factors (Fig. \ref{fig-control-schematic}). 
-If an enzyme was activated by $m$ metabolites, we modeled this activation as:
-\begin{equation}
-	v_{j} = \max\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{mj}\left(\mathcal{Z}\right)\right)
-\end{equation}where $0\leq f_{ij}\left(\mathcal{Z}\right)\leq 1$ was a regulatory transfer function that calculated the influence of metabolite $i$ on the activity of enzyme $j$. 
-Conversely, if enzyme activity was inhibited by a $m$ metabolites, we modeling this inhibition as:
-\begin{equation}
-	v_{j} = 1 - \max\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{mj}\left(\mathcal{Z}\right)\right)
-\end{equation}Lastly, if an enzyme had both $m$ activating and $n$ inhibitory factors, we modeled the regulatory term as:
-\begin{equation}
-	v_{j} = \min\left(u_{j},d_{j}\right)
-\end{equation}where:
-\begin{eqnarray}
-	u_{j} &=& \max_{j^{+}}\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{mj}\left(\mathcal{Z}\right)\right) \\
-	d_{j} &=& 1 - \max_{j^{-}}\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{nj}\left(\mathcal{Z}\right)\right)
-\end{eqnarray}
-The quantities $j^{+}$ and $j^{-}$ denoted the sets of activating and inhibitory factors for enzyme $j$. 
-If an enzyme had no allosteric factors, we set $v_{j} = 1$.
-There are many possible functional forms for $0\leq f_{ij}\left(\mathcal{Z}\right)\leq 1$. 
-However, in this study, each individual transfer function took the form:
-\begin{equation}\label{eqn:control-factor}
-	f_{i}\left(\mathbf{x}\right) = \frac{\kappa_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}{1 + \kappa_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}
-\end{equation}where $\mathcal{Z}_{j}$ denotes the abundance of the $j$ factor (e.g., metabolite abundance), and $\kappa_{ij}$ and $\eta$ are control parameters. 
-The $\kappa_{ij}$ parameter was species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).
-The model equations were encoded using the Octave programming language and solved using the LSODE routine in Octave \citep{Octave:2014}.
-
-\subsection*{Estimation of model parameters and structures from synthetic experimental data.}
-Model parameters were estimated by minimizing the difference between simulations and synthetic experimental data (squared residual):
-\begin{equation}\label{eqn:objective-function}
-	\min_{\mathbf{k}} \sum_{\tau=1}^{\mathcal{T}}\sum_{j=1}^{\mathcal{S}}\left(\frac{\hat{x}_{j}\left(\tau\right) - x_{j}\left(\tau,\mathbf{k}\right)}{\omega_{j}\left(\tau\right)}\right)^{2}
-\end{equation}where $\hat{x}_{j}\left(\tau\right)$ denotes the measured value of species $j$ at time $\tau$, $x_{j}\left(\tau,\mathbf{k}\right)$ denotes the simulated 
-value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$. The outer summation is respect to
-time, while the inner summation is with respect to state. We approximated a realistic model identification scenario, assuming noisy experimental data, 
-limited sampling resolution (approximately 20 minutes per sample) and a limited number of measurable metabolites.
-
-We minimized the model residual using particle swarm optimization (PSO) \citep{PSO}.
-PSO uses a \textit{swarming} metaheuristic to explore parameter spaces. 
-A strength of PSO is its ability to find the global minimum, even in the presence of potentially many local minima, by communicating the local
-error landscape experienced by each particle collectively to the swarm. Thus, PSO acts both as a local and a global search algorithm. 
-For each iteration, particles in the swarm compute their local error by evaluating the model equations using their specific parameter vector realization.
-From each of these local points, a globally best error is identified. Both the local and global error 
-are then used to update the parameter estimates of each particle using the rules:
-\begin{eqnarray}
-	\mathbf{\Delta}_{i} &=&\theta_{1}\mathbf{\Delta}_{i} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{k}_{i}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G} - \mathbf{k}_{i}\right) \\
-	\mathbf{k}_{i} &=& \mathbf{k}_{i} + \mathbf{\Delta}_{i}
-\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes local best solution found by particle $i$, and
-$\mathcal{G}$ denotes the best solution found over the entire population of particles.
-The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
-parameters ($\mathcal{K}\times{1}$). In thus study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$. The quality of parameter
-estimates was measured using two criteria, goodness of fit (model residual) and angle between the estimated parameter vector $\mathbf{k}_{j}$ and the true parameter set $\mathbf{k}^{*}$:
-\begin{equation}
-	\alpha_{j} = cos^{-1}\left(\frac{\mathbf{k}_{j}\cdot{\mathbf{k^{*}}}}{\norm{\mathbf{k}_{j}}\norm{\mathbf{k^{*}}}}\right)
-\end{equation}If the candidate parameter set $\mathbf{k}_{j}$ were perfect, the residual between the model and synthetic data 
-and the angle between $\mathbf{k}_{j}$ and the true parameter set $\mathbf{k}^{*}$ would be equal to zero.
-
-We modified our PSO implementation to simultaneously search over kinetic parameters and putative model control structures.
-In the combined case, each particle potentially carried a different model realization in addition to a different kinetic parameter vector. 
-We kept the update rules the same (along with the update parameters). 
-Thus, each particle competed on the basis of goodness of fit, which allowed different model structures to contribute to the overall behavior of the swarm. 
-We considered five possible model structures (A through E), where network A was the correct formulation (used to generate the synthetic data). 
-We considered a population N = 100 particles, where each particle in the swarm was assigned a model structure, and a random parameter vector.
-The PSO algorithm, model equations, and the objective function were encoded and solved in the Octave programming language \citep{Octave:2014}.
-
-\section*{Acknowledgements}
-This study was supported by an award from the National Science Foundation (MCB \#1411715) 
-and the Army Research Office (ARO \#59155-LS).
-
-\clearpage
-%\bibliographystyle{plain}
-%\bibliographystyle{IEEEbib}
 
@@
 %\begin{comment}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-1-Networks.pdf}
-\caption{Proof of concept cell-free metabolic networks considered in this study. Substrate $S$ is converted to products P$_{1}$ and P$_{2}$ through a series of chemical conversions
-catalyzed by enzyme(s) $E_{j}$. The activity of the pathway enzymes is subject to both positive and negative allosteric regulation. }\label{fig-networks}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-2-ControlSchematic.pdf}
-\caption{Schematic of rule-based allosteric enzyme activity control laws. 
-Traditional enzyme kinetic expressions, e.g., Michaelis–Menten or multiple saturation kinetics, are multiplied by an enzyme activity control variable $0 \leq v_{j} \leq 1 $. 
-Control variables are functions of many possible regulatory factors encoded by arbitrary functions of the form $0\leq f_{j}\left(\mathcal{Z}\right)\leq 1$.
-At each simulation time step, the $v_{j}$ variables are calculated by evaluating integration rules such as the max or min of the set of factors $f_{1},\hdots$ 
-influencing the activity of enzyme $E_{j}$. }\label{fig-control-schematic}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-3-EnzymeKinetics.pdf}
-\caption{Kinetics of simple transformations in the presence of activation and inhibition. 
-\textbf{A}:The conversion of substrate $S$ to product $P$ by enzyme $E$ was activated by $S$. 
-For a fixed control gain parameter $k_{control}$, the reaction rate approached a step for increasing control order $N$. 
-\textbf{B}:The conversion of substrate $S$ to product $P$ by enzyme $E$ with inhibitor $I$. 
-For a fixed control gain parameter $k_{control}$, the reaction rate approximated non-competitive inhibition for increasing control order $N$. 
-}\label{fig-kinetics-simulations}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=0.50\textwidth]{./figs/Figure-4-OnOffSimulations.pdf}
-\caption{ON/OFF control simulations for network A and network B for an ensemble of kinetic parameter sets versus time (N = 100). 
-For each case, N = 100 simulations were conducted using kinetic and initial conditions generated randomly from a hypothetical true parameter set. 
-The gray area represents $\pm$ one standard deviation surrounding the mean. 
-Control parameters were fixed during the ensemble calculations.
-\textbf{A}: End product P$_{1}$ abundance versus time for Network A. 
-The abundance of P$_{1}$ decreased with end product inhibition of $E_{1}$ activity (Control-ON) versus the no inhibition case (Control-OFF). 
-\textbf{B}: End product P$_{2}$ abundance versus time for Network B. Inhibition of branch point $E_{6}$ by end product P$_{1}$ decreased P$_{2}$ abundance (Control-ON) versus the
-no inhibition case (Control-OFF).
-\textbf{C}: End product P$_{1}$ abundance versus time for Network A. 
-Inhibition of branch point $E_{6}$ by end product P$_{1}$ decreased P$_{1}$ abundance (Control-ON) versus the no inhibition case (Control-OFF).
-}\label{fig-onoff-simulations}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-5-ParameterFit.pdf}
-\caption{Parameter estimation from synthetic data for the same and mismatched allosteric control logic using particle swarm optimization (PSO). 
-Synthetic experimental data was generated from a hypothetical parameter set using Network A, 
-where substrate $S$, end product P$_{1}$ and intermediate M$_5$ were sampled approximately every 20 minutes. 
-For cases $\textbf{A,B}$ 20 particles were initialized with randomized parameters and allowed to search for 300 iterations. 
-\textbf{A,B}: PSO estimated an ensemble of parameters sets (N = 20) consistent with the synthetic experimental data assuming the correct
-enzymatic and control connectivity starting from randomized initial parameters.
-\textbf{C,D}: In the presence of control mismatch (Network B control policy simulated with Network A kinetic parameters) 
-the ensemble of models did not describe the synthetic data. 
-}\label{fig-parameter-fit}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure-6-AlternativeNetworks.pdf}
-\caption{Schematic of the alternative allosteric control programs used in the structural particle swarm computation. 
-Each network had the same enzymatic connectivity, initial conditions and kinetic parameters, 
-but alternative feedback control structures for the first enzyme in the pathway.}\label{fig-alternative-networks}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=0.5\textwidth]{./figs/Figure-7-ControlSearch.pdf}
-\caption{Combined control and kinetic parameter search using modified particle swarm optimization (PSO). 
-A population of N = 100 particles was initialized with randomized kinetic parameters and one of five possible control configurations (Network A - E).
-Simulation error was minimized for a synthetic data set 
-($S$, end product P$_{1}$ and intermediate M$_5$ sampled approximately every 20 min) generated using Network A.
-\textbf{A}: Simulation error versus parameter set angle for N = 100 particles biased toward the correct regulatory program (A,B,C,D,E) = (40\%, 10\%, 20\%, 20\% and 10\%).
-\textbf{B}: Simulation error versus parameter set angle for N = 100 uniformly distributed particles (A,B,C,D,E) = (20\%, 20\%, 20\%, 20\% and 20\%).
-\textbf{C}: Simulation error versus parameter set angle for N = 100 negatively biased particles (A,B,C,D,E) = (10\%, 40\%, 10\%, 20\% and 20\%).
-Network A (the correct structure) was preferentially identified for positively and uniform biased particle distributions, but misidentified in the presence of
-a large incorrect bias. 
-}\label{fig-control-search}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth,height=0.6\textheight]{./figs/Figure-8-Flux.pdf}
-\caption{Metabolic flux and control variables as a function of network type and particle index at t = 100 min.
-The control variables governing $E_{1},E_{3}$ and $E_{6}$ activity and the scaled metabolic flux and were calculated for the positively, 
-uniformly and negatively biased particle swarms (N = 100). The particles from each swarm were sorted based upon simulation error (low to high error). 
-\textbf{A}: Model performance for the positively biased particle swarm as a function of particle index.
-\textbf{B}: Model performance for the uniformly biased particle swarm as a function of particle index.
-\textbf{C}: Model performance for the negatively biased particle swarm as a function of particle index. 
-Models with significant control mismatch showed distinct control and flux patterns versus those models with the correct or closely related control policies.
-In particular, models with the correct control policy showed stronger inhibition of $E_{1}$ activity, leading to decreased flux from S$\rightarrow$P$_{1}$.
-Conversely, models with significant mismatch had increased $E_{1}$ activity, leading to an altered flux distribution. 
-This is especially apparent in the negatively biased particle swarm. 
-}\label{fig-flux-pattern}
-\end{figure}
 

### 2015-04-23 Paper_v1.tex
Still updating

@@
 Model calibration or parameter estimation follows the model construction step.
-In this step the model is trained against experimental data to determine the model parameters that give a good fit.
-After model design and calibration, validation is done wherein the model output is validated against unseen experimental data.
-If validation leads to unsatisfactory results model design and calibration are performed once again.
-This process is done iteratively till successful results are obtained from validation step.
-\\
+In this step the model is trained against experimental data to determine the model parameters that give a good fit against experimental data.
+Model simulation using these parameters is then done and the results are validated using hitherto unseen experimental data.
+If validation leads to unsatisfactory results model construction and calibration are repeated iteratively till we get satisfactory results.
 
-The objective function in Equation 5 is minimized using appropriate optimization technique to estimate the model parameters.
+With the capacity to construct or formulate very large biological models, more often than not, parameter estimation has become the bottleneck in large scale mathematical modeling of biological systems.
+The problem of parameter estimation if generally framed as a non-linear optimization problem.
+
+\textit  {\Large Equation needs to be written and explained terms}
+
+\begin{eqnarray}
+	\mathcal{G}_{new}(\mathbf{J})& = &\mathcal{G}(\mathbf{J})+\mathbf{r}_{normal}(\mathbf{J})\sigma(\mathbf{J})
+\end{eqnarray}
+
+The objective function in Equation 1 is minimized using appropriate optimization technique to estimate the model parameters.
 These problems are generally multi-modal i.e.

### 2015-04-27 Paper_v1.tex
Updated table

@@
 
-The objective function in Equation 1 is minimized using appropriate optimization technique to estimate the model parameters.
-These problems are generally multi-modal i.e.
-contain multiple local optima traditional.
-Hence we cannot use local optimization techniques cannot be used to solve these problems.
-In many of these problems derivative information is absent or the objective function may be discontinuous.
+The objective function from equation ---  is minimized using appropriate optimization technique to estimate the model parameters.
+Due to the complex dynamics of biological systems these problems are multi-modal i.e.
+most of them have multiple local minima or maxima.
+The non-linearity of the problem coupled with multi-modality is a substantial challenge and generally renders the local optimization techniques incapable of obtaining an optimal solution.
+\textit {In many of these problems derivative information is absent or the objective function may be discontinuous.
 This hinders the use of deterministic optimization techniques that require the problem to be convex and other derivative/gradient based optimization methods.
-Most of these methods are computationally expensive and for optimization of large-scale ODE systems, this becomes prohibitively expensive.
-Heuristic optimization techniques like genetic algorithm and differential evolution offer an attractive option in finding globally optimum solution or close to global optimum solution.
-However most of the population-based heuristics take a large number of function evaluations and have computationally complex operations, which reduces their effectiveness in finding feasible solutions with limited computational resources and time.
+Most of these methods are computationally expensive and for optimization of large-scale systems, this becomes prohibitively expensive.
+Heuristic approaches have been promising in this regard, however most of the population-based heuristics take a large number of function evaluations and have computationally complex operations, which reduces their effectiveness in finding feasible solutions with limited computational resources and time.
+Hybrid methods that couple heuristics with derivative based methods}
 
-This study tries to overcome the drawbacks mentioned above and presents a novel heuristics based search strategy/algorithm for parameter estimation in large-scale ODE models of biochemical networks with many unknown parameters.
+The problem get further compounded when we have parameter vectors of very high dimensions.
+In this study we developed a novel meta heuristic approach that combines multi-swarm particle swarm optimization with dynamically dimensioned search (DDS)to obtain optimal or near optimal solutions to high-dimensional, non-linear biological models within a relatively few function evaluations.
+
+A bit of history on DDS
+Why coagulation model? feed back loops, parameter identifiability, tight regulation and also high dimensional
 
@@
 
-\subsection *{Modeling methodologies} 
+\subsection *{Modeling methodologies Literature} 
  Shuler and co-workers constructed large-scale dynamic metabolic models of E.coli that were based ordinary differential equations (ODEs).
@@
 
-\section*{Results}\vspace{-12pt}
+\section*{Results}
+\subsection{Coagulation model formulation}
 
@@
 From a computational cost perspective problems they categorized the problems as most expensive, intermediate and least expensive.
-We evaluated the performance of our algorithms on a problem from each category.
-Table zzz shows that
+We evaluated the performance of our algorithms on a problem from the most expensive and least expensive categories.
+Table 1 shows that
 
@@
 \centering
-\begin{tabular}{|c c c c|} 
+\begin{tabular}{|c c c c c c c|} 
  \hline
- ModelID & B1 & B4&\\ [0.5ex] 
+ ModelID & Coagulation & B1 & B4 & Ackley300D & Rastrigin300D&\\ [0.5ex] 
  \hline
- Upper bound & ${5.pnom}$ & ${5.pnom}$ & \\  [0.5ex]
- Lower bound & ${0.2.pnom}$ & ${0.2.pnom}$ &\\ [0.5ex]
- CPU time & $38.308$ ${hours}$ & $6.2$ ${minutes}$& \\ [0.5ex]
- Function Evaluations & $4000$ & $4000$&\\ [0.5ex]
- Initial Objective Value & $1.0589.10 ^{10}$ & $1.4275.10 ^{7}$ &\\ [0.5ex] 
- Final Objective Value & $4.96.10^{5}$ & $38.9375$&\\ [0.5ex]
- Nominal Objective Value & $1.0986·10^{6}$ & $39.0676$&\\ [1ex]
+ Upper bound&$1000.pnom$& ${5.pnom}$ & ${5.pnom}$ &${30}$ &$5.12$& \\  [0.5ex]
+ Lower bound&$0.001.pnom$& ${0.2.pnom}$ & ${0.2.pnom}$&${-15}$&$-5.12$&\\ [0.5ex]
+ CPU time &$5.6$ ${hours}$ &$38.308$ ${hours}$ & $6.2$ ${minutes}$& ${2.8}$${seconds}$&${2.6}$${seconds}$& \\ [0.5ex]
+ Function Evaluations &$4000$&$4000$&$4000$&$4000$&$4000$&\\ [0.5ex]
+ Initial Objective Value & $1.0589.10 ^{10}$ & $1.4275.10 ^{7}$ &$1.8536.10 ^{7}$&$21.12$&$99.985$&\\ [0.5ex] 
+ Final Objective Value & $2.9223.10^{3}$ & $4.96.10^{5}$ & $38.9375$&$0$&$0$&\\ [0.5ex]
+ Nominal Objective Value & $4.7785·10^{6}$& $1.0986·10^{6}$ & $39.0676$&$0$&$0$&\\ [1ex]
  \hline
 \end{tabular}
-\caption{Table to test captions and labels}
+\caption{Table with optimization settings and results for coagulation and different benchmarks using DDSMLSPSO.}
 \label{table:1}

### 2015-05-06 Paper_v1.tex
New Updates

@@
 Mathematical modeling of biological systems is one such area where parameter estimation is a difficult non-linear optimization problem.
-This problem is further compounded when dealing with parameter vectors of high dimensions.
-Such problems generally involve expensive function evaluations and it is hard to obtain optimal or near optimal solutions within finite computational time.
+This difficulty is further compounded when dealing with parameter vectors of high dimensions.
+Such problems generally involve expensive function evaluations and it is hard to obtain optimal solutions within finite computational time.
 In this study we present a novel meta-heuristic approach that combines elements of multi-swarm optimization along with dynamically dimensioned search (DDS) to obtain optimal or near optimal solutions of high dimensional biochemical networks within a relatively few function evaluations.
@@
 
-Using Latin Hyper Squares we randomly initialized a swarm of $\mathcal{K}$-dimensional particles (represented as ${x}_{i}$), wherein each of these particles corresponds to an $\mathcal{K}$-dimensional parameter vector.
+ We randomly initialized a swarm of $\mathcal{K}$-dimensional particles (represented as ${x}_{i}$), wherein each of these particles corresponds to an $\mathcal{K}$-dimensional parameter vector.
 After initialization, the particles were grouped into different sub-swarms randomly.

### 2015-06-10 Paper_v1.tex
Committing first version of parameter estimation paper

@@
 \usepackage{lineno}
+\usepackage{chicago}
+\usepackage{textcomp}
 
+%\usepackage[space]{cite}
 \urlstyle{rm}
@@
 
-%\bibliographystyle{plain}
+%\bibliographystyle{chicago}
 \bibliographystyle{plos2009}
@@
 \begin{titlepage}
-{\par\centering\textbf{\Large A novel metaheuristic approach for parameter estimation in high dimensional biochemical networks}}
+{\par\centering\textbf{\Large A novel metaheuristic approach to obtain near optimal solutions in high dimensional biochemical networks with a few function evaluations}}
 \vspace{0.05in}
@@
 The problem of parameter estimation presents a significant challenge in the modeling of large complex systems.
-Mathematical modeling of biological systems is one such area where parameter estimation is a difficult non-linear optimization problem.
+Mathematical modeling of biological systems with multiple feed back loops is one such area where parameter estimation is a difficult non-linear optimization problem.
 This difficulty is further compounded when dealing with parameter vectors of high dimensions.
 Such problems generally involve expensive function evaluations and it is hard to obtain optimal solutions within finite computational time.
-In this study we present a novel meta-heuristic approach that combines elements of multi-swarm optimization along with dynamically dimensioned search (DDS) to obtain optimal or near optimal solutions of high dimensional biochemical networks within a relatively few function evaluations.
-We use a multi-swarm optimization technique to generate candidate vectors which are then greedily updated using DDS by dynamically varying the perturbed parameter dimensions.
-We tested (25 trials with 4000 function evaluations in each trial) this algorithm on a biochemical network of coagulation (148 parameters and 92 species) and compared it's performance against other metaheuristics like Differential Evolution (DE), Particle Swarm Optimization (PSO), Simulated Annealing (SA) and also against DDS alone.
+In this study we present a novel meta-heuristic approach that combines a variant of particle swarm optimization (PSO) along with dynamically dimensioned search (DDS) to obtain near optimal solutions of high dimensional biochemical networks within a relatively few function evaluations.
+We use a particle swarm optimization technique that uses multi-swarms to generate candidate vectors which are then greedily updated using DDS by dynamically varying the perturbed parameter dimensions.
+We tested this algorithm (25 trials with 4000 function evaluations in each trial) on a biochemical network of coagulation (148 parameters and 92 species) and compared it's performance against other metaheuristics like Differential Evolution (DE), Particle Swarm Optimization (PSO), Simulated Annealing (SA) and also against DDS alone.
 The new algorithm outperforms all the other metaheuristics on the coagulation model.
 The parameter vectors obtained using this approach fit the experimental data well and also make accurate enough predictions on unseen experimental data.
-We also performed this comparison on commonly used test functions for global optimization and found the same behavior.
-Further we used two benchmark problems, a genome wide kinetic model with 1759 parameters and a metabolic model of Chinese Hamster Ovary cells with 117 parameters to evaluate the performance of our approach.
-We  surprisingly outperformed the enhanced scatter search algorithm on these benchmarks and obtained the nominal parameter vector with just 4000 function evaluations.
+We also performed this comparison on commonly used test functions (Ackley and Rosenbrock) for global optimization and found the same behavior.
+Further we used two recently published benchmark problems, a genome wide kinetic model with 1759 parameters and a metabolic model of Chinese Hamster Ovary cells with 117 parameters to evaluate the performance of our approach.
+We  surprisingly performed well on these benchmarks and obtained the nominal parameter vector with just 4000 function evaluations in both cases.
 
@@
 Biological systems interact with one another and with the external world through highly complex biochemical networks.
-The advent of  \textit {omics} era has provided researchers with a deluge of data about these networks.
-Over the last decade, with an increase in computational power, large scale mathematical modeling of biological systems has evolved as a powerful paradigm to understand this data deluge and to effectively characterize these complex biochemical interactions.
+The advent of high throughout technologies has provided researchers with a deluge of data about these networks.
+Over the last decade, with an increase in computational power, large scale mathematical modeling has evolved as a powerful paradigm to characterize these complex biochemical interactions \cite{assmus2006dynamics, Riel:2006aa,Jaqaman:2006aa,kitano2002systems,hood2004systems}.
+Mathematical modeling of these networks is an iterative process that primarily consists of two tasks a) Model construction b) Model calibration or Parameter Estimation \cite{Aldridge:2006aa}\cite{banga2008optimization} \cite{ashyraliyev2009systems}.
+Model construction involves translating existing knowledge about the network (i.e.
+entities in the network, interactions amongst the entities, types of interactions) along with reasonable assumptions into an appropriate mathematical framework.
+Model calibration or parameter estimation follows the model construction step.
+In this step the model is trained against experimental data to determine the model parameters that give a good 'fit' against experimental data.
+Model training is generally framed as a non-linear optimization problem wherein the residual (or objective function) between experimental data and simulation data is minimized using various optimization techniques \cite{moles2003parameter}.
+The optimal parameters obtained from the previous step are then used to validate the model on hitherto unseen experimental data.
+If validation leads to unsatisfactory results, model construction and calibration are repeated iteratively till satisfactory results are obtained.
 
-The development of such mathematical models for biochemical networks can be broadly decomposed into two tasks a) Model construction b) Model calibration or Parameter Estimation.
-Model construction involves translating existing knowledge about the network (i.e.
-entities in the network, interactions amongst the entities, types of interactions) along with reasonable assumptions into an appropriate mathematical
-framework.  Some popular mathematical formulations have been Ordinary Differential Equations (ODEs), Boolean and fuzzy logic, probabilistic graphical models etc. [REFHERE]
+With the capacity to construct very large models using various mathematical formalisms \cite{chen2009input} \cite{tasseff2011modeling}\cite{luan2007computationally} \cite{mo2007genome} \cite{orth2011comprehensive} \cite{karr2012whole}\cite{buchel2013path2models} \cite{smallbone2010towards}, more often than not, parameter estimation is becoming a major challenge if not a bottleneck in large scale mathematical modeling.
+Though parameter estimation has been a well studied problem in engineering \cite{beck1977parameter}\cite{nieman1971review}\cite{young1981parameter} \cite {beck1998inverse}, the complex dynamics of large biological systems and noisy, often incomplete experimental data pose a more difficult and rather unique challenge.
+Most of the optimization problems pertaining to biological systems are non-linear and multi-modal i.e.
+most of them have multiple local minima or maxima \cite{moles2003parameter}\cite{banga2008optimization}.
+The non-linearity of the problem coupled with multi-modality generally renders the local optimization techniques like pattern search method \cite{hooke1961direct}, Nelder-Mead simplex method \cite{nelder1965simplex} or Levenberg-Marquardt method\cite{more1978levenberg} incapable of obtaining an optimal solution.
+Though deterministic global optimization techniques can handle non-linearity and multi-modality \cite{esposito2000deterministic} \cite{horst2013global}, the absence of derivative information, discontinuity of the objective functions, non-smooth regions or the lack of any knowledge about the objective function severely hamper the use of these techniques.
 
-Model calibration or parameter estimation follows the model construction step.
-In this step the model is trained against experimental data to determine the model parameters that give a good fit against experimental data.
-Model simulation using these parameters is then done and the results are validated using hitherto unseen experimental data.
-If validation leads to unsatisfactory results model construction and calibration are repeated iteratively till we get satisfactory results.
+Metaheuristic approaches like Genetic Algorithms, Simulated Annealing \cite{kirkpatrick1983optimization}, Evolutionary Programming and population based searches like Differential Evolution\cite{storn1997differential} have shown promise in this regard \cite{sun2012parameter}.
+They do not make any assumptions about the structure of objective function or require any \textit{a priori} information about the objective function.
+Though they do not guarantee strong convergence, these approaches are effective in finding near optimal solutions.
+Mendes et al.
+\cite{mendes1998non} used Simulated Annealing to estimate rate constants for the irreversible inhibition of HIV proteinase, Modchang et al.
+\cite {modchang2008mathematical} used Genetic Algorithms to estimate parameters for a mathematical model of signal transduction, Differential Evolution based approaches have been an effective approach on various systems \cite{tsai2005evolutionary}, \cite{wang2001hybrid},\cite{noman2007inferring}.
+Tashkova et al.
+\cite{tashkova2011parameter} compared different metaheuristics for parameter estimation on a dynamic model of endocytosis and showed that DE was the most effective.
+Banga and co-workers have applied scatter-search based methods \cite{villaverde2012cooperative} \cite{rodriguez2006novel} \cite{egea2007scatter} to estimate parameters on non-linear biological processes.
+Hybrid approaches that combine a metaheuristic with a local optimization search, wherein a near optimal solution that is obtained with a metaheuristic is further refined using a local search have also become popular.
+Villaverde et al.
+\cite{villaverde2015biopredyn} combined scatter search with local search methods for parameter estimation in large scale systems biology models.
+Fan et al.
+recently showed that population based metaheuristics along with decomposition based methods can be used to model gene circuits from mRNA data \cite{fan2015parameter}.
 
-With the capacity to construct or formulate very large biological models, more often than not, parameter estimation has become the bottleneck in large scale mathematical modeling of biological systems.
-The problem of parameter estimation if generally framed as a non-linear optimization problem.
-
-\textit  {\Large Equation needs to be written and explained terms}
-
-\begin{eqnarray}
-	\mathcal{G}_{new}(\mathbf{J})& = &\mathcal{G}(\mathbf{J})+\mathbf{r}_{normal}(\mathbf{J})\sigma(\mathbf{J})
-\end{eqnarray}
-
-The objective function from equation ---  is minimized using appropriate optimization technique to estimate the model parameters.
-Due to the complex dynamics of biological systems these problems are multi-modal i.e.
-most of them have multiple local minima or maxima.
-The non-linearity of the problem coupled with multi-modality is a substantial challenge and generally renders the local optimization techniques incapable of obtaining an optimal solution.
-\textit {In many of these problems derivative information is absent or the objective function may be discontinuous.
-This hinders the use of deterministic optimization techniques that require the problem to be convex and other derivative/gradient based optimization methods.
-Most of these methods are computationally expensive and for optimization of large-scale systems, this becomes prohibitively expensive.
-Heuristic approaches have been promising in this regard, however most of the population-based heuristics take a large number of function evaluations and have computationally complex operations, which reduces their effectiveness in finding feasible solutions with limited computational resources and time.
-Hybrid methods that couple heuristics with derivative based methods}
-
-The problem get further compounded when we have parameter vectors of very high dimensions.
-In this study we developed a novel meta heuristic approach that combines multi-swarm particle swarm optimization with dynamically dimensioned search (DDS)to obtain optimal or near optimal solutions to high-dimensional, non-linear biological models within a relatively few function evaluations.
-
-A bit of history on DDS
-Why coagulation model? feed back loops, parameter identifiability, tight regulation and also high dimensional
-
-\subsection *{Parameter Estimation Literature}
-
-\subsection *{Modeling methodologies Literature} 
- Shuler and co-workers constructed large-scale dynamic metabolic models of E.coli that were based ordinary differential equations (ODEs).
-Various aspects of E.colis cellular processes like RNA synthesis, chromosome synthesis, cellular energy expenditure etc.
-were modeled using coupled non-linear ordinary differential equations.
-Post genomic era saw the rise of flux balance analysis (FBA), a static, constraint based modeling approach for microbial metabolism.
-FBA assumes a pseudo steady-state and reduces genome-scale kinetic models to a set of static mass balances.
-These mass balances are represented using an underdetermined set of linear algebraic equations.
-The advantage is such a system of equations can be solved efficiently for very large systems.
-
-The dynamics of signal transduction systems are commonly modeled using ODEs [Refs]-----.
-One of the advantages of using ODEs is that the dynamics of large-scale systems can be described with a great degree of mechanistic detail.
-Over the last half a decade we have seen the construction of increasingly large-scale ODE models of signal transduction systems.
-Chen et al.
-built a mass action model of ErbB signaling pathways with 499 ODEs and 229 parameters.
-Tasseff et al analyzed retinoic acid induced differentiation of uncommitted precursor cells using a dynamic ODE model with 729 proteins and protein complexes interconnected with 1356 interconnections.
-Luan et al captured the key aspects of coagulation using an ODE model with 193 proteins and protein complexes with 301 kinetic parameters.
-
-  Boolean logic, probabilistic modeling and fuzzy logic are other mathematical frameworks that are used for mechanistic modeling of signal transduction systems.
-Choi et al modeled the p53 signaling network that affects cellular response to DNA damage using Boolean logic.
-------.
-In Boolean modeling of signal transduction systems the network entities i.e.
-proteins assume a value of 0 or 1 based on their activation levels.
-Fuzzy logic models offer a greater detail by allowing protein activation to take on a range of discrete values between 0 and 1.
-Mitsos et al.
-used the constrained fuzzy logic approach (cFL) wherein protein activation takes real values and a transfer function (TF) is used to propagate the signal along the network.
-Boolean and fuzzy logic models offer a relatively simpler mathematical formulation as compared to ODE models and are not encumbered with a large number of parameters.
-However these models are static in nature and offer a more qualitative view of the system.
-In addition to these deterministic formulations, stochastic modeling is another way of representing biochemical networks.
-Stochastic models are primarily based on stochastic differential equations also called Langevin equations to describe the dynamics of biochemical network ----- [Refs].
-Stochastic models are limited to describing dynamics of small-scale biochemical networks and are computationally very expensive to characterize even in very small biological systems.
+Despite these successes, a  major drawback with most of these approaches is the vast number of objective function evaluations they take in arriving at good search regions.
+As the models grow in size and complexity and the number of dimensions of parameter vector increases,  parameter estimation becomes a computationally expensive task.
+In many of these high dimensional problems approaching an exact solution may not be necessary.
+Gutenkust et al.
+\cite{gutenkunst2007universally} showed that a number of systems biology models are 'sloppy'.
+Sloppy systems have specific parameter combinations that largely define the dynamics of the system.
+Large perturbations to the rest of the parameters does not greatly impact the system dynamics.
+Ensemble approaches \cite{song2010ensembles,luan2010ensembles} have exploited this aspect to describe the dynamics of biological systems including coagulation which can be described using only a set of key species or parameters \cite{sagar2015dynamic}.
+Tolson and Shoemaker \cite{tolson2007dynamically} showed through Dynamically Dimensioned Search (DDS) that high-dimensional watershed models can be calibrated quickly by perturbing only a subset of dimensions.
+Thus in our current study we took into cognizance the fact that it may not be necessary to obtain the exact solution for high dimensional biological systems and that good enough solutions can be quickly obtained without expending a lot of objective function evaluations.
+Hence our current approach uses the power of population based heuristics along with DDS to obtain near optimal or good solutions.
+Though in theory we can combine any metaheuristic with DDS, for the current purpose we used Particle Swarm Optimization (PSO).
+PSO unlike Genetic Algorithm (GA) or Differential Evolution (DE) does not have complex operations like cross over, mutation or recombination.
+It is simple to use and does not have a lot of parameters associated with other heuristics.
+However PSO is known to rapidly converge to a local optimum and thus several variants \cite{peer2003using,zhan2009adaptive,li2007fast} have been developed of which Multi swarm particle swarm optimization approaches (MLSPSO) \cite{zhao2008dynamic,liang2005dynamic} are one of the effective ones.
+We used MLSPSO to preclude bad regions of search and thereafter search within these regions using DDS.
 
@@
 \section*{Results}
-\subsection{Coagulation model formulation}
-
-Coagulation is an archetype biochemical network that is highly interconnected and tightly regulated with multiple positive and negative feedback loops.
-The biochemistry underlying coagulation, though quite complex has been well studied [REFHERE], and reliable experimental coagulation models have been developed [REFHERE].
+\subsection*{Performance on Coagulation model}
+Coagulation is an archetype biochemical network that is highly interconnected and tightly regulated with multiple positive and negative feedback loops (Figure 1).
+The biochemistry underlying coagulation, though quite complex has been well studied \cite{mann2003dynamics,mann2003all,mann2003thrombin,vogler2009contact,diamond2013systems,fogelson2005coagulation,anand2003model}, and reliable experimental coagulation models have been developed \cite{hockin2002model,chatterjee2010systems,mann2006models,luan2007computationally}.
 This makes it an ideal system for mathematical modeling and parameter estimation.
@@
 These zymogens are activated through certain triggers.
-These trigger events like injury or trauma or sepsis expose factors like collagen, tissue factor and von Willebrand factor (vWF) to blood.
+Trigger events like injury or trauma or sepsis expose factors like collagen, tissue factor and von Willebrand factor (vWF) to blood.
 The exposure of these factors to blood kick-starts a series of convergent cascades that lead to conversion of zymogen prothrombinase to thrombin.
 Luan et al.
-modeled coagulation using coupled non-linear ordinary differential equations with 148 reactions and 92 species [REFHERE].
-The model was validated using 21 published datasets.
-The model parameters used in the simulation were drawn from various literature sources.
+modeled coagulation using coupled non-linear ordinary differential equations with 148 reactions and 92 species \cite{luan2007computationally} and validated model using 21 published datasets.
 
-We compared the performance of DDSMLSPSO on this model of coagulation against commonly used meta heuristics like Simulated Annealing (SA), Differential Evolution (DE), Particle Swarm Optimization (PSO) and also against DDS.
-To train the model parameters we used data sets from TF-VIIa initiated coagulation with no anticoagulants.
-The objective error function is a linear combination of two different error functions that used data sets representing coagulation initiated with different concentrations of TF-VIIa (5pm, 5nm).
-Since this is an expensive error function we restricted the number of function evaluations to 4000 for each algorithm.
-We performed 25 trials of this experiment.
+We compared the performance of Dynamically Dimensioned Search with Multi Swarm Particle Swarm Optimization (DDSMLSPSO) on this model of coagulation against commonly used meta heuristics like Simulated Annealing (SA), Differential Evolution (DE), Particle Swarm Optimization (PSO) and also against Dynamically Dimensioned Search (DDS).
+To train the model parameters, we used data sets from TF-VIIa initiated coagulation with no anticoagulants.
+The objective function is a weighted linear combination of two different error functions that used data sets representing coagulation initiated with different concentrations of TF-VIIa (5pM, 5nM) \cite{hockin2002model}.
+We restricted the number of function evaluations to 4000 for each algorithm and performed 25 trials of this experiment.
 DDSMLSPSO exhibits a much faster rate of error convergence and has a much lower final error as compared to the other algorithms (Figure 3).
 Within the first 1000 function evaluations (swarm phase) of DDSMLSPSO there is a very rapid drop in error.
-Subsequently, again after 2500 function evaluations (dynamically dimensioned phase) the error drops quickly.
-Overall at the end of 4000 function evaluations DDSMLSPSO minimizes the error to much a greater extent than any of the other algorithms.
-Amongst the rest of the algorithms DDS and SA have a faster rate of drop in error as compared to PSO or DE.
-However after 1500 function evaluations, the objective error remains nearly constant.
-
-Figure 2 shows the fit between model predictions and experimental data.
+Subsequently, again after 2000 function evaluations (dynamically dimensioned phase) the error drops quickly.
+Overall at the end of 4000 function evaluations DDSMLSPSO minimizes the error (final objective error is 90456.9) to a much greater extent than any of the other algorithms.
+Amongst the rest of the algorithms DDS has a comparable error drop for the first 2000 function evaluations.
+However the drop in error remains constant for DDS after 2000 function evaluations.
+Using the parameters obtained at the end of 4000 function evaluations we examined the 'fits' between models predictions and experimental data (Figure 2).
 The solid lines represent the mean value of prediction over 25 trials and the shaded region represents the 99\% confidence interval.
-Figure 3 shows model predictions on completely 'unseen'  or untrained data sets where coagulation was initiated with 500pm, 50pm, 10pm concentrations of TF-VIIa respectively.
-The parameters that were obtained by simultaneous training on the 2 data sets were able to 'fit' the experimental data well within a few number of function evaluations.
+Subsequently we compared the model predictions against completely 'unseen' or untrained experimental where coagulation was initiated with 500pM,50pM,10pM concentrations of TF-VIIa respectively (Figure 3).
 
@@
 We compared the performance of these algorithms on commonly used test functions for global optimization.
-Figure-xxx shows the rate of error convergence on a 300D rastrigin function where DDSMLSPSO clearly outperforms the other approaches.
-In Figure-yyyy, on a 300D ackley function we see the same trend wherein  DDSMLSPSO reaches the minimum much faster than other algorithms.
+We used a 300 dimensional Rastrigin function and 300 dimensional Ackley function.
+Both Ackley and Rastrigin have multiple local minima and maxima and attain a global minimum value of 0.
+We tested the performance of the 5 different heuristic approaches on these two functions.
+In each experiment we again fixed the number of function evaluations at 4000 and ran 25 experiments.
+In both cases (Figure S6 A and B) we see that the error convergence rate for DDSMLSPSO is much faster as compared to the other four metaheuristics and it finds the global minimum of 0 in both cases.
 
-Villaverde and co-workers recently published a set of benchmark problems to evaluate parameter estimation methods [REFHERE].
+Villaverde and co-workers recently published a set of benchmark problems to evaluate parameter estimation methods \cite{villaverde2015biopredyn}.
 From a computational cost perspective problems they categorized the problems as most expensive, intermediate and least expensive.
 We evaluated the performance of our algorithms on a problem from the most expensive and least expensive categories.
-Table 1 shows that
+The first problem is a genome wide kinetic model of \textit{Saccharomyces cerevisiae} with 1759 parameters (henceforth referred to as problem B1).
+The second problem (henceforth referred to as problem B4) is a metabolic model of Chinese Hamster Ovary (CHO) with 35 metabolites, 32 reactions and 117 parameters.
+In both cases pseudo time series data was generated.
+For problem B1, the time series data consisted of 44 observables and for problem B4 the data corresponded to 13 different metabolites.
+We fixed the number of function evaluations at 4000 for DDSMLSPSO and trained both the models against the pseudo experimental data.
+We recaptured the 'nominal' parameters (Figure S7 A and B)  in both cases within 4000 evaluations.
+The final objective function value (Table 1) is an order smaller for B1 possibly due to overfitting.
+In Table 1 we summarize the results obtained on the coagulation model and the benchmark problems.
 
-\begin{table}[h!]
+\begin{table}[h]
 \centering
-\begin{tabular}{|c c c c c c c|} 
+\caption{Table with optimization settings and results for coagulation and different benchmarks using DDSMLSPSO.}
+\label{table:1}
+\resizebox{\columnwidth}{!}{
+\begin{tabular}{|cccccc|}
 \hline
- ModelID & Coagulation & B1 & B4 & Ackley300D & Rastrigin300D&\\ [0.5ex] 
+                        & Coagulation           & \multicolumn{4}{l}{Benchmarks}\\
  \hline  
- Upper bound&$1000.pnom$& ${5.pnom}$ & ${5.pnom}$ &${30}$ &$5.12$& \\  [0.5ex]
- Lower bound&$0.001.pnom$& ${0.2.pnom}$ & ${0.2.pnom}$&${-15}$&$-5.12$&\\ [0.5ex]
- CPU time &$5.6$ ${hours}$ &$38.308$ ${hours}$ & $6.2$ ${minutes}$& ${2.8}$${seconds}$&${2.6}$${seconds}$& \\ [0.5ex]
- Function Evaluations &$4000$&$4000$&$4000$&$4000$&$4000$&\\ [0.5ex]
- Initial Objective Value & $1.0589.10 ^{10}$ & $1.4275.10 ^{7}$ &$1.8536.10 ^{7}$&$21.12$&$99.985$&\\ [0.5ex] 
- Final Objective Value & $2.9223.10^{3}$ & $4.96.10^{5}$ & $38.9375$&$0$&$0$&\\ [0.5ex]
- Nominal Objective Value & $4.7785·10^{6}$& $1.0986·10^{6}$ & $39.0676$&$0$&$0$&\\ [1ex]
+                        & Experimental Data Set & B1           & B4          & Ackley 300D & Rastrigin 300D \\
+ \hline                        
+Lower Bound             & 0.001.pnom            & 5.pnom       & 5.pnom      & 30          & 5.12           \\
+Upper Bound             & 1000.pnom             & 0.2.pnom     & 0.2.pnom    & -15         & -5.12          \\
+CPU Time                & 10.0835 hours         & 38.308 hours & 6.2 minutes & 2.8 seconds & 2.6 seconds    \\
+Function Evaluations    & 4000                  & 4000         & 4000        & 4000        & 4000           \\
+Initial Objective Value & 1.11837e+07           & 1.4275e+07   & 1.8536e+07  & 21.12       & 99.985         \\
+Final Objective Value   & 90456.9               & 4.96e+05     & 38.9375     & 0           & 0              \\
+Nominal Objective Value & 4.7785e+06            & 1.0986e+06   & 39.0676     & 0           & 0      \\       
 \hline
 \end{tabular}
-\caption{Table with optimization settings and results for coagulation and different benchmarks using DDSMLSPSO.}
-\label{table:1}
+}
 \end{table}
@@
 \section*{Discussion}
+Our study presents a novel approach for high-dimensional parameter estimation in complex biological systems with relatively few function evaluations.
+In this approach we combined a variant of a well known meta heuristic particle swarm optimization with Dynamically Dimensioned Search (DDS).
+We tested our approach on an ODE model of coagulation with 148 parameters and 92 species.
+Coagulation is an ideal system to test our approach since the biology is well known and complex, with multiple feed back loops that are tightly regulated.
+We used experimental data with under different conditions to obtain optimal parameters and used these parameters to make predictions against unseen experimental data.
+We obtained good fits and made accurate enough predictions using parameters obtained from 4000 function evaluations.
+Further, we also used high-dimensional forms of commonly used test functions of global optimization and showed that we were able to find the global minimum for 300 dimensional Ackley and Rastrigin functions faster than other metaheuristics.
+We also considered two recently published benchmark problems to test parameter optimization approaches and showed that we were able to retrieve the nominal parameter vector within 4000 function evaluations.
+
+Metaheuristic approaches are generally effective in finding close to optimum solutions of complex, multi-modal functions.
+In addition, they generally obviate the need for any \textit{a priori} knowledge (like function derivative).
+However they take an exorbitant number of objective function evaluations to come close to an optimum.
+When the objective function evaluations tend to become expensive it is infeasible to take up a large number of evaluations.
+As the dimensionality of parameter space increases, the search region gets widened and thus the problem becomes more challenging.
+In addition, most of these approaches require optimization of  'algorithm parameters' before the actual optimization and also involve computationally expensive update operations.
+Tolson and Shoemaker, through DDS, showed that randomly perturbing a subset of dimensions in high dimensional parameter space is an effective way to obtain near optimal solutions with few function evaluations.
+However their approach was based on a single solution.
+Single solutions based methods may prove ineffective in complex systems when they start from a 'bad' region of search or from a local minimum.
+In our approach, we tried to eliminate the probability of starting from a bad region of search using a variant of particle swarm optimization.
+Particle Swarm Optimization (PSO) is a population based meta heuristic that is known for rapid convergence without any complex operations like recombination, mutation or selection that are associated with other population based metaheuristics like Differential Evolution (DE) or Genetic Algorithm (GA).
+Several particle swarm variants have been proposed to improve the search ability and rate of convergence, that involve different neighborhood structures, multi-swarms or adaptive parameters.
+Multi-swarm PSO with small particle neighborhoods have been shown to better in searching on complex multi-modal solutions \cite{zhao2008dynamic}.
+Multi swarm methods, in addition, avoid rapid convergence to a local optimum or stable point and are able to generate diverse solutions.
+Generation of diverse solutions in the early stage gives a better exploratory capability and thus converge of upon multiple optima.
+
+We utilized this capacity of multi swarms to generate diverse candidate solutions which can be used as initial solutions for a DDS like search.
+Thus the solution vectors obtained at the end of multi swarm search phase have a better propensity to not get stuck in bad regions in DDS phase as opposed to starting with a totally random initial solution.
+Given a fixed number of function evaluations, we have been able to show that we were able to obtain better solutions for the coagulation model and the test functions faster that other commonly known metaheuristics and also DDS alone.
+Choosing the number of function evaluations is largely a function of cost and complexity of the objective function.
+Traditionally the stopping condition for a parameter optimization problem can be the number of function evaluations, percentage of initial error achieved or an absolute error threshold.
+However in case of complex, expensive functions where we desire a value within a certain period of time, the number of evaluations are used as a stopping criterion.
+In our current study we used a value of 4000 which we based upon the time taken (approximately 8-10 seconds) for a single objective function evaluation in the coagulation case.
+We used the same value of 4000 for benchmarks published by Villaverde and co-workers \cite{villaverde2015biopredyn}.
+Quite surprisingly we took a couple of orders lesser number of function evaluations (1e3) to obtain the optimal parameter vector as compared to the enhanced Scatter Search (eSS) with a local optimizer, which took around the order of 1e5 number of evaluations.
+The amount of CPU time taken (on an Intel Xeon processor 2.4 GHz) is lesser as compared to eSS on a similar architecture.
+
+A surprisingly remarkable aspect about our algorithm has been that we did not 'pre optimize' any parameters of the algorithm to suit a specific problem.
+In the swarm phase this includes the number of particles, number of sub-swarms, acceleration constants or the number of generations after which the particles are redistributed and the neighborhood perturbation parameter in DDS phase.
+We used the same parameters for all the problems.
+The same rule was applied to the rest of the metaheuristics barring Simulated Annealing.
+For SA, we optimized the cooling schedule for the coagulation model.
+Thus, in this approach any overhead that usually comes with additional function evaluations to optimize parameters was avoided.
+The only parameter that we adjusted was the fraction of evaluations within swarm search.
+For the coagulation model it was set to be 0.4 and for the benchmarks it is at 0.6.
+This parameter was set based on empirical observations rather than any definitive analysis.
+When the error convergence rates stagnated in the swarm phase we switched to DDS phase.
+A further improvement to our approach could  be to incorporate a meta intelligent way of switching in and switching out of both the phases.
+
+The performance of our approach seems impressive given that it performed well on different complex systems with no specific conditioning being required.
+We comfortably outperformed common metaheuristics and were also able to find minima of high dimensional global optimization test functions.
+Thus this approach may be well suited to large scale global optimization.
+In addition, surprisingly, we were able to obtain optimal parameter vectors for two different large scale systems biology models with a couple of orders fewer number of function evaluations as compared a hybrid approach.
+However it is quite possible that highly optimized versions of common meta heuristics may outperform us on these systems.
+This aspect is currently beyond the scope of this study.
+Our approach can also be combined with local derivative based searches to improve upon the accuracy of the solutions.
+
 \clearpage
 \section*{Materials and Methods}
-\subsection*{Formulation and solution of coagulation model equations.}
 \subsection*{Optimization Strategy}
-Dynamically dimensioned multi-swarm particle swarm optimization (DDSMLSPSO) is a novel optimization approach that combines elements from multi-swarm based PSO methods  with DDS.
+Dynamically dimensioned multi-swarm particle swarm optimization (DDSMLSPSO) is a novel optimization approach that combines elements from multi-swarm based PSO methods  with DDS (Figure 2).
 The goal of this approach is to obtain optimal or near optimal parameters for high-dimensional complex biological systems within a pre-specified number of function evaluations.
 
- We randomly initialized a swarm of $\mathcal{K}$-dimensional particles (represented as ${x}_{i}$), wherein each of these particles corresponds to an $\mathcal{K}$-dimensional parameter vector.
+ We randomly initialized a swarm of $\mathcal{K}$-dimensional particles (represented as ${x}_{i}$), wherein each of these particles corresponds to a $\mathcal{K}$-dimensional parameter vector.
 After initialization, the particles were grouped into different sub-swarms randomly.
@@
 where $\mathbf{N}$ represents the total number of function evaluations, $\mathbf{w}_{max}$ and $\mathbf{w}_{min}$ are the maximum and minimum inertia weights respectively.
-While updating the particle, we made sure all dimensions of the solution represented by the particle was within bounds using a reflection boundary condition.
+While updating the particle, we made sure all dimensions of the solution represented by the particle were within bounds using a reflection boundary condition.
 After every $\mathbf{g}$  function evaluations, the particles within all sub-swarms were mixed and then randomly redistributed to a new sub-swarm.
@@
 
-where $\mathbf{J}$ represents the a vector containing the specific dimensions being perturbed, ${r}_{normal}$ denotes a normal random vector of the same dimensions as $\mathcal{G}$.
+where $\mathbf{J}$ represents the set containing the specific dimensions being perturbed, ${r}_{normal}$ denotes a normal random vector of the same dimensions as $\mathcal{G}$.
 $\sigma$ is the amplitude of perturbation given by following equation
@@
 where $\mathbf{R}$ is the scalar perturbation size parameter, $\mathcal{MAX}$ and $\mathcal{MIN}$ are ($\mathcal{K}\times{1}$) vectors that represent the maximum and minimum bounds on each dimension.
-The probability $\mathcal{P}$ that a specific dimension is perturbed is a  monotonically decreasing function that decreases with the number of function evaluations.
-$\mathcal{P}$ can be any monotonically decreasing function, in our approach we used the following function
+The set $\mathbf{J}$ was constructed using a probability function $\mathcal{P}_{j}$ that represented the probability whether a specific dimension $j$ was perturbed or not.
+This function is a monotonically decreasing function that decreases with the number of function evaluations.
+$\mathcal{P}_{j}$ can be any monotonically decreasing function, in our approach we used the following function
 \begin{eqnarray}
@@
 Thus the number of dimensions of the candidate vector that are updated or perturbed decreases with the as the number of function evaluations increase.
-These updates are greedy in nature that is $\mathcal{G}_{new}$ becomes the new solution vector only if it is better than the old one $\mathcal{G}$.
+These updates are greedy in nature, so $\mathcal{G}_{new}$ becomes the new solution vector only if it is better than the old one $\mathcal{G}$.
+
+%\begin{figure}[H]
+%\centering
+%\includegraphics[width=1.00\textwidth,height=0.7\textheight]{./figs/Figure_2_CoagulationNetwork.pdf}
+%\caption{Schematic of the extrinsic and intrinsic coagulation cascade.
+Inactive zymogens upstream (grey) are activated by exposure to tissue factor (TF) \cite{giesen1999blood} following vessel injury.
+Tissue factor and activated factor VIIa (FVIIa) form a complex that activates factor X (fX) and IX (fIX).
+FXa activates downstream factors including factor VIII (fVIII) and fIX.
+Factor V (fV) is primarily activated by thrombin (FIIa).
+In addition, we included a secondary fV activation route involving FXa.
+FXa and FVa form a complex (prothrombinase) on activated platelets that converts prothrombin (fII) to FIIa.
+FIXa and FVIIIa can also form a complex (tenase) on activated platelets which catalyzes FXa formation.
+Localized platelets are activated by external signals such as adenosine diphosphate (ADP) and thromboxane A2 (TXA2) or thrombin through protease-activated receptors (PARs) \protect\cite{coughlin2000thrombin} \protect \cite{coughlin1999protease} Thrombin also activates upstream coagulation factors, forming a strong positive feedback ensuring rapid activation.
+Tissue factor pathway inhibitor (TFPI) downregulates FXa formation and activity by sequestering free FXa and TF-FVIIa in a FXa-dependent manner.
+Antithrombin III (ATIII)  inhibits all proteases.
+Thrombin inhibits itself binding the surface protein thrombomodulin (TM) \protect\cite{esmon1989roles} The IIa-TM complex catalyzes the conversion of protein C (PC) to activated protein C (APC), which attenuates the coagulation response by the proteolytic cleavage of fV/FVa and fVIII/FVIIIa \protect\cite{esmon1989roles}}\label{fig-network}
+%\end{figure}
 
@@
 \includegraphics[width=1.00\textwidth,height=0.7\textheight]{./figs/Figure_2_CoagulationNetwork.pdf}
-\caption{Schematic of the extrinsic and intrinsic coagulation cascade.Upstream coagulation factors are activated by materials exposed following vessel injury chief among these tissue factor (TF) [Ref] Tissue factor and activated factor VIIa (FVIIa) form a complex that activates factor X (fX) and IX (fIX).
+\caption{Schematic of the extrinsic and intrinsic coagulation cascade.
+Inactive zymogens upstream (grey) are activated by exposure to tissue factor (TF)  following vessel injury.
+Tissue factor and activated factor VIIa (FVIIa) form a complex that activates factor X (fX) and IX (fIX).
 FXa activates downstream factors including factor VIII (fVIII) and fIX.
@@
 In addition, we included a secondary fV activation route involving FXa.
-FXa and FVa form a complex (prothrombinase) on activated platelets that converts
-prothrombin (fII) to FIIa. FIXa and FVIIIa can also form a complex (tenase) on activated platelets which catalyzes FXa formation. Localized
-platelets are activated by external signals such as adenosine diphosphate (ADP) and thromboxane A2 (TXA2) or thrombin through
-protease-activated receptors (PARs)[Ref] Thrombin also activates upstream coagulation factors, forming a strong positive feedback ensuring
-rapid activation. Tissue factor pathway inhibitor (TFPI) downregulates FXa formation and activity by sequestering free FXa and TF-FVIIa in a
-FXa-dependent manner. Antithrombin III (ATIII) neutralizes all proteases, making it perhaps the most powerful control element in the cascade.
-Thrombin itself plays an inadvertent role in its own inhibition by binding the surface protein thrombomodulin (TM), expressed on normal
-vasculature[Ref] The IIa-TM complex catalyzes the conversion of protein C (PC) to activated protein C (APC), which attenuates the coagulation response by the proteolytic cleavage of fV/FVa and fVIII/FVIIIa [Ref] }\label{fig-flux-pattern}
+FXa and FVa form a complex (prothrombinase) on activated platelets that converts prothrombin (fII) to FIIa.
+FIXa and FVIIIa can also form a complex (tenase) on activated platelets which catalyzes FXa formation.
+Thrombin also activates upstream coagulation factors, forming a strong positive feedback ensuring rapid activation.
+Tissue factor pathway inhibitor (TFPI) downregulates FXa formation and activity by sequestering free FXa and TF-FVIIa in a FXa-dependent manner.
+Antithrombin III (ATIII)  inhibits all proteases.
+Thrombin inhibits itself binding the surface protein thrombomodulin (TM) .
+The IIa-TM complex catalyzes the conversion of protein C (PC) to activated protein C (APC), which attenuates the coagulation response by the proteolytic cleavage of fV/FVa and fVIII/FVIIIa }\label{fig-network}
 \end{figure}
@@
 \textbf{A}: Each particle represents an N dimensional parameter vector.
-Particles are randomly initialized and grouped into various sub-swarms.
-The magnitude and direction of the movement a particle is influenced by the position of the best particle in its swarm and also by its own experience.
-After a certain number of function evaluations the particles are mixed and randomly assigned to different swarms.
-At the end of the evaluations assigned to swarm search, the global best particle amongst all sub-swarms is chosen as the candidate parameter vector for Dynamically Dimensioned Search \textbf{B}: The candidate vector does a greedy search in a dynamic neighborhood.
-This done by dynamically adjusting the number of parameter dimensions that are perturbed in each evaluation step.
-The number of dimensions that are perturbed generally decreases as the number of iterations increase.
-This preserves the optimality of the solution as the number of evaluations increases.
+Particles are randomly initialized and grouped into different sub-swarms.
+Within each swarm the magnitude and direction of the movement a particle is influenced by the position of the best particle and also by its own experience.
+After every $\mathbf{g}$ number of function evaluations the particles are mixed and randomly assigned to different swarms.
+At the end of the evaluations (\mathcal{FR}) assigned to swarm search, the global best particle (orange color) amongst all sub-swarms is chosen as the candidate parameter vector for Dynamically Dimensioned Search \textbf{B}: The candidate vector does a greedy search in a dynamic neighborhood.
+The search neighborhood is dynamically adjusted by varying the number of dimensions that are perturbed (in black) in each evaluation step.
+The probability that a dimension is perturbed decreases as the number of function evaluations increase.
+Thus as the evaluations increase the optimality of the solution is preserved.
 }\label{fig-algorithm}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.0\textwidth,height=0.5\textheight]{./figs/Figure_3_Errors_convergence.pdf}
+\caption{Error convergence rates of the 5 different algorithms on the coagulation model.
+The objective error is the mean over $N$= 25 trials.
+DDSMLSPSO and DDS have the steepest drop in error during first 2000 function evaluations.
+Thereafter the error drop in DDS remains nearly constant whereas DDSMLSPSO drops further from around 2000 to 4000 function evaluations.
+At the end of 4000 function evaluations DDSMLSPSO attains the lowest error.
+}\label{fig-train}
 \end{figure}
@@
 The grey shaded region represents the 99\% confidence estimate of the mean simulated thrombin concentration.
-The experimental data is reproduced from the synthetic plasma assays of Mann and co-workers [REFHERE].
+The experimental data is reproduced from the synthetic plasma assays of Mann and co-workers.
 Thrombin generation is initiated by adding Factor VIIa-TF (5nM - Red and 5pM - Green) to synthetic plasma containing 200 $\mu$mol/L of phospholipid vesicles (PCPS) and a mixture of coagulation factors (II,V,VII,VIII,IX,X and XI) at their mean plasma concentrations.
@@
 The grey shaded region represents the 99\% confidence estimate of the mean simulated thrombin concentration.
-The experimental data is reproduced from the synthetic plasma assays of Mann and co-workers [REFHERE].
+The experimental data is reproduced from the synthetic plasma assays of Mann and co-workers.
 Thrombin generation is initiated by adding Factor VIIa-TF (500pM - Blue, 50pM - Pink and 10pM - Orange respectively) to synthetic plasma containing 200 $\mu$mol/L of phospholipid vesicles (PCPS) and a mixture of coagulation factors (II,V,VII,VIII,IX,X and XI) at their mean plasma concentrations.
 }\label{fig-validation}
-\end{figure}
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=1.02\textwidth,height=0.5\textheight]{./figs/Figure6_Ackley_Rast.pdf}
-\caption{Error convergence rates on a 300 dimensional Rastrigin function and Ackley function.
-Ackley and Rastrigin are commonly used test functions for global optimization.
-DDSMLSPSO finds the minimum $ {0} $ much faster than other approaches in both cases within 4000 function evaluations \textbf {A} On a 300-D Rastrigin, DDSMLPSO, DDS and SA perform the best followed by PSO and DE  \textbf {B} On a 300-D Ackley, DDSMLSPSO clearly outperforms the rest of the algorithms.
-}\label{fig-Rast}
-\end{figure}
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=1.02\textwidth,height=0.5\textheight]{./figs/Figure_7_Benchmarks_Parameters}
-\caption{Difference between optimal and nominal parameter vector values on benchmark problems [REFHERE].
-\textbf {(A)} Problem B1: Genome wide kinetic model of E.coli \textit{S.cerevisiae} with 1759 unknown parameters.
-\textbf {(B)} Problem B4: Metabolic model of Chinese  Hamster Ovary Cells (CHO) cells with 117 parameters.
-}\label{fig-benchmark}
 \end{figure}
@@
 \clearpage
+%\bibliography{References_v1}
+%\begin{thebibliography}{9999}
+\bibliography{References_v2}
 
-\bibliography{References_v1}
+%\bibitem[Saltelli \em{et~al.}(2010)Saltelli, Annoni, Azzini, Campolongo, Ratto,
+  %and Tarantola]{Saltelli:2010}
+%Saltelli, A.; Annoni, P.; Azzini, I.; Campolongo, F.; Ratto, M.; Tarantola, S.
+%\newblock Variance based sensitivity analysis of model output. Design and
+  %estimator for the total sensitivity index.
+%\newblock {\em Comput. Phys. Commun.} {\bf 2010}, {\em 181},~259--270.
+
+%\end{thebibliography}
 
@@
 
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.02\textwidth,height=0.5\textheight]{./figs/Figure_6_Ackley_Rast}
+\caption{Error convergence on 300 dimensional Ackley function and Rastrigin function.
+\textbf {(A)} Error convergence of 5 different metaheuristics on 300 dimensional Ackley function \textbf {(B)} Error convergence of 5 different metaheuristics on 300 dimensional Rastrigin function
+}\label{fig-testfunctions}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.02\textwidth,height=0.5\textheight]{./figs/Figure_7_Benchmarks_Parameters}
+\caption{Difference between optimal and nominal parameter vector values on benchmark problems.
+\textbf {(A)} Problem B1: Genome wide kinetic model of E.coli \textit{S.cerevisiae} with 1759 unknown parameters.
+\textbf {(B)} Problem B4: Metabolic model of Chinese  Hamster Ovary Cells (CHO) cells with 117 parameters.
+}\label{fig-benchmark}
+\end{figure}
+
 % Reset the counters -

### 2015-06-12 Paper_v2.tex
Experimental Methods Added

--- 2014-10-30 Paper_v2.tex
@@
 Androgens, such as testosterone, are required for prostate cancer growth. 
-Androgen ablation in combination with radiation or chemotherapy remains the primary non-surgical treatment for androgen-dependent prostate cancer. 
+Androgen ablation in combination with radiation or chemotherapy remains the primary non-surgical treatment for androgen dependent prostate cancer. 
 However, androgen ablation typically fails to permanently arrest cancer progression, often resulting in castration resistant prostate cancer (CRPC). 
@@
 Initially, PCa cells depend upon the activation of cytosolic androgen receptors (AR) by androgen hormones, such as testosterone, for survival and growth. 
-Androgen ablation in combination with radiation or chemotherapy remains the primary non-surgical treatment for androgen-dependent prostate cancer (ADPC) \cite{Huggins1967}. 
+Androgen ablation in combination with radiation or chemotherapy remains the primary non-surgical treatment for androgen dependent prostate cancer (ADPC) \cite{Huggins1967}. 
 However, androgen ablation typically fails to permanently arrest cancer progression as malfunctioning cells eventually lose androgen sensitivity and proliferate without hormone. 
@@
 
+\subsection*{Experimental Validation.}
+
+\subsection*{Cell culture and treatments}
+Androgen dependent LNCaP prostate cancer cells were a gift from Dr.
+Brian Kirby (Cornell University), and the Castrate Resistant C4-2 prostate cancer cell line was purchased from MD Anderson Cancer Center, University of Texas.
+Cell lines were maintained in RPMI 1640 media (Life Technologies, Inc., Grand Island, NY) with 10\% fetal calf serum (FBS; Hyclone) and 1x antibiotic/antimycotic (Sigma, St.
+Louis, MO) in a 5\% CO$_2$ humidified atmosphere at 37$^{\circ}$C.
+The AR inhibitor MDV3100 (enzalutamide) and the Raf inhibitor sorafenib were purchased from SantaCruz Biotechnology (Santa Cruz, CA).
+The PI3K inhibitor LY294002 was purchased from Cell Signaling Technologies (Danvers, MA, USA).
+All stock solutions were diluted in DMSO (Sigma, St.
+Louis, MO).
+
+\subsection*{Protein Extraction and Western Blot}
+
+LNCaP and C4-2 cells were seeded in 60 mm dishes at a density of 4 x 10$^{5}$.
+After 96 and 72 hrs, for LNCaP and C4-2 cells respectively, the media was replaced with fresh media and drug treatments were added.
+After 24 hours, cells were washed twice in PBS buffer, scraped in 250 $\mu$L ice-cold lysis buffer (Pierce, Rockford, IL) supplemented with protease and phosphatase inhibitors (Sigma, St.
+Louis, MO), and lysed for 30 min on ice.
+Lysates were centrifuged at 13,000 rpm for 30 min at 4$^{\circ}$C.
+After quantification of total protein by BCA assay, equal amounts of total protein lysates (25 $\mu$g) were resolved by SDS-PAGE and transferred onto PVDF membranes.
+Membranes were blocked in 5\% fat free milk and then probed with antibodies.
+The primary antibodies used for western blot analysis were pAKT Ser473, AKT, pS6 Ser240/244 , pERK Thr202/Tyr204, ERK, AR, and GAPDH were from Cell Signaling Technologies (Danvers, MA, USA).
+For detection, enhanced chemiluminescence ECL reagent (GE Healthcare, Pittsburgh, PA) was used and signals were visualized using the ChemiDoc XRS system (Bio-Rad).
+
+\subsection*{MTT Assay}
+
+LNCaP and C4-2 cells were seeded at a density of 1x10$^{4}$ cells per well in 96 well plates.
+After 48 hrs the media was refreshed and drug treatments added.
+Cell growth at 24, 48, and 72 hrs was determined using a 3-(4,5-dimethyl thiazol-2-yl)-2,5-diphenyl tetrazolium bromide (MTT) assay.
+At the specified time point 10 $\mu$ MTT reagent (stock of 5 mg/mL in PBS) was added to each well and the cells were further incubated for 4 hrs.
+At 4 hrs, the media was removed and 50 $\mu$L of dissolving reagent DMSO was added to each well.
+After an additional 10 min incubation, the absorbance was measured at 540 nm on a microplate reader.
+Each reading was adjusted by subtracting the absorbance value for the blank (media only) and the results were then scaled to the DMSO-treated (control) case.
+
 \section*{Acknowledgements}

### 2015-06-30 Paper_v1.tex
Some initial edits

@@
 %\usepackage[authoryear,round,comma,sort&compress]{natbib}
-\usepackage[square,sort,comma,numbers]{natbib}
+\usepackage[square,sort,comma,numbers,sort&compress]{natbib}
 %\usepackage[authoryear,round]{natbib}
@@
 \usepackage{simplemargins}
+\usepackage{fullpage}
 \usepackage{comment}
 \usepackage{lineno}
-\usepackage{chicago}
+%\usepackage{chicago}
 \usepackage{textcomp}
@@
 \doublespace
-
-%\usepackage{geometry}
-\usepackage{fullpage}
 
@@
 \begin{titlepage}
-{\par\centering\textbf{\Large A novel metaheuristic approach to obtain near optimal solutions in high dimensional biochemical networks with a few function evaluations}}
+{\par\centering\textbf{\Large {Dynamically dimensioned particle swarm optimization (DdPSO): A meta-heuristic for parameter estimation in large biochemical models.}}}
 \vspace{0.05in}
@@
 \vspace{0.1in}
-{\par \centering \textbf{Running Title:}~Parameter estimation in large scale biological systems}
+{\par \centering \textbf{Running Title:}~Parameter estimation in biochemical models}
 \vspace{0.1in}
@@
 We use a particle swarm optimization technique that uses multi-swarms to generate candidate vectors which are then greedily updated using DDS by dynamically varying the perturbed parameter dimensions.
-We tested this algorithm (25 trials with 4000 function evaluations in each trial) on a biochemical network of coagulation (148 parameters and 92 species) and compared it's performance against other metaheuristics like Differential Evolution (DE), Particle Swarm Optimization (PSO), Simulated Annealing (SA) and also against DDS alone.
-The new algorithm outperforms all the other metaheuristics on the coagulation model.
+We tested this algorithm (25 trials with 4000 function evaluations in each trial) on a biochemical network of coagulation (148 parameters and 92 species) and compared it's performance against other meta-heuristics like Differential Evolution (DE), Particle Swarm Optimization (PSO), Simulated Annealing (SA) and also against DDS alone.
+The new algorithm outperforms all the other meta-heuristics on the coagulation model.
 The parameter vectors obtained using this approach fit the experimental data well and also make accurate enough predictions on unseen experimental data.
@@
 Over the last decade, with an increase in computational power, large scale mathematical modeling has evolved as a powerful paradigm to characterize these complex biochemical interactions \cite{assmus2006dynamics, Riel:2006aa,Jaqaman:2006aa,kitano2002systems,hood2004systems}.
-Mathematical modeling of these networks is an iterative process that primarily consists of two tasks a) Model construction b) Model calibration or Parameter Estimation \cite{Aldridge:2006aa}\cite{banga2008optimization} \cite{ashyraliyev2009systems}.
+Mathematical modeling of these networks is an iterative process that primarily consists of two tasks a) Model construction b) Model calibration or Parameter Estimation \cite{Aldridge:2006aa,banga2008optimization,ashyraliyev2009systems}.
 Model construction involves translating existing knowledge about the network (i.e.
@@
 
-With the capacity to construct very large models using various mathematical formalisms \cite{chen2009input} \cite{tasseff2011modeling}\cite{luan2007computationally} \cite{mo2007genome} \cite{orth2011comprehensive} \cite{karr2012whole}\cite{buchel2013path2models} \cite{smallbone2010towards}, more often than not, parameter estimation is becoming a major challenge if not a bottleneck in large scale mathematical modeling.
-Though parameter estimation has been a well studied problem in engineering \cite{beck1977parameter}\cite{nieman1971review}\cite{young1981parameter} \cite {beck1998inverse}, the complex dynamics of large biological systems and noisy, often incomplete experimental data pose a more difficult and rather unique challenge.
+With the capacity to construct very large models using various mathematical formalisms \cite{chen2009input,tasseff2011modeling,luan2007computationally,mo2007genome,orth2011comprehensive,karr2012whole,buchel2013path2models,smallbone2010towards}, more often than not, parameter estimation is becoming a major challenge if not a bottleneck in large scale mathematical modeling.
+Though parameter estimation has been a well studied problem in engineering \cite{nieman1971review,beck1977parameter,young1981parameter,beck1998inverse}, the complex dynamics of large biological systems and noisy, often incomplete experimental data pose a more difficult and rather unique challenge.
 Most of the optimization problems pertaining to biological systems are non-linear and multi-modal i.e.
-most of them have multiple local minima or maxima \cite{moles2003parameter}\cite{banga2008optimization}.
+most of them have multiple local minima or maxima \cite{moles2003parameter,banga2008optimization}.
 The non-linearity of the problem coupled with multi-modality generally renders the local optimization techniques like pattern search method \cite{hooke1961direct}, Nelder-Mead simplex method \cite{nelder1965simplex} or Levenberg-Marquardt method\cite{more1978levenberg} incapable of obtaining an optimal solution.
-Though deterministic global optimization techniques can handle non-linearity and multi-modality \cite{esposito2000deterministic} \cite{horst2013global}, the absence of derivative information, discontinuity of the objective functions, non-smooth regions or the lack of any knowledge about the objective function severely hamper the use of these techniques.
+Though deterministic global optimization techniques can handle non-linearity and multi-modality \cite{esposito2000deterministic,horst2013global}, the absence of derivative information, discontinuity of the objective functions, non-smooth regions or the lack of any knowledge about the objective function severely hamper the use of these techniques.
 
-Metaheuristic approaches like Genetic Algorithms, Simulated Annealing \cite{kirkpatrick1983optimization}, Evolutionary Programming and population based searches like Differential Evolution\cite{storn1997differential} have shown promise in this regard \cite{sun2012parameter}.
+Meta-heuristic approaches like Genetic Algorithms, Simulated Annealing \cite{kirkpatrick1983optimization}, Evolutionary Programming and population based searches like Differential Evolution\cite{storn1997differential} have shown promise in this regard \cite{sun2012parameter}.
 They do not make any assumptions about the structure of objective function or require any \textit{a priori} information about the objective function.
@@
 \cite{mendes1998non} used Simulated Annealing to estimate rate constants for the irreversible inhibition of HIV proteinase, Modchang et al.
-\cite {modchang2008mathematical} used Genetic Algorithms to estimate parameters for a mathematical model of signal transduction, Differential Evolution based approaches have been an effective approach on various systems \cite{tsai2005evolutionary}, \cite{wang2001hybrid},\cite{noman2007inferring}.
+\cite {modchang2008mathematical} used Genetic Algorithms to estimate parameters for a mathematical model of signal transduction, Differential Evolution based approaches have been an effective approach on various systems \cite{tsai2005evolutionary,wang2001hybrid,noman2007inferring}.
 Tashkova et al.
-\cite{tashkova2011parameter} compared different metaheuristics for parameter estimation on a dynamic model of endocytosis and showed that DE was the most effective.
-Banga and co-workers have applied scatter-search based methods \cite{villaverde2012cooperative} \cite{rodriguez2006novel} \cite{egea2007scatter} to estimate parameters on non-linear biological processes.
-Hybrid approaches that combine a metaheuristic with a local optimization search, wherein a near optimal solution that is obtained with a metaheuristic is further refined using a local search have also become popular.
+\cite{tashkova2011parameter} compared different meta-heuristics for parameter estimation on a dynamic model of endocytosis and showed that DE was the most effective.
+Banga and co-workers have applied scatter-search based methods \cite{villaverde2012cooperative,rodriguez2006novel,egea2007scatter} to estimate parameters on non-linear biological processes.
+Hybrid approaches that combine a meta-heuristic with a local optimization search, wherein a near optimal solution that is obtained with a meta-heuristic is further refined using a local search have also become popular.
 Villaverde et al.
@@
 Fan et al.
-recently showed that population based metaheuristics along with decomposition based methods can be used to model gene circuits from mRNA data \cite{fan2015parameter}.
+recently showed that population based meta-heuristics along with decomposition based methods can be used to model gene circuits from mRNA data \cite{fan2015parameter}.
 
@@
 Hence our current approach uses the power of population based heuristics along with DDS to obtain near optimal or good solutions.
-Though in theory we can combine any metaheuristic with DDS, for the current purpose we used Particle Swarm Optimization (PSO).
+Though in theory we can combine any meta-heuristic with DDS, for the current purpose we used Particle Swarm Optimization (PSO).
 PSO unlike Genetic Algorithm (GA) or Differential Evolution (DE) does not have complex operations like cross over, mutation or recombination.
@@
 \section*{Results}
-\subsection*{Performance on Coagulation model}
-Coagulation is an archetype biochemical network that is highly interconnected and tightly regulated with multiple positive and negative feedback loops (Figure 1).
-The biochemistry underlying coagulation, though quite complex has been well studied \cite{mann2003dynamics,mann2003all,mann2003thrombin,vogler2009contact,diamond2013systems,fogelson2005coagulation,anand2003model}, and reliable experimental coagulation models have been developed \cite{hockin2002model,chatterjee2010systems,mann2006models,luan2007computationally}.
-This makes it an ideal system for mathematical modeling and parameter estimation.
-Coagulation is regulated by a set of serine proteases also known as coagulation factors and blood platelets.
-The coagulation factors are generally in an inactive state and are known as zymogens.
-These zymogens are activated through certain triggers.
-Trigger events like injury or trauma or sepsis expose factors like collagen, tissue factor and von Willebrand factor (vWF) to blood.
-The exposure of these factors to blood kick-starts a series of convergent cascades that lead to conversion of zymogen prothrombinase to thrombin.
-Luan et al.
-modeled coagulation using coupled non-linear ordinary differential equations with 148 reactions and 92 species \cite{luan2007computationally} and validated model using 21 published datasets.
+\subsection*{Dynamically dimensioned particle swarm optimization (DdPSO).}
+DdPSO is a novel meta-heuristic that combines multi-swarm based particle swarm methods with dynamically dimensioned search (Fig. \ref{fig-algorithm}).
+The goal of this approach is to obtain optimal or near optimal parameters for high-dimensional complex biological systems within a pre-specified number of function evaluations.
+We randomly initialized a swarm of $\mathcal{K}$-dimensional particles (represented as ${x}_{i}$), wherein each of these particles corresponds to a $\mathcal{K}$-dimensional parameter vector.
+After initialization, the particles were grouped into different sub-swarms randomly.
+Thereafter within each sub-swarm ${S}_{k}$,  particles were updated according to the following rule.
+\begin{eqnarray}
+	\mathbf{x}_{i,j} &=&\theta_{1,j-1}\mathbf{x}_{i,j-1} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{x}_{i,j-1}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{GL}_{k} - \mathbf{x}_{i,j-1}\right)
+\end{eqnarray}
+where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes the best local solution found by particle $i$ till function evaluation $j$-$1$, and
+$\mathcal{GL}_{k}$ denotes the best local solution found over the population of all particles within the swarm ${S}_{k}$.
+The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
+parameters ($\mathcal{K}\times{1}$). In our algorithm the parameter $\theta_{1,j-1}$ depends on the function evaluations and is controlled according to the following equation
+\begin{eqnarray}
+	\mathbf \theta_{1,j}&=&((\mathbf{N}-{j})*(\mathbf{w}_{max}-\mathbf{w}_{min}))/(\mathbf{N}-{1}) + \mathbf{w}_{min}
+\end{eqnarray}
+where $\mathbf{N}$ represents the total number of function evaluations, $\mathbf{w}_{max}$ and $\mathbf{w}_{min}$ are the maximum and minimum inertia weights respectively.
+While updating the particle, we made sure all dimensions of the solution represented by the particle were within bounds using a reflection boundary condition.
+After every $\mathbf{g}$  function evaluations, the particles within all sub-swarms were mixed and then randomly redistributed to a new sub-swarm.
+The particles were then again updated according Eq.1.
+This process continued till $\mathcal{FR}*\mathbf{N}$ number of functions evaluations, where $\mathcal{FR}$ represents the fraction of evaluations with the multi-swarms.
+At the end of these function evaluations, we froze all the solutions represented by various particles and chose the particle with best solution among $\mathcal{GL}_{1} \cdots \mathcal{GL}_{NS}$ as the initial candidate vector $\mathcal{G}$ for the remaining $({1}-\mathcal{FR})*\mathbf{N}$ number of function evaluations.
 
-We compared the performance of Dynamically Dimensioned Search with Multi Swarm Particle Swarm Optimization (DDSMLSPSO) on this model of coagulation against commonly used meta heuristics like Simulated Annealing (SA), Differential Evolution (DE), Particle Swarm Optimization (PSO) and also against Dynamically Dimensioned Search (DDS).
-To train the model parameters, we used data sets from TF-VIIa initiated coagulation with no anticoagulants.
-The objective function is a weighted linear combination of two different error functions that used data sets representing coagulation initiated with different concentrations of TF-VIIa (5pM, 5nM) \cite{hockin2002model}.
-We restricted the number of function evaluations to 4000 for each algorithm and performed 25 trials of this experiment.
-DDSMLSPSO exhibits a much faster rate of error convergence and has a much lower final error as compared to the other algorithms (Figure 3).
-Within the first 1000 function evaluations (swarm phase) of DDSMLSPSO there is a very rapid drop in error.
-Subsequently, again after 2000 function evaluations (dynamically dimensioned phase) the error drops quickly.
-Overall at the end of 4000 function evaluations DDSMLSPSO minimizes the error (final objective error is 90456.9) to a much greater extent than any of the other algorithms.
-Amongst the rest of the algorithms DDS has a comparable error drop for the first 2000 function evaluations.
-However the drop in error remains constant for DDS after 2000 function evaluations.
-Using the parameters obtained at the end of 4000 function evaluations we examined the 'fits' between models predictions and experimental data (Figure 2).
-The solid lines represent the mean value of prediction over 25 trials and the shaded region represents the 99\% confidence interval.
-Subsequently we compared the model predictions against completely 'unseen' or untrained experimental where coagulation was initiated with 500pM,50pM,10pM concentrations of TF-VIIa respectively (Figure 3).
+This particle was then updated according to the following rule
+\begin{eqnarray}
+	\mathcal{G}_{new}(\mathbf{J})& = &\mathcal{G}(\mathbf{J})+\mathbf{r}_{normal}(\mathbf{J})\sigma(\mathbf{J})
+\end{eqnarray}
+where $\mathbf{J}$ represents the set containing the specific dimensions being perturbed, ${r}_{normal}$ denotes a normal random vector of the same dimensions as $\mathcal{G}$.
+$\sigma$ is the amplitude of perturbation given by following equation:
+\begin{eqnarray}
+	\sigma& = &\mathbf{R}(\mathcal{MAX} -\mathcal{MIN})
+\end{eqnarray}
+where $\mathbf{R}$ is the scalar perturbation size parameter, $\mathcal{MAX}$ and $\mathcal{MIN}$ are ($\mathcal{K}\times{1}$) vectors that represent the maximum and minimum bounds on each dimension.
+The set $\mathbf{J}$ was constructed using a probability function $\mathcal{P}_{j}$ that represented the probability whether a specific dimension $j$ was perturbed or not.
+This function is a monotonically decreasing function that decreases with the number of function evaluations.
+$\mathcal{P}_{j}$ can be any monotonically decreasing function, in our approach we used the following function:
+\begin{eqnarray}
+	\mathcal{P}_{j}&=&{1}-\log(j/ (\beta({1}-\mathcal{FR})*\mathbf{N}))
+\end{eqnarray}
+where $\beta$ is the perturbation frequency probability modulator.
+Thus the number of dimensions of the candidate vector that are updated or perturbed decreases with the as the number of function evaluations increase.
+These updates are greedy in nature, so $\mathcal{G}_{new}$ becomes the new solution vector only if it is better than the old one $\mathcal{G}$.
 
-\subsection*{Performance on Benchmark functions}
-
-We compared the performance of these algorithms on commonly used test functions for global optimization.
+\subsection*{Performance of DdPSO on benchmark problems.}
+We compared the performance of DdPSO with algorithms on commonly used test functions for global optimization, as well as benchmark biochemical problems from literature (Table \ref{table:benchmark-problems}).
 We used a 300 dimensional Rastrigin function and 300 dimensional Ackley function.
@@
 In each experiment we again fixed the number of function evaluations at 4000 and ran 25 experiments.
-In both cases (Figure S6 A and B) we see that the error convergence rate for DDSMLSPSO is much faster as compared to the other four metaheuristics and it finds the global minimum of 0 in both cases.
+In both cases (Figure S6 A and B) we see that the error convergence rate for DDSMLSPSO is much faster as compared to the other four s and it finds the global minimum of 0 in both cases.
 
@@
 
+\subsection*{Performance of DdPSO on a model of the human coagulation cascade.}
+We compared the performance of DdPSO on a model of blood coagulation against commonly used meta-heuristics like simulated annealing (SA), differential evolution (DE), particle swarm optimization (PSO),and dynamically dimensioned search (DDS).
+Coagulation is an archetype biochemical network that is highly interconnected and tightly regulated with multiple positive and negative feedback loops (Fig.
+\ref{fig-coagulation-network}).
+The biochemistry underlying coagulation, though quite complex has been well studied \cite{mann2003dynamics,mann2003all,mann2003thrombin,vogler2009contact,diamond2013systems,fogelson2005coagulation,anand2003model}, and reliable experimental coagulation models have been developed \cite{hockin2002model,chatterjee2010systems,mann2006models,luan2007computationally}.
+This makes it an ideal system for mathematical modeling and parameter estimation.
+Coagulation is regulated by a set of serine proteases also known as coagulation factors and blood platelets.
+The coagulation factors are generally in an inactive state and are known as zymogens.
+These zymogens are activated through certain triggers.
+Trigger events like injury or trauma or sepsis expose factors like collagen, tissue factor and von Willebrand factor (vWF) to blood.
+The exposure of these factors to blood kick-starts a series of convergent cascades that lead to conversion of zymogen prothrombinase to thrombin.
+Luan et al.
+modeled coagulation using coupled non-linear ordinary differential equations with 148 reactions and 92 species \cite{luan2007computationally} and validated model using 21 published datasets.
+
+ To train the model parameters, we used data sets from TF-VIIa initiated coagulation with no anticoagulants.
+The objective function was a weighted linear combination of two different error functions that used data sets representing coagulation initiated with different concentrations of TF-VIIa (5pM, 5nM) \cite{hockin2002model}.
+We restricted the number of function evaluations to 4000 for each algorithm and performed 25 trials of this experiment.
+DdPSO exhibits a much faster rate of error convergence and has a much lower final error as compared to the other algorithms (Figure 3).
+Within the first 1000 function evaluations (swarm phase) of DDSMLSPSO there is a very rapid drop in error.
+Subsequently, again after 2000 function evaluations (dynamically dimensioned phase) the error drops quickly.
+Overall at the end of 4000 function evaluations DDSMLSPSO minimizes the error (final objective error is 90456.9) to a much greater extent than any of the other algorithms.
+Amongst the rest of the algorithms DDS has a comparable error drop for the first 2000 function evaluations.
+However the drop in error remains constant for DDS after 2000 function evaluations.
+Using the parameters obtained at the end of 4000 function evaluations we examined the 'fits' between models predictions and experimental data (Figure 2).
+The solid lines represent the mean value of prediction over 25 trials and the shaded region represents the 99\% confidence interval.
+Subsequently we compared the model predictions against completely 'unseen' or untrained experimental where coagulation was initiated with 500pM,50pM,10pM concentrations of TF-VIIa respectively (Figure 3).
+
+\clearpage
+\section*{Discussion}
+Our study presents a novel approach for high-dimensional parameter estimation in complex biological systems with relatively few function evaluations.
+In this approach we combined a variant of a well known meta heuristic particle swarm optimization with Dynamically Dimensioned Search (DDS).
+We tested our approach on an ODE model of coagulation with 148 parameters and 92 species.
+Coagulation is an ideal system to test our approach since the biology is well known and complex, with multiple feed back loops that are tightly regulated.
+We used experimental data with under different conditions to obtain optimal parameters and used these parameters to make predictions against unseen experimental data.
+We obtained good fits and made accurate enough predictions using parameters obtained from 4000 function evaluations.
+Further, we also used high-dimensional forms of commonly used test functions of global optimization and showed that we were able to find the global minimum for 300 dimensional Ackley and Rastrigin functions faster than other meta-heuristics.
+We also considered two recently published benchmark problems to test parameter optimization approaches and showed that we were able to retrieve the nominal parameter vector within 4000 function evaluations.
+
+meta-heuristic approaches are generally effective in finding close to optimum solutions of complex, multi-modal functions.
+In addition, they generally obviate the need for any \textit{a priori} knowledge (like function derivative).
+However they take an exorbitant number of objective function evaluations to come close to an optimum.
+When the objective function evaluations tend to become expensive it is infeasible to take up a large number of evaluations.
+As the dimensionality of parameter space increases, the search region gets widened and thus the problem becomes more challenging.
+In addition, most of these approaches require optimization of  'algorithm parameters' before the actual optimization and also involve computationally expensive update operations.
+Tolson and Shoemaker, through DDS, showed that randomly perturbing a subset of dimensions in high dimensional parameter space is an effective way to obtain near optimal solutions with few function evaluations.
+However their approach was based on a single solution.
+Single solutions based methods may prove ineffective in complex systems when they start from a 'bad' region of search or from a local minimum.
+In our approach, we tried to eliminate the probability of starting from a bad region of search using a variant of particle swarm optimization.
+Particle Swarm Optimization (PSO) is a population based meta heuristic that is known for rapid convergence without any complex operations like recombination, mutation or selection that are associated with other population based meta-heuristics like Differential Evolution (DE) or Genetic Algorithm (GA).
+Several particle swarm variants have been proposed to improve the search ability and rate of convergence, that involve different neighborhood structures, multi-swarms or adaptive parameters.
+Multi-swarm PSO with small particle neighborhoods have been shown to better in searching on complex multi-modal solutions \cite{zhao2008dynamic}.
+Multi swarm methods, in addition, avoid rapid convergence to a local optimum or stable point and are able to generate diverse solutions.
+Generation of diverse solutions in the early stage gives a better exploratory capability and thus converge of upon multiple optima.
+
+We utilized this capacity of multi swarms to generate diverse candidate solutions which can be used as initial solutions for a DDS like search.
+Thus the solution vectors obtained at the end of multi swarm search phase have a better propensity to not get stuck in bad regions in DDS phase as opposed to starting with a totally random initial solution.
+Given a fixed number of function evaluations, we have been able to show that we were able to obtain better solutions for the coagulation model and the test functions faster that other commonly known meta-heuristics and also DDS alone.
+Choosing the number of function evaluations is largely a function of cost and complexity of the objective function.
+Traditionally the stopping condition for a parameter optimization problem can be the number of function evaluations, percentage of initial error achieved or an absolute error threshold.
+However in case of complex, expensive functions where we desire a value within a certain period of time, the number of evaluations are used as a stopping criterion.
+In our current study we used a value of 4000 which we based upon the time taken (approximately 8-10 seconds) for a single objective function evaluation in the coagulation case.
+We used the same value of 4000 for benchmarks published by Villaverde and co-workers \cite{villaverde2015biopredyn}.
+Quite surprisingly we took a couple of orders lesser number of function evaluations (1e3) to obtain the optimal parameter vector as compared to the enhanced Scatter Search (eSS) with a local optimizer, which took around the order of 1e5 number of evaluations.
+The amount of CPU time taken (on an Intel Xeon processor 2.4 GHz) is lesser as compared to eSS on a similar architecture.
+
+A surprisingly remarkable aspect about our algorithm has been that we did not 'pre optimize' any parameters of the algorithm to suit a specific problem.
+In the swarm phase this includes the number of particles, number of sub-swarms, acceleration constants or the number of generations after which the particles are redistributed and the neighborhood perturbation parameter in DDS phase.
+We used the same parameters for all the problems.
+The same rule was applied to the rest of the meta-heuristics barring Simulated Annealing.
+For SA, we optimized the cooling schedule for the coagulation model.
+Thus, in this approach any overhead that usually comes with additional function evaluations to optimize parameters was avoided.
+The only parameter that we adjusted was the fraction of evaluations within swarm search.
+For the coagulation model it was set to be 0.4 and for the benchmarks it is at 0.6.
+This parameter was set based on empirical observations rather than any definitive analysis.
+When the error convergence rates stagnated in the swarm phase we switched to DDS phase.
+A further improvement to our approach could  be to incorporate a meta intelligent way of switching in and switching out of both the phases.
+
+The performance of our approach seems impressive given that it performed well on different complex systems with no specific conditioning being required.
+We comfortably outperformed common meta-heuristics and were also able to find minima of high dimensional global optimization test functions.
+Thus this approach may be well suited to large scale global optimization.
+In addition, surprisingly, we were able to obtain optimal parameter vectors for two different large scale systems biology models with a couple of orders fewer number of function evaluations as compared a hybrid approach.
+However it is quite possible that highly optimized versions of common meta heuristics may outperform us on these systems.
+This aspect is currently beyond the scope of this study.
+Our approach can also be combined with local derivative based searches to improve upon the accuracy of the solutions.
+
+%\begin{figure}[H]
+%\centering
+%\includegraphics[width=1.00\textwidth,height=0.7\textheight]{./figs/Figure_2_CoagulationNetwork.pdf}
+%\caption{Schematic of the extrinsic and intrinsic coagulation cascade.
+Inactive zymogens upstream (grey) are activated by exposure to tissue factor (TF) \cite{giesen1999blood} following vessel injury.
+Tissue factor and activated factor VIIa (FVIIa) form a complex that activates factor X (fX) and IX (fIX).
+FXa activates downstream factors including factor VIII (fVIII) and fIX.
+Factor V (fV) is primarily activated by thrombin (FIIa).
+In addition, we included a secondary fV activation route involving FXa.
+FXa and FVa form a complex (prothrombinase) on activated platelets that converts prothrombin (fII) to FIIa.
+FIXa and FVIIIa can also form a complex (tenase) on activated platelets which catalyzes FXa formation.
+Localized platelets are activated by external signals such as adenosine diphosphate (ADP) and thromboxane A2 (TXA2) or thrombin through protease-activated receptors (PARs) \protect\cite{coughlin2000thrombin} \protect \cite{coughlin1999protease} Thrombin also activates upstream coagulation factors, forming a strong positive feedback ensuring rapid activation.
+Tissue factor pathway inhibitor (TFPI) downregulates FXa formation and activity by sequestering free FXa and TF-FVIIa in a FXa-dependent manner.
+Antithrombin III (ATIII)  inhibits all proteases.
+Thrombin inhibits itself binding the surface protein thrombomodulin (TM) \protect\cite{esmon1989roles} The IIa-TM complex catalyzes the conversion of protein C (PC) to activated protein C (APC), which attenuates the coagulation response by the proteolytic cleavage of fV/FVa and fVIII/FVIIIa \protect\cite{esmon1989roles}}\label{fig-network}
+%\end{figure}
+
+\section*{Acknowledgements}
+This study was supported by an award from the Army Research Office (ARO \#59155-LS).
+\clearpage
+
+%\bibliography{References_v1}
+%\begin{thebibliography}{9999}
+\bibliography{References_v2}
+
+\clearpage
+
 \begin{table}[h]
 \centering
-\caption{Table with optimization settings and results for coagulation and different benchmarks using DDSMLSPSO.}
-\label{table:1}
+\caption{Table with optimization settings and results for coagulation and different benchmarks using DdPSO.}
+\label{table:benchmark-problems}
 \resizebox{\columnwidth}{!}{
@@
 \clearpage
-\section*{Discussion}
-Our study presents a novel approach for high-dimensional parameter estimation in complex biological systems with relatively few function evaluations.
-In this approach we combined a variant of a well known meta heuristic particle swarm optimization with Dynamically Dimensioned Search (DDS).
-We tested our approach on an ODE model of coagulation with 148 parameters and 92 species.
-Coagulation is an ideal system to test our approach since the biology is well known and complex, with multiple feed back loops that are tightly regulated.
-We used experimental data with under different conditions to obtain optimal parameters and used these parameters to make predictions against unseen experimental data.
-We obtained good fits and made accurate enough predictions using parameters obtained from 4000 function evaluations.
-Further, we also used high-dimensional forms of commonly used test functions of global optimization and showed that we were able to find the global minimum for 300 dimensional Ackley and Rastrigin functions faster than other metaheuristics.
-We also considered two recently published benchmark problems to test parameter optimization approaches and showed that we were able to retrieve the nominal parameter vector within 4000 function evaluations.
 
-Metaheuristic approaches are generally effective in finding close to optimum solutions of complex, multi-modal functions.
-In addition, they generally obviate the need for any \textit{a priori} knowledge (like function derivative).
-However they take an exorbitant number of objective function evaluations to come close to an optimum.
-When the objective function evaluations tend to become expensive it is infeasible to take up a large number of evaluations.
-As the dimensionality of parameter space increases, the search region gets widened and thus the problem becomes more challenging.
-In addition, most of these approaches require optimization of  'algorithm parameters' before the actual optimization and also involve computationally expensive update operations.
-Tolson and Shoemaker, through DDS, showed that randomly perturbing a subset of dimensions in high dimensional parameter space is an effective way to obtain near optimal solutions with few function evaluations.
-However their approach was based on a single solution.
-Single solutions based methods may prove ineffective in complex systems when they start from a 'bad' region of search or from a local minimum.
-In our approach, we tried to eliminate the probability of starting from a bad region of search using a variant of particle swarm optimization.
-Particle Swarm Optimization (PSO) is a population based meta heuristic that is known for rapid convergence without any complex operations like recombination, mutation or selection that are associated with other population based metaheuristics like Differential Evolution (DE) or Genetic Algorithm (GA).
-Several particle swarm variants have been proposed to improve the search ability and rate of convergence, that involve different neighborhood structures, multi-swarms or adaptive parameters.
-Multi-swarm PSO with small particle neighborhoods have been shown to better in searching on complex multi-modal solutions \cite{zhao2008dynamic}.
-Multi swarm methods, in addition, avoid rapid convergence to a local optimum or stable point and are able to generate diverse solutions.
-Generation of diverse solutions in the early stage gives a better exploratory capability and thus converge of upon multiple optima.
-
-We utilized this capacity of multi swarms to generate diverse candidate solutions which can be used as initial solutions for a DDS like search.
-Thus the solution vectors obtained at the end of multi swarm search phase have a better propensity to not get stuck in bad regions in DDS phase as opposed to starting with a totally random initial solution.
-Given a fixed number of function evaluations, we have been able to show that we were able to obtain better solutions for the coagulation model and the test functions faster that other commonly known metaheuristics and also DDS alone.
-Choosing the number of function evaluations is largely a function of cost and complexity of the objective function.
-Traditionally the stopping condition for a parameter optimization problem can be the number of function evaluations, percentage of initial error achieved or an absolute error threshold.
-However in case of complex, expensive functions where we desire a value within a certain period of time, the number of evaluations are used as a stopping criterion.
-In our current study we used a value of 4000 which we based upon the time taken (approximately 8-10 seconds) for a single objective function evaluation in the coagulation case.
-We used the same value of 4000 for benchmarks published by Villaverde and co-workers \cite{villaverde2015biopredyn}.
-Quite surprisingly we took a couple of orders lesser number of function evaluations (1e3) to obtain the optimal parameter vector as compared to the enhanced Scatter Search (eSS) with a local optimizer, which took around the order of 1e5 number of evaluations.
-The amount of CPU time taken (on an Intel Xeon processor 2.4 GHz) is lesser as compared to eSS on a similar architecture.
-
-A surprisingly remarkable aspect about our algorithm has been that we did not 'pre optimize' any parameters of the algorithm to suit a specific problem.
-In the swarm phase this includes the number of particles, number of sub-swarms, acceleration constants or the number of generations after which the particles are redistributed and the neighborhood perturbation parameter in DDS phase.
-We used the same parameters for all the problems.
-The same rule was applied to the rest of the metaheuristics barring Simulated Annealing.
-For SA, we optimized the cooling schedule for the coagulation model.
-Thus, in this approach any overhead that usually comes with additional function evaluations to optimize parameters was avoided.
-The only parameter that we adjusted was the fraction of evaluations within swarm search.
-For the coagulation model it was set to be 0.4 and for the benchmarks it is at 0.6.
-This parameter was set based on empirical observations rather than any definitive analysis.
-When the error convergence rates stagnated in the swarm phase we switched to DDS phase.
-A further improvement to our approach could  be to incorporate a meta intelligent way of switching in and switching out of both the phases.
-
-The performance of our approach seems impressive given that it performed well on different complex systems with no specific conditioning being required.
-We comfortably outperformed common metaheuristics and were also able to find minima of high dimensional global optimization test functions.
-Thus this approach may be well suited to large scale global optimization.
-In addition, surprisingly, we were able to obtain optimal parameter vectors for two different large scale systems biology models with a couple of orders fewer number of function evaluations as compared a hybrid approach.
-However it is quite possible that highly optimized versions of common meta heuristics may outperform us on these systems.
-This aspect is currently beyond the scope of this study.
-Our approach can also be combined with local derivative based searches to improve upon the accuracy of the solutions.
+\begin{figure}[h]
+\centering
+\includegraphics[width=1.0\textwidth,height=0.5\textheight]{./figs/Figure_1_Algorithm.pdf}
+\caption{Multi Swarm Particle Swarm Optimization with Dynamically Dimensioned Search.
+\textbf{A}: Each particle represents an N dimensional parameter vector.
+Particles are randomly initialized and grouped into different sub-swarms.
+Within each swarm the magnitude and direction of the movement a particle is influenced by the position of the best particle and also by its own experience.
+After every $\mathbf{g}$ number of function evaluations the particles are mixed and randomly assigned to different swarms.
+At the end of the evaluations ($\mathcal{FR}$) assigned to swarm search, the global best particle (orange color) amongst all sub-swarms is chosen as the candidate parameter vector for Dynamically Dimensioned Search \textbf{B}: The candidate vector does a greedy search in a dynamic neighborhood.
+The search neighborhood is dynamically adjusted by varying the number of dimensions that are perturbed (in black) in each evaluation step.
+The probability that a dimension is perturbed decreases as the number of function evaluations increase.
+Thus as the evaluations increase the optimality of the solution is preserved.
+}\label{fig-algorithm}
+\end{figure}
 
 \clearpage
-\section*{Materials and Methods}
-\subsection*{Optimization Strategy}
-Dynamically dimensioned multi-swarm particle swarm optimization (DDSMLSPSO) is a novel optimization approach that combines elements from multi-swarm based PSO methods  with DDS (Figure 2).
-The goal of this approach is to obtain optimal or near optimal parameters for high-dimensional complex biological systems within a pre-specified number of function evaluations.
 
- We randomly initialized a swarm of $\mathcal{K}$-dimensional particles (represented as ${x}_{i}$), wherein each of these particles corresponds to a $\mathcal{K}$-dimensional parameter vector.
-After initialization, the particles were grouped into different sub-swarms randomly.
-Thereafter within each sub-swarm ${S}_{k}$,  particles were updated according to the following rule.
-\begin{eqnarray}
-	\mathbf{x}_{i,j} &=&\theta_{1,j-1}\mathbf{x}_{i,j-1} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{x}_{i,j-1}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{GL}_{k} - \mathbf{x}_{i,j-1}\right)
-\end{eqnarray}
-
-where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes the best local solution found by particle $i$ till function evaluation $j$-$1$, and
-$\mathcal{GL}_{k}$ denotes the best local solution found over the population of all particles within the swarm ${S}_{k}$.
-The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
-parameters ($\mathcal{K}\times{1}$). In our algorithm the parameter $\theta_{1,j-1}$ depends on the function evaluations and is controlled according to the following equation
-
-\begin{eqnarray}
-	\mathbf \theta_{1,j}&=&((\mathbf{N}-{j})*(\mathbf{w}_{max}-\mathbf{w}_{min}))/(\mathbf{N}-{1}) + \mathbf{w}_{min}
-\end{eqnarray}
-
-where $\mathbf{N}$ represents the total number of function evaluations, $\mathbf{w}_{max}$ and $\mathbf{w}_{min}$ are the maximum and minimum inertia weights respectively.
-While updating the particle, we made sure all dimensions of the solution represented by the particle were within bounds using a reflection boundary condition.
-After every $\mathbf{g}$  function evaluations, the particles within all sub-swarms were mixed and then randomly redistributed to a new sub-swarm.
-The particles were then again updated according Eq.1.
-This process continued till $\mathcal{FR}*\mathbf{N}$ number of functions evaluations, where $\mathcal{FR}$ represents the fraction of evaluations with the multi-swarms.
-At the end of these function evaluations, we froze all the solutions represented by various particles and chose the particle with best solution among $\mathcal{GL}_{1} \cdots \mathcal{GL}_{NS}$ as the initial candidate vector $\mathcal{G}$ for the remaining $({1}-\mathcal{FR})*\mathbf{N}$ number of function evaluations.
-
-This particle was then updated according to the following rule
-
-\begin{eqnarray}
-	\mathcal{G}_{new}(\mathbf{J})& = &\mathcal{G}(\mathbf{J})+\mathbf{r}_{normal}(\mathbf{J})\sigma(\mathbf{J})
-\end{eqnarray}
-
-where $\mathbf{J}$ represents the set containing the specific dimensions being perturbed, ${r}_{normal}$ denotes a normal random vector of the same dimensions as $\mathcal{G}$.
-$\sigma$ is the amplitude of perturbation given by following equation
-
-\begin{eqnarray}
-	\sigma& = &\mathbf{R}(\mathcal{MAX} -\mathcal{MIN})
-\end{eqnarray}
-
-where $\mathbf{R}$ is the scalar perturbation size parameter, $\mathcal{MAX}$ and $\mathcal{MIN}$ are ($\mathcal{K}\times{1}$) vectors that represent the maximum and minimum bounds on each dimension.
-The set $\mathbf{J}$ was constructed using a probability function $\mathcal{P}_{j}$ that represented the probability whether a specific dimension $j$ was perturbed or not.
-This function is a monotonically decreasing function that decreases with the number of function evaluations.
-$\mathcal{P}_{j}$ can be any monotonically decreasing function, in our approach we used the following function
-\begin{eqnarray}
-	\mathcal{P}_{j}&=&{1}-\log(j/ (\beta({1}-\mathcal{FR})*\mathbf{N}))
-\end{eqnarray}
-
-where $\beta$ is the perturbation frequency probability modulator.
-Thus the number of dimensions of the candidate vector that are updated or perturbed decreases with the as the number of function evaluations increase.
-These updates are greedy in nature, so $\mathcal{G}_{new}$ becomes the new solution vector only if it is better than the old one $\mathcal{G}$.
-
-%\begin{figure}[H]
-%\centering
-%\includegraphics[width=1.00\textwidth,height=0.7\textheight]{./figs/Figure_2_CoagulationNetwork.pdf}
-%\caption{Schematic of the extrinsic and intrinsic coagulation cascade.
-Inactive zymogens upstream (grey) are activated by exposure to tissue factor (TF) \cite{giesen1999blood} following vessel injury.
-Tissue factor and activated factor VIIa (FVIIa) form a complex that activates factor X (fX) and IX (fIX).
-FXa activates downstream factors including factor VIII (fVIII) and fIX.
-Factor V (fV) is primarily activated by thrombin (FIIa).
-In addition, we included a secondary fV activation route involving FXa.
-FXa and FVa form a complex (prothrombinase) on activated platelets that converts prothrombin (fII) to FIIa.
-FIXa and FVIIIa can also form a complex (tenase) on activated platelets which catalyzes FXa formation.
-Localized platelets are activated by external signals such as adenosine diphosphate (ADP) and thromboxane A2 (TXA2) or thrombin through protease-activated receptors (PARs) \protect\cite{coughlin2000thrombin} \protect \cite{coughlin1999protease} Thrombin also activates upstream coagulation factors, forming a strong positive feedback ensuring rapid activation.
-Tissue factor pathway inhibitor (TFPI) downregulates FXa formation and activity by sequestering free FXa and TF-FVIIa in a FXa-dependent manner.
-Antithrombin III (ATIII)  inhibits all proteases.
-Thrombin inhibits itself binding the surface protein thrombomodulin (TM) \protect\cite{esmon1989roles} The IIa-TM complex catalyzes the conversion of protein C (PC) to activated protein C (APC), which attenuates the coagulation response by the proteolytic cleavage of fV/FVa and fVIII/FVIIIa \protect\cite{esmon1989roles}}\label{fig-network}
-%\end{figure}
-
-\begin{figure}[H]
+\begin{figure}[h]
 \centering
@@
 Antithrombin III (ATIII)  inhibits all proteases.
-Thrombin inhibits itself binding the surface protein thrombomodulin (TM) .
-The IIa-TM complex catalyzes the conversion of protein C (PC) to activated protein C (APC), which attenuates the coagulation response by the proteolytic cleavage of fV/FVa and fVIII/FVIIIa }\label{fig-network}
+Thrombin inhibits itself binding the surface protein thrombomodulin (TM).
+The IIa-TM complex catalyzes the conversion of protein C (PC) to activated protein C (APC), which attenuates the coagulation response by the proteolytic cleavage of fV/FVa and fVIII/FVIIIa.
+}\label{fig-coagulation-network}
 \end{figure}
 
-\begin{figure}[H]
-\centering
-\includegraphics[width=1.0\textwidth,height=0.5\textheight]{./figs/Figure_1_Algorithm.pdf}
-\caption{Multi Swarm Particle Swarm Optimization with Dynamically Dimensioned Search.
-\textbf{A}: Each particle represents an N dimensional parameter vector.
-Particles are randomly initialized and grouped into different sub-swarms.
-Within each swarm the magnitude and direction of the movement a particle is influenced by the position of the best particle and also by its own experience.
-After every $\mathbf{g}$ number of function evaluations the particles are mixed and randomly assigned to different swarms.
-At the end of the evaluations (\mathcal{FR}) assigned to swarm search, the global best particle (orange color) amongst all sub-swarms is chosen as the candidate parameter vector for Dynamically Dimensioned Search \textbf{B}: The candidate vector does a greedy search in a dynamic neighborhood.
-The search neighborhood is dynamically adjusted by varying the number of dimensions that are perturbed (in black) in each evaluation step.
-The probability that a dimension is perturbed decreases as the number of function evaluations increase.
-Thus as the evaluations increase the optimality of the solution is preserved.
-}\label{fig-algorithm}
-\end{figure}
+\clearpage
 
-\begin{figure}[H]
+\begin{figure}[h]
 \centering
@@
 
-\begin{figure}[H]
+\clearpage
+
+\begin{figure}[h]
 \centering
@@
 
-\section*{Acknowledgements}
-This study was supported by the National Science Foundation GK12 award (DGE-1045513) 
-and by the National Science Foundation CAREER award (FILLMEIN).
-
-\clearpage
-%\bibliography{References_v1}
-%\begin{thebibliography}{9999}
-\bibliography{References_v2}
-
 %\bibitem[Saltelli \em{et~al.}(2010)Saltelli, Annoni, Azzini, Campolongo, Ratto,
@@
 %\end{thebibliography}
-
-%\begin{comment}
 
@@
 
-\begin{figure}[H]
+% Reset the counters -
+\setcounter{equation}{0}
+\setcounter{table}{0}
+\setcounter{figure}{0}
+\setcounter{page}{1}
+
+\begin{figure}[ht]
 \centering
@@
 \caption{Error convergence on 300 dimensional Ackley function and Rastrigin function.
-\textbf {(A)} Error convergence of 5 different metaheuristics on 300 dimensional Ackley function \textbf {(B)} Error convergence of 5 different metaheuristics on 300 dimensional Rastrigin function
+\textbf {(A)} Error convergence of 5 different meta-heuristics on 300 dimensional Ackley function \textbf {(B)} Error convergence of 5 different meta-heuristics on 300 dimensional Rastrigin function
 }\label{fig-testfunctions}
@@
 
-\begin{figure}[H]
+\begin{figure}[ht]
 \centering
@@
 
-% Reset the counters -
-\setcounter{equation}{0}
-\setcounter{table}{0}
-\setcounter{figure}{0}
-\setcounter{page}{1}
-
 \end{document}

### 2015-06-30 Paper_v2.tex
Experiment Additions

@@
 \vspace{0.05in}
-{\par \centering \large{ Katharine V. Rogers, Joseph A. Wayman, Ryan Tasseff, Timothy Chu, Kyriakos Tsahalis, Spencer Davis, Rohit Yallamraju, Gongcheng Lu, Matthew P. DeLisa, 
+{\par \centering \large{ Katharine V. Rogers, Joseph A. Wayman, Ryan Tasseff, Caitlin Gee, Timothy Chu, Kyriakos Tsahalis, Spencer Davis, Rohit Yallamraju, Gongcheng Lu, Matthew P. DeLisa, 
 and Jeffrey D. Varner$^{*}$}}
@@
 In a constant DHT environment sensitive species were conserved between the cell lines. 
-These results suggest targeting the PI3K and MAPK pathways in addition to anti-androgen therapies as a treatment for AIPC.
+These results suggest targeting the PI3K and MAPK pathways in addition to anti-androgen therapies as a treatment for CRPC.
 
@@
 
+\subsection*{Experimental results confirm the need for dual therapies in prostate cancer.}
+
+The results from our sensitivity analysis indicate that instead of inhibiting solely the AR pathway (enzalutamide), a combination therapy is necessary. 
+To test this hypothesis we used the well characterized ADPC cell line LNCaP as well a LNCaP derived CRPC cell line C4-2 \cite{Thalmann1994}. 
+Three inhibitors were used: the AR inhibitor MDV3100 (enzalutamide), the Raf kinase inhibitor sorafenib, and the PI3K inhibitor LY294002. 
+In both cell lines, inhibition of either the AR or MAPK pathway appears to promote activation of the PI3K pathway, as seen by the increase in pAKT (S473) (Fig. \ref{fg:Experiment}A). 
+The addition of the PI3K inhibitor, LY294002, alone or in combination diminishes PI3K activity. 
+The inhibition of PI3K alone, led to an increase in AR expression in both LNCaP and C4-2 cell lines. 
+Since, AR transcriptionally upregulates its own expression \cite{Lin2003}\cite{Grad1999}, this may indicate an increase in AR activity. 
+The ribosomal protein pS6 was completely inhibited only in the presence of the PI3K inhibitor LY294002. 
+Cell viability results show a large decrease in cell viability at 72 hrs in the dual inhibition cases as well as the triple inhibition case for both LNCaP and C4-2 cell lines (Fig.
+\ref{fg:Experiment}B).
+I
+n both cell lines, MDV3100 (10 $\mu$M), has only a modest effect on cell viability versus control (DMSO). 
+(Fig. \ref{fg:Experiment}C) shows cell viability of both cell lines in varying concentrations of inhibitors at 24 hrs.
+
 \subsection*{Robustness analysis identified key regulators of prostate cancer.} 
@@
 The decrease in p70 (S6) activation due to inhibition of PI3K is shown in both the model and in the Carver \textit{et al.} study (Supplementary Figure), indicating this result is due to the PI3K pathway.
-It would be interesting to repeat the experiment looking at cell cycle proteins and also performing the experiment in CR LNCaP cells, instead of AD cells.
+Our results confirm the Carver \textit{et al.} study in that a dual inhibition of AR and PI3K signaling led to a more prominent decrease in cell viability then each of the inhibitors alone. 
+We extended the study to look at the addition of a third inhibitor, sorafenib, that inhibits Raf kinase in the MAPK pathway.
+
+%It would be interesting to repeat the experiment looking at cell cycle proteins and also performing the experiment in CR LNCaP cells, instead of AD cells.
 
@@
 Cell growth at 24, 48, and 72 hrs was determined using a 3-(4,5-dimethyl thiazol-2-yl)-2,5-diphenyl tetrazolium bromide (MTT) assay.
-At the specified time point 10 $\mu$ MTT reagent (stock of 5 mg/mL in PBS) was added to each well and the cells were further incubated for 4 hrs.
+At the specified time point 10 $\mu$L MTT reagent (stock of 5 mg/mL in PBS) was added to each well and the cells were further incubated for 4 hrs.
 At 4 hrs, the media was removed and 50 $\mu$L of dissolving reagent DMSO was added to each well.
@@
 The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Cancer Institute or the National Institutes of Health.
-This study was also supported by the National Science Foundation GK12 award number DGE-1045513.
+The authors acknowledge the financial support of the National Science Foundation, award no.
+DGE-1045513,for a GK12 training grant entitled "Grass Roots: Advancing Education in Renewable Energy and Cleaner Fuels through collaborative graduate fellow/teacher/grade-school student interactions."
 
@@
 
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure7_ExperimentalResults.pdf}
+\caption{Experimental results for multiple drug combinations on two prostate cancer cell lines, LNCaP and C4-2.
+A.
+Western blot analysis of AR, pS6, AKT, pAKT and ERK in LNCaP and C4-2 cell lines treated for 24 hrs with DMSO (control), sorafenib (5 $\mu$M), LY294002 (40 $\mu$M), and MDV3100 (10 $\mu$M) alone or in combination (at least 3 repeats).
+B.
+Cells (LNCaP and C4-2) were treated for 24, 48 and 72 hrs with sorafenib (5 $\mu$M), LY294002 (40 $\mu$M), and MDV3100 (10 $\mu$M) and cell viability was measured using MTT Assay.
+Values were normalized to DMSO (control).
+C.
+Cell viability results for LNCaP and C4-2 cells at varying concentration of sorafenib, LY294002, and MDV3100 after 24 hrs of treatment.
+Values were normalized to DMSO (control).
+Error bars represent standard error (at least 3 repeats with triplicates performed in each experiment).}
+\label{fg:Experiment}
+\end{figure}
+
+\clearpage
+
 % Supplemental figures -
@@
 \begin{figure}\centering
-%\includegraphics[width = 7in]{Supp_Figure_Sensitivity.pdf}
+\includegraphics[width=1.0\textwidth]{./figs/Fig_6_PI3K_AR_KO_3_Cases_old_new.pdf}
+\caption{Robustness analysis of a population of CR prostate models with PI3K (blue), AR (green), and PI3K and AR (orange) knock-outs (N = 500).
+A log fold change of greater then zero implies that the concentration of the protein increased with the knock-out, while a log fold change of less then zero indicates that the concentration of protein decreased.
+A log of fold change equal to 0, shows no response due to the knock-out.
+ A.,B.,C. Log robustness of cyclin E, PSA, and p70p versus Cyclin D for the three knock-out cases. A CR LNCaP cell was assumed for all knock-out cases.}
+\label{fg:Supp_Dual_Knockout}
+\end{figure}
+
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Supp_Figure_Sensitivity.pdf}
 \caption{Sensitivity analysis of a population of prostate models (N = 500).
@@
 $\bf{A}$ Sensitivity ranking of network species in CR cells in the absence and presence of DHT.
-$\bf{B.}$ Sensitivity ranking of network species in CR cells in the presence and absence of enzalutamide in the absence of a DHT stimulus.
+$\bf{B.}$ Sensitivity ranking of network species in CR cells in the presence of enzalutamide in the presence and absence of a DHT stimulus.
 $\bf{C.,D.}$ Sensitivity ranking of network species in CR cells in the presence and absence of sorafenib and lapatinib, respectively, with a DHT stimulus.}
@@
 \begin{figure}\centering
-%\includegraphics[width = 7in]{Supp_Figure_Robustness.pdf}
+\includegraphics[width=1.0\textwidth]{./figs/Supp_Figure_Robustness.pdf}
 \caption{Robustness analysis of protein markers.
@@
 
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Supp_Figure_KO_new.pdf}
+\caption{Dual knock-out of AR and PI3K leads to decreased expression of activated p70.
+A., B, C.
+Robustness coefficient of activated p70 (S6) in the PI3K knock-out, AR knock-out, and dual knock-out cases, respectively.
+The control was the basil CR LNCaP wild type case.
+Error bars denote plus and minus one standard error above the mean with N = 500.
+Experimental data is from Carver, et al \cite{Carver2011}.
+}
+\label{fg:Supp_KO}
+\end{figure}
+
 %\end{comment}

### 2015-06-30 Paper_v1.tex
Working on the introduction

@@
 \begin{titlepage}
-{\par\centering\textbf{\Large {Dynamically dimensioned particle swarm optimization (DdPSO): A meta-heuristic for parameter estimation in large biochemical models.}}}
+{\par\centering\textbf{\Large {Dynamically dimensioned particle swarm optimization (DdPSO): A meta-heuristic for parameter estimation in biochemical models}}}
 \vspace{0.05in}
-{\par \centering \large{Adithya Sagar, Christine Shoemaker and Jeffrey D. Varner$^{*}$}}
+{\par \centering \large{Adithya Sagar, Christine Shoemaker$^{\dag}$ and Jeffrey D. Varner$^{*}$}}
 \vspace{0.10in}
-{\par \centering \large{School of Chemical and Biomolecular Engineering}}
-{\par \centering \large{Cornell University, Ithaca NY 14853}}
+{\par \centering {School of Chemical and Biomolecular Engineering}}
+{\par \centering {$^{\dag}$School of Civil and Environmental Engineering}}
+{\par \centering {Cornell University, Ithaca NY 14853}}
 \vspace{0.1in}
@@
 \vspace{0.1in}
-{\par \centering \textbf{To be submitted:}~\emph{Biotechnology Journal}}
+{\par \centering \textbf{To be submitted:}~\emph{PLoS ONE}}
 \vspace{0.5in}
@@
 
-Biological systems interact with one another and with the external world through highly complex biochemical networks.
-The advent of high throughout technologies has provided researchers with a deluge of data about these networks.
-Over the last decade, with an increase in computational power, large scale mathematical modeling has evolved as a powerful paradigm to characterize these complex biochemical interactions \cite{assmus2006dynamics, Riel:2006aa,Jaqaman:2006aa,kitano2002systems,hood2004systems}.
-Mathematical modeling of these networks is an iterative process that primarily consists of two tasks a) Model construction b) Model calibration or Parameter Estimation \cite{Aldridge:2006aa,banga2008optimization,ashyraliyev2009systems}.
-Model construction involves translating existing knowledge about the network (i.e.
-entities in the network, interactions amongst the entities, types of interactions) along with reasonable assumptions into an appropriate mathematical framework.
-Model calibration or parameter estimation follows the model construction step.
-In this step the model is trained against experimental data to determine the model parameters that give a good 'fit' against experimental data.
-Model training is generally framed as a non-linear optimization problem wherein the residual (or objective function) between experimental data and simulation data is minimized using various optimization techniques \cite{moles2003parameter}.
-The optimal parameters obtained from the previous step are then used to validate the model on hitherto unseen experimental data.
-If validation leads to unsatisfactory results, model construction and calibration are repeated iteratively till satisfactory results are obtained.
+Cells process nutrients and respond to changes in their environment using complex biochemical networks. These networks contain thousands of components
+interconnected through nonlinear enzyme catalyzed reactions.
+Mathematical modeling has evolved as a powerful paradigm to characterize, and ultimately design these complex networks \cite{assmus2006dynamics, Riel:2006aa,Jaqaman:2006aa,kitano2002systems,hood2004systems}.
+Mathematical modeling of biochemical networks is often an iterative process.
+First, models are formulated from biochemical knowledge, and then trained using using experimental data \cite{Aldridge:2006aa,banga2008optimization,ashyraliyev2009systems}.
+Model training is typically framed as a non-linear optimization problem wherein the residual (or objective function) between experimental data and model simulations is minimized using an optimization strategy \cite{moles2003parameter}.
+The optimal parameters obtained from training are then used to validate the model on hitherto unseen experimental data.
+If validation fails, model construction and calibration are repeated iteratively until satisfactory results are obtained.
 
-With the capacity to construct very large models using various mathematical formalisms \cite{chen2009input,tasseff2011modeling,luan2007computationally,mo2007genome,orth2011comprehensive,karr2012whole,buchel2013path2models,smallbone2010towards}, more often than not, parameter estimation is becoming a major challenge if not a bottleneck in large scale mathematical modeling.
-Though parameter estimation has been a well studied problem in engineering \cite{nieman1971review,beck1977parameter,young1981parameter,beck1998inverse}, the complex dynamics of large biological systems and noisy, often incomplete experimental data pose a more difficult and rather unique challenge.
+Parameter estimation is a major challenge in the development of large biochemical models.
+Though parameter estimation has been a well studied problem in engineering for decades \cite{nieman1971review,beck1977parameter,young1981parameter,beck1998inverse},
+the complex dynamics of large biological systems and noisy, often incomplete experimental data pose a unique estimation challenge.
 Most of the optimization problems pertaining to biological systems are non-linear and multi-modal i.e.
@@
 Fan et al.
-recently showed that population based meta-heuristics along with decomposition based methods can be used to model gene circuits from mRNA data \cite{fan2015parameter}.
-
-Despite these successes, a  major drawback with most of these approaches is the vast number of objective function evaluations they take in arriving at good search regions.
+recently showed that population based meta-heuristics along with decomposition based methods can be used to model gene circuits from mRNA data \cite{fan2015parameter}.Despite these successes, a  major drawback with most of these approaches is the vast number of objective function evaluations they take in arriving at good search regions.
 As the models grow in size and complexity and the number of dimensions of parameter vector increases,  parameter estimation becomes a computationally expensive task.
@@
 Tolson and Shoemaker \cite{tolson2007dynamically} showed through Dynamically Dimensioned Search (DDS) that high-dimensional watershed models can be calibrated quickly by perturbing only a subset of dimensions.
-Thus in our current study we took into cognizance the fact that it may not be necessary to obtain the exact solution for high dimensional biological systems and that good enough solutions can be quickly obtained without expending a lot of objective function evaluations.
+
+In this study, we developed the dynamically dimensioned particle swarm optimization (DdPSO) approach, a novel meta-heurstic that combined the global search capability of multi-swarm particle swarm optimization with dynamic dimensioned search.
+The design objective of DdPSO was to obtain near optimal parameter estimates for large biochemical models within a relatively few function evaluations.
+DdPSO uses a multi-swarm particle swarm optimization technique to generate candidate solution vectors which are then greedily updated using dynamic dimensioned search.
+We tested the performance of DdPSO on commonly used test functions for global optimization (Ackley and Rosenbrock), published biochemical parameter estimation benchmark problems and
+lastly on a model of the human coagulation cascade.
+DdPSO outperformed common meta-heuristic approaches like Differential Evolution (DE), Particle Swarm Optimization (PSO), Simulated Annealing (SA) and dynamically dimensioned search (DDS) on the benchmark and coagulation problems.
+
+% Extra text from the introduction -
+%With the capacity to construct very large models using various mathematical formalisms \cite{chen2009input,tasseff2011modeling,luan2007computationally,mo2007genome,orth2011comprehensive,karr2012whole,buchel2013path2models,smallbone2010towards}, more often than not,
+
+%this algorithm (25 trials with 4000 function evaluations in each trial) on a biochemical network of coagulation (148 parameters and 92 species) and compared it's performance against other meta-heuristics  The new algorithm outperforms all the other meta-heuristics on the coagulation model.
+The parameter vectors obtained using this approach fit the experimental data well and also make accurate enough predictions on unseen experimental data.
+We also performed this comparison on  and found the same behavior.
+Further we used two recently published benchmark problems, a genome wide kinetic model with 1759 parameters and a metabolic model of Chinese Hamster Ovary cells with 117 parameters to evaluate the performance of our approach.
+We  surprisingly performed well on these benchmarks and obtained the nominal parameter vector with just 4000 function evaluations in both cases.
+%took into cognizance the fact that it may not be necessary to obtain the exact solution for high dimensional biological systems and that good enough solutions can be quickly obtained without expending a lot of objective function evaluations.
 Hence our current approach uses the power of population based heuristics along with DDS to obtain near optimal or good solutions.

### 2015-07-01 Paper_v1.tex
Updated the abstract

@@
 \section*{Abstract}
-The problem of parameter estimation presents a significant challenge in the modeling of large complex systems.
-Mathematical modeling of biological systems with multiple feed back loops is one such area where parameter estimation is a difficult non-linear optimization problem.
+Mathematical modeling is as a powerful tool to analyze, and ultimately design biochemical networks.
+However, estimation of biochemical model parameters can be a significant challenge.
+Parameter estimation in biochemical models generally involves expensive function evaluations, making it difficult to quickly obtain optimal solutions.
+Additionally, biochemical models often have many local minima or maxima which further complicates the parameter estimation problem.
+In this study, we developed dynamically dimensioned particle swarm optimization (DdPSO), a novel meta-heurstic that combined the global search capability of multi-swarm particle swarm optimization with dynamic dimensioned search.
+DdPSO uses a multi-swarm particle swarm optimization technique to generate candidate solution vectors which are then greedily updated using dynamic dimensioned search.
+We tested the performance of DdPSO on commonly used test functions for global optimization (Ackley and Rosenbrock), published biochemical parameter estimation benchmark problems and
+lastly on a model of the human coagulation cascade. For each test we performed 25 trials with 4000 function evaluations per trial, and compared the performance of DdPSO with other commonly
+used meta-heuristic approaches such as differential evolution (DE), particle swarm optimization (PSO), simulated annealing (SA) and dynamically dimensioned search (DDS).
+For the wide range of problems that we considered, DdPSO outperformed the other meta-heuristic approaches despite a limited number of function evaluations.
+We further tested the predictive power of the coagulation model parameters against data not used in training, and found good agreement between model simulations and experimental measurements.
+Taken together, DdPSO is a promising meta-heuristic approach for the estimation of biochemical model parameters in relatively few function evaluations.
+
+{\noindent \textbf{Keywords:}~Parameter identification, Meta-heuristic optimization, Biochemical modeling}
+
+% Extra abstract
+% Mathematical modeling of biological systems with multiple feed back loops is one such area where parameter estimation is a difficult non-linear optimization problem.
 This difficulty is further compounded when dealing with parameter vectors of high dimensions.
-Such problems generally involve expensive function evaluations and it is hard to obtain optimal solutions within finite computational time.
-In this study we present a novel meta-heuristic approach that combines a variant of particle swarm optimization (PSO) along with dynamically dimensioned search (DDS) to obtain near optimal solutions of high dimensional biochemical networks within a relatively few function evaluations.
-We use a particle swarm optimization technique that uses multi-swarms to generate candidate vectors which are then greedily updated using DDS by dynamically varying the perturbed parameter dimensions.
+
+%In this study, we present the dynamically dimensioned particle
+%a novel meta-heuristic approach that combines a variant of particle swarm optimization (PSO) along with dynamically dimensioned search (DDS) to obtain near optimal solutions of high dimensional biochemical networks within a relatively few function evaluations.
+%We use a particle swarm optimization technique that uses multi-swarms to generate candidate vectors which are then greedily updated using DDS by dynamically varying the perturbed parameter dimensions.
 We tested this algorithm (25 trials with 4000 function evaluations in each trial) on a biochemical network of coagulation (148 parameters and 92 species) and compared it's performance against other meta-heuristics like Differential Evolution (DE), Particle Swarm Optimization (PSO), Simulated Annealing (SA) and also against DDS alone.
@@
 We  surprisingly performed well on these benchmarks and obtained the nominal parameter vector with just 4000 function evaluations in both cases.
-
-{\noindent \textbf{Keywords:}~Parameter identification, Mathematical modeling}
 
@@
 interconnected through nonlinear enzyme catalyzed reactions.
-Mathematical modeling has evolved as a powerful paradigm to characterize, and ultimately design these complex networks \cite{assmus2006dynamics, Riel:2006aa,Jaqaman:2006aa,kitano2002systems,hood2004systems}.
+Mathematical modeling has evolved as a powerful paradigm to analyze, and ultimately design these complex networks \cite{assmus2006dynamics, Riel:2006aa,Jaqaman:2006aa,kitano2002systems,hood2004systems}.
 Mathematical modeling of biochemical networks is often an iterative process.
@@
 Parameter estimation is a major challenge in the development of large biochemical models.
-Though parameter estimation has been a well studied problem in engineering for decades \cite{nieman1971review,beck1977parameter,young1981parameter,beck1998inverse},
+Although parameter estimation has been a well studied problem in engineering for decades \cite{nieman1971review,beck1977parameter,young1981parameter,beck1998inverse},
 the complex dynamics of large biological systems and noisy, often incomplete experimental data pose a unique estimation challenge.
@@
 
-In this study, we developed the dynamically dimensioned particle swarm optimization (DdPSO) approach, a novel meta-heurstic that combined the global search capability of multi-swarm particle swarm optimization with dynamic dimensioned search.
-The design objective of DdPSO was to obtain near optimal parameter estimates for large biochemical models within a relatively few function evaluations.
+In this study, we developed dynamically dimensioned particle swarm optimization (DdPSO), a novel meta-heurstic that combined the global search capability of multi-swarm particle swarm optimization with dynamic dimensioned search.
+The objective of DdPSO was to obtain near optimal parameter estimates for large biochemical models within a relatively few function evaluations.
 DdPSO uses a multi-swarm particle swarm optimization technique to generate candidate solution vectors which are then greedily updated using dynamic dimensioned search.

### 2015-07-10 Paper_v2.tex
Updates
Latest Updates

@@
 \vspace{0.05in}
-{\par \centering \large{ Katharine V. Rogers, Joseph A. Wayman, Ryan Tasseff, Caitlin Gee, Timothy Chu, Kyriakos Tsahalis, Spencer Davis, Rohit Yallamraju, Gongcheng Lu, Matthew P. DeLisa, 
+{\par \centering \large{ Katharine V. Rogers, Joseph A. Wayman, Ryan Tasseff, Caitlin Gee, Matthew P. DeLisa, 
 and Jeffrey D. Varner$^{*}$}}
@@
 POETs sampled well over a million possible parameter sets, from which we selected N = 5000 sets for further analysis. 
-Over the ensemble, the coefficient of variation (CV) of the kinetic parameters spanned 0.5 - 5.8, with 33\% of the parameters having a CV of less than one (Fig. ZZ). 
+Over the ensemble, the coefficient of variation (CV) of the kinetic parameters spanned 0.59 - 5.8, with 33\% of the parameters having a CV of less than one (Fig. \ref{fg:Supp_CV}). 
 As a control, we also performed simulations for R = 100 random parameter sets to compare against the ensemble generated by POETs.
@@
 We also considered the sensitivity of CR cells following the application of the AR inhibitor enzalutamide in the presence of DHT (Fig. \ref{fg:Sensitivity}D). 
-In an androgen environment with enzalutamide, species which were more fragile included cytosolic AR, cPAcP, and p21Cip1. 
+Species which were more sensitive in an androgen environment with enzalutamide included cytosolic AR, cPAcP, and p21Cip1. 
 As we would expect, AR species found in the nucleus and/or bound to coactivators, were more robust in the absence of enzalutamide. 
@@
 \ref{fg:Experiment}B).
-I
-n both cell lines, MDV3100 (10 $\mu$M), has only a modest effect on cell viability versus control (DMSO). 
-(Fig. \ref{fg:Experiment}C) shows cell viability of both cell lines in varying concentrations of inhibitors at 24 hrs.
+In both cell lines, MDV3100 (10 $\mu$M), has only a modest effect on cell viability versus control (DMSO). 
+Figure \ref{fg:Experiment}C shows cell viability of both cell lines in varying concentrations of inhibitors at 24 hrs.
 
@@
 We calculated the effect of these perturbations on different protein markers, such as PSA, AR, and cyclin D. 
-The results of the model suggest that an inhibition of either the PI3K pathway or the MAPK pathway in combination with an AR inhibitor as a possible therapy for CRPC. 
-Sensitivity analysis revealed no change in the top sensitive species in the presence or absence of the AR inhibitor, enzalutamide.
-PI3K/AKT and MAPK species continued to fall in the top two percent of sensitive species.
-A study by Carver \textit{et al.} looked at dual inhibition of AR and PI3K signaling in LNCaP cells and in a Pten-deficient murine prostate cancer model \cite{Parker2013}. 
-Using both the PI3K inhibitor, BEZ235, and the AR inhibitor, MDV3100 (enzalutamide), the group saw a drastic decrease in the total number of cells. 
-Each inhibitor on it’s own had a much smaller effect on total cell number. 
-They saw an increase in the cell death marker, c-PARP, in the dual inhibition case. 
-The group hypothesized that AKT inhibition leads to increased AR signaling activity through increased protein concentrations of HER3. 
-On the other hand, AR inhibition leads to increased AKT activity due to the down regulation of PHLPP, a protein phosphatase that regulates AKT. 
-For simplicity, the HER3 pathway and also cell death were not included in the model. 
-Dual knock-out studies of PI3K and AR in our model show no additive effect on any cell cycle proteins through the dual knock-out. 
-The pathways in our model appear to be uncoupled and therefore no synergy is shown in the dual knock-out case. 
-This could indicate that the combined decrease in cell population is entirely due to cell death. 
-The Carver \textit{et al.} study did not consider cell cycle proteins or cell growth. 
-Our model does show a decrease in cell cycle proteins in the PI3K knock-out as well as in the dual knock-out case. 
-This result seems to be consistent with the decreased cell growth in the PI3K knock-out case which is not dependent on cell death, as c-PARP levels are low. 
-The decrease in cell cycle proteins is due to a decrease in general translation, including free eIF4E levels and activated 40S ribosome subunit. 
-The decrease in p70 (S6) activation due to inhibition of PI3K is shown in both the model and in the Carver \textit{et al.} study (Supplementary Figure), indicating this result is due to the PI3K pathway.
-It would be interesting to repeat the experiment looking at cell cycle proteins and also performing the experiment in CR LNCaP cells, instead of AD cells.
-
 A knock-out of Raf, MEK or ERK showed an overall increase in cyclin D levels in CR cells (Fig. \ref{fg:Supp_Robustness}). 
@@
 These outlying cell types could be critical for understanding when designing drug targets and combination therapies.
+
+%The results of the model suggest that an inhibition of either the PI3K pathway or the MAPK pathway in combination with an AR inhibitor as a possible therapy for CRPC. 
+%Sensitivity analysis revealed no change in the top sensitive species in the presence or absence of the AR inhibitor, enzalutamide.
+PI3K/AKT and MAPK species continued to fall in the top two percent of sensitive species.
+%A study by Carver \textit{et al.} looked at dual inhibition of AR and PI3K signaling in LNCaP cells and in a Pten-deficient murine prostate cancer model \cite{Parker2013}. 
+%Using both the PI3K inhibitor, BEZ235, and the AR inhibitor, MDV3100 (enzalutamide), the group saw a drastic decrease in the total number of cells. 
+%Each inhibitor on it’s own had a much smaller effect on total cell number. 
+%They saw an increase in the cell death marker, c-PARP, in the dual inhibition case. 
+%The group hypothesized that AKT inhibition leads to increased AR signaling activity through increased protein concentrations of HER3. 
+%On the other hand, AR inhibition leads to increased AKT activity due to the down regulation of PHLPP, a protein phosphatase that regulates AKT. 
+%For simplicity, the HER3 pathway and also cell death were not included in the model. 
+%Dual knock-out studies of PI3K and AR in our model show no additive effect on any cell cycle proteins through the dual knock-out. 
+%The pathways in our model appear to be uncoupled and therefore no synergy is shown in the dual knock-out case. 
+%This could indicate that the combined decrease in cell population is entirely due to cell death. 
+%The Carver \textit{et al.} study did not consider cell cycle proteins or cell growth. 
+%Our model does show a decrease in cell cycle proteins in the PI3K knock-out as well as in the dual knock-out case. 
+%This result seems to be consistent with the decreased cell growth in the PI3K knock-out case which is not dependent on cell death, as c-PARP levels are low. 
+%The decrease in cell cycle proteins is due to a decrease in general translation, including free eIF4E levels and activated 40S ribosome subunit. 
+%The decrease in p70 (S6) activation due to inhibition of PI3K is shown in both the model and in the Carver \textit{et al.} study (Supplementary Figure), indicating this result is due to the PI3K pathway.
+%It would be interesting to repeat the experiment looking at cell cycle proteins and also performing the experiment in CR LNCaP cells, instead of AD cells.
 
@@
 The decrease in cell cycle proteins is due to a decrease in general translation, including free eIF4E levels and activated 40S ribosome subunit. 
-The decrease in p70 (S6) activation due to inhibition of PI3K is shown in both the model and in the Carver \textit{et al.} study (Supplementary Figure), indicating this result is due to the PI3K pathway.
+The decrease in p70 (S6) activation due to inhibition of PI3K is shown in both the model and in the Carver \textit{et al.} study ((Fig.
+\ref{fg:Supp_Figure_KO_new.pdf})), indicating this result is due to the PI3K pathway.
 Our results confirm the Carver \textit{et al.} study in that a dual inhibition of AR and PI3K signaling led to a more prominent decrease in cell viability then each of the inhibitors alone. 
@@
 DGE-1045513,for a GK12 training grant entitled "Grass Roots: Advancing Education in Renewable Energy and Cleaner Fuels through collaborative graduate fellow/teacher/grade-school student interactions."
+We would like to thank Timothy Chu, Kyriakos Tsahalis, Spencer Davis, Rohit Yallamraju, Gongcheng Lu, and Yalin Zhao.
 
@@
 \begin{figure}\centering
-\includegraphics[width=1.0\textwidth]{./figs/Fig_6_PI3K_AR_KO_3_Cases_old_new.pdf}
+\includegraphics[width=1.0\textwidth]{./figs/Supplementary_CV_Subset_Total}
+\caption{Coefficient of variation (CV) of model parameters estimated using POETs.
+The solid line denotes the mean CV calculated over the entire ensemble (N = 5000).
+The points denote the mean CV of the 500 ensemble members used for sensitivity and robustness calculations.
+Over the ensemble, the coefficient of variation (CV) of the kinetic parameters spanned 0.59 - 5.8, with 33\% of the parameters having a CV of less than one.}
+\label{fg:Supp_CV}
+\end{figure}
+
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Supp_Figure_NewModelvsOldModel}
+\caption{Blind model predictions for the ensemble with the original and updated model (EGFR and HER2 heterodimer).
+A,B.
+Time course data for HER phosphorylation due to a stimulus of 1.6 nM EGF (LNCaP C33, P7) for the old and new model, respectively.
+Dark grey shows only parameters improved by the updated model (N=2870) while light grey show all parameter sets (N=5000).
+C,D.
+Shc phosphorylation levels at 16 hrs in the presence and absence of 1.6 nM EGF (LNCaP C33, P29) for the old and new model, respectively.
+Light grey denotes experimental data, mid grey denotes simulation results for all parameters (N=5000), and black denotes only parameters improved by the updated model (N=3142).}
+\label{fg:Exps_NewModelvsOldModel}
+\end{figure}
+
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Fig_6_PI3K_AR_KO_3_Cases.pdf}
 \caption{Robustness analysis of a population of CR prostate models with PI3K (blue), AR (green), and PI3K and AR (orange) knock-outs (N = 500).
@@
 Simulations shown were from CR cells, with indicated perturbation.
-Mean of 500 ensemble members is shown.
-}
+Mean of 500 ensemble members is shown.}
 \label{fg:Supp_Robustness}

### 2015-07-15 Paper_v2.tex
New Updates

@@
 For the simplicity of this model, the HER3 pathway and also cell death were not included in the model. 
-Dual knock-out studies of PI3K and AR in our model show no additive effect on any cell cycle proteins through the dual knock-out. 
-The pathways in our model appear to be uncoupled and therefore no synergy is shown in the dual knock-out case. 
-This could indicate that the combined decrease in cell population is entirely due to cell death. 
+Dual knock-out studies of PI3K and AR in our model show only a slight additive effect on cell cycle proteins through the dual knock-out (Fig. \ref{fg:Dual_Knockout}).
+The pathways in our model appear to be uncoupled and therefore little synergy is shown in the dual knock-out case. 
+This could indicate that the combined decrease in cell population due to the dual inhibition of PI3K and AR is entirely due to cell death. 
 The Carver \textit{et al.} study did not consider cell cycle proteins or cell growth. 
@@
 This result seems to be consistent with the decreased cell growth in the PI3K knock-out case which is not dependent on cell death, as c-PARP levels are low. 
-The decrease in cell cycle proteins is due to a decrease in general translation, including free eIF4E levels and activated 40S ribosome subunit. 
-The decrease in p70 (S6) activation due to inhibition of PI3K is shown in both the model and in the Carver \textit{et al.} study ((Fig.
-\ref{fg:Supp_Figure_KO_new.pdf})), indicating this result is due to the PI3K pathway.
-Our results confirm the Carver \textit{et al.} study in that a dual inhibition of AR and PI3K signaling led to a more prominent decrease in cell viability then each of the inhibitors alone. 
-We extended the study to look at the addition of a third inhibitor, sorafenib, that inhibits Raf kinase in the MAPK pathway.
+The decrease in cell cycle proteins in the model is due to a decrease in general translation, including free eIF4E levels and activated 40S ribosome subunit. 
+The decrease in p70 (S6) activation due to inhibition of PI3K is shown in both the model and in the Carver \textit{et al.} study , indicating this result is due to the PI3K pathway (Fig.
+\ref{fg:Supp_KO}).
+Our experimental results confirm the Carver \textit{et al.} study in that a dual inhibition of AR and PI3K signaling led to a more prominent decrease in cell viability then each of the inhibitors alone in LNCaP cells.
+We extended the study to look at the addition of a third inhibitor, sorafenib, that inhibits Raf kinase in the MAPK pathway and an additional cell line, C4-2, which was CR.
+In both cell lines, there was not a significant decrease in cell viability between the three dual inhibitor cases and the triple inhibition case at 74 hours.
 
@@
 Androgen dependent LNCaP prostate cancer cells were a gift from Dr.
-Brian Kirby (Cornell University), and the Castrate Resistant C4-2 prostate cancer cell line was purchased from MD Anderson Cancer Center, University of Texas.
+Brian Kirby (Cornell University), and the castration resistant C4-2 prostate cancer cell line was purchased from MD Anderson Cancer Center, University of Texas.
 Cell lines were maintained in RPMI 1640 media (Life Technologies, Inc., Grand Island, NY) with 10\% fetal calf serum (FBS; Hyclone) and 1x antibiotic/antimycotic (Sigma, St.
@@
 
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Fig_8_PI3K_AR_KO_3_Cases.pdf}
+\caption{Robustness analysis of a population of CR prostate models with PI3K (blue), AR (green), and PI3K and AR (orange) knock-outs (N = 500).
+A log fold change of greater then zero implies that the concentration of the protein increased with the knock-out, while a log fold change of less then zero indicates that the concentration of protein decreased.
+A log of fold change equal to 0, shows no response due to the knock-out.
+ A.,B.,C. Log robustness of cyclin E, PSA, and p70p versus Cyclin D for the three knock-out cases. A CR LNCaP cell was assumed for all knock-out cases.}
+\label{fg:Dual_Knockout}
+\end{figure}
+
+\clearpage
+
 % Supplemental figures -
@@
 \begin{figure}\centering
-\includegraphics[width=1.0\textwidth]{./figs/Fig_6_PI3K_AR_KO_3_Cases.pdf}
-\caption{Robustness analysis of a population of CR prostate models with PI3K (blue), AR (green), and PI3K and AR (orange) knock-outs (N = 500).
-A log fold change of greater then zero implies that the concentration of the protein increased with the knock-out, while a log fold change of less then zero indicates that the concentration of protein decreased.
-A log of fold change equal to 0, shows no response due to the knock-out.
- A.,B.,C. Log robustness of cyclin E, PSA, and p70p versus Cyclin D for the three knock-out cases. A CR LNCaP cell was assumed for all knock-out cases.}
-\label{fg:Supp_Dual_Knockout}
-\end{figure}
-
-\begin{figure}\centering
 \includegraphics[width=1.0\textwidth]{./figs/Supp_Figure_Sensitivity.pdf}

### 2015-07-22 Paper_v2.tex
New Updates

@@
 This was a reasonable first approximation, as the affinity of HER2/EGFR heterodimerization and EGRF homodimerization is thought to be similar \cite{Hendriks2003}. 
-With the inclusion of HER2-EGFR heterodimerization, we qualitatively fit the EGF-induced HER2 activation case and improved our training for experiments that involved an EGF stimulus, e.g., cyclin D mRNA and protein abundance following an EGF stimulus in C-33 cells (Fig \ref{fg:ColorPlot}A and C, white pixels).
+With the inclusion of HER2-EGFR heterodimerization, we qualitatively fit the EGF-induced HER2 activation case and improved our training for experiments that involved an EGF stimulus, e.g., cyclin D mRNA and protein abundance following an EGF stimulus in C-33 cells (Fig.
+\ref{fg:ColorPlot}A and C, white pixels and Fig.
+\ref{fg:Exps_NewModelvsOldModel}).
 
@@
 Other species such as E2F, cyclin E, and DHT-activated AR were also more sensitive in AD cells. 
-On the other hand, HER2-Grb2-Gab activation of PI3K and AKT inhibition of RAF were more sensitive in CR cells.
+On the other hand, HER2-Grb2-Gab activation of PI3K and AKT inhibition of Raf were more sensitive in CR cells.
 
@@
 For the simplicity of this model, the HER3 pathway and also cell death were not included in the model. 
-Dual knock-out studies of PI3K and AR in our model show only a slight additive effect on cell cycle proteins through the dual knock-out (Fig. \ref{fg:Dual_Knockout}).
-The pathways in our model appear to be uncoupled and therefore little synergy is shown in the dual knock-out case. 
+Dual knock-out studies of PI3K and AR in our model show only a slight additive effect on cell cycle proteins through the dual knock-out compared to solely inhibiting PI3K (Fig.
+\ref{fg:Dual_Knockout}).
+The triple case, with PI3K, AR, and RAF knock-outs, also only showed a slight additive effect (Fig. \ref{fg:Triple_KO}).
+%The pathways in our model appear to be uncoupled and therefore little synergy is shown in the dual knock-out case. 
 This could indicate that the combined decrease in cell population due to the dual inhibition of PI3K and AR is entirely due to cell death. 
 The Carver \textit{et al.} study did not consider cell cycle proteins or cell growth. 
-Our model does show a decrease in cell cycle proteins in the PI3K knock-out as well as in the dual knock-out case. 
-This result seems to be consistent with the decreased cell growth in the PI3K knock-out case which is not dependent on cell death, as c-PARP levels are low. 
+Our model does show a decrease in cell cycle proteins in the PI3K knock-out as well as in the PI3K and AR dual knock-out case (Fig. \ref{fg:Dual_Knockout}). 
+This result seems to be consistent with the decreased cell count in the PI3K knock-out case which is not dependent on cell death, as c-PARP levels are low. 
 The decrease in cell cycle proteins in the model is due to a decrease in general translation, including free eIF4E levels and activated 40S ribosome subunit. 
@@
 We extended the study to look at the addition of a third inhibitor, sorafenib, that inhibits Raf kinase in the MAPK pathway and an additional cell line, C4-2, which was CR.
-In both cell lines, there was not a significant decrease in cell viability between the three dual inhibitor cases and the triple inhibition case at 74 hours.
+In both cell lines, there was not a significant decrease in cell viability between the three dual inhibitor cases and the triple inhibition case at 74 hours, indicating that a dual inhibition (PI3K/AR, AR/MAPK, or PI3K/MAPK) may be a sufficient treatment.
 
@@
 I.
-AR mRNA levels in the presence and absence of E2F over expression (LNCaP C33, O34).}
+AR mRNA levels in the presence and absence of E2F over expression (LNCaP C33, O34).
+Error bars denote plus and minus one standard deviation above the mean}
 \label{fg:Training}
@@
 A.
-Time course data for HER phosphorylation due to a stimulus of 10 nM DHT (LNCaP C33, P1).
+Time course data for HER2 phosphorylation due to a stimulus of 10 nM DHT (LNCaP C33, P1).
 B.
@@
 E.
-Time course data for HER phosphorylation due to a stimulus of 1.6 nM EGF (LNCaP C33, P7).
+Time course data for HER2 phosphorylation due to a stimulus of 1.6 nM EGF (LNCaP C33, P7).
 F.
@@
 G, H, I.
-Fold change in PSA concentration due to drug stimulus: enzalutamide, lapatinib, and sorafenib.}
+Fold change in PSA concentration due to drug stimulus: enzalutamide, lapatinib, and sorafenib.
+Error bars denote plus and minus one standard deviation above the mean.}
 \label{fg:Validation}
@@
 $\bf{C.}$ Sensitivity ranking of network species in AD cells in the absence and presence of DHT.
-$\bf{D.}$ Sensitivity ranking of network species in CR cells in the presence and absence of enzalutamide with a DHT stimulus.}
+$\bf{D.}$ Sensitivity ranking of network species in CR cells in the presence and absence of enzalutamide with a DHT stimulus.
+Error bars denote standard error with N = 500}
 \label{fg:Sensitivity}
@@
 
+\clearpage
+
 \begin{figure}\centering
@@
 A,B.
-Time course data for HER phosphorylation due to a stimulus of 1.6 nM EGF (LNCaP C33, P7) for the old and new model, respectively.
+Time course data for HER2 phosphorylation due to a stimulus of 1.6 nM EGF (LNCaP C33, P7) for the old and new model, respectively.
 Dark grey shows only parameters improved by the updated model (N=2870) while light grey show all parameter sets (N=5000).
@@
 Shc phosphorylation levels at 16 hrs in the presence and absence of 1.6 nM EGF (LNCaP C33, P29) for the old and new model, respectively.
-Light grey denotes experimental data, mid grey denotes simulation results for all parameters (N=5000), and black denotes only parameters improved by the updated model (N=3142).}
+Light grey denotes experimental data, mid grey denotes simulation results for all parameters (N=5000), and black denotes only parameters improved by the updated model (N=3142).
+Error bars denote plus and minus one standard deviation above the mean.}
 \label{fg:Exps_NewModelvsOldModel}
 \end{figure}
+
+\clearpage
 
@@
 $\bf{B.}$ Sensitivity ranking of network species in CR cells in the presence of enzalutamide in the presence and absence of a DHT stimulus.
-$\bf{C.,D.}$ Sensitivity ranking of network species in CR cells in the presence and absence of sorafenib and lapatinib, respectively, with a DHT stimulus.}
+$\bf{C.,D.}$ Sensitivity ranking of network species in CR cells in the presence and absence of sorafenib and lapatinib, respectively, with a DHT stimulus.
+Error bars denote standard error with N = 500.}
 \label{fg:Supp_Sensitivity}
 \end{figure}
+
+\clearpage
 
@@
 \end{figure}
+
+\clearpage
 
@@
 
+\clearpage
+
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Supp_Figure_TripleKOCase.pdf}
+\caption{Robustness analysis of a population of CR prostate models with PI3K.
+AR, and Raf (maroon) knock-outs (N = 500).
+A log fold change of greater then zero implies that the concentration of the protein increased with the knock-out, while a log fold change of less then zero indicates that the concentration of protein decreased.
+A log of fold change equal to 0, shows no response due to the knock-out.
+A.,B.,C.
+Log robustness of cyclin E, PSA, and p70p versus Cyclin D for the three knock-out cases.
+A CR LNCaP cell was assumed for all knock-out cases.}
+\label{fg:Triple_KO}
+\end{figure}
+
+\clearpage
+
 %\end{comment}

### 2015-08-31 Paper_v2.tex
Final Updates
Updated version of the paper-paper_v2

--- 2015-07-10 Paper_v2.tex
@@
 \usepackage{simplemargins}
+\usepackage{fullpage}
 \usepackage{comment}
 \usepackage{lineno}
+%\usepackage{chicago}
+\usepackage{textcomp}
 
+%\usepackage[space]{cite}
 \urlstyle{rm}
@@
 
-%\usepackage{geometry}
-\usepackage{fullpage}
-
-%\bibliographystyle{plain}
-\bibliographystyle{msb}
+%\bibliographystyle{chicago}
+\bibliographystyle{plos2009}
 
@@
 \begin{titlepage}
-{\par\centering\textbf{\Large Modeling and Analysis of Hormone and Mitogenic Signal Integration in Prostate Cancer}}
+{\par\centering\textbf{\Large {Dynamically dimensioned particle swarm optimization (DdPSO): A meta-heuristic for parameter estimation in biochemical models}}}
 \vspace{0.05in}
-{\par \centering \large{ Katharine V. Rogers, Joseph A. Wayman, Ryan Tasseff, Caitlin Gee, Matthew P. DeLisa, 
-and Jeffrey D. Varner$^{*}$}}
+{\par \centering \large{Adithya Sagar, Christine Shoemaker$^{\dag}$ and Jeffrey D. Varner$^{*}$}}
 \vspace{0.10in}
-{\par \centering \large{School of Chemical and Biomolecular Engineering}}
-{\par \centering \large{Cornell University, Ithaca NY 14853}}
+{\par \centering {School of Chemical and Biomolecular Engineering}}
+{\par \centering {$^{\dag}$School of Civil and Environmental Engineering}}
+{\par \centering {Cornell University, Ithaca NY 14853}}
 \vspace{0.1in}
-{\par \centering \textbf{Running Title:}~Modeling signal transduction in prostate cancer}
+{\par \centering \textbf{Running Title:}~Parameter estimation in biochemical models}
 \vspace{0.1in}
-{\par \centering \textbf{To be submitted:}~\emph{Molecular Systems Biology}}
+{\par \centering \textbf{To be submitted:}~\emph{PLoS ONE}}
 \vspace{0.5in}
@@
 \section*{Abstract}
-Prostate cancer is the most common cancer in men and the second leading cause of cancer related death in the United States. 
-Androgens, such as testosterone, are required for prostate cancer growth. 
-Androgen ablation in combination with radiation or chemotherapy remains the primary non-surgical treatment for androgen dependent prostate cancer. 
-However, androgen ablation typically fails to permanently arrest cancer progression, often resulting in castration resistant prostate cancer (CRPC). 
-CRPC is closely related to metastasis and decreased survival. 
-In this study, we developed and analyzed a population of mathematical models describing growth factor and hormone signal integration in androgen dependent, intermediate and resistant prostate cancer cells.
-The model describes the integration of two simultaneous extracellular signaling inputs, namely, androgen and growth factors into a G1/S cell cycle checkpoint decision. 
-Model parameters were identified from 43 studies in androgen dependent and resistant LNCaP cell lines. 
-The model was validated by comparing simulations with an additional 29 data sets from LNCaP cell lines that were not used during training. 
-Additionally, data from four drug trials was also used to evaluate the model’s performance.
-Sensitivity analysis, conducted over an ensemble of prostate signaling models, suggested that in an androgen free environment general translation and transcription was more sensitive in androgen dependent cells, while in androgen independent cells the PI3K and MAPK pathway species were more sensitive.
-In a constant DHT environment sensitive species were conserved between the cell lines. 
-These results suggest targeting the PI3K and MAPK pathways in addition to anti-androgen therapies as a treatment for CRPC.
+Mathematical modeling is as a powerful tool to analyze, and ultimately design biochemical networks.
+However, estimation of biochemical model parameters can be a significant challenge.
+Parameter estimation in biochemical models generally involves expensive function evaluations, making it difficult to quickly obtain optimal solutions.
+Additionally, biochemical models often have many local minima or maxima which further complicates the parameter estimation problem.
+In this study, we developed dynamically dimensioned particle swarm optimization (DdPSO), a novel meta-heurstic that combined the global search capability of multi-swarm particle swarm optimization with dynamically dimensioned search.
+DdPSO uses a multi-swarm particle swarm optimization technique to generate candidate solution vectors which are then greedily updated using dynamically dimensioned search.
+We tested the performance of DdPSO on commonly used test functions for global optimization (Ackley and Rosenbrock), on published biochemical parameter estimation benchmark problems and
+lastly on a model of the human coagulation cascade. For each test we performed 25 trials with 4000 function evaluations per trial, and compared the performance of DdPSO with other commonly
+used meta-heuristic approaches such as differential evolution (DE), particle swarm optimization (PSO), simulated annealing (SA) and dynamically dimensioned search (DDS).
+For the wide range of problems that we considered, DdPSO outperformed the other meta-heuristic approaches despite a limited number of function evaluations.
+We further tested the predictive power of the coagulation model parameters against data not used in training, and found good agreement between model simulations and experimental measurements.
+Taken together, DdPSO is a promising meta-heuristic approach for the estimation of biochemical model parameters in relatively few function evaluations.
 
-\vspace{0.5in}
-{\noindent \textbf{Keywords:}~Prostate cancer, signal transduction, mathematical modeling}
+{\noindent \textbf{Keywords:}~Parameter identification, Meta-heuristic optimization, Biochemical modeling}
+
+% Extra abstract
+% Mathematical modeling of biological systems with multiple feed back loops is one such area where parameter estimation is a difficult non-linear optimization problem.
+This difficulty is further compounded when dealing with parameter vectors of high dimensions.
+
+%In this study, we present the dynamically dimensioned particle
+%a novel meta-heuristic approach that combines a variant of particle swarm optimization (PSO) along with dynamically dimensioned search (DDS) to obtain near optimal solutions of high dimensional biochemical networks within a relatively few function evaluations.
+%We use a particle swarm optimization technique that uses multi-swarms to generate candidate vectors which are then greedily updated using DDS by dynamically varying the perturbed parameter dimensions.
+We tested this algorithm (25 trials with 4000 function evaluations in each trial) on a biochemical network of coagulation (148 parameters and 92 species) and compared it's performance against other meta-heuristics like Differential Evolution (DE), Particle Swarm Optimization (PSO), Simulated Annealing (SA) and also against DDS alone.
+The new algorithm outperforms all the other meta-heuristics on the coagulation model.
+The parameter vectors obtained using this approach fit the experimental data well and also make accurate enough predictions on unseen experimental data.
+We also performed this comparison on commonly used test functions (Ackley and Rosenbrock) for global optimization and found the same behavior.
+Further we used two recently published benchmark problems, a genome wide kinetic model with 1759 parameters and a metabolic model of Chinese Hamster Ovary cells with 117 parameters to evaluate the performance of our approach.
+We  surprisingly performed well on these benchmarks and obtained the nominal parameter vector with just 4000 function evaluations in both cases.
 
@@
 
-%
-%Sipuleucel-T is a first generation cancer vaccine \cite{Kantoff2010}, while radium-223 is a targeted alpha emitter for bone metastasis \cite{Parker2013}. 
-%Abiraterone is an inhibitor of cytochrome P450 17A1 (CYP17), a critical enzyme in androgen biosynthesis \cite{Sartor2013} while enzalutamide inhibits androgen receptor (AR) translocation to the nucleus, as well as DNA binding and coactivator recruitment \cite{Scher2012}.
-%Thus, understanding the operation of the biological pathways which ultimately facilitate the transition to CRPC is vital to the discovery of new therapeutic strategies against this disease.
-%The model focuses on the outlaw pathway, in which constitutively or overly-activated receptor tyrosine kinases (RTKs) stimulate AKT and mitogen-activated protein kinases (MAPKs), which in turn activate cytosolic androgen receptor (AR) in the absence of androgen.
-Interestingly, among the few genes activated AR represses is cellular prostatic acid phosphatase (cPAcP), itself a key regulatory of RTK activation.
-Thus, the outlaw pathway encodes crosstalk between RTK-dependent kinase and AR activation inside of a positive feedback loop.
-The model was validated with multiple experiments in LNCaP cell lines.
-In addition  For example, in response to enzalutamide, an AR inhibitor, the model predicted a decline in PSA expression similar to patient response.
-%and there is limited data on combination therapies or the use of biomarkers to develop personalized treatment strategies.
+Cells process nutrients and respond to changes in their environment using complex biochemical networks. These networks contain thousands of components
+interconnected through nonlinear enzyme catalyzed reactions.
+Mathematical modeling has evolved as a powerful paradigm to analyze, and ultimately design these complex networks \cite{assmus2006dynamics, Riel:2006aa,Jaqaman:2006aa,kitano2002systems,hood2004systems}.
+Mathematical modeling of biochemical networks is often an iterative process.
+First, models are formulated from biochemical knowledge, and then model parameters are estimated using experimental data \cite{Aldridge:2006aa,banga2008optimization,ashyraliyev2009systems}.
+Parameter estimation is typically framed as a non-linear optimization problem wherein the residual (or objective function) between experimental data and model simulations is minimized using an optimization strategy \cite{moles2003parameter}.
+The optimal parameters obtained from training are then used to validate the model on hitherto unseen experimental data.
+If validation fails, model construction and calibration are repeated iteratively until satisfactory results are obtained.
 
-Prostate cancer (PCa) is the most commonly diagnosed cancer and the second leading cause of cancer-related death in men in the United States \cite{Siegel2013}. 
-Initially, PCa cells depend upon the activation of cytosolic androgen receptors (AR) by androgen hormones, such as testosterone, for survival and growth. 
-Androgen ablation in combination with radiation or chemotherapy remains the primary non-surgical treatment for androgen dependent prostate cancer (ADPC) \cite{Huggins1967}. 
-However, androgen ablation typically fails to permanently arrest cancer progression as malfunctioning cells eventually lose androgen sensitivity and proliferate without hormone. 
-The loss of androgen sensitivity results in castration resistant prostate cancer (CRPC), a phenotype closely linked with metastasis and greatly reduced survival \cite{Harris2009}.
-Currently, there are six approved treatments that demonstrate a survival advantage in patients with metastatic CRPC, each of these target diverse aspects of the disease \cite{Sartor2013}.
-The taxane family members docetaxel and cabazitaxel interact with microtubule stability \cite{Tannock2004,Bono2010}, while abiraterone \cite{Sartor2013} or enzalutamide \cite{Scher2012} interfere with androgen signaling by blocking androgen formation or nuclear translocation, respectively.
-Other approved treatments are non-specific to PCa. 
-For example, general treatments such as sipuleucel-T, a first generation cancer vaccine \cite{Kantoff2010}, 
-and radium-223, an alpha emitter which targets bone metastasis \cite{Parker2013}, are both approved to treat CRPC.
-Unfortunately, regardless of the therapeutic approach, the survival advantage of these treatments is typically only a few months. 
-Thus, understanding the molecular basis of the loss of androgen sensitivity in CRPC could be an important step for the development of the next generation of therapies with a prolonged survival
-advantage.
+Parameter estimation is a major challenge in the development of large biochemical models.
+Although parameter estimation has been a well studied problem in engineering for decades \cite{nieman1971review,beck1977parameter,young1981parameter,beck1998inverse},
+the complex dynamics of large biological systems and noisy, often incomplete experimental data pose a unique estimation challenge.
+Most of the optimization problems pertaining to biological systems are non-linear and multi-modal i.e.
+most of them have multiple local minima or maxima \cite{moles2003parameter,banga2008optimization}.
+The non-linearity of the problem coupled with multi-modality generally renders the local optimization techniques like pattern search method \cite{hooke1961direct}, Nelder-Mead simplex method \cite{nelder1965simplex}, steepest descent method or Levenberg-Marquardt method \cite{more1978levenberg} incapable of reliably obtaining an optimal solution since they generally stop at the first local minimum.
+Though deterministic global optimization techniques (for example algorithms based on branch and bound framework) can handle non-linearity and multi-modality \cite{esposito2000deterministic,horst2013global}, the absence of derivative information, discontinuity of the objective functions, non-smooth regions or the lack of any knowledge about the objective function severely hamper the use of these techniques.
 
-Androgen-induced proliferation and survival depends upon many coordinated signal transduction and gene expression events. 
-Androgen Receptor (AR) is part of the nuclear hormone receptor superfamily, which includes other important cancer targets such as progesterone receptor (PR) and estrogen receptor (ER) in breast cancer \cite{Aranda2001}.
-Nuclear hormone receptors act as ligand dependent transcription factors interacting with specific DNA sequences of target genes as either monomers, heterodimers, or homodimers; AR, PR, and ER act as homodimers.
-In the case of AR these specific DNA sequences are known as androgen response elements (ARE) \cite{Mangelsdorf1995}.
-In the absence of androgen, AR is predominately found in the cytoplasm bound to heat shock proteins (HSP) \cite{Prescott2006}.
-%Nuclear Hormone Receptors (Reviewed \cite{Aranda2001}) and AR in disease reviewed in \cite{Matsumoto2013}
-Androgen, either testosterone or testosterone metabolites such as 5$\alpha$-dihydrotestosterone (DHT), enter prostate cells and interact with the cytosolic androgen receptor (AR). 
-The interaction of DHT with AR promotes the dissociation of AR from chaperones such as HSP \citep{Pratt:1997kx} and its subsequent dimerization, phosphorylation and translocation to the nucleus (reviewed by Brinkmann \textit{et al.} \citep{Brinkmann1999}).
-Activated nuclear AR drives a gene expression program broadly referred to as androgen action, that promotes both proliferation and survival.
-In addition to many genes including itself, activated nuclear AR promotes the expression and secretion of prostate specific antigen (PSA), arguably the best known PCa biomarker \citep{Feldman2001}.
-PSA is commonly used as a prostate cancer indicator, although its prognostic ability is controversial \citep{Attard2009,Hoffman2011,Moyer2012}. 
-In CRPC, AR signals in the absence of androgens. 
-Androgen dependent (AD) prostate cells can become castration resistant (CR) through several possible mechanisms, including constitutively amplified AR expression and altered AR sensitivity to testosterone or other non-androgenic molecules \citep{Feldman2001}.
-In this study, we focused on the aberrant activation of AR by kinase signaling cascades, sometimes called the outlaw pathway. 
-Outlaw pathway activation is driven by over-activated receptor tyrosine kinases (RTKs), a common pathology in many cancer types including PCa \citep{Slamon1989,Craft1999}. 
-RTKs stimulate downstream kinases, including the AKT and mitogen-activated protein kinase (MAPK) pathways, which promote AR phosphorylation and dimerization in the absence of an androgen signal \citep{Craft1999,Yeh1999}.
-Interestingly, among the few genes activated AR represses is cellular prostatic acid phosphatase (cPAcP), itself a key regulatory of RTK activation \cite{Veeramani2005}. 
-Thus, in CRPC the androgen program is initiated without the corresponding extracellular hormone cue, potentially from crosstalk between growth factor and hormone receptor pathways.
+Meta-heuristic stochastic optimization approaches like Genetic Algorithms (GAs), Simulated Annealing (SA) \cite{kirkpatrick1983optimization}, Evolutionary Programming and population based searches like Differential Evolution (DE) \cite{storn1997differential} have shown promise in this regard \cite{sun2012parameter}.
+They do not make any assumptions about the structure of objective function or require any \textit{a priori} information about the objective function.
+Though they do not guarantee strong convergence, these approaches are effective in finding near optimal solutions.
+Mendes et al.
+\cite{mendes1998non} used Simulated Annealing to estimate rate constants for the irreversible inhibition of HIV proteinase, Modchang et al.
+\cite {modchang2008mathematical} used Genetic Algorithms to estimate parameters for a mathematical model of signal transduction, Differential Evolution based approaches have been an effective approach on various systems \cite{tsai2005evolutionary,wang2001hybrid,noman2007inferring}.
+Tashkova et al.
+\cite{tashkova2011parameter} compared different meta-heuristics for parameter estimation on a dynamic model of endocytosis and showed that DE was the most effective.
+Banga and co-workers have applied scatter-search based methods \cite{villaverde2012cooperative,rodriguez2006novel,egea2007scatter} to estimate parameters on non-linear biological processes.
+Hybrid approaches that combine a meta-heuristic with a local optimization search, wherein a near globally optimal solution that is obtained with a meta-heuristic is further refined using a local search have also become popular.
+Villaverde et al.
+\cite{villaverde2015biopredyn} combined scatter search with local search methods for parameter estimation in large scale systems biology models.
+Fan et al.
+recently showed that population based meta-heuristics along with decomposition based methods can be used to model gene circuits from mRNA data \cite{fan2015parameter}.Despite these successes, a  major drawback with most of these metaheuristic approaches is the vast number of objective function evaluations they take in arriving at good search regions.
+As the models grow in size and complexity and the number of dimensions of parameter vector increases, evaluation of the objective function becomes computationally expensive.
+Thus performing a large number of evaluations is not computationally feasible.
 
-In this study, we developed a mathematical model of growth factor and hormone signal integration in androgen dependent, intermediate and resistant prostate cancer cells. 
-We used this model to better understand which components and processes were differentially important in AD versus CR cells.
-The new model architecture was a significant advance over our previous prostate signaling model \citep{Tasseff2010}. 
-We added the regulated expression of ten additional proteins, including the cell cycle restriction point proteins cyclin D (and the differentially spliced variants cyclin D1a and cyclin D1b), cyclin E, cyclin-dependent kinase inhibitor 1A (p21Cip1), and cyclin-dependent kinase inhibitor 1B (p27Kip1).
-Also, we included the Rb/E2F pathway, expanded our description of the activation of the mammalian target of rapamycin (mTOR) protein and its role in translation initiation, and included the regulation of AR action by cyclin D1a and E2F.
-However, this upgraded architecture, while increasing the biological scope of the model, also expanded the number of unknown model parameters. 
-To estimate these parameters, we used multiobjective optimization in combination with dynamic and steady-state data sets generated in AD, intermediate and CR LNCaP cell lines. 
-We identified a population of approximately N = 5000 models (from well over a million candidate models) which described both AD and CR data sets using a single model structure.
-Furthermore, we tested the model using an additional 29 LNCaP data sets not used for model training, along with four drug studies.
-We analyzed the model population using sensitivity and robustness analysis to uncover differentially important mechanisms in AD versus CR cell lines. 
-In the presence of androgen, the sensitivity profile was similar between AD and CR cells. 
-Components of the MAPK and PI3K pathways were highly fragile, irrespective of the level of androgen dependence. 
-However, in the absence of androgen, there were 609 statistically significant shifts in species sensitivity between AD and CR cells.
-Of these, 108 were larger than one standard deviation above the mean.
-In CR cells, HER2 activation of the MAPK and PI3K pathways was significantly more important, as was AR activation through the MAPK pathway. 
-On the other hand, components of the translation and transcription infrastructure were differentially more important in AD cells in the absence of androgen.  
-Taken together, our analysis suggested that independently targeting the PI3K or MAPK pathways in combination with anti-androgen therapies could perhaps be an effective treatment strategy for CRPC.
+In many of these high dimensional problems approaching an exact solution may not be necessary.
+Gutenkust et al.
+\cite{gutenkunst2007universally} showed that a number of systems biology models are 'sloppy'.
+Sloppy systems have specific parameter combinations that largely define the dynamics of the system.
+Large perturbations to the rest of the parameters does not greatly impact the system dynamics.
+Ensemble approaches \cite{song2010ensembles,luan2010ensembles} have exploited this aspect to describe the dynamics of biological systems including coagulation which can be described using only a set of key species or parameters \cite{sagar2015dynamic}.
+Tolson and Shoemaker \cite{tolson2007dynamically} showed through Dynamically Dimensioned Search (DDS) that high-dimensional watershed models can be calibrated quickly by perturbing only a subset of dimensions.
 
-%(323 more and 287 less sensitive)
-%In this study, we rank-ordered the importance of the molecular components controlling the transition from ADPC to CRPC.
-%Toward this, we developed and analyzed an ensemble of mechanistic mathematical models of the signal transduction processes governing the 
-%transition from androgen dependent to castration resistant prostate cancer.
-%The objective of this study was to rank-order the importance of components in CRPC and ADPC using computational tools. 
-%We formulated and analyzed an ensemble of mechanistic mathematical models of the signal transduction processes in CRPC and ADPC 
-%that was an extension of our previous work .
-%In particular, we focused on the outlaw pathway and the possible crosstalk between RTK-dependent kinases and AR activation. 
-%First, we expanded the original model by adding ten  
-%Third, we added additional co-activators, such as cyclin E, and inhibitors of AR, such as
-%ODE models typically contain a large number of unknown model parameters, which must be estimated  
-%Due to uncertainty in parameter estimation and to account for differences between individual cells and cell sublines, an ensemble of parameter sets was used to model the system. 
-%Proteins that were not modeled at a gene regulatory level were generated using zeroth order reactions. 
-%All protein and RNA species were degraded with first order reaction rates. 
-%To obtain an ensemble of parameter sets, we used multiobjective optimization. 
-%Relative concentration values were quantified 
-%Training data came from experiments on LNCaP sublines. 
-%We used , previously developed in the lab \cite{Song2010}, to determine a diverse ensemble of models. 
-%C-33 cells, having undergone fewer than 33 passages, are considered AD while C-51 cells, having between 34 and 80 passages, are considered an intermediate prostate cancer cell line. 
-%C-81 (passage numbers beyond 80) display a CR phenotype \cite{Lin2000}. 
-%http://endo.endojournals.org/content/140/11/5054.long or http://endo.endojournals.org/content/140/11/5054.long#ref-19
+In this study, we developed dynamically dimensioned particle swarm optimization (DdPSO), a novel meta-heurstic that combined the global search capability of multi-swarm particle swarm optimization with dynamic dimensioned search.
+The objective of DdPSO was to obtain near optimal parameter estimates for large biochemical models within a relatively few function evaluations.
+DdPSO uses a multi-swarm particle swarm optimization technique to generate candidate solution vectors which are then greedily updated using dynamic dimensioned search.
+We tested the performance of DdPSO on commonly used test functions for global optimization (Ackley and Rosenbrock), published biochemical parameter estimation benchmark problems and
+lastly on a model of the human coagulation cascade.
+DdPSO outperformed common meta-heuristic approaches like Differential Evolution (DE), Particle Swarm Optimization (PSO), Simulated Annealing (SA) and dynamically dimensioned search (DDS) on the benchmark and coagulation problems.
+
+% Extra text from the introduction -
+%With the capacity to construct very large models using various mathematical formalisms \cite{chen2009input,tasseff2011modeling,luan2007computationally,mo2007genome,orth2011comprehensive,karr2012whole,buchel2013path2models,smallbone2010towards}, more often than not,
+
+%this algorithm (25 trials with 4000 function evaluations in each trial) on a biochemical network of coagulation (148 parameters and 92 species) and compared it's performance against other meta-heuristics  The new algorithm outperforms all the other meta-heuristics on the coagulation model.
+The parameter vectors obtained using this approach fit the experimental data well and also make accurate enough predictions on unseen experimental data.
+We also performed this comparison on  and found the same behavior.
+Further we used two recently published benchmark problems, a genome wide kinetic model with 1759 parameters and a metabolic model of Chinese Hamster Ovary cells with 117 parameters to evaluate the performance of our approach.
+We  surprisingly performed well on these benchmarks and obtained the nominal parameter vector with just 4000 function evaluations in both cases.
+%took into cognizance the fact that it may not be necessary to obtain the exact solution for high dimensional biological systems and that good enough solutions can be quickly obtained without expending a lot of objective function evaluations.
+Hence our current approach uses the power of population based heuristics along with DDS to obtain near optimal or good solutions.
+Though in theory we can combine any meta-heuristic with DDS, for the current purpose we used Particle Swarm Optimization (PSO).
+PSO unlike Genetic Algorithm (GA) or Differential Evolution (DE) does not have complex operations like cross over, mutation or recombination.
+It is simple to use and does not have a lot of parameters associated with other heuristics.
+However PSO is known to rapidly converge to a local optimum and thus several variants \cite{peer2003using,zhan2009adaptive,li2007fast} have been developed of which Multi swarm particle swarm optimization approaches (MLSPSO) \cite{zhao2008dynamic,liang2005dynamic} are one of the effective ones.
+We used MLSPSO to preclude bad regions of search and thereafter search within these regions using DDS.
+
+\clearpage
+\section*{Results}
+\subsection*{Dynamically dimensioned particle swarm optimization (DdPSO).}
+DdPSO is a novel meta-heuristic that combines multi-swarm based particle swarm methods with dynamically dimensioned search (Fig. \ref{fig-algorithm}).
+The goal of this approach is to obtain optimal or near optimal parameters for high-dimensional complex biological systems within a pre-specified number of function evaluations.
+We randomly initialized a swarm of $\mathcal{K}$-dimensional particles (represented as ${x}_{i}$), wherein each of these particles corresponds to a $\mathcal{K}$-dimensional parameter vector.
+After initialization, the particles were grouped into different sub-swarms randomly.
+Thereafter within each sub-swarm ${S}_{k}$,  particles were updated according to the following rule.
+\begin{eqnarray}
+	\mathbf{x}_{i,j} &=&\theta_{1,j-1}\mathbf{x}_{i,j-1} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{x}_{i,j-1}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{GL}_{k} - \mathbf{x}_{i,j-1}\right)
+\end{eqnarray}
+where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes the best local solution found by particle $i$ till function evaluation $j$-$1$, and
+$\mathcal{GL}_{k}$ denotes the best local solution found over the population of all particles within the swarm ${S}_{k}$.
+The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
+parameters ($\mathcal{K}\times{1}$). In our algorithm the parameter $\theta_{1,j-1}$ depends on the function evaluations and is controlled according to the following equation
+\begin{eqnarray}
+	\mathbf \theta_{1,j}&=&((\mathbf{N}-{j})*(\mathbf{w}_{max}-\mathbf{w}_{min}))/(\mathbf{N}-{1}) + \mathbf{w}_{min}
+\end{eqnarray}
+where $\mathbf{N}$ represents the total number of function evaluations, $\mathbf{w}_{max}$ and $\mathbf{w}_{min}$ are the maximum and minimum inertia weights respectively.
+While updating the particle, we made sure all dimensions of the solution represented by the particle were within bounds using a reflection boundary condition.
+After every $\mathbf{g}$  function evaluations, the particles within all sub-swarms were mixed and then randomly redistributed to a new sub-swarm.
+The particles were then again updated according Eq.1.
+This process continued till $\mathcal{FR}*\mathbf{N}$ number of functions evaluations, where $\mathcal{FR}$ represents the fraction of evaluations with the multi-swarms.
+At the end of these function evaluations, we froze all the solutions represented by various particles and chose the particle with best solution among $\mathcal{GL}_{1} \cdots \mathcal{GL}_{NS}$ as the initial candidate vector $\mathcal{G}$ for the remaining $({1}-\mathcal{FR})*\mathbf{N}$ number of function evaluations.
+
+This particle was then updated according to the following rule
+\begin{eqnarray}
+	\mathcal{G}_{new}(\mathbf{J})& = &\mathcal{G}(\mathbf{J})+\mathbf{r}_{normal}(\mathbf{J})\sigma(\mathbf{J})
+\end{eqnarray}
+where $\mathbf{J}$ represents the set containing the specific dimensions being perturbed, ${r}_{normal}$ denotes a normal random vector of the same dimensions as $\mathcal{G}$.
+$\sigma$ is the amplitude of perturbation given by following equation:
+\begin{eqnarray}
+	\sigma& = &\mathbf{R}(\mathcal{MAX} -\mathcal{MIN})
+\end{eqnarray}
+where $\mathbf{R}$ is the scalar perturbation size parameter, $\mathcal{MAX}$ and $\mathcal{MIN}$ are ($\mathcal{K}\times{1}$) vectors that represent the maximum and minimum bounds on each dimension.
+The set $\mathbf{J}$ was constructed using a probability function $\mathcal{P}_{j}$ that represented the probability whether a specific dimension $j$ was perturbed or not.
+This function is a monotonically decreasing function that decreases with the number of function evaluations.
+$\mathcal{P}_{j}$ can be any monotonically decreasing function, in our approach we used the following function:
+\begin{eqnarray}
+	\mathcal{P}_{j}&=&{1}-\log(j/ (\beta({1}-\mathcal{FR})*\mathbf{N}))
+\end{eqnarray}
+where $\beta$ is the perturbation frequency probability modulator.
+Thus the number of dimensions of the candidate vector that are updated or perturbed decreases with the as the number of function evaluations increase.
+These updates are greedy in nature, so $\mathcal{G}_{new}$ becomes the new solution vector only if it is better than the old one $\mathcal{G}$.
+
+\subsection*{Performance of DdPSO on benchmark problems.}
+We compared the performance of DdPSO with algorithms on commonly used test functions for global optimization, as well as benchmark biochemical problems from literature (Table \ref{table:benchmark-problems}).
+We used a 300 dimensional Rastrigin function and 300 dimensional Ackley function.
+Both Ackley and Rastrigin have multiple local minima and maxima and attain a global minimum value of 0.
+We tested the performance of the 5 different heuristic approaches on these two functions.
+In each experiment we again fixed the number of function evaluations at 4000 and ran 25 experiments.
+In both cases (Figure S6 A and B) we see that the error convergence rate for DDSMLSPSO is much faster as compared to the other four s and it finds the global minimum of 0 in both cases.
+
+Villaverde and co-workers recently published a set of benchmark problems to evaluate parameter estimation methods \cite{villaverde2015biopredyn}.
+From a computational cost perspective problems they categorized the problems as most expensive, intermediate and least expensive.
+We evaluated the performance of our algorithms on a problem from the most expensive and least expensive categories.
+The first problem is a genome wide kinetic model of \textit{Saccharomyces cerevisiae} with 1759 parameters (henceforth referred to as problem B1).
+The second problem (henceforth referred to as problem B4) is a metabolic model of Chinese Hamster Ovary (CHO) with 35 metabolites, 32 reactions and 117 parameters.
+In both cases pseudo time series data was generated.
+For problem B1, the time series data consisted of 44 observables and for problem B4 the data corresponded to 13 different metabolites.
+We fixed the number of function evaluations at 4000 for DDSMLSPSO and trained both the models against the pseudo experimental data.
+We recaptured the 'nominal' parameters (Figure S7 A and B)  in both cases within 4000 evaluations.
+The final objective function value (Table 1) is an order smaller for B1 possibly due to overfitting.
+In Table 1 we summarize the results obtained on the coagulation model and the benchmark problems.
+
+\subsection*{Performance of DdPSO on a model of the human coagulation cascade.}
+We compared the performance of DdPSO on a model of blood coagulation against commonly used meta-heuristics like simulated annealing (SA), differential evolution (DE), particle swarm optimization (PSO),and dynamically dimensioned search (DDS).
+Coagulation is an archetype biochemical network that is highly interconnected and tightly regulated with multiple positive and negative feedback loops (Fig.
+\ref{fig-coagulation-network}).
+The biochemistry underlying coagulation, though quite complex has been well studied \cite{mann2003dynamics,mann2003all,mann2003thrombin,vogler2009contact,diamond2013systems,fogelson2005coagulation,anand2003model}, and reliable experimental coagulation models have been developed \cite{hockin2002model,chatterjee2010systems,mann2006models,luan2007computationally}.
+This makes it an ideal system for mathematical modeling and parameter estimation.
+Coagulation is regulated by a set of serine proteases also known as coagulation factors and blood platelets.
+The coagulation factors are generally in an inactive state and are known as zymogens.
+These zymogens are activated through certain triggers.
+Trigger events like injury or trauma or sepsis expose factors like collagen, tissue factor and von Willebrand factor (vWF) to blood.
+The exposure of these factors to blood kick-starts a series of convergent cascades that lead to conversion of zymogen prothrombinase to thrombin.
+Luan et al.
+modeled coagulation using coupled non-linear ordinary differential equations with 148 reactions and 92 species \cite{luan2007computationally} and validated model using 21 published datasets.
+
+ To train the model parameters, we used data sets from TF-VIIa initiated coagulation with no anticoagulants.
+The objective function was a weighted linear combination of two different error functions that used data sets representing coagulation initiated with different concentrations of TF-VIIa (5pM, 5nM) \cite{hockin2002model}.
+We restricted the number of function evaluations to 4000 for each algorithm and performed 25 trials of this experiment.
+DdPSO exhibits a much faster rate of error convergence and has a much lower final error as compared to the other algorithms (Figure 3).
+Within the first 1000 function evaluations (swarm phase) of DDSMLSPSO there is a very rapid drop in error.
+Subsequently, again after 2000 function evaluations (dynamically dimensioned phase) the error drops quickly.
+Overall at the end of 4000 function evaluations DDSMLSPSO minimizes the error (final objective error is 90456.9) to a much greater extent than any of the other algorithms.
+Amongst the rest of the algorithms DDS has a comparable error drop for the first 2000 function evaluations.
+However the drop in error remains constant for DDS after 2000 function evaluations.
+Using the parameters obtained at the end of 4000 function evaluations we examined the 'fits' between models predictions and experimental data (Figure 2).
+The solid lines represent the mean value of prediction over 25 trials and the shaded region represents the 99\% confidence interval.
+Subsequently we compared the model predictions against completely 'unseen' or untrained experimental where coagulation was initiated with 500pM,50pM,10pM concentrations of TF-VIIa respectively (Figure 3).
+
+\clearpage
+\section*{Discussion}
+Our study presents a novel approach for high-dimensional parameter estimation in complex biological systems with relatively few function evaluations.
+In this approach we combined a variant of a well known meta heuristic particle swarm optimization with Dynamically Dimensioned Search (DDS).
+We tested our approach on an ODE model of coagulation with 148 parameters and 92 species.
+Coagulation is an ideal system to test our approach since the biology is well known and complex, with multiple feed back loops that are tightly regulated.
+We used experimental data with under different conditions to obtain optimal parameters and used these parameters to make predictions against unseen experimental data.
+We obtained good fits and made accurate enough predictions using parameters obtained from 4000 function evaluations.
+Further, we also used high-dimensional forms of commonly used test functions of global optimization and showed that we were able to find the global minimum for 300 dimensional Ackley and Rastrigin functions faster than other meta-heuristics.
+We also considered two recently published benchmark problems to test parameter optimization approaches and showed that we were able to retrieve the nominal parameter vector within 4000 function evaluations.
+
+meta-heuristic approaches are generally effective in finding close to optimum solutions of complex, multi-modal functions.
+In addition, they generally obviate the need for any \textit{a priori} knowledge (like function derivative).
+However they take an exorbitant number of objective function evaluations to come close to an optimum.
+When the objective function evaluations tend to become expensive it is infeasible to take up a large number of evaluations.
+As the dimensionality of parameter space increases, the search region gets widened and thus the problem becomes more challenging.
+In addition, most of these approaches require optimization of  'algorithm parameters' before the actual optimization and also involve computationally expensive update operations.
+Tolson and Shoemaker, through DDS, showed that randomly perturbing a subset of dimensions in high dimensional parameter space is an effective way to obtain near optimal solutions with few function evaluations.
+However their approach was based on a single solution.
+Single solutions based methods may prove ineffective in complex systems when they start from a 'bad' region of search or from a local minimum.
+In our approach, we tried to eliminate the probability of starting from a bad region of search using a variant of particle swarm optimization.
+Particle Swarm Optimization (PSO) is a population based meta heuristic that is known for rapid convergence without any complex operations like recombination, mutation or selection that are associated with other population based meta-heuristics like Differential Evolution (DE) or Genetic Algorithm (GA).
+Several particle swarm variants have been proposed to improve the search ability and rate of convergence, that involve different neighborhood structures, multi-swarms or adaptive parameters.
+Multi-swarm PSO with small particle neighborhoods have been shown to better in searching on complex multi-modal solutions \cite{zhao2008dynamic}.
+Multi swarm methods, in addition, avoid rapid convergence to a local optimum or stable point and are able to generate diverse solutions.
+Generation of diverse solutions in the early stage gives a better exploratory capability and thus converge of upon multiple optima.
+
+We utilized this capacity of multi swarms to generate diverse candidate solutions which can be used as initial solutions for a DDS like search.
+Thus the solution vectors obtained at the end of multi swarm search phase have a better propensity to not get stuck in bad regions in DDS phase as opposed to starting with a totally random initial solution.
+Given a fixed number of function evaluations, we have been able to show that we were able to obtain better solutions for the coagulation model and the test functions faster that other commonly known meta-heuristics and also DDS alone.
+Choosing the number of function evaluations is largely a function of cost and complexity of the objective function.
+Traditionally the stopping condition for a parameter optimization problem can be the number of function evaluations, percentage of initial error achieved or an absolute error threshold.
+However in case of complex, expensive functions where we desire a value within a certain period of time, the number of evaluations are used as a stopping criterion.
+In our current study we used a value of 4000 which we based upon the time taken (approximately 8-10 seconds) for a single objective function evaluation in the coagulation case.
+We used the same value of 4000 for benchmarks published by Villaverde and co-workers \cite{villaverde2015biopredyn}.
+Quite surprisingly we took a couple of orders lesser number of function evaluations (1e3) to obtain the optimal parameter vector as compared to the enhanced Scatter Search (eSS) with a local optimizer, which took around the order of 1e5 number of evaluations.
+The amount of CPU time taken (on an Intel Xeon processor 2.4 GHz) is lesser as compared to eSS on a similar architecture.
+
+A surprisingly remarkable aspect about our algorithm has been that we did not 'pre optimize' any parameters of the algorithm to suit a specific problem.
+In the swarm phase this includes the number of particles, number of sub-swarms, acceleration constants or the number of generations after which the particles are redistributed and the neighborhood perturbation parameter in DDS phase.
+We used the same parameters for all the problems.
+The same rule was applied to the rest of the meta-heuristics barring Simulated Annealing.
+For SA, we optimized the cooling schedule for the coagulation model.
+Thus, in this approach any overhead that usually comes with additional function evaluations to optimize parameters was avoided.
+The only parameter that we adjusted was the fraction of evaluations within swarm search.
+For the coagulation model it was set to be 0.4 and for the benchmarks it is at 0.6.
+This parameter was set based on empirical observations rather than any definitive analysis.
+When the error convergence rates stagnated in the swarm phase we switched to DDS phase.
+A further improvement to our approach could  be to incorporate a meta intelligent way of switching in and switching out of both the phases.
+
+The performance of our approach seems impressive given that it performed well on different complex systems with no specific conditioning being required.
+We comfortably outperformed common meta-heuristics and were also able to find minima of high dimensional global optimization test functions.
+Thus this approach may be well suited to large scale global optimization.
+In addition, surprisingly, we were able to obtain optimal parameter vectors for two different large scale systems biology models with a couple of orders fewer number of function evaluations as compared a hybrid approach.
+However it is quite possible that highly optimized versions of common meta heuristics may outperform us on these systems.
+This aspect is currently beyond the scope of this study.
+Our approach can also be combined with local derivative based searches to improve upon the accuracy of the solutions.
+
+%\begin{figure}[H]
+%\centering
+%\includegraphics[width=1.00\textwidth,height=0.7\textheight]{./figs/Figure_2_CoagulationNetwork.pdf}
+%\caption{Schematic of the extrinsic and intrinsic coagulation cascade.
+Inactive zymogens upstream (grey) are activated by exposure to tissue factor (TF) \cite{giesen1999blood} following vessel injury.
+Tissue factor and activated factor VIIa (FVIIa) form a complex that activates factor X (fX) and IX (fIX).
+FXa activates downstream factors including factor VIII (fVIII) and fIX.
+Factor V (fV) is primarily activated by thrombin (FIIa).
+In addition, we included a secondary fV activation route involving FXa.
+FXa and FVa form a complex (prothrombinase) on activated platelets that converts prothrombin (fII) to FIIa.
+FIXa and FVIIIa can also form a complex (tenase) on activated platelets which catalyzes FXa formation.
+Localized platelets are activated by external signals such as adenosine diphosphate (ADP) and thromboxane A2 (TXA2) or thrombin through protease-activated receptors (PARs) \protect\cite{coughlin2000thrombin} \protect \cite{coughlin1999protease} Thrombin also activates upstream coagulation factors, forming a strong positive feedback ensuring rapid activation.
+Tissue factor pathway inhibitor (TFPI) downregulates FXa formation and activity by sequestering free FXa and TF-FVIIa in a FXa-dependent manner.
+Antithrombin III (ATIII)  inhibits all proteases.
+Thrombin inhibits itself binding the surface protein thrombomodulin (TM) \protect\cite{esmon1989roles} The IIa-TM complex catalyzes the conversion of protein C (PC) to activated protein C (APC), which attenuates the coagulation response by the proteolytic cleavage of fV/FVa and fVIII/FVIIIa \protect\cite{esmon1989roles}}\label{fig-network}
+%\end{figure}
+
+\section*{Acknowledgements}
+This study was supported by an award from the Army Research Office (ARO \#59155-LS).
+\clearpage
+
+%\bibliography{References_v1}
+%\begin{thebibliography}{9999}
+\bibliography{References_v2}
 
@@
 
-\section*{Results}
-
-\subsection*{Estimating an ensemble of prostate signaling models.} 
-We modeled the integration of growth factor, cell cycle and hormone signaling pathways in AD and CR LNCaP cells (Fig. \ref{fg:NetworkSchematic}). 
-The signaling architecture was hand curated from over 80 primary literature sources in combination with biological databases.
-The model equations were formulated as a system of ordinary differential equations (ODEs), where biochemical reaction rates were modeled using mass action kinetics.
-ODEs and mass action kinetics are common modeling tools \citep{Wayman2013}, however, ODEs have the disadvantage of requiring estimates for unknown model parameters.
-Many techniques have been developed to estimate ODE model parameters, often from noisy and sparse experimental data \citep{Moles:2003dn}.
-Typically these identification problems are underdetermined, hence no unique parameter values can be estimated \citep{Villaverde:2014kl}.
-Thus, instead of estimating a single yet highly uncertain parameter set, we estimated an ensemble of possible parameter sets using the Pareto Optimal Ensemble Techniques (POETs) algorithm \citep{Song2010}.
-POETs uses a combination of simulated annealing and local optimization techniques coupled with Pareto optimality-based ranking to simultaneously optimize multiple objective functions.
-Starting from an initial best fit set, we estimated the 1687 unknown model parameters (1674 kinetic parameters and 13 non-zero initial conditions) using 43 \textit{in vitro} data sets taken from AD, intermediate and CR LNCaP cells (Table~\ref{objective_table}).
-Each of the 43 training data sets was a separate objective in the multiobjective calculation. 
-The training data were steady-state or dynamic immunoblots from which we extracted relative species abundance from their optical density profiles. 
-POETs sampled well over a million possible parameter sets, from which we selected N = 5000 sets for further analysis. 
-Over the ensemble, the coefficient of variation (CV) of the kinetic parameters spanned 0.59 - 5.8, with 33\% of the parameters having a CV of less than one (Fig. \ref{fg:Supp_CV}). 
-As a control, we also performed simulations for R = 100 random parameter sets to compare against the ensemble generated by POETs.
-
-The ensemble of PCa models recapitulated training data in both AD and CR cell lines with only two experimentally justified parameter changes (Fig. \ref{fg:ColorPlot} and Fig. \ref{fg:Training}). 
-Data from the LNCaP clones C-33 (dependent), C-51 (intermediate), and C-81 (resistant) \cite{Lin2000,Igawa2002,Horoszewicz1983} along with the CR LNCaP cell lines LNCaP-Rf \cite{Murillo2001}, LNCaP-AI \cite{Chen2011} and LNAI \cite{Graff2009} were used for model identification.
-To simulate the effective difference between LNCaP cell lines, the parameter controlling the maximum rate of PAcP gene expression was scaled by 0.1 and 0.5, respectively, for the C-81 and C-51 cell-lines compared to C-33.
-This modification was based upon steady-state PAcP data from the three LNCaP clones \cite{Lee2003}. 
-Similarly, the expression of p16INK4 was adjusted in accordance with the study of Lu \textit{et al.} \citep{Lu1997}. 
-These two parameters were the only adjustable parameter differences between AD and CR cells. 
-To simulate an increased mTOR activation in the presence of a DHT stimulus, we added a first order activation term for mTOR activation with a DHT stimulus.
-Androgens have been shown to increase expression of proteins involved in cellular metabolism, which may lead to an increase in mTOR activation \cite{Xu2006}.
-The model fit 36 of the 43 training objectives for greater than 40\% of the ensemble members (Fig. \ref{fg:ColorPlot}A). 
-Conversely, only 10 of the 43 training objectives were captured with the random parameter control (Fig. \ref{fg:ColorPlot}B). 
-The model captured the crosstalk between RTK activation and androgen action (Fig. \ref{fg:Training}). 
-The model described DHT-induced PSA expression in both C-33 (Fig. \ref{fg:Training}A) and C-81 (Fig. \ref{fg:Training}B) cells. 
-Interestingly, simulations with the HER2 inhibitor AG879 recapitulated decreased PSA expression in C-81 cells (Fig. \ref{fg:Training}C) in the absence of androgen stimulation.
-AR action decreased the PAcP mRNA message (Fig. \ref{fg:Training}D), presumably leading to increased HER2 activity. 
-The model also recapitulated the integration of androgen action with AR expression, G1/S cell cycle protein expression and AKT phosphorylation. 
-For example, the model captured AR-induced AR expression following a DHT stimulus (Fig. \ref{fg:Training}H).
-Conversely, the transcription factor E2F inhibits AR transcription in LNCaP cells (Fig. \ref{fg:Training}I). 
-Other cell cycle proteins were also integrated with androgen action.
-For example, the cyclin D1 abundance increased in CR compared to AD cells in the absence of androgen (Fig.
-\ref{fg:Training}E), while DHT induced p21Cip1 expression in C-33 cells (Fig.
-\ref{fg:Training}F).
-The level of phosphorylated AKT was also increased in higher passage number cells (Fig. \ref{fg:Training}G). Taken together, [FINISH ME].
-
-%We used a single model architecture to describe the behavior of both AD and CR LNCaP cells, 
-%Two new additions to the model were the cell cycle protein p21Cip1 and transcriptional regulation of AR. 
-%We were able to capture training data for both mechanisms. 
-%AR transcriptionally upregulates its own expression \cite{Lin2003}\cite{Grad1999}. 
-%We captured the increase in AR concentration due to AR activation, and thus increased transcriptional activity, upon  
-%Higher basal AKT phosphorylation leads to increased activation of the PI3K/AKT pathway in PCa cells with less androgen dependence \cite{Lin2003}. 
-%\cite{Davis2006}
-
-%An additional protein we considered in validation that was present in training was phosphorylated SHC-transforming protein 1 (Shc).
-%Simulations predicted a similar increase in Shc phosphorylation after DHT stimulus (Fig. \ref{fg:Validation}D) \cite{Veeramani2005Onc}. 
-%The Prostate Cancer Working Group (PCWG) updated outcome measurements for castration resistant prostate cancer studies in 2008 \cite{Scher2008}. 
-%The primary endpoint in prostate cancer studies is overall survival, while PSA response is often a secondary endpoint. 
-%Figure \ref{fg:Validation}I shows fold change of PSA of ensemble members due to the application of sorafenib.
-
-%(Fig. \ref{fg:Validation}H)
-
-\subsection*{Validation simulations revealed missing network structure.} 
-The model was validated against 29 \textit{in vitro} and four \textit{in vivo} studies (Table~\ref{prediction_table}). 
-For 15 of the 29 cases, greater than 40\% of the ensemble was qualitatively consistent with the experimental data (Fig. \ref{fg:ColorPlot}C). 
-However, for the random parameter control, only 7 of the 29 cases were satisfied (Fig. \ref{fg:ColorPlot}D). 
-We correctly predicted positive feedback between HER2 auto-activation and androgen action (Fig. \ref{fg:Validation}A and Fig. \ref{fg:Validation}B). 
-We also captured the dose-dependence of AR abundance on DHT (Fig. \ref{fg:Validation}C). 
-In addition to the cell line studies, we simulated the outcome of enzalutamide, lapatinib, and sorafenib clinical trials in AD and CRPC patients. 
-The trial end points were the reduction in PSA expression relative to an untreated baseline.  
-Enzalutamide acts on AR by inhibiting its nuclear translocation, DNA binding, and coactivator recruitment \cite{Scher2012}. 
-In the enzalutamide trial, 54\% of the patients that received the drug showed a PSA decline of $\geq$ 50\% while 25\% showed a decline $\geq$ 90\%. 
-We simulated enzalutamide exposure by reducing the rate constants governing activated AR binding to nuclear importer, cyclin E, and CDK6 to 1\% of their initial values. 
-Consistent with the trial, 62\% of ensemble members showed a $\geq$ 50\% decline in PSA abundance, while 14\% showed a $\geq$ 90\% decline (Fig. \ref{fg:Validation}G). 
-The second trial we simulated involved exposure of CRPC patients to sorafenib. 
-Sorafenib is a kinase inhibitor with activity against Raf, vascular endothelial growth factor receptor (VEGFR), platelet-derived growth factor receptor (PDGFR), c-kit and c-Ret \cite{Dahut2008}. 
-We considered only the effects of sorafenib on the protein kinase Raf, as VEGFR, PDGFR, c-kit and c-Ret were not included in the model. 
-None of the 22 patients in the sorafenib study showed a PSA decline of $>$ 50\%. 
-However, our simulations showed that approximately 55\% of the ensemble members had a PSA decline of $\geq$ 50\%. 
-The last drug we considered was lapatinib, an inhibitor of epidermal growth factor receptor (EGFR) and HER2 tyrosine kinase activity \cite{Liu2013}. 
-Two lapatinib drug trials were considered: one in which patients had CRPC and one in which patients had biochemically relapsed ADPC \cite{Whang2013,Liu2013}. 
-In the CRPC lapatinib drug trial, two of the 21 enrolled patients had a PSA response $\geq$ 47\%  \cite{Whang2013}. 
-For the CRPC case, our model showed 26.5\% of ensemble members with a PSA response $\geq$ 47\%. 
-Of the 35 patients enrolled in the ADPC lapatinib study, no PSA decreases was observed \cite{Liu2013}. 
-In this case, our model showed 9.2\% of ensemble members with a PSA response $\geq$ 50\%.
-Although no response to lapatinib was seen in ADPC clinical trials, \textit{in vitro} AD LNCaP experiments showed decreased PSA expression in response to lapatinib, most notably with the addition of DHT \cite{Liu2005}.
-
-Validation and training failures suggested the original signaling architecture was missing critical components. 
-Several of the failed training and validation simulations involved the response of the network to epidermal growth factor (EGF) stimulation. 
-For example, Chen \textit{et al.} showed that HER2 phosphorylation increased within five minutes following EGF stimulation of LNCaP-AI cells \citep{Chen2011}. 
-We predicted no connection between HER2 phosphorylation and EGF stimulation on this short timescale (Fig. \ref{fg:Validation}E). 
-Interestingly, we initially neglected the heterodimerization of HER2 with other ErbB family members in order to simplify the model.
-However, Chen \textit{et al.} suggested that HER2-EGFR heterodimerization could be an important factor in EGF-driven activation of HER2 \cite{Chen2011}. 
-We tested this hypothesis by developing a new model that included HER2 and EGFR heterodimerization. 
-We set the rate constants governing the assembly of HER2/EGFR heterodimers equal to EGFR homodimer assembly; all other parameters were unchanged.
-This was a reasonable first approximation, as the affinity of HER2/EGFR heterodimerization and EGRF homodimerization is thought to be similar \cite{Hendriks2003}. 
-With the inclusion of HER2-EGFR heterodimerization, we qualitatively fit the EGF-induced HER2 activation case and improved our training for experiments that involved an EGF stimulus, e.g., cyclin D mRNA and protein abundance following an EGF stimulus in C-33 cells (Fig \ref{fg:ColorPlot}A and C, white pixels).
-
-%in diverse environments (i.e., DHT, no DHT, drug) 
-%Stimuli included the presence or absence of DHT, presence of therapeutics, and also EGF. 
-%PIP3 recruits AKT and PDK1 to the membrane where PDK1 activates AKT \cite{Vivanco2002}. 
-%Thus, the PIP3 and AKT interaction is required in the model for AKT phosphorylation to occur.
-
-\subsection*{Sensitivity analysis identified differentially important features of the prostate architecture.} 
-We used sensitivity analysis to identify important signaling components in AD versus CR cells (Fig. \ref{fg:Sensitivity}). 
-We calculated first order steady-state sensitivity coefficients under different stimuli for 500 parameter sets selected from the ensemble. 
-Signaling components were rank-ordered based upon analysis of their sensitivity coefficient values. 
-In the presence of DHT, the sensitivity profile was similar for AD versus CR cells, with only a few differences (Fig. \ref{fg:Sensitivity}B). 
-The top 2\% of sensitive species, regardless of androgen dependence, involved components from the MAPK and PI3K pathways. 
-In particular, activated Ras, Raf, phosphorylated MEK, as well as PIP3 localized AKT, phosphorylated AKT, and PI3K were sensitive in both AD and CR cells. 
-Species involving PAcP and p16INK4 were more sensitive in AD cells, which was expected since the expression of these two proteins were the only parameters changed between AD and CR cells. 
-Other species such as E2F, cyclin E, and DHT-activated AR were also more sensitive in AD cells. 
-On the other hand, HER2-Grb2-Gab activation of PI3K and AKT inhibition of RAF were more sensitive in CR cells.
-
-The importance of signaling components varied with androgen dependence in the absence of DHT (Fig. \ref{fg:Sensitivity}A). 
-There were 609 statistically significant shifts in species sensitivity (318 more and 291 less sensitive) between CR and AD cells in a non-androgen environment. 
-However, only 108 of these shifts were greater than one standard deviation above the mean. 
-In CR cells, HER2 activation of ERK and PI3K was more sensitive, as was AR activation through the MAPK pathway. 
-This was expected, as outlaw pathway activity was elevated in castration resistant cells. 
-Species in the MAPK pathway were in general more sensitive in CR cells (128 out of 140 significant), with all forms of sPAcP more robust in CR cells.
-On the other hand, infrastructure pathways encoding transcription and translation were more sensitive in AD cells.
-PSA  and cyclin D1b (mRNA and mRNA complexes) were the only species involved in translation that were more robust in AD cells (14 out of 116).
-The transcription factor, E2F was more fragile in AD cells, while the transcription factors ETS and AP1 were more robust. 
-ETS and AP1 are activated by phosphorylated ERK, and ETS is also activated by active PKC \cite{Wilkinson2000,Lindemann2003}. 
-E2F is deactivated through binding to Rb, which is deactivated by cyclin D1 and CDK phosphorylation \cite{Lapenna2009}. 
-The model also included AP1 suppression of AR transcriptional activity (more sensitive in CR) \cite{Sato1997}, as well as inhibition of transcription of the AR gene by E2F (more sensitive in AD) \cite{Davis2006}.
-Species in the PI3K pathway that were more fragile in AD cells included Rheb and TOR complexes. 
-Interestingly, these species were included as the last step in the PI3K pathway prior to translation, with the phosphorylation of 4E-BP1 by TOR being considered the beginning of translation in this model.
-This again indicates that in the absence of DHT general translation is more fragile in AD cells.
-
-There was a large shift in sensitive species between an androgen and a non-androgen environment in both AD and CR cell lines (Fig. \ref{fg:Sensitivity}C and Fig. \ref{fg:Supp_Sensitivity}). 
-Of the 664 statistically significant shifts in AD cells, 288 were more sensitive between androgen versus non-androgen environments. 
-However, only 119 shifts were larger than one standard deviation above the mean. 
-Unsurprisingly, AR activation through DHT binding, with and without coactivators, in a DHT environment was more sensitive, as was AR inhibition of PAcP transcription (repressed by AR in the model). 
-Species further upstream, such as HER2 activation of the MAPK and PI3K/AKT pathways, were also more sensitive in a DHT environment. 
-Cell cycle species that were more fragile in the presence of DHT, included complexes involving p21Cip1 and CDC25A.
-In a non-androgen environment, basal transcription (68 out of 72) and translation (114 out of 120) were more fragile. 
-Other fragile species in the absence of DHT included Rb, E2F, Sam68, cyclin D1a complexes, phosphatases in the MAPK pathway, Rheb complexes, and TOR complexes.
-
-We also considered the sensitivity of CR cells following the application of the AR inhibitor enzalutamide in the presence of DHT (Fig. \ref{fg:Sensitivity}D). 
-Species which were more sensitive in an androgen environment with enzalutamide included cytosolic AR, cPAcP, and p21Cip1. 
-As we would expect, AR species found in the nucleus and/or bound to coactivators, were more robust in the absence of enzalutamide. 
-The top two percent of sensitive species with and without enzalutamide were conserved. 
-In a CR cell, enzalutamide had no effect on the sensitivity of PI3K/AKT species as well as many MAPK species (ERK, Raf, and MEK). 
-Next, we looked at the effect of enzalutamide on a CR cell in both a non-androgen and DHT environment (Fig. \ref{fg:Supp_Sensitivity}). 
-More sensitive species in a non-androgen environment included dimerized HER2, ERK, and PAcP. 
-Species which were more robust in the non-androgen environment included, AR activated by DHT, AKT, p70, and AR bound to HSP.
-The results of our sensitivity analysis indicate that instead of inhibiting solely the AR pathway (enzalutamide), a combination therapy targeting the PI3K or MAPK pathways in addition to AR may be more effective.
-
-\subsection*{Experimental results confirm the need for dual therapies in prostate cancer.}
-
-The results from our sensitivity analysis indicate that instead of inhibiting solely the AR pathway (enzalutamide), a combination therapy is necessary. 
-To test this hypothesis we used the well characterized ADPC cell line LNCaP as well a LNCaP derived CRPC cell line C4-2 \cite{Thalmann1994}. 
-Three inhibitors were used: the AR inhibitor MDV3100 (enzalutamide), the Raf kinase inhibitor sorafenib, and the PI3K inhibitor LY294002. 
-In both cell lines, inhibition of either the AR or MAPK pathway appears to promote activation of the PI3K pathway, as seen by the increase in pAKT (S473) (Fig. \ref{fg:Experiment}A). 
-The addition of the PI3K inhibitor, LY294002, alone or in combination diminishes PI3K activity. 
-The inhibition of PI3K alone, led to an increase in AR expression in both LNCaP and C4-2 cell lines. 
-Since, AR transcriptionally upregulates its own expression \cite{Lin2003}\cite{Grad1999}, this may indicate an increase in AR activity. 
-The ribosomal protein pS6 was completely inhibited only in the presence of the PI3K inhibitor LY294002. 
-Cell viability results show a large decrease in cell viability at 72 hrs in the dual inhibition cases as well as the triple inhibition case for both LNCaP and C4-2 cell lines (Fig.
-\ref{fg:Experiment}B).
-In both cell lines, MDV3100 (10 $\mu$M), has only a modest effect on cell viability versus control (DMSO). 
-Figure \ref{fg:Experiment}C shows cell viability of both cell lines in varying concentrations of inhibitors at 24 hrs.
-
-\subsection*{Robustness analysis identified key regulators of prostate cancer.} 
-Robustness analysis was conducted for 80 proteins to quantify the effects of amplifying or knocking down key model components in both AD and CR cells. 
-Gene expression parameters were altered by a factor 10, 0.5, and 0 for knock-in, knock-down, or knock-out perturbations, respectively.
-We calculated the effect of these perturbations on different protein markers, such as PSA, AR, and cyclin D. 
-A knock-out of Raf, MEK or ERK showed an overall increase in cyclin D levels in CR cells (Fig. \ref{fg:Supp_Robustness}). 
-This was unexpected and we saw a similar increase in cyclin D due to the knock-in of Raf, MEK or ERK. 
-We found that individual ensemble members showed different response to a Raf knock-out, in both cyclin D concentration and PSA concentration. 
-Of the 500 ensemble members, 126 members saw an increase in PSA concentration and 62 members saw an increase in cyclin D concentration due to the knock-out of Raf (Fig. \ref{fg:Robustness}). 
-We saw three distinct regions: (1) increase in PSA concentration, (2) increase in cyclin D concentration, and (3) a decrease in both PSA and cyclin D. 
-We explored the flux vectors of the outlying parameter sets to understand the mechanistic effect of Raf knock-out on PSA and cyclin D. 
-Outlying parameter sets in region 1 displayed high activation of PI3K through HER2 signaling as well as high association of AP1 with AR. 
-AP1 binds and suppresses AR transcriptional activity in LNCaP cells \cite{Sato1997}. 
-Knocking out Raf lowered AP1 levels and, therefore, freed AR for increased transcription of PSA.
-In region 2, parameter sets also had high activation of PI3K through HER2. 
-They also had higher association of E2F with Rb and cyclin D1a with AR. 
-Cyclin D levels in this region increase due to an increase in E2F levels caused by the Raf knock-out. 
-Parameters in region 3 have high association of TOR. 
-Interestingly, the drug sorafenib, a multi-kinase inhibitor that has activity against Raf, showed no measurable PSA decline in prostate cancer patients in clinical trials \cite{Dahut2008}. 
-The robustness analysis showed that network perturbation can result in unexpected responses due to cell-to-cell heterogeneity in gene expression. 
-These outlying cell types could be critical for understanding when designing drug targets and combination therapies.
-
-%The results of the model suggest that an inhibition of either the PI3K pathway or the MAPK pathway in combination with an AR inhibitor as a possible therapy for CRPC. 
-%Sensitivity analysis revealed no change in the top sensitive species in the presence or absence of the AR inhibitor, enzalutamide.
-PI3K/AKT and MAPK species continued to fall in the top two percent of sensitive species.
-%A study by Carver \textit{et al.} looked at dual inhibition of AR and PI3K signaling in LNCaP cells and in a Pten-deficient murine prostate cancer model \cite{Parker2013}. 
-%Using both the PI3K inhibitor, BEZ235, and the AR inhibitor, MDV3100 (enzalutamide), the group saw a drastic decrease in the total number of cells. 
-%Each inhibitor on it’s own had a much smaller effect on total cell number. 
-%They saw an increase in the cell death marker, c-PARP, in the dual inhibition case. 
-%The group hypothesized that AKT inhibition leads to increased AR signaling activity through increased protein concentrations of HER3. 
-%On the other hand, AR inhibition leads to increased AKT activity due to the down regulation of PHLPP, a protein phosphatase that regulates AKT. 
-%For simplicity, the HER3 pathway and also cell death were not included in the model. 
-%Dual knock-out studies of PI3K and AR in our model show no additive effect on any cell cycle proteins through the dual knock-out. 
-%The pathways in our model appear to be uncoupled and therefore no synergy is shown in the dual knock-out case. 
-%This could indicate that the combined decrease in cell population is entirely due to cell death. 
-%The Carver \textit{et al.} study did not consider cell cycle proteins or cell growth. 
-%Our model does show a decrease in cell cycle proteins in the PI3K knock-out as well as in the dual knock-out case. 
-%This result seems to be consistent with the decreased cell growth in the PI3K knock-out case which is not dependent on cell death, as c-PARP levels are low. 
-%The decrease in cell cycle proteins is due to a decrease in general translation, including free eIF4E levels and activated 40S ribosome subunit. 
-%The decrease in p70 (S6) activation due to inhibition of PI3K is shown in both the model and in the Carver \textit{et al.} study (Supplementary Figure), indicating this result is due to the PI3K pathway.
-%It would be interesting to repeat the experiment looking at cell cycle proteins and also performing the experiment in CR LNCaP cells, instead of AD cells.
-
-\section*{Discussion}
-In this study, we developed a population of mathematical models describing growth factor and hormone signal integration in androgen dependent, intermediate and resistant prostate cancer cells. 
-These models described the regulation of androgen receptor expression and activation through androgen binding as well as a ligand-independent, MAPK-driven mechanism referred to as the outlaw pathway. 
-An ensemble of model parameters was estimated using 43 steady-state and dynamic data sets taken from androgen dependent, intermediate and independent LNCaP cell lines using multiobjective optimization.
-Further, we tested the predictive power of the model by comparing model predictions against 33 novel data sets (including four \textit{in vivo} drug studies) not used during model training. 
-The model ensemble captured 84\% of the training data and 52\% of the validation data relative to 23\% and 24\% for a random control population. 
-Interestingly, during the the initial round of parameter estimation, we identified several potentially missing structural components not present in the original connectivity. 
-One such component, EGF-induced HER2/EGFR heterodimerization, was added to the current generation model. 
-Inclusion of this structural component significantly improved both training and validation performance using the same rate constants as the EGFR-homodimer case (no additional parameter fitting). 
-We then analyzed the population of signaling models, using both sensitivity and robustness analysis, to identify the critical components controlling network performance in a variety of conditions.
-
-%Hendrick's \textit{et al.} hypothesized that HER2 heterodimerization amplified signals by allowing just one molecule of EGF to activate two receptors \cite{Hendriks2003}. 
-%In our case, we see an amplified cyclin D concentration due to an equal amount of EGF stimulus. %
-%Model connectivity involving EGF and the downstream effect on AR and AR-regulated genes also came up as an open question. 
-%Whether EGF downregulates or upregulates AR activity is debatable and the biology is still unknown \cite{Cai2009,Cinar2005,Lin2003,Lin2001PNAS, Hakariya2006,Gregory2004,Ponguta2008,Guo2006,Weber2004}.
-%The current model assumed no direct effect of EGF on AR, and instead we saw increased AR activity through increased HER2 activation. 
-%Further studies will need to be done to update this section of the model.
-
-In addition, three of the validation cases involved the effect of EGF on AR and AR-activated genes, i.e., PSA. 
-Cai \textit{et al.} showed decreased expression of endogenous AR as well as androgen-regulated PSA in AD LNCaP cells due to an EGF stimulus \cite{Cai2009}. 
-Cinar \textit{et al.} also showed decreased AR protein expression due to EGF, an effect reversed by the mTOR inhibitor, Rapamycin \cite{Cinar2005}. 
-Model simulations show either the opposite trend or no effect due to EGF stimulus (Fig. \ref{fg:Validation}F) \cite{Cinar2005}. 
-These results suggest missing network structure. 
-From additional literature searches, the inhibition of AR activation through EGF is still an open question, with many groups debating the biology involved, predominately in the PI3K/AKT pathway. 
-Lin \textit{et al.} found that in low passage number LNCaP cells (C-33), AKT negatively regulates AR by destabilizing it and marking it for ubiquitylation. 
-In high passage number LNCaP (C-81), AKT levels are high which contribute to AR stability and less degradation \cite{Lin2003}. 
-Wen \textit{et al.} showed that HER2 could induce AKT activation and LNCaP cell growth in the presence and absence of androgen \cite{Wen2000}. 
-Another study shows AKT phosphorylation of AR at S213 and S790 suppresses AR transactivation and AR-mediated apoptosis of LNCaP \cite{Lin2001PNAS}. 
-The study from Cai \textit{et al.} showed the reduction in AR was not due to degradation or PI3K/AKT signaling, but instead was due to decreased AR mRNA levels \cite{Cai2009}. 
-They found that AR protein levels in CR cells were not affected by EGF. 
-Others though have found that PSA expression, even in C-81 cells, is decreased by EGF \cite{Hakariya2006}. 
-In other prostate cell lines, EGF has been shown to increase AR transactivation \cite{Gregory2004,Ponguta2008}. 
-The MAPK pathway, which is downstream of EGFR, may also enhance AR responses to low levels of androgen \cite{Guo2006,Weber2004}. 
-Due to the discrepancies in the literature, experiments should be performed before adding additional network connectivity to the model.
-
-The population of PCa models was analyzed using sensitivity analysis to identify key signaling components and processes in both AD and CR cells. 
-There was very little difference between sensitive and robust components in AD versus CR cells in the presence of androgen. 
-MAPK and PI3K pathway components were consistently ranked in the top 2\% of sensitive species in the presence of androgen for both AD and CR cells.
-On the other hand, cell cycle species, such as cyclin D-CDK4/6 complexes bound to cell cycle inhibitors (p27Kip1, p21Cip1, p16INK4), were consistently robust.
-However, this profile changed considerably in the absence of androgen.
-The activation of PI3K and ERK by HER2 dimerization and autophosphorylation was significantly more important in CR versus AD cells.
-Interestingly, AR activation by ERK was also more sensitive in CR versus AD cells in the absence of androgen.
-Lastly, although AR-regulated transcriptional processes were equally sensitive between the cell types, general translational and transcriptional components were more robust in CR versus AD cells.
-This evidence supports the current theory that CR cells will still respond to androgen and, thus, AR is still an active target in therapeutic against CRPC \cite{Karantanos2013}. 
-Supporting the argument that AR can be activated in the absence of androgens by MAPK activation \cite{Feldman2001}. 
-Advanced prostate cancers often have higher levels of E2F and other transcription factors \cite{Davis2006}.
-Interestingly, E2F was more sensitive in AD cells, while other transcription factors (ETS and AP1) were more robust. 
-The drug enzalutamide had no effect on the top 2\% of sensitive species. 
-Species in the PI3K/AKT and MAPK pathways in the presence of enzalutamide were still highly sensitive. 
-The application of enzalutamide, increased sensitivity of AR species found outside of the nucleus as well as PAcP species.
-Robustness analysis indicated diverse effects of Raf knock-out on PSA and cyclin D concentrations. 
-Clinical studies of sorafenib, a multi-kinase inhibitor that has activity against Raf, showed increase PSA levels in patients \cite{Dahut2008}. 
-Our results indicate that cell-to cell heterogeneity in gene expression can play a significant role in determining cell response. 
-Thus, combination therapies need to be considered even in the case of a Raf knock-out.
-
-The results of the model suggest that an inhibition of either the PI3K pathway or the MAPK pathway in combination with an AR inhibitor as a possible therapy for CRPC. 
-Sensitivity analysis revealed no change in the top sensitive species in the presence or absence of the AR inhibitor, enzalutamide. 
-PI3K/AKT and MAPK species continued to fall in the top two percent of sensitive species. 
-A study by Carver \textit{et al.} looked at dual inhibition of AR and PI3K signaling in LNCaP cells and in a Pten-deficient murine prostate cancer model \cite{Parker2013}. 
-Using both the PI3K inhibitor, BEZ235, and the AR inhibitor, MDV3100 (enzalutamide), the group saw a drastic decrease in the total number of cells. 
-Each inhibitor on it’s own had a much smaller effect on total cell number. They saw an increase in the cell death marker, c-PARP, in the dual inhibition case. 
-The group hypothesized that AKT inhibition leads to increased AR signaling activity through increased protein concentrations of HER3. 
-On the other hand AR inhibition leads to increased AKT activity due to the down regulation of PHLPP, a protein phosphatase that regulates AKT. 
-For the simplicity of this model, the HER3 pathway and also cell death were not included in the model. 
-Dual knock-out studies of PI3K and AR in our model show no additive effect on any cell cycle proteins through the dual knock-out. 
-The pathways in our model appear to be uncoupled and therefore no synergy is shown in the dual knock-out case. 
-This could indicate that the combined decrease in cell population is entirely due to cell death. 
-The Carver \textit{et al.} study did not consider cell cycle proteins or cell growth. 
-Our model does show a decrease in cell cycle proteins in the PI3K knock-out as well as in the dual knock-out case. 
-This result seems to be consistent with the decreased cell growth in the PI3K knock-out case which is not dependent on cell death, as c-PARP levels are low. 
-The decrease in cell cycle proteins is due to a decrease in general translation, including free eIF4E levels and activated 40S ribosome subunit. 
-The decrease in p70 (S6) activation due to inhibition of PI3K is shown in both the model and in the Carver \textit{et al.} study ((Fig.
-\ref{fg:Supp_Figure_KO_new.pdf})), indicating this result is due to the PI3K pathway.
-Our results confirm the Carver \textit{et al.} study in that a dual inhibition of AR and PI3K signaling led to a more prominent decrease in cell viability then each of the inhibitors alone. 
-We extended the study to look at the addition of a third inhibitor, sorafenib, that inhibits Raf kinase in the MAPK pathway.
-
-%It would be interesting to repeat the experiment looking at cell cycle proteins and also performing the experiment in CR LNCaP cells, instead of AD cells.
-
-%and  in the CR CWR-R1 prostate cell line.
-%a steroid receptor coactivator family member along with other important AR coregulators such as 
-%transcriptional intermediary factor 2 (TIF-2)
-
-The PCa signaling architecture was assembled after extensive literature review and hand curation of the biochemical interactions.
-However, there are a number of areas where model connectivity could be refined, e.g., the regulation of AR phosphorylation.
-We assumed a single canonical activating AR phosphorylation site (S515), with ERK being the major kinase and PP2A or PP1 being the major phosphatases responsible for regulating this site.  
-MAPK activation following EGF treatment increases AR transcription and cell growth, partially through AR phosphorylation on MAPK consensus site S515 \cite{Ponguta2008}.
-However, there are at least 13 phosphorylation sites identified on AR, with phosphorylation at six of these being androgen induced \cite{Gioeli2012}. 
-Moreover, other kinases such as AKT, protein kinase C (PKC) family members, as well as Src-family kinases can all phosphorylate AR in prostate cells \cite{Guo2006,Ponguta2008}. 
-For example, AKT activation leads to AR  phosphorylation at both S213 and S791 (however, the role of these sites remains unclear) \cite{Wen2000,Lin2001PNAS,Taneja2005,Lin2003}.  
-AKT effects on AR may also be passage number dependent, with AKT repressing AR transcription in low passage number cells and enhancing transcription in higher passage numbers \cite{Lin2003}. 
-Androgen independent phosphorylation of AR by Src family kinases (not currently in model) at Y534 \cite{Guo2006} or by protein kinase C (PKC) family members at the consensus site S578 could also be important for understanding the regulation of AR activity.
-A second area we will revisit is the gene expression program associated with androgen action, and particularly the role of AR coregulators.
-Currently, we included only two AR coactivators, cyclin E and CDK6 \cite{Yamamoto2000,Lim2005} and three corepressors AP1, Cdc25A, and cyclin D1a in the model \cite{Sato1997,Chiu2009,Petre-Draviam2003}.
-However, there are at least 169 proteins classified as potential AR coregulators \cite{Heinlein:2002qy,Heemers2007} with many of these being differentially expressed in malignant cells.
-For example, the expression of steroid receptor coactivator-1 (Src-1) and transcriptional intermediary factor 2 (Tif-2), both members of the steroid receptor coactivator family, are elevated in prostate cancer \cite{Gregory2001,Gregory2004}.
-Src-1 is phosphorylated by MAPK and interacts directly with AR to enhance AR-mediated transcription \cite{Heemers2007}. 
-Another class of potentially important AR coregulators are the cell cycle proteins Cdc25 and Rb.
-Unlike Cdc25A, Cdc25B (not in the model) can act as an AR coactivator leading to enhanced AR transcription activity \cite{Ngan2003}. 
-The Rb protein, in addition to being a key cell cycle regulator, has been shown to be an AR coactivator in an androgen-independent manner in DU145 cells \cite{Yeh1998}.
-However, there is some uncertainty about the role of Rb as Sharma \textit{et al.} showed that Rb decreased AR activation in multiple prostate cancer cell lines and xenografts \cite{Sharma2010}. 
-Forkhead proteins have also been shown to activate as well as repress AR function. 
-In prostate cancer, AKT suppresses AFX/Forkhead proteins, which diminishes expression of AFX target genes, such as p27Kip1 \cite{Graff2000, Brunet1999, Medema2000, Takaishi1999}. 
-Lastly, undoubtedly there are several other signaling axes important in PCa, such as cytokine or insulin- and insulin-like growth factor signaling \citep{Cardillo2003,Heinlein2004,Tam2007,Seaton2008}.
-Understanding the pathways associated with these signals and how they relate to the current model, may give us a more complete picture of CR prostate cancer.
-
-%Lastly, with an extended model, it would be beneficial to include possible biomarkers for drug outcomes, as currently there is no biomarker that is used to select treatments for patients  \cite{Sartor2013}.
+\begin{table}[h]
+\centering
+\caption{Table with optimization settings and results for coagulation and different benchmarks using DdPSO.}
+\label{table:benchmark-problems}
+\resizebox{\columnwidth}{!}{
+\begin{tabular}{|cccccc|}
+\hline
+                        & Coagulation           & \multicolumn{4}{l}{Benchmarks}\\
+ \hline
+                        & Experimental Data Set & B1           & B4          & Ackley 300D & Rastrigin 300D \\
+ \hline
+Lower Bound             & 0.001.pnom            & 5.pnom       & 5.pnom      & 30          & 5.12           \\
+Upper Bound             & 1000.pnom             & 0.2.pnom     & 0.2.pnom    & -15         & -5.12          \\
+CPU Time                & 10.0835 hours         & 38.308 hours & 6.2 minutes & 2.8 seconds & 2.6 seconds    \\
+Function Evaluations    & 4000                  & 4000         & 4000        & 4000        & 4000           \\
+Initial Objective Value & 1.11837e+07           & 1.4275e+07   & 1.8536e+07  & 21.12       & 99.985         \\
+Final Objective Value   & 90456.9               & 4.96e+05     & 38.9375     & 0           & 0              \\
+Nominal Objective Value & 4.7785e+06            & 1.0986e+06   & 39.0676     & 0           & 0      \\
+\hline
+\end{tabular}
+}
+\end{table}
 
@@
 
-\section*{Materials and Methods}
-
-\subsection*{Prostate model signaling architecture.}
-We modeled the expression, translation and post-translational modifications of key components of the signaling architecture.  
-The model, which consisted of 780 protein, lipid or mRNA species interconnected by 1674 interactions, was a significant extension to our previous model \cite{Tasseff2010} in several important areas. 
-First, we included well-mixed nuclear, cytosolic, membrane and extracellular compartments (including transfer terms between compartments).
-Next, we expanded the description of growth factor receptor signaling, considering both homo- and heterodimer formation between ErbB family members and the role of cellular and secreted prostatic acid phosphatase (cPAcP and sPAcP, respectively).
-Both forms of PAcP were included because cPAcP downregulates HER2 activity, while sPAcP promotes modest HER2 activation \cite{Veeramani2005}. 
-Third, we expanded the description of the G1/S transition of the cell cycle (restriction point). 
-The previous model used the abundance of cyclin D as a proliferation marker, but did not include other proteins or interactions potentially important to the restriction point.
-Toward this shortcoming, we included cyclin E expression (and its role as a coregulator of androgen receptor expression), enhanced the description of cyclin D expression and the alternative splicing of cyclin D mRNA (including the role of the splice variants in androgen action), included the Rb/E2F pathway as well as E2F inhibition of androgen receptor expression \cite{Davis2006}, and the cyclin-dependent kinases cyclin-dependent kinase 4 (CDK4) and cyclin-dependent kinase 6 (CDK6).
-We also included key inhibitors of the restriction point including cyclin-dependent kinase inhibitor 1 (p21Cip1), cyclin-dependent kinase inhibitor 1B (p27Kip1), and cyclin-dependent kinase inhibitor 2A (p16INK4) \cite{Sherr1999}.
-Fourth, we enhanced the description of growth factor induced translation initiation.
-One of the key findings of the previous model was that growth factor induced translation initiation was globally sensitive (important in both androgen dependent and independent conditions). 
-However, the description of this important subsystem was simplified in the previous model. 
-Here, we expanded this subsystem, using connectivity similar to previous study of Lequieu \textit{et al.} \cite{Lequieu:2011eu}, and re-examined the importance of key components of this axis, such as mammalian target of rapamycin (mTOR), phosphatidylinositide 3-kinase (PI3K) and AKT.
-Lastly, we significantly expanded the description of the role of androgen receptor. 
-The previous model assumed constant AR expression, consistent with studies in androgen dependent and independent LNCaP sublines \cite{Lee2003}.
-However, other prostate cancer cell lines vary in their AR expression \cite{Sobel2005}. 
-Thus, to capture androgen signaling in a variety of prostate cancer cells, we included the transcriptional regulation governing androgen receptor expression, updated our description of the regulation of androgen receptor activity and androgen action (gene expression program driven by activated androgen receptor).
-At the expression level, we included AR auto-regulation in combination with the co-activators cyclin E and CDK6 \cite{Yamamoto2000,Lim2005}. 
-We also assumed androgen receptor could be activated through androgen binding or a ligand-independent, MAPK-driven mechanism referred to as the outlaw pathway \cite{Feldman2001,Yeh1999}. 
-We assumed a single canonical activating AR phosphorylation site (S515), with phosphorylated extracellular-signal-regulated kinase 1/2 (ppERK1/2) being the major kinase and protein phosphatase 2 (PP2A) or phosphoprotein phosphatase 1 (PP1) being the major phosphatases responsible for regulating this site.
-Finally, we modeled androgen receptor induced gene expression, including prostate specific antigen (PSA), cPAcP and p21Cip1.
-
-%This description includes  were added to the model because of their importance in co-regulating AR transcription regulation activity in prostate cells . 
-%With the addition of cell cycle proteins, we also included 
-%only considered cyclin D as a cell marker and 
-%In both models we considered nuclear hormone (AR) and transmembrane growth factor receptor (homo- and heterodimers of ErbB1/2) activation through both ligand-dependent and independent mechanisms. 
-%The model investigated the activation of AR through androgen binding or a ligand-independent, MAPK-driven mechanism. 
-%In addition to MAPK-dependent outlaw activation of AR, the model also included transcriptional activation via MAPK signaling \cite{Feldman2001,Yeh1999}. 
-%The PI3K/AKT/mTOR pathway for translation initiation was also included in the model, as well as the regulated expression of PSA, cyclin D,  
-%Thus, 
-%The model of Tasseff \textit{et al.}  
-%In order 
-%Additionally, expression of transcription factors AP-1 and ETS were also added to the model. 
-%Another significant addition to the model was the inclusion of a nuclear compartment separate from the cytosol, membrane and extracellular environments. 
-%The prostate model was developed to determine the shift in important signaling components between CR and AD prostate cancer cells.
-
-\subsection*{Formulation and solution of the model equations.}
-The prostate model was formulated as a coupled set of non-linear ordinary differential equations (ODEs):
-\begin{equation}\label{asseassertain_massbalance}
-\frac{d\mathbf{x}}{dt}=\mathbf{S}\cdot\mathbf{r}\left(\mathbf{x},\mathbf{k}\right)\qquad\mathbf{x}\left(t_{o}\right)=\mathbf{x}_{o}
-\end{equation}
-The quantity $\mathbf{x}$ denotes the vector describing the abundance of protein, mRNA, and other species in the model ($780\times 1$).
-The stoichiometric matrix $\mathbf{S}$ encodes the signaling architecture considered in the model ($780\times 1674$).  
-Each row of $\mathbf{S}$ describes a signaling component while each column describes a particular interaction.
-The $(i,j)$ element of $\mathbf{S}$, denoted by $\sigma_{ij}$, describes how species $i$ is involved with interaction $j$. 
-If $\sigma_{ij}>0$, species $i$ is produced by interaction $j$. Conversely, If $\sigma_{ij}<0$, then species $i$ is consumed in interaction $j$. 
-Lastly, if $\sigma_{ij}=0$, then species $i$ is not involved in interaction $j$.
-The term  $\mathbf{r}\left(\mathbf{x},\mathbf{k}\right)$ denotes the vector of interactions rates ($1674\times{1}$).
-Gene expression and translation processes as well as all biochemical transformations were decomposed into simple elementary steps, 
-where all reversible interactions were split into two irreversible steps (supplemental materials). 
-We modeled each network interaction using elementary rate laws where all reversible interactions were split into two irreversible steps. 
-Thus, the rate expression for interaction $q$ was given by:
-\begin{equation}\label{eq-mass-action}
-r_{q}\left(\mathbf{x},k_{q}\right)=k_{q}\prod_{j\in\left\{\mathbf{R}_{q}\right\}}x_{j}^{-\sigma_{jq}}
-\end{equation}
-The set $\left\{\mathbf{R}_{q}\right\}$ denotes reactants for reaction $q$, while $\sigma_{jq}$ denotes the stoichiometric coefficient (element of the matrix $\mathbf{S}$) governing species $j$ in reaction $q$.
-The quantity $k_{q}$ denotes the rate constant (unknown) governing reaction $q$.
-Model equations were generated in the C-programming language using the UNIVERSAL code generator, starting from an text-based input file (available in supplemental materials). 
-UNIVERSAL, an open source Objective-C/Java code generator, is freely available as a Google Code project (http://code.google.com/p/universal-code-generator/). 
-Model equations were solved using the CVODE solver in the SUNDIALS library \cite{Hindmarsh2005} on an Apple workstation (Apple, Cupertino, CA; OS X v10.6.8).
-
-We ran the model to steady-state before calculating the response to DHT or growth factor inputs.
-The steady-state was estimated numerically by repeatedly solving the model equations and estimating the difference between subsequent time points:
-\begin{equation}
-\Vert\mathbf{x}\left(t+\Delta{t}\right) - \mathbf{x}\left(t\right)\Vert_{2}\leq\gamma
-\end{equation}
-The quantities $\mathbf{x}\left(t\right)$ and $\mathbf{x}\left(t+\Delta{t}\right)$ denote the simulated abundance vector at time $t$ and $t + \Delta{t}$, respectively.
-The $L_{2}$ vector-norm was used as the distance metric, where $\Delta{t} = 100$ hr of simulated time and $\gamma$ = 0.001 for all simulations.
-
-We estimated an ensemble of model parameter sets using the Pareto Optimal Ensemble Techniques (POETs) multiobjective optimization routine \cite{Song:2009dq,Song2010,Lequieu:2011eu}. 
-POETs minimized the residual between model simulations and 43 separate training objectives taken from protein and mRNA signaling data generated in androgen dependent, intermediate and independent LNCaP cell lines (Table \ref{objective_table}).
-From these training objectives, POETs generated $>10^6$ candidate parameter vectors from which we selected N = 5000 Pareto rank-zero vectors for further analysis. 
-The set-to-set correlation between selected sets was approximately 0.60, suggesting only modest similarity between ensemble members.
-Approximately 33\%, or 560 of the 1674 parameters had a coefficient of variation (CV) of less than 1.0, where the CV ranged from 0.59 to 5.8 over the ensemble.
-Details of the parameter estimation problem and POETs are given in the supplemental materials.
-
-\subsection*{Sensitivity and robustness analysis.}
-Steady-state sensitivity coefficients were calculated for N = 500 parameter sets selected from the ensemble by solving the augmented kinetic-sensitivity equations \cite{Dickinson1976}:
-\begin{equation}
-	\begin{bmatrix}
-		\mathbf{S}\cdot\mathbf{r}(\mathbf{x},\mathbf{k}) \\
-		\mathbf{A}(t_s)\mathbf{s}_j + \mathbf{b}_j(t_s)
-	\end{bmatrix} = \begin{pmatrix}
-		\mathbf{0} \\
-		\mathbf{0}
-	\end{pmatrix}
-	   \qquad j = 1,2,\dots,\mathcal{P}
-\end{equation}
-where
-\begin{equation}
-s_{ij}(t_s)= \left. \frac{\partial{x_i}}{\partial{k_j}} \right|_{t_s}
-\end{equation} for each parameter set.
-Steady-state was calculated as described previously.
-The quantity $j$ denotes the parameter index, $\mathbf{A}$ denotes the Jacobian matrix, and $\mathcal{P}$ denotes the number of parameters in the model. 
-The vector $\mathbf{b}_j$ denotes the $j$th column of the matrix of first-derivatives of the mass balances with respect to the parameters. 
-Steady-state sensitivity coefficients were used because of the computational burden associated with sampling several hundred parameters sets for each of the 1674 parameters. 
-The steady-state sensitivity coefficients $\mathcal{N}_{ij} \equiv s_{ij}$ were organized into an array for each parameter set in the ensemble: 
-\begin{equation}
-	\mathcal{N}^{\left(\epsilon\right)} = 
-	\begin{pmatrix}
-		\mathcal{N}_{11}^{\left(\epsilon\right)} & \mathcal{N}_{12}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{1j}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{1P}^{\left(\epsilon\right)} \\
-		\mathcal{N}_{21}^{\left(\epsilon\right)} & \mathcal{N}_{22}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{2j}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{2P}^{\left(\epsilon\right)} \\
-		\vdots & \vdots & & \vdots & & \vdots \\
-		\mathcal{N}_{M1}^{\left(\epsilon\right)} & \mathcal{N}_{M2}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{Mj}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{MP}^{\left(\epsilon\right)} \\
-	\end{pmatrix}\qquad\epsilon = 1,2,\hdots,N_{\epsilon}
-\end{equation}
-where $\epsilon$ denotes the index of the ensemble member, $P$ denotes the number of parameters, $N_{\epsilon}$ denotes the number of parameter sets sampled (N = 500) and $M$ denotes the number of model species.
-To estimate the relative fragility or robustness of species and reactions in the network, we decomposed $\mathcal{N}^{\left(\epsilon\right)}$ 
-using Singular Value Decomposition (SVD):
-
-\begin{equation}
-	\mathcal{N}^{\left(\epsilon\right)} = \mathbf{U}^{\left(\epsilon\right)}\Sigma^{\left(\epsilon\right)}\mathbf{V}^{T,{\left(\epsilon\right)}}
-\end{equation}
-Coefficients of the left singular vectors corresponding to largest $\theta$ $\leq$ 15 singular values of $\mathcal{N}^{\left(\epsilon\right)}$ were rank-ordered to estimate important species combinations, while coefficients of the right singular vectors were used to rank important reaction combinations.
-Only coefficients with magnitude greater than a threshold ($\delta$ = 0.001) were considered. 
-The fraction of the $\theta$ vectors in which a reaction or species index occurred was used to quantify its importance (sensitivity ranking). 
-We compared the sensitivity ranking between different conditions to understand how control in the network shifted in different cellular environments.
-
-Robustness coefficients were calculated as shown previously \cite{Tasseff2011}. 
-Robustness coefficients denoted by $\alpha\left(i,j,t_{o},t_{f}\right)$ are defined as:
-\begin{equation}
-\alpha\left(i,j,t_{o},t_{f}\right)=\left(\displaystyle\int_{t_{o}}^{t_{f}}x_{i}\left(t\right)dt\right)^{-1}\left(\displaystyle\int_{t_{o}}^{t_{f}}x^{(j)}_{i}\left(t\right)dt\right)
-\end{equation}
-Robustness coefficients quantify the response of a marker to a structural or operational perturbation to the network architecture.
-Here $t_o$ and $t_f$ denote the initial and final simulation time respectively, while $i$ and $j$ denote the indices for the marker and the perturbation respectively. 
-A value of $\alpha\left(i,j,t_{o},t_{f}\right)>1$, indicates increased marker abundance, while $\alpha\left(i,j,t_{o},t_{f}\right)<1$ indicates decreased marker abundance following perturbation $j$. 
-If $\alpha\left(i,j,t_{o},t_{f}\right)\sim{1}$ the $j$th perturbation does not influence the abundance of marker $i$.
-Robustness coefficients were calculated (starting from steady-state) from $t_{o} = 0$ hr to $t_{f} = 72$ hr following the addition of 10nM DHT at $t_{o}$.
-Robustness coefficients were calculated for the same N = 500 models selected for sensitivity analysis.
-
-\subsection*{Experimental Validation.}
-
-\subsection*{Cell culture and treatments}
-Androgen dependent LNCaP prostate cancer cells were a gift from Dr.
-Brian Kirby (Cornell University), and the Castrate Resistant C4-2 prostate cancer cell line was purchased from MD Anderson Cancer Center, University of Texas.
-Cell lines were maintained in RPMI 1640 media (Life Technologies, Inc., Grand Island, NY) with 10\% fetal calf serum (FBS; Hyclone) and 1x antibiotic/antimycotic (Sigma, St.
-Louis, MO) in a 5\% CO$_2$ humidified atmosphere at 37$^{\circ}$C.
-The AR inhibitor MDV3100 (enzalutamide) and the Raf inhibitor sorafenib were purchased from SantaCruz Biotechnology (Santa Cruz, CA).
-The PI3K inhibitor LY294002 was purchased from Cell Signaling Technologies (Danvers, MA, USA).
-All stock solutions were diluted in DMSO (Sigma, St.
-Louis, MO).
-
-\subsection*{Protein Extraction and Western Blot}
-
-LNCaP and C4-2 cells were seeded in 60 mm dishes at a density of 4 x 10$^{5}$.
-After 96 and 72 hrs, for LNCaP and C4-2 cells respectively, the media was replaced with fresh media and drug treatments were added.
-After 24 hours, cells were washed twice in PBS buffer, scraped in 250 $\mu$L ice-cold lysis buffer (Pierce, Rockford, IL) supplemented with protease and phosphatase inhibitors (Sigma, St.
-Louis, MO), and lysed for 30 min on ice.
-Lysates were centrifuged at 13,000 rpm for 30 min at 4$^{\circ}$C.
-After quantification of total protein by BCA assay, equal amounts of total protein lysates (25 $\mu$g) were resolved by SDS-PAGE and transferred onto PVDF membranes.
-Membranes were blocked in 5\% fat free milk and then probed with antibodies.
-The primary antibodies used for western blot analysis were pAKT Ser473, AKT, pS6 Ser240/244 , pERK Thr202/Tyr204, ERK, AR, and GAPDH were from Cell Signaling Technologies (Danvers, MA, USA).
-For detection, enhanced chemiluminescence ECL reagent (GE Healthcare, Pittsburgh, PA) was used and signals were visualized using the ChemiDoc XRS system (Bio-Rad).
-
-\subsection*{MTT Assay}
-
-LNCaP and C4-2 cells were seeded at a density of 1x10$^{4}$ cells per well in 96 well plates.
-After 48 hrs the media was refreshed and drug treatments added.
-Cell growth at 24, 48, and 72 hrs was determined using a 3-(4,5-dimethyl thiazol-2-yl)-2,5-diphenyl tetrazolium bromide (MTT) assay.
-At the specified time point 10 $\mu$L MTT reagent (stock of 5 mg/mL in PBS) was added to each well and the cells were further incubated for 4 hrs.
-At 4 hrs, the media was removed and 50 $\mu$L of dissolving reagent DMSO was added to each well.
-After an additional 10 min incubation, the absorbance was measured at 540 nm on a microplate reader.
-Each reading was adjusted by subtracting the absorbance value for the blank (media only) and the results were then scaled to the DMSO-treated (control) case.
-
-\section*{Acknowledgements}
-This study was supported by award number U54CA143876 from the National Cancer Institute.
-The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Cancer Institute or the National Institutes of Health.
-The authors acknowledge the financial support of the National Science Foundation, award no.
-DGE-1045513,for a GK12 training grant entitled "Grass Roots: Advancing Education in Renewable Energy and Cleaner Fuels through collaborative graduate fellow/teacher/grade-school student interactions."
-We would like to thank Timothy Chu, Kyriakos Tsahalis, Spencer Davis, Rohit Yallamraju, Gongcheng Lu, and Yalin Zhao.
-
-\clearpage
-%\bibliographystyle{plain}
-%\bibliographystyle{IEEEbib}
-
-\bibliography{References_v1}
-
-%\begin{comment}
-
-\clearpage
-
-\begin{figure}\centering
-\includegraphics[width=1.0\textwidth]{./figs/Fig_1_LNCAPModel.pdf}
-\caption{Schematic overview of the prostate signaling network.
-The model describes hormone and growth factor induced expression of several proteins, including PSA.
-In the absence of outside hormones/growth factors, overactive HER2 can stimulate the MAPK and AKT pathways.
-AR can be activated directly by the MAPK pathway.
-}
-\label{fg:NetworkSchematic}
+\begin{figure}[h]
+\centering
+\includegraphics[width=1.0\textwidth,height=0.5\textheight]{./figs/Figure_1_Algorithm.pdf}
+\caption{Multi Swarm Particle Swarm Optimization with Dynamically Dimensioned Search.
+\textbf{A}: Each particle represents an N dimensional parameter vector.
+Particles are randomly initialized and grouped into different sub-swarms.
+Within each swarm the magnitude and direction of the movement a particle is influenced by the position of the best particle and also by its own experience.
+After every $\mathbf{g}$ number of function evaluations the particles are mixed and randomly assigned to different swarms.
+At the end of the evaluations ($\mathcal{FR}$) assigned to swarm search, the global best particle (orange color) amongst all sub-swarms is chosen as the candidate parameter vector for Dynamically Dimensioned Search \textbf{B}: The candidate vector does a greedy search in a dynamic neighborhood.
+The search neighborhood is dynamically adjusted by varying the number of dimensions that are perturbed (in black) in each evaluation step.
+The probability that a dimension is perturbed decreases as the number of function evaluations increase.
+Thus as the evaluations increase the optimality of the solution is preserved.
+}\label{fig-algorithm}
 \end{figure}
@@
 
-\begin{figure}\centering
-\includegraphics[width=1.0\textwidth]{./figs/Fig_3_ColorPlot.pdf}
-\caption{Simulation results versus experimental results for training and validation data.
-Experiment numbers 1 through 43 were used for training, while experiments 44 through 72 were validation.
-Gray means the ensemble member qualitatively fit experimental data in both models.
-White means the the ensemble member only fit the data using the new model that included HER2 heterodimerization.
-Red means the ensemble member fit using only the old model.
-Black corresponds to an incorrect cellular response in both models.
-$\bf{A.,C.}$ Training and validation results, respectively, for entire ensemble population using both the original model and an updated model including HER2 heterodimerization (N = 5000).
-$\bf{B.,D.}$ Simulation results for training and validation of a random set of 100 members using both models.
- }
-\label{fg:ColorPlot}
+\begin{figure}[h]
+\centering
+\includegraphics[width=1.00\textwidth,height=0.7\textheight]{./figs/Figure_2_CoagulationNetwork.pdf}
+\caption{Schematic of the extrinsic and intrinsic coagulation cascade.
+Inactive zymogens upstream (grey) are activated by exposure to tissue factor (TF)  following vessel injury.
+Tissue factor and activated factor VIIa (FVIIa) form a complex that activates factor X (fX) and IX (fIX).
+FXa activates downstream factors including factor VIII (fVIII) and fIX.
+Factor V (fV) is primarily activated by thrombin (FIIa).
+In addition, we included a secondary fV activation route involving FXa.
+FXa and FVa form a complex (prothrombinase) on activated platelets that converts prothrombin (fII) to FIIa.
+FIXa and FVIIIa can also form a complex (tenase) on activated platelets which catalyzes FXa formation.
+Thrombin also activates upstream coagulation factors, forming a strong positive feedback ensuring rapid activation.
+Tissue factor pathway inhibitor (TFPI) downregulates FXa formation and activity by sequestering free FXa and TF-FVIIa in a FXa-dependent manner.
+Antithrombin III (ATIII)  inhibits all proteases.
+Thrombin inhibits itself binding the surface protein thrombomodulin (TM).
+The IIa-TM complex catalyzes the conversion of protein C (PC) to activated protein C (APC), which attenuates the coagulation response by the proteolytic cleavage of fV/FVa and fVIII/FVIIIa.
+}\label{fig-coagulation-network}
 \end{figure}
@@
 
-\begin{figure}\centering
-\includegraphics[width=1.0\textwidth]{./figs/Fig_2_Training.pdf}
-\caption{Ensemble performance against selected training objectives (N = 5000).
-A, B.
-Time course data for PSA concentration due to a stimulus of 10 nM DHT in LNCaP C33 cells and LNCaP C81 cells, respectively (O2, O3).
-C.
-PSA levels in the presence and absence of a HER2 inhibitor (LNCaP C81 cells, O7).
-D.
-PAcP mRNA levels at 72 hours in the presence and absence of DHT (LNCaP C51 cells, O14).
-E.
-Steady-state cyclin D levels in LNCaP C33 vs.
-C81 (O17).
-F.
-p21Cip1 levels at 48 hrs in the presence and absence of DHT (LNCaP C33, O25).
-G.
-Steady-state AKT phosphorylation levels in LNCaP C33 vs.
-C51 (O30).
-H.
-AR levels at 24 hours in the presence and absence of DHT (LNCaP C51, O31).
-I.
-AR mRNA levels in the presence and absence of E2F over expression (LNCaP C33, O34).}
-\label{fg:Training}
+\begin{figure}[h]
+\centering
+\includegraphics[width=1.0\textwidth,height=0.5\textheight]{./figs/Figure_3_Errors_convergence.pdf}
+\caption{Error convergence rates of the 5 different algorithms on the coagulation model.
+The objective error is the mean over $N$= 25 trials.
+DDSMLSPSO and DDS have the steepest drop in error during first 2000 function evaluations.
+Thereafter the error drop in DDS remains nearly constant whereas DDSMLSPSO drops further from around 2000 to 4000 function evaluations.
+At the end of 4000 function evaluations DDSMLSPSO attains the lowest error.
+}\label{fig-train}
 \end{figure}
@@
 
-\begin{figure}\centering
-\includegraphics[width=1.0\textwidth]{./figs/Fig_4_Validation.pdf}
-\caption{Blind model predictions for the ensemble (N = 5000).
-The model ensemble’s predictive ability was assessed by comparing simulation versus experimental data not used for training.
-A.
-Time course data for HER phosphorylation due to a stimulus of 10 nM DHT (LNCaP C33, P1).
-B.
-ERK phosphorylation levels in the presence and absence of a PAcP inhibitor (LNCaP C33 cells, P3).
-C.
-AR levels at 24 hrs in varying levels of DHT (LNCaP C33, P17).
-D.
-Shc phosphorylation levels at 24 hrs in the presence and absence of DHT (LNCaP C33, P22).
-E.
-Time course data for HER phosphorylation due to a stimulus of 1.6 nM EGF (LNCaP C33, P7).
-F.
-AR levels in varying levels of EGF (LNCaP C33, P14).
-G, H, I.
-Fold change in PSA concentration due to drug stimulus: enzalutamide, lapatinib, and sorafenib.}
-\label{fg:Validation}
+\begin{figure}[h]
+\centering
+\includegraphics[width=1.0\textwidth,height=0.5\textheight]{./figs/Figure_4_Sim_Train_E1_E5.pdf}
+\caption{Model fits on experimental data using DDSMLSPSO.
+The model parameters were estimated using DDSMLSPSO.
+Solid black lines indicate the simulated mean thrombin concentration using parameter vectors from $N$= 25 trials.
+The grey shaded region represents the 99\% confidence estimate of the mean simulated thrombin concentration.
+The experimental data is reproduced from the synthetic plasma assays of Mann and co-workers.
+Thrombin generation is initiated by adding Factor VIIa-TF (5nM - Red and 5pM - Green) to synthetic plasma containing 200 $\mu$mol/L of phospholipid vesicles (PCPS) and a mixture of coagulation factors (II,V,VII,VIII,IX,X and XI) at their mean plasma concentrations.
+}\label{fig-train}
 \end{figure}
@@
 
-\begin{figure}\centering
-\includegraphics[width=1.0\textwidth]{./figs/Fig_5_Sensitivity.pdf}
-\caption{Sensitivity analysis of a population of prostate models (N = 500).
-Species with a low sensitivity are considered robust, while species with a high sensitivity ranking are considered fragile.
-$\bf{A,B.}$ Sensitivity ranking of network species in AD versus CR cells in the absence (presence) of DHT.
-$\bf{C.}$ Sensitivity ranking of network species in AD cells in the absence and presence of DHT.
-$\bf{D.}$ Sensitivity ranking of network species in CR cells in the presence and absence of enzalutamide with a DHT stimulus.}
-\label{fg:Sensitivity}
+\begin{figure}[h]
+\centering
+\includegraphics[width=1.0\textwidth,height=0.5\textheight]{./figs/Figure_5_Sim_Validate_E2_E4_E6.pdf}
+\caption{Model predictions on unseen experimental data using parameters obtained from DDSMLSPSO.
+The parameter estimates that were obtained using DDSMLSPSO were tested against data that was not used in the model training.
+Solid black lines indicate the simulated mean thrombin concentration using parameter vectors from $N$= 25 trials.
+The grey shaded region represents the 99\% confidence estimate of the mean simulated thrombin concentration.
+The experimental data is reproduced from the synthetic plasma assays of Mann and co-workers.
+Thrombin generation is initiated by adding Factor VIIa-TF (500pM - Blue, 50pM - Pink and 10pM - Orange respectively) to synthetic plasma containing 200 $\mu$mol/L of phospholipid vesicles (PCPS) and a mixture of coagulation factors (II,V,VII,VIII,IX,X and XI) at their mean plasma concentrations.
+}\label{fig-validation}
 \end{figure}
 
-\clearpage
+%\bibitem[Saltelli \em{et~al.}(2010)Saltelli, Annoni, Azzini, Campolongo, Ratto,
+  %and Tarantola]{Saltelli:2010}
+%Saltelli, A.; Annoni, P.; Azzini, I.; Campolongo, F.; Ratto, M.; Tarantola, S.
+%\newblock Variance based sensitivity analysis of model output. Design and
+  %estimator for the total sensitivity index.
+%\newblock {\em Comput. Phys. Commun.} {\bf 2010}, {\em 181},~259--270.
 
-\begin{figure}\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure6_RAF_K0_C81.pdf}
-\caption{Robustness analysis of a population of CR prostate models with Raf knock-out (N = 500).
-A log fold change of greater then zero implies that the concentration of the protein increased with the knock-out of Raf, while a log fold change of less then zero indicates that the concentration of protein decreased.
-A log of fold change equal to 0, shows no response due to Raf knock-out.
-Three distinct regions emerge in Raf knock-out case: (1) PSA increases, (2) cyclin D concentration increases, and (3) PSA and cyclin D concentration decrease.
-}
-\label{fg:Robustness}
-\end{figure}
-
-\clearpage
-
-\begin{figure}\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure7_ExperimentalResults.pdf}
-\caption{Experimental results for multiple drug combinations on two prostate cancer cell lines, LNCaP and C4-2.
-A.
-Western blot analysis of AR, pS6, AKT, pAKT and ERK in LNCaP and C4-2 cell lines treated for 24 hrs with DMSO (control), sorafenib (5 $\mu$M), LY294002 (40 $\mu$M), and MDV3100 (10 $\mu$M) alone or in combination (at least 3 repeats).
-B.
-Cells (LNCaP and C4-2) were treated for 24, 48 and 72 hrs with sorafenib (5 $\mu$M), LY294002 (40 $\mu$M), and MDV3100 (10 $\mu$M) and cell viability was measured using MTT Assay.
-Values were normalized to DMSO (control).
-C.
-Cell viability results for LNCaP and C4-2 cells at varying concentration of sorafenib, LY294002, and MDV3100 after 24 hrs of treatment.
-Values were normalized to DMSO (control).
-Error bars represent standard error (at least 3 repeats with triplicates performed in each experiment).}
-\label{fg:Experiment}
-\end{figure}
+%\end{thebibliography}
 
@@
 
-\section*{Supplementary materials}
-
-\subsection*{Estimation of a population of models using Pareto Optimal Ensemble Techniques (POETs).}
-
-We used multiobjective optimization to estimate an ensemble of prostate models.
-Although computationally more complex than single-objective formulations, multiobjective optimization can be used to address qualitative conflicts in training data arising from experimental error or cell-line artifacts \cite{Handl2007}.
-In this study we used the Pareto Optimal Ensemble Technique (POETs) to perform the optimization.
-POETs integrates standard search strategies, e.g., Simulated Annealing (SA) or Local Pattern Search (PS) with a Pareto-rank fitness assignment \cite{Song2010}.
-The mean squared error, $\eta$, of parameter set $\mathbf{k}$ for training objective $j$ was defined as:
-
-\begin{equation}
-\eta_j(\mathbf{p}_k) = \frac{1}{N}\sum_{i}^{N}\frac{(\hat{x}_{i,j} - \beta_j x(\mathbf{p_k})_{i,j})^2}{\hat{\sigma}^2_{i,j}}
-\end{equation}
-The symbol $\hat{x}_{i,j}$ denotes scaled experimental observations (from training objective j) while $x(\mathbf{p_k})_{i,j}$ denotes the simulation output (from training objective j).
-The quantity $i$ denotes the sampled time-index or condition, and $N$ denotes the number of time points or conditions for experiment j.
-The standard deviation, $\hat{\sigma}_{i,j}$, was assumed to be equal to 10\% of the reported observation, if no experimental error was reported.
-$\beta_j$ is a scaling factor which is required when considering experimental data that is accurate only to a multiplicative constant.
-In this study, the experimental data used for training and validation was typically band intensity from immunoblots, where intensity was estimated using the ImageJ software package \cite{Abramoff2004}.
-The scaling factor used was chosen to minimize the normalized squared error \cite{Brown2003}:
-
-\begin{equation}
-\beta_j = \frac{\sum_{i}(\hat{x}_{i,j}x_{i,j}/\hat{\sigma}_{i,j}^2)}{\sum_{i}(x_{i,j}/\hat{\sigma}_{i,j})^2}
-\end{equation} 
-By using the scaling factor, the concentration units on simulation results were arbitrary, which was consistent with the arbitrary units on the experimental training data.
-All simulation data was scaled by the corresponding $\beta_j$.
-
-We computed the Pareto rank of parameter set $\mathbf{k}_{i+1}$ by comparing the simulation error at iteration $i+1$ against the simulation archive, denoted as $\mathbf{K}_i$.
-We used the Fonseca and Fleming ranking scheme \cite{Fonseca1993} to estimate the rank of the parameter set $\mathbf{k}_{i+1}$.
-Parameter sets with increasing rank are progressively further away from the optimal trade-off surface.
-The parameter set $\mathbf{k}_{i+1}$ was accepted or rejected by the SA with probability $\mathcal{P}\left(\mathbf{k}_{i+1}\right)$:
-\begin{equation}\label{eqn_costMOSA}
-\mathcal{P}(\mathbf{k}_{i+1}) \equiv \exp{\{-rank\left(\mathbf{k}_{i+1} \mid \mathbf{K}_{i} \right)/T\}}
-\end{equation}
-where $T$ is the computational annealing temperature.
-The Pareto rank for $\mathbf{k}_{i+1}$ is denoted by $rank\left(\mathbf{k}_{i+1}\mid \mathbf{K}_{i}\right)$.
-The annealing temperature was  adjusted according to the schedule
-$T_k = \beta^k T_0$
-where $\beta$ was defined as $\beta = \left(\frac{T_{f}}{T_{o}}\right)^{1/10}$.
-The initial temperature was given by $T_0 = n/log(2)$, with $n=4$ and the final temperature $T_f = 0.1$ used in this study.
-The epoch-counter $k$ was incremented after the addition of 50 members to the ensemble.
-As the ensemble grew, the likelihood of accepting a high rank set decreased.
-Parameter sets were generated by applying a random perturbation in log space:
-
-\begin{equation}
-\log\mathbf{k}_{i+1} = \log\mathbf{k}_i + \mathcal{N}(0,\nu)
-\end{equation}
-where $\mathcal{N}(0,\nu)$ is a normally distributed random number with zero mean and variance $\nu$, set as 0.1 in this model.
-The perturbation was applied in log space to account for large variation in parameter scales and to ensure positive parameter values.
-We used a local pattern search every $q$ steps, in our case 20, to minimize error for a single randomly selected objective.
-The local pattern-search algorithm used has been described previously \cite{Gadkar2003}.
-
-\input{Prostate_Obj_Table}
-\clearpage
-
-\input{Prostate_Pred_Table}
-\clearpage
-
-\begin{figure}\centering
-\includegraphics[width=1.0\textwidth]{./figs/Supplementary_CV_Subset_Total}
-\caption{Coefficient of variation (CV) of model parameters estimated using POETs.
-The solid line denotes the mean CV calculated over the entire ensemble (N = 5000).
-The points denote the mean CV of the 500 ensemble members used for sensitivity and robustness calculations.
-Over the ensemble, the coefficient of variation (CV) of the kinetic parameters spanned 0.59 - 5.8, with 33\% of the parameters having a CV of less than one.}
-\label{fg:Supp_CV}
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1.02\textwidth,height=0.5\textheight]{./figs/Figure_6_Ackley_Rast}
+\caption{Error convergence on 300 dimensional Ackley function and Rastrigin function.
+\textbf {(A)} Error convergence of 5 different meta-heuristics on 300 dimensional Ackley function \textbf {(B)} Error convergence of 5 different meta-heuristics on 300 dimensional Rastrigin function
+}\label{fig-testfunctions}
 \end{figure}
 
-\begin{figure}\centering
-\includegraphics[width=1.0\textwidth]{./figs/Supp_Figure_NewModelvsOldModel}
-\caption{Blind model predictions for the ensemble with the original and updated model (EGFR and HER2 heterodimer).
-A,B.
-Time course data for HER phosphorylation due to a stimulus of 1.6 nM EGF (LNCaP C33, P7) for the old and new model, respectively.
-Dark grey shows only parameters improved by the updated model (N=2870) while light grey show all parameter sets (N=5000).
-C,D.
-Shc phosphorylation levels at 16 hrs in the presence and absence of 1.6 nM EGF (LNCaP C33, P29) for the old and new model, respectively.
-Light grey denotes experimental data, mid grey denotes simulation results for all parameters (N=5000), and black denotes only parameters improved by the updated model (N=3142).}
-\label{fg:Exps_NewModelvsOldModel}
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1.02\textwidth,height=0.5\textheight]{./figs/Figure_7_Benchmarks_Parameters}
+\caption{Difference between optimal and nominal parameter vector values on benchmark problems.
+\textbf {(A)} Problem B1: Genome wide kinetic model of E.coli \textit{S.cerevisiae} with 1759 unknown parameters.
+\textbf {(B)} Problem B4: Metabolic model of Chinese  Hamster Ovary Cells (CHO) cells with 117 parameters.
+}\label{fig-benchmark}
 \end{figure}
 
-\begin{figure}\centering
-\includegraphics[width=1.0\textwidth]{./figs/Fig_6_PI3K_AR_KO_3_Cases.pdf}
-\caption{Robustness analysis of a population of CR prostate models with PI3K (blue), AR (green), and PI3K and AR (orange) knock-outs (N = 500).
-A log fold change of greater then zero implies that the concentration of the protein increased with the knock-out, while a log fold change of less then zero indicates that the concentration of protein decreased.
-A log of fold change equal to 0, shows no response due to the knock-out.
- A.,B.,C. Log robustness of cyclin E, PSA, and p70p versus Cyclin D for the three knock-out cases. A CR LNCaP cell was assumed for all knock-out cases.}
-\label{fg:Supp_Dual_Knockout}
-\end{figure}
-
-\begin{figure}\centering
-\includegraphics[width=1.0\textwidth]{./figs/Supp_Figure_Sensitivity.pdf}
-\caption{Sensitivity analysis of a population of prostate models (N = 500).
-Species with a low sensitivity are considered robust, while species with a high sensitivity ranking are considered fragile.
-$\bf{A}$ Sensitivity ranking of network species in CR cells in the absence and presence of DHT.
-$\bf{B.}$ Sensitivity ranking of network species in CR cells in the presence of enzalutamide in the presence and absence of a DHT stimulus.
-$\bf{C.,D.}$ Sensitivity ranking of network species in CR cells in the presence and absence of sorafenib and lapatinib, respectively, with a DHT stimulus.}
-\label{fg:Supp_Sensitivity}
-\end{figure}
-
-\begin{figure}\centering
-\includegraphics[width=1.0\textwidth]{./figs/Supp_Figure_Robustness.pdf}
-\caption{Robustness analysis of protein markers.
-Expression level of key proteins was altered by a factor of 2, 0.1, or 0 (knock-in, knock-down, or knock-out) and robustness coefficients were calculated for five key protein markers.
-Simulations shown were from CR cells, with indicated perturbation.
-Mean of 500 ensemble members is shown.}
-\label{fg:Supp_Robustness}
-\end{figure}
-
-\begin{figure}\centering
-\includegraphics[width=1.0\textwidth]{./figs/Supp_Figure_KO_new.pdf}
-\caption{Dual knock-out of AR and PI3K leads to decreased expression of activated p70.
-A., B, C.
-Robustness coefficient of activated p70 (S6) in the PI3K knock-out, AR knock-out, and dual knock-out cases, respectively.
-The control was the basil CR LNCaP wild type case.
-Error bars denote plus and minus one standard error above the mean with N = 500.
-Experimental data is from Carver, et al \cite{Carver2011}.
-}
-\label{fg:Supp_KO}
-\end{figure}
-
-%\end{comment}
-
 \end{document}

### 2015-09-12 Paper_v2.tex
changes

@@
 Fan et al.
-recently showed that population based meta-heuristics along with decomposition based methods can be used to model gene circuits from mRNA data \cite{fan2015parameter}.Despite these successes, a  major drawback with most of these metaheuristic approaches is the vast number of objective function evaluations they take in arriving at good search regions.
+recently showed that population based meta-heuristics along with decomposition based methods can be used to model gene circuits from mRNA data \cite{fan2015parameter}.
+Despite these successes, a  major drawback with most of these metaheuristic approaches is the vast number of objective function evaluations they take in arriving at good search regions.
 As the models grow in size and complexity and the number of dimensions of parameter vector increases, evaluation of the objective function becomes computationally expensive.
@@
 \subsection*{Dynamically dimensioned particle swarm optimization (DdPSO).}
-DdPSO is a novel meta-heuristic that combines multi-swarm based particle swarm methods with dynamically dimensioned search (Fig. \ref{fig-algorithm}).
+DdPSO is a novel meta-heuristic that combines multi-swarm based particle swarm methods with dynamically dimensioned search (Fig.\ref{fig-algorithm}).
 The goal of this approach is to obtain optimal or near optimal parameters for high-dimensional complex biological systems within a pre-specified number of function evaluations.
@@
 While updating the particle, we made sure all dimensions of the solution represented by the particle were within bounds using a reflection boundary condition.
+
+\[
+    {x}_{i,j}^{new}= 
+\begin{cases}
+    {x}_{i,j}^{old}+(x_{i}^{min} - {x}_{i,j}^{old}),&\text{if} \ x_{i,j}^{old} < x_{i}^{min}\\
+    		0,              & \text{if}
+\end{cases}
+\]
+
 After every $\mathbf{g}$  function evaluations, the particles within all sub-swarms were mixed and then randomly redistributed to a new sub-swarm.
@@
 Coagulation is an ideal system to test our approach since the biology is well known and complex, with multiple feed back loops that are tightly regulated.
-We used experimental data with under different conditions to obtain optimal parameters and used these parameters to make predictions against unseen experimental data.
-We obtained good fits and made accurate enough predictions using parameters obtained from 4000 function evaluations.
+We used experimental data under different conditions to obtain optimal parameters and used these parameters to make predictions against unseen experimental data.
+We obtained good fits and made sufficiently accurate enough predictions using parameters obtained from 4000 function evaluations.
 Further, we also used high-dimensional forms of commonly used test functions of global optimization and showed that we were able to find the global minimum for 300 dimensional Ackley and Rastrigin functions faster than other meta-heuristics.
 We also considered two recently published benchmark problems to test parameter optimization approaches and showed that we were able to retrieve the nominal parameter vector within 4000 function evaluations.
-
-meta-heuristic approaches are generally effective in finding close to optimum solutions of complex, multi-modal functions.
+Meta-heuristic approaches are generally effective in finding close to optimum solutions of complex, multi-modal functions.
 In addition, they generally obviate the need for any \textit{a priori} knowledge (like function derivative).

### 2015-09-12 Paper_v2.tex
error change
probform

@@
 \usepackage{textcomp}
-
+\usepackage{multirow}
+\usepackage{amsmath}
+\DeclareMathOperator*{\argmin}{\arg\!\min}
 %\usepackage[space]{cite}
@@
 
-\clearpage
+\section*{Problem Formulation}
+
+The problem of parameter estimation in dynamic biological systems consists of finding an optimal parameter vector that results in the best fit between the model predictions and observed experimental data.The cost function or objective that measures the goodness of fit is generally the Euclidean norm which is minimized under certain constraints.
+This can be mathematically expressed as follows.
+
+\begin{equation}\label{eq:static_opt}
+\begin{aligned}
+& \text{minimize}
+& & K(\mathbf{p})=(\sum_{i=1}^N g_i(t_i,\mathbf{x,p,u})-y_i)^2) \\
+& \text{subject to}
+&&\dot{x}=f(t,\mathbf{x(t,\mathbf{p}),p,u(\mathbf{t})})\\
+&&&x(t_0) = x_0\\
+&&&c(t,\mathbf{x,p,u}) \geqslant 0 \\
+&&& \mathbf{p}^L \leqslant \mathbf{p} \leqslant \mathbf{p}^U\\
+\end{aligned}
+\end{equation}
+
+where t is time, x is the state variable with an initial state x\_0 u is the input vector \, p is the parameter vector and f is the system of equations (like differential equations or set of algebraic constraints), c in the set of linear and non-linear constraints, g is the set of observables and p\textsuperscript{L} and p\textsuperscript{U} are the lower and upper bounds.
+Thus the problem eventually is to find {p}\textsuperscript{*} where
+
+\begin{equation}\label{eq:static_opt_2}
+\mathbf{p^{*}}= \arg\min_{p} K(\mathbf{p})
+\end{equation}
+
 \section*{Results}
@@
 \centering
+\caption{Error Analysis.}
+\label{table:Error-Analysis}
+\begin{center}
+\begin{tabular}{ |c|c|c|c| } 
+\hline
+Simulation&Concentration& Normalized Standard Error & R-Squared Value \\
+\hline
+\multirow{3}{4em}{FVIIa-TF level} & 5nM & 0.0367 & 0.7680\\ 
+& 500pM & 0.0636 & 0.8939\\
+& 50pM & 0.1204 & 0.8384\\ 
+& 10pM & 0.0813 & 0.9492\\
+& 5pM & 0.0574 & 0.9794\\ 
+\hline
+\end{tabular}
+\end{center}
+\end{table}
+
+\begin{table}[h]
+\centering
 \caption{Table with optimization settings and results for coagulation and different benchmarks using DdPSO.}

### 2015-09-23 Paper_v2.tex
FigureUpdates

--- 2015-08-31 Paper_v2.tex
@@
 %\usepackage[authoryear,round,comma,sort&compress]{natbib}
-\usepackage[square,sort,comma,numbers,sort&compress]{natbib}
+\usepackage[square,sort,comma,numbers]{natbib}
 %\usepackage[authoryear,round]{natbib}
@@
 \usepackage{simplemargins}
-\usepackage{fullpage}
 \usepackage{comment}
 \usepackage{lineno}
-%\usepackage{chicago}
-\usepackage{textcomp}
 
-%\usepackage[space]{cite}
 \urlstyle{rm}
@@
 
-%\bibliographystyle{chicago}
-\bibliographystyle{plos2009}
+%\usepackage{geometry}
+\usepackage{fullpage}
+
+%\bibliographystyle{plain}
+\bibliographystyle{msb}
 
@@
 \begin{titlepage}
-{\par\centering\textbf{\Large {Dynamically dimensioned particle swarm optimization (DdPSO): A meta-heuristic for parameter estimation in biochemical models}}}
+{\par\centering\textbf{\Large Modeling and Analysis of Hormone and Mitogenic Signal Integration in Prostate Cancer}}
 \vspace{0.05in}
-{\par \centering \large{Adithya Sagar, Christine Shoemaker$^{\dag}$ and Jeffrey D. Varner$^{*}$}}
+{\par \centering \large{ Katharine V. Rogers, Joseph A. Wayman, Ryan Tasseff, Caitlin Gee, Matthew P. DeLisa, 
+and Jeffrey D. Varner$^{*}$}}
 \vspace{0.10in}
-{\par \centering {School of Chemical and Biomolecular Engineering}}
-{\par \centering {$^{\dag}$School of Civil and Environmental Engineering}}
-{\par \centering {Cornell University, Ithaca NY 14853}}
+{\par \centering \large{School of Chemical and Biomolecular Engineering}}
+{\par \centering \large{Cornell University, Ithaca NY 14853}}
 \vspace{0.1in}
-{\par \centering \textbf{Running Title:}~Parameter estimation in biochemical models}
+{\par \centering \textbf{Running Title:}~Modeling signal transduction in prostate cancer}
 \vspace{0.1in}
-{\par \centering \textbf{To be submitted:}~\emph{PLoS ONE}}
+{\par \centering \textbf{To be submitted:}~\emph{Molecular Systems Biology}}
 \vspace{0.5in}
@@
 \section*{Abstract}
-Mathematical modeling is as a powerful tool to analyze, and ultimately design biochemical networks.
-However, estimation of biochemical model parameters can be a significant challenge.
-Parameter estimation in biochemical models generally involves expensive function evaluations, making it difficult to quickly obtain optimal solutions.
-Additionally, biochemical models often have many local minima or maxima which further complicates the parameter estimation problem.
-In this study, we developed dynamically dimensioned particle swarm optimization (DdPSO), a novel meta-heurstic that combined the global search capability of multi-swarm particle swarm optimization with dynamically dimensioned search.
-DdPSO uses a multi-swarm particle swarm optimization technique to generate candidate solution vectors which are then greedily updated using dynamically dimensioned search.
-We tested the performance of DdPSO on commonly used test functions for global optimization (Ackley and Rosenbrock), on published biochemical parameter estimation benchmark problems and
-lastly on a model of the human coagulation cascade. For each test we performed 25 trials with 4000 function evaluations per trial, and compared the performance of DdPSO with other commonly
-used meta-heuristic approaches such as differential evolution (DE), particle swarm optimization (PSO), simulated annealing (SA) and dynamically dimensioned search (DDS).
-For the wide range of problems that we considered, DdPSO outperformed the other meta-heuristic approaches despite a limited number of function evaluations.
-We further tested the predictive power of the coagulation model parameters against data not used in training, and found good agreement between model simulations and experimental measurements.
-Taken together, DdPSO is a promising meta-heuristic approach for the estimation of biochemical model parameters in relatively few function evaluations.
+Prostate cancer is the most common cancer in men and the second leading cause of cancer related death in the United States. 
+Androgens, such as testosterone, are required for prostate cancer growth. 
+Androgen ablation in combination with radiation or chemotherapy remains the primary non-surgical treatment for androgen dependent prostate cancer. 
+However, androgen ablation typically fails to permanently arrest cancer progression, often resulting in castration resistant prostate cancer (CRPC). 
+CRPC is closely related to metastasis and decreased survival. 
+In this study, we developed and analyzed a population of mathematical models describing growth factor and hormone signal integration in androgen dependent, intermediate and resistant prostate cancer cells.
+The model describes the integration of two simultaneous extracellular signaling inputs, namely, androgen and growth factors into a G1/S cell cycle checkpoint decision. 
+Model parameters were identified from 43 studies in androgen dependent and resistant LNCaP cell lines. 
+The model was validated by comparing simulations with an additional 29 data sets from LNCaP cell lines that were not used during training. 
+Additionally, data from four drug trials was also used to evaluate the model’s performance.
+Sensitivity analysis, conducted over an ensemble of prostate signaling models, suggested that in an androgen free environment general translation and transcription was more sensitive in androgen dependent cells, while in androgen independent cells the PI3K and MAPK pathway species were more sensitive.
+In a constant DHT environment sensitive species were conserved between the cell lines. 
+These results suggest targeting the PI3K and MAPK pathways in addition to anti-androgen therapies as a treatment for CRPC.
 
-{\noindent \textbf{Keywords:}~Parameter identification, Meta-heuristic optimization, Biochemical modeling}
-
-% Extra abstract
-% Mathematical modeling of biological systems with multiple feed back loops is one such area where parameter estimation is a difficult non-linear optimization problem.
-This difficulty is further compounded when dealing with parameter vectors of high dimensions.
-
-%In this study, we present the dynamically dimensioned particle
-%a novel meta-heuristic approach that combines a variant of particle swarm optimization (PSO) along with dynamically dimensioned search (DDS) to obtain near optimal solutions of high dimensional biochemical networks within a relatively few function evaluations.
-%We use a particle swarm optimization technique that uses multi-swarms to generate candidate vectors which are then greedily updated using DDS by dynamically varying the perturbed parameter dimensions.
-We tested this algorithm (25 trials with 4000 function evaluations in each trial) on a biochemical network of coagulation (148 parameters and 92 species) and compared it's performance against other meta-heuristics like Differential Evolution (DE), Particle Swarm Optimization (PSO), Simulated Annealing (SA) and also against DDS alone.
-The new algorithm outperforms all the other meta-heuristics on the coagulation model.
-The parameter vectors obtained using this approach fit the experimental data well and also make accurate enough predictions on unseen experimental data.
-We also performed this comparison on commonly used test functions (Ackley and Rosenbrock) for global optimization and found the same behavior.
-Further we used two recently published benchmark problems, a genome wide kinetic model with 1759 parameters and a metabolic model of Chinese Hamster Ovary cells with 117 parameters to evaluate the performance of our approach.
-We  surprisingly performed well on these benchmarks and obtained the nominal parameter vector with just 4000 function evaluations in both cases.
+\vspace{0.5in}
+{\noindent \textbf{Keywords:}~Prostate cancer, signal transduction, mathematical modeling}
 
@@
 
-Cells process nutrients and respond to changes in their environment using complex biochemical networks. These networks contain thousands of components
-interconnected through nonlinear enzyme catalyzed reactions.
-Mathematical modeling has evolved as a powerful paradigm to analyze, and ultimately design these complex networks \cite{assmus2006dynamics, Riel:2006aa,Jaqaman:2006aa,kitano2002systems,hood2004systems}.
-Mathematical modeling of biochemical networks is often an iterative process.
-First, models are formulated from biochemical knowledge, and then model parameters are estimated using experimental data \cite{Aldridge:2006aa,banga2008optimization,ashyraliyev2009systems}.
-Parameter estimation is typically framed as a non-linear optimization problem wherein the residual (or objective function) between experimental data and model simulations is minimized using an optimization strategy \cite{moles2003parameter}.
-The optimal parameters obtained from training are then used to validate the model on hitherto unseen experimental data.
-If validation fails, model construction and calibration are repeated iteratively until satisfactory results are obtained.
+%
+%Sipuleucel-T is a first generation cancer vaccine \cite{Kantoff2010}, while radium-223 is a targeted alpha emitter for bone metastasis \cite{Parker2013}. 
+%Abiraterone is an inhibitor of cytochrome P450 17A1 (CYP17), a critical enzyme in androgen biosynthesis \cite{Sartor2013} while enzalutamide inhibits androgen receptor (AR) translocation to the nucleus, as well as DNA binding and coactivator recruitment \cite{Scher2012}.
+%Thus, understanding the operation of the biological pathways which ultimately facilitate the transition to CRPC is vital to the discovery of new therapeutic strategies against this disease.
+%The model focuses on the outlaw pathway, in which constitutively or overly-activated receptor tyrosine kinases (RTKs) stimulate AKT and mitogen-activated protein kinases (MAPKs), which in turn activate cytosolic androgen receptor (AR) in the absence of androgen.
+Interestingly, among the few genes activated AR represses is cellular prostatic acid phosphatase (cPAcP), itself a key regulatory of RTK activation.
+Thus, the outlaw pathway encodes crosstalk between RTK-dependent kinase and AR activation inside of a positive feedback loop.
+The model was validated with multiple experiments in LNCaP cell lines.
+In addition  For example, in response to enzalutamide, an AR inhibitor, the model predicted a decline in PSA expression similar to patient response.
+%and there is limited data on combination therapies or the use of biomarkers to develop personalized treatment strategies.
 
-Parameter estimation is a major challenge in the development of large biochemical models.
-Although parameter estimation has been a well studied problem in engineering for decades \cite{nieman1971review,beck1977parameter,young1981parameter,beck1998inverse},
-the complex dynamics of large biological systems and noisy, often incomplete experimental data pose a unique estimation challenge.
-Most of the optimization problems pertaining to biological systems are non-linear and multi-modal i.e.
-most of them have multiple local minima or maxima \cite{moles2003parameter,banga2008optimization}.
-The non-linearity of the problem coupled with multi-modality generally renders the local optimization techniques like pattern search method \cite{hooke1961direct}, Nelder-Mead simplex method \cite{nelder1965simplex}, steepest descent method or Levenberg-Marquardt method \cite{more1978levenberg} incapable of reliably obtaining an optimal solution since they generally stop at the first local minimum.
-Though deterministic global optimization techniques (for example algorithms based on branch and bound framework) can handle non-linearity and multi-modality \cite{esposito2000deterministic,horst2013global}, the absence of derivative information, discontinuity of the objective functions, non-smooth regions or the lack of any knowledge about the objective function severely hamper the use of these techniques.
+Prostate cancer (PCa) is the most commonly diagnosed cancer and the second leading cause of cancer-related death in men in the United States \cite{Siegel2013}. 
+Initially, PCa cells depend upon the activation of cytosolic androgen receptors (AR) by androgen hormones, such as testosterone, for survival and growth. 
+Androgen ablation in combination with radiation or chemotherapy remains the primary non-surgical treatment for androgen dependent prostate cancer (ADPC) \cite{Huggins1967}. 
+However, androgen ablation typically fails to permanently arrest cancer progression as malfunctioning cells eventually lose androgen sensitivity and proliferate without hormone. 
+The loss of androgen sensitivity results in castration resistant prostate cancer (CRPC), a phenotype closely linked with metastasis and greatly reduced survival \cite{Harris2009}.
+Currently, there are six approved treatments that demonstrate a survival advantage in patients with metastatic CRPC, each of these target diverse aspects of the disease \cite{Sartor2013}.
+The taxane family members docetaxel and cabazitaxel interact with microtubule stability \cite{Tannock2004,Bono2010}, while abiraterone \cite{Sartor2013} or enzalutamide \cite{Scher2012} interfere with androgen signaling by blocking androgen formation or nuclear translocation, respectively.
+Other approved treatments are non-specific to PCa. 
+For example, general treatments such as sipuleucel-T, a first generation cancer vaccine \cite{Kantoff2010}, 
+and radium-223, an alpha emitter which targets bone metastasis \cite{Parker2013}, are both approved to treat CRPC.
+Unfortunately, regardless of the therapeutic approach, the survival advantage of these treatments is typically only a few months. 
+Thus, understanding the molecular basis of the loss of androgen sensitivity in CRPC could be an important step for the development of the next generation of therapies with a prolonged survival
+advantage.
 
-Meta-heuristic stochastic optimization approaches like Genetic Algorithms (GAs), Simulated Annealing (SA) \cite{kirkpatrick1983optimization}, Evolutionary Programming and population based searches like Differential Evolution (DE) \cite{storn1997differential} have shown promise in this regard \cite{sun2012parameter}.
-They do not make any assumptions about the structure of objective function or require any \textit{a priori} information about the objective function.
-Though they do not guarantee strong convergence, these approaches are effective in finding near optimal solutions.
-Mendes et al.
-\cite{mendes1998non} used Simulated Annealing to estimate rate constants for the irreversible inhibition of HIV proteinase, Modchang et al.
-\cite {modchang2008mathematical} used Genetic Algorithms to estimate parameters for a mathematical model of signal transduction, Differential Evolution based approaches have been an effective approach on various systems \cite{tsai2005evolutionary,wang2001hybrid,noman2007inferring}.
-Tashkova et al.
-\cite{tashkova2011parameter} compared different meta-heuristics for parameter estimation on a dynamic model of endocytosis and showed that DE was the most effective.
-Banga and co-workers have applied scatter-search based methods \cite{villaverde2012cooperative,rodriguez2006novel,egea2007scatter} to estimate parameters on non-linear biological processes.
-Hybrid approaches that combine a meta-heuristic with a local optimization search, wherein a near globally optimal solution that is obtained with a meta-heuristic is further refined using a local search have also become popular.
-Villaverde et al.
-\cite{villaverde2015biopredyn} combined scatter search with local search methods for parameter estimation in large scale systems biology models.
-Fan et al.
-recently showed that population based meta-heuristics along with decomposition based methods can be used to model gene circuits from mRNA data \cite{fan2015parameter}.Despite these successes, a  major drawback with most of these metaheuristic approaches is the vast number of objective function evaluations they take in arriving at good search regions.
-As the models grow in size and complexity and the number of dimensions of parameter vector increases, evaluation of the objective function becomes computationally expensive.
-Thus performing a large number of evaluations is not computationally feasible.
+Androgen-induced proliferation and survival depends upon many coordinated signal transduction and gene expression events. 
+Androgen Receptor (AR) is part of the nuclear hormone receptor superfamily, which includes other important cancer targets such as progesterone receptor (PR) and estrogen receptor (ER) in breast cancer \cite{Aranda2001}.
+Nuclear hormone receptors act as ligand dependent transcription factors interacting with specific DNA sequences of target genes as either monomers, heterodimers, or homodimers; AR, PR, and ER act as homodimers.
+In the case of AR these specific DNA sequences are known as androgen response elements (ARE) \cite{Mangelsdorf1995}.
+In the absence of androgen, AR is predominately found in the cytoplasm bound to heat shock proteins (HSP) \cite{Prescott2006}.
+%Nuclear Hormone Receptors (Reviewed \cite{Aranda2001}) and AR in disease reviewed in \cite{Matsumoto2013}
+Androgen, either testosterone or testosterone metabolites such as 5$\alpha$-dihydrotestosterone (DHT), enter prostate cells and interact with the cytosolic androgen receptor (AR). 
+The interaction of DHT with AR promotes the dissociation of AR from chaperones such as HSP \citep{Pratt:1997kx} and its subsequent dimerization, phosphorylation and translocation to the nucleus (reviewed by Brinkmann \textit{et al.} \citep{Brinkmann1999}).
+Activated nuclear AR drives a gene expression program broadly referred to as androgen action, that promotes both proliferation and survival.
+In addition to many genes including itself, activated nuclear AR promotes the expression and secretion of prostate specific antigen (PSA), arguably the best known PCa biomarker \citep{Feldman2001}.
+PSA is commonly used as a prostate cancer indicator, although its prognostic ability is controversial \citep{Attard2009,Hoffman2011,Moyer2012}. 
+In CRPC, AR signals in the absence of androgens. 
+Androgen dependent (AD) prostate cells can become castration resistant (CR) through several possible mechanisms, including constitutively amplified AR expression and altered AR sensitivity to testosterone or other non-androgenic molecules \citep{Feldman2001}.
+In this study, we focused on the aberrant activation of AR by kinase signaling cascades, sometimes called the outlaw pathway. 
+Outlaw pathway activation is driven by over-activated receptor tyrosine kinases (RTKs), a common pathology in many cancer types including PCa \citep{Slamon1989,Craft1999}. 
+RTKs stimulate downstream kinases, including the AKT and mitogen-activated protein kinase (MAPK) pathways, which promote AR phosphorylation and dimerization in the absence of an androgen signal \citep{Craft1999,Yeh1999}.
+Interestingly, among the few genes activated AR represses is cellular prostatic acid phosphatase (cPAcP), itself a key regulatory of RTK activation \cite{Veeramani2005}. 
+Thus, in CRPC the androgen program is initiated without the corresponding extracellular hormone cue, potentially from crosstalk between growth factor and hormone receptor pathways.
 
-In many of these high dimensional problems approaching an exact solution may not be necessary.
-Gutenkust et al.
-\cite{gutenkunst2007universally} showed that a number of systems biology models are 'sloppy'.
-Sloppy systems have specific parameter combinations that largely define the dynamics of the system.
-Large perturbations to the rest of the parameters does not greatly impact the system dynamics.
-Ensemble approaches \cite{song2010ensembles,luan2010ensembles} have exploited this aspect to describe the dynamics of biological systems including coagulation which can be described using only a set of key species or parameters \cite{sagar2015dynamic}.
-Tolson and Shoemaker \cite{tolson2007dynamically} showed through Dynamically Dimensioned Search (DDS) that high-dimensional watershed models can be calibrated quickly by perturbing only a subset of dimensions.
+In this study, we developed a mathematical model of growth factor and hormone signal integration in androgen dependent, intermediate and resistant prostate cancer cells. 
+We used this model to better understand which components and processes were differentially important in AD versus CR cells.
+The new model architecture was a significant advance over our previous prostate signaling model \citep{Tasseff2010}. 
+We added the regulated expression of ten additional proteins, including the cell cycle restriction point proteins cyclin D (and the differentially spliced variants cyclin D1a and cyclin D1b), cyclin E, cyclin-dependent kinase inhibitor 1A (p21Cip1), and cyclin-dependent kinase inhibitor 1B (p27Kip1).
+Also, we included the Rb/E2F pathway, expanded our description of the activation of the mammalian target of rapamycin (mTOR) protein and its role in translation initiation, and included the regulation of AR action by cyclin D1a and E2F.
+However, this upgraded architecture, while increasing the biological scope of the model, also expanded the number of unknown model parameters. 
+To estimate these parameters, we used multiobjective optimization in combination with dynamic and steady-state data sets generated in AD, intermediate and CR LNCaP cell lines. 
+We identified a population of approximately N = 5000 models (from well over a million candidate models) which described both AD and CR data sets using a single model structure.
+Furthermore, we tested the model using an additional 29 LNCaP data sets not used for model training, along with four drug studies.
+We analyzed the model population using sensitivity and robustness analysis to uncover differentially important mechanisms in AD versus CR cell lines. 
+In the presence of androgen, the sensitivity profile was similar between AD and CR cells. 
+Components of the MAPK and PI3K pathways were highly fragile, irrespective of the level of androgen dependence. 
+However, in the absence of androgen, there were 609 statistically significant shifts in species sensitivity between AD and CR cells.
+Of these, 108 were larger than one standard deviation above the mean.
+In CR cells, HER2 activation of the MAPK and PI3K pathways was significantly more important, as was AR activation through the MAPK pathway. 
+On the other hand, components of the translation and transcription infrastructure were differentially more important in AD cells in the absence of androgen.  
+Taken together, our analysis suggested that independently targeting the PI3K or MAPK pathways in combination with anti-androgen therapies could perhaps be an effective treatment strategy for CRPC.
 
-In this study, we developed dynamically dimensioned particle swarm optimization (DdPSO), a novel meta-heurstic that combined the global search capability of multi-swarm particle swarm optimization with dynamic dimensioned search.
-The objective of DdPSO was to obtain near optimal parameter estimates for large biochemical models within a relatively few function evaluations.
-DdPSO uses a multi-swarm particle swarm optimization technique to generate candidate solution vectors which are then greedily updated using dynamic dimensioned search.
-We tested the performance of DdPSO on commonly used test functions for global optimization (Ackley and Rosenbrock), published biochemical parameter estimation benchmark problems and
-lastly on a model of the human coagulation cascade.
-DdPSO outperformed common meta-heuristic approaches like Differential Evolution (DE), Particle Swarm Optimization (PSO), Simulated Annealing (SA) and dynamically dimensioned search (DDS) on the benchmark and coagulation problems.
-
-% Extra text from the introduction -
-%With the capacity to construct very large models using various mathematical formalisms \cite{chen2009input,tasseff2011modeling,luan2007computationally,mo2007genome,orth2011comprehensive,karr2012whole,buchel2013path2models,smallbone2010towards}, more often than not,
-
-%this algorithm (25 trials with 4000 function evaluations in each trial) on a biochemical network of coagulation (148 parameters and 92 species) and compared it's performance against other meta-heuristics  The new algorithm outperforms all the other meta-heuristics on the coagulation model.
-The parameter vectors obtained using this approach fit the experimental data well and also make accurate enough predictions on unseen experimental data.
-We also performed this comparison on  and found the same behavior.
-Further we used two recently published benchmark problems, a genome wide kinetic model with 1759 parameters and a metabolic model of Chinese Hamster Ovary cells with 117 parameters to evaluate the performance of our approach.
-We  surprisingly performed well on these benchmarks and obtained the nominal parameter vector with just 4000 function evaluations in both cases.
-%took into cognizance the fact that it may not be necessary to obtain the exact solution for high dimensional biological systems and that good enough solutions can be quickly obtained without expending a lot of objective function evaluations.
-Hence our current approach uses the power of population based heuristics along with DDS to obtain near optimal or good solutions.
-Though in theory we can combine any meta-heuristic with DDS, for the current purpose we used Particle Swarm Optimization (PSO).
-PSO unlike Genetic Algorithm (GA) or Differential Evolution (DE) does not have complex operations like cross over, mutation or recombination.
-It is simple to use and does not have a lot of parameters associated with other heuristics.
-However PSO is known to rapidly converge to a local optimum and thus several variants \cite{peer2003using,zhan2009adaptive,li2007fast} have been developed of which Multi swarm particle swarm optimization approaches (MLSPSO) \cite{zhao2008dynamic,liang2005dynamic} are one of the effective ones.
-We used MLSPSO to preclude bad regions of search and thereafter search within these regions using DDS.
-
-\clearpage
-\section*{Results}
-\subsection*{Dynamically dimensioned particle swarm optimization (DdPSO).}
-DdPSO is a novel meta-heuristic that combines multi-swarm based particle swarm methods with dynamically dimensioned search (Fig. \ref{fig-algorithm}).
-The goal of this approach is to obtain optimal or near optimal parameters for high-dimensional complex biological systems within a pre-specified number of function evaluations.
-We randomly initialized a swarm of $\mathcal{K}$-dimensional particles (represented as ${x}_{i}$), wherein each of these particles corresponds to a $\mathcal{K}$-dimensional parameter vector.
-After initialization, the particles were grouped into different sub-swarms randomly.
-Thereafter within each sub-swarm ${S}_{k}$,  particles were updated according to the following rule.
-\begin{eqnarray}
-	\mathbf{x}_{i,j} &=&\theta_{1,j-1}\mathbf{x}_{i,j-1} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{x}_{i,j-1}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{GL}_{k} - \mathbf{x}_{i,j-1}\right)
-\end{eqnarray}
-where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes the best local solution found by particle $i$ till function evaluation $j$-$1$, and
-$\mathcal{GL}_{k}$ denotes the best local solution found over the population of all particles within the swarm ${S}_{k}$.
-The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
-parameters ($\mathcal{K}\times{1}$). In our algorithm the parameter $\theta_{1,j-1}$ depends on the function evaluations and is controlled according to the following equation
-\begin{eqnarray}
-	\mathbf \theta_{1,j}&=&((\mathbf{N}-{j})*(\mathbf{w}_{max}-\mathbf{w}_{min}))/(\mathbf{N}-{1}) + \mathbf{w}_{min}
-\end{eqnarray}
-where $\mathbf{N}$ represents the total number of function evaluations, $\mathbf{w}_{max}$ and $\mathbf{w}_{min}$ are the maximum and minimum inertia weights respectively.
-While updating the particle, we made sure all dimensions of the solution represented by the particle were within bounds using a reflection boundary condition.
-After every $\mathbf{g}$  function evaluations, the particles within all sub-swarms were mixed and then randomly redistributed to a new sub-swarm.
-The particles were then again updated according Eq.1.
-This process continued till $\mathcal{FR}*\mathbf{N}$ number of functions evaluations, where $\mathcal{FR}$ represents the fraction of evaluations with the multi-swarms.
-At the end of these function evaluations, we froze all the solutions represented by various particles and chose the particle with best solution among $\mathcal{GL}_{1} \cdots \mathcal{GL}_{NS}$ as the initial candidate vector $\mathcal{G}$ for the remaining $({1}-\mathcal{FR})*\mathbf{N}$ number of function evaluations.
-
-This particle was then updated according to the following rule
-\begin{eqnarray}
-	\mathcal{G}_{new}(\mathbf{J})& = &\mathcal{G}(\mathbf{J})+\mathbf{r}_{normal}(\mathbf{J})\sigma(\mathbf{J})
-\end{eqnarray}
-where $\mathbf{J}$ represents the set containing the specific dimensions being perturbed, ${r}_{normal}$ denotes a normal random vector of the same dimensions as $\mathcal{G}$.
-$\sigma$ is the amplitude of perturbation given by following equation:
-\begin{eqnarray}
-	\sigma& = &\mathbf{R}(\mathcal{MAX} -\mathcal{MIN})
-\end{eqnarray}
-where $\mathbf{R}$ is the scalar perturbation size parameter, $\mathcal{MAX}$ and $\mathcal{MIN}$ are ($\mathcal{K}\times{1}$) vectors that represent the maximum and minimum bounds on each dimension.
-The set $\mathbf{J}$ was constructed using a probability function $\mathcal{P}_{j}$ that represented the probability whether a specific dimension $j$ was perturbed or not.
-This function is a monotonically decreasing function that decreases with the number of function evaluations.
-$\mathcal{P}_{j}$ can be any monotonically decreasing function, in our approach we used the following function:
-\begin{eqnarray}
-	\mathcal{P}_{j}&=&{1}-\log(j/ (\beta({1}-\mathcal{FR})*\mathbf{N}))
-\end{eqnarray}
-where $\beta$ is the perturbation frequency probability modulator.
-Thus the number of dimensions of the candidate vector that are updated or perturbed decreases with the as the number of function evaluations increase.
-These updates are greedy in nature, so $\mathcal{G}_{new}$ becomes the new solution vector only if it is better than the old one $\mathcal{G}$.
-
-\subsection*{Performance of DdPSO on benchmark problems.}
-We compared the performance of DdPSO with algorithms on commonly used test functions for global optimization, as well as benchmark biochemical problems from literature (Table \ref{table:benchmark-problems}).
-We used a 300 dimensional Rastrigin function and 300 dimensional Ackley function.
-Both Ackley and Rastrigin have multiple local minima and maxima and attain a global minimum value of 0.
-We tested the performance of the 5 different heuristic approaches on these two functions.
-In each experiment we again fixed the number of function evaluations at 4000 and ran 25 experiments.
-In both cases (Figure S6 A and B) we see that the error convergence rate for DDSMLSPSO is much faster as compared to the other four s and it finds the global minimum of 0 in both cases.
-
-Villaverde and co-workers recently published a set of benchmark problems to evaluate parameter estimation methods \cite{villaverde2015biopredyn}.
-From a computational cost perspective problems they categorized the problems as most expensive, intermediate and least expensive.
-We evaluated the performance of our algorithms on a problem from the most expensive and least expensive categories.
-The first problem is a genome wide kinetic model of \textit{Saccharomyces cerevisiae} with 1759 parameters (henceforth referred to as problem B1).
-The second problem (henceforth referred to as problem B4) is a metabolic model of Chinese Hamster Ovary (CHO) with 35 metabolites, 32 reactions and 117 parameters.
-In both cases pseudo time series data was generated.
-For problem B1, the time series data consisted of 44 observables and for problem B4 the data corresponded to 13 different metabolites.
-We fixed the number of function evaluations at 4000 for DDSMLSPSO and trained both the models against the pseudo experimental data.
-We recaptured the 'nominal' parameters (Figure S7 A and B)  in both cases within 4000 evaluations.
-The final objective function value (Table 1) is an order smaller for B1 possibly due to overfitting.
-In Table 1 we summarize the results obtained on the coagulation model and the benchmark problems.
-
-\subsection*{Performance of DdPSO on a model of the human coagulation cascade.}
-We compared the performance of DdPSO on a model of blood coagulation against commonly used meta-heuristics like simulated annealing (SA), differential evolution (DE), particle swarm optimization (PSO),and dynamically dimensioned search (DDS).
-Coagulation is an archetype biochemical network that is highly interconnected and tightly regulated with multiple positive and negative feedback loops (Fig.
-\ref{fig-coagulation-network}).
-The biochemistry underlying coagulation, though quite complex has been well studied \cite{mann2003dynamics,mann2003all,mann2003thrombin,vogler2009contact,diamond2013systems,fogelson2005coagulation,anand2003model}, and reliable experimental coagulation models have been developed \cite{hockin2002model,chatterjee2010systems,mann2006models,luan2007computationally}.
-This makes it an ideal system for mathematical modeling and parameter estimation.
-Coagulation is regulated by a set of serine proteases also known as coagulation factors and blood platelets.
-The coagulation factors are generally in an inactive state and are known as zymogens.
-These zymogens are activated through certain triggers.
-Trigger events like injury or trauma or sepsis expose factors like collagen, tissue factor and von Willebrand factor (vWF) to blood.
-The exposure of these factors to blood kick-starts a series of convergent cascades that lead to conversion of zymogen prothrombinase to thrombin.
-Luan et al.
-modeled coagulation using coupled non-linear ordinary differential equations with 148 reactions and 92 species \cite{luan2007computationally} and validated model using 21 published datasets.
-
- To train the model parameters, we used data sets from TF-VIIa initiated coagulation with no anticoagulants.
-The objective function was a weighted linear combination of two different error functions that used data sets representing coagulation initiated with different concentrations of TF-VIIa (5pM, 5nM) \cite{hockin2002model}.
-We restricted the number of function evaluations to 4000 for each algorithm and performed 25 trials of this experiment.
-DdPSO exhibits a much faster rate of error convergence and has a much lower final error as compared to the other algorithms (Figure 3).
-Within the first 1000 function evaluations (swarm phase) of DDSMLSPSO there is a very rapid drop in error.
-Subsequently, again after 2000 function evaluations (dynamically dimensioned phase) the error drops quickly.
-Overall at the end of 4000 function evaluations DDSMLSPSO minimizes the error (final objective error is 90456.9) to a much greater extent than any of the other algorithms.
-Amongst the rest of the algorithms DDS has a comparable error drop for the first 2000 function evaluations.
-However the drop in error remains constant for DDS after 2000 function evaluations.
-Using the parameters obtained at the end of 4000 function evaluations we examined the 'fits' between models predictions and experimental data (Figure 2).
-The solid lines represent the mean value of prediction over 25 trials and the shaded region represents the 99\% confidence interval.
-Subsequently we compared the model predictions against completely 'unseen' or untrained experimental where coagulation was initiated with 500pM,50pM,10pM concentrations of TF-VIIa respectively (Figure 3).
-
-\clearpage
-\section*{Discussion}
-Our study presents a novel approach for high-dimensional parameter estimation in complex biological systems with relatively few function evaluations.
-In this approach we combined a variant of a well known meta heuristic particle swarm optimization with Dynamically Dimensioned Search (DDS).
-We tested our approach on an ODE model of coagulation with 148 parameters and 92 species.
-Coagulation is an ideal system to test our approach since the biology is well known and complex, with multiple feed back loops that are tightly regulated.
-We used experimental data with under different conditions to obtain optimal parameters and used these parameters to make predictions against unseen experimental data.
-We obtained good fits and made accurate enough predictions using parameters obtained from 4000 function evaluations.
-Further, we also used high-dimensional forms of commonly used test functions of global optimization and showed that we were able to find the global minimum for 300 dimensional Ackley and Rastrigin functions faster than other meta-heuristics.
-We also considered two recently published benchmark problems to test parameter optimization approaches and showed that we were able to retrieve the nominal parameter vector within 4000 function evaluations.
-
-meta-heuristic approaches are generally effective in finding close to optimum solutions of complex, multi-modal functions.
-In addition, they generally obviate the need for any \textit{a priori} knowledge (like function derivative).
-However they take an exorbitant number of objective function evaluations to come close to an optimum.
-When the objective function evaluations tend to become expensive it is infeasible to take up a large number of evaluations.
-As the dimensionality of parameter space increases, the search region gets widened and thus the problem becomes more challenging.
-In addition, most of these approaches require optimization of  'algorithm parameters' before the actual optimization and also involve computationally expensive update operations.
-Tolson and Shoemaker, through DDS, showed that randomly perturbing a subset of dimensions in high dimensional parameter space is an effective way to obtain near optimal solutions with few function evaluations.
-However their approach was based on a single solution.
-Single solutions based methods may prove ineffective in complex systems when they start from a 'bad' region of search or from a local minimum.
-In our approach, we tried to eliminate the probability of starting from a bad region of search using a variant of particle swarm optimization.
-Particle Swarm Optimization (PSO) is a population based meta heuristic that is known for rapid convergence without any complex operations like recombination, mutation or selection that are associated with other population based meta-heuristics like Differential Evolution (DE) or Genetic Algorithm (GA).
-Several particle swarm variants have been proposed to improve the search ability and rate of convergence, that involve different neighborhood structures, multi-swarms or adaptive parameters.
-Multi-swarm PSO with small particle neighborhoods have been shown to better in searching on complex multi-modal solutions \cite{zhao2008dynamic}.
-Multi swarm methods, in addition, avoid rapid convergence to a local optimum or stable point and are able to generate diverse solutions.
-Generation of diverse solutions in the early stage gives a better exploratory capability and thus converge of upon multiple optima.
-
-We utilized this capacity of multi swarms to generate diverse candidate solutions which can be used as initial solutions for a DDS like search.
-Thus the solution vectors obtained at the end of multi swarm search phase have a better propensity to not get stuck in bad regions in DDS phase as opposed to starting with a totally random initial solution.
-Given a fixed number of function evaluations, we have been able to show that we were able to obtain better solutions for the coagulation model and the test functions faster that other commonly known meta-heuristics and also DDS alone.
-Choosing the number of function evaluations is largely a function of cost and complexity of the objective function.
-Traditionally the stopping condition for a parameter optimization problem can be the number of function evaluations, percentage of initial error achieved or an absolute error threshold.
-However in case of complex, expensive functions where we desire a value within a certain period of time, the number of evaluations are used as a stopping criterion.
-In our current study we used a value of 4000 which we based upon the time taken (approximately 8-10 seconds) for a single objective function evaluation in the coagulation case.
-We used the same value of 4000 for benchmarks published by Villaverde and co-workers \cite{villaverde2015biopredyn}.
-Quite surprisingly we took a couple of orders lesser number of function evaluations (1e3) to obtain the optimal parameter vector as compared to the enhanced Scatter Search (eSS) with a local optimizer, which took around the order of 1e5 number of evaluations.
-The amount of CPU time taken (on an Intel Xeon processor 2.4 GHz) is lesser as compared to eSS on a similar architecture.
-
-A surprisingly remarkable aspect about our algorithm has been that we did not 'pre optimize' any parameters of the algorithm to suit a specific problem.
-In the swarm phase this includes the number of particles, number of sub-swarms, acceleration constants or the number of generations after which the particles are redistributed and the neighborhood perturbation parameter in DDS phase.
-We used the same parameters for all the problems.
-The same rule was applied to the rest of the meta-heuristics barring Simulated Annealing.
-For SA, we optimized the cooling schedule for the coagulation model.
-Thus, in this approach any overhead that usually comes with additional function evaluations to optimize parameters was avoided.
-The only parameter that we adjusted was the fraction of evaluations within swarm search.
-For the coagulation model it was set to be 0.4 and for the benchmarks it is at 0.6.
-This parameter was set based on empirical observations rather than any definitive analysis.
-When the error convergence rates stagnated in the swarm phase we switched to DDS phase.
-A further improvement to our approach could  be to incorporate a meta intelligent way of switching in and switching out of both the phases.
-
-The performance of our approach seems impressive given that it performed well on different complex systems with no specific conditioning being required.
-We comfortably outperformed common meta-heuristics and were also able to find minima of high dimensional global optimization test functions.
-Thus this approach may be well suited to large scale global optimization.
-In addition, surprisingly, we were able to obtain optimal parameter vectors for two different large scale systems biology models with a couple of orders fewer number of function evaluations as compared a hybrid approach.
-However it is quite possible that highly optimized versions of common meta heuristics may outperform us on these systems.
-This aspect is currently beyond the scope of this study.
-Our approach can also be combined with local derivative based searches to improve upon the accuracy of the solutions.
-
-%\begin{figure}[H]
-%\centering
-%\includegraphics[width=1.00\textwidth,height=0.7\textheight]{./figs/Figure_2_CoagulationNetwork.pdf}
-%\caption{Schematic of the extrinsic and intrinsic coagulation cascade.
-Inactive zymogens upstream (grey) are activated by exposure to tissue factor (TF) \cite{giesen1999blood} following vessel injury.
-Tissue factor and activated factor VIIa (FVIIa) form a complex that activates factor X (fX) and IX (fIX).
-FXa activates downstream factors including factor VIII (fVIII) and fIX.
-Factor V (fV) is primarily activated by thrombin (FIIa).
-In addition, we included a secondary fV activation route involving FXa.
-FXa and FVa form a complex (prothrombinase) on activated platelets that converts prothrombin (fII) to FIIa.
-FIXa and FVIIIa can also form a complex (tenase) on activated platelets which catalyzes FXa formation.
-Localized platelets are activated by external signals such as adenosine diphosphate (ADP) and thromboxane A2 (TXA2) or thrombin through protease-activated receptors (PARs) \protect\cite{coughlin2000thrombin} \protect \cite{coughlin1999protease} Thrombin also activates upstream coagulation factors, forming a strong positive feedback ensuring rapid activation.
-Tissue factor pathway inhibitor (TFPI) downregulates FXa formation and activity by sequestering free FXa and TF-FVIIa in a FXa-dependent manner.
-Antithrombin III (ATIII)  inhibits all proteases.
-Thrombin inhibits itself binding the surface protein thrombomodulin (TM) \protect\cite{esmon1989roles} The IIa-TM complex catalyzes the conversion of protein C (PC) to activated protein C (APC), which attenuates the coagulation response by the proteolytic cleavage of fV/FVa and fVIII/FVIIIa \protect\cite{esmon1989roles}}\label{fig-network}
-%\end{figure}
-
-\section*{Acknowledgements}
-This study was supported by an award from the Army Research Office (ARO \#59155-LS).
-\clearpage
-
-%\bibliography{References_v1}
-%\begin{thebibliography}{9999}
-\bibliography{References_v2}
+%(323 more and 287 less sensitive)
+%In this study, we rank-ordered the importance of the molecular components controlling the transition from ADPC to CRPC.
+%Toward this, we developed and analyzed an ensemble of mechanistic mathematical models of the signal transduction processes governing the 
+%transition from androgen dependent to castration resistant prostate cancer.
+%The objective of this study was to rank-order the importance of components in CRPC and ADPC using computational tools. 
+%We formulated and analyzed an ensemble of mechanistic mathematical models of the signal transduction processes in CRPC and ADPC 
+%that was an extension of our previous work .
+%In particular, we focused on the outlaw pathway and the possible crosstalk between RTK-dependent kinases and AR activation. 
+%First, we expanded the original model by adding ten  
+%Third, we added additional co-activators, such as cyclin E, and inhibitors of AR, such as
+%ODE models typically contain a large number of unknown model parameters, which must be estimated  
+%Due to uncertainty in parameter estimation and to account for differences between individual cells and cell sublines, an ensemble of parameter sets was used to model the system. 
+%Proteins that were not modeled at a gene regulatory level were generated using zeroth order reactions. 
+%All protein and RNA species were degraded with first order reaction rates. 
+%To obtain an ensemble of parameter sets, we used multiobjective optimization. 
+%Relative concentration values were quantified 
+%Training data came from experiments on LNCaP sublines. 
+%We used , previously developed in the lab \cite{Song2010}, to determine a diverse ensemble of models. 
+%C-33 cells, having undergone fewer than 33 passages, are considered AD while C-51 cells, having between 34 and 80 passages, are considered an intermediate prostate cancer cell line. 
+%C-81 (passage numbers beyond 80) display a CR phenotype \cite{Lin2000}. 
+%http://endo.endojournals.org/content/140/11/5054.long or http://endo.endojournals.org/content/140/11/5054.long#ref-19
 
@@
 
-\begin{table}[h]
-\centering
-\caption{Table with optimization settings and results for coagulation and different benchmarks using DdPSO.}
-\label{table:benchmark-problems}
-\resizebox{\columnwidth}{!}{
-\begin{tabular}{|cccccc|}
-\hline
-                        & Coagulation           & \multicolumn{4}{l}{Benchmarks}\\
- \hline
-                        & Experimental Data Set & B1           & B4          & Ackley 300D & Rastrigin 300D \\
- \hline
-Lower Bound             & 0.001.pnom            & 5.pnom       & 5.pnom      & 30          & 5.12           \\
-Upper Bound             & 1000.pnom             & 0.2.pnom     & 0.2.pnom    & -15         & -5.12          \\
-CPU Time                & 10.0835 hours         & 38.308 hours & 6.2 minutes & 2.8 seconds & 2.6 seconds    \\
-Function Evaluations    & 4000                  & 4000         & 4000        & 4000        & 4000           \\
-Initial Objective Value & 1.11837e+07           & 1.4275e+07   & 1.8536e+07  & 21.12       & 99.985         \\
-Final Objective Value   & 90456.9               & 4.96e+05     & 38.9375     & 0           & 0              \\
-Nominal Objective Value & 4.7785e+06            & 1.0986e+06   & 39.0676     & 0           & 0      \\
-\hline
-\end{tabular}
-}
-\end{table}
+\section*{Results}
+
+\subsection*{Estimating an ensemble of prostate signaling models.} 
+We modeled the integration of growth factor, cell cycle and hormone signaling pathways in AD and CR LNCaP cells (Fig. \ref{fg:NetworkSchematic}). 
+The signaling architecture was hand curated from over 80 primary literature sources in combination with biological databases.
+The model equations were formulated as a system of ordinary differential equations (ODEs), where biochemical reaction rates were modeled using mass action kinetics.
+ODEs and mass action kinetics are common modeling tools \citep{Wayman2013}, however, ODEs have the disadvantage of requiring estimates for unknown model parameters.
+Many techniques have been developed to estimate ODE model parameters, often from noisy and sparse experimental data \citep{Moles:2003dn}.
+Typically these identification problems are underdetermined, hence no unique parameter values can be estimated \citep{Villaverde:2014kl}.
+Thus, instead of estimating a single yet highly uncertain parameter set, we estimated an ensemble of possible parameter sets using the Pareto Optimal Ensemble Techniques (POETs) algorithm \citep{Song2010}.
+POETs uses a combination of simulated annealing and local optimization techniques coupled with Pareto optimality-based ranking to simultaneously optimize multiple objective functions.
+Starting from an initial best fit set, we estimated the 1687 unknown model parameters (1674 kinetic parameters and 13 non-zero initial conditions) using 43 \textit{in vitro} data sets taken from AD, intermediate and CR LNCaP cells (Table~\ref{objective_table}).
+Each of the 43 training data sets was a separate objective in the multiobjective calculation. 
+The training data were steady-state or dynamic immunoblots from which we extracted relative species abundance from their optical density profiles. 
+POETs sampled well over a million possible parameter sets, from which we selected N = 5000 sets for further analysis. 
+Over the ensemble, the coefficient of variation (CV) of the kinetic parameters spanned 0.59 - 5.8, with 33\% of the parameters having a CV of less than one (Fig. \ref{fg:Supp_CV}). 
+As a control, we also performed simulations for R = 100 random parameter sets to compare against the ensemble generated by POETs.
+
+The ensemble of PCa models recapitulated training data in both AD and CR cell lines with only two experimentally justified parameter changes (Fig. \ref{fg:ColorPlot} and Fig. \ref{fg:Training}). 
+Data from the LNCaP clones C-33 (dependent), C-51 (intermediate), and C-81 (resistant) \cite{Lin2000,Igawa2002,Horoszewicz1983} along with the CR LNCaP cell lines LNCaP-Rf \cite{Murillo2001}, LNCaP-AI \cite{Chen2011} and LNAI \cite{Graff2009} were used for model identification.
+To simulate the effective difference between LNCaP cell lines, the parameter controlling the maximum rate of PAcP gene expression was scaled by 0.1 and 0.5, respectively, for the C-81 and C-51 cell-lines compared to C-33.
+This modification was based upon steady-state PAcP data from the three LNCaP clones \cite{Lee2003}. 
+Similarly, the expression of p16INK4 was adjusted in accordance with the study of Lu \textit{et al.} \citep{Lu1997}. 
+These two parameters were the only adjustable parameter differences between AD and CR cells. 
+To simulate an increased mTOR activation in the presence of a DHT stimulus, we added a first order activation term for mTOR activation with a DHT stimulus.
+Androgens have been shown to increase expression of proteins involved in cellular metabolism, which may lead to an increase in mTOR activation \cite{Xu2006}.
+The model fit 36 of the 43 training objectives for greater than 40\% of the ensemble members (Fig. \ref{fg:ColorPlot}A). 
+Conversely, only 10 of the 43 training objectives were captured with the random parameter control (Fig. \ref{fg:ColorPlot}B). 
+The model captured the crosstalk between RTK activation and androgen action (Fig. \ref{fg:Training}). 
+The model described DHT-induced PSA expression in both C-33 (Fig. \ref{fg:Training}A) and C-81 (Fig. \ref{fg:Training}B) cells. 
+Interestingly, simulations with the HER2 inhibitor AG879 recapitulated decreased PSA expression in C-81 cells (Fig. \ref{fg:Training}C) in the absence of androgen stimulation.
+AR action decreased the PAcP mRNA message (Fig. \ref{fg:Training}D), presumably leading to increased HER2 activity. 
+The model also recapitulated the integration of androgen action with AR expression, G1/S cell cycle protein expression and AKT phosphorylation. 
+For example, the model captured AR-induced AR expression following a DHT stimulus (Fig. \ref{fg:Training}H).
+Conversely, the transcription factor E2F inhibits AR transcription in LNCaP cells (Fig. \ref{fg:Training}I). 
+Other cell cycle proteins were also integrated with androgen action.
+For example, the cyclin D1 abundance increased in CR compared to AD cells in the absence of androgen (Fig.
+\ref{fg:Training}E), while DHT induced p21Cip1 expression in C-33 cells (Fig.
+\ref{fg:Training}F).
+The level of phosphorylated AKT was also increased in higher passage number cells (Fig. \ref{fg:Training}G). Taken together, [FINISH ME].
+
+%We used a single model architecture to describe the behavior of both AD and CR LNCaP cells, 
+%Two new additions to the model were the cell cycle protein p21Cip1 and transcriptional regulation of AR. 
+%We were able to capture training data for both mechanisms. 
+%AR transcriptionally upregulates its own expression \cite{Lin2003}\cite{Grad1999}. 
+%We captured the increase in AR concentration due to AR activation, and thus increased transcriptional activity, upon  
+%Higher basal AKT phosphorylation leads to increased activation of the PI3K/AKT pathway in PCa cells with less androgen dependence \cite{Lin2003}. 
+%\cite{Davis2006}
+
+%An additional protein we considered in validation that was present in training was phosphorylated SHC-transforming protein 1 (Shc).
+%Simulations predicted a similar increase in Shc phosphorylation after DHT stimulus (Fig. \ref{fg:Validation}D) \cite{Veeramani2005Onc}. 
+%The Prostate Cancer Working Group (PCWG) updated outcome measurements for castration resistant prostate cancer studies in 2008 \cite{Scher2008}. 
+%The primary endpoint in prostate cancer studies is overall survival, while PSA response is often a secondary endpoint. 
+%Figure \ref{fg:Validation}I shows fold change of PSA of ensemble members due to the application of sorafenib.
+
+%(Fig. \ref{fg:Validation}H)
+
+\subsection*{Validation simulations revealed missing network structure.} 
+The model was validated against 29 \textit{in vitro} and four \textit{in vivo} studies (Table~\ref{prediction_table}). 
+For 15 of the 29 cases, greater than 40\% of the ensemble was qualitatively consistent with the experimental data (Fig. \ref{fg:ColorPlot}C). 
+However, for the random parameter control, only 7 of the 29 cases were satisfied (Fig. \ref{fg:ColorPlot}D). 
+We correctly predicted positive feedback between HER2 auto-activation and androgen action (Fig. \ref{fg:Validation}A and Fig. \ref{fg:Validation}B). 
+We also captured the dose-dependence of AR abundance on DHT (Fig. \ref{fg:Validation}C). 
+In addition to the cell line studies, we simulated the outcome of enzalutamide, lapatinib, and sorafenib clinical trials in AD and CRPC patients. 
+The trial end points were the reduction in PSA expression relative to an untreated baseline.  
+Enzalutamide acts on AR by inhibiting its nuclear translocation, DNA binding, and coactivator recruitment \cite{Scher2012}. 
+In the enzalutamide trial, 54\% of the patients that received the drug showed a PSA decline of $\geq$ 50\% while 25\% showed a decline $\geq$ 90\%. 
+We simulated enzalutamide exposure by reducing the rate constants governing activated AR binding to nuclear importer, cyclin E, and CDK6 to 1\% of their initial values. 
+Consistent with the trial, 62\% of ensemble members showed a $\geq$ 50\% decline in PSA abundance, while 14\% showed a $\geq$ 90\% decline (Fig. \ref{fg:Validation}G). 
+The second trial we simulated involved exposure of CRPC patients to sorafenib. 
+Sorafenib is a kinase inhibitor with activity against Raf, vascular endothelial growth factor receptor (VEGFR), platelet-derived growth factor receptor (PDGFR), c-kit and c-Ret \cite{Dahut2008}. 
+We considered only the effects of sorafenib on the protein kinase Raf, as VEGFR, PDGFR, c-kit and c-Ret were not included in the model. 
+None of the 22 patients in the sorafenib study showed a PSA decline of $>$ 50\%. 
+However, our simulations showed that approximately 55\% of the ensemble members had a PSA decline of $\geq$ 50\%. 
+The last drug we considered was lapatinib, an inhibitor of epidermal growth factor receptor (EGFR) and HER2 tyrosine kinase activity \cite{Liu2013}. 
+Two lapatinib drug trials were considered: one in which patients had CRPC and one in which patients had biochemically relapsed ADPC \cite{Whang2013,Liu2013}. 
+In the CRPC lapatinib drug trial, two of the 21 enrolled patients had a PSA response $\geq$ 47\%  \cite{Whang2013}. 
+For the CRPC case, our model showed 26.5\% of ensemble members with a PSA response $\geq$ 47\%. 
+Of the 35 patients enrolled in the ADPC lapatinib study, no PSA decreases was observed \cite{Liu2013}. 
+In this case, our model showed 9.2\% of ensemble members with a PSA response $\geq$ 50\%.
+Although no response to lapatinib was seen in ADPC clinical trials, \textit{in vitro} AD LNCaP experiments showed decreased PSA expression in response to lapatinib, most notably with the addition of DHT \cite{Liu2005}.
+
+Validation and training failures suggested the original signaling architecture was missing critical components. 
+Several of the failed training and validation simulations involved the response of the network to epidermal growth factor (EGF) stimulation. 
+For example, Chen \textit{et al.} showed that HER2 phosphorylation increased within five minutes following EGF stimulation of LNCaP-AI cells \citep{Chen2011}. 
+We predicted no connection between HER2 phosphorylation and EGF stimulation on this short timescale (Fig. \ref{fg:Validation}E). 
+Interestingly, we initially neglected the heterodimerization of HER2 with other ErbB family members in order to simplify the model.
+However, Chen \textit{et al.} suggested that HER2-EGFR heterodimerization could be an important factor in EGF-driven activation of HER2 \cite{Chen2011}. 
+We tested this hypothesis by developing a new model that included HER2 and EGFR heterodimerization. 
+We set the rate constants governing the assembly of HER2/EGFR heterodimers equal to EGFR homodimer assembly; all other parameters were unchanged.
+This was a reasonable first approximation, as the affinity of HER2/EGFR heterodimerization and EGRF homodimerization is thought to be similar \cite{Hendriks2003}. 
+With the inclusion of HER2-EGFR heterodimerization, we qualitatively fit the EGF-induced HER2 activation case and improved our training for experiments that involved an EGF stimulus, e.g., cyclin D mRNA and protein abundance following an EGF stimulus in C-33 cells (Fig.
+\ref{fg:ColorPlot}A and C, white pixels and Fig.
+\ref{fg:Exps_NewModelvsOldModel}).
+
+%in diverse environments (i.e., DHT, no DHT, drug) 
+%Stimuli included the presence or absence of DHT, presence of therapeutics, and also EGF. 
+%PIP3 recruits AKT and PDK1 to the membrane where PDK1 activates AKT \cite{Vivanco2002}. 
+%Thus, the PIP3 and AKT interaction is required in the model for AKT phosphorylation to occur.
+
+\subsection*{Sensitivity analysis identified differentially important features of the prostate architecture.} 
+We used sensitivity analysis to identify important signaling components in AD versus CR cells (Fig. \ref{fg:Sensitivity}). 
+We calculated first order steady-state sensitivity coefficients under different stimuli for 500 parameter sets selected from the ensemble. 
+Signaling components were rank-ordered based upon analysis of their sensitivity coefficient values. 
+In the presence of DHT, the sensitivity profile was similar for AD versus CR cells, with only a few differences (Fig. \ref{fg:Sensitivity}B). 
+The top 2\% of sensitive species, regardless of androgen dependence, involved components from the MAPK and PI3K pathways. 
+In particular, activated Ras, Raf, phosphorylated MEK, as well as PIP3 localized AKT, phosphorylated AKT, and PI3K were sensitive in both AD and CR cells. 
+Species involving PAcP and p16INK4 were more sensitive in AD cells, which was expected since the expression of these two proteins were the only parameters changed between AD and CR cells. 
+Other species such as E2F, cyclin E, and DHT-activated AR were also more sensitive in AD cells. 
+On the other hand, HER2-Grb2-Gab activation of PI3K and AKT inhibition of Raf were more sensitive in CR cells.
+
+The importance of signaling components varied with androgen dependence in the absence of DHT (Fig. \ref{fg:Sensitivity}A). 
+There were 609 statistically significant shifts in species sensitivity (318 more and 291 less sensitive) between CR and AD cells in a non-androgen environment. 
+However, only 108 of these shifts were greater than one standard deviation above the mean. 
+In CR cells, HER2 activation of ERK and PI3K was more sensitive, as was AR activation through the MAPK pathway. 
+This was expected, as outlaw pathway activity was elevated in castration resistant cells. 
+Species in the MAPK pathway were in general more sensitive in CR cells (128 out of 140 significant), with all forms of sPAcP more robust in CR cells.
+On the other hand, infrastructure pathways encoding transcription and translation were more sensitive in AD cells.
+PSA  and cyclin D1b (mRNA and mRNA complexes) were the only species involved in translation that were more robust in AD cells (14 out of 116).
+The transcription factor, E2F was more fragile in AD cells, while the transcription factors ETS and AP1 were more robust. 
+ETS and AP1 are activated by phosphorylated ERK, and ETS is also activated by active PKC \cite{Wilkinson2000,Lindemann2003}. 
+E2F is deactivated through binding to Rb, which is deactivated by cyclin D1 and CDK phosphorylation \cite{Lapenna2009}. 
+The model also included AP1 suppression of AR transcriptional activity (more sensitive in CR) \cite{Sato1997}, as well as inhibition of transcription of the AR gene by E2F (more sensitive in AD) \cite{Davis2006}.
+Species in the PI3K pathway that were more fragile in AD cells included Rheb and TOR complexes. 
+Interestingly, these species were included as the last step in the PI3K pathway prior to translation, with the phosphorylation of 4E-BP1 by TOR being considered the beginning of translation in this model.
+This again indicates that in the absence of DHT general translation is more fragile in AD cells.
+
+There was a large shift in sensitive species between an androgen and a non-androgen environment in both AD and CR cell lines (Fig. \ref{fg:Sensitivity}C and Fig. \ref{fg:Supp_Sensitivity}). 
+Of the 664 statistically significant shifts in AD cells, 288 were more sensitive between androgen versus non-androgen environments. 
+However, only 119 shifts were larger than one standard deviation above the mean. 
+Unsurprisingly, AR activation through DHT binding, with and without coactivators, in a DHT environment was more sensitive, as was AR inhibition of PAcP transcription (repressed by AR in the model). 
+Species further upstream, such as HER2 activation of the MAPK and PI3K/AKT pathways, were also more sensitive in a DHT environment. 
+Cell cycle species that were more fragile in the presence of DHT, included complexes involving p21Cip1 and CDC25A.
+In a non-androgen environment, basal transcription (68 out of 72) and translation (114 out of 120) were more fragile. 
+Other fragile species in the absence of DHT included Rb, E2F, Sam68, cyclin D1a complexes, phosphatases in the MAPK pathway, Rheb complexes, and TOR complexes.
+
+We also considered the sensitivity of CR cells following the application of the AR inhibitor enzalutamide in the presence of DHT (Fig. \ref{fg:Sensitivity}D). 
+Species which were more sensitive in an androgen environment with enzalutamide included cytosolic AR, cPAcP, and p21Cip1. 
+As we would expect, AR species found in the nucleus and/or bound to coactivators, were more robust in the absence of enzalutamide. 
+The top two percent of sensitive species with and without enzalutamide were conserved. 
+In a CR cell, enzalutamide had no effect on the sensitivity of PI3K/AKT species as well as many MAPK species (ERK, Raf, and MEK). 
+Next, we looked at the effect of enzalutamide on a CR cell in both a non-androgen and DHT environment (Fig. \ref{fg:Supp_Sensitivity}). 
+More sensitive species in a non-androgen environment included dimerized HER2, ERK, and PAcP. 
+Species which were more robust in the non-androgen environment included, AR activated by DHT, AKT, p70, and AR bound to HSP.
+The results of our sensitivity analysis indicate that instead of inhibiting solely the AR pathway (enzalutamide), a combination therapy targeting the PI3K or MAPK pathways in addition to AR may be more effective.
+
+\subsection*{Experimental results confirm the need for dual therapies in prostate cancer.}
+
+The results from our sensitivity analysis indicate that instead of inhibiting solely the AR pathway (enzalutamide), a combination therapy is necessary. 
+To test this hypothesis we used the well characterized ADPC cell line LNCaP as well a LNCaP derived CRPC cell line C4-2 \cite{Thalmann1994}. 
+Three inhibitors were used: the AR inhibitor MDV3100 (enzalutamide), the Raf kinase inhibitor sorafenib, and the PI3K inhibitor LY294002. 
+In both cell lines, inhibition of either the AR or MAPK pathway appears to promote activation of the PI3K pathway, as seen by the increase in pAKT (S473) (Fig. \ref{fg:Experiment}A). 
+The addition of the PI3K inhibitor, LY294002, alone or in combination diminishes PI3K activity. 
+The inhibition of PI3K alone, led to an increase in AR expression in both LNCaP and C4-2 cell lines. 
+Since, AR transcriptionally upregulates its own expression \cite{Lin2003}\cite{Grad1999}, this may indicate an increase in AR activity. 
+The ribosomal protein pS6 was completely inhibited only in the presence of the PI3K inhibitor LY294002. 
+Cell viability results show a large decrease in cell viability at 72 hrs in the dual inhibition cases as well as the triple inhibition case for both LNCaP and C4-2 cell lines (Fig.
+\ref{fg:Experiment}B).
+In both cell lines, MDV3100 (10 $\mu$M), has only a modest effect on cell viability versus control (DMSO). 
+Figure \ref{fg:Experiment}C shows cell viability of both cell lines in varying concentrations of inhibitors at 24 hrs.
+
+\subsection*{Robustness analysis identified key regulators of prostate cancer.} 
+Robustness analysis was conducted for 80 proteins to quantify the effects of amplifying or knocking down key model components in both AD and CR cells. 
+Gene expression parameters were altered by a factor 10, 0.5, and 0 for knock-in, knock-down, or knock-out perturbations, respectively.
+We calculated the effect of these perturbations on different protein markers, such as PSA, AR, and cyclin D. 
+A knock-out of Raf, MEK or ERK showed an overall increase in cyclin D levels in CR cells (Fig. \ref{fg:Supp_Robustness}). 
+This was unexpected and we saw a similar increase in cyclin D due to the knock-in of Raf, MEK or ERK. 
+We found that individual ensemble members showed different response to a Raf knock-out, in both cyclin D concentration and PSA concentration. 
+Of the 500 ensemble members, 126 members saw an increase in PSA concentration and 62 members saw an increase in cyclin D concentration due to the knock-out of Raf (Fig. \ref{fg:Robustness}). 
+We saw three distinct regions: (1) increase in PSA concentration, (2) increase in cyclin D concentration, and (3) a decrease in both PSA and cyclin D. 
+We explored the flux vectors of the outlying parameter sets to understand the mechanistic effect of Raf knock-out on PSA and cyclin D. 
+Outlying parameter sets in region 1 displayed high activation of PI3K through HER2 signaling as well as high association of AP1 with AR. 
+AP1 binds and suppresses AR transcriptional activity in LNCaP cells \cite{Sato1997}. 
+Knocking out Raf lowered AP1 levels and, therefore, freed AR for increased transcription of PSA.
+In region 2, parameter sets also had high activation of PI3K through HER2. 
+They also had higher association of E2F with Rb and cyclin D1a with AR. 
+Cyclin D levels in this region increase due to an increase in E2F levels caused by the Raf knock-out. 
+Parameters in region 3 have high association of TOR. 
+Interestingly, the drug sorafenib, a multi-kinase inhibitor that has activity against Raf, showed no measurable PSA decline in prostate cancer patients in clinical trials \cite{Dahut2008}. 
+The robustness analysis showed that network perturbation can result in unexpected responses due to cell-to-cell heterogeneity in gene expression. 
+These outlying cell types could be critical for understanding when designing drug targets and combination therapies.
+
+%The results of the model suggest that an inhibition of either the PI3K pathway or the MAPK pathway in combination with an AR inhibitor as a possible therapy for CRPC. 
+%Sensitivity analysis revealed no change in the top sensitive species in the presence or absence of the AR inhibitor, enzalutamide.
+PI3K/AKT and MAPK species continued to fall in the top two percent of sensitive species.
+%A study by Carver \textit{et al.} looked at dual inhibition of AR and PI3K signaling in LNCaP cells and in a Pten-deficient murine prostate cancer model \cite{Parker2013}. 
+%Using both the PI3K inhibitor, BEZ235, and the AR inhibitor, MDV3100 (enzalutamide), the group saw a drastic decrease in the total number of cells. 
+%Each inhibitor on it’s own had a much smaller effect on total cell number. 
+%They saw an increase in the cell death marker, c-PARP, in the dual inhibition case. 
+%The group hypothesized that AKT inhibition leads to increased AR signaling activity through increased protein concentrations of HER3. 
+%On the other hand, AR inhibition leads to increased AKT activity due to the down regulation of PHLPP, a protein phosphatase that regulates AKT. 
+%For simplicity, the HER3 pathway and also cell death were not included in the model. 
+%Dual knock-out studies of PI3K and AR in our model show no additive effect on any cell cycle proteins through the dual knock-out. 
+%The pathways in our model appear to be uncoupled and therefore no synergy is shown in the dual knock-out case. 
+%This could indicate that the combined decrease in cell population is entirely due to cell death. 
+%The Carver \textit{et al.} study did not consider cell cycle proteins or cell growth. 
+%Our model does show a decrease in cell cycle proteins in the PI3K knock-out as well as in the dual knock-out case. 
+%This result seems to be consistent with the decreased cell growth in the PI3K knock-out case which is not dependent on cell death, as c-PARP levels are low. 
+%The decrease in cell cycle proteins is due to a decrease in general translation, including free eIF4E levels and activated 40S ribosome subunit. 
+%The decrease in p70 (S6) activation due to inhibition of PI3K is shown in both the model and in the Carver \textit{et al.} study (Supplementary Figure), indicating this result is due to the PI3K pathway.
+%It would be interesting to repeat the experiment looking at cell cycle proteins and also performing the experiment in CR LNCaP cells, instead of AD cells.
+
+\section*{Discussion}
+In this study, we developed a population of mathematical models describing growth factor and hormone signal integration in androgen dependent, intermediate and resistant prostate cancer cells. 
+These models described the regulation of androgen receptor expression and activation through androgen binding as well as a ligand-independent, MAPK-driven mechanism referred to as the outlaw pathway. 
+An ensemble of model parameters was estimated using 43 steady-state and dynamic data sets taken from androgen dependent, intermediate and independent LNCaP cell lines using multiobjective optimization.
+Further, we tested the predictive power of the model by comparing model predictions against 33 novel data sets (including four \textit{in vivo} drug studies) not used during model training. 
+The model ensemble captured 84\% of the training data and 52\% of the validation data relative to 23\% and 24\% for a random control population. 
+Interestingly, during the the initial round of parameter estimation, we identified several potentially missing structural components not present in the original connectivity. 
+One such component, EGF-induced HER2/EGFR heterodimerization, was added to the current generation model. 
+Inclusion of this structural component significantly improved both training and validation performance using the same rate constants as the EGFR-homodimer case (no additional parameter fitting). 
+We then analyzed the population of signaling models, using both sensitivity and robustness analysis, to identify the critical components controlling network performance in a variety of conditions.
+
+%Hendrick's \textit{et al.} hypothesized that HER2 heterodimerization amplified signals by allowing just one molecule of EGF to activate two receptors \cite{Hendriks2003}. 
+%In our case, we see an amplified cyclin D concentration due to an equal amount of EGF stimulus. %
+%Model connectivity involving EGF and the downstream effect on AR and AR-regulated genes also came up as an open question. 
+%Whether EGF downregulates or upregulates AR activity is debatable and the biology is still unknown \cite{Cai2009,Cinar2005,Lin2003,Lin2001PNAS, Hakariya2006,Gregory2004,Ponguta2008,Guo2006,Weber2004}.
+%The current model assumed no direct effect of EGF on AR, and instead we saw increased AR activity through increased HER2 activation. 
+%Further studies will need to be done to update this section of the model.
+
+In addition, three of the validation cases involved the effect of EGF on AR and AR-activated genes, i.e., PSA. 
+Cai \textit{et al.} showed decreased expression of endogenous AR as well as androgen-regulated PSA in AD LNCaP cells due to an EGF stimulus \cite{Cai2009}. 
+Cinar \textit{et al.} also showed decreased AR protein expression due to EGF, an effect reversed by the mTOR inhibitor, Rapamycin \cite{Cinar2005}. 
+Model simulations show either the opposite trend or no effect due to EGF stimulus (Fig. \ref{fg:Validation}F) \cite{Cinar2005}. 
+These results suggest missing network structure. 
+From additional literature searches, the inhibition of AR activation through EGF is still an open question, with many groups debating the biology involved, predominately in the PI3K/AKT pathway. 
+Lin \textit{et al.} found that in low passage number LNCaP cells (C-33), AKT negatively regulates AR by destabilizing it and marking it for ubiquitylation. 
+In high passage number LNCaP (C-81), AKT levels are high which contribute to AR stability and less degradation \cite{Lin2003}. 
+Wen \textit{et al.} showed that HER2 could induce AKT activation and LNCaP cell growth in the presence and absence of androgen \cite{Wen2000}. 
+Another study shows AKT phosphorylation of AR at S213 and S790 suppresses AR transactivation and AR-mediated apoptosis of LNCaP \cite{Lin2001PNAS}. 
+The study from Cai \textit{et al.} showed the reduction in AR was not due to degradation or PI3K/AKT signaling, but instead was due to decreased AR mRNA levels \cite{Cai2009}. 
+They found that AR protein levels in CR cells were not affected by EGF. 
+Others though have found that PSA expression, even in C-81 cells, is decreased by EGF \cite{Hakariya2006}. 
+In other prostate cell lines, EGF has been shown to increase AR transactivation \cite{Gregory2004,Ponguta2008}. 
+The MAPK pathway, which is downstream of EGFR, may also enhance AR responses to low levels of androgen \cite{Guo2006,Weber2004}. 
+Due to the discrepancies in the literature, experiments should be performed before adding additional network connectivity to the model.
+
+The population of PCa models was analyzed using sensitivity analysis to identify key signaling components and processes in both AD and CR cells. 
+There was very little difference between sensitive and robust components in AD versus CR cells in the presence of androgen. 
+MAPK and PI3K pathway components were consistently ranked in the top 2\% of sensitive species in the presence of androgen for both AD and CR cells.
+On the other hand, cell cycle species, such as cyclin D-CDK4/6 complexes bound to cell cycle inhibitors (p27Kip1, p21Cip1, p16INK4), were consistently robust.
+However, this profile changed considerably in the absence of androgen.
+The activation of PI3K and ERK by HER2 dimerization and autophosphorylation was significantly more important in CR versus AD cells.
+Interestingly, AR activation by ERK was also more sensitive in CR versus AD cells in the absence of androgen.
+Lastly, although AR-regulated transcriptional processes were equally sensitive between the cell types, general translational and transcriptional components were more robust in CR versus AD cells.
+This evidence supports the current theory that CR cells will still respond to androgen and, thus, AR is still an active target in therapeutic against CRPC \cite{Karantanos2013}. 
+Supporting the argument that AR can be activated in the absence of androgens by MAPK activation \cite{Feldman2001}. 
+Advanced prostate cancers often have higher levels of E2F and other transcription factors \cite{Davis2006}.
+Interestingly, E2F was more sensitive in AD cells, while other transcription factors (ETS and AP1) were more robust. 
+The drug enzalutamide had no effect on the top 2\% of sensitive species. 
+Species in the PI3K/AKT and MAPK pathways in the presence of enzalutamide were still highly sensitive. 
+The application of enzalutamide, increased sensitivity of AR species found outside of the nucleus as well as PAcP species.
+Robustness analysis indicated diverse effects of Raf knock-out on PSA and cyclin D concentrations. 
+Clinical studies of sorafenib, a multi-kinase inhibitor that has activity against Raf, showed increase PSA levels in patients \cite{Dahut2008}. 
+Our results indicate that cell-to cell heterogeneity in gene expression can play a significant role in determining cell response. 
+Thus, combination therapies need to be considered even in the case of a Raf knock-out.
+
+The results of the model suggest that an inhibition of either the PI3K pathway or the MAPK pathway in combination with an AR inhibitor as a possible therapy for CRPC. 
+Sensitivity analysis revealed no change in the top sensitive species in the presence or absence of the AR inhibitor, enzalutamide. 
+PI3K/AKT and MAPK species continued to fall in the top two percent of sensitive species. 
+A study by Carver \textit{et al.} looked at dual inhibition of AR and PI3K signaling in LNCaP cells and in a Pten-deficient murine prostate cancer model \cite{Parker2013}. 
+Using both the PI3K inhibitor, BEZ235, and the AR inhibitor, MDV3100 (enzalutamide), the group saw a drastic decrease in the total number of cells. 
+Each inhibitor on it’s own had a much smaller effect on total cell number. They saw an increase in the cell death marker, c-PARP, in the dual inhibition case. 
+The group hypothesized that AKT inhibition leads to increased AR signaling activity through increased protein concentrations of HER3. 
+On the other hand AR inhibition leads to increased AKT activity due to the down regulation of PHLPP, a protein phosphatase that regulates AKT. 
+For the simplicity of this model, the HER3 pathway and also cell death were not included in the model. 
+Dual knock-out studies of PI3K and AR in our model show only a slight additive effect on cell cycle protein cyclin D through the dual knock-out compared to solely inhibiting PI3K (Fig.
+\ref{fg:Dual_Knockout}).
+The triple case, with PI3K, AR, and RAF knock-outs, also only showed a slight additive effect (Fig. \ref{fg:Triple_KO}).
+%The pathways in our model appear to be uncoupled and therefore little synergy is shown in the dual knock-out case. 
+This could indicate that the combined decrease in cell population due to the dual inhibition of PI3K and AR is entirely due to cell death. 
+The Carver \textit{et al.} study did not consider cell cycle proteins or cell growth. 
+Our model does show a decrease in cell cycle proteins in the PI3K knock-out as well as in the PI3K and AR dual knock-out case in some ensemble members (Fig. \ref{fg:Dual_Knockout}). 
+This result seems to be consistent with the decreased cell count in the PI3K knock-out case which is not dependent on cell death, as c-PARP levels are low. 
+The decrease in cell cycle proteins in the model is due to a decrease in general translation, including free eIF4E levels and activated 40S ribosome subunit. 
+The decrease in p70 (S6) activation due to inhibition of PI3K is shown in both the model and in the Carver \textit{et al.} study , indicating this result is due to the PI3K pathway (Fig.
+\ref{fg:Supp_KO}).
+Our experimental results confirm the Carver \textit{et al.} study in that a dual inhibition of AR and PI3K signaling led to a more prominent decrease in cell viability then each of the inhibitors alone in LNCaP cells.
+We extended the study to look at the addition of a third inhibitor, sorafenib, that inhibits Raf kinase in the MAPK pathway and an additional cell line, C4-2, which was CR.
+In both cell lines, there was not a significant decrease in cell viability between the three dual inhibitor cases and the triple inhibition case at 74 hours, indicating that a dual inhibition (PI3K/AR, AR/MAPK, or PI3K/MAPK) may be a sufficient treatment.
+
+%It would be interesting to repeat the experiment looking at cell cycle proteins and also performing the experiment in CR LNCaP cells, instead of AD cells.
+
+%and  in the CR CWR-R1 prostate cell line.
+%a steroid receptor coactivator family member along with other important AR coregulators such as 
+%transcriptional intermediary factor 2 (TIF-2)
+
+The PCa signaling architecture was assembled after extensive literature review and hand curation of the biochemical interactions.
+However, there are a number of areas where model connectivity could be refined, e.g., the regulation of AR phosphorylation.
+We assumed a single canonical activating AR phosphorylation site (S515), with ERK being the major kinase and PP2A or PP1 being the major phosphatases responsible for regulating this site.  
+MAPK activation following EGF treatment increases AR transcription and cell growth, partially through AR phosphorylation on MAPK consensus site S515 \cite{Ponguta2008}.
+However, there are at least 13 phosphorylation sites identified on AR, with phosphorylation at six of these being androgen induced \cite{Gioeli2012}. 
+Moreover, other kinases such as AKT, protein kinase C (PKC) family members, as well as Src-family kinases can all phosphorylate AR in prostate cells \cite{Guo2006,Ponguta2008}. 
+For example, AKT activation leads to AR  phosphorylation at both S213 and S791 (however, the role of these sites remains unclear) \cite{Wen2000,Lin2001PNAS,Taneja2005,Lin2003}.  
+AKT effects on AR may also be passage number dependent, with AKT repressing AR transcription in low passage number cells and enhancing transcription in higher passage numbers \cite{Lin2003}. 
+Androgen independent phosphorylation of AR by Src family kinases (not currently in model) at Y534 \cite{Guo2006} or by protein kinase C (PKC) family members at the consensus site S578 could also be important for understanding the regulation of AR activity.
+A second area we will revisit is the gene expression program associated with androgen action, and particularly the role of AR coregulators.
+Currently, we included only two AR coactivators, cyclin E and CDK6 \cite{Yamamoto2000,Lim2005} and three corepressors AP1, Cdc25A, and cyclin D1a in the model \cite{Sato1997,Chiu2009,Petre-Draviam2003}.
+However, there are at least 169 proteins classified as potential AR coregulators \cite{Heinlein:2002qy,Heemers2007} with many of these being differentially expressed in malignant cells.
+For example, the expression of steroid receptor coactivator-1 (Src-1) and transcriptional intermediary factor 2 (Tif-2), both members of the steroid receptor coactivator family, are elevated in prostate cancer \cite{Gregory2001,Gregory2004}.
+Src-1 is phosphorylated by MAPK and interacts directly with AR to enhance AR-mediated transcription \cite{Heemers2007}. 
+Another class of potentially important AR coregulators are the cell cycle proteins Cdc25 and Rb.
+Unlike Cdc25A, Cdc25B (not in the model) can act as an AR coactivator leading to enhanced AR transcription activity \cite{Ngan2003}. 
+The Rb protein, in addition to being a key cell cycle regulator, has been shown to be an AR coactivator in an androgen-independent manner in DU145 cells \cite{Yeh1998}.
+However, there is some uncertainty about the role of Rb as Sharma \textit{et al.} showed that Rb decreased AR activation in multiple prostate cancer cell lines and xenografts \cite{Sharma2010}. 
+Forkhead proteins have also been shown to activate as well as repress AR function. 
+In prostate cancer, AKT suppresses AFX/Forkhead proteins, which diminishes expression of AFX target genes, such as p27Kip1 \cite{Graff2000, Brunet1999, Medema2000, Takaishi1999}. 
+Lastly, undoubtedly there are several other signaling axes important in PCa, such as cytokine or insulin- and insulin-like growth factor signaling \citep{Cardillo2003,Heinlein2004,Tam2007,Seaton2008}.
+Understanding the pathways associated with these signals and how they relate to the current model, may give us a more complete picture of CR prostate cancer.
+
+%Lastly, with an extended model, it would be beneficial to include possible biomarkers for drug outcomes, as currently there is no biomarker that is used to select treatments for patients  \cite{Sartor2013}.
 
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=1.0\textwidth,height=0.5\textheight]{./figs/Figure_1_Algorithm.pdf}
-\caption{Multi Swarm Particle Swarm Optimization with Dynamically Dimensioned Search.
-\textbf{A}: Each particle represents an N dimensional parameter vector.
-Particles are randomly initialized and grouped into different sub-swarms.
-Within each swarm the magnitude and direction of the movement a particle is influenced by the position of the best particle and also by its own experience.
-After every $\mathbf{g}$ number of function evaluations the particles are mixed and randomly assigned to different swarms.
-At the end of the evaluations ($\mathcal{FR}$) assigned to swarm search, the global best particle (orange color) amongst all sub-swarms is chosen as the candidate parameter vector for Dynamically Dimensioned Search \textbf{B}: The candidate vector does a greedy search in a dynamic neighborhood.
-The search neighborhood is dynamically adjusted by varying the number of dimensions that are perturbed (in black) in each evaluation step.
-The probability that a dimension is perturbed decreases as the number of function evaluations increase.
-Thus as the evaluations increase the optimality of the solution is preserved.
-}\label{fig-algorithm}
+\section*{Materials and Methods}
+
+\subsection*{Prostate model signaling architecture.}
+We modeled the expression, translation and post-translational modifications of key components of the signaling architecture.  
+The model, which consisted of 780 protein, lipid or mRNA species interconnected by 1674 interactions, was a significant extension to our previous model \cite{Tasseff2010} in several important areas. 
+First, we included well-mixed nuclear, cytosolic, membrane and extracellular compartments (including transfer terms between compartments).
+Next, we expanded the description of growth factor receptor signaling, considering both homo- and heterodimer formation between ErbB family members and the role of cellular and secreted prostatic acid phosphatase (cPAcP and sPAcP, respectively).
+Both forms of PAcP were included because cPAcP downregulates HER2 activity, while sPAcP promotes modest HER2 activation \cite{Veeramani2005}. 
+Third, we expanded the description of the G1/S transition of the cell cycle (restriction point). 
+The previous model used the abundance of cyclin D as a proliferation marker, but did not include other proteins or interactions potentially important to the restriction point.
+Toward this shortcoming, we included cyclin E expression (and its role as a coregulator of androgen receptor expression), enhanced the description of cyclin D expression and the alternative splicing of cyclin D mRNA (including the role of the splice variants in androgen action), included the Rb/E2F pathway as well as E2F inhibition of androgen receptor expression \cite{Davis2006}, and the cyclin-dependent kinases cyclin-dependent kinase 4 (CDK4) and cyclin-dependent kinase 6 (CDK6).
+We also included key inhibitors of the restriction point including cyclin-dependent kinase inhibitor 1 (p21Cip1), cyclin-dependent kinase inhibitor 1B (p27Kip1), and cyclin-dependent kinase inhibitor 2A (p16INK4) \cite{Sherr1999}.
+Fourth, we enhanced the description of growth factor induced translation initiation.
+One of the key findings of the previous model was that growth factor induced translation initiation was globally sensitive (important in both androgen dependent and independent conditions). 
+However, the description of this important subsystem was simplified in the previous model. 
+Here, we expanded this subsystem, using connectivity similar to previous study of Lequieu \textit{et al.} \cite{Lequieu:2011eu}, and re-examined the importance of key components of this axis, such as mammalian target of rapamycin (mTOR), phosphatidylinositide 3-kinase (PI3K) and AKT.
+Lastly, we significantly expanded the description of the role of androgen receptor. 
+The previous model assumed constant AR expression, consistent with studies in androgen dependent and independent LNCaP sublines \cite{Lee2003}.
+However, other prostate cancer cell lines vary in their AR expression \cite{Sobel2005}. 
+Thus, to capture androgen signaling in a variety of prostate cancer cells, we included the transcriptional regulation governing androgen receptor expression, updated our description of the regulation of androgen receptor activity and androgen action (gene expression program driven by activated androgen receptor).
+At the expression level, we included AR auto-regulation in combination with the co-activators cyclin E and CDK6 \cite{Yamamoto2000,Lim2005}. 
+We also assumed androgen receptor could be activated through androgen binding or a ligand-independent, MAPK-driven mechanism referred to as the outlaw pathway \cite{Feldman2001,Yeh1999}. 
+We assumed a single canonical activating AR phosphorylation site (S515), with phosphorylated extracellular-signal-regulated kinase 1/2 (ppERK1/2) being the major kinase and protein phosphatase 2 (PP2A) or phosphoprotein phosphatase 1 (PP1) being the major phosphatases responsible for regulating this site.
+Finally, we modeled androgen receptor induced gene expression, including prostate specific antigen (PSA), cPAcP and p21Cip1.
+
+%This description includes  were added to the model because of their importance in co-regulating AR transcription regulation activity in prostate cells . 
+%With the addition of cell cycle proteins, we also included 
+%only considered cyclin D as a cell marker and 
+%In both models we considered nuclear hormone (AR) and transmembrane growth factor receptor (homo- and heterodimers of ErbB1/2) activation through both ligand-dependent and independent mechanisms. 
+%The model investigated the activation of AR through androgen binding or a ligand-independent, MAPK-driven mechanism. 
+%In addition to MAPK-dependent outlaw activation of AR, the model also included transcriptional activation via MAPK signaling \cite{Feldman2001,Yeh1999}. 
+%The PI3K/AKT/mTOR pathway for translation initiation was also included in the model, as well as the regulated expression of PSA, cyclin D,  
+%Thus, 
+%The model of Tasseff \textit{et al.}  
+%In order 
+%Additionally, expression of transcription factors AP-1 and ETS were also added to the model. 
+%Another significant addition to the model was the inclusion of a nuclear compartment separate from the cytosol, membrane and extracellular environments. 
+%The prostate model was developed to determine the shift in important signaling components between CR and AD prostate cancer cells.
+
+\subsection*{Formulation and solution of the model equations.}
+The prostate model was formulated as a coupled set of non-linear ordinary differential equations (ODEs):
+\begin{equation}\label{asseassertain_massbalance}
+\frac{d\mathbf{x}}{dt}=\mathbf{S}\cdot\mathbf{r}\left(\mathbf{x},\mathbf{k}\right)\qquad\mathbf{x}\left(t_{o}\right)=\mathbf{x}_{o}
+\end{equation}
+The quantity $\mathbf{x}$ denotes the vector describing the abundance of protein, mRNA, and other species in the model ($780\times 1$).
+The stoichiometric matrix $\mathbf{S}$ encodes the signaling architecture considered in the model ($780\times 1674$).  
+Each row of $\mathbf{S}$ describes a signaling component while each column describes a particular interaction.
+The $(i,j)$ element of $\mathbf{S}$, denoted by $\sigma_{ij}$, describes how species $i$ is involved with interaction $j$. 
+If $\sigma_{ij}>0$, species $i$ is produced by interaction $j$. Conversely, If $\sigma_{ij}<0$, then species $i$ is consumed in interaction $j$. 
+Lastly, if $\sigma_{ij}=0$, then species $i$ is not involved in interaction $j$.
+The term  $\mathbf{r}\left(\mathbf{x},\mathbf{k}\right)$ denotes the vector of interactions rates ($1674\times{1}$).
+Gene expression and translation processes as well as all biochemical transformations were decomposed into simple elementary steps, 
+where all reversible interactions were split into two irreversible steps (supplemental materials). 
+We modeled each network interaction using elementary rate laws where all reversible interactions were split into two irreversible steps. 
+Thus, the rate expression for interaction $q$ was given by:
+\begin{equation}\label{eq-mass-action}
+r_{q}\left(\mathbf{x},k_{q}\right)=k_{q}\prod_{j\in\left\{\mathbf{R}_{q}\right\}}x_{j}^{-\sigma_{jq}}
+\end{equation}
+The set $\left\{\mathbf{R}_{q}\right\}$ denotes reactants for reaction $q$, while $\sigma_{jq}$ denotes the stoichiometric coefficient (element of the matrix $\mathbf{S}$) governing species $j$ in reaction $q$.
+The quantity $k_{q}$ denotes the rate constant (unknown) governing reaction $q$.
+Model equations were generated in the C-programming language using the UNIVERSAL code generator, starting from an text-based input file (available in supplemental materials). 
+UNIVERSAL, an open source Objective-C/Java code generator, is freely available as a Google Code project (http://code.google.com/p/universal-code-generator/). 
+Model equations were solved using the CVODE solver in the SUNDIALS library \cite{Hindmarsh2005} on an Apple workstation (Apple, Cupertino, CA; OS X v10.6.8).
+
+We ran the model to steady-state before calculating the response to DHT or growth factor inputs.
+The steady-state was estimated numerically by repeatedly solving the model equations and estimating the difference between subsequent time points:
+\begin{equation}
+\Vert\mathbf{x}\left(t+\Delta{t}\right) - \mathbf{x}\left(t\right)\Vert_{2}\leq\gamma
+\end{equation}
+The quantities $\mathbf{x}\left(t\right)$ and $\mathbf{x}\left(t+\Delta{t}\right)$ denote the simulated abundance vector at time $t$ and $t + \Delta{t}$, respectively.
+The $L_{2}$ vector-norm was used as the distance metric, where $\Delta{t} = 100$ hr of simulated time and $\gamma$ = 0.001 for all simulations.
+
+We estimated an ensemble of model parameter sets using the Pareto Optimal Ensemble Techniques (POETs) multiobjective optimization routine \cite{Song:2009dq,Song2010,Lequieu:2011eu}. 
+POETs minimized the residual between model simulations and 43 separate training objectives taken from protein and mRNA signaling data generated in androgen dependent, intermediate and independent LNCaP cell lines (Table \ref{objective_table}).
+From these training objectives, POETs generated $>10^6$ candidate parameter vectors from which we selected N = 5000 Pareto rank-zero vectors for further analysis. 
+The set-to-set correlation between selected sets was approximately 0.60, suggesting only modest similarity between ensemble members.
+Approximately 33\%, or 560 of the 1674 parameters had a coefficient of variation (CV) of less than 1.0, where the CV ranged from 0.59 to 5.8 over the ensemble.
+Details of the parameter estimation problem and POETs are given in the supplemental materials.
+
+\subsection*{Sensitivity and robustness analysis.}
+Steady-state sensitivity coefficients were calculated for N = 500 parameter sets selected from the ensemble by solving the augmented kinetic-sensitivity equations \cite{Dickinson1976}:
+\begin{equation}
+	\begin{bmatrix}
+		\mathbf{S}\cdot\mathbf{r}(\mathbf{x},\mathbf{k}) \\
+		\mathbf{A}(t_s)\mathbf{s}_j + \mathbf{b}_j(t_s)
+	\end{bmatrix} = \begin{pmatrix}
+		\mathbf{0} \\
+		\mathbf{0}
+	\end{pmatrix}
+	   \qquad j = 1,2,\dots,\mathcal{P}
+\end{equation}
+where
+\begin{equation}
+s_{ij}(t_s)= \left. \frac{\partial{x_i}}{\partial{k_j}} \right|_{t_s}
+\end{equation} for each parameter set.
+Steady-state was calculated as described previously.
+The quantity $j$ denotes the parameter index, $\mathbf{A}$ denotes the Jacobian matrix, and $\mathcal{P}$ denotes the number of parameters in the model. 
+The vector $\mathbf{b}_j$ denotes the $j$th column of the matrix of first-derivatives of the mass balances with respect to the parameters. 
+Steady-state sensitivity coefficients were used because of the computational burden associated with sampling several hundred parameters sets for each of the 1674 parameters. 
+The steady-state sensitivity coefficients $\mathcal{N}_{ij} \equiv s_{ij}$ were organized into an array for each parameter set in the ensemble: 
+\begin{equation}
+	\mathcal{N}^{\left(\epsilon\right)} = 
+	\begin{pmatrix}
+		\mathcal{N}_{11}^{\left(\epsilon\right)} & \mathcal{N}_{12}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{1j}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{1P}^{\left(\epsilon\right)} \\
+		\mathcal{N}_{21}^{\left(\epsilon\right)} & \mathcal{N}_{22}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{2j}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{2P}^{\left(\epsilon\right)} \\
+		\vdots & \vdots & & \vdots & & \vdots \\
+		\mathcal{N}_{M1}^{\left(\epsilon\right)} & \mathcal{N}_{M2}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{Mj}^{\left(\epsilon\right)} & \hdots & \mathcal{N}_{MP}^{\left(\epsilon\right)} \\
+	\end{pmatrix}\qquad\epsilon = 1,2,\hdots,N_{\epsilon}
+\end{equation}
+where $\epsilon$ denotes the index of the ensemble member, $P$ denotes the number of parameters, $N_{\epsilon}$ denotes the number of parameter sets sampled (N = 500) and $M$ denotes the number of model species.
+To estimate the relative fragility or robustness of species and reactions in the network, we decomposed $\mathcal{N}^{\left(\epsilon\right)}$ 
+using Singular Value Decomposition (SVD):
+
+\begin{equation}
+	\mathcal{N}^{\left(\epsilon\right)} = \mathbf{U}^{\left(\epsilon\right)}\Sigma^{\left(\epsilon\right)}\mathbf{V}^{T,{\left(\epsilon\right)}}
+\end{equation}
+Coefficients of the left singular vectors corresponding to largest $\theta$ $\leq$ 15 singular values of $\mathcal{N}^{\left(\epsilon\right)}$ were rank-ordered to estimate important species combinations, while coefficients of the right singular vectors were used to rank important reaction combinations.
+Only coefficients with magnitude greater than a threshold ($\delta$ = 0.001) were considered. 
+The fraction of the $\theta$ vectors in which a reaction or species index occurred was used to quantify its importance (sensitivity ranking). 
+We compared the sensitivity ranking between different conditions to understand how control in the network shifted in different cellular environments.
+
+Robustness coefficients were calculated as shown previously \cite{Tasseff2011}. 
+Robustness coefficients denoted by $\alpha\left(i,j,t_{o},t_{f}\right)$ are defined as:
+\begin{equation}
+\alpha\left(i,j,t_{o},t_{f}\right)=\left(\displaystyle\int_{t_{o}}^{t_{f}}x_{i}\left(t\right)dt\right)^{-1}\left(\displaystyle\int_{t_{o}}^{t_{f}}x^{(j)}_{i}\left(t\right)dt\right)
+\end{equation}
+Robustness coefficients quantify the response of a marker to a structural or operational perturbation to the network architecture.
+Here $t_o$ and $t_f$ denote the initial and final simulation time respectively, while $i$ and $j$ denote the indices for the marker and the perturbation respectively. 
+A value of $\alpha\left(i,j,t_{o},t_{f}\right)>1$, indicates increased marker abundance, while $\alpha\left(i,j,t_{o},t_{f}\right)<1$ indicates decreased marker abundance following perturbation $j$. 
+If $\alpha\left(i,j,t_{o},t_{f}\right)\sim{1}$ the $j$th perturbation does not influence the abundance of marker $i$.
+Robustness coefficients were calculated (starting from steady-state) from $t_{o} = 0$ hr to $t_{f} = 72$ hr following the addition of 10nM DHT at $t_{o}$.
+Robustness coefficients were calculated for the same N = 500 models selected for sensitivity analysis.
+
+\subsection*{Experimental Validation.}
+
+\subsection*{Cell culture and treatments}
+Androgen dependent LNCaP prostate cancer cells were a gift from Dr.
+Brian Kirby (Cornell University), and the castration resistant C4-2 prostate cancer cell line was purchased from MD Anderson Cancer Center, University of Texas.
+Cell lines were maintained in RPMI 1640 media (Life Technologies, Inc., Grand Island, NY) with 10\% fetal calf serum (FBS; Hyclone) and 1x antibiotic/antimycotic (Sigma, St.
+Louis, MO) in a 5\% CO$_2$ humidified atmosphere at 37$^{\circ}$C.
+The AR inhibitor MDV3100 (enzalutamide) and the Raf inhibitor sorafenib were purchased from SantaCruz Biotechnology (Santa Cruz, CA).
+The PI3K inhibitor LY294002 was purchased from Cell Signaling Technologies (Danvers, MA, USA).
+All stock solutions were diluted in DMSO (Sigma, St.
+Louis, MO).
+
+\subsection*{Protein Extraction and Western Blot}
+
+LNCaP and C4-2 cells were seeded in 60 mm dishes at a density of 4 x 10$^{5}$.
+After 96 and 72 hrs, for LNCaP and C4-2 cells respectively, the media was replaced with fresh media and drug treatments were added.
+After 24 hours, cells were washed twice in PBS buffer, scraped in 250 $\mu$L ice-cold lysis buffer (Pierce, Rockford, IL) supplemented with protease and phosphatase inhibitors (Sigma, St.
+Louis, MO), and lysed for 30 min on ice.
+Lysates were centrifuged at 13,000 rpm for 30 min at 4$^{\circ}$C.
+After quantification of total protein by BCA assay, equal amounts of total protein lysates (25 $\mu$g) were resolved by SDS-PAGE and transferred onto PVDF membranes.
+Membranes were blocked in 5\% fat free milk and then probed with antibodies.
+The primary antibodies used for western blot analysis were pAKT Ser473, AKT, pS6 Ser240/244 , pERK Thr202/Tyr204, ERK, AR, cleaved PARP, and GAPDH were from Cell Signaling Technologies (Danvers, MA, USA).
+For detection, enhanced chemiluminescence ECL reagent (GE Healthcare, Pittsburgh, PA) was used and signals were visualized using the ChemiDoc XRS system (Bio-Rad).
+
+\subsection*{MTT Assay}
+
+LNCaP and C4-2 cells were seeded at a density of 1x10$^{4}$ cells per well in 96 well plates.
+After 48 hrs the media was refreshed and drug treatments added.
+Cell growth at 24, 48, and 72 hrs was determined using a 3-(4,5-dimethyl thiazol-2-yl)-2,5-diphenyl tetrazolium bromide (MTT) assay.
+At the specified time point 10 $\mu$L MTT reagent (stock of 5 mg/mL in PBS) was added to each well and the cells were further incubated for 4 hrs.
+At 4 hrs, the media was removed and 50 $\mu$L of dissolving reagent DMSO was added to each well.
+After an additional 10 min incubation, the absorbance was measured at 540 nm on a microplate reader.
+Each reading was adjusted by subtracting the absorbance value for the blank (media only) and the results were then scaled to the DMSO-treated (control) case.
+
+\section*{Acknowledgements}
+This study was supported by award number U54CA143876 from the National Cancer Institute.
+The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Cancer Institute or the National Institutes of Health.
+The authors acknowledge the financial support of the National Science Foundation, award no.
+DGE-1045513,for a GK12 training grant entitled "Grass Roots: Advancing Education in Renewable Energy and Cleaner Fuels through collaborative graduate fellow/teacher/grade-school student interactions."
+We would like to thank Timothy Chu, Kyriakos Tsahalis, Spencer Davis, Rohit Yallamraju, Gongcheng Lu, and Yalin Zhao.
+
+\clearpage
+%\bibliographystyle{plain}
+%\bibliographystyle{IEEEbib}
+
+\bibliography{References_v1}
+
+%\begin{comment}
+
+\clearpage
+
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Fig_1_LNCAPModel.pdf}
+\caption{Schematic overview of the prostate signaling network.
+The model describes hormone and growth factor induced expression of several proteins, including PSA.
+In the absence of outside hormones/growth factors, overactive HER2 can stimulate the MAPK and AKT pathways.
+AR can be activated directly by the MAPK pathway.
+}
+\label{fg:NetworkSchematic}
 \end{figure}
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=1.00\textwidth,height=0.7\textheight]{./figs/Figure_2_CoagulationNetwork.pdf}
-\caption{Schematic of the extrinsic and intrinsic coagulation cascade.
-Inactive zymogens upstream (grey) are activated by exposure to tissue factor (TF)  following vessel injury.
-Tissue factor and activated factor VIIa (FVIIa) form a complex that activates factor X (fX) and IX (fIX).
-FXa activates downstream factors including factor VIII (fVIII) and fIX.
-Factor V (fV) is primarily activated by thrombin (FIIa).
-In addition, we included a secondary fV activation route involving FXa.
-FXa and FVa form a complex (prothrombinase) on activated platelets that converts prothrombin (fII) to FIIa.
-FIXa and FVIIIa can also form a complex (tenase) on activated platelets which catalyzes FXa formation.
-Thrombin also activates upstream coagulation factors, forming a strong positive feedback ensuring rapid activation.
-Tissue factor pathway inhibitor (TFPI) downregulates FXa formation and activity by sequestering free FXa and TF-FVIIa in a FXa-dependent manner.
-Antithrombin III (ATIII)  inhibits all proteases.
-Thrombin inhibits itself binding the surface protein thrombomodulin (TM).
-The IIa-TM complex catalyzes the conversion of protein C (PC) to activated protein C (APC), which attenuates the coagulation response by the proteolytic cleavage of fV/FVa and fVIII/FVIIIa.
-}\label{fig-coagulation-network}
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Fig_3_ColorPlot.pdf}
+\caption{Simulation results versus experimental results for training and validation data.
+Experiment numbers 1 through 43 were used for training, while experiments 44 through 72 were validation.
+Gray means the ensemble member qualitatively fit experimental data in both models.
+White means the the ensemble member only fit the data using the new model that included HER2 heterodimerization.
+Red means the ensemble member fit using only the old model.
+Black corresponds to an incorrect cellular response in both models.
+$\bf{A.,C.}$ Training and validation results, respectively, for entire ensemble population using both the original model and an updated model including HER2 heterodimerization (N = 5000).
+$\bf{B.,D.}$ Simulation results for training and validation of a random set of 100 members using both models.
+ }
+\label{fg:ColorPlot}
 \end{figure}
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=1.0\textwidth,height=0.5\textheight]{./figs/Figure_3_Errors_convergence.pdf}
-\caption{Error convergence rates of the 5 different algorithms on the coagulation model.
-The objective error is the mean over $N$= 25 trials.
-DDSMLSPSO and DDS have the steepest drop in error during first 2000 function evaluations.
-Thereafter the error drop in DDS remains nearly constant whereas DDSMLSPSO drops further from around 2000 to 4000 function evaluations.
-At the end of 4000 function evaluations DDSMLSPSO attains the lowest error.
-}\label{fig-train}
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Fig_2_Training.pdf}
+\caption{Ensemble performance against selected training objectives (N = 5000).
+A, B.
+Time course data for PSA concentration due to a stimulus of 10 nM DHT in LNCaP C33 cells and LNCaP C81 cells, respectively (O2, O3).
+C.
+PSA levels in the presence and absence of a HER2 inhibitor (LNCaP C81 cells, O7).
+D.
+PAcP mRNA levels at 72 hours in the presence and absence of DHT (LNCaP C51 cells, O14).
+E.
+Steady-state cyclin D levels in LNCaP C33 vs.
+C81 (O17).
+F.
+p21Cip1 levels at 48 hrs in the presence and absence of DHT (LNCaP C33, O25).
+G.
+Steady-state AKT phosphorylation levels in LNCaP C33 vs.
+C51 (O30).
+H.
+AR levels at 24 hours in the presence and absence of DHT (LNCaP C51, O31).
+I.
+AR mRNA levels in the presence and absence of E2F over expression (LNCaP C33, O34).
+Error bars denote plus and minus one standard deviation above the mean.}
+\label{fg:Training}
 \end{figure}
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=1.0\textwidth,height=0.5\textheight]{./figs/Figure_4_Sim_Train_E1_E5.pdf}
-\caption{Model fits on experimental data using DDSMLSPSO.
-The model parameters were estimated using DDSMLSPSO.
-Solid black lines indicate the simulated mean thrombin concentration using parameter vectors from $N$= 25 trials.
-The grey shaded region represents the 99\% confidence estimate of the mean simulated thrombin concentration.
-The experimental data is reproduced from the synthetic plasma assays of Mann and co-workers.
-Thrombin generation is initiated by adding Factor VIIa-TF (5nM - Red and 5pM - Green) to synthetic plasma containing 200 $\mu$mol/L of phospholipid vesicles (PCPS) and a mixture of coagulation factors (II,V,VII,VIII,IX,X and XI) at their mean plasma concentrations.
-}\label{fig-train}
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Fig_4_Validation.pdf}
+\caption{Blind model predictions for the ensemble (N = 5000).
+The model ensemble’s predictive ability was assessed by comparing simulation versus experimental data not used for training.
+A.
+Time course data for HER2 phosphorylation due to a stimulus of 10 nM DHT (LNCaP C33, P1).
+B.
+ERK phosphorylation levels in the presence and absence of a PAcP inhibitor (LNCaP C33 cells, P3).
+C.
+AR levels at 24 hrs in varying levels of DHT (LNCaP C33, P17).
+D.
+Shc phosphorylation levels at 24 hrs in the presence and absence of DHT (LNCaP C33, P22).
+E.
+Time course data for HER2 phosphorylation due to a stimulus of 1.6 nM EGF (LNCaP C33, P7).
+F.
+AR levels in varying levels of EGF (LNCaP C33, P14).
+G, H, I.
+Fold change in PSA concentration due to drug stimulus: enzalutamide, lapatinib, and sorafenib.
+Error bars denote plus and minus one standard deviation above the mean.}
+\label{fg:Validation}
 \end{figure}
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=1.0\textwidth,height=0.5\textheight]{./figs/Figure_5_Sim_Validate_E2_E4_E6.pdf}
-\caption{Model predictions on unseen experimental data using parameters obtained from DDSMLSPSO.
-The parameter estimates that were obtained using DDSMLSPSO were tested against data that was not used in the model training.
-Solid black lines indicate the simulated mean thrombin concentration using parameter vectors from $N$= 25 trials.
-The grey shaded region represents the 99\% confidence estimate of the mean simulated thrombin concentration.
-The experimental data is reproduced from the synthetic plasma assays of Mann and co-workers.
-Thrombin generation is initiated by adding Factor VIIa-TF (500pM - Blue, 50pM - Pink and 10pM - Orange respectively) to synthetic plasma containing 200 $\mu$mol/L of phospholipid vesicles (PCPS) and a mixture of coagulation factors (II,V,VII,VIII,IX,X and XI) at their mean plasma concentrations.
-}\label{fig-validation}
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Fig_5_Sensitivity.pdf}
+\caption{Sensitivity analysis of a population of prostate models (N = 500).
+Species with a low sensitivity are considered robust, while species with a high sensitivity ranking are considered fragile.
+$\bf{A,B.}$ Sensitivity ranking of network species in AD versus CR cells in the absence (presence) of DHT.
+$\bf{C.}$ Sensitivity ranking of network species in AD cells in the absence and presence of DHT.
+$\bf{D.}$ Sensitivity ranking of network species in CR cells in the presence and absence of enzalutamide with a DHT stimulus.
+Error bars denote standard error with N = 500.}
+\label{fg:Sensitivity}
 \end{figure}
 
-%\bibitem[Saltelli \em{et~al.}(2010)Saltelli, Annoni, Azzini, Campolongo, Ratto,
-  %and Tarantola]{Saltelli:2010}
-%Saltelli, A.; Annoni, P.; Azzini, I.; Campolongo, F.; Ratto, M.; Tarantola, S.
-%\newblock Variance based sensitivity analysis of model output. Design and
-  %estimator for the total sensitivity index.
-%\newblock {\em Comput. Phys. Commun.} {\bf 2010}, {\em 181},~259--270.
+\clearpage
 
-%\end{thebibliography}
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure6_RAF_K0_C81.pdf}
+\caption{Robustness analysis of a population of CR prostate models with Raf knock-out (N = 500).
+A log fold change of greater then zero implies that the concentration of the protein increased with the knock-out of Raf, while a log fold change of less then zero indicates that the concentration of protein decreased.
+A log of fold change equal to 0, shows no response due to Raf knock-out.
+Three distinct regions emerge in Raf knock-out case: (1) PSA increases, (2) cyclin D concentration increases, and (3) PSA and cyclin D concentration decrease.
+}
+\label{fg:Robustness}
+\end{figure}
+
+\clearpage
+
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure7_ExperimentalResults.pdf}
+\caption{Experimental results for multiple drug combinations on two prostate cancer cell lines, LNCaP and C4-2.
+A.
+Western blot analysis of AR, pS6, AKT, pAKT, ERK and cleaved PARP in LNCaP and C4-2 cell lines treated for 24 hrs with DMSO (control), sorafenib (5 $\mu$M), LY294002 (40 $\mu$M), and MDV3100 (10 $\mu$M) alone or in combination (at least 3 repeats).
+B.
+Cells (LNCaP and C4-2) were treated for 24, 48 and 72 hrs with sorafenib (5 $\mu$M), LY294002 (40 $\mu$M), and MDV3100 (10 $\mu$M) and cell viability was measured using MTT Assay.
+Values were normalized to DMSO (control).
+C.
+Cell viability results for LNCaP and C4-2 cells at varying concentration of sorafenib, LY294002, and MDV3100 after 24 hrs of treatment.
+Values were normalized to DMSO (control).
+Error bars represent standard error (at least 3 repeats with triplicates performed in each experiment).}
+\label{fg:Experiment}
+\end{figure}
+
+\clearpage
+
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Fig_8_BoxPlot_KnockOut_Cases_new.pdf}
+\caption{Robustness analysis of a population of CR prostate models with seven protein knock-out cases (N = 500).
+A log fold change of greater then zero implies that the concentration of the protein increased with the knock-out, while a log fold change of less then zero indicates that the concentration of protein decreased.
+A log of fold change equal to 0, shows no response due to the knock-out.
+ A.,B.,C.,D. Log robustness of PSA, p70p, Cyclin D, and AKTp versus protein knock-out. A CR LNCaP cell was assumed for all knock-out cases.}
+\label{fg:Dual_Knockout}
+\end{figure}
 
@@
 
-\begin{figure}[ht]
-\centering
-\includegraphics[width=1.02\textwidth,height=0.5\textheight]{./figs/Figure_6_Ackley_Rast}
-\caption{Error convergence on 300 dimensional Ackley function and Rastrigin function.
-\textbf {(A)} Error convergence of 5 different meta-heuristics on 300 dimensional Ackley function \textbf {(B)} Error convergence of 5 different meta-heuristics on 300 dimensional Rastrigin function
-}\label{fig-testfunctions}
+\section*{Supplementary materials}
+
+\subsection*{Estimation of a population of models using Pareto Optimal Ensemble Techniques (POETs).}
+
+We used multiobjective optimization to estimate an ensemble of prostate models.
+Although computationally more complex than single-objective formulations, multiobjective optimization can be used to address qualitative conflicts in training data arising from experimental error or cell-line artifacts \cite{Handl2007}.
+In this study we used the Pareto Optimal Ensemble Technique (POETs) to perform the optimization.
+POETs integrates standard search strategies, e.g., Simulated Annealing (SA) or Local Pattern Search (PS) with a Pareto-rank fitness assignment \cite{Song2010}.
+The mean squared error, $\eta$, of parameter set $\mathbf{k}$ for training objective $j$ was defined as:
+
+\begin{equation}
+\eta_j(\mathbf{p}_k) = \frac{1}{N}\sum_{i}^{N}\frac{(\hat{x}_{i,j} - \beta_j x(\mathbf{p_k})_{i,j})^2}{\hat{\sigma}^2_{i,j}}
+\end{equation}
+The symbol $\hat{x}_{i,j}$ denotes scaled experimental observations (from training objective j) while $x(\mathbf{p_k})_{i,j}$ denotes the simulation output (from training objective j).
+The quantity $i$ denotes the sampled time-index or condition, and $N$ denotes the number of time points or conditions for experiment j.
+The standard deviation, $\hat{\sigma}_{i,j}$, was assumed to be equal to 10\% of the reported observation, if no experimental error was reported.
+$\beta_j$ is a scaling factor which is required when considering experimental data that is accurate only to a multiplicative constant.
+In this study, the experimental data used for training and validation was typically band intensity from immunoblots, where intensity was estimated using the ImageJ software package \cite{Abramoff2004}.
+The scaling factor used was chosen to minimize the normalized squared error \cite{Brown2003}:
+
+\begin{equation}
+\beta_j = \frac{\sum_{i}(\hat{x}_{i,j}x_{i,j}/\hat{\sigma}_{i,j}^2)}{\sum_{i}(x_{i,j}/\hat{\sigma}_{i,j})^2}
+\end{equation} 
+By using the scaling factor, the concentration units on simulation results were arbitrary, which was consistent with the arbitrary units on the experimental training data.
+All simulation data was scaled by the corresponding $\beta_j$.
+
+We computed the Pareto rank of parameter set $\mathbf{k}_{i+1}$ by comparing the simulation error at iteration $i+1$ against the simulation archive, denoted as $\mathbf{K}_i$.
+We used the Fonseca and Fleming ranking scheme \cite{Fonseca1993} to estimate the rank of the parameter set $\mathbf{k}_{i+1}$.
+Parameter sets with increasing rank are progressively further away from the optimal trade-off surface.
+The parameter set $\mathbf{k}_{i+1}$ was accepted or rejected by the SA with probability $\mathcal{P}\left(\mathbf{k}_{i+1}\right)$:
+\begin{equation}\label{eqn_costMOSA}
+\mathcal{P}(\mathbf{k}_{i+1}) \equiv \exp{\{-rank\left(\mathbf{k}_{i+1} \mid \mathbf{K}_{i} \right)/T\}}
+\end{equation}
+where $T$ is the computational annealing temperature.
+The Pareto rank for $\mathbf{k}_{i+1}$ is denoted by $rank\left(\mathbf{k}_{i+1}\mid \mathbf{K}_{i}\right)$.
+The annealing temperature was  adjusted according to the schedule
+$T_k = \beta^k T_0$
+where $\beta$ was defined as $\beta = \left(\frac{T_{f}}{T_{o}}\right)^{1/10}$.
+The initial temperature was given by $T_0 = n/log(2)$, with $n=4$ and the final temperature $T_f = 0.1$ used in this study.
+The epoch-counter $k$ was incremented after the addition of 50 members to the ensemble.
+As the ensemble grew, the likelihood of accepting a high rank set decreased.
+Parameter sets were generated by applying a random perturbation in log space:
+
+\begin{equation}
+\log\mathbf{k}_{i+1} = \log\mathbf{k}_i + \mathcal{N}(0,\nu)
+\end{equation}
+where $\mathcal{N}(0,\nu)$ is a normally distributed random number with zero mean and variance $\nu$, set as 0.1 in this model.
+The perturbation was applied in log space to account for large variation in parameter scales and to ensure positive parameter values.
+We used a local pattern search every $q$ steps, in our case 20, to minimize error for a single randomly selected objective.
+The local pattern-search algorithm used has been described previously \cite{Gadkar2003}.
+
+\input{Prostate_Obj_Table}
+\clearpage
+
+\input{Prostate_Pred_Table}
+\clearpage
+
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Supplementary_CV_Subset_Total}
+\caption{Coefficient of variation (CV) of model parameters estimated using POETs.
+The solid line denotes the mean CV calculated over the entire ensemble (N = 5000).
+The points denote the mean CV of the 500 ensemble members used for sensitivity and robustness calculations.
+Over the ensemble, the coefficient of variation (CV) of the kinetic parameters spanned 0.59 - 5.8, with 33\% of the parameters having a CV of less than one.}
+\label{fg:Supp_CV}
 \end{figure}
 
-\begin{figure}[ht]
-\centering
-\includegraphics[width=1.02\textwidth,height=0.5\textheight]{./figs/Figure_7_Benchmarks_Parameters}
-\caption{Difference between optimal and nominal parameter vector values on benchmark problems.
-\textbf {(A)} Problem B1: Genome wide kinetic model of E.coli \textit{S.cerevisiae} with 1759 unknown parameters.
-\textbf {(B)} Problem B4: Metabolic model of Chinese  Hamster Ovary Cells (CHO) cells with 117 parameters.
-}\label{fig-benchmark}
+\clearpage
+
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Supp_Figure_NewModelvsOldModel}
+\caption{Blind model predictions for the ensemble with the original and updated model (EGFR and HER2 heterodimer).
+A,B.
+Time course data for HER2 phosphorylation due to a stimulus of 1.6 nM EGF (LNCaP C33, P7) for the old and new model, respectively.
+Dark grey shows only parameters improved by the updated model (N=2870) while light grey show all parameter sets (N=5000).
+C,D.
+Shc phosphorylation levels at 16 hrs in the presence and absence of 1.6 nM EGF (LNCaP C33, P29) for the old and new model, respectively.
+Light grey denotes experimental data, mid grey denotes simulation results for all parameters (N=5000), and black denotes only parameters improved by the updated model (N=3142).
+Error bars denote plus and minus one standard deviation above the mean.}
+\label{fg:Exps_NewModelvsOldModel}
 \end{figure}
 
+\clearpage
+
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Supp_Figure_Sensitivity.pdf}
+\caption{Sensitivity analysis of a population of prostate models (N = 500).
+Species with a low sensitivity are considered robust, while species with a high sensitivity ranking are considered fragile.
+$\bf{A}$ Sensitivity ranking of network species in CR cells in the absence and presence of DHT.
+$\bf{B.}$ Sensitivity ranking of network species in CR cells in the presence of enzalutamide in the presence and absence of a DHT stimulus.
+$\bf{C.,D.}$ Sensitivity ranking of network species in CR cells in the presence and absence of sorafenib and lapatinib, respectively, with a DHT stimulus.
+Error bars denote standard error with N = 500.}
+\label{fg:Supp_Sensitivity}
+\end{figure}
+
+\clearpage
+
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Supp_Figure_Robustness.pdf}
+\caption{Robustness analysis of protein markers.
+Expression level of key proteins was altered by a factor of 2, 0.1, or 0 (knock-in, knock-down, or knock-out) and robustness coefficients were calculated for five key protein markers.
+Simulations shown were from CR cells, with indicated perturbation.
+Mean of 500 ensemble members is shown.}
+\label{fg:Supp_Robustness}
+\end{figure}
+
+\clearpage
+
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Supp_Figure_KO_new.pdf}
+\caption{Dual knock-out of AR and PI3K leads to decreased expression of activated p70.
+A., B, C.
+Robustness coefficient of activated p70 (S6) in the PI3K knock-out, AR knock-out, and dual knock-out cases, respectively.
+The control was the basil CR LNCaP wild type case.
+Error bars denote plus and minus one standard error above the mean with N = 500.
+Experimental data is from Carver, et al \cite{Carver2011}.
+}
+\label{fg:Supp_KO}
+\end{figure}
+
+\clearpage
+
+%\end{comment}
+
 \end{document}

### 2015-09-29 Paper_v2.tex
Methods_Update

@@
 Robustness coefficients were calculated (starting from steady-state) from $t_{o} = 0$ hr to $t_{f} = 72$ hr following the addition of 10nM DHT at $t_{o}$.
+For scaled log fold change we used the following equation:
+
+\begin{equation}
+\alpha_{scaled}\left(i,j\right)=
+\begin{cases}
+	\frac{\log_{10}(\alpha(i,j))}{\max \log_{10}(\alpha(i))},             & \text{if } \log_{10}(\alpha(i,j))\geq 0\\
+    	-\frac{\log_{10}(\alpha(i,j))}{\min \log_{10}(\alpha(i))},              & \text{if } \log_{10}(\alpha(i,j))< 0
+\end{cases}	
+\
+\end{equation}
+A value of $\alpha_{scaled}\left(i,j\right)>0$, indicates increased marker abundance, while $\alpha_{scaled}\left(i,j\right)<0$ indicates decreased marker abundance following perturbation $j$. 
+If $\alpha_{scaled}\left(i,j\right)\sim{0}$ the $j$th perturbation does not influence the abundance of marker $i$.
+A value of $\alpha_{scaled}\left(i,j\right)=1$, indicates max increase of marker abundance, while $\alpha_{scaled}\left(i,j\right)=-1$ indicates the max decrease of marker abundance. 
 Robustness coefficients were calculated for the same N = 500 models selected for sensitivity analysis.
@@
 The PI3K inhibitor LY294002 was purchased from Cell Signaling Technologies (Danvers, MA, USA).
-All stock solutions were diluted in DMSO (Sigma, St.
+All stock solutions were diluted in DMSO and stored at -20$^{\circ}$C (Sigma, St.
 Louis, MO).
+Stock solution concentrations for western blotting experiments were 10 mM Sorafenib, 50 mM LY294002, and 10 mM MDV3100.
+For the cell viability assays, stock solution concentrations were 0.5 mM Sorafenib, 4 mM LY294002, and 1 mM MDV3100.
 

### 2015-10-16 Paper_v2.tex
Robustness Updates

@@
 
-\subsection*{Experimental results confirm the need for dual therapies in prostate cancer.}
-
-The results from our sensitivity analysis indicate that instead of inhibiting solely the AR pathway (enzalutamide), a combination therapy is necessary. 
-To test this hypothesis we used the well characterized ADPC cell line LNCaP as well a LNCaP derived CRPC cell line C4-2 \cite{Thalmann1994}. 
-Three inhibitors were used: the AR inhibitor MDV3100 (enzalutamide), the Raf kinase inhibitor sorafenib, and the PI3K inhibitor LY294002. 
-In both cell lines, inhibition of either the AR or MAPK pathway appears to promote activation of the PI3K pathway, as seen by the increase in pAKT (S473) (Fig. \ref{fg:Experiment}A). 
-The addition of the PI3K inhibitor, LY294002, alone or in combination diminishes PI3K activity. 
-The inhibition of PI3K alone, led to an increase in AR expression in both LNCaP and C4-2 cell lines. 
-Since, AR transcriptionally upregulates its own expression \cite{Lin2003}\cite{Grad1999}, this may indicate an increase in AR activity. 
-The ribosomal protein pS6 was completely inhibited only in the presence of the PI3K inhibitor LY294002. 
-Cell viability results show a large decrease in cell viability at 72 hrs in the dual inhibition cases as well as the triple inhibition case for both LNCaP and C4-2 cell lines (Fig.
-\ref{fg:Experiment}B).
-In both cell lines, MDV3100 (10 $\mu$M), has only a modest effect on cell viability versus control (DMSO). 
-Figure \ref{fg:Experiment}C shows cell viability of both cell lines in varying concentrations of inhibitors at 24 hrs.
-
 \subsection*{Robustness analysis identified key regulators of prostate cancer.} 
@@
 
+We also used robustness analysis to observe the effect of knocking out combinations of key species found in our sensitivity analysis. 
+We calculated the effect of perturbation in CR cells on different protein markers in the following 7 cases: (1) Raf knock-out, (2) PI3K knock-out, (3) AR knock-out, (4) Raf and PI3K knock-outs, (5) Raf and AR knock-outs, (6) PI3K and AR knock-outs, and (7) Raf, PI3K and AR knock-outs.
+Over the 500 ensemble member set, the overall decrease in PSA expression levels was greatest in all cases which included the knock-out of AR (Fig.
+\ref{fg:Dual_Knockout}A).
+The decreased expression of activated p70  was most prominent in the double PI3K/AR knock-out case and the triple Raf/PI3K/AR knock-out case (Fig. \ref{fg:Dual_Knockout}B). 
+Although the change in expression level of cyclin D varied for individual ensemble members (increased and decreased), the median and mean behavior for all knock-out cases is around zero suggesting no change due to perturbation (Fig.\ref{fg:Dual_Knockout}C).
+Individual ensemble members do show a decline in cyclin D expression in cases involving PI3K in double and triple knock-outs and in these cases we also see fewer ensemble members that have an increased expression of cyclin D.
+Fig. \ref{fg:Dual_Knockout}D shows that in the PI3K knock-out case we see a complete inhibition of AKTp expression (a value of -1 is essentially zero expression due to knock-out). 
+This is a trivial result in the model because AKTp only occurs through PIP3 recruitment to the membrane, and PIP3 is only activated by PI3K in the model. 
+All other knock-out cases show a median response of no fold change in AKTp expression, but we do see a group of ensemble members that have an increased expression of AKTp in the RAF knock-out and RAF/AR knock-out cases.
+Overall, these results are consistent with our hypothesis that a combination therapy will be more effective.
+
+\subsection*{Experimental results confirm the need for dual therapies in prostate cancer.}
+
+The results from our sensitivity analysis indicate that instead of inhibiting solely the AR pathway (enzalutamide), a combination therapy is necessary. 
+To test this hypothesis we used the well characterized ADPC cell line LNCaP as well a LNCaP derived CRPC cell line C4-2 \cite{Thalmann1994}. 
+Three inhibitors were used: the AR inhibitor MDV3100 (enzalutamide), the Raf kinase inhibitor sorafenib, and the PI3K inhibitor LY294002. 
+In both cell lines, inhibition of either the AR or MAPK pathway appears to promote activation of the PI3K pathway, as seen by the increase in pAKT (S473) (Fig. \ref{fg:Experiment}A). 
+The addition of the PI3K inhibitor, LY294002, alone or in combination diminishes PI3K activity. 
+The inhibition of PI3K alone, led to an increase in AR expression in both LNCaP and C4-2 cell lines. 
+Since, AR transcriptionally upregulates its own expression \cite{Lin2003}\cite{Grad1999}, this may indicate an increase in AR activity. 
+The ribosomal protein pS6 was completely inhibited only in the presence of the PI3K inhibitor LY294002. 
+Cell viability results show a large decrease in cell viability at 72 hrs in the dual inhibition cases as well as the triple inhibition case for both LNCaP and C4-2 cell lines (Fig.
+\ref{fg:Experiment}B).
+In both cell lines, MDV3100 (10 $\mu$M), has only a modest effect on cell viability versus control (DMSO). 
+Figure \ref{fg:Experiment}C shows cell viability of both cell lines in varying concentrations of inhibitors at 24 hrs.
+
 %The results of the model suggest that an inhibition of either the PI3K pathway or the MAPK pathway in combination with an AR inhibitor as a possible therapy for CRPC. 
@@
 \ref{fg:Dual_Knockout}).
-The triple case, with PI3K, AR, and RAF knock-outs, also only showed a slight additive effect (Fig. \ref{fg:Triple_KO}).
+The triple case, with PI3K, AR, and RAF knock-outs, also only showed a slight additive effect (Fig. \ref{fg:Dual_Knockout}).
 %The pathways in our model appear to be uncoupled and therefore little synergy is shown in the dual knock-out case. 
@@
 \begin{figure}\centering
-\includegraphics[width=1.0\textwidth]{./figs/Figure7_ExperimentalResults.pdf}
+\includegraphics[width=1.0\textwidth]{./figs/Fig_7_BoxPlot_KnockOut_Cases_new.pdf}
+\caption{Robustness analysis of a population of CR prostate models with seven protein knock-out cases (N = 500).
+A log fold change of greater then zero implies that the concentration of the protein increased with the knock-out, while a log fold change of less then zero indicates that the concentration of protein decreased.
+A log of fold change equal to 0, shows no response due to the knock-out.
+ A.,B.,C.,D. Log robustness of PSA, p70p, Cyclin D, and AKTp versus protein knock-out. A CR LNCaP cell was assumed for all knock-out cases.}
+\label{fg:Dual_Knockout}
+\end{figure}
+
+\clearpage
+
+\begin{figure}\centering
+\includegraphics[width=1.0\textwidth]{./figs/Figure8_ExperimentalResults.pdf}
 \caption{Experimental results for multiple drug combinations on two prostate cancer cell lines, LNCaP and C4-2.
@@
 
-\begin{figure}\centering
-\includegraphics[width=1.0\textwidth]{./figs/Fig_8_BoxPlot_KnockOut_Cases_new.pdf}
-\caption{Robustness analysis of a population of CR prostate models with seven protein knock-out cases (N = 500).
-A log fold change of greater then zero implies that the concentration of the protein increased with the knock-out, while a log fold change of less then zero indicates that the concentration of protein decreased.
-A log of fold change equal to 0, shows no response due to the knock-out.
- A.,B.,C.,D. Log robustness of PSA, p70p, Cyclin D, and AKTp versus protein knock-out. A CR LNCaP cell was assumed for all knock-out cases.}
-\label{fg:Dual_Knockout}
-\end{figure}
-
-\clearpage
-
 % Supplemental figures -

### 2016-01-25 Paper_v4.tex
New Version of Paper

--- 2015-07-01 Paper_v1.tex
@@
 \usepackage{textcomp}
+\usepackage{multirow}
+\usepackage{amsmath}
+\DeclareMathOperator*{\argmin}{\arg\!\min}
 
+\usepackage{algorithm2e}
 %\usepackage[space]{cite}
@@
 \begin{titlepage}
-{\par\centering\textbf{\Large {Dynamically dimensioned particle swarm optimization (DdPSO): A meta-heuristic for parameter estimation in biochemical models}}}
+{\par\centering\textbf{\Large {Dynamic Optimization with Particle Swarms (DOPS): A meta-heuristic for parameter estimation in biochemical models}}}
 \vspace{0.05in}
@@
 \section*{Abstract}
-Mathematical modeling is as a powerful tool to analyze, and ultimately design biochemical networks.
-However, estimation of biochemical model parameters can be a significant challenge.
+Mathematical modeling is a powerful tool to analyze, and ultimately design biochemical networks.
+However, estimation of biochemical model parameters is a significant challenge.
 Parameter estimation in biochemical models generally involves expensive function evaluations, making it difficult to quickly obtain optimal solutions.
 Additionally, biochemical models often have many local minima or maxima which further complicates the parameter estimation problem.
-In this study, we developed dynamically dimensioned particle swarm optimization (DdPSO), a novel meta-heurstic that combined the global search capability of multi-swarm particle swarm optimization with dynamic dimensioned search.
-DdPSO uses a multi-swarm particle swarm optimization technique to generate candidate solution vectors which are then greedily updated using dynamic dimensioned search.
-We tested the performance of DdPSO on commonly used test functions for global optimization (Ackley and Rosenbrock), published biochemical parameter estimation benchmark problems and
-lastly on a model of the human coagulation cascade. For each test we performed 25 trials with 4000 function evaluations per trial, and compared the performance of DdPSO with other commonly
-used meta-heuristic approaches such as differential evolution (DE), particle swarm optimization (PSO), simulated annealing (SA) and dynamically dimensioned search (DDS).
-For the wide range of problems that we considered, DdPSO outperformed the other meta-heuristic approaches despite a limited number of function evaluations.
+In this study, we developed Dynamic Optimization with Particle Swarms (DOPS), a novel meta-heuristic that combined features of multi-swarm particle swarm optimization with dynamically dimensioned search (DDS).
+DOPS uses a multi-swarm particle swarm optimization technique to generate candidate solution vectors of which the best one is greedily updated using dynamically dimensioned search.
+We first tested the performance of DOPS on a model of human coagulation cascade.
+We performed 25 trials with 4000 function evaluations per trial, and compared the performance of DOPS with other commonly
+used meta-heuristic approaches such as differential evolution (DE), simulated annealing (SA) and dynamically dimensioned search (DDS).
 We further tested the predictive power of the coagulation model parameters against data not used in training, and found good agreement between model simulations and experimental measurements.
-Taken together, DdPSO is a promising meta-heuristic approach for the estimation of biochemical model parameters in relatively few function evaluations.
+Thereafter we tested the performance of DOPS on commonly used test functions for global optimization (Ackley and Rosenbrock) and on published biochemical parameter estimation benchmark problems.
+For the wide range of problems that we considered, DOPS outperformed other meta-heuristic approaches despite a limited number of function evaluations.
+Taken together, DOPS is a promising meta-heuristic approach for the estimation of biochemical model parameters in relatively few function evaluations.
 
@@
 Mathematical modeling of biochemical networks is often an iterative process.
-First, models are formulated from biochemical knowledge, and then trained using using experimental data \cite{Aldridge:2006aa,banga2008optimization,ashyraliyev2009systems}.
-Model training is typically framed as a non-linear optimization problem wherein the residual (or objective function) between experimental data and model simulations is minimized using an optimization strategy \cite{moles2003parameter}.
+First, models are formulated from biochemical knowledge, and then model parameters are estimated using experimental data \cite{Aldridge:2006aa,banga2008optimization,ashyraliyev2009systems}.
+Parameter estimation is typically framed as a non-linear optimization problem wherein the residual (or objective function) between experimental data and model simulations is minimized using an optimization strategy \cite{moles2003parameter}.
 The optimal parameters obtained from training are then used to validate the model on hitherto unseen experimental data.
@@
 most of them have multiple local minima or maxima \cite{moles2003parameter,banga2008optimization}.
-The non-linearity of the problem coupled with multi-modality generally renders the local optimization techniques like pattern search method \cite{hooke1961direct}, Nelder-Mead simplex method \cite{nelder1965simplex} or Levenberg-Marquardt method\cite{more1978levenberg} incapable of obtaining an optimal solution.
-Though deterministic global optimization techniques can handle non-linearity and multi-modality \cite{esposito2000deterministic,horst2013global}, the absence of derivative information, discontinuity of the objective functions, non-smooth regions or the lack of any knowledge about the objective function severely hamper the use of these techniques.
+The non-linearity of the problem coupled with multi-modality generally renders the local optimization techniques like pattern search method \cite{hooke1961direct}, Nelder-Mead simplex method \cite{nelder1965simplex}, steepest descent method or Levenberg-Marquardt method \cite{more1978levenberg} incapable of reliably obtaining an optimal solution since they generally stop at the first local minimum.
+Though deterministic global optimization techniques (for example algorithms based on branch and bound framework) can handle non-linearity and multi-modality \cite{esposito2000deterministic,horst2013global}, the absence of derivative information, discontinuity of the objective functions, non-smooth regions or the lack of any knowledge about the objective function severely hamper the use of these techniques.
 
-Meta-heuristic approaches like Genetic Algorithms, Simulated Annealing \cite{kirkpatrick1983optimization}, Evolutionary Programming and population based searches like Differential Evolution\cite{storn1997differential} have shown promise in this regard \cite{sun2012parameter}.
+Meta-heuristic stochastic optimization approaches like Genetic Algorithms (GAs), Simulated Annealing (SA) \cite{kirkpatrick1983optimization}, Evolutionary Programming and population based searches like Differential Evolution (DE) \cite{storn1997differential} have shown promise in this regard \cite{sun2012parameter}.
 They do not make any assumptions about the structure of objective function or require any \textit{a priori} information about the objective function.
@@
 Banga and co-workers have applied scatter-search based methods \cite{villaverde2012cooperative,rodriguez2006novel,egea2007scatter} to estimate parameters on non-linear biological processes.
-Hybrid approaches that combine a meta-heuristic with a local optimization search, wherein a near optimal solution that is obtained with a meta-heuristic is further refined using a local search have also become popular.
+Hybrid approaches that combine a meta-heuristic with a local optimization search, wherein a near globally optimal solution that is obtained with a meta-heuristic is further refined using a local search have also become popular.
 Villaverde et al.
@@
 Fan et al.
-recently showed that population based meta-heuristics along with decomposition based methods can be used to model gene circuits from mRNA data \cite{fan2015parameter}.Despite these successes, a  major drawback with most of these approaches is the vast number of objective function evaluations they take in arriving at good search regions.
-As the models grow in size and complexity and the number of dimensions of parameter vector increases,  parameter estimation becomes a computationally expensive task.
+recently showed that population based meta-heuristics along with decomposition based methods can be used to model gene circuits from mRNA data \cite{fan2015parameter}.
+Despite these successes, a  major drawback with most of these metaheuristic approaches is the vast number of objective function evaluations they take in arriving at good search regions.
+As the models grow in size and complexity and the number of dimensions of parameter vector increases, evaluation of the objective function becomes computationally expensive.
+Thus performing a large number of evaluations is not computationally feasible.
+
 In many of these high dimensional problems approaching an exact solution may not be necessary.
@@
 
-In this study, we developed dynamically dimensioned particle swarm optimization (DdPSO), a novel meta-heurstic that combined the global search capability of multi-swarm particle swarm optimization with dynamic dimensioned search.
-The objective of DdPSO was to obtain near optimal parameter estimates for large biochemical models within a relatively few function evaluations.
-DdPSO uses a multi-swarm particle swarm optimization technique to generate candidate solution vectors which are then greedily updated using dynamic dimensioned search.
-We tested the performance of DdPSO on commonly used test functions for global optimization (Ackley and Rosenbrock), published biochemical parameter estimation benchmark problems and
-lastly on a model of the human coagulation cascade.
-DdPSO outperformed common meta-heuristic approaches like Differential Evolution (DE), Particle Swarm Optimization (PSO), Simulated Annealing (SA) and dynamically dimensioned search (DDS) on the benchmark and coagulation problems.
+In this study, we developed Dynamic Optimization with Particle Swarms (DOPS), a novel meta-heuristic that combines the global search capability of multi-swarm particle swarm optimization and dynamically dimensioned search (DDS).
+The objective of DOPS is to obtain near optimal parameter estimates for large biochemical models within a relatively few function evaluations.
+DOPS uses a multi-swarm particle swarm optimization technique to generate candidate solution vectors which are then greedily updated using dynamically dimensioned search.
+We first considered a model of human  coagulation cascade to test the performance of DOPS.
+Coagulation is a large, complex biochemical network that is tightly regulated with several feedback loops.
+We then tested the performance of DOPS on commonly used test functions for global optimization (Ackley and Rosenbrock), published biochemical parameter estimation benchmark problems \cite{villaverde2015biopredyn}.
+DOPS outperformed common meta-heuristic approaches like Differential Evolution (DE), Simulated Annealing (SA) and dynamically dimensioned search (DDS) on the test functions and the coagulation model.
+It also performed very well on the benchmark problems where it outperformed enhanced scatter search (eSS) and recovered the nominal parameters with only 4000 function evaluations across all the benchmark problems considered.
 
@@
 
-\clearpage
+\section*{Problem Formulation}
+
+The problem of parameter estimation in dynamic biological systems consists of finding an optimal parameter vector that results in the best fit between the model predictions and observed experimental data.The cost function or objective that measures the goodness of fit is generally the Euclidean norm which is minimized under certain constraints.
+This can be mathematically expressed as follows.
+
+\begin{equation}\label{eq:static_opt}
+\begin{aligned}
+& \text{minimize}
+& & K(\mathbf{p})=(\sum_{i=1}^N (g_i(t_i,\mathbf{x,p,u})-y_i)^2) \\
+& \text{subject to}
+&&\dot{x}=f(t,\mathbf{x(t,\mathbf{p}),p,u(\mathbf{t})})\\
+&&&x(t_0) = x_0\\
+&&&c(t,\mathbf{x,p,u}) \geqslant 0 \\
+&&& \mathbf{p}^L \leqslant \mathbf{p} \leqslant \mathbf{p}^U\\
+\end{aligned}
+\end{equation}
+
+where t is time, x is the state variable with an initial state x\textsubscript{0}, u is the input vector \, p is the parameter vector and f is the system of equations (like differential equations or set of algebraic constraints), c in the set of linear and non-linear constraints, g is the set of observables and p\textsuperscript{L} and p\textsuperscript{U} are the lower and upper bounds.
+Thus the problem eventually is to find {p}\textsuperscript{*} where
+
+\begin{equation}\label{eq:static_opt_2}
+\mathbf{p^{*}}= \arg\min_{p} K(\mathbf{p})
+\end{equation}
 
 \section*{Results}
-\subsection*{Dynamically dimensioned particle swarm optimization (DdPSO).}
-DdPSO is a novel meta-heuristic that combines multi-swarm based particle swarm methods with dynamically dimensioned search (Fig. \ref{fig-algorithm}).
+\subsection*{Dynamic Optimization with Particle Swarms (DOPS).}
+DOPS is a novel meta-heuristic that combines multi-swarm based particle swarm methods with dynamically dimensioned search (Fig.\ref{fig-algorithm}).
 The goal of this approach is to obtain optimal or near optimal parameters for high-dimensional complex biological systems within a pre-specified number of function evaluations.
@@
 \end{eqnarray}
-where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes the best local solution found by particle $i$ till function evaluation $j$-$1$, and
-$\mathcal{GL}_{k}$ denotes the best local solution found over the population of all particles within the swarm ${S}_{k}$.
+where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes the best solution found by particle $i$ within the sub-swarm till function evaluation $j$-$1$, and
+$\mathcal{GL}_{k}$ denotes the best solution found over the population of all particles within the sub-swarm ${S}_{k}$.
 The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
-parameters ($\mathcal{K}\times{1}$). In our algorithm the parameter $\theta_{1,j-1}$ depends on the function evaluations and is controlled according to the following equation
+parameters ($\mathcal{K}\times{1}$).
+Equation (3) is suggested by the general form of particle swarm equations.
+However this equation does not contain the velocity terms that are generally associated with particle swarm optimization.
+In our algorithm the parameter $\theta_{1,j-1}$ depends on the function evaluations and is controlled according to the following equation
 \begin{eqnarray}
@@
 where $\mathbf{N}$ represents the total number of function evaluations, $\mathbf{w}_{max}$ and $\mathbf{w}_{min}$ are the maximum and minimum inertia weights respectively.
-While updating the particle, we made sure all dimensions of the solution represented by the particle were within bounds using a reflection boundary condition.
+While updating the particle, we made sure all dimensions of the solution represented by the particle were within bounds using a set of reflection boundary conditions shown below.
+
+\begin{eqnarray}
+\end{eqnarray}
+
+\begin{algorithm}
+\If{$x_{i,j}^{old} < x_{i}^{min}$}{  	
+  ${x}_{i,j}^{new}={x}_{i,j}^{old}+(x_{i}^{min} - {x}_{i,j}^{old})$
+  \If{${x}_{i,j}^{new}>x_{i}^{max}$}{
+    ${x}_{i,j}^{new}=x_{i}^{max}$
+  }
+}
+\If{$x_{i,j}^{old} > x_{i}^{max}$}{  	
+  ${x}_{i,j}^{new}={x}_{i,j}^{old}+({x}_{i,j}^{old}-x_{i}^{max})$
+  \If{${x}_{i,j}^{new}<x_{i}^{min}$}{
+    ${x}_{i,j}^{new}=x_{i}^{min}$
+  }
+}
+\end{algorithm}
+
 After every $\mathbf{g}$  function evaluations, the particles within all sub-swarms were mixed and then randomly redistributed to a new sub-swarm.
@@
 \begin{eqnarray}
-	\mathcal{P}_{j}&=&{1}-\log(j/ (\beta({1}-\mathcal{FR})*\mathbf{N}))
+	%\mathcal{P}_{j}&=&{1}-\log(j/ (\beta({1}-\mathcal{FR})*\mathbf{N}))
+	\mathcal{P}_{j}&=&{1}-\log(j/ (\({1}-\mathcal{FR})*\mathbf{N}))
 \end{eqnarray}
-where $\beta$ is the perturbation frequency probability modulator.
+%where $\beta$ is the perturbation frequency probability modulator. 
 Thus the number of dimensions of the candidate vector that are updated or perturbed decreases with the as the number of function evaluations increase.
 These updates are greedy in nature, so $\mathcal{G}_{new}$ becomes the new solution vector only if it is better than the old one $\mathcal{G}$.
+The decrease of weight function in equation 4, reflection boundary conditions in equation 5, the update function described in equations 6 and 7, the selection probability in equation 8 are the means by which DDS ideas are incorporated into DOPS.
+%Equation 5 describing the reflection boundary conditions, equations 6,7 and 8 are the means by which DDS ideas are incorporated into DOPS. 
+The fraction of evaluations $\mathcal{FR}$  within the swarm phase and DDS phase is based on a switching strategy wherein the switch from swarm phase to DDS phase happens when the error due to the best solution does not drop more than 1\% of the original error, continuously for more than a prescribed number of function evaluations.
+This allows the solution to quickly jump out of local optima and avoid any convergence issues that are generally associated with swarm based searches.
 
-\subsection*{Performance of DdPSO on benchmark problems.}
-We compared the performance of DdPSO with algorithms on commonly used test functions for global optimization, as well as benchmark biochemical problems from literature (Table \ref{table:benchmark-problems}).
-We used a 300 dimensional Rastrigin function and 300 dimensional Ackley function.
-Both Ackley and Rastrigin have multiple local minima and maxima and attain a global minimum value of 0.
-We tested the performance of the 5 different heuristic approaches on these two functions.
-In each experiment we again fixed the number of function evaluations at 4000 and ran 25 experiments.
-In both cases (Figure S6 A and B) we see that the error convergence rate for DDSMLSPSO is much faster as compared to the other four s and it finds the global minimum of 0 in both cases.
-
-Villaverde and co-workers recently published a set of benchmark problems to evaluate parameter estimation methods \cite{villaverde2015biopredyn}.
-From a computational cost perspective problems they categorized the problems as most expensive, intermediate and least expensive.
-We evaluated the performance of our algorithms on a problem from the most expensive and least expensive categories.
-The first problem is a genome wide kinetic model of \textit{Saccharomyces cerevisiae} with 1759 parameters (henceforth referred to as problem B1).
-The second problem (henceforth referred to as problem B4) is a metabolic model of Chinese Hamster Ovary (CHO) with 35 metabolites, 32 reactions and 117 parameters.
-In both cases pseudo time series data was generated.
-For problem B1, the time series data consisted of 44 observables and for problem B4 the data corresponded to 13 different metabolites.
-We fixed the number of function evaluations at 4000 for DDSMLSPSO and trained both the models against the pseudo experimental data.
-We recaptured the 'nominal' parameters (Figure S7 A and B)  in both cases within 4000 evaluations.
-The final objective function value (Table 1) is an order smaller for B1 possibly due to overfitting.
-In Table 1 we summarize the results obtained on the coagulation model and the benchmark problems.
-
-\subsection*{Performance of DdPSO on a model of the human coagulation cascade.}
-We compared the performance of DdPSO on a model of blood coagulation against commonly used meta-heuristics like simulated annealing (SA), differential evolution (DE), particle swarm optimization (PSO),and dynamically dimensioned search (DDS).
+\subsection*{Performance of DOPS on a model of the human coagulation cascade.}
+We compared the performance of DOPS on a model of blood coagulation against commonly used meta-heuristics like simulated annealing (SA), differential evolution (DE), particle swarm optimization (PSO),and dynamically dimensioned search (DDS).
 Coagulation is an archetype biochemical network that is highly interconnected and tightly regulated with multiple positive and negative feedback loops (Fig.
@@
 
+\subsection*{Performance of DOPS on benchmark problems and test functions.}
+Villaverde and co-workers recently published a set of benchmark biochemical problems to evaluate parameter estimation methods \cite{villaverde2015biopredyn}.
+From a computational cost perspective problems they categorized the problems as most expensive, intermediate and least expensive.
+We evaluated the performance of our algorithms on a problem from the most expensive and least expensive categories.
+The first problem is a genome wide kinetic model of \textit{Saccharomyces cerevisiae} with 1759 parameters (henceforth referred to as problem B1).
+The second problem (henceforth referred to as problem B4) is a metabolic model of Chinese Hamster Ovary (CHO) with 35 metabolites, 32 reactions and 117 parameters.
+In both cases pseudo time series data was generated.
+For problem B1, the time series data consisted of 44 observables and for problem B4 the data corresponded to 13 different metabolites.
+We fixed the number of function evaluations at 4000 for DOPS and trained both the models against the pseudo experimental data.
+We recaptured the 'nominal' parameters (Figure S7 A and B)  in both cases within 4000 evaluations.
+The final objective function value (Table 1) is an order smaller for B1 possibly due to overfitting.
+In Table 1 we summarize the results obtained on the coagulation model and the benchmark problems.
+
+We compared the performance of DdPSO with algorithms on commonly used test functions for global optimization, as well as benchmark biochemical problems from literature (Table \ref{table:benchmark-problems}).
+We used a 300 dimensional Rastrigin function and 300 dimensional Ackley function.
+Both Ackley and Rastrigin have multiple local minima and maxima and attain a global minimum value of 0.
+We tested the performance of the 5 different heuristic approaches on these two functions.
+In each experiment we again fixed the number of function evaluations at 4000 and ran 25 experiments.
+In both cases (Figure S6 A and B) we see that the error convergence rate for DDSMLSPSO is much faster as compared to the other four s and it finds the global minimum of 0 in both cases.
+
 \clearpage
@@
  \hline
-                        & Experimental Data Set & B1           & B4          & Ackley 300D & Rastrigin 300D \\
+                        & Experimental Data Set & B1       &B2    & B4   &B5       & Ackley 300D & Rastrigin 300D \\
  \hline
-Lower Bound             & 0.001.pnom            & 5.pnom       & 5.pnom      & 30          & 5.12           \\
-Upper Bound             & 1000.pnom             & 0.2.pnom     & 0.2.pnom    & -15         & -5.12          \\
-CPU Time                & 10.0835 hours         & 38.308 hours & 6.2 minutes & 2.8 seconds & 2.6 seconds    \\
-Function Evaluations    & 4000                  & 4000         & 4000        & 4000        & 4000           \\
-Initial Objective Value & 1.11837e+07           & 1.4275e+07   & 1.8536e+07  & 21.12       & 99.985         \\
-Final Objective Value   & 90456.9               & 4.96e+05     & 38.9375     & 0           & 0              \\
-Nominal Objective Value & 4.7785e+06            & 1.0986e+06   & 39.0676     & 0           & 0      \\
+Lower Bound             & 0.001.pnom            & 5.pnom    &10.pnom   & 5.pnom  &varying    & 30          & 5.12           \\
+Upper Bound             & 1000.pnom             & 0.2.pnom    &0.1.pnom & 0.2.pnom  &varying  & -15         & -5.12          \\
+CPU Time                & 10.0835 hours         & 65.26 hours &17.9 minutes & 6.2 minutes &32.686 minutes& 2.8 seconds & 2.6 seconds    \\
+Function Evaluations    & 4000                  & 4000              &4000& 4000        & 4000      &4000  & 4000           \\
+Initial Objective Value & 1.11837e+07           & 1.3824e+07 &3.8848e+03  & 1.8536e+07 &1.4288e+04  & 21.12       & 99.985         \\
+Final Objective Value   & 90456.9               & 3.5348e+04     &324.4182 & 38.9375    &3.4495e+03  & 0           & 0              \\
+Nominal Objective Value & 4.7785e+06            & 1.0986e+06   &- & 39.0676    &4.2737e+03 & 0           & 0      \\
 \hline

### 2016-01-27 Paper_v4.tex
Figs Updated

@@
 	%\mathcal{P}_{j}&=&{1}-\log(j/ (\beta({1}-\mathcal{FR})*\mathbf{N}))
-	\mathcal{P}_{j}&=&{1}-\log(j/ (\({1}-\mathcal{FR})*\mathbf{N}))
+	\mathcal{P}_{j}&=&{1}-\log(j/ (({1}-\mathcal{FR})*\mathbf{N}))
 \end{eqnarray}
@@
 \subsection*{Performance of DOPS on a model of the human coagulation cascade.}
-We compared the performance of DOPS on a model of blood coagulation against commonly used meta-heuristics like simulated annealing (SA), differential evolution (DE), particle swarm optimization (PSO),and dynamically dimensioned search (DDS).
+We compared the performance of DOPS on a model of blood coagulation against commonly used meta-heuristics like simulated annealing (SA), differential evolution (DE), and dynamically dimensioned search (DDS).
 Coagulation is an archetype biochemical network that is highly interconnected and tightly regulated with multiple positive and negative feedback loops (Fig.
@@
 Coagulation is regulated by a set of serine proteases also known as coagulation factors and blood platelets.
+
+%Added a little more description here
 The coagulation factors are generally in an inactive state and are known as zymogens.
@@
 The exposure of these factors to blood kick-starts a series of convergent cascades that lead to conversion of zymogen prothrombinase to thrombin.
+For example when coagulation is initiated through tissue factor pathway, tissue factor and activated factor VIIa (FVIIa) form a complex that activates factors X (fX) and IX (fIX).
+Activated factor X (fXa) thereafter activates downstream factors VIII and IX.
+The initial activation leads to the production of picomolar amounts of thrombin (fIIa) which activates platelets and amplifies its own production through the formation of a prothrombinase complex (FXa-FVa) on the surface of the activated platelets.
+Thrombin also  downregulates its own production by forming a complex with thrombomodulin which then activates protein C (PC).
+PC inhibits the formation of prothrombinase complex.
+In addition, Tissue factor pathway inhibitor (TFPI) downregulates FXa formation and activity by sequestering free FXa and TF-FVIIa in a FXa-dependent manner.
+Antithrombin III (ATIII)  inhibits all proteases.
+surface protein thrombomodulin (TM).
 Luan et al.
@@
 The objective function was a weighted linear combination of two different error functions that used data sets representing coagulation initiated with different concentrations of TF-VIIa (5pM, 5nM) \cite{hockin2002model}.
+This choice of using a linear combination of two different error functions was motivated by poor validation results while using a single error function.
 We restricted the number of function evaluations to 4000 for each algorithm and performed 25 trials of this experiment.
-DdPSO exhibits a much faster rate of error convergence and has a much lower final error as compared to the other algorithms (Figure 3).
-Within the first 1000 function evaluations (swarm phase) of DDSMLSPSO there is a very rapid drop in error.
-Subsequently, again after 2000 function evaluations (dynamically dimensioned phase) the error drops quickly.
-Overall at the end of 4000 function evaluations DDSMLSPSO minimizes the error (final objective error is 90456.9) to a much greater extent than any of the other algorithms.
-Amongst the rest of the algorithms DDS has a comparable error drop for the first 2000 function evaluations.
-However the drop in error remains constant for DDS after 2000 function evaluations.
-Using the parameters obtained at the end of 4000 function evaluations we examined the 'fits' between models predictions and experimental data (Figure 2).
+DOPS exhibits a much faster rate of error convergence and has a much lower final error as compared to the other algorithms (Figure 3).
+
+ Within the first 1000 function evaluations  of DOPS there is a very rapid drop in error.
+Approximately between 500 -1000 function evaluations a switch to dynamically dimensioned search phase happens (this switch varies from trial to trial since the switch is based on the error from the swarm phase).
+Overall at the end of 4000 function evaluations DOPS minimizes the error (final objective error is 90456.9) to a much greater extent than any of the other algorithms.
+Using the parameters obtained at the end of 4000 function evaluations we examined the 'fits' between models predictions and experimental data (Figure 4).
 The solid lines represent the mean value of prediction over 25 trials and the shaded region represents the 99\% confidence interval.
-Subsequently we compared the model predictions against completely 'unseen' or untrained experimental where coagulation was initiated with 500pM,50pM,10pM concentrations of TF-VIIa respectively (Figure 3).
+Subsequently we compared the model predictions against completely 'unseen' or untrained experimental where coagulation was initiated with 500pM,50pM,10pM concentrations of TF-VIIa respectively (Figure 5).
 

### 2016-01-28 Paper_v4.tex
Changes to text

@@
  Within the first 1000 function evaluations  of DOPS there is a very rapid drop in error.
-Approximately between 500 -1000 function evaluations a switch to dynamically dimensioned search phase happens (this switch varies from trial to trial since the switch is based on the error from the swarm phase).
-Overall at the end of 4000 function evaluations DOPS minimizes the error (final objective error is 90456.9) to a much greater extent than any of the other algorithms.
+Approximately between 500-1000 function evaluations a switch to dynamically dimensioned search phase happens (this switch varies from trial to trial since the switch is based on the error from the swarm phase).
+Overall at the end of 4000 function evaluations DOPS minimizes the error (final objective error is 5.5916e+04) to a much greater extent than any of the other algorithms.
 Using the parameters obtained at the end of 4000 function evaluations we examined the 'fits' between models predictions and experimental data (Figure 4).
@@
 We evaluated the performance of our algorithms on a problem from the most expensive and least expensive categories.
-The first problem is a genome wide kinetic model of \textit{Saccharomyces cerevisiae} with 1759 parameters (henceforth referred to as problem B1).
-The second problem (henceforth referred to as problem B4) is a metabolic model of Chinese Hamster Ovary (CHO) with 35 metabolites, 32 reactions and 117 parameters.
-In both cases pseudo time series data was generated.
+The first problem (most expensive) is a genome wide kinetic model of \textit{Saccharomyces cerevisiae} with 261 reactions, 262 variables and 1759 parameters (henceforth referred to as problem B1).
+The reactions were modeled using modular rate law and generalized form of Michaelis-Menten kinetics.
+The second problem (henceforth referred to as problem B4, least expensive) is a metabolic model of Chinese Hamster Ovary (CHO) with 35 metabolites, 32 reactions and 117 parameters.
+In both cases pseudo time series data was generated by the authors.
 For problem B1, the time series data consisted of 44 observables and for problem B4 the data corresponded to 13 different metabolites.
+For a detailed description about the model architectures please refer to the work of Villaverde et al.
+\cite{villaverde2015biopredyn}
+
 We fixed the number of function evaluations at 4000 for DOPS and trained both the models against the pseudo experimental data.
-We recaptured the 'nominal' parameters (Figure S7 A and B)  in both cases within 4000 evaluations.
-The final objective function value (Table 1) is an order smaller for B1 possibly due to overfitting.
-In Table 1 we summarize the results obtained on the coagulation model and the benchmark problems.
-
-We compared the performance of DdPSO with algorithms on commonly used test functions for global optimization, as well as benchmark biochemical problems from literature (Table \ref{table:benchmark-problems}).
+In both cases we found good fits  (Figures 6 and Figures 7) to the problems within 4000 evaluations.
+We recaptured the 'nominal' parameters' in both cases within 4000 evaluations (Figure 8).
+The final objective function value (Table 1) is an order smaller than nominal error for B1 possibly due to overfitting against the noise.
+Thereafter we compared the performance of DOPS with algorithms on commonly used test functions for global optimization.
 We used a 300 dimensional Rastrigin function and 300 dimensional Ackley function.
 Both Ackley and Rastrigin have multiple local minima and maxima and attain a global minimum value of 0.
-We tested the performance of the 5 different heuristic approaches on these two functions.
+We tested the performance of the 4 different heuristic approaches on these two functions.
 In each experiment we again fixed the number of function evaluations at 4000 and ran 25 experiments.
-In both cases (Figure S6 A and B) we see that the error convergence rate for DDSMLSPSO is much faster as compared to the other four s and it finds the global minimum of 0 in both cases.
+In both cases (Figure 9) we see that the error convergence rate for DOPS is much faster as compared to the other three heuristics and it finds the global minimum of 0 in both cases.
+In Table 1 (Table \ref{table:benchmark-problems}) we summarize the results obtained on the coagulation model, the benchmark problems and the test functions for the best trial.
 
@@
 For SA, we optimized the cooling schedule for the coagulation model.
-Thus, in this approach any overhead that usually comes with additional function evaluations to optimize parameters was avoided.
-The only parameter that we adjusted was the fraction of evaluations within swarm search.
-For the coagulation model it was set to be 0.4 and for the benchmarks it is at 0.6.
-This parameter was set based on empirical observations rather than any definitive analysis.
-When the error convergence rates stagnated in the swarm phase we switched to DDS phase.
-A further improvement to our approach could  be to incorporate a meta intelligent way of switching in and switching out of both the phases.
+Thus, in this approach any overhead that usually comes with additional function evaluations to pre optimize parameters was avoided.
 
@@
 
-\begin{table}[h]
-\centering
-\caption{Table with optimization settings and results for coagulation and different benchmarks using DdPSO.}
-\label{table:benchmark-problems}
-\resizebox{\columnwidth}{!}{
-\begin{tabular}{|cccccc|}
-\hline
-                        & Coagulation           & \multicolumn{4}{l}{Benchmarks}\\
- \hline
-                        & Experimental Data Set & B1       &B2    & B4   &B5       & Ackley 300D & Rastrigin 300D \\
- \hline
-Lower Bound             & 0.001.pnom            & 5.pnom    &10.pnom   & 5.pnom  &varying    & 30          & 5.12           \\
-Upper Bound             & 1000.pnom             & 0.2.pnom    &0.1.pnom & 0.2.pnom  &varying  & -15         & -5.12          \\
-CPU Time                & 10.0835 hours         & 65.26 hours &17.9 minutes & 6.2 minutes &32.686 minutes& 2.8 seconds & 2.6 seconds    \\
-Function Evaluations    & 4000                  & 4000              &4000& 4000        & 4000      &4000  & 4000           \\
-Initial Objective Value & 1.11837e+07           & 1.3824e+07 &3.8848e+03  & 1.8536e+07 &1.4288e+04  & 21.12       & 99.985         \\
-Final Objective Value   & 90456.9               & 3.5348e+04     &324.4182 & 38.9375    &3.4495e+03  & 0           & 0              \\
-Nominal Objective Value & 4.7785e+06            & 1.0986e+06   &- & 39.0676    &4.2737e+03 & 0           & 0      \\
-\hline
-\end{tabular}
-}
-\end{table}
+%\begin{table}[h]
+%\centering
+%\caption{Table with optimization settings and results for coagulation and different benchmarks using DdPSO.}
+%\label{table:benchmark-problems}
+%\resizebox{\columnwidth}{!}{
+%\begin{tabular}{|cccccc|}
+%\hline
+%                        & Coagulation           & \multicolumn{4}{l}{Benchmarks}\\
+% \hline
+%                        & Experimental Data Set & B1       &B2    & B4   &B5       & Ackley 300D & Rastrigin 300D \\
+% \hline
+%Lower Bound             & 0.001.pnom            & 5.pnom    &10.pnom   & 5.pnom  &varying    & 30          & 5.12           \\
+%Upper Bound             & 1000.pnom             & 0.2.pnom    &0.1.pnom & 0.2.pnom  &varying  & -15         & -5.12          \\
+%CPU Time                & 10.0835 hours         & 65.26 hours &17.9 minutes & 6.2 minutes &32.686 minutes& 2.8 seconds & 2.6 seconds    \\
+%Function Evaluations    & 4000                  & 4000              &4000& 4000        & 4000      &4000  & 4000           \\
+%Initial Objective Value & 1.11837e+07           & 1.3824e+07 &3.8848e+03  & 1.8536e+07 &1.4288e+04  & 21.12       & 99.985         \\
+%Final Objective Value   & 90456.9               & 3.5348e+04     &324.4182 & 38.9375    &3.4495e+03  & 0           & 0              \\
+%Nominal Objective Value & 4.7785e+06            & 1.0986e+06   &- & 39.0676    &4.2737e+03 & 0           & 0      \\
+%\hline
+%\end{tabular}
+%}
+%\end{table}
 
